### 2020-01-29 notebook/bad-forecast-good-decision.Rmd
branch off from forecasting-limits

---
title: "bad forecast good decision"
author: "Carl Boettiger"
output: github_document
---

```{r message=FALSE, warning=FALSE}
library(ggthemes)
library(hrbrthemes)
library(ggplot2)
library(Cairo)
library(extrafont)
extrafont::loadfonts()
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
#theme_set(theme_solarized(base_size=16))

scale_colour_discrete <- function(...) scale_colour_solarized()
scale_fill_discrete <- function(...) scale_fill_solarized()
pal <- solarized_pal()(6)
txtcolor <- "#586e75"
```

```{r}
library(tidyverse)
library(MDPtoolbox)
library(expm)
```

```{r}
states <- seq(0,20, length.out = 200)
actions <- states
obs <- states
sigma_g <- 0.04
reward_fn <- function(x,h) pmin(x,h)
discount <- 0.99

# K is at twice max of f3; 8 * K_3 / 5
f1 <- function(x, h = 0, r = 1, K = 10 * 8 / 5){
  s <- pmax(x - h, 0)
  s + s * (r * (1 - s / K) )
}
f2 <- function(x, h = 0, r = 1, K = 10){
  s <- pmax(x - h, 0)
  s + s * (r * (1 - s / K) )
}

# max is at 4 * K / 5 
f3  <- function(x, h = 0, r = .002, K = 10){
  s <- pmax(x - h, 0)
  s + s ^ 4 * r * (1 - s / K)
}

f4  <- function(x, h = 0, r = 4e-7, K = 10){
  s <- pmax(x - h, 0)
  s + s ^ 8 * r * (1 - s / K) + s * (1 - s / K)
}

models <- list(f1 = f1, f2 = f2, f3 = f3)

true_model <- "f3"

## Could consider ricker-style versions of these models. need to take care to set limits correctly
## to get correspondence (recall x * e^y ~ x *(1 + y) for small y), e.g. something like
#f1 <- function(x, h = 0, r = .1, K = 15){
#  s <- pmax(x - h, 0)
#  s * exp(r * (1 - s / K) )
#}
#f3  <- function(x, h = 0, r = .01, K = 10){
#  s <- pmax(x - h, 0)
#  s * exp(s ^ 3 * r * (1 - s / K))
#}

```

```{r}
d <- map_dfc(models, function(f) f(states) - states) %>% mutate(state = states)
d %>% pivot_longer(names(models), "model") %>%
  ggplot(aes(state, value, col=model)) +
  geom_point() + 
  geom_hline(aes(yintercept = 0)) + 
  coord_cartesian(ylim = c(-5, 5), xlim = c(0,16))
```

Comparing forecasts

```{r}

```

```{r}
# A function to compute the transition matrices for each model:
transition_matrices <- function(f,
                      states,
                      actions,
                      sigma_g){

  n_s <- length(states)
  n_a <- length(actions)

  transition <- array(0, dim = c(n_s, n_s, n_a))
  for (k in 1:n_s) {
    for (i in 1:n_a) {
      nextpop <- f(states[k], actions[i])
      if(nextpop <= 0){
        transition[k, , i] <- c(1, rep(0, n_s - 1))
      } else if(sigma_g > 0){
        x <- dlnorm(states, log(nextpop), sdlog = sigma_g)
        if(sum(x) == 0){ ## nextpop is computationally zero
          transition[k, , i] <- c(1, rep(0, n_s - 1))
        } else {
          x <- x / sum(x) # normalize evenly. 
          ## pile excess on boundary
          #N <- plnorm(states[n_s], log(nextpop), sigma_g)
          #x <- x * N / sum(x)
          #x[n_s] <- 1 - N + x[n_s]
          transition[k, , i] <- x
        }
      } else {
        stop("sigma_g not > 0")
      }
      reward[k, i] <- reward_fn(states[k], actions[i])
    }
  }
  transition
}

## Reward matrix is shared by each model
n_s <- length(states)
n_a <- length(actions)
reward <- array(0, dim = c(n_s, n_a))
for (k in 1:n_s) {
  for (i in 1:n_a) {
    reward[k, i] <- reward_fn(states[k], actions[i])
  }
}

transitions <- lapply(models, function(f) transition_matrices(f, states, actions, sigma_g))

```

## Forecasting performance

```{r}

sim <- function(f, x0, Tmax, reps = 1){
  map_dfr(1:reps, 
          function(i){
            x <- numeric(length(Tmax))
            x[1] <- x0
            for(t in 2:Tmax){
              mu <- f(x[t-1])
              x[t] <-  rlnorm(1, log(mu), sdlog = sigma_g)
            }
            tibble(t = 1:Tmax, x= x)
          },
          .id="rep")
}
models[names(models) != true_model]

true_sim <- sim(models[[true_model]], 5, 100, 1) %>%
  mutate(model = true_model, mean = x, ymin = NA, ymax = NA)
df <- map_dfr(models[1:2], sim, 5, 100, 100, .id = "model")
forecast <- df %>% 
  group_by(model, t) %>% summarise(mean = mean(x), sd = sd(x))

forecast %>%
  ggplot(aes(t, mean, col=model)) +
  geom_line() + 
  geom_line(data = true_sim) + 
  geom_ribbon(aes(fill= model, ymin = mean - 2*sd, ymax = mean + 2*sd), alpha=0.1, col=NA)

```

```{r}
d <- forecast %>% group_by(model) %>% 
  mutate(true = true_sim$x) %>%
  summarise(rmsd = sqrt(mean( (mean - true) ^ 2 )),
            r2 = 1 - sum( (mean - true) ^ 2  ) / sum( mean ^ 2 )
            )
rmsd <- d %>% pull(rmsd)
names(rmsd) <- d %>% pull(model)

opt <- sqrt(mean((true_sim$x - 10)^2))
```

Model 2 clearly outperforms model 1 in predictive capacity. The root mean squared deviation between the model 2 mean and the realization is `r rmsd[["f2"]]`, far lower than model 1 at `r rmsd[["f1"]]`.

```{r}
# Simulate the evolution of probability distribution directly using a matrix exponent
prob_dynamics <- function(M, X, Tmax){
  probability <- t(M) %^% Tmax %*% X 
  data.frame(state = states, probability)
}

x0 <- which.min(abs(states - 6))
X <- numeric(length=length(states))
X[x0] <- 1

map_dfr(transitions, 
        function(m) prob_dynamics(m[,,1], X, 100),
        .id = "model") %>%
  group_by(model) %>%
  ggplot(aes(state, probability, col=model)) +
  geom_line()
```

## Model analysis

```{r}
d <- map_dfc(models, function(f) f(states) - states) %>% mutate(state = states)
d %>% pivot_longer(names(models), "model") %>%
  ggplot(aes(state, value, col=model)) +
  geom_point() + 
  geom_hline(aes(yintercept = 0)) + 
  coord_cartesian(ylim = c(-1, 5), xlim = c(0,16))
```

```{r}

policies <- map_dfr(transitions, function(P){
  soln <- mdp_value_iteration(P, reward, discount = discount)
  tibble(states, policy = soln$policy, escapement = states - actions[soln$policy])
}, .id = "model")

```

```{r}
policies %>%
  ggplot(aes(states,escapement, col=model, lty=model)) + geom_line(lwd=2)
```

Management under the wrong model (`f3`)

```{r}
library(mdplearning)
Tmax <- 100
x0 <- which.min(abs(states - 6))
reps <- 5
set.seed(12345)

## Simulate each policy reps times, with `f3` as the true model:

sims <- map_dfr(names(transitions), 
                function(m){
                  policy <- policies %>% filter(model == m) %>% pull(policy)
                  map_dfr(1:reps, 
                          function(i){

                            mdp_planning(transitions[[true_model]], reward, discount,
                                     policy = policy, x0 = x0, Tmax = Tmax) %>%

                              select(value, state, time)  %>% 
                              mutate(state = states[state]) # index->value
                            },
                          .id = "reps")
                },
                .id = "model")

```

```{r}
fig_ts <- 
  sims %>%
  filter(time < 25) %>%
  ggplot(aes(time, state, col=model, group = interaction(model,reps))) + 
  geom_line(alpha=0.5) 
fig_ts
```

```{r fig.cap = "Corresponding utility (measured as mean net present value, that is: cumulative value, discounting future values by $\\delta^t$, averaged across replicates). Note that the exepcted utility under model 1, which has the worst forecast, is nearly identical to the optimal utility achieved by managing under the correct model, 3.  The utility derived from model 2 is far smaller, despite it's overall better performance in long term forecasts."}
##  Net Present Value accumulates over time, equivalent for models with near-identical management stategy
npv_df <- sims %>% 
  group_by(model, reps) %>%
  mutate(npv = cumsum(value * discount ^ time)) %>%
  group_by(time, model)  %>% 
  summarise(mean_npv = mean(npv))
npv_df %>%
  ggplot(aes(time, mean_npv, fill=model)) +
  ## Area plots are weird but avoids overplotting
  geom_area(alpha = 0.9) +
   #geom_ribbon(aes(ymin = 0, ymax = mean_npv), alpha = 0.6) + 
  ylab("Net present value") + xlab("time")

```

## Conclusions

- A good forecast does not mean good management
- A bad forecast does not mean the model is bad for management
- A model permitting successful management does not imply the model is "correct" or generally "good at forecasting"

Model with the egregiously optimistic long-term forecast for the stock size of an unexploited fishery nevertheless actually leads to more conservative management.  In contrast, the model which correctly predicts the long-term average stock size without fishing nevertheless leads to substantial overharvesting.

### 2020-07-02 manuscript/manuscript.Rmd
use large r to highlight poor forecast performance

---
title: "Forecast accuracy can be a poor metric for decision-making"
date: "`r Sys.Date()`"
author:
  - name: Carl Boettiger
    email: cboettig@berkeley.edu
    affiliation: ucb
#    footnote: Corresponding Author
address:
  - code: ucb
    address: "ESPM Department, University of California, 130 Mulford Hall Berkeley, CA 94720-3114, USA"
abstract: |

journal: Theoretical ecology
layout: 3p
bibliography: refs.bib
output: rticles::elsevier_article
keywords: 
  - transients, 
  - optimal control, 
  - adaptive management, 
  - stochasticity, 
  - uncertainty, 
  - ecological management
preamble: |
  \journal{preprint}
  ## EditorialManager adds its own linenumbers to pdf already
  \usepackage{lineno}
  \linenumbers

---

```{r setup, message = FALSE}
library(tidyverse)   # plotting and data manipulation
```

Is forecasting is the ultimate test of theory?

# Acknowledgements

This work was supported in part by NSF CAREER () and computational resources from NSF's XSEDE Jetstream (DEB160003) and Chameleon cloud platforms.

\pagebreak

# References

### 2020-07-02 notebook/bad-forecast-good-decision.Rmd
re-run
use large r to highlight poor forecast performance

@@
 library(expm)
+library(mdplearning) # remotes::install_github("boettiger-lab/mdplearning")
 ```
@@
 # K is at twice max of f3; 8 * K_3 / 5
-f1 <- function(x, h = 0, r = 1, K = 10 * 8 / 5){
+f1 <- function(x, h = 0, r = 2, K = 10 * 8 / 5){
   s <- pmax(x - h, 0)
@@
   geom_hline(aes(yintercept = 0)) + 
-  coord_cartesian(ylim = c(-5, 5), xlim = c(0,16))
-```
-
-Comparing forecasts
-
-```{r}
-
+  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16))
 ```
@@
 opt <- sqrt(mean((true_sim$x - 10)^2))
+opt
 ```
@@
 ## Model analysis
-
-```{r}
-d <- map_dfc(models, function(f) f(states) - states) %>% mutate(state = states)
-d %>% pivot_longer(names(models), "model") %>%
-  ggplot(aes(state, value, col=model)) +
-  geom_point() + 
-  geom_hline(aes(yintercept = 0)) + 
-  coord_cartesian(ylim = c(-1, 5), xlim = c(0,16))
-```
 
@@
                           function(i){
-
                             mdp_planning(transitions[[true_model]], reward, discount,
                                      policy = policy, x0 = x0, Tmax = Tmax) %>%
-
-                              select(value, state, time)  %>% 
-                              mutate(state = states[state]) # index->value
+                              select(value, state_index = state, time, action_index = action)  %>% 
+                              mutate(state = states[state_index]) # index->value
                             },
@@
 fig_ts
+```
+
+Note that the pattern is not driven merely be differences in the unfished equilibrium.  Compare forecast performance under management:
+
+```{r}
+scores <- sims %>%
+  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
+  select(time, state_index, next_state, action_index, model, reps) %>% 
+  rowwise() %>%
+  mutate(prob =  transitions[[model]][state_index, next_state, action_index] )
+
+scores %>% 
+  group_by(reps, model) %>%
+  summarise(score = sum(log(prob))) %>%
+  group_by(model) %>% summarise(mean_score = mean(score))
+
 ```
@@
 In contrast, the model which correctly predicts the long-term average stock size without fishing nevertheless leads to substantial overharvesting.
+
+```{r}
+```

### 2020-07-03 notebook/bad-forecast-good-decision.Rmd
more reps, plot distribution

@@
 x0 <- which.min(abs(states - 6))
-reps <- 5
+reps <- 100
 set.seed(12345)
@@
   sims %>%
-  filter(time < 25) %>%
+  filter(time < 25, reps < 5) %>%
   ggplot(aes(time, state, col=model, group = interaction(model,reps))) + 
@@
 
-Note that the pattern is not driven merely be differences in the unfished equilibrium.  Compare forecast performance under management:
+Note that the pattern is not driven merely be differences in the unfished equilibrium.
+Compare forecast performance under management.
+Score is the sum of the log probabilities of observed outcomes, given the
 
@@
   mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
+  filter(model < 3) %>%
   select(time, state_index, next_state, action_index, model, reps) %>% 
   rowwise() %>%
-  mutate(prob =  transitions[[model]][state_index, next_state, action_index] )
+  mutate(prob =  transitions[[model]][state_index, next_state, action_index] ) %>% 
+  group_by(reps, model) %>%
+  summarise(score = sum(log(prob)))
 
-scores %>% 
-  group_by(reps, model) %>%
-  summarise(score = sum(log(prob))) %>%
-  group_by(model) %>% summarise(mean_score = mean(score))
+scores %>% group_by(model) %>% summarise(mean_score = mean(score))
+
+scores %>% mutate(model = as.factor(model)) %>% 
+  ggplot(aes(x = score, group = model, fill = model)) +
+  geom_histogram(binwidth = 70)
 

### 2020-07-08 notebook/bad-forecast-good-decision.Rmd
exploring more

@@
 ```{r}
-states <- seq(0,20, length.out = 200)
+states <- seq(0,25, length.out = 200)
 actions <- states
 obs <- states
-sigma_g <- 0.04
+sigma_g <- 0.05
 reward_fn <- function(x,h) pmin(x,h)
@@
 }
-f2 <- function(x, h = 0, r = 1, K = 10){
+f2 <- function(x, h = 0, r = 0.5, K = 10){
   s <- pmax(x - h, 0)
@@
 
-transitions <- lapply(models, function(f) transition_matrices(f, states, actions, sigma_g))
+model_sigmas <- c(sigma_g, sigma_g, sigma_g)
+transitions <- lapply(seq_along(models), function(i) transition_matrices(models[[i]], states, actions, model_sigmas[[i]]))
 
@@
 
-Note that the pattern is not driven merely be differences in the unfished equilibrium.
-Compare forecast performance under management.
-Score is the sum of the log probabilities of observed outcomes, given the
+```{r}
+compare <- sims %>%
+  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
+  filter(reps == 2, time < 10) %>%
+  rowwise() %>%
+  mutate(prob =  transitions[[model]][state_index, next_state, action_index],
+         expected = transitions[[model]][state_index, , action_index]  %*% states,
+         low = states[ max(which(cumsum(transitions[[model]][state_index,,action_index]) < 0.025)) ],
+         high = states[ min(which(cumsum(transitions[[model]][state_index,,action_index]) > 0.975)) ],
+         true = states[ next_state],
+         model = as.character(model))
+
+compare %>% filter(model < 3) %>% 
+  select(time, model, true, expected, low, high) %>% 
+  ggplot(aes(time, col = model, fill = model)) + 
+  geom_point(aes(y = true), pch = "*", size = 10) +
+  geom_point(aes(y = expected)) + 
+  geom_errorbar(aes(ymin = low, ymax = high), alpha = 0.5)
+```
 
@@
   summarise(score = sum(log(prob))) 
+```
 
+````{r}
 scores %>% group_by(model) %>% summarise(mean_score = mean(score))

### 2020-07-08 notebook/bad-forecast-good-decision.Rmd
maintain names on models

@@
 models <- list(f1 = f1, f2 = f2, f3 = f3)
+model_sigmas <- c(sigma_g, sigma_g, sigma_g)
 
@@
 
-model_sigmas <- c(sigma_g, sigma_g, sigma_g)
-transitions <- lapply(seq_along(models), function(i) transition_matrices(models[[i]], states, actions, model_sigmas[[i]]))
+transitions <- lapply(seq_along(models), 
+                      function(i) transition_matrices(models[[i]], 
+                                                      states, 
+                                                      actions, 
+                                                      model_sigmas[[i]]))
+names(transitions) <- c("f1", "f2", "f3")
 

### 2020-07-08 notebook/bad-forecast-good-decision.Rmd
reknit

--- 2020-07-03 notebook/bad-forecast-good-decision.Rmd
@@
 
-```{r message=FALSE, warning=FALSE}
+```{r graphics_setup, message=FALSE, warning=FALSE}
+## Plotting theme settings -- aesthetics only, all can be omitted without consequence
 library(ggthemes)
@@
 
-scale_colour_discrete <- function(...) scale_colour_solarized()
-scale_fill_discrete <- function(...) scale_fill_solarized()
-pal <- solarized_pal()(6)
+scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
+scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
+pal <- ggthemes::solarized_pal()(6)
 txtcolor <- "#586e75"
@@
 
-```{r}
+```{r setup, message=FALSE}
+## Required computational libraries
 library(tidyverse)
@@
 
-```{r}
-states <- seq(0,20, length.out = 200)
+```{r model_definitions}
+states <- seq(0,24, length.out = 240)
 actions <- states
 obs <- states
-sigma_g <- 0.04
+sigma_g <- 0.05
 reward_fn <- function(x,h) pmin(x,h)
@@
 
-f4  <- function(x, h = 0, r = 4e-7, K = 10){
-  s <- pmax(x - h, 0)
-  s + s ^ 8 * r * (1 - s / K) + s * (1 - s / K)
-}
+# looks even closer to f2; not used
+#f4  <- function(x, h = 0, r = 4e-7, K = 10){
+#  s <- pmax(x - h, 0)
+#  s + s ^ 8 * r * (1 - s / K) + s * (1 - s / K)
+#}
 
 models <- list(f1 = f1, f2 = f2, f3 = f3)
+model_sigmas <- c(sigma_g, 1.5 * sigma_g, sigma_g)
 
 true_model <- "f3"
-
-## Could consider ricker-style versions of these models. need to take care to set limits correctly
-## to get correspondence (recall x * e^y ~ x *(1 + y) for small y), e.g. something like
-#f1 <- function(x, h = 0, r = .1, K = 15){
-#  s <- pmax(x - h, 0)
-#  s * exp(r * (1 - s / K) )
-#}
-#f3  <- function(x, h = 0, r = .01, K = 10){
-#  s <- pmax(x - h, 0)
-#  s * exp(s ^ 3 * r * (1 - s / K))
-#}
-
 ```
 
-```{r}
-d <- map_dfc(models, function(f) f(states) - states) %>% mutate(state = states)
-d %>% pivot_longer(names(models), "model") %>%
-  ggplot(aes(state, value, col=model)) +
-  geom_point() + 
-  geom_hline(aes(yintercept = 0)) + 
-  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16))
-```
-
-```{r}
+```{r transition_matrices}
 # A function to compute the transition matrices for each model:
@@
         } else {
-          x <- x / sum(x) # normalize evenly. 
-          ## pile excess on boundary
-          #N <- plnorm(states[n_s], log(nextpop), sigma_g)
-          #x <- x * N / sum(x)
-          #x[n_s] <- 1 - N + x[n_s]
+          x <- x / sum(x) # normalize evenly
           transition[k, , i] <- x
@@
 
-## Reward matrix is shared by each model
+## Compute reward matrix (shared across all models)
 n_s <- length(states)
@@
 
-transitions <- lapply(models, function(f) transition_matrices(f, states, actions, sigma_g))
+## Now compute the transition matrices for each model
+transitions <- lapply(seq_along(models), 
+                      function(i) transition_matrices(models[[i]], 
+                                                      states, 
+                                                      actions, 
+                                                      model_sigmas[[i]]))
+names(transitions) <- c("f1", "f2", "f3")
 
@@
 
-## Forecasting performance
+```{r sdp, results="hide"}
+## This is the most (only) computationally expensive code chunk:
+policies <- 
+  map_dfr(transitions, 
+          function(P){
+    soln <- mdp_value_iteration(P, reward, discount = discount)
+    escapement <- states - actions[soln$policy]
+    tibble(states, policy = soln$policy, escapement)
+    }, 
+  .id = "model")
+```
+
+```{r simulations}
+library(mdplearning)
+Tmax <- 100
+x0 <- which.min(abs(states - 6))
+reps <- 100
+set.seed(12345)
+
+## Simulate each policy reps times, with `f3` as the true model:
+
+sims <- map_dfr(names(transitions), 
+                function(m){
+                  policy <- policies %>% filter(model == m) %>% pull(policy)
+                  map_dfr(1:reps, 
+                          function(i){
+                            mdp_planning(transitions[[true_model]], reward, discount,
+                                     policy = policy, x0 = x0, Tmax = Tmax) %>%
+                              select(value, state_index = state, time, action_index = action)  %>% 
+                              mutate(state = states[state_index]) # index->value
+                            },
+                          .id = "reps")
+                },
+                .id = "model")
+
+```
+
+## Forecast skill
+
+Note the low variability in the predicted value is a result of the decision strategy, which tweaks harvest up or down with variation.
+
+```{r plot_stepahead}
+compare <- sims %>%
+  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
+  rowwise() %>%
+  mutate(prob =  transitions[[model]][state_index, next_state, action_index],
+         expected = transitions[[model]][state_index, , action_index]  %*% states,
+         low = states[ max(which(cumsum(transitions[[model]][state_index,,action_index]) < 0.025)) ],
+         high = states[ min(which(cumsum(transitions[[model]][state_index,,action_index]) > 0.975)) ],
+         true = states[ next_state],
+         model = as.character(model)) %>%
+ select(time, model, true, expected, low, high, prob, reps)
+
+compare %>% filter(model != "3") %>%  filter(reps == 3, time < 10) %>%
+  ggplot(aes(time, col = model, fill = model)) + 
+  geom_point(aes(y = true), pch = "*", size = 12) +
+  geom_point(aes(y = expected)) + 
+  geom_errorbar(aes(ymin = low, ymax = high))
+```
+
+```{r plot_stepahead_unfished}
+compare_unfished <- sims
+compare_unfished$state_index <- rep(sims$state_index[sims$model == "1"],3)
+compare_unfished <- compare_unfished %>% 
+  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
+  rowwise() %>%
+  mutate(expected = transitions[[model]][state_index, , 1]  %*% states,
+         low = states[ max(which(cumsum(transitions[[model]][state_index,,1]) < 0.025)) ],
+         high = states[ min(which(cumsum(transitions[[model]][state_index,,1]) > 0.975)) ],
+         true = states[ next_state ],
+         model = as.character(model)) %>% filter(reps == 2, time < 10, model != "3")
+
+compare_unfished %>%
+  ggplot(aes(time, col = model, fill = model)) + 
+  geom_point(aes(y = expected)) + 
+  geom_errorbar(aes(ymin = low, ymax = high)) +
+  geom_point(aes(y = true), pch = "*", size = 12, col = "grey20", show.legend = FALSE)
+```
+
+```{r prob_scores, message = FALSE}
+scores <- sims %>%
+  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
+  filter(model < 3) %>%
+  rowwise() %>%
+  mutate(prob =  transitions[[model]][state_index, next_state, action_index] ) %>% 
+  group_by(reps, model) %>%
+  summarise(score = sum(log(prob)))
+
+scores %>% group_by(model) %>% summarise(mean_score = mean(score))
+
+scores %>% mutate(model = as.factor(model)) %>% 
+  ggplot(aes(x = score, group = model, fill = model)) +
+  geom_histogram(binwidth = 70)
+
+```
+
+## Simulations
+
+```{r plot_sims}
+fig_ts <- 
+  sims %>%
+  filter(time < 25, reps < 5) %>%
+  ggplot(aes(time, state, col=model, group = interaction(model,reps))) + 
+  geom_line(alpha=0.3) 
+fig_ts
+```
+
+```{r plot_npv, message = FALSE, fig.cap = "Corresponding utility (measured as mean net present value, that is: cumulative value, discounting future values by $\\delta^t$, averaged across replicates).
+Note that the exepcted utility under model 1, which has the worst forecast, is nearly identical to the optimal utility achieved by managing under the correct model, 3.
+The utility derived from model 2 is far smaller, despite it's overall better performance in long term forecasts."}
+##  Net Present Value accumulates over time, equivalent for models with near-identical management stategy
+npv_df <- sims %>% 
+  group_by(model, reps) %>%
+  mutate(npv = cumsum(value * discount ^ time)) %>%
+  group_by(time, model)  %>% 
+  summarise(mean_npv = mean(npv)) %>% 
+  arrange(model, time) %>% ungroup()
+
+optimal <- select(filter(npv_df, model == "3"), time, mean_npv)
+
+npv_df %>%
+  filter(model != "3", time %in% seq(1,100, by = 5)) %>%
+  ggplot(aes(time, mean_npv)) +
+  geom_line(data = optimal, lwd = 1.5, col = "grey20") +
+  geom_point(aes(col=model), size = 4, alpha = 0.8) + 
+  ylab("Net present value") + xlab("time")
+
+```
+
+```{r plot_models}
+d <- map_dfc(models, function(f) f(states) - states) %>% mutate(state = states)
+d %>% pivot_longer(names(models), "model") %>%
+  ggplot(aes(state, value, col=model)) +
+  geom_point() + 
+  geom_hline(aes(yintercept = 0)) + 
+  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16))
+```
+
+```{r plot_policies}
+policies %>%
+  ggplot(aes(states,escapement, col=model, lty=model)) + geom_line(lwd=2)
+```
+
+## Conclusions
+
+- A good forecast does not mean good management
+- A bad forecast does not mean the model is bad for management
+- A model permitting successful management does not imply the model is "correct" or generally "good at forecasting"
+
+Model with the egregiously optimistic long-term forecast for the stock size of an unexploited fishery nevertheless actually leads to more conservative management.
+In contrast, the model which correctly predicts the long-term average stock size without fishing nevertheless leads to substantial over-harvesting.
+
+```{r}
+```
+
+# Appendix
+
+### Unharvested forecasting performance
+
+Here we examine the forecasts under each model, `Tmax` steps into the future, 
+ignoring harvesting.  This largely highlights only the differences in the 
+unfished equilibria of the models.  It less surprising that bad performance
+in predicting the long-term equilibrium would not necessarily mean bad management.
 
@@
 ```
-
-## Model analysis
-
-```{r}
-
-policies <- map_dfr(transitions, function(P){
-  soln <- mdp_value_iteration(P, reward, discount = discount)
-  tibble(states, policy = soln$policy, escapement = states - actions[soln$policy])
-}, .id = "model")
-
-```
-
-```{r}
-policies %>%
-  ggplot(aes(states,escapement, col=model, lty=model)) + geom_line(lwd=2)
-```
-
-Management under the wrong model (`f3`)
-
-```{r}
-library(mdplearning)
-Tmax <- 100
-x0 <- which.min(abs(states - 6))
-reps <- 100
-set.seed(12345)
-
-## Simulate each policy reps times, with `f3` as the true model:
-
-sims <- map_dfr(names(transitions), 
-                function(m){
-                  policy <- policies %>% filter(model == m) %>% pull(policy)
-                  map_dfr(1:reps, 
-                          function(i){
-                            mdp_planning(transitions[[true_model]], reward, discount,
-                                     policy = policy, x0 = x0, Tmax = Tmax) %>%
-                              select(value, state_index = state, time, action_index = action)  %>% 
-                              mutate(state = states[state_index]) # index->value
-                            },
-                          .id = "reps")
-                },
-                .id = "model")
-
-```
-
-```{r}
-fig_ts <- 
-  sims %>%
-  filter(time < 25, reps < 5) %>%
-  ggplot(aes(time, state, col=model, group = interaction(model,reps))) + 
-  geom_line(alpha=0.5) 
-fig_ts
-```
-
-Note that the pattern is not driven merely be differences in the unfished equilibrium.
-Compare forecast performance under management.
-Score is the sum of the log probabilities of observed outcomes, given the
-
-```{r}
-scores <- sims %>%
-  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
-  filter(model < 3) %>%
-  select(time, state_index, next_state, action_index, model, reps) %>% 
-  rowwise() %>%
-  mutate(prob =  transitions[[model]][state_index, next_state, action_index] ) %>% 
-  group_by(reps, model) %>%
-  summarise(score = sum(log(prob)))
-
-scores %>% group_by(model) %>% summarise(mean_score = mean(score))
-
-scores %>% mutate(model = as.factor(model)) %>% 
-  ggplot(aes(x = score, group = model, fill = model)) +
-  geom_histogram(binwidth = 70)
-
-```
-
-```{r fig.cap = "Corresponding utility (measured as mean net present value, that is: cumulative value, discounting future values by $\\delta^t$, averaged across replicates).
-Note that the exepcted utility under model 1, which has the worst forecast, is nearly identical to the optimal utility achieved by managing under the correct model, 3.
-The utility derived from model 2 is far smaller, despite it's overall better performance in long term forecasts."}
-##  Net Present Value accumulates over time, equivalent for models with near-identical management stategy
-npv_df <- sims %>% 
-  group_by(model, reps) %>%
-  mutate(npv = cumsum(value * discount ^ time)) %>%
-  group_by(time, model)  %>% 
-  summarise(mean_npv = mean(npv))
-npv_df %>%
-  ggplot(aes(time, mean_npv, fill=model)) +
-  ## Area plots are weird but avoids overplotting
-  geom_area(alpha = 0.9) +
-   #geom_ribbon(aes(ymin = 0, ymax = mean_npv), alpha = 0.6) + 
-  ylab("Net present value") + xlab("time")
-
-```
-
-## Conclusions
-
-- A good forecast does not mean good management
-- A bad forecast does not mean the model is bad for management
-- A model permitting successful management does not imply the model is "correct" or generally "good at forecasting"
-
-Model with the egregiously optimistic long-term forecast for the stock size of an unexploited fishery nevertheless actually leads to more conservative management.
-In contrast, the model which correctly predicts the long-term average stock size without fishing nevertheless leads to substantial overharvesting.
-
-```{r}
-```

### 2020-09-10 manuscript/manuscript.Rmd
reknit
:pen:

@@
 ---
-title: "Forecast accuracy can be a poor metric for decision-making"
+title: "Bad forecast, good decision: accuracy can be a poor metric for policy"
 date: "`r Sys.Date()`"
@@
 
-```{r setup, message = FALSE}
-library(tidyverse)   # plotting and data manipulation
+This result will be considered by some as so preposterous that it can surely be ignored in any real world situations, and by others as so inherently obvious and commonplace as not to merit publication.
+It cannot be both.
+Forecasting is a well developed and rapidly growing field, accelerated by the explosion of available data and the convergence of statistical techniques with those now emerging from artificial intelligence.
+
+A primary purpose of statistical analyses and ecological modeling is to make forecasts for the future.
+Accurate forecasts are important both in assessing the accuracy of our models and understanding of natural processes, and can also underpin policy and decision making.
+Yet the model that leads to the best decisions is not always the model that makes the most accurate forecasts, as I illustrate here.
+Ecological processes are intrinsically complex, so that even our best models can only ever be approximations of underlying processes.
+In choosing the best model for decision-making, it can be more important to capture a single key feature of the process than it is to make the most accurate prediction about future states.
+Surprisingly, this is true not only in conceptual models, but also when a decision policy is derived directly from a complex optimization routine of a probablistic predictive model, such as the stochastic dynamic programming (SDP) algorithms used to solve Markov Decision Processes [@Marescot2013].
+
+Many ecological management problems are sequential decision problems, in which each year (or other interval) a manager must observe the state of the system and choose a course of action to maximize long term objectives.
+Such problems inherently depend on forecasts: each possible action can result in a different forecast for the future state, typically reflecting some uncertainty as well.
+The utility the manager derives may depend on both the choice of action and the state of the system, reflecting the costs and benefits associated with each.
+Sequential decision-making problems are distinguished by the need to think more than one move ahead.
+For instance, harvesting as many fish as possible in year one may maximize the market value that year, but if too few fish are left to reproduce then harvest in future years will suffer.
+The same calculus of thinking ahead frequently applies to rebuilding species populations as well [e.g.
+@Lambert;  @Chades2008].
+
+Here I focus on a classic problem in conservation and resource management: optimal harvest of a fishery [@Schaefer1954; @Clark1974; @Reed1979].
+Fisheries are a significant economic and conservation concern world wide and their management remains an important debate [e.g.
+@Worm2006; @Costello2016].
+Moreover, their management has been both a proving grounds for theoretical and practical decision-making issues which are widely applicable in other areas of ecology and conservation.
+For simplicity, I will focus on the case of a single-species model whose population is observed annually without error in a stationary environment without age structure.
+I will further assume that the manager seeks only to maximize the net present value of the harvest.
+These assumptions are not at all necessary for the analysis and results presented here, and a large literature already exists relaxing many of these assumptions (e.g.
+@Holden2015; @Memarzadeh2018; ..., though in practice, large scale analyses such as @Costello2016 still make these same assumptions).
+Rather, I have kept the model assumptions simple only for clarity of presentation, which will allow the reader to see that this counter-intuitive result does not arise from the complexity of any particular parametric assumptions
+
+More precisely, the decision problem in question can be stated as follows: The fish stock is observed to be in state `X_t` at time `t`, and is then subjected to some harvest `H_t` before recruiting new fish, subject to stochastic environmental noise `\xi_t`, to bring the stock to $X_t+1$,
+
+\begin{equation}
+X_{t+1} = f(X_t - H_t, \xi_t) 
+\end{equation}
+
+and the manager seeks to maximize the sum of the utility derived from such a harvest and such a state, $U(X_t,H_t)$, over all time, subject to discount rate $\delta$:
+
+\begin{equation}
+\sum_{t=0}^{t=\infty} U(X_t, H_t) \delta^t
+\end{equation}
+
+While in principle the utility could reflect many things, including the cost of fishing, market responses to supply and demand, the value of recreational fishing, the intrinsic value fish left in the sea [see @Halpern], for simplicity we will assume utility is merely a linear function of the harvest, i.e.
+a fixed price $p$ per kilogram of fish harvested: $U(X_t, H_t) = p H_t$.
+This problem is already well studied, and it is worth noting that even under such a pessimistic assumption, the optimal strategy still seeks to sustain the fish population indefinitely; as @Clark1974 shows for the deterministic function $f$ and @Reed1979 extended to the stochastic case.
+Given the function $f$ with known parameters, it is straight forward to determine the optimal harvest policy by stochastic dynamic programming (SDP, see @Mangel; @Marescot2013).
+
+```{r graphics_setup, message=FALSE, warning=FALSE}
+knitr::opts_chunk$set(cache=TRUE)
+
+## Plotting theme settings -- aesthetics only, all can be omitted without consequence
+library(ggthemes)
+library(hrbrthemes)
+library(ggplot2)
+library(Cairo)
+library(extrafont)
+library(patchwork)
+extrafont::loadfonts()
+ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
+#theme_set(theme_solarized(base_size=16))
+
+scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
+scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
+pal <- ggthemes::solarized_pal()(6)
+txtcolor <- "#586e75"
 ```
 
-Is forecasting is the ultimate test of theory?
+```{r setup, message=FALSE}
+## Required computational libraries
+library(tidyverse)
+library(MDPtoolbox)
+library(expm)
+library(mdplearning) # remotes::install_github("boettiger-lab/mdplearning")
+```
+
+\begin{equation}
+f_i(Y) = Y + Y r_i \left( 1 - \frac{Y}{K_i} \right)
+\end{equation}
+
+Model 1 is given by `r = 2,
+
+```{r model_definitions}
+states <- seq(0,24, length.out = 240)
+actions <- states
+obs <- states
+sigma_g <- 0.05
+reward_fn <- function(x,h) pmin(x,h)
+discount <- 0.99
+
+# K is at twice max of f3; 8 * K_3 / 5
+f1 <- function(x, h = 0, r = 2, K = 10 * 8 / 5){
+  s <- pmax(x - h, 0)
+  s + s * (r * (1 - s / K) )
+}
+f2 <- function(x, h = 0, r = 0.5, K = 10){
+  s <- pmax(x - h, 0)
+  s + s * (r * (1 - s / K) )
+}
+
+# max is at 4 * K / 5 
+f3  <- function(x, h = 0, r = .002, K = 10){
+  s <- pmax(x - h, 0)
+  s + s ^ 4 * r * (1 - s / K)
+}
+
+# looks even closer to f2; not used
+#f4  <- function(x, h = 0, r = 4e-7, K = 10){
+#  s <- pmax(x - h, 0)
+#  s + s ^ 8 * r * (1 - s / K) + s * (1 - s / K)
+#}
+
+models <- list(f1 = f1, f2 = f2, f3 = f3)
+model_sigmas <- c(sigma_g, 1.5 * sigma_g, sigma_g)
+
+true_model <- "f3"
+```
+
+```{r transition_matrices}
+# A function to compute the transition matrices for each model:
+transition_matrices <- function(f,
+                      states,
+                      actions,
+                      sigma_g){
+
+  n_s <- length(states)
+  n_a <- length(actions)
+
+  transition <- array(0, dim = c(n_s, n_s, n_a))
+  for (k in 1:n_s) {
+    for (i in 1:n_a) {
+      nextpop <- f(states[k], actions[i])
+      if(nextpop <= 0){
+        transition[k, , i] <- c(1, rep(0, n_s - 1))
+      } else if(sigma_g > 0){
+        x <- dlnorm(states, log(nextpop), sdlog = sigma_g)
+        if(sum(x) == 0){ ## nextpop is computationally zero
+          transition[k, , i] <- c(1, rep(0, n_s - 1))
+        } else {
+          x <- x / sum(x) # normalize evenly
+          transition[k, , i] <- x
+        }
+      } else {
+        stop("sigma_g not > 0")
+      }
+      reward[k, i] <- reward_fn(states[k], actions[i])
+    }
+  }
+  transition
+}
+
+## Compute reward matrix (shared across all models)
+n_s <- length(states)
+n_a <- length(actions)
+reward <- array(0, dim = c(n_s, n_a))
+for (k in 1:n_s) {
+  for (i in 1:n_a) {
+    reward[k, i] <- reward_fn(states[k], actions[i])
+  }
+}
+
+## Now compute the transition matrices for each model
+transitions <- lapply(seq_along(models), 
+                      function(i) transition_matrices(models[[i]], 
+                                                      states, 
+                                                      actions, 
+                                                      model_sigmas[[i]]))
+names(transitions) <- c("f1", "f2", "f3")
+
+```
+
+```{r sdp, results="hide"}
+## This is the most (only) computationally expensive code chunk. use memoization to cache
+mdp <- memoise::memoise(mdp_value_iteration, cache = memoise::cache_filesystem("cache/"))
+
+policies <- 
+  map_dfr(transitions, 
+          function(P){
+    soln <- mdp(P, reward, discount = discount, 
+                epsilon = 0.01, max_iter = 1000, V0 = rep(0,dim(P)[[1]]))
+    escapement <- states - actions[soln$policy]
+    tibble(states, policy = soln$policy, escapement)
+    }, 
+  .id = "model")
+```
+
+```{r simulations}
+library(mdplearning)
+Tmax <- 100
+x0 <- which.min(abs(states - 6))
+reps <- 100
+set.seed(12345)
+
+## Simulate each policy reps times, with `f3` as the true model:
+
+sims <- map_dfr(names(transitions), 
+                function(m){
+                  policy <- policies %>% filter(model == m) %>% pull(policy)
+                  map_dfr(1:reps, 
+                          function(i){
+                            mdp_planning(transitions[[true_model]], reward, discount,
+                                     policy = policy, x0 = x0, Tmax = Tmax) %>%
+                              select(value, state_index = state, time, action_index = action)  %>% 
+                              mutate(state = states[state_index]) # index->value
+                            },
+                          .id = "reps")
+                },
+                .id = "model")
+
+```
+
+```{r stepahead_unfished}
+stepahead_unfished <- sims
+stepahead_unfished$state_index <- rep(sims$state_index[sims$model == "1"],3)
+stepahead_unfished <- stepahead_unfished  %>% 
+  filter(model != "3") %>% 
+  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
+  rowwise() %>%
+  mutate(expected = transitions[[model]][state_index, , 1]  %*% states,
+         var = transitions[[model]][state_index, , 1]  %*% states ^ 2 - expected ^ 2,
+         low = states[ max(which(cumsum(transitions[[model]][state_index,,1]) < 0.025)) ],
+         high = states[ min(which(cumsum(transitions[[model]][state_index,,1]) > 0.975)) ],
+         true = states[ next_state ],
+         model = as.character(model)) 
+```
+
+```{r fig1a}
+fig1A <- stepahead_unfished %>% 
+  filter(reps == 2, time < 10) %>%
+  ggplot(aes(time, col = model, fill = model)) + 
+  geom_point(aes(y = expected), show.legend = FALSE) + 
+  geom_errorbar(aes(ymin = low, ymax = high), show.legend = FALSE) +
+  geom_point(aes(y = true), pch = "*", size = 12, col = "grey40", show.legend = FALSE)
+
+```
+
+```{r stepahead_fished}
+stepahead_fished <- sims %>% 
+  filter(model != "3") %>%
+  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
+  rowwise() %>%
+  mutate(prob =  transitions[[model]][state_index, next_state, action_index],
+         expected = transitions[[model]][state_index, , action_index]  %*% states,
+         var = transitions[[model]][state_index, , action_index]  %*% states ^ 2 - expected ^ 2,
+         low = states[ max(which(cumsum(transitions[[model]][state_index,,action_index]) < 0.025)) ],
+         high = states[ min(which(cumsum(transitions[[model]][state_index,,action_index]) > 0.975)) ],
+         true = states[ next_state],
+         model = as.character(model)) %>%
+ select(time, model, true, expected, low, high, var, prob, reps)
+```
+
+```{r fig1b}
+fig1B <- stepahead_fished  %>%  
+  filter(reps == 3, time < 10) %>%
+  ggplot(aes(time, col = model, fill = model)) + 
+  geom_point(aes(y = true), pch = "*", size = 12) +
+  geom_point(aes(y = expected)) + 
+  geom_errorbar(aes(ymin = low, ymax = high))
+ # geom_errorbar(aes(ymin = expected - 2 * sqrt(var), ymax=expected + 2 * sqrt(var)))
+
+```
+
+```{r proper_scores, message = FALSE}
+
+scoring_fn <- function(x, mu, sigma){ -(mu - x )^2 / sigma^2  - log(sigma)}
+
+proper_scores_unfished <- stepahead_unfished %>%
+  mutate(sd = sqrt(var),
+         score = scoring_fn(expected, true, sd)) %>%
+  select(model, reps, time, score)
+
+fig1C <- proper_scores_unfished %>%
+  ggplot(aes(x = score, group = model, fill = model)) +
+  geom_histogram(binwidth = 2, show.legend = FALSE) +
+  coord_cartesian(xlim = c(-100, 1))
+
+proper_scores_fished <- stepahead_fished %>%
+  mutate(sd = sqrt(var),
+         score = scoring_fn(expected, true, sd)) %>%
+  select(model, reps, time, score)
+
+fig1D <- proper_scores_fished %>%
+  ggplot(aes(x = score, group = model, fill = model)) +
+  geom_histogram(binwidth = 2) +
+  coord_cartesian(xlim = c(-100, 1))
+
+```
+
+# Results
+
+```{r figure1, fig.width=7, fig.height=6}
+fig1A / fig1B
+```
+
+```{r}
+ (fig1C + fig1D)
+```
+
+```{r prob_scores, message = FALSE}
+scores <- sims %>%
+  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
+  filter(model < 3) %>%
+  rowwise() %>%
+  mutate(prob =  transitions[[model]][state_index, next_state, action_index] ) %>% 
+  group_by(reps, model) %>%
+  summarise(score = sum(log(prob)))
+
+score_table <- scores %>% group_by(model) %>% summarise(mean_score = mean(score))
+
+scores %>% mutate(model = as.factor(model)) %>% 
+  ggplot(aes(x = score, group = model, fill = model)) +
+  geom_histogram(binwidth = 70)
+
+```
+
+```{r plot_sims}
+fig_ts <- 
+  sims %>%
+  filter(time < 25, reps < 5) %>%
+  ggplot(aes(time, state, col=model, group = interaction(model,reps))) + 
+  geom_line(alpha=0.3) 
+fig_ts
+```
+
+```{r plot_npv, message = FALSE, fig.cap = "Corresponding utility (measured as mean net present value, that is: cumulative value, discounting future values by $\\delta^t$, averaged across replicates).
+Note that the exepcted utility under model 1, which has the worst forecast, is nearly identical to the optimal utility achieved by managing under the correct model, 3.
+The utility derived from model 2 is far smaller, despite it's overall better performance in long term forecasts."}
+##  Net Present Value accumulates over time, equivalent for models with near-identical management stategy
+npv_df <- sims %>% 
+  group_by(model, reps) %>%
+  mutate(npv = cumsum(value * discount ^ time)) %>%
+  group_by(time, model)  %>% 
+  summarise(mean_npv = mean(npv)) %>% 
+  arrange(model, time) %>% ungroup()
+
+optimal <- select(filter(npv_df, model == "3"), time, mean_npv)
+
+npv_df %>%
+  filter(model != "3", time %in% seq(1,100, by = 5)) %>%
+  ggplot(aes(time, mean_npv)) +
+  geom_line(data = optimal, lwd = 1.5, col = "grey20") +
+  geom_point(aes(col=model), size = 4, alpha = 0.8) + 
+  ylab("Net present value") + xlab("time")
+
+```
+
+```{r plot_models}
+d <- map_dfc(models, function(f) f(states) - states) %>% mutate(state = states)
+d %>% pivot_longer(names(models), "model") %>%
+  ggplot(aes(state, value, col=model)) +
+  geom_point() + 
+  geom_hline(aes(yintercept = 0)) + 
+  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16))
+```
+
+```{r plot_policies}
+policies %>%
+  ggplot(aes(states,escapement, col=model, lty=model)) + geom_line(lwd=2)
+```
+
+# Discussion
 

### 2020-09-15 manuscript/manuscript.Rmd
:pen:

@@
 
-While in principle the utility could reflect many things, including the cost of fishing, market responses to supply and demand, the value of recreational fishing, the intrinsic value fish left in the sea [see @Halpern], for simplicity we will assume utility is merely a linear function of the harvest, i.e.
+While in principle the utility could reflect many things, including the cost of fishing, market responses to supply and demand, the value of recreational fishing, the intrinsic value fish left in the sea [see @Halpern2013], for simplicity we will assume utility is merely a linear function of the harvest, i.e.
 a fixed price $p$ per kilogram of fish harvested: $U(X_t, H_t) = p H_t$.
-This problem is already well studied, and it is worth noting that even under such a pessimistic assumption, the optimal strategy still seeks to sustain the fish population indefinitely; as @Clark1974 shows for the deterministic function $f$ and @Reed1979 extended to the stochastic case.
-Given the function $f$ with known parameters, it is straight forward to determine the optimal harvest policy by stochastic dynamic programming (SDP, see @Mangel; @Marescot2013).
+This problem is already well studied, and it is worth noting that even under such a pessimistic assumption, the optimal strategy still seeks to sustain the fish population indefinitely; as @Clark1973 shows for the deterministic function $f$ and @Reed1979 extended to the stochastic case.
+Given the function $f$ with known parameters, it is straight forward to determine the optimal harvest policy by stochastic dynamic programming (SDP, see @Mangel1985; @Marescot2013).
 

### 2020-10-16 manuscript/appendix.Rmd
WIP :pen:

--- 2020-07-02 manuscript/manuscript.Rmd
@@
----
-title: "Forecast accuracy can be a poor metric for decision-making"
-date: "`r Sys.Date()`"
-author:
-  - name: Carl Boettiger
-    email: cboettig@berkeley.edu
-    affiliation: ucb
-#    footnote: Corresponding Author
-address:
-  - code: ucb
-    address: "ESPM Department, University of California, 130 Mulford Hall Berkeley, CA 94720-3114, USA"
-abstract: |
+## Decision Context
 
-journal: Theoretical ecology
-layout: 3p
-bibliography: refs.bib
-output: rticles::elsevier_article
-keywords: 
-  - transients, 
-  - optimal control, 
-  - adaptive management, 
-  - stochasticity, 
-  - uncertainty, 
-  - ecological management
-preamble: |
-  \journal{preprint}
-  ## EditorialManager adds its own linenumbers to pdf already
-  \usepackage{lineno}
-  \linenumbers
+and the manager seeks to maximize the sum of the utility derived from such a harvest and such a state, $U(X_t,H_t)$, over all time, subject to discount rate $\delta$:
 
----
+\begin{equation}
+\sum_{t=0}^{t=\infty} U(X_t, H_t) \delta^t
+\end{equation}
 
-```{r setup, message = FALSE}
-library(tidyverse)   # plotting and data manipulation
+While in principle the utility could reflect many things, including the cost of fishing, market responses to supply and demand, the value of recreational fishing, the intrinsic value fish left in the sea [see @Halpern2013], for simplicity we will assume utility is merely a linear function of the harvest, i.e.
+a fixed price $p$ per kilogram of fish harvested: $U(X_t, H_t) = p H_t$.
+This problem is already well studied, and it is worth noting that even under such a pessimistic assumption, the optimal strategy still seeks to sustain the fish population indefinitely; as @Clark1973 shows for the deterministic function $f$ and @Reed1979 extended to the stochastic case.
+Given the function $f$ with known parameters, it is straight forward to determine the optimal harvest policy by stochastic dynamic programming (SDP, see @Mangel1985; @Marescot2013).
+
+```{r prob_scores, message = FALSE}
+## Probability scores instead of proper scores
+scores <- sims %>%
+  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
+  filter(model < 3) %>%
+  rowwise() %>%
+  mutate(prob =  transitions[[model]][state_index, next_state, action_index] ) %>% 
+  group_by(reps, model) %>%
+  summarise(score = sum(log(prob)))
+
+score_table <- scores %>% group_by(model) %>% summarise(mean_score = mean(score))
+
+scores %>% mutate(model = as.factor(model)) %>% 
+  ggplot(aes(x = score, group = model, fill = model)) +
+  geom_histogram(binwidth = 70)
+
 ```
-
-Is forecasting is the ultimate test of theory?
-
-# Acknowledgements
-
-This work was supported in part by NSF CAREER () and computational resources from NSF's XSEDE Jetstream (DEB160003) and Chameleon cloud platforms.
-
-\pagebreak
-
-# References

### 2020-10-16 manuscript/manuscript.Rmd
WIP :pen:

--- 2020-09-10 manuscript/manuscript.Rmd
@@
 abstract: |
-
+    With the rapid expansion of available data, it is frequently suggested that _forecasting performance_ is the ultimate test of a scientific model [e.g.
+@Dietz2018].
+Yet the model that makes the most accurate forecasts is not always the model that yields the best decisions.
+This is no accident of statistics, even formal criteria such as strictly proper scoring rules [@Gneiting2007] cannot determine the best model for decision-making because they do not consider the context of the decision process itself.
+I illustrate this paradox using a classic example of ecological decision making in fisheries management.
+This example also reveals how an understanding of the decision problem can both explain the poor management outcomes of model which produces the best forecast, and also see how to avoid making such a mistake.
+As forecasting becomes an increasing attractive and viable possibility for assessing ecological models and applying them to real-world decisions, I hope this illustration serves as a reminder that modeling can not be done in a vacuum, but must reflect the context in which the model will be applied.
 journal: Theoretical ecology
@@
 
-This result will be considered by some as so preposterous that it can surely be ignored in any real world situations, and by others as so inherently obvious and commonplace as not to merit publication.
-It cannot be both.
-Forecasting is a well developed and rapidly growing field, accelerated by the explosion of available data and the convergence of statistical techniques with those now emerging from artificial intelligence.
-
-A primary purpose of statistical analyses and ecological modeling is to make forecasts for the future.
+A primary purpose of statistical analyses and ecological modeling is to make forecasts for the future [].
 Accurate forecasts are important both in assessing the accuracy of our models and understanding of natural processes, and can also underpin policy and decision making.
@@
 In choosing the best model for decision-making, it can be more important to capture a single key feature of the process than it is to make the most accurate prediction about future states.
-Surprisingly, this is true not only in conceptual models, but also when a decision policy is derived directly from a complex optimization routine of a probablistic predictive model, such as the stochastic dynamic programming (SDP) algorithms used to solve Markov Decision Processes [@Marescot2013].
+Even strictly proper scoring rules for probablistic forecasts [@Gneiting2007] will not always select the best model to guide decision-making.
+Surprisingly, this is true not only in conceptual models that may guide policy choices, but also when a decision policy is derived directly from a complex optimization routine of a probablistic predictive model, such as the stochastic dynamic programming (SDP) algorithms used to solve Markov Decision Processes [@Marescot2013].
+Here, I use a classical, well-understood example from fisheries management [@Schaefer1954; @Clark1974; @Reed1979] to illustrate both the paradox of how a model with the worst forecast provides the best decision outcomes, as well as show how we can avoid selecting models that are poorly suited for management by considering the management context more explicitly.
 
+<!--
 Many ecological management problems are sequential decision problems, in which each year (or other interval) a manager must observe the state of the system and choose a course of action to maximize long term objectives.
@@
 @Lambert;  @Chades2008].
+-->
 
-Here I focus on a classic problem in conservation and resource management: optimal harvest of a fishery [@Schaefer1954; @Clark1974; @Reed1979].
 Fisheries are a significant economic and conservation concern world wide and their management remains an important debate [e.g.
 @Worm2006; @Costello2016].
-Moreover, their management has been both a proving grounds for theoretical and practical decision-making issues which are widely applicable in other areas of ecology and conservation.
-For simplicity, I will focus on the case of a single-species model whose population is observed annually without error in a stationary environment without age structure.
-I will further assume that the manager seeks only to maximize the net present value of the harvest.
-These assumptions are not at all necessary for the analysis and results presented here, and a large literature already exists relaxing many of these assumptions (e.g.
-@Holden2015; @Memarzadeh2018; ..., though in practice, large scale analyses such as @Costello2016 still make these same assumptions).
-Rather, I have kept the model assumptions simple only for clarity of presentation, which will allow the reader to see that this counter-intuitive result does not arise from the complexity of any particular parametric assumptions
-
-More precisely, the decision problem in question can be stated as follows: The fish stock is observed to be in state `X_t` at time `t`, and is then subjected to some harvest `H_t` before recruiting new fish, subject to stochastic environmental noise `\xi_t`, to bring the stock to $X_t+1$,
-
-\begin{equation}
-X_{t+1} = f(X_t - H_t, \xi_t) 
-\end{equation}
-
-and the manager seeks to maximize the sum of the utility derived from such a harvest and such a state, $U(X_t,H_t)$, over all time, subject to discount rate $\delta$:
-
-\begin{equation}
-\sum_{t=0}^{t=\infty} U(X_t, H_t) \delta^t
-\end{equation}
-
-While in principle the utility could reflect many things, including the cost of fishing, market responses to supply and demand, the value of recreational fishing, the intrinsic value fish left in the sea [see @Halpern], for simplicity we will assume utility is merely a linear function of the harvest, i.e.
-a fixed price $p$ per kilogram of fish harvested: $U(X_t, H_t) = p H_t$.
-This problem is already well studied, and it is worth noting that even under such a pessimistic assumption, the optimal strategy still seeks to sustain the fish population indefinitely; as @Clark1974 shows for the deterministic function $f$ and @Reed1979 extended to the stochastic case.
-Given the function $f$ with known parameters, it is straight forward to determine the optimal harvest policy by stochastic dynamic programming (SDP, see @Mangel; @Marescot2013).
+Moreover, their management has been both a proving grounds for theoretical and practical decision-making issues which are widely applicable in other areas of ecology and conservation [e.g.
+@Mangel1985; @Clark1990; @Marescot2013].
+The decision-making problem is characterized by the need for a manager to set an acceptable harvest quota $H_t$ each year given some stock assessment estimate of the current stock size (population abundance) of the species in question [@Clark1974].
+Such a decision problem appears to hinge on an accurate forecast: if we can predict to what size the stock will increase next year, $X_t+1$, knowing the current stock, $X_t$, then we can safely harvest $X_{t+1} - X_t$.
+Overestimating or underestimating such recruitment will result in over-harvesting or under-harvesting, respectively.
+Thus it may seem natural that our first step would be to select the model that makes the most accurate forecast of next year's stock, $X_{t+1}$.
+I illustrate how we do this using strictly proper scoring criteria [@Gneiting2007] for a set of candidate models,
+and show that it leads to worse decisions.
 
@@
 ```
-
-\begin{equation}
-f_i(Y) = Y + Y r_i \left( 1 - \frac{Y}{K_i} \right)
-\end{equation}
-
-Model 1 is given by `r = 2,
 
@@
 ```
+
+# Ecological Models
+
+For simplicity, I will focus on the classic case case of a single-species model whose population is observed annually without error in a stochastic but stationary environment without age structure [@Reed1979; @Clark1990; @Costello2016].
+These are not necessary assumptions -- in fact, the more complex the models become, the easier it is find examples in which the best forecast does not produce the best decision.
+Rather, using a simple model merely reflects the famous compromise of Richard Levins [@Levins1969] in choosing generality over precision.
+More precisely, the decision problem in question can be stated as follows: The fish stock is observed to be in state `X_t` at time `t`, and is then subjected to some harvest `H_t` before recruiting new fish, subject to stochastic environmental noise `\xi_t`, to bring the stock to $X_t+1$,
+
+\begin{equation}
+X_{t+1} = f(X_t - H_t, \xi_t) 
+\end{equation}
+
+Further we imagine that the function $f$ is not known precisely, and so we will rely on an evaluation of forecasting skill across a set of candidate models to determine which one to use to manage the fishery.
+Again for simplicity, we will restrict ourselves to two simple candidate models $f_1$ and $f_2$.
+Both share the same underlying structure of logistic recruitment:
+
+\begin{equation}
+f_i(Y) = Y + Y r_i \left( 1 - \frac{Y}{K_i} \right)
+\end{equation}
+
+Model 1 is given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, Model 2 by $r_2 = 0.5$,  $K_2 = 10$, $\sigma = 0.075$.
+Having both the larger growth rate and the larger carrying capacity, Model 1 is clearly the more optimistic of the two choices.
+
+Mathematical models are, at best, approximations of the underlying processes.
+Ecological processes are much too complex to ever be modeled exactly down to the last atom.
+For illustrative purposes, we will assume the "true" process to be given by a third model which we will leave unspecified for the moment and will examine in more detail later on.
 
@@
 ```{r fig1a}
-fig1A <- stepahead_unfished %>% 
+fig1a <- stepahead_unfished %>% 
   filter(reps == 2, time < 10) %>%
@@
 ```{r fig1b}
-fig1B <- stepahead_fished  %>%  
+fig1b <- stepahead_fished  %>%  
   filter(reps == 3, time < 10) %>%
@@
 
-fig1C <- proper_scores_unfished %>%
+fig1c <- proper_scores_unfished %>%
   ggplot(aes(x = score, group = model, fill = model)) +
@@
 
-fig1D <- proper_scores_fished %>%
+fig1d <- proper_scores_fished %>%
   ggplot(aes(x = score, group = model, fill = model)) +
@@
 
-```{r figure1, fig.width=7, fig.height=6}
-fig1A / fig1B
-```
+## Forecast performance
 
-```{r}
- (fig1C + fig1D)
-```
-
-```{r prob_scores, message = FALSE}
-scores <- sims %>%
-  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
-  filter(model < 3) %>%
-  rowwise() %>%
-  mutate(prob =  transitions[[model]][state_index, next_state, action_index] ) %>% 
-  group_by(reps, model) %>%
-  summarise(score = sum(log(prob)))
-
-score_table <- scores %>% group_by(model) %>% summarise(mean_score = mean(score))
-
-scores %>% mutate(model = as.factor(model)) %>% 
-  ggplot(aes(x = score, group = model, fill = model)) +
-  geom_histogram(binwidth = 70)
+```{r figure1, fig.width=7, fig.height=8}
+(fig1a / fig1b) /
+(fig1c + fig1d)
 
@@
 
+Figure 1A shows the step-ahead prediction performance of each model in a simulation of an un-fished environment, with error bars indicating the 95% confidence intervals around each prediction, while stars denote the observed value in that year.
+Model 1 predictions appear far too optimistic, with the true value falling well below the 95% confidence intervals.
+In contrast, all observed values fall easily within the confidence intervals produced by model 2.
+
+Predictive performance of the unfished population does not give us the full picture, since it reflects predictive accuracy only in the region of the true carrying capacity, while an actively harvested stock will be at a lower size.
+The model that predicts the equilibrium size may not be the one that best forecasts stock recovery.
+Further, this comparison does not yet implement any decisions that might be made from either model.
+To address these concerns, we consider the case where our fishery is managed according to the optimal harvest predicted by each model in turn.
+Each year the model produces both a forecast and a decision about the harvest quota.
+(Mechanics of determining a harvest quota given the model follow standard methods, e.g.
+@Reed1979; @Marescot2013, see appendix for details.) We then implement that harvest and compare the observed stock size the following year to that which the model has predicted, Figure 1B.
+Because the models make different decisions each year, the stock size in year 2, 3, etc under management of model 1 (blue stars) is different from that under model 2 (red stars).
+Again we observe that the observations under model 1 consistently fall well outside of the 95% confidence intervals it predicts, while under model 2, stock sizes consistently fall within the predicted intervals.
+Once again, model 2 shows a higher forecast accuracy while model 1 appears hopelessly optimistic.
+
+**Proper scores**.
+
+## Ecological and economic performance
+
+Given this evidence, model 2 clearly provides the more accurate forecast and we would no doubt conclude that model 2 was thus a better approximation of the true model and thus the better choice to inform decision making about harvest quotas.
+Yet if we revisit our experiment of managing the fishery under each model in turn, and focus not on _predictive accuracy_ but on _ecological_ and _economic_ outcomes, a different story emerges.
+Figure 2A shows the stock sizes across five replicate simulations that have been managed according to each model.
+For comparison, we have also included the results of optimal management given the true model.
+Despite its optimistic predictions, model 1 does not result in over-fishing, but holds the stock near the same level as the optimal management strategy.
+In contrast, model 2 suppresses the stock to a much lower level.
+The over-fishing in model 2 is not economically efficient either, as shown in Figure 2B.
+The net present value of the fishery, as calculated as the cumulative, discounted value of the harvest (assuming a fixed unit price for fish with negligible cost for harvest, see appendix) under the fishing regime of model 1 falls precisely along that of the optimal solution, while the value derived under model 2 is consistently lower.
+
 ```{r plot_sims}
-fig_ts <- 
+fig2a <- 
   sims %>%
@@
   geom_line(alpha=0.3) 
-fig_ts
 ```
 
-```{r plot_npv, message = FALSE, fig.cap = "Corresponding utility (measured as mean net present value, that is: cumulative value, discounting future values by $\\delta^t$, averaged across replicates).
-Note that the exepcted utility under model 1, which has the worst forecast, is nearly identical to the optimal utility achieved by managing under the correct model, 3.
-The utility derived from model 2 is far smaller, despite it's overall better performance in long term forecasts."}
+```{r plot_npv}
 ##  Net Present Value accumulates over time, equivalent for models with near-identical management stategy
@@
 
+fig2b <- 
 npv_df %>%
@@
 ```
+```{r figure2, message = FALSE, fig.cap = "Corresponding utility (measured as mean net present value, that is: cumulative value, discounting future values by $\\delta^t$, averaged across replicates).
+Note that the exepcted utility under model 1, which has the worst forecast, is nearly identical to the optimal utility achieved by managing under the correct model, 3.
+The utility derived from model 2 is far smaller, despite it's overall better performance in long term forecasts."}
+fig2a + fig2b
+```
+
+## Discussion
+
+This paradox in performance of forecasting vs performance in decision making can be easily resolved by considering the context of the decision problem more closely.
+Starting back in the origins of modern fisheries management in the 1950s, @Schaefer1954 demonstrates that the Maximum Sustainable Yield (MSY) is maintained by harvesting a stock down to the size at which it achieves its maximum growth rate, i.e.
+$K/2$ for a logistic model (or any other symmetric growth model).
+The details of this optimal harvest have been refined to allow for dynamic optimization with discounted future profits [@Clark1974] stochastic population growth [@Reed1979], and even imperfect observations [@Memarzadeh2018], but Schaefer's observation is sufficient in this case to see that the decision problem depends very much on the value of $K$ while being essentially independent of the other model parameters, including the growth rate, $r$ and stochasticity, $\sigma$.
+However, the growth rate (and magnitude of the stochasticity) matter very much to forecast accuracy.
+Figure 3 shows the functional form of our two logistic-curve models, compared to the functional form of the "true" model used to drive the simulations.
+From this graph, it is clear to see that Model 2 does indeed lie closer to the true model throughout the state space, and agrees precisely with the carrying capacity (where both functions cross zero with negative slope).
+However, because the model 3, the true model, is not symmetric, it reaches its maximum growth rate very near the stock size that also maximizes the growth rate of model 1, which is considerably higher stock size than that of model 2.
+Consequently, model 2 leads to over-fishing, while model 1 leads to nearly-optimal management, despite its poor performance as a forecaster.
+
+If we had access to model 3, we would no doubt find that it outperformed model 2 in forecast accuracy as well as ecological and economic performance.
+But in real ecological decision making, we never know the true model -- we will always be comparing among approximations.
+
+Within fisheries, even in age-structured models, recruitment approximations with symmetric growth functions (logistic, Ricker, Beverton-Holt, etc) still dominate.
 
@@
 d <- map_dfc(models, function(f) f(states) - states) %>% mutate(state = states)
+
+fig3a <- 
   d %>% pivot_longer(names(models), "model") %>%
@@
 ```{r plot_policies}
-policies %>%
+fig3b <- policies %>%
   ggplot(aes(states,escapement, col=model, lty=model)) + geom_line(lwd=2)
@@
 
-# Discussion
+```{r figure3, fig.width=5, fig.height=7}
+fig3a / fig3b
+```
 

### 2020-10-16 manuscript/manuscript.Rmd
:pen:

@@
 ---
-title: "Bad forecast, good decision: accuracy can be a poor metric for policy"
+title: "Bad forecast, good decision: what makes a model relevant for determining policy"
 date: "`r Sys.Date()`"
@@
 From this graph, it is clear to see that Model 2 does indeed lie closer to the true model throughout the state space, and agrees precisely with the carrying capacity (where both functions cross zero with negative slope).
-However, because the model 3, the true model, is not symmetric, it reaches its maximum growth rate very near the stock size that also maximizes the growth rate of model 1, which is considerably higher stock size than that of model 2.
-Consequently, model 2 leads to over-fishing, while model 1 leads to nearly-optimal management, despite its poor performance as a forecaster.
+However, the peak of model 3 very nearly matches the peak of model 1, and falls higher stock size than that of model 2.
 
+Thus, each year our model 1 managers are again chagrined to see the stock size estimates come in far below their rosy predictions, but nevertheless manage to set a nearly optimal quota by comparing the observed stock size to the model's predicted optimal escapement level [@Reed1979].
+Meanwhile, model 2 managers could only congratulate themselves that each year's observations fall neatly within their predicted interval, unaware that the they were over-exploiting the fishery by both economic and ecological metrics.
 If we had access to model 3, we would no doubt find that it outperformed model 2 in forecast accuracy as well as ecological and economic performance.
 But in real ecological decision making, we never know the true model -- we will always be comparing among approximations.
+Within fisheries, even in today's parameter-rich age-structured models, recruitment approximations with symmetric growth functions (Logistic, Ricker, Beverton-Holt, etc) still dominate [@ramlegacy].
 
-Within fisheries, even in age-structured models, recruitment approximations with symmetric growth functions (logistic, Ricker, Beverton-Holt, etc) still dominate.
+This issue is by no means unique to fisheries.
+Throughout resource management and conservation decision-making, as is no doubt true in other fields, decisions about which model to use are guided by which model best fits the data [].
+As this example illustrates, such metrics, no matter how rigorously defined, may select entirely the wrong model.
+And once we are managing with the wrong model, no amount of comparing model predictions to outcomes will guarantee we discover our mistake.
+Only by comparing to the outcomes we actually care about -- in this case, economic and ecological performance -- can we see which model is best for decision-making.
+Conversely, our example reminds us that just because a model continually makes poor predictions does not mean that the model leads to poor decisions.
+Model 1, despite it's many mistakes, is right about one key feature: the position of the peak growth rate -- and that is enough to guarentee nearly optimal performance.
+This should also be reassuring to modelers and decision makers, for it reminds us that effective models need be correct or even close to correct in every aspect, as long as they capture the key features of the decision context.
+Research into decision theory, such as the work by @Schaefer1954; @Clark1974; @Reed1979; @Memarzadeh2018 help us better understand that context.
+Adaptive management approaches [@Walters1978] can apply that theory to compare management outcomes between models directly, rather than comparing to forecast performance.
 

### 2020-10-17 manuscript/manuscript.Rmd
:pen:

@@
   sims %>%
-  filter(time < 25, reps < 5) %>%
+  filter(time < 25, reps < 5, model != 3) %>%
   ggplot(aes(time, state, col=model, group = interaction(model,reps))) + 

### 2020-10-26 manuscript/appendix.Rmd
:pen:

@@
-## Decision Context
+---
+title: 'Appendix for: "Bad Forecast, Good Decision..."'
+date: "`r Sys.Date()`"
+author: Carl Boettiger
+journal: Appendix
+layout: 3p
+bibliography: refs.bib
+#output:
+#  tufte::tufte_handout:
+#    dev: cairo_pdf
+#    latex_engine: xelatex
+output: 
+  hrbrthemes::ipsum_pdf:
+    dev: cairo_pdf
+    latex_engine: xelatex
 
-and the manager seeks to maximize the sum of the utility derived from such a harvest and such a state, $U(X_t,H_t)$, over all time, subject to discount rate $\delta$:
+---
+
+```{r graphics_setup, message=FALSE, warning=FALSE, include = FALSE}
+## Plotting themes, colors, and fonts 
+## aesthetics choices only, all can be omitted without consequence
+library(ggthemes)
+library(hrbrthemes)
+library(ggplot2)
+library(Cairo)
+library(extrafont)
+library(patchwork)
+library(styler)
+extrafont::loadfonts(quiet = TRUE)
+ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
+
+scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
+scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
+pal <- ggthemes::solarized_pal()(6)
+txtcolor <- "#586e75"
+
+knitr::opts_chunk$set(cache=FALSE, tidy = "styler", message = FALSE, warning = FALSE, echo = TRUE)
+
+```
+
+Here we provide annotated code necessary to completely reproduce all of the analysis
+presented in the main paper.
+This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving 
+Markov Decision Processes (MDP) using stochastic dynamic programming functionality,
+`expm` for matrix exponentials [@expm], and a few custom MDP functions provided
+by our package, `mdplearning` [@mdplearning].  We will also use `tidyverse` 
+packages for basic manipulation and plotting [@tidyverse]. This file is also
+available as an RMarkdown document [@rmarkdown] at 
+<https://github.com/cboettig/bad-forecast-good-decision>
+
+```{r setup, message=FALSE}
+library(tidyverse)
+library(MDPtoolbox)
+library(expm)
+# remotes::install_github("boettiger-lab/mdplearning")
+library(mdplearning)
+```
+
+Our decision problem is defined as follows: 
+A manager seeks to maximize the sum of the utility derived from such a harvest
+and such a state, $U(X_t,H_t)$, over all time, subject to discount rate $\delta$:
 
@@
 
-```{r prob_scores, message = FALSE}
-## Probability scores instead of proper scores
-scores <- sims %>%
-  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
-  filter(model < 3) %>%
-  rowwise() %>%
-  mutate(prob =  transitions[[model]][state_index, next_state, action_index] ) %>% 
-  group_by(reps, model) %>%
-  summarise(score = sum(log(prob)))
+To solve the decision problem using SDP, we define
+the state space and the action space on a discrete grid of 240 points.  The maximum
+state is set well above the largest carrying capacity used in the models,
+which limits the influence of boundary effects introduced by the transformation to
+a discrete, finite grid.  Available harvest actions match the state space, effectively 
+allowing any harvest level to be possible.
 
-score_table <- scores %>% group_by(model) %>% summarise(mean_score = mean(score))
+```{r model_definitions}
+states <- seq(0,24, length.out = 240)
+actions <- states
+obs <- states
+```
 
-scores %>% mutate(model = as.factor(model)) %>% 
-  ggplot(aes(x = score, group = model, fill = model)) +
-  geom_histogram(binwidth = 70)
+The utility (reward) of an action is set to the (realized) harvest (e.g.
+a fixed price per unit fish, with no cost applied to harvest effort).  Future
+utility is discounted by fixed factor of $0.99$. Classic maximum sustainable 
+yield models [@Schaefer1954] ignore discounting, while modern economic optimization
+models insist on it, so a small discount conforms to the latter while reasonably
+approximating the former.  The qualitative conclusion is not sensitive to the 
+discounting rate.
+
+```{r}
+reward_fn <- function(x,h) pmin(x,h)
+discount <- 0.99
+```
+
+Alternate reward functions with varying price and cost structures are possible but do not qualitatively
+impact the conclusions.  This reward function is a limiting case of any more 
+complex reward, and corresponds both to classic work that does not model utility
+explicitly [e.g. MSY theory, @Schaefer1954], as well as explicit assumptions
+typically made in more recent models [@Reed1979]. Moreover, using a simple reward function
+makes it clear that model that performs best is not doing so merely because of particular
+features baked into the a carefully chosen reward rule.
+
+# Ecological Models
+
+The manager chooses between two different logistic growth models, each of which
+can be thought of as an approximation to the underlying "true" population model:
+
+\begin{equation}
+f_i(Y) = Y + Y r_i \left( 1 - \frac{Y}{K_i} \right)
+\end{equation}
+
+Model 1 has $r_1 = 2$, $K = 16$, and $\sigma_1 = 0.05$.
+Model 2 has $r_2 = 0.5, K_2 = 10$, and $\sigma_2 = 0.075$.
+
+Meanwhile, the true population growth rate is simulated using a function with
+non-linear per-capita growth:
+
+\begin{equation}
+f_i(Y) = Y + Y^4 r_i \left( 1 - \frac{Y}{K_i} \right)
+\end{equation}
+
+with $r_3 = 0.002$, $K_3 = 10$ and $\sigma_3 = 0.05$:
+
+```{r}
+# K is at twice max of f3; 8 * K_3 / 5
+f1 <- function(x, h = 0, r = 2, K = 10 * 8 / 5){
+  s <- pmax(x - h, 0)
+  s + s * (r * (1 - s / K) )
+}
+f2 <- function(x, h = 0, r = 0.5, K = 10){
+  s <- pmax(x - h, 0)
+  s + s * (r * (1 - s / K) )
+}
+
+# max is at 4 * K / 5 
+f3  <- function(x, h = 0, r = .002, K = 10){
+  s <- pmax(x - h, 0)
+  s + s ^ 4 * r * (1 - s / K)
+}
+
+## gather models together, indicate true model
+sigma_g <- 0.05
+models <- list(f1 = f1, f2 = f2, f3 = f3)
+model_sigmas <- c(sigma_g, 1.5 * sigma_g, sigma_g)
+true_model <- "f3"
+```
+
+On a discrete grid of possible states and actions, we can define the growth rate of a given state $X_t$ subject to harvest $H_t$,
+$f(X_t,H_t)$ as set of matrices.  Each matrix $i$ gives the transition probabilities for any current state to any future state, 
+given that action $i$ is taken.
+
+```{r transition_matrices}
+transition_matrices <- function(f, states, actions, sigma_g){
+  n_s <- length(states)
+  n_a <- length(actions)
+  transition <- array(0, dim = c(n_s, n_s, n_a))
+  for (k in 1:n_s) {
+    for (i in 1:n_a) {
+      nextpop <- f(states[k], actions[i])
+      if(nextpop <= 0){
+        transition[k, , i] <- c(1, rep(0, n_s - 1))
+      } else if(sigma_g > 0){
+        x <- dlnorm(states, log(nextpop), sdlog = sigma_g)
+        if(sum(x) == 0){ ## nextpop is computationally zero
+          transition[k, , i] <- c(1, rep(0, n_s - 1))
+        } else {
+          x <- x / sum(x) # normalize evenly
+          transition[k, , i] <- x
+        }
+      }
+    }
+  }
+  transition
+}
+```
+
+This follows the standard setup for standard stochastic dynamic programming, see @Marescot2013.
+Having defined a function to compute the transition matrix, we can use it to create matrices corresponding to each of the three models:
+
+```{r}
+transitions <- lapply(seq_along(models), 
+                      function(i) transition_matrices(models[[i]], 
+                                                      states, 
+                                                      actions, 
+                                                      model_sigmas[[i]]))
+names(transitions) <- c("f1", "f2", "f3")
 
 ```
+
+Likewise, a corresponding matrix defining the rewards associated with each state $X$ and each harvest action $H$ can also be defined.
+
+````{r}
+## Compute reward matrix (shared across all models)
+n_s <- length(states)
+n_a <- length(actions)
+reward <- array(0, dim = c(n_s, n_a))
+for (k in 1:n_s) {
+  for (i in 1:n_a) {
+    reward[k, i] <- reward_fn(states[k], actions[i])
+  }
+}
+```
+
+# Optimal control solutions
+
+We use value iteration to solve the stochastic dynamic program [@Marescot2013; @MDPtoolbox] for each model.  
+This determines the optimal harvest policy for each possible state, given each model. Because this step is the
+most computationally intensive routine, we cache the results using memosization conditioned on the transition
+matrices [@memoise].  Running this code with alternate transition matrices automatically invalidates that
+cache, reducing the risk of loading spurious results.
+
+```{r sdp, results="hide"}
+mdp <- memoise::memoise(mdp_value_iteration,
+                        cache = memoise::cache_filesystem("cache/"))
+
+policies <- 
+  map_dfr(transitions, 
+          function(P){
+    soln <- mdp(P, reward, discount = discount, 
+                epsilon = 0.01, max_iter = 1000, V0 = rep(0,dim(P)[[1]]))
+    escapement <- states - actions[soln$policy]
+    tibble(states, policy = soln$policy, escapement)
+    }, 
+  .id = "model")
+```
+
+# Simulations and step-ahead forecasts
+
+We simulate fishing dynamics under the optimal policy for each model, using a simple
+helper function from the `mdplearning` package. Because growth dynamics are stochastic,
+we perform 100 simulations of each model from identical starting condition to
+ensure results are not the result of chance alone.
+
+```{r simulations}
+library(mdplearning)
+Tmax <- 100
+x0 <- which.min(abs(states - 6))
+reps <- 100
+set.seed(12345)
+
+## Simulate each policy reps times, with `f3` as the true model:
+simulate_policy <- function(i, policy){
+  mdp_planning(transitions[[true_model]], reward, discount,
+           policy = policy, x0 = x0, Tmax = Tmax) %>%
+    select(value, state_index = state, time, action_index = action)  %>% 
+    mutate(state = states[state_index])
+}
+
+sims <- 
+  map_dfr(names(transitions), 
+          function(m){
+            policy <- policies %>% filter(model == m) %>% pull(policy)
+            map_dfr(1:reps, simulate_policy, policy = policy, .id = "reps")
+          },
+          .id = "model"
+         )
+```
+
+Using the transition matrices directly, we can examine what each model would 
+have forecast the future stock size to be in the following year when no fishing
+occurs (note that for each model, we use the transition matrix that corresponds
+to 'no fishing', `model[[state_index, ,1]]`) (Fig 1a, main text).
+
+The transition matrices give the full (discretized) probability distribution,
+from which we can easily calculate both the expected value and the 95% 
+confidence interval.
+
+```{r stepahead_unfished}
+stepahead_unfished <- sims
+stepahead_unfished$state_index <- rep(sims$state_index[sims$model == "1"],3)
+
+stepahead_unfished <- stepahead_unfished  %>% 
+  filter(model != "3") %>% 
+  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
+  rowwise() %>%
+  mutate(expected = transitions[[model]][state_index, , 1]  %*% states,
+         var = transitions[[model]][state_index, , 1]  %*% states ^ 2 - expected ^ 2,
+         low = states[max(which(cumsum(transitions[[model]][state_index,,1]) < 0.025)) ],
+         high = states[min(which(cumsum(transitions[[model]][state_index,,1]) > 0.975)) ],
+         true = states[next_state ],
+         model = paste0("f", model)) 
+```
+
+We also look at the forecast each model makes when implementing the corresponding
+optimal harvest:
+
+```{r stepahead_fished}
+stepahead_fished <- sims %>% 
+  filter(model != "3") %>%
+  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
+  rowwise() %>%
+  mutate(prob =  transitions[[model]][state_index, next_state, action_index],
+         expected = transitions[[model]][state_index, , action_index]  %*% states,
+         var = transitions[[model]][state_index, , action_index]  %*% states ^ 2 - expected ^ 2,
+         low = states[max(which(cumsum(transitions[[model]][state_index,,action_index]) < 0.025)) ],
+         high = states[min(which(cumsum(transitions[[model]][state_index,,action_index]) > 0.975)) ],
+         true = states[next_state],
+         model = paste0("f", model)) %>%
+ select(time, model, true, expected, low, high, var, prob, reps)
+```
+
+# Proper scores
+
+It is straight forward to apply the proper scoring formula of @Gneiting2007 
+based on the first two moments of the distribution to score the respective
+forecasts under both the unfished and actively managed scenario sfor each model:
+
+```{r proper_scores, message = FALSE}
+# Gneiting & Raferty (2007), eq24
+scoring_fn <- function(x, mu, sigma){ -(mu - x )^2 / sigma^2  - log(sigma)}
+
+stepahead_unfished <- stepahead_unfished %>%
+  mutate(sd = sqrt(var),
+         score = scoring_fn(expected, true, sd))
+
+stepahead_fished <- stepahead_fished %>%
+  mutate(sd = sqrt(var),
+         score = scoring_fn(expected, true, sd))
+```
+
+```{r}
+predictions <- 
+  stepahead_unfished %>% 
+  select(time, model, reps, expected, low, high, true, score) %>% 
+  mutate(scenario = "A_unfished") %>% 
+  bind_rows(stepahead_fished  %>% 
+          select(time, model, reps, expected, low, high, true, score) %>% 
+          mutate(scenario = "B_fished")
+  ) 
+```
+
+# Plotting
+
+Having computed a single data frame of simulation data under both optimally fished
+(according to each model) and unfished scenarios for each model, along with
+the proper scores for the step-ahead forecasts a la @Gneiting2007, we have all 
+the results necessary to generate the figures presented in the main paper.
+
+```{r fig1a}
+fig1ab <- predictions %>%
+  filter(reps == "2", time < 10) %>%
+  ggplot(aes(time, col = model, fill = model)) + 
+  geom_point(aes(y = expected)) + 
+  geom_errorbar(aes(ymin = low, ymax = high)) +
+  geom_point(aes(y = true), pch = "*", size = 12, 
+             alpha = 0.6) + 
+  facet_wrap(~scenario, ncol = 1, labeller =as_labeller(c(
+    A_unfished = "A. Without fishing",
+    B_fished = "B. Managed harvest"))) +
+  ylab("stock size")
+
+```
+
+Likewise we can plot the data on proper scores associated with
+each prediction of each model:
+
+```{r}
+fig1cd <- predictions %>%
+  ggplot(aes(x = score, group = model, fill = model)) +
+  geom_histogram(binwidth = 2, show.legend = FALSE) +
+  coord_cartesian(xlim = c(-100, 1), ylim = c(0,4000)) +
+  xlab("Proper score") +
+  facet_wrap(~scenario, ncol = 1, labeller =as_labeller(c(
+    A_unfished = "C",
+    B_fished = "D"))) + 
+  scale_x_continuous(breaks = c(-100, 0)) +
+  theme(axis.text.y = element_blank(),
+        plot.margin = margin(0, 0, 0, 0, "cm"),
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank()
+        )
+
+```
+
+## Forecast ecological and economic performance
+
+The ecological performance of the model can easily be visualized by plotting
+the results of each simulation.
+
+```{r plot_sims}
+fig2a <- 
+  sims %>%
+  filter(time < 25, reps < 5, model != 3) %>%
+  ggplot(aes(time, state, col=model, group = interaction(model,reps))) + 
+  geom_line(alpha=0.3) 
+```
+
+To plot the economic value over time, we must sum up the discounted values
+at each time step, and then average over replicate simulations of each model:
+
+```{r plot_npv}
+##  Net Present Value accumulates over time
+npv_df <- sims %>% 
+  group_by(model, reps) %>%
+  mutate(npv = cumsum(value * discount ^ time)) %>%
+  group_by(time, model)  %>% 
+  summarise(mean_npv = mean(npv), .groups="drop") %>% 
+  arrange(model, time)
+
+optimal <- select(filter(npv_df, model == "3"), time, mean_npv)
+
+fig2b <- 
+npv_df %>%
+  filter(model != "3", time %in% seq(1,100, by = 5)) %>%
+  ggplot(aes(time, mean_npv)) +
+  geom_line(data = optimal, lwd = 1.5, col = "grey20") +
+  geom_point(aes(col=model), size = 4, alpha = 0.8) + 
+  ylab("Net present value") + xlab("time")
+
+```
+
+Plotting the model functions themselves are shown in Fig 3a:
+
+```{r plot_models}
+d <- 
+  map_dfc(models, function(f) f(states) - states) %>%
+  mutate(state = states)
+
+fig3a <- 
+  d %>% pivot_longer(names(models), "model") %>%
+  ggplot(aes(state, value, col=model)) +
+  geom_point() + 
+  geom_hline(aes(yintercept = 0)) + 
+  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16))
+```
+
+We can easily plot the optimal policy derived from each model, as shown
+in Fig3b.
+
+```{r plot_policies}
+fig3b <- policies %>%
+  ggplot(aes(states, escapement, col=model, lty=model)) + 
+  geom_line(lwd=2)
+```
+
+# References

### 2020-10-26 manuscript/manuscript.Rmd
:pen:

--- 2020-10-16 manuscript/manuscript.Rmd
@@
 As forecasting becomes an increasing attractive and viable possibility for assessing ecological models and applying them to real-world decisions, I hope this illustration serves as a reminder that modeling can not be done in a vacuum, but must reflect the context in which the model will be applied.
-journal: Theoretical ecology
+journal: TBD
 layout: 3p
 bibliography: refs.bib
-output: rticles::elsevier_article
+output: 
+  rticles::elsevier_article:
+    dev: cairo_pdf
+    keep_tex: true
 keywords: 
@@
   \journal{preprint}
-  ## EditorialManager adds its own linenumbers to pdf already
   \usepackage{lineno}
@@
 
-```{r graphics_setup, message=FALSE, warning=FALSE}
-knitr::opts_chunk$set(cache=TRUE)
-
-## Plotting theme settings -- aesthetics only, all can be omitted without consequence
+```{r, message=FALSE, warning=FALSE, include = FALSE}
+## Plotting themes, colors, and fonts 
+## aesthetics choices only, all can be omitted without consequence
 library(ggthemes)
@@
 library(patchwork)
-extrafont::loadfonts()
+library(styler)
+extrafont::loadfonts(quiet = TRUE)
 ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
-#theme_set(theme_solarized(base_size=16))
 
@@
 
-```{r setup, message=FALSE}
-## Required computational libraries
-library(tidyverse)
-library(MDPtoolbox)
-library(expm)
-library(mdplearning) # remotes::install_github("boettiger-lab/mdplearning")
-```
-
-```{r model_definitions}
-states <- seq(0,24, length.out = 240)
-actions <- states
-obs <- states
-sigma_g <- 0.05
-reward_fn <- function(x,h) pmin(x,h)
-discount <- 0.99
-
-# K is at twice max of f3; 8 * K_3 / 5
-f1 <- function(x, h = 0, r = 2, K = 10 * 8 / 5){
-  s <- pmax(x - h, 0)
-  s + s * (r * (1 - s / K) )
-}
-f2 <- function(x, h = 0, r = 0.5, K = 10){
-  s <- pmax(x - h, 0)
-  s + s * (r * (1 - s / K) )
-}
-
-# max is at 4 * K / 5 
-f3  <- function(x, h = 0, r = .002, K = 10){
-  s <- pmax(x - h, 0)
-  s + s ^ 4 * r * (1 - s / K)
-}
-
-# looks even closer to f2; not used
-#f4  <- function(x, h = 0, r = 4e-7, K = 10){
-#  s <- pmax(x - h, 0)
-#  s + s ^ 8 * r * (1 - s / K) + s * (1 - s / K)
-#}
-
-models <- list(f1 = f1, f2 = f2, f3 = f3)
-model_sigmas <- c(sigma_g, 1.5 * sigma_g, sigma_g)
-
-true_model <- "f3"
+```{r, include=FALSE}
+rmarkdown::render("appendix.Rmd")
+knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, warning = FALSE, dev="cairo_pdf")
 ```
@@
 
-```{r transition_matrices}
-# A function to compute the transition matrices for each model:
-transition_matrices <- function(f,
-                      states,
-                      actions,
-                      sigma_g){
-
-  n_s <- length(states)
-  n_a <- length(actions)
-
-  transition <- array(0, dim = c(n_s, n_s, n_a))
-  for (k in 1:n_s) {
-    for (i in 1:n_a) {
-      nextpop <- f(states[k], actions[i])
-      if(nextpop <= 0){
-        transition[k, , i] <- c(1, rep(0, n_s - 1))
-      } else if(sigma_g > 0){
-        x <- dlnorm(states, log(nextpop), sdlog = sigma_g)
-        if(sum(x) == 0){ ## nextpop is computationally zero
-          transition[k, , i] <- c(1, rep(0, n_s - 1))
-        } else {
-          x <- x / sum(x) # normalize evenly
-          transition[k, , i] <- x
-        }
-      } else {
-        stop("sigma_g not > 0")
-      }
-      reward[k, i] <- reward_fn(states[k], actions[i])
-    }
-  }
-  transition
-}
-
-## Compute reward matrix (shared across all models)
-n_s <- length(states)
-n_a <- length(actions)
-reward <- array(0, dim = c(n_s, n_a))
-for (k in 1:n_s) {
-  for (i in 1:n_a) {
-    reward[k, i] <- reward_fn(states[k], actions[i])
-  }
-}
-
-## Now compute the transition matrices for each model
-transitions <- lapply(seq_along(models), 
-                      function(i) transition_matrices(models[[i]], 
-                                                      states, 
-                                                      actions, 
-                                                      model_sigmas[[i]]))
-names(transitions) <- c("f1", "f2", "f3")
-
-```
-
-```{r sdp, results="hide"}
-## This is the most (only) computationally expensive code chunk. use memoization to cache
-mdp <- memoise::memoise(mdp_value_iteration, cache = memoise::cache_filesystem("cache/"))
-
-policies <- 
-  map_dfr(transitions, 
-          function(P){
-    soln <- mdp(P, reward, discount = discount, 
-                epsilon = 0.01, max_iter = 1000, V0 = rep(0,dim(P)[[1]]))
-    escapement <- states - actions[soln$policy]
-    tibble(states, policy = soln$policy, escapement)
-    }, 
-  .id = "model")
-```
-
-```{r simulations}
-library(mdplearning)
-Tmax <- 100
-x0 <- which.min(abs(states - 6))
-reps <- 100
-set.seed(12345)
-
-## Simulate each policy reps times, with `f3` as the true model:
-
-sims <- map_dfr(names(transitions), 
-                function(m){
-                  policy <- policies %>% filter(model == m) %>% pull(policy)
-                  map_dfr(1:reps, 
-                          function(i){
-                            mdp_planning(transitions[[true_model]], reward, discount,
-                                     policy = policy, x0 = x0, Tmax = Tmax) %>%
-                              select(value, state_index = state, time, action_index = action)  %>% 
-                              mutate(state = states[state_index]) # index->value
-                            },
-                          .id = "reps")
-                },
-                .id = "model")
-
-```
-
-```{r stepahead_unfished}
-stepahead_unfished <- sims
-stepahead_unfished$state_index <- rep(sims$state_index[sims$model == "1"],3)
-stepahead_unfished <- stepahead_unfished  %>% 
-  filter(model != "3") %>% 
-  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
-  rowwise() %>%
-  mutate(expected = transitions[[model]][state_index, , 1]  %*% states,
-         var = transitions[[model]][state_index, , 1]  %*% states ^ 2 - expected ^ 2,
-         low = states[ max(which(cumsum(transitions[[model]][state_index,,1]) < 0.025)) ],
-         high = states[ min(which(cumsum(transitions[[model]][state_index,,1]) > 0.975)) ],
-         true = states[ next_state ],
-         model = as.character(model)) 
-```
-
-```{r fig1a}
-fig1a <- stepahead_unfished %>% 
-  filter(reps == 2, time < 10) %>%
-  ggplot(aes(time, col = model, fill = model)) + 
-  geom_point(aes(y = expected), show.legend = FALSE) + 
-  geom_errorbar(aes(ymin = low, ymax = high), show.legend = FALSE) +
-  geom_point(aes(y = true), pch = "*", size = 12, col = "grey40", show.legend = FALSE)
-
-```
-
-```{r stepahead_fished}
-stepahead_fished <- sims %>% 
-  filter(model != "3") %>%
-  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
-  rowwise() %>%
-  mutate(prob =  transitions[[model]][state_index, next_state, action_index],
-         expected = transitions[[model]][state_index, , action_index]  %*% states,
-         var = transitions[[model]][state_index, , action_index]  %*% states ^ 2 - expected ^ 2,
-         low = states[ max(which(cumsum(transitions[[model]][state_index,,action_index]) < 0.025)) ],
-         high = states[ min(which(cumsum(transitions[[model]][state_index,,action_index]) > 0.975)) ],
-         true = states[ next_state],
-         model = as.character(model)) %>%
- select(time, model, true, expected, low, high, var, prob, reps)
-```
-
-```{r fig1b}
-fig1b <- stepahead_fished  %>%  
-  filter(reps == 3, time < 10) %>%
-  ggplot(aes(time, col = model, fill = model)) + 
-  geom_point(aes(y = true), pch = "*", size = 12) +
-  geom_point(aes(y = expected)) + 
-  geom_errorbar(aes(ymin = low, ymax = high))
- # geom_errorbar(aes(ymin = expected - 2 * sqrt(var), ymax=expected + 2 * sqrt(var)))
-
-```
-
-```{r proper_scores, message = FALSE}
-
-scoring_fn <- function(x, mu, sigma){ -(mu - x )^2 / sigma^2  - log(sigma)}
-
-proper_scores_unfished <- stepahead_unfished %>%
-  mutate(sd = sqrt(var),
-         score = scoring_fn(expected, true, sd)) %>%
-  select(model, reps, time, score)
-
-fig1c <- proper_scores_unfished %>%
-  ggplot(aes(x = score, group = model, fill = model)) +
-  geom_histogram(binwidth = 2, show.legend = FALSE) +
-  coord_cartesian(xlim = c(-100, 1))
-
-proper_scores_fished <- stepahead_fished %>%
-  mutate(sd = sqrt(var),
-         score = scoring_fn(expected, true, sd)) %>%
-  select(model, reps, time, score)
-
-fig1d <- proper_scores_fished %>%
-  ggplot(aes(x = score, group = model, fill = model)) +
-  geom_histogram(binwidth = 2) +
-  coord_cartesian(xlim = c(-100, 1))
-
-```
-
 # Results
@@
 
-```{r figure1, fig.width=7, fig.height=8}
-(fig1a / fig1b) /
-(fig1c + fig1d)
-
+```{r figure1, fig.width=7, fig.height=5, include=TRUE, message=TRUE, warning=TRUE}
+fig1ab + fig1cd + 
+  plot_layout(widths =  c(3,1), guides = 'collect') + 
+  theme(plot.margin = margin(0, 0, .1, .1, "cm"))
 ```
@@
 
-```{r plot_sims}
-fig2a <- 
-  sims %>%
-  filter(time < 25, reps < 5) %>%
-  ggplot(aes(time, state, col=model, group = interaction(model,reps))) + 
-  geom_line(alpha=0.3) 
-```
-
-```{r plot_npv}
-##  Net Present Value accumulates over time, equivalent for models with near-identical management stategy
-npv_df <- sims %>% 
-  group_by(model, reps) %>%
-  mutate(npv = cumsum(value * discount ^ time)) %>%
-  group_by(time, model)  %>% 
-  summarise(mean_npv = mean(npv)) %>% 
-  arrange(model, time) %>% ungroup()
-
-optimal <- select(filter(npv_df, model == "3"), time, mean_npv)
-
-fig2b <- 
-npv_df %>%
-  filter(model != "3", time %in% seq(1,100, by = 5)) %>%
-  ggplot(aes(time, mean_npv)) +
-  geom_line(data = optimal, lwd = 1.5, col = "grey20") +
-  geom_point(aes(col=model), size = 4, alpha = 0.8) + 
-  ylab("Net present value") + xlab("time")
-
-```
 ```{r figure2, message = FALSE, fig.cap = "Corresponding utility (measured as mean net present value, that is: cumulative value, discounting future values by $\\delta^t$, averaged across replicates).
@@
 
-```{r plot_models}
-d <- map_dfc(models, function(f) f(states) - states) %>% mutate(state = states)
-
-fig3a <- 
-  d %>% pivot_longer(names(models), "model") %>%
-  ggplot(aes(state, value, col=model)) +
-  geom_point() + 
-  geom_hline(aes(yintercept = 0)) + 
-  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16))
-```
-
-```{r plot_policies}
-fig3b <- policies %>%
-  ggplot(aes(states,escapement, col=model, lty=model)) + geom_line(lwd=2)
-```
-
 ```{r figure3, fig.width=5, fig.height=7}

### 2020-10-26 scratch.Rmd
:pen:

```{r prob_scores, message = FALSE}
## Probability scores instead of proper scores
scores <- sims %>%
  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
  filter(model < 3) %>%
  rowwise() %>%
  mutate(prob =  transitions[[model]][state_index, next_state, action_index] ) %>% 
  group_by(reps, model) %>%
  summarise(score = sum(log(prob)))

score_table <- scores %>% group_by(model) %>% summarise(mean_score = mean(score))

scores %>% mutate(model = as.factor(model)) %>% 
  ggplot(aes(x = score, group = model, fill = model)) +
  geom_histogram(binwidth = 70)
```

### 2020-10-27 manuscript/appendix.Rmd
:pen:

@@
 sigma_g <- 0.05
-models <- list(f1 = f1, f2 = f2, f3 = f3)
+models <- list("1" = f1, "2" = f2, "3" = f3)
 model_sigmas <- c(sigma_g, 1.5 * sigma_g, sigma_g)
-true_model <- "f3"
+true_model <- "3"
 ```
@@
                                                       model_sigmas[[i]]))
-names(transitions) <- c("f1", "f2", "f3")
+names(transitions) <- c("1", "2", "3")
 
@@
 
-## Simulate each policy reps times, with `f3` as the true model:
+## Simulate each policy reps times, with `3` as the true model:
 simulate_policy <- function(i, policy){
@@
          high = states[min(which(cumsum(transitions[[model]][state_index,,1]) > 0.975)) ],
-         true = states[next_state ],
-         model = paste0("f", model)) 
+         true = states[next_state ]) 
 ```
@@
          high = states[min(which(cumsum(transitions[[model]][state_index,,action_index]) > 0.975)) ],
-         true = states[next_state],
-         model = paste0("f", model)) %>%
+         true = states[next_state]) %>%
  select(time, model, true, expected, low, high, var, prob, reps)
@@
   facet_wrap(~scenario, ncol = 1, labeller =as_labeller(c(
-    A_unfished = "C",
-    B_fished = "D"))) + 
+    A_unfished = "C.",
+    B_fished = "D."))) + 
   scale_x_continuous(breaks = c(-100, 0)) +
@@
   sims %>% 
-  filter(time < 25, reps < 5, model != 3) %>%
-  ggplot(aes(time, state, col=model, group = interaction(model,reps))) + 
-  geom_line(alpha=0.3) 
+  group_by(model, time) %>%
+  summarise(mean_state = mean(state), sd = sd(state), .groups = "drop") %>%
+  filter(time < 25, model != "3") %>%
+  ggplot(aes(time, mean_state)) + 
+  geom_line(aes(col = model), lwd=1.5, show.legend = FALSE) +
+    geom_ribbon(aes(ymin = mean_state - 2*sd, 
+                    ymax = mean_state + 2*sd,
+                    fill = model),
+                alpha = 0.2, show.legend = FALSE) +
+  ylab("state") + 
+  ggtitle("Ecological performance")
 ```
@@
   geom_point(aes(col=model), size = 4, alpha = 0.8) + 
-  ylab("Net present value") + xlab("time")
+  ylab("Net present value") + xlab("time")  + 
+  ggtitle("Economic performance")
 
@@
   d %>% pivot_longer(names(models), "model") %>%
-  ggplot(aes(state, value, col=model)) +
-  geom_point() + 
-  geom_hline(aes(yintercept = 0)) + 
-  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16))
+  ggplot(aes(state, value, col=model, lty=model)) +
+  geom_hline(aes(yintercept = 0), lwd=1) + 
+  geom_line(lwd=2, show.legend = FALSE) + 
+  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
+  ylab(bquote(f(x))) + xlab("x")
 ```
@@
   ggplot(aes(states, escapement, col=model, lty=model)) + 
-  geom_line(lwd=2)
+  geom_line(lwd=2) + xlab("state")
 ```

### 2020-10-27 manuscript/manuscript.Rmd
:pen:

@@
 
-```{r figure1, fig.width=7, fig.height=5, include=TRUE, message=TRUE, warning=TRUE}
+```{r figure1, fig.width=7, fig.height=5, include=TRUE, message=TRUE, warning=TRUE,, fig.cap = "Forecast performance of each model."}
 fig1ab + fig1cd + 
@@
 
-```{r figure2, message = FALSE, fig.cap = "Corresponding utility (measured as mean net present value, that is: cumulative value, discounting future values by $\\delta^t$, averaged across replicates).
-Note that the exepcted utility under model 1, which has the worst forecast, is nearly identical to the optimal utility achieved by managing under the correct model, 3.
-The utility derived from model 2 is far smaller, despite it's overall better performance in long term forecasts."}
+```{r figure2, message = FALSE, fig.cap = "Ecological and economic performance of each forecast."}
 fig2a + fig2b
@@
 
-```{r figure3, fig.width=5, fig.height=7}
-fig3a / fig3b
+```{r figure3, fig.width=7, fig.height=4}
+fig3a + fig3b
 ```

### 2020-11-05 manuscript/appendix.Rmd
WIP

@@
           mutate(scenario = "B_fished")
-  ) 
+  ) %>% 
+  mutate(model = as.character(model))
 ```
@@
   ylab("state") + 
-  ggtitle("Ecological performance")
+  ggtitle("A") + labs(subtitle = "Ecological performance")
 ```
@@
   ylab("Net present value") + xlab("time")  + 
-  ggtitle("Economic performance")
+  ggtitle("B.") + labs(subtitle = "Economic performance")
 

### 2020-11-05 manuscript/manuscript.Rmd
WIP

@@
     affiliation: ucb
-#    footnote: Corresponding Author
 address:
@@
 abstract: |
-    With the rapid expansion of available data, it is frequently suggested that _forecasting performance_ is the ultimate test of a scientific model [e.g.
-@Dietz2018].
+    With the rapid expansion of available data, it is frequently suggested that _forecasting performance_ is the ultimate test of a scientific model.
 Yet the model that makes the most accurate forecasts is not always the model that yields the best decisions.
-This is no accident of statistics, even formal criteria such as strictly proper scoring rules [@Gneiting2007] cannot determine the best model for decision-making because they do not consider the context of the decision process itself.
+This is no accident of statistics, even formal criteria such as strictly proper scoring rules cannot determine the best model for decision-making because they do not consider the context of the decision process itself.
 I illustrate this paradox using a classic example of ecological decision making in fisheries management.
@@
 bibliography: refs.bib
+csl: nature.csl
 output: 
@@
 
-A primary purpose of statistical analyses and ecological modeling is to make forecasts for the future [].
-Accurate forecasts are important both in assessing the accuracy of our models and understanding of natural processes, and can also underpin policy and decision making.
+A primary purpose of statistical analyses and ecological modeling is to make forecasts for the future [@Clark2001; @Gneiting2014].
+Accurate forecasts are important both in assessing the accuracy of our models and understanding of natural processes, they also underpin policy and decision making [@Dietze2018].
 Yet the model that leads to the best decisions is not always the model that makes the most accurate forecasts, as I illustrate here.
@@
 In choosing the best model for decision-making, it can be more important to capture a single key feature of the process than it is to make the most accurate prediction about future states.
-Even strictly proper scoring rules for probablistic forecasts [@Gneiting2007] will not always select the best model to guide decision-making.
-Surprisingly, this is true not only in conceptual models that may guide policy choices, but also when a decision policy is derived directly from a complex optimization routine of a probablistic predictive model, such as the stochastic dynamic programming (SDP) algorithms used to solve Markov Decision Processes [@Marescot2013].
-Here, I use a classical, well-understood example from fisheries management [@Schaefer1954; @Clark1974; @Reed1979] to illustrate both the paradox of how a model with the worst forecast provides the best decision outcomes, as well as show how we can avoid selecting models that are poorly suited for management by considering the management context more explicitly.
+Even strictly proper scoring rules for probabilistic forecasts [@Gneiting2007] will not always select the best model to guide decision-making.
+Surprisingly, this is true not only in conceptual models that may guide policy choices, but also when a decision policy is derived directly from a complex optimization routine of a probabilistic predictive model, such as the stochastic dynamic programming (SDP) algorithms used to derive optimal policies to sequential decision problems [@Marescot2013].
+Here, I use a classical, well-understood example from fisheries management [@Schaefer1954; @Clark1973; @Reed1979] to illustrate both the paradox of how a model with the worst forecast provides the best decision outcomes, as well as show how we can avoid selecting models that are poorly suited for management by considering the management context more explicitly.
 
@@
 Fisheries are a significant economic and conservation concern world wide and their management remains an important debate [e.g.
-@Worm2006; @Costello2016].
+@Worm2006; @Worm2009; @Costello2016].
 Moreover, their management has been both a proving grounds for theoretical and practical decision-making issues which are widely applicable in other areas of ecology and conservation [e.g.
 @Mangel1985; @Clark1990; @Marescot2013].
-The decision-making problem is characterized by the need for a manager to set an acceptable harvest quota $H_t$ each year given some stock assessment estimate of the current stock size (population abundance) of the species in question [@Clark1974].
+The decision-making problem is characterized by the need for a manager to set an acceptable harvest quota $H_t$ each year given some stock assessment estimate of the current stock size (population abundance) of the species in question [@Clark1973].
 Such a decision problem appears to hinge on an accurate forecast: if we can predict to what size the stock will increase next year, $X_t+1$, knowing the current stock, $X_t$, then we can safely harvest $X_{t+1} - X_t$.
@@
 These are not necessary assumptions -- in fact, the more complex the models become, the easier it is find examples in which the best forecast does not produce the best decision.
-Rather, using a simple model merely reflects the famous compromise of Richard Levins [@Levins1969] in choosing generality over precision.
+Rather, using a simple model merely reflects the famous compromise of Richard Levins [@Levins1966] in choosing generality over precision.
 More precisely, the decision problem in question can be stated as follows: The fish stock is observed to be in state `X_t` at time `t`, and is then subjected to some harvest `H_t` before recruiting new fish, subject to stochastic environmental noise `\xi_t`, to bring the stock to $X_t+1$,
@@
 \begin{equation}
-f_i(Y) = Y + Y r_i \left( 1 - \frac{Y}{K_i} \right)
+f_i(Y) = Y + r_i Y \left( 1 - \frac{Y}{K_i} \right)
 \end{equation}
 
-Model 1 is given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, Model 2 by $r_2 = 0.5$,  $K_2 = 10$, $\sigma = 0.075$.
+Model 1 is given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, Model 2 by $r_2 = 0.5$,  $K_2 = 10$, $\sigma = 0.075$ (in non-dimensionalized units).
 Having both the larger growth rate and the larger carrying capacity, Model 1 is clearly the more optimistic of the two choices.
@@
 Ecological processes are much too complex to ever be modeled exactly down to the last atom.
-For illustrative purposes, we will assume the "true" process to be given by a third model which we will leave unspecified for the moment and will examine in more detail later on.
+For illustrative purposes, we will thus assume the "true" process to be given by a third model, which is unknown to the decision-maker:
+
+\begin{equation}
+f_3(Y) = Y + r_3 Y^4 \left( 1 - \frac{Y}{K_3} \right)
+\end{equation}
+
+with $r_3 = 0.002$, $K_3 = 10$ and $\sigma_3 = 0.05$.
 
@@
 
-```{r figure1, fig.width=7, fig.height=5, include=TRUE, message=TRUE, warning=TRUE,, fig.cap = "Forecast performance of each model."}
+```{r figure1, fig.width=7, fig.height=5, fig.cap = "Forecast performance of each model.
+Panels A, B: Step ahead predictions under unfished (A) and fished (B) scenarios.
+Error bars indicating the 95% confidence intervals around each prediction, while stars denote the observed value in that year.
+Because the models make different decisions each year, the stock size in year 2, 3, etc under management of model 1 (blue stars) is different from that under model 2 (red stars).
+Panels C, D: corresponding distribution of proper scores across all predictions.
+Higher scores are better."}
 fig1ab + fig1cd + 
@@
 
-Figure 1A shows the step-ahead prediction performance of each model in a simulation of an un-fished environment, with error bars indicating the 95% confidence intervals around each prediction, while stars denote the observed value in that year.
+The one-step-ahead prediction performance of each model in a simulation of an un-fished environment show consistently better performance of Model 2 (Fig.
+1A).
 Model 1 predictions appear far too optimistic, with the true value falling well below the 95% confidence intervals.
@@
 
-Predictive performance of the unfished population does not give us the full picture, since it reflects predictive accuracy only in the region of the true carrying capacity, while an actively harvested stock will be at a lower size.
+Predictive performance of the un-fished population does not give us the full picture, since it reflects predictive accuracy only in the region of the true carrying capacity, while an actively harvested stock will be at a lower size.
 The model that predicts the equilibrium size may not be the one that best forecasts stock recovery.
-Further, this comparison does not yet implement any decisions that might be made from either model.
-To address these concerns, we consider the case where our fishery is managed according to the optimal harvest predicted by each model in turn.
+This comparison does not reflect the influence of any decisions that might be made based on the model forecast.
+To address these concerns, we consider a second scenario where our fishery is managed according to the optimal harvest predicted by each model in turn.
 Each year the model produces both a forecast and a decision about the harvest quota.
-(Mechanics of determining a harvest quota given the model follow standard methods, e.g.
-@Reed1979; @Marescot2013, see appendix for details.) We then implement that harvest and compare the observed stock size the following year to that which the model has predicted, Figure 1B.
-Because the models make different decisions each year, the stock size in year 2, 3, etc under management of model 1 (blue stars) is different from that under model 2 (red stars).
+(The mechanics of determining a harvest quota given the model follow standard methods for Markov Decision Process, which depend on step-ahead predictions [@Marescot2013], see appendix for details.) We then implement that harvest and compare the observed stock size the following year to that which the model has predicted, (Fig.
+1B).
 Again we observe that the observations under model 1 consistently fall well outside of the 95% confidence intervals it predicts, while under model 2, stock sizes consistently fall within the predicted intervals.
-Once again, model 2 shows a higher forecast accuracy while model 1 appears hopelessly optimistic.
+Once again, model 2 shows a higher forecast accuracy while model 1 appears consistently over-optimistic.
 
-**Proper scores**.
+Interest in assessing probabilistic forecasts has led to the development of rigorous methods of scoring [e.g.
+@Gneiting2014; @Raftery2016].
+A scoring rule is "proper" if it has the convenient property that no other prediction can achieve a higher score, on average, than we would if we used the true distribution [@Gneiting2007].
+The distribution of proper scores across 100 replicate simulations for both un-fished and actively managed scenarios (Fig.
+1C, 1D respectively) show consistently higher scores of model 2, using a proper scoring rule (Eq (27) of Ref.
+@Gneiting2007).
 
@@
 Given this evidence, model 2 clearly provides the more accurate forecast and we would no doubt conclude that model 2 was thus a better approximation of the true model and thus the better choice to inform decision making about harvest quotas.
-Yet if we revisit our experiment of managing the fishery under each model in turn, and focus not on _predictive accuracy_ but on _ecological_ and _economic_ outcomes, a different story emerges.
-Figure 2A shows the stock sizes across five replicate simulations that have been managed according to each model.
+Yet if we revisit our experiment of managing the fishery under each model in turn, and focus not on _predictive accuracy_ but on _ecological_ and _economic_ outcomes, it quickly becomes clear that model 1 gives much better results (Fig.
+2A).
 For comparison, we have also included the results of optimal management given the true model.
@@
 In contrast, model 2 suppresses the stock to a much lower level.
-The over-fishing in model 2 is not economically efficient either, as shown in Figure 2B.
+The over-fishing in model 2 is not economically efficient either (Fig.
+2B).
 The net present value of the fishery, as calculated as the cumulative, discounted value of the harvest (assuming a fixed unit price for fish with negligible cost for harvest, see appendix) under the fishing regime of model 1 falls precisely along that of the optimal solution, while the value derived under model 2 is consistently lower.
@@
 This paradox in performance of forecasting vs performance in decision making can be easily resolved by considering the context of the decision problem more closely.
-Starting back in the origins of modern fisheries management in the 1950s, @Schaefer1954 demonstrates that the Maximum Sustainable Yield (MSY) is maintained by harvesting a stock down to the size at which it achieves its maximum growth rate, i.e.
-$K/2$ for a logistic model (or any other symmetric growth model).
-The details of this optimal harvest have been refined to allow for dynamic optimization with discounted future profits [@Clark1974] stochastic population growth [@Reed1979], and even imperfect observations [@Memarzadeh2018], but Schaefer's observation is sufficient in this case to see that the decision problem depends very much on the value of $K$ while being essentially independent of the other model parameters, including the growth rate, $r$ and stochasticity, $\sigma$.
-However, the growth rate (and magnitude of the stochasticity) matter very much to forecast accuracy.
-Figure 3 shows the functional form of our two logistic-curve models, compared to the functional form of the "true" model used to drive the simulations.
-From this graph, it is clear to see that Model 2 does indeed lie closer to the true model throughout the state space, and agrees precisely with the carrying capacity (where both functions cross zero with negative slope).
-However, the peak of model 3 very nearly matches the peak of model 1, and falls higher stock size than that of model 2.
+Comparing plots of the functional form of our two logistic-curve models, compared to the functional form of the "true" model used to drive the simulations (Fig.
+3A), it is clear to see that model 2 does indeed lie closer to the true model throughout the state space, agreeing precisely with the true carrying capacity (where both functions cross zero with negative slope).
+However, the peak of model 3 very nearly matches the peak of model 1.
+The optimal decision literature, dating back to the 1950s [@Schaefer1954], demonstrates that the Maximum Sustainable Yield (MSY) is maintained by harvesting a stock down to the size at which it achieves its maximum growth rate, i.e.
+50% of the un-fished equilibrium size for a symmetric growth model ($K/2$).
+Model 1, while being very wrong about both the growth rate and the un-fished equilibrium, is nevertheless nearly perfect in estimating the stock size at which maximum growth rate is achieved, and this gives nearly optimal decisions (Fig 3B) despite its terrible forecasts.
 
@@
 But in real ecological decision making, we never know the true model -- we will always be comparing among approximations.
-Within fisheries, even in today's parameter-rich age-structured models, recruitment approximations with symmetric growth functions (Logistic, Ricker, Beverton-Holt, etc) still dominate [@ramlegacy].
+Within fisheries, even in today's parameter-rich age-structured models, recruitment approximations with symmetric growth functions (Logistic, Ricker, Beverton-Holt, etc) still dominate [@ramlegacy2012; @ramlegacy2018].
 
 This issue is by no means unique to fisheries.
-Throughout resource management and conservation decision-making, as is no doubt true in other fields, decisions about which model to use are guided by which model best fits the data [].
-As this example illustrates, such metrics, no matter how rigorously defined, may select entirely the wrong model.
-And once we are managing with the wrong model, no amount of comparing model predictions to outcomes will guarantee we discover our mistake.
-Only by comparing to the outcomes we actually care about -- in this case, economic and ecological performance -- can we see which model is best for decision-making.
-Conversely, our example reminds us that just because a model continually makes poor predictions does not mean that the model leads to poor decisions.
-Model 1, despite it's many mistakes, is right about one key feature: the position of the peak growth rate -- and that is enough to guarentee nearly optimal performance.
-This should also be reassuring to modelers and decision makers, for it reminds us that effective models need be correct or even close to correct in every aspect, as long as they capture the key features of the decision context.
-Research into decision theory, such as the work by @Schaefer1954; @Clark1974; @Reed1979; @Memarzadeh2018 help us better understand that context.
-Adaptive management approaches [@Walters1978] can apply that theory to compare management outcomes between models directly, rather than comparing to forecast performance.
+Throughout resource management and conservation, and no doubt other fields, decisions about which model to use are guided by which model best fits available data [@Clark1990].
+Increasingly, these are joined by calls to assess _forecast accuracy_ [@Clark2001; @Dietze2018; @White2019] as the ultimate test of a model.
+Yet as this example illustrates, such metrics, no matter how rigorously defined, may select entirely the wrong model for the task at hand.
+A decision maker has other objectives than prediction accuracy, and approaches which ignore these considerations do so at their peril.
+This example has also shown that once we are managing with the wrong model, no amount of comparing predictions from that model to actual outcomes will guarantee we discover our mistake.
+Despite its consistently good predictions, model 1 is in fact over-fishing to a dangerous level.
 
-```{r figure3, fig.width=7, fig.height=4}
-fig3a + fig3b
+Because we will never know the "true" model, we must never forget that our choice of models must reflect the context for which those models will be used [@Levins1966].
+Model 2 would indeed be a better choice than model 1 if our objective was to determine the natural size of our fish stock in the absence of fishing.
+Only when we focus on the outcomes we actually care about -- in this case, economic and ecological performance -- can we see which model is best for decision-making.
+Model 1, despite its many mistakes, is right about one key feature: the biomass for peak growth -- and that is enough to guarantee nearly optimal performance.
+This conclusion should also be reassuring to both modelers and decision makers, for it reminds us that effective models need be perfect or even all that close in every aspect, as long as they capture the key features of the decision context.
+Decision theory [e.g.
+@Clark1973; @Reed1979; @pomdp-intro] and research into the socio-ecological models [e.g.
+@Kareiva2006] helps us better understand that context.
+Adaptive management approaches [@Walters1978] can apply that theory to compare management outcomes between models directly.
+It is not true that we need good forecasts to make good decisions.
+
+```{r figure3, fig.width=7, fig.height=4, fig.cap="Panel A: Population recruitment curves for each model, compared to that of the true model.
+Panel B: The computed optimal policy of each model, derived by SDP, expressed in terms of the target escapement (population size remaining after harvest) for each possible stock size.
+Model 1 overharvests consistently, while the target escapement under model 2 is nearly identical to that of model 3."}
+fig3a + ggtitle("A") + fig3b + ggtitle("B")
 ```
@@
 
-This work was supported in part by NSF CAREER () and computational resources from NSF's XSEDE Jetstream (DEB160003) and Chameleon cloud platforms.
+This work was supported in part by NSF CAREER (#1942280) and computational resources from NSF's XSEDE Jetstream (DEB160003) and Chameleon cloud platforms.
 

### 2020-11-05 manuscript/appendix.Rmd
here we go

@@
 ---
-title: 'Appendix for: "Bad Forecast, Good Decision..."'
+title: 'Appendix for: "Bad Forecast, Good Decision"'
 date: "`r Sys.Date()`"
@@
 
-Here we provide annotated code necessary to completely reproduce all of the analysis
-presented in the main paper.
-This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving 
-Markov Decision Processes (MDP) using stochastic dynamic programming functionality,
-`expm` for matrix exponentials [@expm], and a few custom MDP functions provided
-by our package, `mdplearning` [@mdplearning].  We will also use `tidyverse` 
-packages for basic manipulation and plotting [@tidyverse]. This file is also
-available as an RMarkdown document [@rmarkdown] at 
-<https://github.com/cboettig/bad-forecast-good-decision>
+Here we provide annotated code necessary to completely reproduce all of the analysis presented in the main paper.
+This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving Markov Decision Processes (MDP) using stochastic dynamic programming functionality, `expm` for matrix exponentials [@expm], and a few custom MDP functions provided by our package, `mdplearning` [@mdplearning].
+We will also use `tidyverse`  packages for basic manipulation and plotting [@tidyverse].
+This file is also available as an RMarkdown document [@rmarkdown] at  <https://github.com/cboettig/bad-forecast-good-decision>.
 
@@
 
-While in principle the utility could reflect many things, including the cost of fishing, market responses to supply and demand, the value of recreational fishing, the intrinsic value fish left in the sea [see @Halpern2013], for simplicity we will assume utility is merely a linear function of the harvest, i.e.
-a fixed price $p$ per kilogram of fish harvested: $U(X_t, H_t) = p H_t$.
+While in principle the utility could reflect many things, including the cost of fishing, market responses to supply and demand, the value of recreational fishing, the intrinsic value fish left in the sea [see @Halpern2013], for simplicity we will assume utility is merely a linear function of the harvest quota set by the manager, i.e.
+a fixed price $p$ per kilogram of fish harvested: $U(X_t, H_t) = p \min(H_t, X_t)$ (noting that realized harvest cannot exceed the available stock).
+As units are considered already non-dimensionalized in this example, without loss of generality we will set $p=1$.
 This problem is already well studied, and it is worth noting that even under such a pessimistic assumption, the optimal strategy still seeks to sustain the fish population indefinitely; as @Clark1973 shows for the deterministic function $f$ and @Reed1979 extended to the stochastic case.
@@
 
-To solve the decision problem using SDP, we define
-the state space and the action space on a discrete grid of 240 points.  The maximum
-state is set well above the largest carrying capacity used in the models,
-which limits the influence of boundary effects introduced by the transformation to
-a discrete, finite grid.  Available harvest actions match the state space, effectively 
-allowing any harvest level to be possible.
+To solve the decision problem using SDP, we define the state space and the action space on a discrete grid of 240 points.
+The maximum state is set well above the largest carrying capacity used in the models, which limits the influence of boundary effects introduced by the transformation to a discrete, finite grid.
+Available harvest actions match the state space, effectively allowing any harvest level to be possible.
 
@@
 
-Alternate reward functions with varying price and cost structures are possible but do not qualitatively
-impact the conclusions.  This reward function is a limiting case of any more 
-complex reward, and corresponds both to classic work that does not model utility
-explicitly [e.g. MSY theory, @Schaefer1954], as well as explicit assumptions
-typically made in more recent models [@Reed1979]. Moreover, using a simple reward function
-makes it clear that model that performs best is not doing so merely because of particular
-features baked into the a carefully chosen reward rule.
+Alternate reward functions with varying price and cost structures are possible but do not qualitatively impact the conclusions.
+This reward function is a limiting case of any more complex reward, and corresponds both to classic work that does not model utility explicitly [e.g.
+MSY theory, @Schaefer1954], as well as explicit assumptions typically made in more recent models [@Reed1979].
+Moreover, using a simple reward function makes it clear that model that performs best is not doing so merely because of particular features baked into the a carefully chosen reward rule.
 
@@
 We use value iteration to solve the stochastic dynamic program [@Marescot2013; @MDPtoolbox] for each model.
-This determines the optimal harvest policy for each possible state, given each model. Because this step is the
-most computationally intensive routine, we cache the results using memosization conditioned on the transition
-matrices [@memoise].  Running this code with alternate transition matrices automatically invalidates that
-cache, reducing the risk of loading spurious results.
+This determines the optimal harvest policy for each possible state, given each model.
+Because this step is the most computationally intensive routine, we cache the results using memosization conditioned on the transition matrices [@memoise].
+Running this code with alternate transition matrices automatically invalidates that cache, reducing the risk of loading spurious results.
 
@@
 
-We simulate fishing dynamics under the optimal policy for each model, using a simple
-helper function from the `mdplearning` package. Because growth dynamics are stochastic,
-we perform 100 simulations of each model from identical starting condition to
-ensure results are not the result of chance alone.
+We simulate fishing dynamics under the optimal policy for each model, using a simple helper function from the `mdplearning` package.
+Because growth dynamics are stochastic, we perform 100 simulations of each model from identical starting condition to ensure results are not the result of chance alone.
 
@@
 ```{r proper_scores, message = FALSE}
-# Gneiting & Raferty (2007), eq24
+# Gneiting & Raferty (2007), eq27
 scoring_fn <- function(x, mu, sigma){ -(mu - x )^2 / sigma^2  - log(sigma)}
@@
 
-Having computed a single data frame of simulation data under both optimally fished
-(according to each model) and unfished scenarios for each model, along with
-the proper scores for the step-ahead forecasts a la @Gneiting2007, we have all 
-the results necessary to generate the figures presented in the main paper.
+Having computed a single data frame of simulation data under both optimally fished (according to each model) and unfished scenarios for each model, along with the proper scores for the step-ahead forecasts a la @Gneiting2007, we have all the results necessary to generate the figures presented in the main paper.
 
@@
 
-Likewise we can plot the data on proper scores associated with
-each prediction of each model:
+Likewise we can plot the data on proper scores associated with each prediction of each model:
 
@@
   ylab("state") + 
-  ggtitle("A") + labs(subtitle = "Ecological performance")
+  ggtitle("A") + labs(subtitle = "Ecological")
 ```
@@
   ylab("Net present value") + xlab("time")  + 
-  ggtitle("B.") + labs(subtitle = "Economic performance")
+  ggtitle("B.") + labs(subtitle = "Economic")
 

### 2020-11-05 manuscript/manuscript.Rmd
here we go

@@
 Ecological processes are intrinsically complex, so that even our best models can only ever be approximations of underlying processes.
-In choosing the best model for decision-making, it can be more important to capture a single key feature of the process than it is to make the most accurate prediction about future states.
 Even strictly proper scoring rules for probabilistic forecasts [@Gneiting2007] will not always select the best model to guide decision-making.
-Surprisingly, this is true not only in conceptual models that may guide policy choices, but also when a decision policy is derived directly from a complex optimization routine of a probabilistic predictive model, such as the stochastic dynamic programming (SDP) algorithms used to derive optimal policies to sequential decision problems [@Marescot2013].
-Here, I use a classical, well-understood example from fisheries management [@Schaefer1954; @Clark1973; @Reed1979] to illustrate both the paradox of how a model with the worst forecast provides the best decision outcomes, as well as show how we can avoid selecting models that are poorly suited for management by considering the management context more explicitly.
+Surprisingly, this can even happen when a decision is derived directly from a complex optimization routine of a probabilistic predictive models [@Marescot2013].
+Here, I use a classic, well-understood example from fisheries management [@Schaefer1954; @Clark1973; @Reed1979] to illustrate both the paradox of how a model with the worst forecast provides the best decision outcomes, as well as show how we can avoid selecting models that are poorly suited for management by considering the management context more explicitly.
+These results underscore that in choosing the best model for decision-making, it can be more important to capture a single key feature of the process than it is to make the most accurate prediction about future states.
 
@@
 Rather, using a simple model merely reflects the famous compromise of Richard Levins [@Levins1966] in choosing generality over precision.
-More precisely, the decision problem in question can be stated as follows: The fish stock is observed to be in state `X_t` at time `t`, and is then subjected to some harvest `H_t` before recruiting new fish, subject to stochastic environmental noise `\xi_t`, to bring the stock to $X_t+1$,
+More precisely, the decision problem in question can be stated as follows: The fish stock is observed to be in state $X_t$ at time $t$, and is then subjected to some harvest $H_t$ before recruiting new fish, subject to stochastic environmental noise $\xi_t$, to bring the stock to $X_t+1$,
 
@@
 ```{r figure1, fig.width=7, fig.height=5, fig.cap = "Forecast performance of each model.
-Panels A, B: Step ahead predictions under unfished (A) and fished (B) scenarios.
+Panels A, B: Step ahead predictions of stock size under unfished (A) and fished (B) scenarios.
 Error bars indicating the 95% confidence intervals around each prediction, while stars denote the observed value in that year.
-Because the models make different decisions each year, the stock size in year 2, 3, etc under management of model 1 (blue stars) is different from that under model 2 (red stars).
-Panels C, D: corresponding distribution of proper scores across all predictions.
-Higher scores are better."}
+Because the models make different decisions each year in the fished scenario, the observed stock size in year 2, 3, etc under the management of model 1 (blue stars) is different from that under model 2 (red stars).
+Panels C, D: corresponding distribution of proper scores across all predictions (100 replicates of 100 timesteps).
+Higher scores are better, confirming that model 2 makes the better forecasts."}
 fig1ab + fig1cd + 
@@
 
-```{r figure2, message = FALSE, fig.cap = "Ecological and economic performance of each forecast."}
+```{r figure2, message = FALSE, fig.cap = "Ecological and economic performance of each forecast.
+Harvest quotas derived from model 1 result in a significantly higher stock size than under model 2 (panel A).
+Despite this less agressive harvest policy, economic returns under model 1 are substantially higher (panel B)."}
 fig2a + fig2b
@@
 
+# Methods
+
+Stochastic transition matrices are defined for models 1-3 on a discrete grid of
+240 possible states spaced uniformly from 0 to 24. A discrete action space 
+enumerating possible harvest quotas is set to the same grid.
+The utility of a harvest quota $H_t$ given a population state $X_t$ is given by $U(X_t, H_t) = \min(X_t, H_t)$ (i.e.
+a fixed price for realized harvest).
+A modest discount of
+$\gamma = 0.99$ allows comparisons to approaches that ignore [@Schaefer1954] or include [@Clark1973; @Reed1979] discounting; results are not sensitive to this choice.
+The optimal policy for each model is determined by stochastic dynamic programming [@Marescot2013].
+Details of the implementation, including fully reproducible R code, have been included in the appendix.
+
 # Acknowledgements

### 2020-11-05 manuscript/manuscript.Rmd
re-title

@@
 ---
-title: "Bad forecast, good decision: what makes a model relevant for determining policy"
+title: "Bad forecast, good decision: predictive accuracy is not everything"
 date: "`r Sys.Date()`"

### 2020-11-19 manuscript/manuscript.Rmd
edits

@@
 abstract: |
-    With the rapid expansion of available data, it is frequently suggested that _forecasting performance_ is the ultimate test of a scientific model.
-Yet the model that makes the most accurate forecasts is not always the model that yields the best decisions.
-This is no accident of statistics, even formal criteria such as strictly proper scoring rules cannot determine the best model for decision-making because they do not consider the context of the decision process itself.
-I illustrate this paradox using a classic example of ecological decision making in fisheries management.
-This example also reveals how an understanding of the decision problem can both explain the poor management outcomes of model which produces the best forecast, and also see how to avoid making such a mistake.
-As forecasting becomes an increasing attractive and viable possibility for assessing ecological models and applying them to real-world decisions, I hope this illustration serves as a reminder that modeling can not be done in a vacuum, but must reflect the context in which the model will be applied.
-journal: TBD
+     Model-based forecasts have been enabled by recent explosion of data and computational methods and spurred on by decision-makers appetite for forecasts in everything from elections to pandemic response.
+It is taken for granted that the model which makes the most accurate forecast, accounting for uncertainty, will also be the best model to inform decision-making.
+Using a classic example from fisheries management, I demonstrate that selecting the model that produces the best forecast can lead to decidedly worse outcomes.
+This situation can arise whenever the models are only approximations of the reality, underscoring the risk of evaluating models only by prediction accuracy and not decision outcomes
+journal: " "
 layout: 3p
 bibliography: refs.bib
-csl: nature.csl
-output: 
-  rticles::elsevier_article:
-    dev: cairo_pdf
-    keep_tex: true
+csl: science.csl
+output: word_document
+#  rticles::elsevier_article:
+#    dev: cairo_pdf
+#    keep_tex: true
+
 keywords: 
@@
 Accurate forecasts are important both in assessing the accuracy of our models and understanding of natural processes, they also underpin policy and decision making [@Dietze2018].
+Strictly proper scoring rules provide mathematical guarantees that they will, on average, select the true model over any other[@Gneiting2007], paving the way for ever more complex and more accurate forecasting models.
 Yet the model that leads to the best decisions is not always the model that makes the most accurate forecasts, as I illustrate here.
-Ecological processes are intrinsically complex, so that even our best models can only ever be approximations of underlying processes.
-Even strictly proper scoring rules for probabilistic forecasts [@Gneiting2007] will not always select the best model to guide decision-making.
 Surprisingly, this can even happen when a decision is derived directly from a complex optimization routine of a probabilistic predictive models [@Marescot2013].
+Reality is complex; even our best models can only ever be approximations of underlying processes.
 Here, I use a classic, well-understood example from fisheries management [@Schaefer1954; @Clark1973; @Reed1979] to illustrate both the paradox of how a model with the worst forecast provides the best decision outcomes, as well as show how we can avoid selecting models that are poorly suited for management by considering the management context more explicitly.
@@
 ```{r figure2, message = FALSE, fig.cap = "Ecological and economic performance of each forecast.
-Harvest quotas derived from model 1 result in a significantly higher stock size than under model 2 (panel A).
-Despite this less agressive harvest policy, economic returns under model 1 are substantially higher (panel B)."}
+Harvest quotas derived from Model 1 result in a significantly higher fish stock size than under Model 2 (panel A).
+Economic returns under Model 1 are also substantially higher (panel B)"}
 fig2a + fig2b
@@
 
-```{r figure3, fig.width=7, fig.height=4, fig.cap="Panel A: Population recruitment curves for each model, compared to that of the true model.
+```{r figure3, fig.width=7, fig.height=4, fig.cap="Panel A: Population recruitment curves for each model, compared to that of the true Model.
+3 Model 2 more closely approximates the true Model 3, but note that maximum value Model 1 and Model 3 occur at nearly the same value for the state, x.
 Panel B: The computed optimal policy of each model, derived by SDP, expressed in terms of the target escapement (population size remaining after harvest) for each possible stock size.
-Model 1 overharvests consistently, while the target escapement under model 2 is nearly identical to that of model 3."}
+Model 1 over-harvests consistently, while the target escapement under Model 2 is nearly identical to that of the true Model 3."}
 fig3a + ggtitle("A") + fig3b + ggtitle("B")

### 2020-11-20 manuscript/manuscript.Rmd
formatting options

@@
 ---
-title: "Bad forecast, good decision: predictive accuracy is not everything"
-date: "`r Sys.Date()`"
+title: 'Bad forecast, good decision: predictive accuracy is not everything'
 author:
@@
   affiliation: ucb
-address:
-  - code: ucb
-    address: "ESPM Department, University of California, 130 Mulford Hall Berkeley, CA 94720-3114, USA"
+date: "`r Sys.Date()`"
+output:
+  html_document:
+    df_print: paged
 abstract: |
@@
 This situation can arise whenever the models are only approximations of the reality, underscoring the risk of evaluating models only by prediction accuracy and not decision outcomes
-journal: " "
+journal: ' '
 layout: 3p
@@
 csl: science.csl
-output: word_document
-#  rticles::elsevier_article:
-#    dev: cairo_pdf
-#    keep_tex: true
-
+address:
+- code: ucb
+  address: ESPM Department, University of California, 130 Mulford Hall Berkeley, CA
+    94720-3114, USA
 keywords:
@@
   \linenumbers
-
 ---

### 2020-12-04 manuscript/manuscript.Rmd
reformatting

@@
   email: cboettig@berkeley.edu
-  affiliation: ucb
+  affiliation: ucb,corr
 date: "`r Sys.Date()`"
-output:
-  html_document:
-    df_print: paged
+output: rticles::pnas_article
 abstract: |
@@
 Using a classic example from fisheries management, I demonstrate that selecting the model that produces the best forecast can lead to decidedly worse outcomes.
-This situation can arise whenever the models are only approximations of the reality, underscoring the risk of evaluating models only by prediction accuracy and not decision outcomes
-journal: ' '
-layout: 3p
+This situation can arise whenever the models are only approximations of the reality, underscoring the risk of evaluating models only by prediction accuracy and not decision objectives using formal tools of decision theory.
+
+corresponding_author:
+  code: corr
+  text: "To whom correspondence should be addressed. E-mail: cboettig@berkeley.edu"
+
 bibliography: refs.bib
-csl: science.csl
 address:
@@
     94720-3114, USA
+
 keywords:
@@
 - ecological management
-preamble: |
-  \journal{preprint}
-  \usepackage{lineno}
-  \linenumbers
+
+# For footer text
+lead_author_surname: Boettiger
+
+significance: |
+  ...
+
+acknowledgements: |
+  This work was supported in part by NSF CAREER (#1942280) and computational resources from NSF's XSEDE Jetstream (DEB160003) and Chameleon cloud platforms.
+
+keywords:
+  - one
+  - two
+  - optional
+  - optional
+  - optional
+
+## must be one of: pnasresearcharticle (usual two-column layout), pnasmathematics (one column layout), or pnasinvited (invited submissions only)
+pnas_type: pnasresearcharticle
+csl: pnas.csl
+
+## change to true to add optional line numbering
+lineno: true
+
+output: rticles::pnas_article
+
 ---
@@
 
-# Acknowledgements
-
-This work was supported in part by NSF CAREER (#1942280) and computational resources from NSF's XSEDE Jetstream (DEB160003) and Chameleon cloud platforms.
-
 \pagebreak

### 2021-01-01 manuscript/appendix.Rmd
am

@@
 
+# Adaptive Management
+
+The adaptive management solution does not rely on forecast skill alone to select a model.
+Instead, the manager assigns prior probabilities across all possible models, and updates those priors in response to future observations.
+Decisions reflect uncertainty over all models, and the resulting uncertainty about the net present value that would result.
+
+```{r}
+adaptive_management <- memoise::memoise(mdp_learning, cache = memoise::cache_filesystem("cache/"))
+
+am1 <- adaptive_management(transitions[1:2], reward, discount, model_prior = c(0.5, 0.5), x0 = 10, 
+               Tmax = 20, true_transition = transitions[[3]])
+```
+
+```{r}
+## Skeptic of model-1
+am2 <- adaptive_management(transitions[1:2], reward, discount, model_prior = c(0.1, 0.9), x0 = 10, 
+               Tmax = 20, true_transition = transitions[[3]])
+```
+
 # Simulations and step-ahead forecasts

### 2021-01-01 manuscript/manuscript.Rmd

@@
   email: cboettig@berkeley.edu
-  affiliation: ucb,corr
+  affiliation: a,1
 date: "`r Sys.Date()`"
-output: rticles::pnas_article
+
 abstract: |
@@
 It is taken for granted that the model which makes the most accurate forecast, accounting for uncertainty, will also be the best model to inform decision-making.
-Using a classic example from fisheries management, I demonstrate that selecting the model that produces the best forecast can lead to decidedly worse outcomes.
+Using a classic example from fisheries management, I demonstrate that selecting the model that produces the most accurate and precise forecast can lead to decidedly worse outcomes.
 This situation can arise whenever the models are only approximations of the reality, underscoring the risk of evaluating models only by prediction accuracy and not decision objectives using formal tools of decision theory.
@@
 corresponding_author:
-  code: corr
+  code: 1
   text: "To whom correspondence should be addressed. E-mail: cboettig@berkeley.edu"
@@
 address:
-- code: ucb
-  address: ESPM Department, University of California, 130 Mulford Hall Berkeley, CA
+- code: a
+  address: Department of Environmental Science, Policy, and Management, University of California, 130 Mulford Hall Berkeley, CA
     94720-3114, USA
@@
 significance: |
-  ...
+  It is natural to assume that a more accurate and precise forecast will lead to better decisions
+  than a less accurate one.  I demonstrate that this premise is false, and illustrate how a 
+  decision-theoretic approach provides better criteria for chosing a forecast model than forecast skill.
+  As the expansion of machine learning methods and big data lead to increasing more accurate and
+  precise forecasts in many fields, we should remember that alone is not enough to jettison earlier 
+  models.
 
@@
   This work was supported in part by NSF CAREER (#1942280) and computational resources from NSF's XSEDE Jetstream (DEB160003) and Chameleon cloud platforms.
-
-keywords:
-  - one
-  - two
-  - optional
-  - optional
-  - optional
 
@@
 output: rticles::pnas_article
-
 ---

### 2021-01-25 manuscript/appendix.Rmd
appendix

--- 2020-11-05 manuscript/appendix.Rmd
@@
 scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
-pal <- ggthemes::solarized_pal()(6)
+pal <- ggthemes::solarized_pal()(8)
 txtcolor <- "#586e75"
@@
     soln <- mdp(P, reward, discount = discount, 
-                epsilon = 0.01, max_iter = 1000, V0 = rep(0,dim(P)[[1]]))
+                epsilon = 0.01, max_iter = 2000, V0 = rep(0,dim(P)[[1]]))
     escapement <- states - actions[soln$policy]
@@
   .id = "model")
+```
+
+# Adaptive Management
+
+Passive adaptive management using a Bayesian learning scheme still learns the wrong model.
+
+```{r}
+adaptive_management <- memoise::memoise(mdp_learning, cache = memoise::cache_filesystem("cache/"))
+x0 <- which.min(abs(states - 6))
+
+am1 <- adaptive_management(transitions[1:2], reward, discount, model_prior = c(0.99, 0.01), x0 = x0, 
+               Tmax = 50, true_transition = transitions[[3]], 
+                epsilon = 0.001, max_iter = 2000)
+```
+
+```{r}
+am <- am1$df %>% mutate(belief = am1$posterior$V2)
+am %>%  ggplot(aes(time, states[state], col = belief)) + geom_line() + geom_point() +  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4])
+```
+
+```{r include = FALSE}
+as_tibble(am1$posterior) %>% 
+  mutate(time = seq_along(V1)) %>%
+  select(time = time, "1" = V1, "2" = V2) %>% 
+  pivot_longer(c("1", "2"), values_to="belief", names_to = "model") %>%
+  filter(time < 20) %>%
+  ggplot(aes(time, belief, col=model)) + geom_line(col = pal[4])
 ```

### 2021-01-25 manuscript/manuscript.Rmd
plotting

@@
 
+Uncertainty in decision-making
+forecasting as the answer
+forecasting is not always the answer
+
 A primary purpose of statistical analyses and ecological modeling is to make forecasts for the future [@Clark2001; @Gneiting2014].
 Accurate forecasts are important both in assessing the accuracy of our models and understanding of natural processes, they also underpin policy and decision making [@Dietze2018].
-Strictly proper scoring rules provide mathematical guarantees that they will, on average, select the true model over any other[@Gneiting2007], paving the way for ever more complex and more accurate forecasting models.
+"Strictly proper" scoring rules [sensu @Gneiting2007] provide mathematical guarantee that they will, on average, select the true model over any other, paving the way for ever more complex and more accurate forecasting models.
+Unlike alternative measures of 'goodness of fit' forecasts are
+
 Yet the model that leads to the best decisions is not always the model that makes the most accurate forecasts, as I illustrate here.

### 2021-01-26 manuscript/appendix.Rmd
plotting
revise, more background / context

@@
 
-Our decision problem is defined as follows: 
+The question of optimal harvest given a population growth model can be posed as a Markov Decision Process (MDP).
+@Marescot2013 provides an accessible overview of MDP problems and their solutions in conservation context.
+@Reed1979 provides a formal proof of optimal harvest strategy for stochastic dynamics under the assumption of any concave growth function.
+The problem can be summarized as follows:
+
 A manager seeks to maximize the sum of the utility derived from such a harvest
@@
 a fixed price $p$ per kilogram of fish harvested: $U(X_t, H_t) = p \min(H_t, X_t)$ (noting that realized harvest cannot exceed the available stock).
-As units are considered already non-dimensionalized in this example, without loss of generality we will set $p=1$.
-This problem is already well studied, and it is worth noting that even under such a pessimistic assumption, the optimal strategy still seeks to sustain the fish population indefinitely; as @Clark1973 shows for the deterministic function $f$ and @Reed1979 extended to the stochastic case.
+As units are already specified in dimensionless quantities in this example, so without loss of generality we will set $p=1$.
+
+To solve this optimization, we must also know the dynamics of $X_t$, describing how future stock $X_{t+1}$ changes from the present state, $X_t$, after to harvest $H_t$:
+
+\begin{equation}
+X_{t+1} = f(X_t, H_t)
+\end{equation}
+
+This problem is already well studied, and it is worth noting that even under the simple utility function which assigns no value to fish that are not harvested, the optimal strategy still seeks to sustain the fish population indefinitely; as @Clark1973 shows for the deterministic function $f$ and @Reed1979 extended to the stochastic case.
+As @Reed1979 proves, the optimal strategy can be characterized by a policy of "constant escapement," in which the manager adjusts the harvest effort each year in an effort to ensure the same number or biomass of fish are left in the sea each year.
+
+The optimization problem can then be solved by dynamic programming, using the recursive Bellman equation as ,
+
+\begin{equation}
+\end{equation}
+
+This technique, known as stochastic dynamic programming (SDP) is widely used in conservation decision problems [e.g. @Mangel1985, @Marescot2013].
+
+It is worth noting that while constant escapement is used in the management of several important fish stocks such as salmon, most marine fisheries are not managed using this optimal solution for a fluctuating population, but instead rely on a 'constant harvest effort' (or constant "yield") policy.
+In the case of a deterministic model at equilibrium, these policies are identical.
+Though constant-effort policies are technically sub-optimal for the problem as stated above with stochastic growth, in practice they can be more robust, such as when measurements of the current stock size $X_t$ are uncertain.
+These issues are discussed in detail in @pomdp-intro.
+For our purposes, it suffices to note that constant escapement policy is relevant both from a theoretical standpoint as the optimal solution for the problem under consideration, and an applied standpoint as a policy that guides management of many salmon fisheries.
+The optimal control approach using SDP as illustrated here is used in a wide and growing number of conservation problems.
+
+Bear in mind that in the example presented here, SDP merely serves the role of a convenient and established way for a manager to determine the optimal action, given some model estimation.
+Whether we use SDP or some other way to determine the optimal policy, given the model, is immaterial to the conclusion.
+The issue at hand is how we determine the model in the first place.
+
 Given the function $f$ with known parameters, it is straight forward to determine the optimal harvest policy by stochastic dynamic programming (SDP, see @Mangel1985; @Marescot2013).
@@
 \begin{equation}
-f_i(Y) = Y + Y r_i \left( 1 - \frac{Y}{K_i} \right)
+f_i(Y) = Y + Y r_i \left( 1 - \frac{Y}{K_i} \right) \xi_i(t)
 \end{equation}
+
+where $\xi_i(t)$ is a multiplicative log-normal noise term with mean 1 and log-standar-deviation $\sigma_i$.
 
@@
 am <- am1$df %>% mutate(belief = am1$posterior$V2)
-am %>%  ggplot(aes(time, states[state], col = belief)) + geom_line() + geom_point() +  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4])
+am %>%  ggplot(aes(time, states[state], col = belief)) + 
+  geom_line() + 
+  geom_point() +
+  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4])
 ```

### 2021-02-05 manuscript/appendix.Rmd
WIP

@@
 ---
-
-```{r graphics_setup, message=FALSE, warning=FALSE, include = FALSE}
-## Plotting themes, colors, and fonts 
-## aesthetics choices only, all can be omitted without consequence
-library(ggthemes)
-library(hrbrthemes)
-library(ggplot2)
-library(Cairo)
-library(extrafont)
-library(patchwork)
-library(styler)
-extrafont::loadfonts(quiet = TRUE)
-ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
-
-scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
-scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
-pal <- ggthemes::solarized_pal()(8)
-txtcolor <- "#586e75"
-
-knitr::opts_chunk$set(cache=FALSE, tidy = "styler", message = FALSE, warning = FALSE, echo = TRUE)
-
-```
-
-Here we provide annotated code necessary to completely reproduce all of the analysis presented in the main paper.
-This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving Markov Decision Processes (MDP) using stochastic dynamic programming functionality, `expm` for matrix exponentials [@expm], and a few custom MDP functions provided by our package, `mdplearning` [@mdplearning].
-We will also use `tidyverse`  packages for basic manipulation and plotting [@tidyverse].
-This file is also available as an RMarkdown document [@rmarkdown] at  <https://github.com/cboettig/bad-forecast-good-decision>.
-
-```{r setup, message=FALSE}
-library(tidyverse)
-library(MDPtoolbox)
-library(expm)
-# remotes::install_github("boettiger-lab/mdplearning")
-library(mdplearning)
-```
 
@@
 
-with $r_3 = 0.002$, $K_3 = 10$ and $\sigma_3 = 0.05$:
+with $r_3 = 0.002$, $K_3 = 10$ and $\sigma_3 = 0.05$.
+
+Here we provide annotated code necessary to completely reproduce all of the analysis presented in the main paper.
+This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving Markov Decision Processes (MDP) using stochastic dynamic programming functionality, `expm` for matrix exponentials [@expm], and a few custom MDP functions provided by our package, `mdplearning` [@mdplearning].
+We will also use `tidyverse`  packages for basic manipulation and plotting [@tidyverse].
+This file is also available as an RMarkdown document [@rmarkdown] at  <https://github.com/cboettig/bad-forecast-good-decision>.
+
+```{r graphics_setup, message=FALSE, warning=FALSE, include = FALSE}
+## Plotting themes, colors, and fonts 
+## aesthetics choices only, all can be omitted without consequence
+library(ggthemes)
+library(hrbrthemes)
+library(ggplot2)
+library(Cairo)
+library(extrafont)
+library(patchwork)
+library(styler)
+extrafont::loadfonts(quiet = TRUE)
+ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
+
+scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
+scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
+pal <- ggthemes::solarized_pal()(8)
+txtcolor <- "#586e75"
+
+knitr::opts_chunk$set(cache=FALSE, tidy = "styler", message = FALSE, warning = FALSE, echo = TRUE)
+
+```
+
+```{r setup, message=FALSE}
+library(tidyverse)
+library(MDPtoolbox)
+library(expm)
+# remotes::install_github("boettiger-lab/mdplearning")
+library(mdplearning)
+```
 
@@
   coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
-  ylab(bquote(f(x))) + xlab("x")
+  ylab(bquote(f(x) - x)) + xlab("x")
 ```

### 2021-02-05 manuscript/manuscript.Rmd
:pen: WIP

--- 2021-01-01 manuscript/manuscript.Rmd
@@
 ---
-title: 'Bad forecast, good decision: predictive accuracy is not everything'
+title: ''
 author:
@@
 abstract: |
+
   Model-based forecasts have been enabled by recent explosion of data and computational methods and spurred on by decision-makers appetite for forecasts in everything from elections to pandemic response.
@@
 significance: |
-  It is natural to assume that a more accurate and precise forecast will lead to better decisions
-  than a less accurate one.  I demonstrate that this premise is false, and illustrate how a 
-  decision-theoretic approach provides better criteria for chosing a forecast model than forecast skill.
-  As the expansion of machine learning methods and big data lead to increasing more accurate and
-  precise forecasts in many fields, we should remember that alone is not enough to jettison earlier 
-  models.
 
@@
 
-A primary purpose of statistical analyses and ecological modeling is to make forecasts for the future [@Clark2001; @Gneiting2014].
-Accurate forecasts are important both in assessing the accuracy of our models and understanding of natural processes, they also underpin policy and decision making [@Dietze2018].
-Strictly proper scoring rules provide mathematical guarantees that they will, on average, select the true model over any other[@Gneiting2007], paving the way for ever more complex and more accurate forecasting models.
+Global change issues are complex and outcomes are difficult to predict, making approaches to uncertainty a central part
+of effective decision-making [@ @]. While some uncertainty is intrinsic to the underlying processes (stochasticity) or by
+limits to what variables we can observe and how accurately we measure them (measurement uncertainty), much uncertainty 
+is the result of our imperfect knowledge of the processes involved, as expressed by the structure and
+parameters of mathematical models used to approximate those processes (model uncertainty). Model uncertainty can be 
+reduced by gathering additional data, which may rule out certain models or parameter values as implausible. 
+When model parameters are estimated directly from available data, there is a risk that the best-fitting models may 
+_overfit_ to patterns arising from by chance from the stochasticity or measurement error which do not reflect the
+underlying process. Researchers have often favored simpler models which are less prone to over-fitting, sometimes
+through explicit penalties such as information criteria [@], though these can be misleading [@]. With the rapid 
+expansion of available ecological and environmental data [@], it is increasingly possible to rely instead on comparisons
+between model predictions and data purposely excluded from model estimation (cross-validation), or better, 
+comparisons between model predictions and data collected in the future (forecasting, @Clark2001; @Dietze2018]). 
+Over sufficiently long timescales, comparing model forecasts to future observations will select the models with
+greatest predictive accuracy.  Because model-based forecasts frequently play an important role in decision-making,
+it is commonly assumed [e.g. @Walters1981; @Clark2001; @Dietze2018] that this approach for addressing model uncertainty
+will also improve decision-making outcomes. Here, I illustrate that this need not be the case: it is quiet possible
+that any approach with improves the forecast accuracy of the model(s) over time can simultaneously lead to worse 
+decision outcomes. This example illustrates that knowing when increased forecast accuracy will or will not also improve 
+decision outcomes in general remains an open problem, as is the challenge of what to do about it.
+
+<!-- 
+It may be easy to dismiss this example on the argument that both models are (or more precisely,
+the any logistic growth model, regardless of parameterization would be) extremely naive
+and unrealistic, and any solution would surely begin with replacing both models with something
+more realistic.  Yet such objections overlook the simple fact that for all its faults in 
+prediction, model 1 already provides nearly optimal performance.  That performance arises
+because model 1 has very accurately reflected one key aspect of the true model (the position of the
+maximum growth rate), even while it is wrong about every other aspect of the biology.
+
+The open problem is simply to propose an approach to model uncertainty which would select model 1
+over model 2 (i.e. choosing among the suite of available models, the one that gives the better 
+decisions.)
+
+An important aspect of this example is that model 1 performs nearly optimally.
+
+-->
+
+I illustrate this problem using an example from fisheries management. 
+Fisheries are a significant economic and conservation concern world
+wide and their management remains an important debate [e.g. @Worm2006;
+@Worm2009; @Costello2016]. Moreover, their management has been a proving 
+grounds for theoretical and practical decision-making issues which are widely
+applicable in other areas of ecology and conservation [@Ludwig1993; @Lande1994],
+and one that has long wrestled with issues of uncertainty in the context
+of management decisions [e.g. @Clark1973; @Reed1979; @Walters1981; @Ludwig1982].
+While modern fisheries management frequently relies on complex models which may
+contain scores of parameters to reflect the specific age or stage structure of a
+specific fish stock [@ram], I will rely on simple, well-studied models which
+permit greater intuition and generalization [@Levins1966]. Consistent with 
+such previous work [@Schaefer1954; @Clark1973; @Reed1979; @Walters1981; @Ludwig1982; @Costello2016],
+let us consider the problem of determining the optimal harvest policy given a 
+measurement of the current stock size.
+
+# Ecological Models
+
+The sustainable harvest decision problem can be stated as follows: The fish stock is observed to be in state $X_t$ at time $t$, and is then subjected to some harvest $H_t$ before recruiting new fish, subject to stochastic environmental noise $\xi_t$, to bring the stock to $X_t+1$,
+
+\begin{equation}
+X_{t+1} = f(X_t - H_t, \xi_t) 
+\end{equation}
+
+A manager seeks each year to select the harveset $H_t$ which will maximize the sum
+of the utility derived from such a harvest and such a state, $U(X_t,H_t)$, over all time,
+subject to discount rate $\delta$ [@Clark1973]:
+
+\begin{equation}
+\sum_{t=0}^{t=\infty} U(X_t, H_t) \delta^t
+\end{equation}
+
+Further we imagine that the function $f$ is not known precisely, and so we will rely on an evaluation of forecasting skill across a set of candidate models to determine which one to use to manage the fishery.
+Again for simplicity, we will restrict ourselves to two simple candidate models $f_1$ and $f_2$.
+Both share the same underlying structure of logistic recruitment (known as the Gordon-Schaefer model in fisheries context owing to groundbreaking work independently by @Schafer1954 and @Gordon1954), differing only in their choice of certain parameters:
+
+\begin{equation}
+f_i(Y) = Y + r_i Y \left( 1 - \frac{Y}{K_i} \right) * \xi_t(\sigma)
+\end{equation}
+
+Where $\xi_t(\sigma)$ represents log-normal random noise with mean of unity and log-standard-deviation $\sigma$.
+Model 1 is given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, Model 2 by $r_2 = 0.5$,  $K_2 = 10$, $\sigma = 0.075$ (in dimensionless units).
+Having both the larger growth rate and the larger carrying capacity, Model 1 is clearly the more optimistic of the two choices.
+
+Selecting between Model 1 and Model 2 can thus be considered the simplest illustration of the model uncertainty problem.
+This is a subset of the more general problem of selecting model parameters, assuming a logistic growth, which itself is a subset of estimating the best structural from (e.g.
+Ricker, Beverton-Holt, etc).
+There is no need to consider these more complicated versions of the model uncertainty problem here, since they all inherit the same issue.
+Reducing the model selection problem to these two models simplifies the presentation and will aid intuition at no loss of generality.
+
+The only additional assumption we will need is that the "true" model is not among the suite of models under consideration. 
+Mathematical models are, at best, approximations of the underlying processes.
+Ecological processes are much too complex to ever be modeled exactly.
+For illustrative purposes, we will thus assume the "true" process to be given by a third model, which is unknown to the decision-maker, but similar enough to at least one of the candidate models might be considered a reasonable approximation.
+
+\begin{equation}
+f_3(Y) = Y + r_3 Y^4 \left( 1 - \frac{Y}{K_3} \right)
+\end{equation}
+
+with $r_3 = 0.002$, $K_3 = 10$ and $\sigma_3 = 0.05$.
+
+# Managing Under Model Uncertainty
+
+A wide range of paradigms are available for approaching the issue of decision-making under uncertainty.
+These approaches can roughly be divided into two groups: the first group treats the issue of model uncertainty independently from the decision itself, while the second integrates the process of reducing model uncertainty into the process of decision making.
+There are a wide range of techniques within each group.
+
+**Forecasts and Strictly Proper Scoring**
+
+Like many decision problems, this one appears to hinge on having an accurate forecast: if we can predict to what size the stock will increase next year, $X_t+1$, and we know the current stock, $X_t$, then we can safely harvest $X_{t+1} - X_t$.
+Overestimating or underestimating such recruitment will result in over-harvesting or under-harvesting, respectively.
+Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of over-fitting by comparing model predictions to later observations that were not used to estimate the model.
+@Gneiting2007 provides a rigorous proof for "strictly proper" scoring criteria, which have the desirable property which no model predicting the distribution of future outcomes, $Q(x)$ can achieve a better average score than the true model $P(x)$ (That is, unlike likelihood or other goodness-of-fit scores, it is impossible to overfit when conditioning on a strictly proper score).
+
+Iterative updating, with model priors.
+
 Yet the model that leads to the best decisions is not always the model that makes the most accurate forecasts, as I illustrate here.
@@
 -->
-
-Fisheries are a significant economic and conservation concern world wide and their management remains an important debate [e.g.
-@Worm2006; @Worm2009; @Costello2016].
-Moreover, their management has been both a proving grounds for theoretical and practical decision-making issues which are widely applicable in other areas of ecology and conservation [e.g.
-@Mangel1985; @Clark1990; @Marescot2013].
-The decision-making problem is characterized by the need for a manager to set an acceptable harvest quota $H_t$ each year given some stock assessment estimate of the current stock size (population abundance) of the species in question [@Clark1973].
-Such a decision problem appears to hinge on an accurate forecast: if we can predict to what size the stock will increase next year, $X_t+1$, knowing the current stock, $X_t$, then we can safely harvest $X_{t+1} - X_t$.
-Overestimating or underestimating such recruitment will result in over-harvesting or under-harvesting, respectively.
-Thus it may seem natural that our first step would be to select the model that makes the most accurate forecast of next year's stock, $X_{t+1}$.
-I illustrate how we do this using strictly proper scoring criteria [@Gneiting2007] for a set of candidate models,
-and show that it leads to worse decisions.
 
@@
 ```
-
-# Ecological Models
-
-For simplicity, I will focus on the classic case case of a single-species model whose population is observed annually without error in a stochastic but stationary environment without age structure [@Reed1979; @Clark1990; @Costello2016].
-These are not necessary assumptions -- in fact, the more complex the models become, the easier it is find examples in which the best forecast does not produce the best decision.
-Rather, using a simple model merely reflects the famous compromise of Richard Levins [@Levins1966] in choosing generality over precision.
-More precisely, the decision problem in question can be stated as follows: The fish stock is observed to be in state $X_t$ at time $t$, and is then subjected to some harvest $H_t$ before recruiting new fish, subject to stochastic environmental noise $\xi_t$, to bring the stock to $X_t+1$,
-
-\begin{equation}
-X_{t+1} = f(X_t - H_t, \xi_t) 
-\end{equation}
-
-Further we imagine that the function $f$ is not known precisely, and so we will rely on an evaluation of forecasting skill across a set of candidate models to determine which one to use to manage the fishery.
-Again for simplicity, we will restrict ourselves to two simple candidate models $f_1$ and $f_2$.
-Both share the same underlying structure of logistic recruitment:
-
-\begin{equation}
-f_i(Y) = Y + r_i Y \left( 1 - \frac{Y}{K_i} \right)
-\end{equation}
-
-Model 1 is given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, Model 2 by $r_2 = 0.5$,  $K_2 = 10$, $\sigma = 0.075$ (in non-dimensionalized units).
-Having both the larger growth rate and the larger carrying capacity, Model 1 is clearly the more optimistic of the two choices.
-
-Mathematical models are, at best, approximations of the underlying processes.
-Ecological processes are much too complex to ever be modeled exactly down to the last atom.
-For illustrative purposes, we will thus assume the "true" process to be given by a third model, which is unknown to the decision-maker:
-
-\begin{equation}
-f_3(Y) = Y + r_3 Y^4 \left( 1 - \frac{Y}{K_3} \right)
-\end{equation}
-
-with $r_3 = 0.002$, $K_3 = 10$ and $\sigma_3 = 0.05$.
 

### 2021-02-09 manuscript/appendix_B.Rmd
:pen:

--- 2021-01-25 manuscript/appendix.Rmd
@@
 ---
+
+```{r model_definitions}
+states <- seq(0,24, length.out = 240)
+actions <- states
+obs <- states
+```
+
+```{r}
+reward_fn <- function(x,h) pmin(x,h)
+discount <- 0.99
+```
 
@@
 
-Here we provide annotated code necessary to completely reproduce all of the analysis presented in the main paper.
-This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving Markov Decision Processes (MDP) using stochastic dynamic programming functionality, `expm` for matrix exponentials [@expm], and a few custom MDP functions provided by our package, `mdplearning` [@mdplearning].
-We will also use `tidyverse`  packages for basic manipulation and plotting [@tidyverse].
-This file is also available as an RMarkdown document [@rmarkdown] at  <https://github.com/cboettig/bad-forecast-good-decision>.
-
 ```{r setup, message=FALSE}
@@
 ```
-
-Our decision problem is defined as follows: 
-A manager seeks to maximize the sum of the utility derived from such a harvest
-and such a state, $U(X_t,H_t)$, over all time, subject to discount rate $\delta$:
-
-\begin{equation}
-\sum_{t=0}^{t=\infty} U(X_t, H_t) \delta^t
-\end{equation}
-
-While in principle the utility could reflect many things, including the cost of fishing, market responses to supply and demand, the value of recreational fishing, the intrinsic value fish left in the sea [see @Halpern2013], for simplicity we will assume utility is merely a linear function of the harvest quota set by the manager, i.e.
-a fixed price $p$ per kilogram of fish harvested: $U(X_t, H_t) = p \min(H_t, X_t)$ (noting that realized harvest cannot exceed the available stock).
-As units are considered already non-dimensionalized in this example, without loss of generality we will set $p=1$.
-This problem is already well studied, and it is worth noting that even under such a pessimistic assumption, the optimal strategy still seeks to sustain the fish population indefinitely; as @Clark1973 shows for the deterministic function $f$ and @Reed1979 extended to the stochastic case.
-Given the function $f$ with known parameters, it is straight forward to determine the optimal harvest policy by stochastic dynamic programming (SDP, see @Mangel1985; @Marescot2013).
-
-To solve the decision problem using SDP, we define the state space and the action space on a discrete grid of 240 points.
-The maximum state is set well above the largest carrying capacity used in the models, which limits the influence of boundary effects introduced by the transformation to a discrete, finite grid.
-Available harvest actions match the state space, effectively allowing any harvest level to be possible.
-
-```{r model_definitions}
-states <- seq(0,24, length.out = 240)
-actions <- states
-obs <- states
-```
-
-The utility (reward) of an action is set to the (realized) harvest (e.g.
-a fixed price per unit fish, with no cost applied to harvest effort).  Future
-utility is discounted by fixed factor of $0.99$. Classic maximum sustainable 
-yield models [@Schaefer1954] ignore discounting, while modern economic optimization
-models insist on it, so a small discount conforms to the latter while reasonably
-approximating the former.  The qualitative conclusion is not sensitive to the 
-discounting rate.
-
-```{r}
-reward_fn <- function(x,h) pmin(x,h)
-discount <- 0.99
-```
-
-Alternate reward functions with varying price and cost structures are possible but do not qualitatively impact the conclusions.
-This reward function is a limiting case of any more complex reward, and corresponds both to classic work that does not model utility explicitly [e.g.
-MSY theory, @Schaefer1954], as well as explicit assumptions typically made in more recent models [@Reed1979].
-Moreover, using a simple reward function makes it clear that model that performs best is not doing so merely because of particular features baked into the a carefully chosen reward rule.
-
-# Ecological Models
-
-The manager chooses between two different logistic growth models, each of which
-can be thought of as an approximation to the underlying "true" population model:
-
-\begin{equation}
-f_i(Y) = Y + Y r_i \left( 1 - \frac{Y}{K_i} \right)
-\end{equation}
-
-Model 1 has $r_1 = 2$, $K = 16$, and $\sigma_1 = 0.05$.
-Model 2 has $r_2 = 0.5, K_2 = 10$, and $\sigma_2 = 0.075$.
-
-Meanwhile, the true population growth rate is simulated using a function with
-non-linear per-capita growth:
-
-\begin{equation}
-f_i(Y) = Y + Y^4 r_i \left( 1 - \frac{Y}{K_i} \right)
-\end{equation}
-
-with $r_3 = 0.002$, $K_3 = 10$ and $\sigma_3 = 0.05$:
 
@@
 
+n_models <- 32
+rs <- seq(0.4, 2, length.out = n_models)
+Ks <- seq(8, 10*8/5, length.out = n_models)
+sigmas <- seq(0.05, 1.5*0.05, length.out = n_models)
+pars <- data.frame(r = rs, K = Ks, sigma = sigmas)
+
+closure <- function(r, K){
+  function(x, h = 0){
+  s <- pmax(x - h, 0)
+  s + s * (r * (1 - s / K) )
+  }
+}
+
+models <- lapply(1:n_models, function(i) closure(pars[i,"r"], pars[i,"K"]))
+
 ## gather models together, indicate true model
 sigma_g <- 0.05
-models <- list("1" = f1, "2" = f2, "3" = f3)
-model_sigmas <- c(sigma_g, 1.5 * sigma_g, sigma_g)
-true_model <- "3"
+models <- c(f3, models)
+model_sigmas <- c(sigma_g, pars$sigma)
+true_model <- 1
 ```
@@
                                                       model_sigmas[[i]]))
-names(transitions) <- c("1", "2", "3")
+names(transitions) <- as.character(seq_along(models))
 
@@
 
-# Optimal control solutions
+# Plot Models
 
-We use value iteration to solve the stochastic dynamic program [@Marescot2013; @MDPtoolbox] for each model.
-This determines the optimal harvest policy for each possible state, given each model.
-Because this step is the most computationally intensive routine, we cache the results using memosization conditioned on the transition matrices [@memoise].
-Running this code with alternate transition matrices automatically invalidates that cache, reducing the risk of loading spurious results.
+```{r plot_models}
+model_set <- models
+names(model_set) <- c("true", as.character(round(Ks, 2)))
+d <- 
+  map_dfc(model_set, function(f) f(states) - states) %>%
+  mutate(state = states)
 
-```{r sdp, results="hide"}
-mdp <- memoise::memoise(mdp_value_iteration,
-                        cache = memoise::cache_filesystem("cache/"))
+  d %>% pivot_longer(names(model_set[-1]), "model") %>%
+  ggplot(aes(state, value, col=as.numeric(model))) +
+  geom_hline(aes(yintercept = 0), lwd=1) + 
+  geom_point(alpha = .5) + 
+  geom_line(aes(state, true), col = "red") +
+  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
+  ylab(bquote(f(x) - x)) + xlab("x") + viridis::scale_color_viridis()
+```
 
-policies <- 
-  map_dfr(transitions, 
-          function(P){
-    soln <- mdp(P, reward, discount = discount, 
-                epsilon = 0.01, max_iter = 2000, V0 = rep(0,dim(P)[[1]]))
-    escapement <- states - actions[soln$policy]
-    tibble(states, policy = soln$policy, escapement)
-    }, 
-  .id = "model")
+We can easily plot the optimal policy derived from each model, as shown
+in Fig3b.
+
+```{r plot_policies}
+fig3b <- policies %>%
+  ggplot(aes(states, escapement, col=model, lty=model)) + 
+  geom_line(lwd=2) + xlab("state")
 ```
@@
 
-am1 <- adaptive_management(transitions[1:2], reward, discount, model_prior = c(0.99, 0.01), x0 = x0, 
-               Tmax = 50, true_transition = transitions[[3]], 
-                epsilon = 0.001, max_iter = 2000)
+am1 <- adaptive_management(transitions[-1], 
+                           reward, 
+                           discount, 
+                           x0 = x0, 
+                           Tmax = 50, 
+                           true_transition = transitions[[true_model]], 
+                           epsilon = 0.001, 
+                           max_iter = 2000)
 ```
@@
 ```{r}
-am <- am1$df %>% mutate(belief = am1$posterior$V2)
-am %>%  ggplot(aes(time, states[state], col = belief)) + geom_line() + geom_point() +  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4])
+names(am1$posterior) <- Ks
+belief <- am1$posterior %>%
+  mutate(time = 1:dim(am1$posterior)[1]) %>% 
+  pivot_longer(as.character(Ks), "K", values_to = "belief") %>%
+  mutate(K = as.numeric(K))
+
+belief %>% filter(time %in% seq(1, 11, by = 2)) %>%
+  ggplot(aes(K, belief, col=time, group=as.character(time))) +
+  geom_line()
+
+belief %>% filter(K < 10) %>%
+  ggplot(aes(time, belief, col=K, group=as.character(K))) +
+  geom_line()
+
+```
+
+```{r}
+am <- am1$df %>% mutate(belief = am1$posterior[,1])
+am %>%  ggplot(aes(time, states[state], col = belief)) + 
+  geom_line() + 
+  geom_point() +
+  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4])
 ```
@@
 
-Using the transition matrices directly, we can examine what each model would 
-have forecast the future stock size to be in the following year when no fishing
-occurs (note that for each model, we use the transition matrix that corresponds
-to 'no fishing', `model[[state_index, ,1]]`) (Fig 1a, main text).
-
-The transition matrices give the full (discretized) probability distribution,
-from which we can easily calculate both the expected value and the 95% 
-confidence interval.
-
 ```{r stepahead_unfished}
@@
 ```
-
-We also look at the forecast each model makes when implementing the corresponding
-optimal harvest:
 
@@
 # Proper scores
-
-It is straight forward to apply the proper scoring formula of @Gneiting2007 
-based on the first two moments of the distribution to score the respective
-forecasts under both the unfished and actively managed scenario sfor each model:
 
@@
 
-```{r plot_models}
-d <- 
-  map_dfc(models, function(f) f(states) - states) %>%
-  mutate(state = states)
-
-fig3a <- 
-  d %>% pivot_longer(names(models), "model") %>%
-  ggplot(aes(state, value, col=model, lty=model)) +
-  geom_hline(aes(yintercept = 0), lwd=1) + 
-  geom_line(lwd=2, show.legend = FALSE) + 
-  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
-  ylab(bquote(f(x))) + xlab("x")
-```
-
-We can easily plot the optimal policy derived from each model, as shown
-in Fig3b.
-
-```{r plot_policies}
-fig3b <- policies %>%
-  ggplot(aes(states, escapement, col=model, lty=model)) + 
-  geom_line(lwd=2) + xlab("state")
-```
-
 # References

### 2021-02-09 manuscript/manuscript.Rmd
:pen:

@@
 is the result of our imperfect knowledge of the processes involved, as expressed by the structure and
-parameters of mathematical models used to approximate those processes (model uncertainty). Model uncertainty can be 
-reduced by gathering additional data, which may rule out certain models or parameter values as implausible. 
+parameters of mathematical models used to approximate those processes (model uncertainty). Unlike the other forms of
+uncertainty, model uncertainty can be 
+reduced by gathering additional data, which may rule out certain models or parameter values as implausible (). 
 When model parameters are estimated directly from available data, there is a risk that the best-fitting models may 
@@
 A wide range of paradigms are available for approaching the issue of decision-making under uncertainty.
-These approaches can roughly be divided into two groups: the first group treats the issue of model uncertainty independently from the decision itself, while the second integrates the process of reducing model uncertainty into the process of decision making.
-There are a wide range of techniques within each group.
+These approaches can roughly be divided into two groups: the first group treats the issue of model uncertainty independently from the decision itself, while the second integrates the process of reducing model uncertainty into the process of decision making to maximize the value of some objective.
+There are a wide range of techniques within each, and it is also possible to blend approaches.
+The key distinction is that methods in the first group do not involve any direct consideration of the possible actions or the utility that may result from those actions in how they select models [statistical approaches such as information criteria @BurnhamAnderson; and in particular, forecasting evaluation @Clark2001; @Dietze2018], while those in the second group require a more explicit statement of possible actions and the desired objectives [Decision theoretic approaches, for which @Polasky2011 provides an excellent and accessible review].
+I illustrate how the most promising techniques from each of these groups would be applied to this simple problem, and demonstrate that it both cases they lead us away from the model that produces the most desirable decisions towards worse outcomes.
+In retrospect, it will become obvious that neither these nor any other widely applied methods will select the model that leads to the best outcomes from the set of models considered.
 
@@
 
-Like many decision problems, this one appears to hinge on having an accurate forecast: if we can predict to what size the stock will increase next year, $X_t+1$, and we know the current stock, $X_t$, then we can safely harvest $X_{t+1} - X_t$.
+Like many decision problems, the task of setting a sustainable harvest quota appears to hinge on having an accurate forecast: if we can predict to what size the fish stock will increase next year, $X_t+1$, and we know the current stock, $X_t$, then we can safely harvest $X_{t+1} - X_t$.
 Overestimating or underestimating such recruitment will result in over-harvesting or under-harvesting, respectively.
 Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of over-fitting by comparing model predictions to later observations that were not used to estimate the model.
-@Gneiting2007 provides a rigorous proof for "strictly proper" scoring criteria, which have the desirable property which no model predicting the distribution of future outcomes, $Q(x)$ can achieve a better average score than the true model $P(x)$ (That is, unlike likelihood or other goodness-of-fit scores, it is impossible to overfit when conditioning on a strictly proper score).
+@Gneiting2007 provides a rigorous proof for "strictly proper" scoring criteria, which have the desirable property which no model predicting the distribution of future outcomes, $Q(x)$ can achieve a better average score than the true model $P(x)$.
+That is, unlike likelihood or other goodness-of-fit scores, it is impossible to overfit when conditioning on a strictly proper score -- since no model model can out-perform the true model.
+Not that strictly proper scoring rules score _probabilistic forecasts_ and not just point predictions, favoring models which accurately reflect the uncertainty over those which under-estimate it.
+These features have made proper scoring rules for probabilistic forecasts a successful and popular approach for addressing model uncertainty in many other areas [@Gneiting2014; @Raftery2016] and a promising tool for evaluation of ecological forecasts [@Dietze2018].
 
-Iterative updating, with model priors.
+I illustrate the process of model selection by strictly proper scoring rules using two scenarios.
+
+**Decision-Theoretic Approaches**
+
+Decision-theoretic approaches include scenario analysis, resilience thinking, optimal control and related methods [@Polasky2011], which have a long history in ecology and particularly in fisheries management.
+One of the most influential of these is _adaptive management_ [sensu @Walters1978], in which a manager seeks to both reduce uncertainty over time while also achieving the best outcomes given current knowledge.
+While the term today is frequently used in a looser sense, adaptive management as originally developed can be considered an example of optimal control under model uncertainty
+
+[@Walters1981; @Smith1981]
+
+# A way forward?
+
+note that the model set considered here already contains a model which captures the key feature.
+
+- Greedy strategy will not work.  Could stick with a model for 50 years then switch...
+- A mechanistic understanding of the decision proces, not the ecological one.
+Analyzing the decision problem, we know that in this situation, the optimal stategy depends only on the location of the peak of the growth rate.
+This could suggest a very different way forward
+-
 

### 2021-02-16 manuscript/appendix.Rmd
:pen:

@@
   geom_point() +
-  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4])
+  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4]) +
+  ylab("stock size")
 ```
@@
 ```{r plot_sims}
-fig2a <- 
+plot_ecology <- 
   sims %>% 
@@
   ylab("state") + 
-  ggtitle("A") + labs(subtitle = "Ecological")
+  labs(subtitle = "Ecological")
 ```
@@
 
-fig2b <- 
+plot_econ <- 
 npv_df %>%
@@
 ```
+```{r}
+# tabular comparisons
+npv <- npv_df %>% group_by(model) %>% summarize(npv = max(mean_npv))
+
+am_npv <- sum(am$value * discount ^ am$time)
+am_npv / npv[[1,"npv"]]
+
+mean_state <- sims %>% group_by(model) %>% summarize(state = mean(state))
+mean(states[am$state]) / mean_state[[1, "state"]]
+```
 
@@
 
-fig3a <- 
+plot_models <- 
   d %>% pivot_longer(names(models), "model") %>%
@@
   geom_hline(aes(yintercept = 0), lwd=1) + 
-  geom_line(lwd=2, show.legend = FALSE) + 
+  geom_line(lwd=2) + 
   coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
@@
 ```{r plot_policies}
-fig3b <- policies %>%
+plot_policies <- policies %>%
   ggplot(aes(states, escapement, col=model, lty=model)) + 

### 2021-02-16 manuscript/appendix_B.Rmd
:pen:

@@
 ```{r}
-names(am1$posterior) <- Ks
-belief <- am1$posterior %>%
-  mutate(time = 1:dim(am1$posterior)[1]) %>% 
-  pivot_longer(as.character(Ks), "K", values_to = "belief") %>%
-  mutate(K = as.numeric(K))
-
-belief %>% filter(time %in% seq(1, 11, by = 2)) %>%
-  ggplot(aes(K, belief, col=time, group=as.character(time))) +
-  geom_line()
-
-belief %>% filter(K < 10) %>%
-  ggplot(aes(time, belief, col=K, group=as.character(K))) +
-  geom_line()
-
-```
-
-```{r}
 am <- am1$df %>% mutate(belief = am1$posterior[,1])
@@
   scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4])
+```
+
+```{r}
+
+am_npv <- sum(am$value * discount ^ am$time)
+am_npv / npv[[1,"npv"]]
 ```
@@
 ```
-
-# Simulations and step-ahead forecasts
-
-We simulate fishing dynamics under the optimal policy for each model, using a simple helper function from the `mdplearning` package.
-Because growth dynamics are stochastic, we perform 100 simulations of each model from identical starting condition to ensure results are not the result of chance alone.
-
-```{r simulations}
-library(mdplearning)
-Tmax <- 100
-x0 <- which.min(abs(states - 6))
-reps <- 100
-set.seed(12345)
-
-## Simulate each policy reps times, with `3` as the true model:
-simulate_policy <- function(i, policy){
-  mdp_planning(transitions[[true_model]], reward, discount,
-           policy = policy, x0 = x0, Tmax = Tmax) %>%
-    select(value, state_index = state, time, action_index = action)  %>% 
-    mutate(state = states[state_index])
-}
-
-sims <- 
-  map_dfr(names(transitions), 
-          function(m){
-            policy <- policies %>% filter(model == m) %>% pull(policy)
-            map_dfr(1:reps, simulate_policy, policy = policy, .id = "reps")
-          },
-          .id = "model"
-         )
-```
-
-```{r stepahead_unfished}
-stepahead_unfished <- sims
-stepahead_unfished$state_index <- rep(sims$state_index[sims$model == "1"],3)
-
-stepahead_unfished <- stepahead_unfished  %>% 
-  filter(model != "3") %>% 
-  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
-  rowwise() %>%
-  mutate(expected = transitions[[model]][state_index, , 1]  %*% states,
-         var = transitions[[model]][state_index, , 1]  %*% states ^ 2 - expected ^ 2,
-         low = states[max(which(cumsum(transitions[[model]][state_index,,1]) < 0.025)) ],
-         high = states[min(which(cumsum(transitions[[model]][state_index,,1]) > 0.975)) ],
-         true = states[next_state ]) 
-```
-
-```{r stepahead_fished}
-stepahead_fished <- sims %>% 
-  filter(model != "3") %>%
-  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
-  rowwise() %>%
-  mutate(prob =  transitions[[model]][state_index, next_state, action_index],
-         expected = transitions[[model]][state_index, , action_index]  %*% states,
-         var = transitions[[model]][state_index, , action_index]  %*% states ^ 2 - expected ^ 2,
-         low = states[max(which(cumsum(transitions[[model]][state_index,,action_index]) < 0.025)) ],
-         high = states[min(which(cumsum(transitions[[model]][state_index,,action_index]) > 0.975)) ],
-         true = states[next_state]) %>%
- select(time, model, true, expected, low, high, var, prob, reps)
-```
-
-# Proper scores
-
-```{r proper_scores, message = FALSE}
-# Gneiting & Raferty (2007), eq27
-scoring_fn <- function(x, mu, sigma){ -(mu - x )^2 / sigma^2  - log(sigma)}
-
-stepahead_unfished <- stepahead_unfished %>%
-  mutate(sd = sqrt(var),
-         score = scoring_fn(expected, true, sd))
-
-stepahead_fished <- stepahead_fished %>%
-  mutate(sd = sqrt(var),
-         score = scoring_fn(expected, true, sd))
-```
-
-```{r}
-predictions <- 
-  stepahead_unfished %>% 
-  select(time, model, reps, expected, low, high, true, score) %>% 
-  mutate(scenario = "A_unfished") %>% 
-  bind_rows(stepahead_fished  %>% 
-          select(time, model, reps, expected, low, high, true, score) %>% 
-          mutate(scenario = "B_fished")
-  ) %>% 
-  mutate(model = as.character(model))
-```
-
-# Plotting
-
-Having computed a single data frame of simulation data under both optimally fished (according to each model) and unfished scenarios for each model, along with the proper scores for the step-ahead forecasts a la @Gneiting2007, we have all the results necessary to generate the figures presented in the main paper.
-
-```{r fig1a}
-fig1ab <- predictions %>%
-  filter(reps == "2", time < 10) %>%
-  ggplot(aes(time, col = model, fill = model)) + 
-  geom_point(aes(y = expected)) + 
-  geom_errorbar(aes(ymin = low, ymax = high)) +
-  geom_point(aes(y = true), pch = "*", size = 12, 
-             alpha = 0.6) + 
-  facet_wrap(~scenario, ncol = 1, labeller =as_labeller(c(
-    A_unfished = "A. Without fishing",
-    B_fished = "B. Managed harvest"))) +
-  ylab("stock size")
-
-```
-
-Likewise we can plot the data on proper scores associated with each prediction of each model:
-
-```{r}
-fig1cd <- predictions %>%
-  ggplot(aes(x = score, group = model, fill = model)) +
-  geom_histogram(binwidth = 2, show.legend = FALSE) +
-  coord_cartesian(xlim = c(-100, 1), ylim = c(0,4000)) +
-  xlab("Proper score") +
-  facet_wrap(~scenario, ncol = 1, labeller =as_labeller(c(
-    A_unfished = "C.",
-    B_fished = "D."))) + 
-  scale_x_continuous(breaks = c(-100, 0)) +
-  theme(axis.text.y = element_blank(),
-        plot.margin = margin(0, 0, 0, 0, "cm"),
-        panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank()
-        )
-
-```
-
-## Forecast ecological and economic performance
-
-The ecological performance of the model can easily be visualized by plotting
-the results of each simulation.
-
-```{r plot_sims}
-fig2a <- 
-  sims %>% 
-  group_by(model, time) %>%
-  summarise(mean_state = mean(state), sd = sd(state), .groups = "drop") %>%
-  filter(time < 25, model != "3") %>%
-  ggplot(aes(time, mean_state)) + 
-  geom_line(aes(col = model), lwd=1.5, show.legend = FALSE) +
-    geom_ribbon(aes(ymin = mean_state - 2*sd, 
-                    ymax = mean_state + 2*sd,
-                    fill = model),
-                alpha = 0.2, show.legend = FALSE) +
-  ylab("state") + 
-  ggtitle("A") + labs(subtitle = "Ecological")
-```
-
-To plot the economic value over time, we must sum up the discounted values
-at each time step, and then average over replicate simulations of each model:
-
-```{r plot_npv}
-##  Net Present Value accumulates over time
-npv_df <- sims %>% 
-  group_by(model, reps) %>%
-  mutate(npv = cumsum(value * discount ^ time)) %>%
-  group_by(time, model)  %>% 
-  summarise(mean_npv = mean(npv), .groups="drop") %>% 
-  arrange(model, time)
-
-optimal <- select(filter(npv_df, model == "3"), time, mean_npv)
-
-fig2b <- 
-npv_df %>%
-  filter(model != "3", time %in% seq(1,100, by = 5)) %>%
-  ggplot(aes(time, mean_npv)) +
-  geom_line(data = optimal, lwd = 1.5, col = "grey20") +
-  geom_point(aes(col=model), size = 4, alpha = 0.8) + 
-  ylab("Net present value") + xlab("time")  + 
-  ggtitle("B.") + labs(subtitle = "Economic")
-
-```
-
-Plotting the model functions themselves are shown in Fig 3a:
-
-# References

### 2021-02-16 manuscript/manuscript.Rmd
:pen:

@@
 ---
-title: ''
+title: 'Learning to be wrong:'
 author:
@@
 abstract: |
-
   Model-based forecasts have been enabled by recent explosion of data and computational methods and spurred on by decision-makers appetite for forecasts in everything from elections to pandemic response.
@@
 Using a classic example from fisheries management, I demonstrate that selecting the model that produces the most accurate and precise forecast can lead to decidedly worse outcomes.
-This situation can arise whenever the models are only approximations of the reality, underscoring the risk of evaluating models only by prediction accuracy and not decision objectives using formal tools of decision theory.
 
@@
 
-acknowledgements: |
-  This work was supported in part by NSF CAREER (#1942280) and computational resources from NSF's XSEDE Jetstream (DEB160003) and Chameleon cloud platforms.
-
-## must be one of: pnasresearcharticle (usual two-column layout), pnasmathematics (one column layout), or pnasinvited (invited submissions only)
-pnas_type: pnasresearcharticle
-csl: pnas.csl
-
 ## change to true to add optional line numbering
@@
 ---
+
+```{r, message=FALSE, warning=FALSE, include = FALSE}
+## Plotting themes, colors, and fonts 
+## aesthetics choices only, all can be omitted without consequence
+library(ggthemes)
+library(hrbrthemes)
+library(ggplot2)
+library(Cairo)
+library(extrafont)
+library(patchwork)
+library(styler)
+extrafont::loadfonts(quiet = TRUE)
+ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
+
+scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
+scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
+pal <- ggthemes::solarized_pal()(8)
+txtcolor <- "#586e75"
+```
+```{r, include=FALSE}
+rmarkdown::render("appendix.Rmd")
+knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, warning = FALSE, dev="cairo_pdf")
+```
 
@@
 uncertainty, model uncertainty can be 
-reduced by gathering additional data, which may rule out certain models or parameter values as implausible (). 
+reduced by gathering additional data, which may rule out certain models or parameter values as implausible[^1]. 
 When model parameters are estimated directly from available data, there is a risk that the best-fitting models may 
@@
 decision outcomes in general remains an open problem, as is the challenge of what to do about it.
+
+[^1]: While some authors distinguish between model uncertainty and parameter uncertainty, I will refer collectively to
+any uncertainty that arises from our imperfect knowledge of the system as "model uncertainty".
+Uncertainty in model structure can often be reflected by appropriate parameterization, just as models with different parameter values can be treated distinctly.
+Whereas stochasticity is inherent to the process (or rather, to our choice of state variables, @Boettiger2018) and measurement uncertainty can only be reduced by more accurate tools for measurement, model uncertainty is reducible by learning from additional data.
 
@@
 
-A manager seeks each year to select the harveset $H_t$ which will maximize the sum
+A manager seeks each year to select the harvest $H_t$ which will maximize the sum
 of the utility derived from such a harvest and such a state, $U(X_t,H_t)$, over all time,
@@
 \begin{equation}
-\sum_{t=0}^{t=\infty} U(X_t, H_t) \delta^t
+\sum_{t=0}^{t=\infty} U(X_t, H_t) \delta^t \label{utility}
 \end{equation}
@@
 Ecological processes are much too complex to ever be modeled exactly.
-For illustrative purposes, we will thus assume the "true" process to be given by a third model, which is unknown to the decision-maker, but similar enough to at least one of the candidate models might be considered a reasonable approximation.
+For illustrative purposes, we will thus assume the "true" process to be given by Model 3, which is unknown to the decision-maker, but similar enough to at least one of the candidate models might be considered a reasonable approximation:
 
@@
 
-# Managing Under Model Uncertainty
+The task of deciding whether Model 1 or Model 2 would be the better choice for decision making is 
+thus perhaps the simplest example of the much studied issue of model uncertainty that we can pose. 
+As in any real world scenario, neither model is the true model, but nevertheless this model set
+contains a good enough approximation of the true model to make good decisions.  However, any of
+the well-developed approaches for decision-making under model uncertainty will prefer Model 2 over
+Model 1, despite the fact that the optimal policy under Model 2 leads to much worse outcomes
+ecologically and economically.
+
+# Methods for Managing Under Model Uncertainty
 
@@
 
-**Forecasts and Strictly Proper Scoring**
+## Forecasts and Proper Scores
 
@@
 Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of over-fitting by comparing model predictions to later observations that were not used to estimate the model.
-@Gneiting2007 provides a rigorous proof for "strictly proper" scoring criteria, which have the desirable property which no model predicting the distribution of future outcomes, $Q(x)$ can achieve a better average score than the true model $P(x)$.
+@Gneiting2007 provides a rigorous proof for "proper" scoring criteria, which have the desirable property which no model predicting the distribution of future outcomes, $Q(x)$ can achieve a better average score than the true model $P(x)$.
 That is, unlike likelihood or other goodness-of-fit scores, it is impossible to overfit when conditioning on a strictly proper score -- since no model model can out-perform the true model.
@@
 I illustrate the process of model selection by strictly proper scoring rules using two scenarios.
+In Scenario A (passive observation) the fish stock is unharvested and allowed to recover towards carrying capacity (as simulated under our "true" model, Model 3) while comparing the observed stock size in each subsequent time step to the distribution predicted under Model 1 and Model 2 respectively [Fig].
+The mean, $\mu$ and variance, $\sigma$ of the forecast are compared against the true observation $x$ using a Proper scoring rule of @Geiting2007,
 
-**Decision-Theoretic Approaches**
+\bein{equation}
+-(\mu - x )^2 / \sigma^2  - \log(\sigma) \label{proper}
+\end{equation}
 
-Decision-theoretic approaches include scenario analysis, resilience thinking, optimal control and related methods [@Polasky2011], which have a long history in ecology and particularly in fisheries management.
-One of the most influential of these is _adaptive management_ [sensu @Walters1978], in which a manager seeks to both reduce uncertainty over time while also achieving the best outcomes given current knowledge.
-While the term today is frequently used in a looser sense, adaptive management as originally developed can be considered an example of optimal control under model uncertainty
-
-[@Walters1981; @Smith1981]
-
-# A way forward?
-
-note that the model set considered here already contains a model which captures the key feature.
-
-- Greedy strategy will not work.  Could stick with a model for 50 years then switch...
-- A mechanistic understanding of the decision proces, not the ecological one.
-Analyzing the decision problem, we know that in this situation, the optimal stategy depends only on the location of the peak of the growth rate.
-This could suggest a very different way forward
--
-
-Yet the model that leads to the best decisions is not always the model that makes the most accurate forecasts, as I illustrate here.
-Surprisingly, this can even happen when a decision is derived directly from a complex optimization routine of a probabilistic predictive models [@Marescot2013].
-Reality is complex; even our best models can only ever be approximations of underlying processes.
-Here, I use a classic, well-understood example from fisheries management [@Schaefer1954; @Clark1973; @Reed1979] to illustrate both the paradox of how a model with the worst forecast provides the best decision outcomes, as well as show how we can avoid selecting models that are poorly suited for management by considering the management context more explicitly.
-These results underscore that in choosing the best model for decision-making, it can be more important to capture a single key feature of the process than it is to make the most accurate prediction about future states.
-
-<!--
-Many ecological management problems are sequential decision problems, in which each year (or other interval) a manager must observe the state of the system and choose a course of action to maximize long term objectives.
-Such problems inherently depend on forecasts: each possible action can result in a different forecast for the future state, typically reflecting some uncertainty as well.
-The utility the manager derives may depend on both the choice of action and the state of the system, reflecting the costs and benefits associated with each.
-Sequential decision-making problems are distinguished by the need to think more than one move ahead.
-For instance, harvesting as many fish as possible in year one may maximize the market value that year, but if too few fish are left to reproduce then harvest in future years will suffer.
-The same calculus of thinking ahead frequently applies to rebuilding species populations as well [e.g.
-@Lambert;  @Chades2008].
--->
-
-```{r, message=FALSE, warning=FALSE, include = FALSE}
-## Plotting themes, colors, and fonts 
-## aesthetics choices only, all can be omitted without consequence
-library(ggthemes)
-library(hrbrthemes)
-library(ggplot2)
-library(Cairo)
-library(extrafont)
-library(patchwork)
-library(styler)
-extrafont::loadfonts(quiet = TRUE)
-ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
-
-scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
-scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
-pal <- ggthemes::solarized_pal()(6)
-txtcolor <- "#586e75"
-```
-
-```{r, include=FALSE}
-rmarkdown::render("appendix.Rmd")
-knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, warning = FALSE, dev="cairo_pdf")
-```
-
-# Results
-
-## Forecast performance
+for each prediction over 100 replicate simulations of 100 time steps each [Fig].
 
@@
 
-The one-step-ahead prediction performance of each model in a simulation of an un-fished environment show consistently better performance of Model 2 (Fig.
-1A).
-Model 1 predictions appear far too optimistic, with the true value falling well below the 95% confidence intervals.
-In contrast, all observed values fall easily within the confidence intervals produced by model 2.
+In Scenario B (actively harvest), I have first solved for the optimal management strategy for both Model 1 and Model 2 [Fig], which for small noise and concave functions with linear reward structure can be done analytically [see proof in @Reed1979], or can be solved more generally by stochastic dynamic programming [see review by @Marescot2013, details in Appendix].
+Under this scenario, replicate simulations of the stock are harvested at each time step using the optimal quota dictated by either Model 1 and Model 2.
+The resulting stock sizes in the time-step following this harvest are once again compared to the probabilities predicted by each model using Eq \eqref{proper}.
+Model 2 unequivocally outperforms Model 1 in both scenarios of passive observation and active harvest.
 
-Predictive performance of the un-fished population does not give us the full picture, since it reflects predictive accuracy only in the region of the true carrying capacity, while an actively harvested stock will be at a lower size.
-The model that predicts the equilibrium size may not be the one that best forecasts stock recovery.
-This comparison does not reflect the influence of any decisions that might be made based on the model forecast.
-To address these concerns, we consider a second scenario where our fishery is managed according to the optimal harvest predicted by each model in turn.
-Each year the model produces both a forecast and a decision about the harvest quota.
-(The mechanics of determining a harvest quota given the model follow standard methods for Markov Decision Process, which depend on step-ahead predictions [@Marescot2013], see appendix for details.) We then implement that harvest and compare the observed stock size the following year to that which the model has predicted, (Fig.
-1B).
-Again we observe that the observations under model 1 consistently fall well outside of the 95% confidence intervals it predicts, while under model 2, stock sizes consistently fall within the predicted intervals.
-Once again, model 2 shows a higher forecast accuracy while model 1 appears consistently over-optimistic.
-
-Interest in assessing probabilistic forecasts has led to the development of rigorous methods of scoring [e.g.
-@Gneiting2014; @Raftery2016].
-A scoring rule is "proper" if it has the convenient property that no other prediction can achieve a higher score, on average, than we would if we used the true distribution [@Gneiting2007].
-The distribution of proper scores across 100 replicate simulations for both un-fished and actively managed scenarios (Fig.
-1C, 1D respectively) show consistently higher scores of model 2, using a proper scoring rule (Eq (27) of Ref.
-@Gneiting2007).
-
-## Ecological and economic performance
-
-Given this evidence, model 2 clearly provides the more accurate forecast and we would no doubt conclude that model 2 was thus a better approximation of the true model and thus the better choice to inform decision making about harvest quotas.
-Yet if we revisit our experiment of managing the fishery under each model in turn, and focus not on _predictive accuracy_ but on _ecological_ and _economic_ outcomes, it quickly becomes clear that model 1 gives much better results (Fig.
-2A).
-For comparison, we have also included the results of optimal management given the true model.
-Despite its optimistic predictions, model 1 does not result in over-fishing, but holds the stock near the same level as the optimal management strategy.
-In contrast, model 2 suppresses the stock to a much lower level.
-The over-fishing in model 2 is not economically efficient either (Fig.
-2B).
-The net present value of the fishery, as calculated as the cumulative, discounted value of the harvest (assuming a fixed unit price for fish with negligible cost for harvest, see appendix) under the fishing regime of model 1 falls precisely along that of the optimal solution, while the value derived under model 2 is consistently lower.
+Despite the clearly superior predictive accuracy of Model 2 in both scenarios, the outcomes from management under Model 2 are substantially worse.
+We can assess such outcomes in less abstract terms than forecasting skill, such as economic value (in dollars) or the ecological value (unharvested biomass).
+In our simple formulation of the decision problem, the "utility" the manager seeks to maximize is simply the economic value (net present value: the discounted sum of all profits from future harvests, Eq \eqref{utility}) of harvested fish.
+This formulation ignores any utility provided by fish that are not harvested, beyond their contribution to future potential harvests.
+While it is possible to include such contributions directly in the utility function being optimized [e.g.
+@Halpern2013], even without doing so, Model 1 maintains both a higher unharvested biomass and also leads to higher economic returns throughout [Fig ].
 
@@
 ```
+
+Had we been able to include Model 3 in our forecast comparisons, it would equal or outperform the forecasting skill of both Model 1 and Model 2 (as guaranteed by the theorem of @Gneiting2007), while also matching or out-performing their economic utility (as guaranteed by the theorem of @Reed1979).
+In practice, we never have access to the generating model.
+
+One obvious limitation in this comparison is that scenario B treats each model as fixed over the entire course of the simulation.
+Research has long emphasized the importance of learning and adaption in the face of new data whenever we are dealing with model uncertainty [@Polasky2011].
+A related limitation is that in Scenario B, decisions were based either on assuming Model 1 is correct or assuming Model 2 is correct.
+A more robust approach [e.g.
+@Walters1978] incorporates the uncertainty over models directly into the decision calculations by integrating over the probability each model $M$ (or parameter value) is correct when calculating the utility; $U(x_t, a_t) = \int U(x_t, a_t | M) P(M)$.
+The probability assigned to each model can then be updated after each subsequent action in light of the resulting outcomes [@Ludgwig1982, @Smith1981].
+While a the wide range of strategies for such iterative updating are known [@Polasky2011]; all will fail to select the higher-performing of the two simple models considered  here.
+
+<!-- replacing step-ahead with n-step ahead would not matter-->
+
+## Decision-Theoretic Approaches
+
+Decision-theoretic approaches include scenario analysis, resilience thinking, optimal control and related methods [@Polasky2011], which have a long history in ecology and particularly in fisheries management.
+One of the most influential of these is _adaptive management_ [sensu @Walters1978], in which a manager seeks to both reduce uncertainty over time while also achieving the best outcomes given current knowledge.
+While the term today is frequently used in a looser sense, adaptive management as originally developed can be considered an example of optimal control under model uncertainty [@Walters1981; @Smith1981].
+In this approach, a manager assigns probabilities to each of the possible models under consideration, i.e.
+models 1 & 2.
+The expected utility of any action reflects both the intrinsic uncertainty over future states under either model (process uncertainty) and the manager's uncertainty over the choice of models (or parameter values).
+Unlike the forecast comparison, an adaptive management approach explicitly considers the utility function, Eqn \eqref{utility}, which it seeks to maximize by intergrating over model uncertainty.
+
+In passive adaptive management, the manager chooses the action which maximizes that expected future utility.
+After observing the consequences of the action, the manager updates the probabilities each model is correct, typically by Bayes rule [e.g.
+@Walters1981; @Smith1981; @Ludwig1982].
+In active adaptive management, the key difference is that the manager may chose an action which does not maximize the utility _given the current model uncertainty_, but instead, takes an action whose consequences will more quickly discriminate between the models and thus more rapidly reduce the uncertainty.
+After observing the outcome of this exploratory step, model probabilities are updated as before.
+Under some scenarios the cost of this 'exploration' step can be made up later by actions based on the more precise knowledge gained as a result ('exploitation').
+Alternative approaches such as scenario analysis can be considered approximations of this approach.
+A scenario analysis does not rely on quantitative estimates of model probabilities or expected utility, but still consider possible outcomes over a suite of models (scenarios), select actions to minimize the risk or or other suitable objective, and update or eliminate scenarios deemed implausible over time [@Polasky2011].
+
+Each of these approaches swiftly decides that model 1 in our example is implausible, and settles for making decisions driven almost exclusively by model 2.
+I illustrate this using a passive adaptive management strategy following classic examples for parameter [@Ludwig1982] or structural [@Smith1981] model uncertainty.
+To demonstrate that the behavior is not driven by failure to explore sufficiently, (which might be addressed by an active adaptive management), I will assign initial probability that model 2 is true at 1%.
+After a single iteration of learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct  [Fig].
+As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating [Fig , achieving a net present value over 100 time steps that only 31% that expected under management using Model 1 alone].
+If instead of considering only these two models we consider a whole suite of Gordon-Schaefer-type models with varying $r$ and $K$ parameters, learning is slightly slower but no less counter-productive: mean stock size is $46%$ and net present value is $20%$ of what would be achieved under Model 1   (Appendix, Fig S1).
+
+```{r}
+am <- am1$df %>% mutate(belief = am1$posterior$V2)
+am %>%  ggplot(aes(time, states[state], col = belief)) + 
+  geom_line() + 
+  geom_point() +
+  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4]) +
+  ylab("stock size")
+```
+
+# Discussion
+
+Given this simplified version of model uncertainty in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariable choose the other.
+The reason for this may appear obvious once we compare both curves to that of the underlying model, Model 3.
+Plotting the growth rate functions of each model, [Fig 4a], it is hardly surprising that no method which would prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3.
+Seeing this example, it is natural to pose any of four objections: (1) this problem is an artifact of overly-simplified models, (2) this problem is an artifact of the optimal control solution (3) the problem arises from different objectives, (4) the problem is already solved.
+I address these in turn.
+
+```{r fig.cap="Panel A: Population growth curves of each model.
+The positive equilibrium of each model occurs where the curve crosses the horizontal axis.
+Note that while Model 2 is a better approximation to the truth (Model 3), Model 1 better approximates the stock size which leads to maximum growth."}
+# Fig 4
+plot_models + plot_policies
+```
+
+Is this result an artifact of the simple models considered here?
+A more realistic approach would consider uncertainty over model parameters, as well as models with different structural forms, such as Beverton Holt [@], Ricker [@], Shepard [@].
+Such complexity would not be sufficient to resolve the paradox of model uncertainty illustrated here.
+Simplicity facilitates intuition: Model 1 performs best because it correctly approximates the stock size which achieves maximum growth rate under the true model.
+In contrast, each of those widely used recruitment relationships are symmetric, and their best-fitting parameter estimates will always have a maximum growth rate at a lower stock size.
+It is easy to point out that even this wider array of parameter and model values is too simple, since fisheries models in practice are much more complicated still (indeed, too complicated for the formal integration over model uncertainty considered here).
+It is indeed true that either the forecasting or adaptive management example would work as expected if the set of candidate models included the true model, or at least a functional form that could better approximate the peak of the true model while also fitting better than model 2.
+But we never have the true model in real systems.
+If the candidate models are only caricatures, so too is the "true" model.
+The candidate models considered here are no doubt much closer approximations to model 3 (all being single species, unstructured in age, stage, time or space, etc) than any model used in practice is to reality.
+Morever, a more accurate model is not required to reach optimal outcomes -- Model 1 already provides this by correctly approximating the key feature in this problem, even while it is wrong about much else.
+How do proceed when our best methods fail in such a simple case?
+
+<!-- move to appendix?  too confusing/too long?-->
+Is this result an artifact of the optimal control?
+The analysis has relied on optimal control theory to determine the optimal decision associated with each model or distribution of models.
+The resulting policies are mathematically optimal under the assumptions given [see @Reed1979 for stochastic harvest problem in particular and @Marescot2013 for discussion of Markov Decision Problems more generally].
+In most realistic decision problems, model complexity currently precludes the calculation of an optimal control solution at all ["the curse of dimensionality", e.g.
+@Polasky2011; @Marescot2013].
+Under the assumptions of any of the models considered here, the optimal control solution leads to what is known as a "constant escapement" policy, which seeks to bring biomass to the target escapement level, $B_{MSY}$ as quickly as possible, $H_t = \min(X_t - B_{MSY}, 0)$.
+If the assumptions are not met, a constant-escapement policy may not be optimal, as in the case when observations of the stock $X_t$ are not perfect [@pomdp-intro].
+While constant escapement is used as a basis of management in some real world cases such as salmon fisheries, most marine fisheries are based on a target of "constant mortality", ($F_{MSY}$).
+Notably, a constant mortality policy, implemented as $H_t = F_{MSY} X_t$, would not perform better under Model 1 than Model 2.
+Constant mortality is optimal only under deterministic growth at equilibrium (where the policy is the same under constant escapement).
+Does that mean the result is an artifact of assuming constant escapement rather than constant mortality as the basis for management?
+No, it does not.
+It is just as easy to construct a model under which the constant mortality matches the constant mortality of the true model while being a poor predictor.
+Moreover, using a non-optimal policy would always beg the question of whether the poor outcomes were merely the result of using a non-optimal policy rather than a consequence of model uncertainty.
+
+The optimal policy does not depend equally on all aspects of the underlying model.
+In this simple case, the policy depends only on the biomass which provides the maximum growth rate, and it is precisely this feature which allows Model 1 to match the policy of the true model despite over-estimating both the per-capita growth rate and unharvested equilbrium stock size so dramatically.
+The ability to capture certain key features rather than replicate every detail is the objective of any modeling effort, and the reason why scientific decision making is possible in the first place.
+Many optimal control solutions are similarly insensitive to getting all the details right.
+If we seek to improve decisions in face of model uncertainty, we must be able to exploit this aspect to find models fit to purpose, and be cautious about employing methods that would reject such models out of hand.
+
+Does this problem merely reflect the fact that different objective functions lead to different solutions? In selecting models based on forecast accuracy
+
+ Of course no statistic favors Model 1 over Model 2.
+Nevertheless, the optimal policy derived under Model 1 is virtually indistinguishable from the optimal policy for the true model, [Fig 4b].
+It is considerably less obvious that the adaptive management solution would also fail to ever select Model 1.
+After all, that approach integrates over the uncertainty in the models, and explicitly seeks to maximize utility (here assumed to be economic net present value), where Model 1 does so much better.
+Nevertheless, the "adaptive" part of the approach is actually its downfall.
+If the manager never learned, never adapted by updating the probabilities that each model was correct, the overall economic and ecological benefits would have been much higher.
+What if the learning step was not based on comparing model expectations to observed outcomes, but was instead conditioned only on utility derived by that action?
+Clearly such a strategy would take the largest harvest possible, crashing the stock and forfeiting all future harvest value.
+Such optimization techniques are known as greedy algorithms, and clearly fail in fisheries management.
+
+There are techniques for approaching model uncertainty which condition on utility alone which are not greedy algorithms.
+For example, in model-free deep reinforcement learning, a neural network is trained on cumulative utility to select the best action.
+
+If the problem as posed is not solved by existing approaches, then perhaps the paradox is only a result of
+
+- This result is driven by the unrealistic simplicity, and would not occur in more realistic examples.
+
+An example based on a constant-mortality policy would be easier to critique by observing that the policy was not in fact optimal in the first place.
+Recall that the purpose of this modeling exercise is not to precisely mimic the decision process of a specific quota in specific fish stock, but rather to illustrate in more general terms how carefully considered theory for approaching decision making under model uncertainty, from rigorous assessment of forecast skill to adaptive management, fail to handle this very simple example.
+
+No measure of forecast skill would prefer predictions drawn from Model 1 over those drawn from Model 2.
+Any process of updating model probabilities will lead away from Model 1 and towards Model 2.
+Simpler methods such as information criteria, $r^2$ or other goodness-of-fit metrics will also all clearly prefer model 2.
+The problem does not arise from a lack of data, or from over-fitting, which given sufficient data would be addressed by the use of @Gneiting2007's proper scoring rules.
+Nor does the problem resolved by approaches explicitly designed to maximize utility (e.g.
+economic value) over model uncertainty [@Polasky2011].
+
+- This result is simply an example of how different objectives lead to different strategies.
+
+<!-- -->
+# Discussion take 2
+
+Some readers will find the presentation of forecasting
+
+addressing potential critiques:
+
+Making this problem more complicated might better approximate real-world decision making but would
+by no means resolve the issue.  For example, increasing the suite of models under consideration to
+include all possible parameter values Gordon-Schaefer model, or better, adding
+in many other well studied structural forms such as Ricker [@], Beverton-Holt [@], or Shepard equation
+to our set of candidate models would give the same result.  Moreover, focusing on expanding the set of
+possible models ignores the simple fact that our most stripped down version of the problem already 
+contains a model which can obtain nearly optimal performance, just as it already contains a model which
+can make pretty accurate forecasts.  If well considered approaches make the wrong choice in this simple
+problem, the issue requires more attention.
+
+Note that this problem is not merely a case of alternative objectives leading to alternate outcomes. 
+Unlike model selection based on forecasting performance, the adaptive management approach explicitly
+considers the objective of maximizing the utility, and yet still fails to select the correct model.
 
@@
 
+acknowledgements: |
+  This work was supported in part by NSF CAREER (#1942280) and computational resources from NSF's XSEDE Jetstream (DEB160003) and Chameleon cloud platforms.
+
+# A way forward?
+
+note that the model set considered here already contains a model which captures the key feature.
+
+- Greedy strategy will not work.  Could stick with a model for 50 years then switch...
+- A mechanistic understanding of the decision process, not the ecological one.
+Analyzing the decision problem, we know that in this situation, the optimal stategy depends only on the location of the peak of the growth rate.
+This could suggest a very different way forward
+-
+
+Yet the model that leads to the best decisions is not always the model that makes the most accurate forecasts, as I illustrate here.
+Surprisingly, this can even happen when a decision is derived directly from a complex optimization routine of a probabilistic predictive models [@Marescot2013].
+Reality is complex; even our best models can only ever be approximations of underlying processes.
+Here, I use a classic, well-understood example from fisheries management [@Schaefer1954; @Clark1973; @Reed1979] to illustrate both the paradox of how a model with the worst forecast provides the best decision outcomes, as well as show how we can avoid selecting models that are poorly suited for management by considering the management context more explicitly.
+These results underscore that in choosing the best model for decision-making, it can be more important to capture a single key feature of the process than it is to make the most accurate prediction about future states.
+
+<!--
+Many ecological management problems are sequential decision problems, in which each year (or other interval) a manager must observe the state of the system and choose a course of action to maximize long term objectives.
+Such problems inherently depend on forecasts: each possible action can result in a different forecast for the future state, typically reflecting some uncertainty as well.
+The utility the manager derives may depend on both the choice of action and the state of the system, reflecting the costs and benefits associated with each.
+Sequential decision-making problems are distinguished by the need to think more than one move ahead.
+For instance, harvesting as many fish as possible in year one may maximize the market value that year, but if too few fish are left to reproduce then harvest in future years will suffer.
+The same calculus of thinking ahead frequently applies to rebuilding species populations as well [e.g.
+@Lambert;  @Chades2008].
+-->
+
 \pagebreak

### 2021-02-27 manuscript/manuscript.Rmd
:pen:

@@
 
-Given this simplified version of model uncertainty in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariable choose the other.
+Given this simplified version of model uncertainty in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariably choose the other.
 The reason for this may appear obvious once we compare both curves to that of the underlying model, Model 3.
 Plotting the growth rate functions of each model, [Fig 4a], it is hardly surprising that no method which would prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3.
-Seeing this example, it is natural to pose any of four objections: (1) this problem is an artifact of overly-simplified models, (2) this problem is an artifact of the optimal control solution (3) the problem arises from different objectives, (4) the problem is already solved.
+Nevertheless, decisions based on model 1 are nearly indistinguishable from those based on the true model [Fig4b], while Model 2 leads to over-harvesting.
+So why do our best approaches fail such a simple test case?
+It is tempting to dismiss this example as having little relevance to real-world decision problems on a variety of grounds: (1) this problem is an artifact of overly-simplified models, (2) this problem is an artifact of an unrealistic optimal control solution, (3) the problem arises from having started with different objectives, (4) the problem is already solved.
 I address these in turn.
@@
 <!-- move to appendix?  too confusing/too long?-->
-Is this result an artifact of the optimal control?
+Is this result an artifact of optimal control?
 The analysis has relied on optimal control theory to determine the optimal decision associated with each model or distribution of models.
@@
 Moreover, using a non-optimal policy would always beg the question of whether the poor outcomes were merely the result of using a non-optimal policy rather than a consequence of model uncertainty.
+Because the space of possible actions is smaller than that of possible models, we should expect most decision problems to have their own "Model 1," a model that would lead to correct actions while being a poor predictor of future outcomes.
+The trouble is that we have no theory which would not reject such a model.
 
-The optimal policy does not depend equally on all aspects of the underlying model.
-In this simple case, the policy depends only on the biomass which provides the maximum growth rate, and it is precisely this feature which allows Model 1 to match the policy of the true model despite over-estimating both the per-capita growth rate and unharvested equilbrium stock size so dramatically.
-The ability to capture certain key features rather than replicate every detail is the objective of any modeling effort, and the reason why scientific decision making is possible in the first place.
-Many optimal control solutions are similarly insensitive to getting all the details right.
-If we seek to improve decisions in face of model uncertainty, we must be able to exploit this aspect to find models fit to purpose, and be cautious about employing methods that would reject such models out of hand.
+Is this result merely an artifact of differing objectives?
+@Gneiting2007's proper scoring rules are designed to select models which make the best forecast, and Model 2 really is the best choice for that task.
+However, the adaptive management approach explicitly seeks to maximize expected utility, integrating over the uncertainty in the model, and yet it evolves away from Model 1 which would achieve that goal and settles on Model 2.
+Note that this issue is not resolved by updating the learning step to reflect only the immediate reward -- it is in fact easy to see that so-called greedy algorithms would take the largest harvest possible and never reach the sustainable yield achieved by lower harvest rates of Model 1.
 
-Does this problem merely reflect the fact that different objective functions lead to different solutions? In selecting models based on forecast accuracy
-
- Of course no statistic favors Model 1 over Model 2.
-Nevertheless, the optimal policy derived under Model 1 is virtually indistinguishable from the optimal policy for the true model, [Fig 4b].
-It is considerably less obvious that the adaptive management solution would also fail to ever select Model 1.
-After all, that approach integrates over the uncertainty in the models, and explicitly seeks to maximize utility (here assumed to be economic net present value), where Model 1 does so much better.
-Nevertheless, the "adaptive" part of the approach is actually its downfall.
-If the manager never learned, never adapted by updating the probabilities that each model was correct, the overall economic and ecological benefits would have been much higher.
-What if the learning step was not based on comparing model expectations to observed outcomes, but was instead conditioned only on utility derived by that action?
-Clearly such a strategy would take the largest harvest possible, crashing the stock and forfeiting all future harvest value.
-Such optimization techniques are known as greedy algorithms, and clearly fail in fisheries management.
-
-There are techniques for approaching model uncertainty which condition on utility alone which are not greedy algorithms.
-For example, in model-free deep reinforcement learning, a neural network is trained on cumulative utility to select the best action.
-
-If the problem as posed is not solved by existing approaches, then perhaps the paradox is only a result of
-
-- This result is driven by the unrealistic simplicity, and would not occur in more realistic examples.
-
-An example based on a constant-mortality policy would be easier to critique by observing that the policy was not in fact optimal in the first place.
-Recall that the purpose of this modeling exercise is not to precisely mimic the decision process of a specific quota in specific fish stock, but rather to illustrate in more general terms how carefully considered theory for approaching decision making under model uncertainty, from rigorous assessment of forecast skill to adaptive management, fail to handle this very simple example.
-
-No measure of forecast skill would prefer predictions drawn from Model 1 over those drawn from Model 2.
-Any process of updating model probabilities will lead away from Model 1 and towards Model 2.
-Simpler methods such as information criteria, $r^2$ or other goodness-of-fit metrics will also all clearly prefer model 2.
-The problem does not arise from a lack of data, or from over-fitting, which given sufficient data would be addressed by the use of @Gneiting2007's proper scoring rules.
-Nor does the problem resolved by approaches explicitly designed to maximize utility (e.g.
-economic value) over model uncertainty [@Polasky2011].
-
-- This result is simply an example of how different objectives lead to different strategies.
-
-<!-- -->
-# Discussion take 2
-
-Some readers will find the presentation of forecasting
-
-addressing potential critiques:
-
-Making this problem more complicated might better approximate real-world decision making but would
-by no means resolve the issue.  For example, increasing the suite of models under consideration to
-include all possible parameter values Gordon-Schaefer model, or better, adding
-in many other well studied structural forms such as Ricker [@], Beverton-Holt [@], or Shepard equation
-to our set of candidate models would give the same result.  Moreover, focusing on expanding the set of
-possible models ignores the simple fact that our most stripped down version of the problem already 
-contains a model which can obtain nearly optimal performance, just as it already contains a model which
-can make pretty accurate forecasts.  If well considered approaches make the wrong choice in this simple
-problem, the issue requires more attention.
-
-Note that this problem is not merely a case of alternative objectives leading to alternate outcomes. 
-Unlike model selection based on forecasting performance, the adaptive management approach explicitly
-considers the objective of maximizing the utility, and yet still fails to select the correct model.
-
-## Discussion
-
-This paradox in performance of forecasting vs performance in decision making can be easily resolved by considering the context of the decision problem more closely.
-Comparing plots of the functional form of our two logistic-curve models, compared to the functional form of the "true" model used to drive the simulations (Fig.
-3A), it is clear to see that model 2 does indeed lie closer to the true model throughout the state space, agreeing precisely with the true carrying capacity (where both functions cross zero with negative slope).
-However, the peak of model 3 very nearly matches the peak of model 1.
-The optimal decision literature, dating back to the 1950s [@Schaefer1954], demonstrates that the Maximum Sustainable Yield (MSY) is maintained by harvesting a stock down to the size at which it achieves its maximum growth rate, i.e.
-50% of the un-fished equilibrium size for a symmetric growth model ($K/2$).
-Model 1, while being very wrong about both the growth rate and the un-fished equilibrium, is nevertheless nearly perfect in estimating the stock size at which maximum growth rate is achieved, and this gives nearly optimal decisions (Fig 3B) despite its terrible forecasts.
-
-Thus, each year our model 1 managers are again chagrined to see the stock size estimates come in far below their rosy predictions, but nevertheless manage to set a nearly optimal quota by comparing the observed stock size to the model's predicted optimal escapement level [@Reed1979].
-Meanwhile, model 2 managers could only congratulate themselves that each year's observations fall neatly within their predicted interval, unaware that the they were over-exploiting the fishery by both economic and ecological metrics.
-If we had access to model 3, we would no doubt find that it outperformed model 2 in forecast accuracy as well as ecological and economic performance.
-But in real ecological decision making, we never know the true model -- we will always be comparing among approximations.
-Within fisheries, even in today's parameter-rich age-structured models, recruitment approximations with symmetric growth functions (Logistic, Ricker, Beverton-Holt, etc) still dominate [@ramlegacy2012; @ramlegacy2018].
-
-This issue is by no means unique to fisheries.
-Throughout resource management and conservation, and no doubt other fields, decisions about which model to use are guided by which model best fits available data [@Clark1990].
-Increasingly, these are joined by calls to assess _forecast accuracy_ [@Clark2001; @Dietze2018; @White2019] as the ultimate test of a model.
-Yet as this example illustrates, such metrics, no matter how rigorously defined, may select entirely the wrong model for the task at hand.
-A decision maker has other objectives than prediction accuracy, and approaches which ignore these considerations do so at their peril.
-This example has also shown that once we are managing with the wrong model, no amount of comparing predictions from that model to actual outcomes will guarantee we discover our mistake.
-Despite its consistently good predictions, model 1 is in fact over-fishing to a dangerous level.
-
-Because we will never know the "true" model, we must never forget that our choice of models must reflect the context for which those models will be used [@Levins1966].
-Model 2 would indeed be a better choice than model 1 if our objective was to determine the natural size of our fish stock in the absence of fishing.
-Only when we focus on the outcomes we actually care about -- in this case, economic and ecological performance -- can we see which model is best for decision-making.
-Model 1, despite its many mistakes, is right about one key feature: the biomass for peak growth -- and that is enough to guarantee nearly optimal performance.
-This conclusion should also be reassuring to both modelers and decision makers, for it reminds us that effective models need be perfect or even all that close in every aspect, as long as they capture the key features of the decision context.
-Decision theory [e.g.
-@Clark1973; @Reed1979; @pomdp-intro] and research into the socio-ecological models [e.g.
-@Kareiva2006] helps us better understand that context.
-Adaptive management approaches [@Walters1978] can apply that theory to compare management outcomes between models directly.
-It is not true that we need good forecasts to make good decisions.
-
-```{r figure3, fig.width=7, fig.height=4, fig.cap="Panel A: Population recruitment curves for each model, compared to that of the true Model.
-3 Model 2 more closely approximates the true Model 3, but note that maximum value Model 1 and Model 3 occur at nearly the same value for the state, x.
-Panel B: The computed optimal policy of each model, derived by SDP, expressed in terms of the target escapement (population size remaining after harvest) for each possible stock size.
-Model 1 over-harvests consistently, while the target escapement under Model 2 is nearly identical to that of the true Model 3."}
-fig3a + ggtitle("A") + fig3b + ggtitle("B")
-```
-
-# Methods
-
-Stochastic transition matrices are defined for models 1-3 on a discrete grid of
-240 possible states spaced uniformly from 0 to 24. A discrete action space 
-enumerating possible harvest quotas is set to the same grid.
-The utility of a harvest quota $H_t$ given a population state $X_t$ is given by $U(X_t, H_t) = \min(X_t, H_t)$ (i.e.
-a fixed price for realized harvest).
-A modest discount of
-$\gamma = 0.99$ allows comparisons to approaches that ignore [@Schaefer1954] or include [@Clark1973; @Reed1979] discounting; results are not sensitive to this choice.
-The optimal policy for each model is determined by stochastic dynamic programming [@Marescot2013].
-Details of the implementation, including fully reproducible R code, have been included in the appendix.
-
-acknowledgements: |
-  This work was supported in part by NSF CAREER (#1942280) and computational resources from NSF's XSEDE Jetstream (DEB160003) and Chameleon cloud platforms.
-
-# A way forward?
-
-note that the model set considered here already contains a model which captures the key feature.
-
-- Greedy strategy will not work.  Could stick with a model for 50 years then switch...
-- A mechanistic understanding of the decision process, not the ecological one.
-Analyzing the decision problem, we know that in this situation, the optimal stategy depends only on the location of the peak of the growth rate.
-This could suggest a very different way forward
--
-
-Yet the model that leads to the best decisions is not always the model that makes the most accurate forecasts, as I illustrate here.
-Surprisingly, this can even happen when a decision is derived directly from a complex optimization routine of a probabilistic predictive models [@Marescot2013].
-Reality is complex; even our best models can only ever be approximations of underlying processes.
-Here, I use a classic, well-understood example from fisheries management [@Schaefer1954; @Clark1973; @Reed1979] to illustrate both the paradox of how a model with the worst forecast provides the best decision outcomes, as well as show how we can avoid selecting models that are poorly suited for management by considering the management context more explicitly.
-These results underscore that in choosing the best model for decision-making, it can be more important to capture a single key feature of the process than it is to make the most accurate prediction about future states.
-
-<!--
-Many ecological management problems are sequential decision problems, in which each year (or other interval) a manager must observe the state of the system and choose a course of action to maximize long term objectives.
-Such problems inherently depend on forecasts: each possible action can result in a different forecast for the future state, typically reflecting some uncertainty as well.
-The utility the manager derives may depend on both the choice of action and the state of the system, reflecting the costs and benefits associated with each.
-Sequential decision-making problems are distinguished by the need to think more than one move ahead.
-For instance, harvesting as many fish as possible in year one may maximize the market value that year, but if too few fish are left to reproduce then harvest in future years will suffer.
-The same calculus of thinking ahead frequently applies to rebuilding species populations as well [e.g.
-@Lambert;  @Chades2008].
--->
+Perhaps Model 1 is a unicorn, a mythically simple model that gives optimal results but has no analogue in the real world.
+This is not an easy assertion to prove either way.
+Future observations consistently fall outside the 95% confidence intervals predicted under Model 1.
+Unfortunately, such innaccurate predictions are only too common in the real world.
+Have any of those led to effective policy despite the innnaccuracy?
+Have any been replaced with more accurate models that have also driven worse decisions?
+That is more difficult to answer.
+What this example demonstrates is that if such unicorn models exist, our current paradigm will surely filter them out eventually, to be replaced with worse outcomes.
+<!-- Omit this?  Does it beg an example? -->
+There is at least one technique which might address the problem presented here.
+Recently developed techniques based on model-free deep reinforcement learning train multi-layer neural networks to approximate the optimal control policy without assuming or attempting to model the underlying dynamic process at all.
+Feedback is based on cumulative reward, not short-term reward or predictive accuracy.
 

### 2021-02-27 manuscript/scratch.Rmd
:pen:

```{r prob_scores, message = FALSE}
## Probability scores instead of proper scores
scores <- sims %>%
  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
  filter(model < 3) %>%
  rowwise() %>%
  mutate(prob =  transitions[[model]][state_index, next_state, action_index] ) %>% 
  group_by(reps, model) %>%
  summarise(score = sum(log(prob)))

score_table <- scores %>% group_by(model) %>% summarise(mean_score = mean(score))

scores %>% mutate(model = as.factor(model)) %>% 
  ggplot(aes(x = score, group = model, fill = model)) +
  geom_histogram(binwidth = 70)
```

There are techniques for approaching model uncertainty which condition on utility alone which are not greedy algorithms. For example, in model-free deep reinforcement learning, a neural network is trained on cumulative utility to select the best action.

If the problem as posed is not solved by existing approaches, then perhaps the paradox is only a result of

- This result is driven by the unrealistic simplicity, and would not occur in more realistic examples.

An example based on a constant-mortality policy would be easier to critique by observing that the policy was not in fact optimal in the first place.  Recall that the purpose of this modeling exercise is not to precisely mimic the decision process of a specific quota in specific fish stock, but rather to illustrate in more general terms how carefully considered theory for approaching decision making under model uncertainty, from rigorous assessment of forecast skill to adaptive management, fail to handle this very simple example.

No measure of forecast skill would prefer predictions drawn from Model 1 over those drawn from Model 2. Any process of updating model probabilities will lead away from Model 1 and towards Model 2.  Simpler methods such as information criteria, $r^2$ or other goodness-of-fit metrics will also all clearly prefer model 2.  The problem does not arise from a lack of data, or from over-fitting, which given sufficient data would be addressed by the use of @Gneiting2007's proper scoring rules.  Nor does the problem resolved by approaches explicitly designed to maximize utility (e.g. economic value) over model uncertainty [@Polasky2011].

- This result is simply an example of how different objectives lead to different strategies.

<!-- -->
# Discussion take 2

Some readers will find the presentation of forecasting

addressing potential critiques:

Making this problem more complicated might better approximate real-world decision making but would
by no means resolve the issue.  For example, increasing the suite of models under consideration to
include all possible parameter values Gordon-Schaefer model, or better, adding
in many other well studied structural forms such as Ricker [@], Beverton-Holt [@], or Shepard equation
to our set of candidate models would give the same result.  Moreover, focusing on expanding the set of
possible models ignores the simple fact that our most stripped down version of the problem already 
contains a model which can obtain nearly optimal performance, just as it already contains a model which
can make pretty accurate forecasts.  If well considered approaches make the wrong choice in this simple
problem, the issue requires more attention.

Note that this problem is not merely a case of alternative objectives leading to alternate outcomes. 
Unlike model selection based on forecasting performance, the adaptive management approach explicitly
considers the objective of maximizing the utility, and yet still fails to select the correct model.

## Discussion

This paradox in performance of forecasting vs performance in decision making can be easily resolved by considering the context of the decision problem more closely. Comparing plots of the functional form of our two logistic-curve models, compared to the functional form of the "true" model used to drive the simulations (Fig. 3A), it is clear to see that model 2 does indeed lie closer to the true model throughout the state space, agreeing precisely with the true carrying capacity (where both functions cross zero with negative slope).  However, the peak of model 3 very nearly matches the peak of model 1. The optimal decision literature, dating back to the 1950s [@Schaefer1954], demonstrates that the Maximum Sustainable Yield (MSY) is maintained by harvesting a stock down to the size at which it achieves its maximum growth rate, i.e. 50% of the un-fished equilibrium size for a symmetric growth model ($K/2$).  Model 1, while being very wrong about both the growth rate and the un-fished equilibrium, is nevertheless nearly perfect in estimating the stock size at which maximum growth rate is achieved, and this gives nearly optimal decisions (Fig 3B) despite its terrible forecasts.

Thus, each year our model 1 managers are again chagrined to see the stock size estimates come in far below their rosy predictions, but nevertheless manage to set a nearly optimal quota by comparing the observed stock size to the model's predicted optimal escapement level [@Reed1979].  Meanwhile, model 2 managers could only congratulate themselves that each year's observations fall neatly within their predicted interval, unaware that the they were over-exploiting the fishery by both economic and ecological metrics.  If we had access to model 3, we would no doubt find that it outperformed model 2 in forecast accuracy as well as ecological and economic performance.  But in real ecological decision making, we never know the true model -- we will always be comparing among approximations.  Within fisheries, even in today's parameter-rich age-structured models, recruitment approximations with symmetric growth functions (Logistic, Ricker, Beverton-Holt, etc) still dominate [@ramlegacy2012; @ramlegacy2018].

This issue is by no means unique to fisheries.  Throughout resource management and conservation, and no doubt other fields, decisions about which model to use are guided by which model best fits available data [@Clark1990]. Increasingly, these are joined by calls to assess _forecast accuracy_ [@Clark2001; @Dietze2018; @White2019] as the ultimate test of a model.  Yet as this example illustrates, such metrics, no matter how rigorously defined, may select entirely the wrong model for the task at hand.  A decision maker has other objectives than prediction accuracy, and approaches which ignore these considerations do so at their peril. This example has also shown that once we are managing with the wrong model, no amount of comparing predictions from that model to actual outcomes will guarantee we discover our mistake.  Despite its consistently good predictions, model 1 is in fact over-fishing to a dangerous level.

Because we will never know the "true" model, we must never forget that our choice of models must reflect the context for which those models will be used [@Levins1966].  Model 2 would indeed be a better choice than model 1 if our objective was to determine the natural size of our fish stock in the absence of fishing.  Only when we focus on the outcomes we actually care about -- in this case, economic and ecological performance -- can we see which model is best for decision-making.  Model 1, despite its many mistakes, is right about one key feature: the biomass for peak growth -- and that is enough to guarantee nearly optimal performance.   This conclusion should also be reassuring to both modelers and decision makers, for it reminds us that effective models need be perfect or even all that close in every aspect, as long as they capture the key features of the decision context.  Decision theory [e.g. @Clark1973; @Reed1979; @pomdp-intro] and research into the socio-ecological models [e.g. @Kareiva2006] helps us better understand that context.  Adaptive management approaches [@Walters1978] can apply that theory to compare management outcomes between models directly.  It is not true that we need good forecasts to make good decisions.

```{r figure3, fig.width=7, fig.height=4, fig.cap="Panel A: Population recruitment curves for each model, compared to that of the true Model. 3 Model 2 more closely approximates the true Model 3, but note that maximum value Model 1 and Model 3 occur at nearly the same value for the state, x.  Panel B: The computed optimal policy of each model, derived by SDP, expressed in terms of the target escapement (population size remaining after harvest) for each possible stock size. Model 1 over-harvests consistently, while the target escapement under Model 2 is nearly identical to that of the true Model 3."}
fig3a + ggtitle("A") + fig3b + ggtitle("B")
```

# Methods

Stochastic transition matrices are defined for models 1-3 on a discrete grid of
240 possible states spaced uniformly from 0 to 24. A discrete action space 
enumerating possible harvest quotas is set to the same grid. The utility of a harvest quota $H_t$ given a population state $X_t$ is given by $U(X_t, H_t) = \min(X_t, H_t)$ (i.e. a fixed price for realized harvest). A modest discount of
$\gamma = 0.99$ allows comparisons to approaches that ignore [@Schaefer1954] or include [@Clark1973; @Reed1979] discounting; results are not sensitive to this choice. The optimal policy for each model is determined by stochastic dynamic programming [@Marescot2013].  Details of the implementation, including fully reproducible R code, have been included in the appendix.

acknowledgements: |
  This work was supported in part by NSF CAREER (#1942280) and computational resources from NSF's XSEDE Jetstream (DEB160003) and Chameleon cloud platforms.

# A way forward?

note that the model set considered here already contains a model which captures the key feature.

- Greedy strategy will not work.  Could stick with a model for 50 years then switch...
- A mechanistic understanding of the decision process, not the ecological one.  Analyzing the decision problem, we know that in this situation, the optimal stategy depends only on the location of the peak of the growth rate.  This could suggest a very different way forward
-

Yet the model that leads to the best decisions is not always the model that makes the most accurate forecasts, as I illustrate here. Surprisingly, this can even happen when a decision is derived directly from a complex optimization routine of a probabilistic predictive models [@Marescot2013].  Reality is complex; even our best models can only ever be approximations of underlying processes.  Here, I use a classic, well-understood example from fisheries management [@Schaefer1954; @Clark1973; @Reed1979] to illustrate both the paradox of how a model with the worst forecast provides the best decision outcomes, as well as show how we can avoid selecting models that are poorly suited for management by considering the management context more explicitly. These results underscore that in choosing the best model for decision-making, it can be more important to capture a single key feature of the process than it is to make the most accurate prediction about future states.

<!--
Many ecological management problems are sequential decision problems, in which each year (or other interval) a manager must observe the state of the system and choose a course of action to maximize long term objectives. Such problems inherently depend on forecasts: each possible action can result in a different forecast for the future state, typically reflecting some uncertainty as well.  The utility the manager derives may depend on both the choice of action and the state of the system, reflecting the costs and benefits associated with each. Sequential decision-making problems are distinguished by the need to think more than one move ahead. For instance, harvesting as many fish as possible in year one may maximize the market value that year, but if too few fish are left to reproduce then harvest in future years will suffer.  The same calculus of thinking ahead frequently applies to rebuilding species populations as well [e.g. @Lambert;  @Chades2008].  
-->

### 2021-02-27 manuscript/manuscript.Rmd
:pen:

@@
 Note that this issue is not resolved by updating the learning step to reflect only the immediate reward -- it is in fact easy to see that so-called greedy algorithms would take the largest harvest possible and never reach the sustainable yield achieved by lower harvest rates of Model 1.
+To select Model 1 over alternatives like Model 2 which make better predictions but lead to worse outcomes requires a more long-term approach that forgoes iterative updating of probabilities as new data becomes available.
+This is the opposite recommendation of _adaptive_ management  @Dietze2018
 
@@
 What this example demonstrates is that if such unicorn models exist, our current paradigm will surely filter them out eventually, to be replaced with worse outcomes.
+
 <!-- Omit this?  Does it beg an example? -->
+<!--
 There is at least one technique which might address the problem presented here.
@@
 Feedback is based on cumulative reward, not short-term reward or predictive accuracy.
+-->
 

### 2021-03-01 manuscript/appendix.Rmd
data cache

--- 2021-02-05 manuscript/appendix.Rmd
@@
 ---
+
+This appendix provides a fully reproducible walk-through of the implementation of the analyses considered in this paper.
+None of the methods applied here: Stochastic Dynamic Programming, Iterative Forecasts assessed with Proper Scoring Rules, or Adaptive Management, are novel
+
+This appendix includes computer code in the R language, which can also be found in the corresponding R-Markdown document in the paper's GitHub repository, <https://github.com/cboettig/bad-forecast-good-decision>.
+While computer code can be more verbose and difficult to read than mathematical formulae presented here, it is also less ambiguous.
 
@@
 
-The optimization problem can then be solved by dynamic programming, using the recursive Bellman equation as ,
+The optimization problem can then be solved by dynamic programming, using the recursive Bellman equation as,
 
@@
 Whether we use SDP or some other way to determine the optimal policy, given the model, is immaterial to the conclusion.
-The issue at hand is how we determine the model in the first place.
-
-Given the function $f$ with known parameters, it is straight forward to determine the optimal harvest policy by stochastic dynamic programming (SDP, see @Mangel1985; @Marescot2013).
 
@@
 
-On a discrete grid of possible states and actions, we can define the growth rate of a given state $X_t$ subject to harvest $H_t$,
-$f(X_t,H_t)$ as set of matrices.  Each matrix $i$ gives the transition probabilities for any current state to any future state, 
+On a discrete grid of possible states and actions, we can define the growth rate of a given state $X_t$ subject to harvest $H_t$, $f(X_t,H_t)$ as set of matrices.
+Each matrix $i$ gives the transition probabilities for any current state to any future state,
 given that action $i$ is taken.
@@
   .id = "model")
-```
 
-# Adaptive Management
-
-Passive adaptive management using a Bayesian learning scheme still learns the wrong model.
-
-```{r}
-adaptive_management <- memoise::memoise(mdp_learning, cache = memoise::cache_filesystem("cache/"))
-x0 <- which.min(abs(states - 6))
-
-am1 <- adaptive_management(transitions[1:2], reward, discount, model_prior = c(0.99, 0.01), x0 = x0, 
-               Tmax = 50, true_transition = transitions[[3]], 
-                epsilon = 0.001, max_iter = 2000)
-```
-
-```{r}
-am <- am1$df %>% mutate(belief = am1$posterior$V2)
-am %>%  ggplot(aes(time, states[state], col = belief)) + 
-  geom_line() + 
-  geom_point() +
-  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4])
-```
-
-```{r include = FALSE}
-as_tibble(am1$posterior) %>% 
-  mutate(time = seq_along(V1)) %>%
-  select(time = time, "1" = V1, "2" = V2) %>% 
-  pivot_longer(c("1", "2"), values_to="belief", names_to = "model") %>%
-  filter(time < 20) %>%
-  ggplot(aes(time, belief, col=model)) + geom_line(col = pal[4])
+write_csv(policies, "../data/policies.csv")
 ```
@@
          )
+
+write_csv(sims, "../data/sims.csv")
 ```
@@
   mutate(model = as.character(model))
+
+write_csv(predictions, "../data/predictions.csv")
 ```
 
-# Plotting
+# Adaptive Management
 
-Having computed a single data frame of simulation data under both optimally fished (according to each model) and unfished scenarios for each model, along with the proper scores for the step-ahead forecasts a la @Gneiting2007, we have all the results necessary to generate the figures presented in the main paper.
+Passive adaptive management using a Bayesian learning scheme still learns the wrong model.
 
-```{r fig1a}
-fig1ab <- predictions %>%
-  filter(reps == "2", time < 10) %>%
-  ggplot(aes(time, col = model, fill = model)) + 
-  geom_point(aes(y = expected)) + 
-  geom_errorbar(aes(ymin = low, ymax = high)) +
-  geom_point(aes(y = true), pch = "*", size = 12, 
-             alpha = 0.6) + 
-  facet_wrap(~scenario, ncol = 1, labeller =as_labeller(c(
-    A_unfished = "A. Without fishing",
-    B_fished = "B. Managed harvest"))) +
-  ylab("stock size")
+```{r}
+adaptive_management <- memoise::memoise(mdp_learning, cache = memoise::cache_filesystem("cache/"))
+x0 <- which.min(abs(states - 6))
+
+am1 <- adaptive_management(transitions[1:2], reward, discount, 
+                           model_prior = c(0.99, 0.01), x0 = x0, 
+                           Tmax = 50, true_transition = transitions[[3]], 
+                            epsilon = 0.001, max_iter = 2000)
+```
+
+```{r}
+am <- am1$df %>% mutate(belief = am1$posterior$V2)
+write_csv(am, "../data/am.csv")
 
@@
 
-Likewise we can plot the data on proper scores associated with each prediction of each model:
-
-```{r}
-fig1cd <- predictions %>%
-  ggplot(aes(x = score, group = model, fill = model)) +
-  geom_histogram(binwidth = 2, show.legend = FALSE) +
-  coord_cartesian(xlim = c(-100, 1), ylim = c(0,4000)) +
-  xlab("Proper score") +
-  facet_wrap(~scenario, ncol = 1, labeller =as_labeller(c(
-    A_unfished = "C.",
-    B_fished = "D."))) + 
-  scale_x_continuous(breaks = c(-100, 0)) +
-  theme(axis.text.y = element_blank(),
-        plot.margin = margin(0, 0, 0, 0, "cm"),
-        panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank()
-        )
-
-```
-
-## Forecast ecological and economic performance
-
-The ecological performance of the model can easily be visualized by plotting
-the results of each simulation.
-
-```{r plot_sims}
-fig2a <- 
-  sims %>% 
-  group_by(model, time) %>%
-  summarise(mean_state = mean(state), sd = sd(state), .groups = "drop") %>%
-  filter(time < 25, model != "3") %>%
-  ggplot(aes(time, mean_state)) + 
-  geom_line(aes(col = model), lwd=1.5, show.legend = FALSE) +
-    geom_ribbon(aes(ymin = mean_state - 2*sd, 
-                    ymax = mean_state + 2*sd,
-                    fill = model),
-                alpha = 0.2, show.legend = FALSE) +
-  ylab("state") + 
-  ggtitle("A") + labs(subtitle = "Ecological")
-```
+### Additional calculations for plots
 
@@
 
-optimal <- select(filter(npv_df, model == "3"), time, mean_npv)
+write_csv(npv_df, "../data/npv_df.csv")
+```
 
-fig2b <- 
-npv_df %>%
-  filter(model != "3", time %in% seq(1,100, by = 5)) %>%
-  ggplot(aes(time, mean_npv)) +
-  geom_line(data = optimal, lwd = 1.5, col = "grey20") +
-  geom_point(aes(col=model), size = 4, alpha = 0.8) + 
-  ylab("Net present value") + xlab("time")  + 
-  ggtitle("B.") + labs(subtitle = "Economic")
+```{r}
+# tabular comparisons
+npv <- npv_df %>% group_by(model) %>% summarize(npv = max(mean_npv))
+
+am_npv <- sum(am$value * discount ^ am$time)
+am_economics_percent <- round(am_npv / npv[[1,"npv"]] * 100)
+
+mean_state <- sims %>% group_by(model) %>% summarize(state = mean(state))
+am_ecology_percent <- round(mean(states[am$state])/mean_state[[1, "state"]]*100)
+```
+
+Under adaptive management, the manager realizes only `r am_economics_percent`%
+of the economic value that would be achieved under Model 1 alone, and only 
+`r am_ecology_percent`% of the spawning stock biomass that would have been 
+achieved under Model 1 alone.
+
+To plot growth curves of individual models (Fig 4a), we evaluate
+
+$$\Delta x = x_{t+1} - x_t = f(x_t) - x_t$$
+
+for each model for all possible states $x_t$.
+
+```{r plot_models}
+model_curves <- 
+  map_dfc(models, function(f) f(states) - states) %>%
+  mutate(state = states) %>%  pivot_longer(names(models), "model")
+
+write_csv(model_curves, "../data/model_curves.csv")
 
@@
 
-Plotting the model functions themselves are shown in Fig 3a:
-
-```{r plot_models}
-d <- 
-  map_dfc(models, function(f) f(states) - states) %>%
-  mutate(state = states)
-
-fig3a <- 
-  d %>% pivot_longer(names(models), "model") %>%
-  ggplot(aes(state, value, col=model, lty=model)) +
-  geom_hline(aes(yintercept = 0), lwd=1) + 
-  geom_line(lwd=2, show.legend = FALSE) + 
-  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
-  ylab(bquote(f(x) - x)) + xlab("x")
-```
-
-We can easily plot the optimal policy derived from each model, as shown
-in Fig3b.
-
-```{r plot_policies}
-fig3b <- policies %>%
-  ggplot(aes(states, escapement, col=model, lty=model)) + 
-  geom_line(lwd=2) + xlab("state")
-```
-
 # References

### 2021-03-01 manuscript/manuscript.Rmd
data cache

@@
 ---
-title: 'Learning to be wrong:'
+title: 'Are we learning to build better models at the cost of better decisions?'
 author:
@@
 
-# For footer text
-lead_author_surname: Boettiger
-
-significance: |
-
 ## change to true to add optional line numbering
@@
 
-output: rticles::pnas_article
+output: rticles::elsevier_article
+layout: 3p
 ---
@@
 rmarkdown::render("appendix.Rmd")
-knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, warning = FALSE, dev="cairo_pdf")
+knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
+                      warning = FALSE, dev="cairo_pdf")
 ```
@@
 contain scores of parameters to reflect the specific age or stage structure of a
-specific fish stock [@ram], I will rely on simple, well-studied models which
+specific fish stock [@ramlegacy2018; @ram], I will rely on simple, well-studied models which
 permit greater intuition and generalization [@Levins1966]. Consistent with 
@@
 Again for simplicity, we will restrict ourselves to two simple candidate models $f_1$ and $f_2$.
-Both share the same underlying structure of logistic recruitment (known as the Gordon-Schaefer model in fisheries context owing to groundbreaking work independently by @Schafer1954 and @Gordon1954), differing only in their choice of certain parameters:
+Both share the same underlying structure of logistic recruitment (known as the Gordon-Schaefer model in fisheries context owing to groundbreaking work independently by @Schaefer1954 and @Gordon1954), differing only in their choice of certain parameters:
 
@@
 There are a wide range of techniques within each, and it is also possible to blend approaches.
-The key distinction is that methods in the first group do not involve any direct consideration of the possible actions or the utility that may result from those actions in how they select models [statistical approaches such as information criteria @BurnhamAnderson; and in particular, forecasting evaluation @Clark2001; @Dietze2018], while those in the second group require a more explicit statement of possible actions and the desired objectives [Decision theoretic approaches, for which @Polasky2011 provides an excellent and accessible review].
+The key distinction is that methods in the first group do not involve any direct consideration of the possible actions or the utility that may result from those actions in how they select models [statistical approaches such as information criteria e.g.
+@Burnham1998]; and in particular, forecasting evaluation @Clark2001; @Dietze2018], while those in the second group require a more explicit statement of possible actions and the desired objectives [Decision theoretic approaches, for which @Polasky2011 provides an excellent and accessible review].
 I illustrate how the most promising techniques from each of these groups would be applied to this simple problem, and demonstrate that it both cases they lead us away from the model that produces the most desirable decisions towards worse outcomes.
@@
 In Scenario A (passive observation) the fish stock is unharvested and allowed to recover towards carrying capacity (as simulated under our "true" model, Model 3) while comparing the observed stock size in each subsequent time step to the distribution predicted under Model 1 and Model 2 respectively [Fig].
-The mean, $\mu$ and variance, $\sigma$ of the forecast are compared against the true observation $x$ using a Proper scoring rule of @Geiting2007,
+The mean, $\mu$ and variance, $\sigma$ of the forecast are compared against the true observation $x$ using a Proper scoring rule of @Gneiting2007,
 
-\bein{equation}
+\begin{equation}
 -(\mu - x )^2 / \sigma^2  - \log(\sigma) \label{proper}
@@
 
-for each prediction over 100 replicate simulations of 100 time steps each [Fig].
+for each prediction over 100 replicate simulations of 100 time steps each [Fig 1].
 
@@
 Higher scores are better, confirming that model 2 makes the better forecasts."}
+predictions <- read_csv("../data/predictions.csv") %>% mutate(model = as.character(model))
+
+fig1cd <- predictions %>%
+  ggplot(aes(x = score, group = model, fill = model)) +
+  geom_histogram(binwidth = 2, show.legend = FALSE) +
+  coord_cartesian(xlim = c(-100, 1), ylim = c(0,4000)) +
+  xlab("Proper score") +
+  facet_wrap(~scenario, ncol = 1, labeller =as_labeller(c(
+    A_unfished = "C.",
+    B_fished = "D."))) + 
+  scale_x_continuous(breaks = c(-100, 0)) +
+  theme(axis.text.y = element_blank(),
+        plot.margin = margin(0, 0, 0, 0, "cm"),
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank()
+        )
+
+ fig1ab <- predictions %>% 
+  filter(reps == "2", time < 10) %>%
+  ggplot(aes(time, col = model, fill = model)) + 
+  geom_point(aes(y = expected)) + 
+  geom_errorbar(aes(ymin = low, ymax = high)) +
+  geom_point(aes(y = true), pch = "*", size = 12, 
+             alpha = 0.6) + 
+  facet_wrap(~scenario, ncol = 1, labeller =as_labeller(c(
+    A_unfished = "A. Without fishing",
+    B_fished = "B. Managed harvest"))) +
+  ylab("stock size")
+
  fig1ab + fig1cd + 
@@
 
-In Scenario B (actively harvest), I have first solved for the optimal management strategy for both Model 1 and Model 2 [Fig], which for small noise and concave functions with linear reward structure can be done analytically [see proof in @Reed1979], or can be solved more generally by stochastic dynamic programming [see review by @Marescot2013, details in Appendix].
+In Scenario B (actively harvest), I have first solved for the optimal management strategy for both Model 1 and Model 2 [Fig 1b].
+For small noise and concave functions with linear reward structure this can be done analytically [see proof in @Reed1979], or solved more generally by stochastic dynamic programming [see review by @Marescot2013, details in Appendix].
 Under this scenario, replicate simulations of the stock are harvested at each time step using the optimal quota dictated by either Model 1 and Model 2.
@@
 While it is possible to include such contributions directly in the utility function being optimized [e.g.
-@Halpern2013], even without doing so, Model 1 maintains both a higher unharvested biomass and also leads to higher economic returns throughout [Fig ].
+@Halpern2013], even without doing so, Model 1 maintains both a higher unharvested biomass and also leads to higher economic returns throughout [Fig 2].
 
@@
 Economic returns under Model 1 are also substantially higher (panel B)"}
-fig2a + fig2b
+
+sims <- read_csv("../data/sims.csv")
+npv_df <- read_csv("../data/npv_df.csv")
+
+plot_ecology <- 
+  sims %>%
+  mutate(model = as.character(model), resp = as.character(reps)) %>%
+  group_by(model, time) %>%
+  summarise(mean_state = mean(state), sd = sd(state), .groups = "drop") %>%
+  filter(time < 25, model != "3") %>%
+  ggplot(aes(time, mean_state)) + 
+  geom_line(aes(col = model), lwd=1.5, show.legend = FALSE) +
+    geom_ribbon(aes(ymin = mean_state - 2*sd, 
+                    ymax = mean_state + 2*sd,
+                    fill = model),
+                alpha = 0.2, show.legend = FALSE) +
+  ylab("state") + 
+  labs(subtitle = "A. Ecological outcomes")
+
+optimal <- select(filter(npv_df, model == "3"), time, mean_npv)
+
+plot_econ <- 
+npv_df %>%
+  mutate(model = as.character(model), resp = as.character(reps)) %>%
+  filter(model != "3", time %in% seq(1,100, by = 5)) %>%
+  ggplot(aes(time, mean_npv)) +
+  geom_line(data = optimal, lwd = 1.5, col = "grey20") +
+  geom_point(aes(col=model), size = 4, alpha = 0.8) + 
+  ylab("Net present value") + xlab("time")  + 
+  labs(subtitle = "B. Economic outcomes")
+
+plot_ecology + plot_econ
 ```
@@
 @Walters1978] incorporates the uncertainty over models directly into the decision calculations by integrating over the probability each model $M$ (or parameter value) is correct when calculating the utility; $U(x_t, a_t) = \int U(x_t, a_t | M) P(M)$.
-The probability assigned to each model can then be updated after each subsequent action in light of the resulting outcomes [@Ludgwig1982, @Smith1981].
+The probability assigned to each model can then be updated after each subsequent action in light of the resulting outcomes [@Ludwig1982, @Smith1981].
 While a the wide range of strategies for such iterative updating are known [@Polasky2011]; all will fail to select the higher-performing of the two simple models considered  here.
@@
 After a single iteration of learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct  [Fig].
-As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating [Fig , achieving a net present value over 100 time steps that only 31% that expected under management using Model 1 alone].
+As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating [Fig 2, achieving a net present value over 100 time steps that only 31% that expected under management using Model 1 alone].
 If instead of considering only these two models we consider a whole suite of Gordon-Schaefer-type models with varying $r$ and $K$ parameters, learning is slightly slower but no less counter-productive: mean stock size is $46%$ and net present value is $20%$ of what would be achieved under Model 1 (Appendix, Fig S1).
 
-```{r}
-am <- am1$df %>% mutate(belief = am1$posterior$V2)
-am %>%  ggplot(aes(time, states[state], col = belief)) + 
+```{r figure3, fig.cap="Adaptive management under model uncertainty, color indicates the belief that Model 2 is correct (red).
+Each model is assigned an initial prior belief.
+To ensure that any issues with passive exploration are not the cause of preferring Model 2, the initial belief in Model 2 is set to 1%.
+Within a single iteration of adaptive management, the belief over models is updated to near certainty in Model 2, resulting in higher harvests and lower stock sizes similar to managing under Model 2 alone, Fig 2."}
+am <- read_csv("../data/am.csv")
+am %>%  filter(time <30) %>%
+  ggplot(aes(time, states[state], col = belief)) + 
   geom_line() + 
@@
 
-```{r fig.cap="Panel A: Population growth curves of each model.
+```{r figure4,  fig.cap="Panel A: Population growth curves of each model.
 The positive equilibrium of each model occurs where the curve crosses the horizontal axis.
-Note that while Model 2 is a better approximation to the truth (Model 3), Model 1 better approximates the stock size which leads to maximum growth."}
+Note that while Model 2 is a better approximation to the truth (Model 3), Model 1 better approximates the stock size which leads to maximum growth.", fig.width=8}
 # Fig 4
-plot_models + plot_policies
+model_curves <- read_csv("../data/model_curves.csv") %>% mutate(model = as.character(model))
+policies <- read_csv("../data/policies.csv") %>% mutate(model = as.character(model))
+
+plot_models <- 
+  model_curves %>% 
+  ggplot(aes(state, value, col=model, lty=model)) +
+  geom_hline(aes(yintercept = 0), lwd=1) + 
+  geom_line(lwd=2, show.legend = FALSE) + 
+  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
+  ylab(bquote(f(x) - x)) + xlab("x")
+
+plot_policies <- policies %>%
+  ggplot(aes(states, escapement, col=model, lty=model)) + 
+  geom_line(lwd=2) + xlab("state")
+
+plot_models + labs(subtitle="A") +  plot_policies +  labs(subtitle="B")
 ```

### 2021-03-02 manuscript/manuscript.Rmd
:pen:

@@
 I illustrate the process of model selection by strictly proper scoring rules using two scenarios.
-In Scenario A (passive observation) the fish stock is unharvested and allowed to recover towards carrying capacity (as simulated under our "true" model, Model 3) while comparing the observed stock size in each subsequent time step to the distribution predicted under Model 1 and Model 2 respectively [Fig].
+In Scenario A (passive observation) the fish stock is unharvested and allowed to recover towards carrying capacity (as simulated under our "true" model, Model 3) while comparing the observed stock size in each subsequent time step to the distribution predicted under Model 1 and Model 2 respectively [Fig 1].
 The mean, $\mu$ and variance, $\sigma$ of the forecast are compared against the true observation $x$ using a Proper scoring rule of @Gneiting2007,
@@
 To demonstrate that the behavior is not driven by failure to explore sufficiently, (which might be addressed by an active adaptive management), I will assign initial probability that model 2 is true at 1%.
-After a single iteration of learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct  [Fig].
+After a single iteration of learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct  [Fig 4].
 As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating [Fig 2, achieving a net present value over 100 time steps that only 31% that expected under management using Model 1 alone].
@@
 Plotting the growth rate functions of each model, [Fig 4a], it is hardly surprising that no method which would prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3.
-Nevertheless, decisions based on model 1 are nearly indistinguishable from those based on the true model [Fig4b], while Model 2 leads to over-harvesting.
+Nevertheless, decisions based on model 1 are nearly indistinguishable from those based on the true model [Fig 4b], while Model 2 leads to over-harvesting.
 So why do our best approaches fail such a simple test case?
@@
 Is this result an artifact of the simple models considered here?
-A more realistic approach would consider uncertainty over model parameters, as well as models with different structural forms, such as Beverton Holt [@], Ricker [@], Shepard [@].
+A more realistic approach would consider uncertainty over model parameters, as well as models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980].
 Such complexity would not be sufficient to resolve the paradox of model uncertainty illustrated here.
@@
 To select Model 1 over alternatives like Model 2 which make better predictions but lead to worse outcomes requires a more long-term approach that forgoes iterative updating of probabilities as new data becomes available.
-This is the opposite recommendation of _adaptive_ management  @Dietze2018
+Note that this is the opposite recommendation of _adaptive_ management [@Walters1978] or _iterative_ forecasting [@Dietze2018].
 
@@
 Unfortunately, such innaccurate predictions are only too common in the real world.
-Have any of those led to effective policy despite the innnaccuracy?
+Have any of those led to effective policy despite the innaccuracy?
 Have any been replaced with more accurate models that have also driven worse decisions?

### 2021-03-05 manuscript/manuscript.Rmd
millie comments

@@
 ```
+
+<!-- I think the introduction could have a clearer storyline and better introduce some of the ideas you bring up later on in the methods/results. A quick outline idea:
+
+  - Paragraph1: Model-based forecasts have become a cornerstone of applied science in disciplines as diverse as atmospheric science [weather?], economics [stock markets?], and ecology [dietz].
+Seeking to inform a variety of decisions, near-term forecasting in these diverse fields has  simultaneously been encouraged by policy-makers [CITE?] and enabled by the rapid and recent increase in data availability and computational methods leading to a rapid expansion of thier use (or something...)
+
+   - Paragraph 2: Introduce ecological decision making under uncertainty, rapid global change, etc. (kind of like your current opening bit)
+
+  - Paragraph 3: But do better forecast always lead to better decisions?
+Because model-based forecasts frequently play an important role in decision-making,it is commonly assumed [e.g.
+@Walters1981; @Clark2001; @Dietze2018] that this approach for addressing model uncertainty will also improve decision-making outcomes.
+            - While the literature has largely addressed issues of overfittting that might lead forecasts to misguide decisions, proper scoring methods blah blah [...introduce this concept]. 
+            - Beyond overfitted models,... [maybe add something about Value of Information literature here?]
+
+  - Paragraph 4: Your current paragraph on "I illustrate..." (I'd add into that paragraph something about the fact that you explore adaptive management?)
+
+A second idea on framing: this maybe is a to divergent from the main point (or leads you into a post-modernist narrative that you'd prefer to avoid...) but it seems like you could reframe the whole thing as an argument for embracing an inclusive process of resource management that acknowledges pluralism.
+  (1) Policy makers often think "scientists" hold knowledge of "an environmental reality"
+  (2) Despite this not being true (scientist disagree on a lot of things), even when "science" points to a more likely reality, that might be off in ways that can fundementally misguide decisions. 
+  (3) The problem is deeper than acknowledging a "rashamon set" of models or multiple possible descriptions of of reality.
+Sometimes what appears closer to reality is actually off in a way that is critical to the decision.
+-->
+
+<!-- Open on real-world issues, uncertainty, particularly model uncertainty as key theme.  -->
 
@@
 
+<!-- I feel like parts of the below paragrph should be in the abstract (i.e.
+"I illustrate how the most promising techniques from statistical approaches and decision theoretic approaches to decision making under uncertainty would be applied to this simple problem, and demonstrate that it both cases they lead us away from the model that produces the most desirable decisions towards worse outcome") -->
+
 A wide range of paradigms are available for approaching the issue of decision-making under uncertainty.
@@
 
+<!-- I think this subsection should be renamed.
+it makes it a bit hard to follow the structure of the larger section.
+Seems like something like: "Statistical approaches: forecasts and proper scores" idk, but something that more closely aligns with the above paragraph and the second subsection -->
+
 Like many decision problems, the task of setting a sustainable harvest quota appears to hinge on having an accurate forecast: if we can predict to what size the fish stock will increase next year, $X_t+1$, and we know the current stock, $X_t$, then we can safely harvest $X_{t+1} - X_t$.
+
+<!-- it is unclear what you mean by safely harvest?
+as in the stock wont crash?
+or it won't be worse than it is this year?
+You could imaging that we could safetly harvest more (or only less) than that.-->
+
 Overestimating or underestimating such recruitment will result in over-harvesting or under-harvesting, respectively.
+
 Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of over-fitting by comparing model predictions to later observations that were not used to estimate the model.
+
 @Gneiting2007 provides a rigorous proof for "proper" scoring criteria, which have the desirable property which no model predicting the distribution of future outcomes, $Q(x)$ can achieve a better average score than the true model $P(x)$.
@@
 These features have made proper scoring rules for probabilistic forecasts a successful and popular approach for addressing model uncertainty in many other areas [@Gneiting2014; @Raftery2016] and a promising tool for evaluation of ecological forecasts [@Dietze2018].
+
+<!-- This scoring bit comes abruptly (and it feels like could be in the introduction) -->
 
@@
 
+I think it would be effec
+
 # Discussion
+<!--
+hit bigger themes first & objections first
+- Learning can be counter-productive
+- Value of information can be negative in practice
+
+Objections:
+- we already know simple models can be better, overfit models can lead us astray
+- we already know Knightian uncertainty matters.  (underestimating uncertainty isn't the same thing)
+
+-->
 
@@
 I address these in turn.
+
+<!-- (3) THIS problem (4) THIS problem. I like that you outline the points you are going to cover! -->
 
@@
 
-<!-- move to appendix?  too confusing/too long?-->
+<!-- move to appendix?  too confusing/too long?
+
+Not sure - it feels a little extraneous (I did think this was an interesting point brought up by reviewer 1 but I also don't necessarily think it warrants a long paragraph that might confuse the main point rather than clarify something) Could this be a couple sentences and reference an appendix that walks through what you write out (potentially even with a simple constant mortality example)
+-->
 Is this result an artifact of optimal control?
@@
 
+<!-- I really like the last paragraph!! -->
+
 Perhaps Model 1 is a unicorn, a mythically simple model that gives optimal results but has no analogue in the real world.
@@
 
-<!-- Omit this?  Does it beg an example? -->
-<!--
+<!-- Omit this?  Does it beg an example?
+
 There is at least one technique which might address the problem presented here.
@@
 
+<!--I was thinking about exactly this while I was reading the paper.
+I do think this would likely beg an example, which might be a can of worms that you don't want to open with this piece...
+I also don't think that I think it would address the presented problem.
+(Kind of repeating my second framing point but) It reminds me of the "Stop Explaining Black Box Machine Learning Models for High Stakes Decisions and use Interpretable Models Instead" paper where she talks about how, for a given data set, you can define the Rashomon set as the set of reasonably accurate predictive models-  Because the data are finite, the data could admit many close-to-optimal models that predict differently from each other: a large Rashomon set.
+I suspect this happens often in practice because sometimes many different ML algorithms perform similarly on the same data set, despite having different functional forms (for example, random forests, neural networks, support vector machines).
+As long as the Rashomon set contains a large enough set of models with diverse predictions, it probably contains functions that can be approximated well by simpler functions, and so the Rashomon set can also contain these simpler functions.
+Said another way, uncertainty arising from the data leads to a Rashomon set; a larger Rashomon set probably contains interpretable models, thus interpretable accurate models often exist.
+If this theory holds, we should expect to see interpretable models exist across domains.
+These interpretable models may be hard to find through optimization, but at least there is a reason we might expect that such models exist.
+
+The problem is deeper than acknowleding a "rashamon set of models" or a plurality of reality (relevant to that Bioscience paper I sent last week about the Rashamon effect in conservation).
+Sometimes what appears closer to reality is actually fundementally wrong in some way that is misguiding to a decision maker...more opaque approaches could lessen our capacity to trouble shoot that?) -->
+
 \pagebreak

### 2021-03-05 manuscript/manuscript.Rmd
wip

@@
 Global change issues are complex and outcomes are difficult to predict, making approaches to uncertainty a central part
-of effective decision-making [@ @]. While some uncertainty is intrinsic to the underlying processes (stochasticity) or by
+of effective decision-making [@ @]. Reducing that uncertainty by improving models in response to more data
+becomes available
+
+While some uncertainty is intrinsic to the underlying processes (stochasticity) or by
 limits to what variables we can observe and how accurately we measure them (measurement uncertainty), much uncertainty 
@@
 _overfit_ to patterns arising from by chance from the stochasticity or measurement error which do not reflect the
-underlying process. Researchers have often favored simpler models which are less prone to over-fitting, sometimes
+underlying process [@Ginzburg2004]. Researchers have often favored simpler models which are less prone to over-fitting, sometimes
 through explicit penalties such as information criteria [@], though these can be misleading [@]. With the rapid 
@@
 In passive adaptive management, the manager chooses the action which maximizes that expected future utility.
-After observing the consequences of the action, the manager updates the probabilities each model is correct, typically by Bayes rule [e.g.
-@Walters1981; @Smith1981; @Ludwig1982].
+After observing the consequences of the action, the manager updates the posterior probabilities that each model is correct, typically by Bayes rule [e.g.
+@Walters1981; @Smith1981; @Ludwig1982; @Punt2016].
+The decision strategy is then re-evaluated in light of this revised uncertainty.
+In many real-world applications, it may be impossible or computationally prohibitive to solve the optimal control policy given all sources of uncertainty, as is typical in fisheries management practice [@Punt2016].
+Management Strategy Evaluation (MSE) evaluates
+
 In active adaptive management, the key difference is that the manager may chose an action which does not maximize the utility _given the current model uncertainty_, but instead, takes an action whose consequences will more quickly discriminate between the models and thus more rapidly reduce the uncertainty.
@@
 Under some scenarios the cost of this 'exploration' step can be made up later by actions based on the more precise knowledge gained as a result ('exploitation').
-Alternative approaches such as scenario analysis can be considered approximations of this approach.
+In practice, passive adaptive management approaches are more common, such as Management Strategy Evaluation (MSE @Punt2016) or Scenario Analysis [@Polasky2011].
+These approaches all share the same adaptive aspect,
+
+Owing to the computational complexity involved, these approaches typically evaluate a suite of proposed policies, rather than seeking to derive the optimal policy.
 A scenario analysis does not rely on quantitative estimates of model probabilities or expected utility, but still consider possible outcomes over a suite of models (scenarios), select actions to minimize the risk or or other suitable objective, and update or eliminate scenarios deemed implausible over time [@Polasky2011].

### 2021-03-06 manuscript/manuscript.Rmd
:pen:

@@
 The expected utility of any action reflects both the intrinsic uncertainty over future states under either model (process uncertainty) and the manager's uncertainty over the choice of models (or parameter values).
-Unlike the forecast comparison, an adaptive management approach explicitly considers the utility function, Eqn \eqref{utility}, which it seeks to maximize by intergrating over model uncertainty.
-
-In passive adaptive management, the manager chooses the action which maximizes that expected future utility.
+Unlike the forecast comparison, an adaptive management approach explicitly considers the utility function, Eqn \eqref{utility}, which it seeks to maximize by integrating over model uncertainty.
 After observing the consequences of the action, the manager updates the posterior probabilities that each model is correct, typically by Bayes rule [e.g.
@@
 The decision strategy is then re-evaluated in light of this revised uncertainty.
-In many real-world applications, it may be impossible or computationally prohibitive to solve the optimal control policy given all sources of uncertainty, as is typical in fisheries management practice [@Punt2016].
-Management Strategy Evaluation (MSE) evaluates
+Adaptive management may be either passive or active.
+Under passive adaptive management, the decision-maker chooses the action which maximizes that expected future utility.
+Under active adaptive management, the decision-maker may chose 'exploratory' actions which are not optimal under the current uncertainty but may lead to more rapidly decreasing the uncertainty in the future [@Walters1978].
+In practice, many managers are reluctant to employ exploratory strategies, so passive adaptive management approaches such as as Management Strategy Evaluation (MSE) are more common [@Punt2016].
 
-In active adaptive management, the key difference is that the manager may chose an action which does not maximize the utility _given the current model uncertainty_, but instead, takes an action whose consequences will more quickly discriminate between the models and thus more rapidly reduce the uncertainty.
-After observing the outcome of this exploratory step, model probabilities are updated as before.
-Under some scenarios the cost of this 'exploration' step can be made up later by actions based on the more precise knowledge gained as a result ('exploitation').
-In practice, passive adaptive management approaches are more common, such as Management Strategy Evaluation (MSE @Punt2016) or Scenario Analysis [@Polasky2011].
-These approaches all share the same adaptive aspect,
+<!--
+Any of these adaptive management approaches proves to be maladaptive under the simple example of model uncertainty considered here.
+That is, the manager derives lower utility (here, economic yield), and the fish stock is also surpressed to a lower biomass than would be achieved if an identical management approach without any adaptive updating to model uncertainty were employed.
+Put another way, the "value of information" gained from future observations of stock in response to any management action is actually negative.
 
-Owing to the computational complexity involved, these approaches typically evaluate a suite of proposed policies, rather than seeking to derive the optimal policy.
-A scenario analysis does not rely on quantitative estimates of model probabilities or expected utility, but still consider possible outcomes over a suite of models (scenarios), select actions to minimize the risk or or other suitable objective, and update or eliminate scenarios deemed implausible over time [@Polasky2011].
+Owing to the greater complexity involved in many fisheries models, a typical MSE would evaluate a fixed set of possible strategies rather than solve a sequential decision problem [@Marescot2013] for the optimal sequence of actions, but the central premise of iteratively updating model posterior probabilities remains the same.
+By limiting ourselves to the simpler models considered here, we are able to perform a passive adaptive management analysis of the full sequential decision problem, analgous to earlier examples [@Walters1981; @Smith1981; @Ludwig1982], which ensures the result is not an artifact of considering too narrow a set of possible policies (e.g.
+only constant-mortality policies).
 
-Each of these approaches swiftly decides that model 1 in our example is implausible, and settles for making decisions driven almost exclusively by model 2.
-I illustrate this using a passive adaptive management strategy following classic examples for parameter [@Ludwig1982] or structural [@Smith1981] model uncertainty.
+-->
+
+Unfortunately, any adaptive management strategy which updates posterior distributions that characterize model undertainty leads to worse outcomes than an approach which does no updating.
+I illustrate the application of a passive adaptive management strategy to this simple example, following classic examples for parameter [@Ludwig1982] or structural [@Smith1981] model uncertainty.
 To demonstrate that the behavior is not driven by failure to explore sufficiently, (which might be addressed by an active adaptive management), I will assign initial probability that model 2 is true at 1%.
@@
 
-I think it would be effec
-
 # Discussion
@@
 
+- is this the result of underestimating the uncertainty?
+
 -->
@@
 Given this simplified version of model uncertainty in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariably choose the other.
+Importantly, a decision maker employing an adaptive management or iterative forecasting assessment such as those considered here would have no way of realizing that the outcomes they experienced were in fact sub-optimal.
+The manager quickly concludes that Model 1 is entirely implausible, but also finds Model 2 to be remarkably accurate at predicting future values, and is entirely unaware that the model is leading to such poor outcomes.
+A more diligent manager may seek to improve Model 2 further by re-estimating parameters, but this in fact leads to a model with slightly lower $K$ and thus even higher level of over-fishing and under-performance than Model 2.
+The only way to avoid this trap is not to learn with each new observation, not to apply iterative updates over the models.
+A manager must follow the decisions proposed by model 1 for many iterations to realize that it leads to more desirable outcomes.
+This raises several questions.
+Do such methods of learning and reducing model uncertainty ever lead us towards worse long-term outcomes in the real world?
+How would we know?
+And what should we do about it?
+
+It is tempting to argue that this situation would almost never arise in real world scenarios.
+In this argument, Model 1 is a unicorn, a mythical model that somehow gives optimal performance while at the same time making predictions that are wildly wrong.
+Could such unicorns really exist?
+This is a difficult assertion to prove either way, but it is worth noting that neither of these features by themselves are especially rare.
+Examples of serious forecasts that widely miss future observations abound wherever forecasts are common, from elections to economics to environmental change [@].
+Meanwhile, the premise that model need not perfectly capture reality to be useful is at the very heart of modeling.
+Model 1 successfully captures the one key feature driving decisions in optimal harvest control problems: the stock size at which the maximum growth rate occurs [Fig 4a].
+Capturing only the essential aspects as simply as possible is the goal of any model building exercise.
+Moreover, rigorous updating of models and parameters through approaches such as adaptive management or iterative forecasting are not yet widespread practices in ecological management [@Dietze2018].
+If such unicorn models exist, that lack of adoption would make it easier for them to persist at least until now.
+
+If these unicorns do exist in real world management, then what do we do about it?
+Our step may be to acknowledging that we do not yet have the answer.
+There are at least three common ideas which are not the answer, which might succinctly be termed (1) don't care, (2) don't learn or (3) don't model.
+
+The first response, don't care, suggests that iterative updating is always desirable.
+Previous work has acknowledged that future observations may be relatively uninformative, failing to reduce uncertainty between models by much.
+This is the motivation for active over passive adaptive management, [@Walters1978], and has also acknowledge that even significant improvements in predictive accuracy may have negligible improvement on the quality of the decision.
+This is the motivation for quantifying the value of information, [@Katz1987].
+However, the conclusion from such observations is that additional information is at worst harmless to the quality of the decision, while in this example it is actively detrimental.
+The realized 'value of information' contained in future measurements is negative.
+
+The
+
+Why then do our best available approaches so readily reject the model?
+
 The reason for this may appear obvious once we compare both curves to that of the underlying model, Model 3.
 Plotting the growth rate functions of each model, [Fig 4a], it is hardly surprising that no method which would prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3.
-Nevertheless, decisions based on model 1 are nearly indistinguishable from those based on the true model [Fig 4b], while Model 2 leads to over-harvesting.
+Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 4b], while Model 2 leads to over-harvesting.
+We have no methods of reducing model uncertainty which do not perform significantly worse on this management problem than an approach which does not attempt to learn.
+Put another way, the value of information contained in future observations [@] is always negative in this example.
+
 So why do our best approaches fail such a simple test case?
-It is tempting to dismiss this example as having little relevance to real-world decision problems on a variety of grounds: (1) this problem is an artifact of overly-simplified models, (2) this problem is an artifact of an unrealistic optimal control solution, (3) the problem arises from having started with different objectives, (4) the problem is already solved.
+It is tempting to dismiss this example as having little relevance to real-world decision problems on a variety of grounds:   (3) Don't we already know this?
+
+(1) this problem is an artifact of overly-simplified models, (2) this problem is an artifact of an unrealistic optimal control solution, (3) the problem arises from having started with different objectives, (4) the problem is already solved.
 I address these in turn.
@@
 ```
+
+This issue is not unique to sequential decision problems or optimal control solutions.
+For example, Management Strategy Evaluation [@Punt2016] in fisheries seeks to evaluate pre-specified strategies rather solve for the best possible strategy through optimal control.
+This approach is well justified -- optimal control techniques such as stochastic dynamic programming illustrated here quickly become intractable for more complex models [@Marescot2013] typically used in fisheries stock assessments.
+Constraining the search to only strategies that impose a constant mortality (defined as harvest per unit biomass, $F = H/B$) means we only have to evaluate those $N$ strategies each iteration, not solve a sequential decision problem.
+Doing so, we would find the best constant-mortality solution under Model 1 performs much worse than the best constant-mortality solution under Model 2.
+Does this mean MSE is not suseptible to this issue?
+No, it does not.
+It is just as easy to construct an alternative Model 1 with the same properties of leading to nearly optimal decisions while being rejected by any method of iterative learning, such as forecasting or adaptive management.
+Just as the optimal control policy for a Gordon-Schaefer (logistic) model depends only on parameter $K$ [@Reed1979], the optimum constant escapement policy depends only on parameter $r$ [@Schaefer1954].
+If Model 1 has a value of $r$ such that it happens to match the best possible constant-mortality solution for the (unobserved) true model, then $K$ can be set arbitrarily high, ensuring the any model selection, forecasting, or adaptive management approach would lead away from Model 1 and towards worse-performing models once again.
 

### 2021-03-08 manuscript/manuscript.Rmd
:pen:

@@
   Model-based forecasts have been enabled by recent explosion of data and computational methods and spurred on by decision-makers appetite for forecasts in everything from elections to pandemic response.
-It is taken for granted that the model which makes the most accurate forecast, accounting for uncertainty, will also be the best model to inform decision-making.
 Using a classic example from fisheries management, I demonstrate that selecting the model that produces the most accurate and precise forecast can lead to decidedly worse outcomes.
@@
 keywords:
-- transients,
-- optimal control,
+- forecasting,
 - adaptive management,
@@
 - uncertainty,
-- ecological management
+- optimal control
 
@@
 
-Given this simplified version of model uncertainty in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariably choose the other.
-Importantly, a decision maker employing an adaptive management or iterative forecasting assessment such as those considered here would have no way of realizing that the outcomes they experienced were in fact sub-optimal.
-The manager quickly concludes that Model 1 is entirely implausible, but also finds Model 2 to be remarkably accurate at predicting future values, and is entirely unaware that the model is leading to such poor outcomes.
-A more diligent manager may seek to improve Model 2 further by re-estimating parameters, but this in fact leads to a model with slightly lower $K$ and thus even higher level of over-fishing and under-performance than Model 2.
-The only way to avoid this trap is not to learn with each new observation, not to apply iterative updates over the models.
-A manager must follow the decisions proposed by model 1 for many iterations to realize that it leads to more desirable outcomes.
+Given this simple decision problem in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariably choose the other.
+Moreover, a decision maker employing an adaptive management or iterative forecasting assessment such as those considered here would have no way of realizing that the outcomes they experienced were in fact sub-optimal.
+In both approaches, the manager quickly concludes that Model 1 is entirely implausible, while finding that Model 2 is remarkably accurate at predicting future values.
+This problem is not addressed by re-estimating parameters (equivalently, considering a larger suite of models with all possible $K$ and $r$ values): in fact leads to a model with slightly lower $K$ and even higher level of over-fishing and under-performance than Model 2 (Appendix B, Fig S1).
+
+The reason for Model 1's seemingly contradictory ability to make good decisions but bad forecasts becomes obvious once we compare both curves to that of the underlying model, Model 3.
+Plotting the growth rate functions of each model, [Fig 4a], it is hardly surprising that no method exists which would not prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3.
+However, Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 4b], while Model 2 leads to over-harvesting.
+The explanation comes from noticing that the stock size corresponding to the maximum growth rate under Model 1 (the peak of the curve) falls at almost exactly the same stock size as that of the peak growth rate for Model 3.
+Meanwhile, the peak of Model 2 occurs at a substantially lower stock size.
+While the optimal control solution appears to depend only on a step-ahead forecast accuracy (indeed, the SDP solution method used here takes only step-ahead forecast probabilities as input, [@Marescot2013]), mathematical analysis showed long ago [@Reed1979] that the optimal solution for this problem depends only on keeping the biomass at the value responsible for the maximum growth rate.
+This realization is quite general: for most decision problems, simple models will exist under which the optimal decision is the same as it would be for the true model, even when that simple model is wrong in most other ways.
+
+This phenomenon is not unique to sequential decision problems or optimal control solutions.
+For example, Management Strategy Evaluation [@Punt2016] in fisheries seeks to evaluate pre-specified strategies rather solve for the best possible strategy through optimal control.
+This approach is well justified -- optimal control techniques such as stochastic dynamic programming illustrated here quickly become intractable for more complex models [@Marescot2013] typically used in fisheries stock assessments.
+Constraining the search to only strategies that impose a constant mortality (defined as harvest per unit biomass, $F = H/B$) means we only have to evaluate those $N$ strategies each iteration, not solve a sequential decision problem.
+Doing so, we would find the best constant-mortality solution under Model 1 performs much worse than the best constant-mortality solution under Model 2.
+Does this mean MSE is not susceptible to this issue?
+No, it does not.
+It is just as easy to construct an alternative Model 1 with the same properties of leading to nearly optimal decisions while being rejected by any method of iterative learning, such as forecasting or adaptive management.
+Just as the optimal control policy for a Gordon-Schaefer (logistic) model depends only on parameter $K$ [@Reed1979], the optimum constant escapement policy depends only on parameter $r$ [@Schaefer1954].
+If Model 1 has a value of $r$ such that it happens to match the best possible constant-mortality solution for the (unobserved) true model, then $K$ can be set arbitrarily high, ensuring the any model selection, forecasting, or adaptive management approach would lead away from Model 1 and towards worse-performing models once again.
+
+Note that the mechanism shown here has nothing to do with the much more familiar issue of over-fitting, which also involves a better-fitting model leading to worse outcomes [].
+In fact, an model which has been over-fit will gradually be rejected by either the iterative forecasting or adaptive management approaches shown here, as these approaches continually confront the models with new data to which they had not been previously fit.
+The ability to avoid over-fitting is one of the greatest appeals of any iterative management strategy.
+But in the example here, both Model 1 and Model 2 have the same structural complexity.
+The better decision performance of Model 1 is a consequence of capturing the key attribute needed for a good decision, which is perhaps a surprisingly an all-together different criteria than fit.
+
+Previous work has acknowledged that iterative learning may not always be feasible and that it may not always be beneficial, but has failed to recognize the possibility that it can potentially be detrimental.
+For example, active adaptive management is premised on the observation that future observations may be too uninformative to distinguish between alternatives [@Walters1978].
+Others have also noted that even big improvements in predictive accuracy could have negligible improvement on the decision outcomes, which motivates quantifying the Value of Information (VOI) [@Katz1987].
+But in both cases, the worse outcome is wasted effort.
+In the scenario considered here, the VOI is negative: any learning over the proposed model uncertainty leads to lower expected net utility than not learning.
+The only way to avoid this trap is to _forego learning_ with each new observation, to avoid iterative updates.
+A manager must follow the decisions proposed by model 1 for many iterations to realize that it leads to more desirable outcomes than alternative models.
 This raises several questions.
 Do such methods of learning and reducing model uncertainty ever lead us towards worse long-term outcomes in the real world?
-How would we know?
-And what should we do about it?
+If so, How would we know, and what should we do about it?
+I address these in turn.
 
-It is tempting to argue that this situation would almost never arise in real world scenarios.
-In this argument, Model 1 is a unicorn, a mythical model that somehow gives optimal performance while at the same time making predictions that are wildly wrong.
-Could such unicorns really exist?
-This is a difficult assertion to prove either way, but it is worth noting that neither of these features by themselves are especially rare.
-Examples of serious forecasts that widely miss future observations abound wherever forecasts are common, from elections to economics to environmental change [@].
-Meanwhile, the premise that model need not perfectly capture reality to be useful is at the very heart of modeling.
+<!-- **Is it just a unicorn?** -->
+Do models which give nearly optimal performance while at the same time making wildly wrong predictions really exist?
+It is tempting to argue that Model 1 is a kind of unicorn, a mythical creature existing only in theory.
+But that is a difficult assertion to prove.
+Examples of serious forecasts that widely fail to predict future observations abound wherever forecasts are common, from elections to economics to environmental change [@].
+If there is no shortage of bad forecasts, then could any of them be useful?
+The premise that model need not perfectly capture reality to be useful is at the very heart of modeling.
 Model 1 successfully captures the one key feature driving decisions in optimal harvest control problems: the stock size at which the maximum growth rate occurs [Fig 4a].
 Capturing only the essential aspects as simply as possible is the goal of any model building exercise.
-Moreover, rigorous updating of models and parameters through approaches such as adaptive management or iterative forecasting are not yet widespread practices in ecological management [@Dietze2018].
-If such unicorn models exist, that lack of adoption would make it easier for them to persist at least until now.
+Thus it should be no surprise that a model can drive good decisions while making poor predictions.
+It is also important to note that in many cases, any models being used in decision-making are not necessarily being subjected to the rigorous evaluation and updating steps proposed by adaptive management or iterative forecasting [@Dietze2018].
+This makes it more likely that such models could persist in practice until now.
+While iterations that revise this models will no doubt improve decision outcomes in many case, it is worth bearing in mind from the example here that such a connection is not guaranteed.
 
-If these unicorns do exist in real world management, then what do we do about it?
-Our step may be to acknowledging that we do not yet have the answer.
-There are at least three common ideas which are not the answer, which might succinctly be termed (1) don't care, (2) don't learn or (3) don't model.
+<!-- So what? -->
+If these unicorns do exist in real world management, then what do we do about them?
+I believe this is an open question, but that our first step must be to recognize it as such.
+I contrast that position with two common but diametrically opposed objections: (1) that the issue is easily resolved by including more or better models, or (2) this issue is easily resolved by relying less on models.
+These are both important points.
 
-The first response, don't care, suggests that iterative updating is always desirable.
-Previous work has acknowledged that future observations may be relatively uninformative, failing to reduce uncertainty between models by much.
-This is the motivation for active over passive adaptive management, [@Walters1978], and has also acknowledge that even significant improvements in predictive accuracy may have negligible improvement on the quality of the decision.
-This is the motivation for quantifying the value of information, [@Katz1987].
-However, the conclusion from such observations is that additional information is at worst harmless to the quality of the decision, while in this example it is actively detrimental.
-The realized 'value of information' contained in future measurements is negative.
+The union of two simple models considered here (Models 1 & 2) certainly fails to fully capture the inherent uncertainty in the underlying dynamics, and the consequences of underestimating model uncertainty are well understood [].
+Had we included the true model in the set of possibilities, the techniques illustrated would have had little difficulty in distinguishing it from Models 1 and 2 after sufficient iterations.
+However, it must also be noted that we could have included a much more realistic approach of uncertainty over many possible values of $r$, $K$ (see Appendix B, Fig S1), or even among common alternative models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980], without altering the results at all (these models also all have symmetric growth rates with a peak growth rate at half the steady-state population size).
+Models will always be simpler than reality.
+Surely it would be preferable if our methods could prefer the nearly optimally-performing Model 1 which we already have than require the issue be solved only by adding more and more models?
 
-The
-
-Why then do our best available approaches so readily reject the model?
-
-The reason for this may appear obvious once we compare both curves to that of the underlying model, Model 3.
-Plotting the growth rate functions of each model, [Fig 4a], it is hardly surprising that no method which would prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3.
-Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 4b], while Model 2 leads to over-harvesting.
-We have no methods of reducing model uncertainty which do not perform significantly worse on this management problem than an approach which does not attempt to learn.
-Put another way, the value of information contained in future observations [@] is always negative in this example.
-
-So why do our best approaches fail such a simple test case?
-It is tempting to dismiss this example as having little relevance to real-world decision problems on a variety of grounds:   (3) Don't we already know this?
-
-(1) this problem is an artifact of overly-simplified models, (2) this problem is an artifact of an unrealistic optimal control solution, (3) the problem arises from having started with different objectives, (4) the problem is already solved.
-I address these in turn.
+If we cannot always resolve the issue by adding more models, is this just another example of the risks of an over-reliance on models at all?
 
@@
 ```
-
-This issue is not unique to sequential decision problems or optimal control solutions.
-For example, Management Strategy Evaluation [@Punt2016] in fisheries seeks to evaluate pre-specified strategies rather solve for the best possible strategy through optimal control.
-This approach is well justified -- optimal control techniques such as stochastic dynamic programming illustrated here quickly become intractable for more complex models [@Marescot2013] typically used in fisheries stock assessments.
-Constraining the search to only strategies that impose a constant mortality (defined as harvest per unit biomass, $F = H/B$) means we only have to evaluate those $N$ strategies each iteration, not solve a sequential decision problem.
-Doing so, we would find the best constant-mortality solution under Model 1 performs much worse than the best constant-mortality solution under Model 2.
-Does this mean MSE is not suseptible to this issue?
-No, it does not.
-It is just as easy to construct an alternative Model 1 with the same properties of leading to nearly optimal decisions while being rejected by any method of iterative learning, such as forecasting or adaptive management.
-Just as the optimal control policy for a Gordon-Schaefer (logistic) model depends only on parameter $K$ [@Reed1979], the optimum constant escapement policy depends only on parameter $r$ [@Schaefer1954].
-If Model 1 has a value of $r$ such that it happens to match the best possible constant-mortality solution for the (unobserved) true model, then $K$ can be set arbitrarily high, ensuring the any model selection, forecasting, or adaptive management approach would lead away from Model 1 and towards worse-performing models once again.
-
-Is this result an artifact of the simple models considered here?
-A more realistic approach would consider uncertainty over model parameters, as well as models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980].
-Such complexity would not be sufficient to resolve the paradox of model uncertainty illustrated here.
-Simplicity facilitates intuition: Model 1 performs best because it correctly approximates the stock size which achieves maximum growth rate under the true model.
-In contrast, each of those widely used recruitment relationships are symmetric, and their best-fitting parameter estimates will always have a maximum growth rate at a lower stock size.
-It is easy to point out that even this wider array of parameter and model values is too simple, since fisheries models in practice are much more complicated still (indeed, too complicated for the formal integration over model uncertainty considered here).
-It is indeed true that either the forecasting or adaptive management example would work as expected if the set of candidate models included the true model, or at least a functional form that could better approximate the peak of the true model while also fitting better than model 2.
-But we never have the true model in real systems.
-If the candidate models are only caricatures, so too is the "true" model.
-The candidate models considered here are no doubt much closer approximations to model 3 (all being single species, unstructured in age, stage, time or space, etc) than any model used in practice is to reality.
-Morever, a more accurate model is not required to reach optimal outcomes -- Model 1 already provides this by correctly approximating the key feature in this problem, even while it is wrong about much else.
-How do proceed when our best methods fail in such a simple case?
 

### 2021-03-10 manuscript/manuscript.Rmd
:pen:

@@
 # Discussion
+
 <!-- 
-hit bigger themes first & objections first
-- Learning can be counter-productive
-- Value of information can be negative in practice
-
-Objections:
-- we already know simple models can be better, overfit models can lead us astray
-- we already know Knightian uncertainty matters.  (underestimating uncertainty isn't the same thing)
-
-- is this the result of underestimating the uncertainty?
+First: Intuition for the result.
+  - See that model 1 captures the key feature, while model 2 is just 'closer to the curve'
+  - See that this is a generic truth about decision-making: there's always a Model 1
+  - This is not about over-fitting.
 
 -->
-
 Given this simple decision problem in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariably choose the other.
@@
 
-Note that the mechanism shown here has nothing to do with the much more familiar issue of over-fitting, which also involves a better-fitting model leading to worse outcomes [].
+Note that the mechanism shown here has nothing to do with the much more familiar issue of over-fitting, in which a better-fitting model will also lead to worse outcomes [].
 In fact, an model which has been over-fit will gradually be rejected by either the iterative forecasting or adaptive management approaches shown here, as these approaches continually confront the models with new data to which they had not been previously fit.
@@
 The better decision performance of Model 1 is a consequence of capturing the key attribute needed for a good decision, which is perhaps a surprisingly an all-together different criteria than fit.
+
+This case cannot be dismissed merely as being dealt a bad hand in having to pick only between Model 1 and Model 2.
+Model 1 approximates the key feature, giving nearly optimal outcomes.
+Previous literature has underscored the importance of Knightian "unknown unknowns" two alternative models examined here fail to fully reflect our uncertainty in the underlying dynamics, and the consequences of underestimating model uncertainty are well understood [].
+Had we included the true model in the set of possibilities, the techniques illustrated would have had little difficulty in distinguishing it from Models 1 and 2 after sufficient iterations.
+In practice, we never have the true model, and even a much more thorough suite of candidate models, such as considering any possible values of $r$, $K$ (see Appendix B, Fig S1), or even among common alternative models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980], would fail to do any better, since all these models still have symmetric growth rates with a peak growth rate at half the steady-state population size).
+Models will always be simpler than reality.
+To insist that this issue can and will be avoided by always including more and better models of the process misses the point.
+
+<!--
+So what?
+
+- This is different than previous work: e.g. VOI is not zero, it's negative here.
+- Model 1 is not a unicorn: we cannot rule out these situations in real world (particularly ecology where we don't actually do much iterative updating in practice)
+- 
+-->
 
@@
 In the scenario considered here, the VOI is negative: any learning over the proposed model uncertainty leads to lower expected net utility than not learning.
+
 The only way to avoid this trap is to _forego learning_ with each new observation, to avoid iterative updates.
-A manager must follow the decisions proposed by model 1 for many iterations to realize that it leads to more desirable outcomes than alternative models.
+It may seem that the adaptive management approach fails because posterior probabilities are updated according to Bayes rule.
+Like iterative forecasting, this favors models which better predict the data, despite the fact that under the adaptive management approach, the overall optimization is conditioned on actual utility and not model fit.
+However, this issue is not easily avoided.
+For example, so-called greedy optimization techniques that favor actions with higher immediate reward do even worse in this context, since such an algorithm would obviously maximize the immediate harvest and therefore collapse the stock.
+Only by comparing the net utility derived from management under model 1 for many iterations to the net utility derived under model 2 after many iterations can we deterime that model 1 leads to better outcomes.
 This raises several questions.
@@
 If so, How would we know, and what should we do about it?
-I address these in turn.
 
@@
 I believe this is an open question, but that our first step must be to recognize it as such.
-I contrast that position with two common but diametrically opposed objections: (1) that the issue is easily resolved by including more or better models, or (2) this issue is easily resolved by relying less on models.
-These are both important points.
+(1) that the issue is just an example of the already-understood problem of underspecifying model uncertainty, which is resolved by including more or better models, or (2) this issue is just an example of the already-understood limitations of relying too much on models to shape decision-making.
+Either we need more models, or we need fewer models.
 
-The union of two simple models considered here (Models 1 & 2) certainly fails to fully capture the inherent uncertainty in the underlying dynamics, and the consequences of underestimating model uncertainty are well understood [].
-Had we included the true model in the set of possibilities, the techniques illustrated would have had little difficulty in distinguishing it from Models 1 and 2 after sufficient iterations.
-However, it must also be noted that we could have included a much more realistic approach of uncertainty over many possible values of $r$, $K$ (see Appendix B, Fig S1), or even among common alternative models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980], without altering the results at all (these models also all have symmetric growth rates with a peak growth rate at half the steady-state population size).
-Models will always be simpler than reality.
-Surely it would be preferable if our methods could prefer the nearly optimally-performing Model 1 which we already have than require the issue be solved only by adding more and more models?
-
+If to advocates of iterative model updating approaches the solution is always more models, then to skeptics the solution is fewer models and algorithms.  
 If we cannot always resolve the issue by adding more models, is this just another example of the risks of an over-reliance on models at all?
-
-<!-- (3) THIS problem (4) THIS problem. I like that you outline the points you are going to cover! -->
 

### 2021-03-11 manuscript/appendix_B.Rmd
appendix_B and notes

@@
 d %>% pivot_longer(names(model_set[-1]), "model") %>%
-  ggplot(aes(state, value, col=as.numeric(model))) +
+  mutate(model = as.numeric(model)) %>%
+  ggplot(aes(state, value, col = model)) +
   geom_hline(aes(yintercept = 0), lwd=1) + 
@@
 
-am1 <- adaptive_management(transitions[-1], 
+am_many <- adaptive_management(transitions[-1], 
                            reward, 
@@
 ```{r}
-am <- am1$df %>% mutate(belief = am1$posterior[,1])
-am %>%  ggplot(aes(time, states[state], col = belief)) + 
+i <- 1:5
+posteriors <- data.frame(r = rs, K = Ks, t(am_many$posterior[i,])) %>%
+  pivot_longer(starts_with("X"), 
+               values_to = "probability",
+               names_to = "time") %>% 
+  mutate(time = as.integer(as.factor(time))) %>%
+  pivot_longer(c("r", "K"), names_to="parameter")
+
+posteriors %>%
+  ggplot(aes(value, probability, group=time, col=time)) +
+  geom_line(show.legend = FALSE) +
+  facet_grid(~parameter, scales="free")
+
+```
+
+```{r, fig.cap = ""}
+am_many$df %>% 
+  mutate(belief = am_many$posterior[,1]) %>%
+  ggplot(aes(time, states[state], col = belief)) + 
   geom_line() + 
   geom_point() +
-  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4])
+  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4]) + 
+  xlab("fish stock")
 ```
@@
 ```{r}
-
 am_npv <- sum(am$value * discount ^ am$time)

### 2021-03-11 manuscript/manuscript.Rmd
appendix_B and notes

@@
   - This is not about over-fitting.
+-->
 
--->
+<!-- quick summary of what was demonstrated -->
 Given this simple decision problem in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariably choose the other.
@@
 
+<!-- Intuition, Figure 4 -->
 The reason for Model 1's seemingly contradictory ability to make good decisions but bad forecasts becomes obvious once we compare both curves to that of the underlying model, Model 3.
@@
 
+<!-- Inttuition for it's not just this example -->
 This phenomenon is not unique to sequential decision problems or optimal control solutions.
@@
 
-Note that the mechanism shown here has nothing to do with the much more familiar issue of over-fitting, in which a better-fitting model will also lead to worse outcomes [].
+```{r figure4,  fig.cap="Panel A: Population growth curves of each model.
+The positive equilibrium of each model occurs where the curve crosses the horizontal axis.
+Note that while Model 2 is a better approximation to the truth (Model 3), Model 1 better approximates the stock size which leads to maximum growth.
+Panel B: The optimal control policy under Model 1 is nearly identical to that under the true Model 3, while the optimal policy under Model 2 supresses stock to a much lower escapement level.", fig.width=8}
+# Fig 4
+model_curves <- read_csv("../data/model_curves.csv") %>% mutate(model = as.character(model))
+policies <- read_csv("../data/policies.csv") %>% mutate(model = as.character(model))
+
+plot_models <- 
+  model_curves %>% 
+  ggplot(aes(state, value, col=model, lty=model)) +
+  geom_hline(aes(yintercept = 0), lwd=1) + 
+  geom_line(lwd=2, show.legend = FALSE) + 
+  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
+  ylab(bquote(f(x) - x)) + xlab("x")
+
+plot_policies <- policies %>%
+  ggplot(aes(states, escapement, col=model, lty=model)) + 
+  geom_line(lwd=2) + xlab("state")
+
+plot_models + labs(subtitle="A") +  plot_policies +  labs(subtitle="B")
+```
+
+<!-- Not just overfitting -->
+Note that the mechanism shown here has nothing to do with the much more familiar issue of over-fitting, in which a better-fitting model will also lead to worse outcomes [@overfit].
 In fact, an model which has been over-fit will gradually be rejected by either the iterative forecasting or adaptive management approaches shown here, as these approaches continually confront the models with new data to which they had not been previously fit.
@@
 
+<!-- Not just "get more models" -->
 This case cannot be dismissed merely as being dealt a bad hand in having to pick only between Model 1 and Model 2.
@@
 So what?
-
 - This is different than previous work: e.g. VOI is not zero, it's negative here.
@@
 Thus it should be no surprise that a model can drive good decisions while making poor predictions.
-It is also important to note that in many cases, any models being used in decision-making are not necessarily being subjected to the rigorous evaluation and updating steps proposed by adaptive management or iterative forecasting [@Dietze2018].
+It is also important to note that in many cases, any models being used in decision-making are not necessarily being subjected to the rigorous evaluation and updating steps proposed by adaptive management [@Walters1978] or iterative forecasting [@Dietze2018].
 This makes it more likely that such models could persist in practice until now.
@@
 I believe this is an open question, but that our first step must be to recognize it as such.
-(1) that the issue is just an example of the already-understood problem of underspecifying model uncertainty, which is resolved by including more or better models, or (2) this issue is just an example of the already-understood limitations of relying too much on models to shape decision-making.
-Either we need more models, or we need fewer models.
+We have seen that the problem cannot be resolved by more data, and is not the result of overfitting.
+Nor is "creating more models" the answer: when we have a model that is good enough to get optimal results, we cannot always insist on more models.
+We have also seen how not updating our uncertainty estimates, or doing so less frequently, can reveal a unicorn model before it is discounted by further iterations.
+Because value of information can be negative, we must revise model uncertainty only cautiously.
+More importantly, we have seen that decision theory, such as the intuition offered by Fig 4, can help us better understand what features of a model are essential to decision outcomes and what are not.
+Building on such understanding, it may possible to identify strategies for learning that are more agnostic to the details of the model that are not important.
 
-If to advocates of iterative model updating approaches the solution is always more models, then to skeptics the solution is fewer models and algorithms.  
-If we cannot always resolve the issue by adding more models, is this just another example of the risks of an over-reliance on models at all?
+<!-- Model free management -->
 
-```{r figure4,  fig.cap="Panel A: Population growth curves of each model.
-The positive equilibrium of each model occurs where the curve crosses the horizontal axis.
-Note that while Model 2 is a better approximation to the truth (Model 3), Model 1 better approximates the stock size which leads to maximum growth.", fig.width=8}
-# Fig 4
-model_curves <- read_csv("../data/model_curves.csv") %>% mutate(model = as.character(model))
-policies <- read_csv("../data/policies.csv") %>% mutate(model = as.character(model))
-
-plot_models <- 
-  model_curves %>% 
-  ggplot(aes(state, value, col=model, lty=model)) +
-  geom_hline(aes(yintercept = 0), lwd=1) + 
-  geom_line(lwd=2, show.legend = FALSE) + 
-  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
-  ylab(bquote(f(x) - x)) + xlab("x")
-
-plot_policies <- policies %>%
-  ggplot(aes(states, escapement, col=model, lty=model)) + 
-  geom_line(lwd=2) + xlab("state")
-
-plot_models + labs(subtitle="A") +  plot_policies +  labs(subtitle="B")
-```
-
-<!-- move to appendix?  too confusing/too long?
-
-Not sure - it feels a little extraneous (I did think this was an interesting point brought up by reviewer 1 but I also don't necessarily think it warrants a long paragraph that might confuse the main point rather than clarify something) Could this be a couple sentences and reference an appendix that walks through what you write out (potentially even with a simple constant mortality example)
--->
-Is this result an artifact of optimal control?
-The analysis has relied on optimal control theory to determine the optimal decision associated with each model or distribution of models.
-The resulting policies are mathematically optimal under the assumptions given [see @Reed1979 for stochastic harvest problem in particular and @Marescot2013 for discussion of Markov Decision Problems more generally].
-In most realistic decision problems, model complexity currently precludes the calculation of an optimal control solution at all ["the curse of dimensionality", e.g.
-@Polasky2011; @Marescot2013].
-Under the assumptions of any of the models considered here, the optimal control solution leads to what is known as a "constant escapement" policy, which seeks to bring biomass to the target escapement level, $B_{MSY}$ as quickly as possible, $H_t = \min(X_t - B_{MSY}, 0)$.
-If the assumptions are not met, a constant-escapement policy may not be optimal, as in the case when observations of the stock $X_t$ are not perfect [@pomdp-intro].
-While constant escapement is used as a basis of management in some real world cases such as salmon fisheries, most marine fisheries are based on a target of "constant mortality", ($F_{MSY}$).
-Notably, a constant mortality policy, implemented as $H_t = F_{MSY} X_t$, would not perform better under Model 1 than Model 2.
-Constant mortality is optimal only under deterministic growth at equilibrium (where the policy is the same under constant escapement).
-Does that mean the result is an artifact of assuming constant escapement rather than constant mortality as the basis for management?
-No, it does not.
-It is just as easy to construct a model under which the constant mortality matches the constant mortality of the true model while being a poor predictor.
-Moreover, using a non-optimal policy would always beg the question of whether the poor outcomes were merely the result of using a non-optimal policy rather than a consequence of model uncertainty.
-Because the space of possible actions is smaller than that of possible models, we should expect most decision problems to have their own "Model 1," a model that would lead to correct actions while being a poor predictor of future outcomes.
-The trouble is that we have no theory which would not reject such a model.
-
+<!--
 Is this result merely an artifact of differing objectives?
@@
 Note that this is the opposite recommendation of _adaptive_ management [@Walters1978] or _iterative_ forecasting [@Dietze2018].
-
-<!-- I really like the last paragraph!! -->
-
-Perhaps Model 1 is a unicorn, a mythically simple model that gives optimal results but has no analogue in the real world.
-This is not an easy assertion to prove either way.
-Future observations consistently fall outside the 95% confidence intervals predicted under Model 1.
-Unfortunately, such innaccurate predictions are only too common in the real world.
-Have any of those led to effective policy despite the innaccuracy?
-Have any been replaced with more accurate models that have also driven worse decisions?
-That is more difficult to answer.
-What this example demonstrates is that if such unicorn models exist, our current paradigm will surely filter them out eventually, to be replaced with worse outcomes.
-
-<!-- Omit this?  Does it beg an example?
-
-There is at least one technique which might address the problem presented here.
-Recently developed techniques based on model-free deep reinforcement learning train multi-layer neural networks to approximate the optimal control policy without assuming or attempting to model the underlying dynamic process at all.
-Feedback is based on cumulative reward, not short-term reward or predictive accuracy.
 -->
-
-<!--I was thinking about exactly this while I was reading the paper.
-I do think this would likely beg an example, which might be a can of worms that you don't want to open with this piece...
-I also don't think that I think it would address the presented problem.
-(Kind of repeating my second framing point but) It reminds me of the "Stop Explaining Black Box Machine Learning Models for High Stakes Decisions and use Interpretable Models Instead" paper where she talks about how, for a given data set, you can define the Rashomon set as the set of reasonably accurate predictive models-  Because the data are finite, the data could admit many close-to-optimal models that predict differently from each other: a large Rashomon set.
-I suspect this happens often in practice because sometimes many different ML algorithms perform similarly on the same data set, despite having different functional forms (for example, random forests, neural networks, support vector machines).
-As long as the Rashomon set contains a large enough set of models with diverse predictions, it probably contains functions that can be approximated well by simpler functions, and so the Rashomon set can also contain these simpler functions.
-Said another way, uncertainty arising from the data leads to a Rashomon set; a larger Rashomon set probably contains interpretable models, thus interpretable accurate models often exist.
-If this theory holds, we should expect to see interpretable models exist across domains.
-These interpretable models may be hard to find through optimization, but at least there is a reason we might expect that such models exist.
-
-The problem is deeper than acknowleding a "rashamon set of models" or a plurality of reality (relevant to that Bioscience paper I sent last week about the Rashamon effect in conservation).
-Sometimes what appears closer to reality is actually fundementally wrong in some way that is misguiding to a decision maker...more opaque approaches could lessen our capacity to trouble shoot that?) -->
 

### 2021-03-11 manuscript/manuscript.Rmd
:pen:

@@
 
-## Forecasts and Proper Scores
+## Statistical approaches: Forecasting under "Proper" Scoring Rules
 
-<!-- I think this subsection should be renamed.
-it makes it a bit hard to follow the structure of the larger section.
-Seems like something like: "Statistical approaches: forecasts and proper scores" idk, but something that more closely aligns with the above paragraph and the second subsection -->
+Like many decision problems, the task of setting a sustainable harvest quota appears to hinge on having an accurate forecast: if we can predict to what size the fish stock will increase next year, $X_t+1$, and we know the current stock, $X_t$, then we can sustainably harvest $X_{t+1} - X_t$ without decreasing the biomass over the long term.
+Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of over-fitting by comparing model predictions to later observations that were not used to estimate the model [@Gneiting2014].
 
-Like many decision problems, the task of setting a sustainable harvest quota appears to hinge on having an accurate forecast: if we can predict to what size the fish stock will increase next year, $X_t+1$, and we know the current stock, $X_t$, then we can safely harvest $X_{t+1} - X_t$.
-
-<!-- it is unclear what you mean by safely harvest?
-as in the stock wont crash?
-or it won't be worse than it is this year?
-You could imaging that we could safetly harvest more (or only less) than that.-->
-
-Overestimating or underestimating such recruitment will result in over-harvesting or under-harvesting, respectively.
-
-Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of over-fitting by comparing model predictions to later observations that were not used to estimate the model.
-
-@Gneiting2007 provides a rigorous proof for "proper" scoring criteria, which have the desirable property which no model predicting the distribution of future outcomes, $Q(x)$ can achieve a better average score than the true model $P(x)$.
-That is, unlike likelihood or other goodness-of-fit scores, it is impossible to overfit when conditioning on a strictly proper score -- since no model model can out-perform the true model.
+@Gneiting2007 provides a rigorous proof for such "proper" scoring criteria, which have the desirable property which no model predicting the distribution of future outcomes, $Q(x)$ can achieve a better average score than the true model $P(x)$.
+That is, unlike likelihood or other goodness-of-fit scores, it is impossible to overfit when conditioning on a strictly proper score -- no model model can out-perform the true model.
 Not that strictly proper scoring rules score _probabilistic forecasts_ and not just point predictions, favoring models which accurately reflect the uncertainty over those which under-estimate it.
@@
 If Model 1 has a value of $r$ such that it happens to match the best possible constant-mortality solution for the (unobserved) true model, then $K$ can be set arbitrarily high, ensuring the any model selection, forecasting, or adaptive management approach would lead away from Model 1 and towards worse-performing models once again.
+Many optimization problems share this feature in which the optimal policy depends only on a subset or ratio of model parameters, such that it is usually easy to find a similar Model 1.
+This example also illustrates the importance of decision constraints -- a model that gives very good outcomes when we are free to vary the havest quota each year (the optimal control problem here) may give very poor results under the constraint of constant mortality, and vice versa.
 
@@
 We have also seen how not updating our uncertainty estimates, or doing so less frequently, can reveal a unicorn model before it is discounted by further iterations.
-Because value of information can be negative, we must revise model uncertainty only cautiously.
+Approaches such as iterative forecasting or adaptive management that can reduce model uncertainty over time remain promising and important techniques, but because value of information can be negative, we must revise model uncertainty only cautiously.
 More importantly, we have seen that decision theory, such as the intuition offered by Fig 4, can help us better understand what features of a model are essential to decision outcomes and what are not.
-Building on such understanding, it may possible to identify strategies for learning that are more agnostic to the details of the model that are not important.
+Building on such understanding, it may possible to identify strategies for learning that are more agnostic to the details of the model that are not important, or perhaps not reliant on model-based predictions at all.
+How that is done is an open question.
 
 <!-- Model free management -->
+<!-- For example, a simple approach could gradually increase harvest from 0 as long as the observed recruitment increases relative to previous values.
+This corresponds to a local search for the peak growth-rate from above.
+This strategy is optimal for the model considered here, at least under small noise, but is not robust to growth rates with local optima, nor is it optimal during a possibly-very-long transient period of the search.
+-->
 

### 2021-03-12 manuscript/appendix_B.Rmd
new run cached

--- 2021-02-16 manuscript/appendix_B.Rmd
@@
 
-n_models <- 32
-rs <- seq(0.4, 2, length.out = n_models)
-Ks <- seq(8, 10*8/5, length.out = n_models)
+n_models <- 20
+rs <- seq(0.1, 2, length.out = n_models)
+Ks <- seq(1, 10*8/5, length.out = n_models)
 sigmas <- seq(0.05, 1.5*0.05, length.out = n_models)
@@
 
-  d %>% pivot_longer(names(model_set[-1]), "model") %>%
-  ggplot(aes(state, value, col=as.numeric(model))) +
+d  %>%
+  pivot_longer(names(model_set[-1]), "model") %>%
+  mutate(model = as.numeric(model)) %>%
+  ggplot(aes(state, value, col = model, group=model)) +
   geom_hline(aes(yintercept = 0), lwd=1) + 
-  geom_point(alpha = .5) + 
+  geom_line(alpha = .8) + 
   geom_line(aes(state, true), col = "red") +
@@
   ylab(bquote(f(x) - x)) + xlab("x") + viridis::scale_color_viridis()
-```
-
-We can easily plot the optimal policy derived from each model, as shown
-in Fig3b.
-
-```{r plot_policies}
-fig3b <- policies %>%
-  ggplot(aes(states, escapement, col=model, lty=model)) + 
-  geom_line(lwd=2) + xlab("state")
 ```
@@
                            x0 = x0, 
-                           Tmax = 50, 
+                           Tmax = 20, 
                            true_transition = transitions[[true_model]], 
@@
 ```{r}
-am <- am1$df %>% mutate(belief = am1$posterior[,1])
-am %>%  ggplot(aes(time, states[state], col = belief)) + 
-  geom_line() + 
-  geom_point() +
-  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4])
+i <- 1:20
+posteriors <- data.frame(r = rs, K = Ks, t(am_many$posterior[i,])) %>%
+  pivot_longer(starts_with("X"), 
+               values_to = "probability",
+               names_to = "time") %>% 
+  mutate(time = as.integer(as.factor(time))) %>%
+  pivot_longer(c("r", "K"), names_to="parameter")
+
+posteriors %>%
+  ggplot(aes(value, probability, group=time, col=time)) +
+  geom_line(show.legend = FALSE) +
+  facet_grid(~parameter, scales="free")
 ```
@@
 ```{r}
+best_model <- which.max(am_many$posterior[20,])
+```
 
-am_npv <- sum(am$value * discount ^ am$time)
+```{r, fig.cap = ""}
+am_many$df %>% 
+  mutate(belief = am_many$posterior[,best_model]) %>%
+  ggplot(aes(time, states[state], col = belief)) + 
+  geom_line() + 
+  geom_point() +
+  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4]) + 
+  xlab("fish stock")
+```
+
+```{r}
+am_npv <- sum(am_many$value * discount ^ am$time)
 am_npv / npv[[1,"npv"]]
@@
   pivot_longer(c("1", "2"), values_to="belief", names_to = "model") %>%
-  filter(time < 20) %>%
   ggplot(aes(time, belief, col=model)) + geom_line(col = pal[4])

### 2021-03-15 manuscript/manuscript.Rmd
:pen:

@@
 Global change issues are complex and outcomes are difficult to predict, making approaches to uncertainty a central part
-of effective decision-making [@ @]. Reducing that uncertainty by improving models in response to more data
-becomes available
+of effective decision-making [@ @]. Reducing that uncertainty through additional data collection and evaluation of outcomes
+is essential for effective management [@].
 
@@
 
-<!-- 
-It may be easy to dismiss this example on the argument that both models are (or more precisely,
-the any logistic growth model, regardless of parameterization would be) extremely naive
-and unrealistic, and any solution would surely begin with replacing both models with something
-more realistic.  Yet such objections overlook the simple fact that for all its faults in 
-prediction, model 1 already provides nearly optimal performance.  That performance arises
-because model 1 has very accurately reflected one key aspect of the true model (the position of the
-maximum growth rate), even while it is wrong about every other aspect of the biology.
+<!--Gonna be two examples.  Shorten me? -->
+A wide range of paradigms are available for approaching the issue of decision-making under uncertainty.
+These approaches can roughly be divided into two groups: statistical approaches and adaptive management approaches.
+The first group, "statistical approaches," seeks to maximize some statistical metric (such as goodness-of-fit or predictive accuracy) in determining which model(s) will drive decisions.
+In contrast, adaptive management approaches seek to select actions which maximize the decision-maker's expected net utility over
 
-The open problem is simply to propose an approach to model uncertainty which would select model 1
-over model 2 (i.e. choosing among the suite of available models, the one that gives the better 
-decisions.)
+Models are assessed by comparison to observation and the best model or models are used to inform management decisions.
+This process of model evaluation may be repeated iteratively after observing the outcome of the management decision compared to the predictions under the model.
+This approach is implicit to most statistical methods of model estimation and assessment [e.g.
+information criteria, @Burnham1998]
 
-An important aspect of this example is that model 1 performs nearly optimally.
+There are a wide range of techniques within each, and it is also possible to blend approaches.
+The key distinction is that methods in the first group do not involve any direct consideration of the possible actions or the utility that may result from those actions in how they select models [statistical approaches such as information criteria e.g.
+@Burnham1998]; and in particular, forecasting evaluation @Clark2001; @Dietze2018], while those in the second group require a more explicit statement of possible actions and the desired objectives [Decision theoretic approaches, for which @Polasky2011 provides an excellent and accessible review].
+I illustrate how the most promising techniques from each of these groups would be applied to this simple problem, and demonstrate that it both cases they lead us away from the model that produces the most desirable decisions towards worse outcomes.
+In retrospect, it will become obvious that neither these nor any other widely applied methods will select the model that leads to the best outcomes from the set of models considered.
 
--->
+<!-- Proper scores-->
+In the first group, I focus on evaluation using "proper" scoring rules as defined by @Gneiting2007, rather than potentially more familiar model comparisons such as $r^2$, likelihood, or information criteria [@Burnham1998] because proper scoring rules are more robust  (see @Gneiting2007 for comparison and how the approaches relate under appropriate assumptions).
+@Gneiting2007 provides a rigorous proof of scoring rules that have the desirable property of being "proper", that is, rules for which no model predicting the distribution of future outcomes, $Q(x)$ can achieve a better average score than the true model $P(x)$.
+Unlike likelihood or other goodness-of-fit scores alone, it is impossible (in expectation) to overfit the proper score -- no model model can out-perform the true model.
+Note that strictly proper scoring rules score _probabilistic forecasts_ and not just point predictions, favoring models which accurately reflect the uncertainty over those which under-estimate it.
+These features have made proper scoring rules for probabilistic forecasts a successful and popular approach for addressing model uncertainty in many other areas [@Gneiting2014; @Raftery2016] and a promising tool for evaluation of ecological forecasts [@Dietze2018].
+
+<!--Adaptive management-->
+For the second group, decision-theoretic approaches, I focus on the example of adaptive management
+Decision-theoretic approaches include scenario analysis, resilience thinking, optimal control and related methods [@Polasky2011], which have a long history in ecology and particularly in fisheries management.
+One of the most influential of these is _adaptive management_ [sensu @Walters1978], in which a manager seeks to both reduce uncertainty over time while also achieving the best outcomes given current knowledge.
+While the term today is frequently used in a looser sense, adaptive management as originally developed can be considered an example of optimal control under model uncertainty [@Walters1981; @Smith1981].
+In this approach, a manager assigns probabilities to each of the possible models under consideration, i.e.
+models 1 & 2.
+The expected utility of any action reflects both the intrinsic uncertainty over future states under either model (process uncertainty) and the manager's uncertainty over the choice of models (or parameter values).
+Unlike the forecast comparison, an adaptive management approach explicitly considers the utility function, Eqn \eqref{utility}, which it seeks to maximize by integrating over model uncertainty.
+After observing the consequences of the action, the manager updates the posterior probabilities that each model is correct, typically by Bayes rule [e.g.
+@Walters1981; @Smith1981; @Ludwig1982; @Punt2016].
+The decision strategy is then re-evaluated in light of this revised uncertainty.
+Adaptive management may be either passive or active.
+Under passive adaptive management, the decision-maker chooses the action which maximizes that expected future utility.
+Under active adaptive management, the decision-maker may chose 'exploratory' actions which are not optimal under the current uncertainty but may lead to more rapidly decreasing the uncertainty in the future [@Walters1978].
+In practice, many managers are reluctant to employ exploratory strategies, so passive adaptive management approaches such as as Management Strategy Evaluation (MSE) are more common [@Punt2016].
 
@@
 
-A wide range of paradigms are available for approaching the issue of decision-making under uncertainty.
-These approaches can roughly be divided into two groups: the first group treats the issue of model uncertainty independently from the decision itself, while the second integrates the process of reducing model uncertainty into the process of decision making to maximize the value of some objective.
-There are a wide range of techniques within each, and it is also possible to blend approaches.
-The key distinction is that methods in the first group do not involve any direct consideration of the possible actions or the utility that may result from those actions in how they select models [statistical approaches such as information criteria e.g.
-@Burnham1998]; and in particular, forecasting evaluation @Clark2001; @Dietze2018], while those in the second group require a more explicit statement of possible actions and the desired objectives [Decision theoretic approaches, for which @Polasky2011 provides an excellent and accessible review].
-I illustrate how the most promising techniques from each of these groups would be applied to this simple problem, and demonstrate that it both cases they lead us away from the model that produces the most desirable decisions towards worse outcomes.
-In retrospect, it will become obvious that neither these nor any other widely applied methods will select the model that leads to the best outcomes from the set of models considered.
-
 ## Statistical approaches: Forecasting under "Proper" Scoring Rules
@@
 Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of over-fitting by comparing model predictions to later observations that were not used to estimate the model [@Gneiting2014].
-
-@Gneiting2007 provides a rigorous proof for such "proper" scoring criteria, which have the desirable property which no model predicting the distribution of future outcomes, $Q(x)$ can achieve a better average score than the true model $P(x)$.
-That is, unlike likelihood or other goodness-of-fit scores, it is impossible to overfit when conditioning on a strictly proper score -- no model model can out-perform the true model.
-Not that strictly proper scoring rules score _probabilistic forecasts_ and not just point predictions, favoring models which accurately reflect the uncertainty over those which under-estimate it.
-These features have made proper scoring rules for probabilistic forecasts a successful and popular approach for addressing model uncertainty in many other areas [@Gneiting2014; @Raftery2016] and a promising tool for evaluation of ecological forecasts [@Dietze2018].
-
-<!-- This scoring bit comes abruptly (and it feels like could be in the introduction) -->
 
@@
 One obvious limitation in this comparison is that scenario B treats each model as fixed over the entire course of the simulation.
-Research has long emphasized the importance of learning and adaption in the face of new data whenever we are dealing with model uncertainty [@Polasky2011].
-A related limitation is that in Scenario B, decisions were based either on assuming Model 1 is correct or assuming Model 2 is correct.
-A more robust approach [e.g.
-@Walters1978] incorporates the uncertainty over models directly into the decision calculations by integrating over the probability each model $M$ (or parameter value) is correct when calculating the utility; $U(x_t, a_t) = \int U(x_t, a_t | M) P(M)$.
-The probability assigned to each model can then be updated after each subsequent action in light of the resulting outcomes [@Ludwig1982, @Smith1981].
-While a the wide range of strategies for such iterative updating are known [@Polasky2011]; all will fail to select the higher-performing of the two simple models considered  here.
+It would be possible to generate forecasts for the next $n$ steps, but these forecasts would also be conditional on the management action (i.e.
+harvest quota) selected at each step.
+To evaluate two-step-ahead predictions we must consider each possible action under each possible state predicted by the step-ahead forecast, weighted by probability of the state given the model and the probability of the model.
+This rapidly expanding set of possibilities is addressed by adaptive management for sequential decision problems [e.g.
+@Smith1981], which I employ in the next section.
 
@@
 ## Decision-Theoretic Approaches
-
-Decision-theoretic approaches include scenario analysis, resilience thinking, optimal control and related methods [@Polasky2011], which have a long history in ecology and particularly in fisheries management.
-One of the most influential of these is _adaptive management_ [sensu @Walters1978], in which a manager seeks to both reduce uncertainty over time while also achieving the best outcomes given current knowledge.
-While the term today is frequently used in a looser sense, adaptive management as originally developed can be considered an example of optimal control under model uncertainty [@Walters1981; @Smith1981].
-In this approach, a manager assigns probabilities to each of the possible models under consideration, i.e.
-models 1 & 2.
-The expected utility of any action reflects both the intrinsic uncertainty over future states under either model (process uncertainty) and the manager's uncertainty over the choice of models (or parameter values).
-Unlike the forecast comparison, an adaptive management approach explicitly considers the utility function, Eqn \eqref{utility}, which it seeks to maximize by integrating over model uncertainty.
-After observing the consequences of the action, the manager updates the posterior probabilities that each model is correct, typically by Bayes rule [e.g.
-@Walters1981; @Smith1981; @Ludwig1982; @Punt2016].
-The decision strategy is then re-evaluated in light of this revised uncertainty.
-Adaptive management may be either passive or active.
-Under passive adaptive management, the decision-maker chooses the action which maximizes that expected future utility.
-Under active adaptive management, the decision-maker may chose 'exploratory' actions which are not optimal under the current uncertainty but may lead to more rapidly decreasing the uncertainty in the future [@Walters1978].
-In practice, many managers are reluctant to employ exploratory strategies, so passive adaptive management approaches such as as Management Strategy Evaluation (MSE) are more common [@Punt2016].
 
@@
 
-Unfortunately, any adaptive management strategy which updates posterior distributions that characterize model undertainty leads to worse outcomes than an approach which does no updating.
+Any adaptive management strategy updates posterior distributions over model uncertainty [@Punt2016; @Ludwig1982].
+Unfortunately, any such adaptive updating leads to worse outcomes than the equivalent non-adaptive strategy, in which model uncertainty is held fixed.
 I illustrate the application of a passive adaptive management strategy to this simple example, following classic examples for parameter [@Ludwig1982] or structural [@Smith1981] model uncertainty.
+Passive adaptive management for a simple sequential decision problem is straight forward to implement over a discrete set of states and actions using dynamic programming with iterative updates [@Smith1981, example code in Appendix].
 To demonstrate that the behavior is not driven by failure to explore sufficiently, (which might be addressed by an active adaptive management), I will assign initial probability that model 2 is true at 1%.
-After a single iteration of learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct  [Fig 4].
+After a single iteration of learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct  [Fig 3].
 As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating [Fig 2, achieving a net present value over 100 time steps that only 31% that expected under management using Model 1 alone].
-If instead of considering only these two models we consider a whole suite of Gordon-Schaefer-type models with varying $r$ and $K$ parameters, learning is slightly slower but no less counter-productive: mean stock size is $46%$ and net present value is $20%$ of what would be achieved under Model 1 (Appendix, Fig S1).
 
@@
 
+Increasing the space of possible models to cover a whole plausible range of parameters $r$ and $K$ does nothing to resolve the problem [Fig 4].
+
 # Discussion
@@
 In both approaches, the manager quickly concludes that Model 1 is entirely implausible, while finding that Model 2 is remarkably accurate at predicting future values.
-This problem is not addressed by re-estimating parameters (equivalently, considering a larger suite of models with all possible $K$ and $r$ values): in fact leads to a model with slightly lower $K$ and even higher level of over-fishing and under-performance than Model 2 (Appendix B, Fig S1).
+This problem is not addressed by re-estimating parameters (equivalently, considering a larger suite of models with all possible $K$ and $r$ values): in fact leads to a model with slightly lower $K$ and even higher level of over-fishing and under-performance than Model 2 (Fig 4).
 
-<!-- Intuition, Figure 4 -->
+<!-- Intuition, Figure 5 -->
 The reason for Model 1's seemingly contradictory ability to make good decisions but bad forecasts becomes obvious once we compare both curves to that of the underlying model, Model 3.
-Plotting the growth rate functions of each model, [Fig 4a], it is hardly surprising that no method exists which would not prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3.
-However, Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 4b], while Model 2 leads to over-harvesting.
+Plotting the growth rate functions of each model, [Fig 5A], it is hardly surprising that no method exists which would not prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3.
+However, Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 5B], while Model 2 leads to over-harvesting.
 The explanation comes from noticing that the stock size corresponding to the maximum growth rate under Model 1 (the peak of the curve) falls at almost exactly the same stock size as that of the peak growth rate for Model 3.
@@
 
-<!-- Inttuition for it's not just this example -->
+<!-- Intuition for it's not just this example -->
 This phenomenon is not unique to sequential decision problems or optimal control solutions.
@@
 
-```{r figure4,  fig.cap="Panel A: Population growth curves of each model.
+```{r figure5,  fig.cap="Panel A: Population growth curves of each model.
 The positive equilibrium of each model occurs where the curve crosses the horizontal axis.
@@
 Panel B: The optimal control policy under Model 1 is nearly identical to that under the true Model 3, while the optimal policy under Model 2 supresses stock to a much lower escapement level.", fig.width=8}
-# Fig 4
+# Fig 5
 model_curves <- read_csv("../data/model_curves.csv") %>% mutate(model = as.character(model))
@@
 The ability to avoid over-fitting is one of the greatest appeals of any iterative management strategy.
-But in the example here, both Model 1 and Model 2 have the same structural complexity.
+In the example here, both Model 1 and Model 2 have the same structural complexity.
 The better decision performance of Model 1 is a consequence of capturing the key attribute needed for a good decision, which is perhaps a surprisingly an all-together different criteria than fit.
+Note that other methods of model selection not considered here, including goodness-of-fit metrics such as $r^2$  or information criteria [@Burnham], will all prefer Model 2 over Model 1 for the same reason.
 
@@
 Had we included the true model in the set of possibilities, the techniques illustrated would have had little difficulty in distinguishing it from Models 1 and 2 after sufficient iterations.
-In practice, we never have the true model, and even a much more thorough suite of candidate models, such as considering any possible values of $r$, $K$ (see Appendix B, Fig S1), or even among common alternative models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980], would fail to do any better, since all these models still have symmetric growth rates with a peak growth rate at half the steady-state population size).
+In practice, we never have the true model, and even a much more thorough suite of candidate models, such as considering any possible values of $r$, $K$ [Fig 4], or even among common alternative models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980], would fail to do any better, since all these models still have symmetric growth rates with a peak growth rate at half the steady-state population size).
 Models will always be simpler than reality.
@@
 The premise that model need not perfectly capture reality to be useful is at the very heart of modeling.
-Model 1 successfully captures the one key feature driving decisions in optimal harvest control problems: the stock size at which the maximum growth rate occurs [Fig 4a].
+Model 1 successfully captures the one key feature driving decisions in optimal harvest control problems: the stock size at which the maximum growth rate occurs [Fig 5A].
 Capturing only the essential aspects as simply as possible is the goal of any model building exercise.
@@
 Approaches such as iterative forecasting or adaptive management that can reduce model uncertainty over time remain promising and important techniques, but because value of information can be negative, we must revise model uncertainty only cautiously.
-More importantly, we have seen that decision theory, such as the intuition offered by Fig 4, can help us better understand what features of a model are essential to decision outcomes and what are not.
+More importantly, we have seen that  the intuition offered by decision theory [Fig 5] can help us better understand what features of a model are essential to decision outcomes and what are not.
+Conversely, models which appear to lead to accurate predictions (like Model 2) can result in outcomes that are far from optimal.
 Building on such understanding, it may possible to identify strategies for learning that are more agnostic to the details of the model that are not important, or perhaps not reliant on model-based predictions at all.
@@
 
-<!-- Model free management -->
+<!-- Model free management: what would it look like? -->
 <!-- For example, a simple approach could gradually increase harvest from 0 as long as the observed recruitment increases relative to previous values.
@@
 -->
+
+<!-- Correlary: Neither accurate forecasting nor effective outcomes are sufficient grounds to assert model validity. (meh, that's just 'all models are wrong, some are useful' for certain things)   -->
+
+## Acknowledgements
+
+The author acknowledges support from NSF CAREER Award #1942280 and helpful discussions with Melissa Chapman and Jeremy Fox.
 
@@
 
+<!--
+
+The open problem is simply to propose an approach to model uncertainty which would select model 1
+over model 2 (i.e. choosing among the suite of available models, the one that gives the better 
+decisions.)
+
+-->
+
 \pagebreak

### 2021-03-18 manuscript/manuscript.Rmd
:pen: drafting intro

@@
 <!-- Open on real-world issues, uncertainty, particularly model uncertainty as key theme.  -->
+Global change issues are complex and outcomes are difficult to predict  [@Clark2001]. 
+To guide decisions in an uncertainty world, researchers and decision makers 
+may consider a range of alternative plausible models[^1] to better reflect 
+what we do and do not know about the processes involved [@Polasky2011].  Forecasts or
+predictions from possible models can indicate what outcomes are most likely to result
+under what decisions or actions.  Further, by comparing the outcomes predicted by a model
+to future observations, a decision maker can not only _plan_ for the uncertainty, but 
+also _learn_ which models are most trustworthy.  This iterative learning process is
+particularly appealing because the models which best fit currently available data are not 
+always the models which lead to the best decisions.  Such outcomes are often characterized
+as the result of "over-fitting" [@Burnham2002]. While many methods (e.g. information criteria @Burnham2002)
+are developed seeking to avoid this
 
-Global change issues are complex and outcomes are difficult to predict, making approaches to uncertainty a central part
-of effective decision-making [@ @]. Reducing that uncertainty through additional data collection and evaluation of outcomes
-is essential for effective management [@].
+While models
+may also reflect uncertainties intrinsic to the process (stochasticity) or the limits 
+of measurement (observation uncertainty),
+
+Why then do some models which most closely match available data lead to worse outcomes than models which appear less accurate?
+This situation is almost universally considered to be the result of "over-fitting" [@], which much effort is dedicated to avoiding 
+in selecting models. Because the defining flaw of an over-fit model is the failure to correctly predict additional data,
+
+<!-- even in a world absent of power politics, conflicting incentives, imperfect implementation, this is not an issue that can be addressed by statistics and data alone.  The
+
+-->
+
+making approaches to uncertainty a central part
+of effective decision-making [@Polasky2011; @Clark2001]. Reducing that uncertainty through additional data collection 
+and evaluation of outcomes is thus essential for effective management [@walters1978].
 
@@
 [^1]: While some authors distinguish between model uncertainty and parameter uncertainty, I will refer collectively to
-any uncertainty that arises from our imperfect knowledge of the system as "model uncertainty".
-Uncertainty in model structure can often be reflected by appropriate parameterization, just as models with different parameter values can be treated distinctly.
-Whereas stochasticity is inherent to the process (or rather, to our choice of state variables, @Boettiger2018) and measurement uncertainty can only be reduced by more accurate tools for measurement, model uncertainty is reducible by learning from additional data.
+any uncertainty that arises from our imperfect knowledge of the system as "model uncertainty," because the distinction
+is largely a consequence of notation rather something more intrinsic.  For example, two structurally different models, $f(x) = a x$
+and $f(x) = a x^2$ can be considered merely different parameterizations of $f(x) = a x^b$, or vice versa.
 

### 2021-03-18 manuscript/appendix_B.Rmd

--- 2021-02-16 manuscript/appendix_B.Rmd
@@
 
-knitr::opts_chunk$set(cache=FALSE, tidy = "styler", message = FALSE, warning = FALSE, echo = TRUE)
+knitr::opts_chunk$set(cache=FALSE, tidy = "styler", message = FALSE, warning = FALSE, echo = FALSE)
 
@@
 }
+```
 
-n_models <- 32
-rs <- seq(0.4, 2, length.out = n_models)
-Ks <- seq(8, 10*8/5, length.out = n_models)
-sigmas <- seq(0.05, 1.5*0.05, length.out = n_models)
-pars <- data.frame(r = rs, K = Ks, sigma = sigmas)
+```{r}
+rs <- c(0.01, 0.1, 0.2, 0.3, 0.4, 0.5, 2)
+Ks <- c(0.5, 1, 1.5, 4, 10, 16)
+n_models <- length(rs)*length(Ks)
+sigmas <- 0.1
+pars <- expand.grid(rs, Ks)
+names(pars) <- c("r", "K")
+pars$sigma <- sigmas
 
@@
 
-This follows the standard setup for standard stochastic dynamic programming, see @Marescot2013.
-Having defined a function to compute the transition matrix, we can use it to create matrices corresponding to each of the three models:
-
 ```{r}
@@
 ```
-
-Likewise, a corresponding matrix defining the rewards associated with each state $X$ and each harvest action $H$ can also be defined.
 
@@
 
-```{r plot_models}
+```{r plot_models, fig.width=6, fig.height=4}
 model_set <- models
-names(model_set) <- c("true", as.character(round(Ks, 2)))
+names(model_set) <- c("true", 1:n_models)
 d <- 
@@
   ylab(bquote(f(x) - x)) + xlab("x") + viridis::scale_color_viridis()
-```
 
-We can easily plot the optimal policy derived from each model, as shown
-in Fig3b.
-
-```{r plot_policies}
-fig3b <- policies %>%
-  ggplot(aes(states, escapement, col=model, lty=model)) + 
-  geom_line(lwd=2) + xlab("state")
 ```
@@
 x0 <- which.min(abs(states - 6))
+## start with 90% weight on 'model 1'
+model_prior <- c(rep(.1, length(models)-2) / (length(models) - 2), .9)
 
-am1 <- adaptive_management(transitions[-1], 
-                           reward, 
-                           discount, 
+am_many <- adaptive_management(transitions[-1],  
+                           reward = reward, 
+                           discount = discount,
+                           model_prior = model_prior,
                            x0 = x0, 
-                           Tmax = 50, 
+                           Tmax = 20, 
                            true_transition = transitions[[true_model]], 
@@
 ```{r}
-am <- am1$df %>% mutate(belief = am1$posterior[,1])
-am %>%  ggplot(aes(time, states[state], col = belief)) + 
-  geom_line() + 
-  geom_point() +
-  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4])
+posteriors <- data.frame(r = rs, K = Ks, t(am_many$posterior)) %>%
+  pivot_longer(starts_with("X"), 
+               values_to = "probability",
+               names_to = "time") %>% 
+  mutate(time = as.integer(as.factor(time)))
+
+i <- 10:20
+r_marginals <- posteriors %>% group_by(time,r) %>% 
+  summarise(prob = sum(probability), .groups = "drop") %>% 
+  filter(time %in% i)
+
+r_marginals %>%  ggplot(aes(r, prob, group=time)) + geom_line() + facet_wrap(~time)
+```
+
+## Marginal Posterior Probability evolution over K
+
+```{r}
+i = 19
+K_marginals <- posteriors %>% group_by(time,K) %>% 
+  summarise(prob = sum(probability), .groups = "drop") %>% 
+  filter(time %in% i) 
+K_marginals %>%
+  ggplot(aes(K, prob, group=time)) + geom_path() + geom_point() + facet_wrap(~time)
+
 ```
@@
 ```{r}
-
-am_npv <- sum(am$value * discount ^ am$time)
-am_npv / npv[[1,"npv"]]
+Tmax <- max(am_many$df$time)
+best_i <- which.max(am_many$posterior[Tmax,])
+good_i <- am_many$posterior[Tmax,] > 0.1
+pars[good_i,] %>% mutate(final_prob = as.numeric(am_many$posterior[Tmax,good_i])) %>% knitr::kable("latex")
 ```
 
-```{r include = FALSE}
-as_tibble(am1$posterior) %>% 
-  mutate(time = seq_along(V1)) %>%
-  select(time = time, "1" = V1, "2" = V2) %>% 
-  pivot_longer(c("1", "2"), values_to="belief", names_to = "model") %>%
-  filter(time < 20) %>%
-  ggplot(aes(time, belief, col=model)) + geom_line(col = pal[4])
+```{r, fig.cap = ""}
+model1 <- dim(am_many$posterior)[2]
+am_many$df %>% 
+  mutate(belief = am_many$posterior[,model1]) %>%
+  ggplot(aes(time, states[state], col = belief)) + 
+  geom_line() + 
+  geom_point() +
+  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) + 
+  ylab("fish stock") + labs(caption = "Belief in model 1 falls off rapidly as magagement pushes stocks to lower and lower sizes.")
+
 ```
+
+```{r}
+am_npv <- sum(am_many$df$value * discount ^ am_many$df$time)
+am_npv
+```

### 2021-03-19 manuscript/manuscript.Rmd
intro :pen:

@@
 predictions from possible models can indicate what outcomes are most likely to result
-under what decisions or actions.  Further, by comparing the outcomes predicted by a model
-to future observations, a decision maker can not only _plan_ for the uncertainty, but 
-also _learn_ which models are most trustworthy.  This iterative learning process is
-particularly appealing because the models which best fit currently available data are not 
-always the models which lead to the best decisions.  Such outcomes are often characterized
-as the result of "over-fitting" [@Burnham2002]. While many methods (e.g. information criteria @Burnham2002)
-are developed seeking to avoid this
+under what decisions or actions. This has made model-based forecasts a cornerstone for scientifically
+based decision making.  But, do better models always lead to better decisions?
 
-While models
-may also reflect uncertainties intrinsic to the process (stochasticity) or the limits 
-of measurement (observation uncertainty),
+Whenever models are estimated from the data they seek to describe, there is a concern that
+the model may over-fit the data: making the data seem even more likely than it would be
+under the true process [@Burnham2002].  Over-fitting is a central concern of any model comparison
+because over-fit models may underestimate uncertainty and make poor predictions about future
+outcomes [@Ginzburg2004]. By comparing outcomes predicted by a model to future observations, 
+a decision maker can not only _plan_ for the uncertainty, but also _learn_ which models are most trustworthy.  
+This iterative learning process of learning will not only decrease uncertainty by narrowing 
+beliefs over plausible model or parameter values, but can also eliminate models which overfit the data.
+Because any iterative learning compares models to future outcomes, overfitting those outcomes
+is impossible: a concept formalized rigorously by @Gneiting2007's proof of "proper" scoring rules.
+By definition, under a proper scoring rule, no probabilistic prediction $Q(x)$ can score better, on
+average, than that of the underlying process, $P(x)$.  @Gneiting2007 further shows which
+techniques for evaluating forecast predictions do an do not satisfy this condition, i.e. which
+cannot be over-fit.  The ability to decrease uncertainty while avoiding over-fitting makes
+such iterative learning particularly appealing approach to decision making in an uncertain
+world [@Deitze2018], and the approach is already well established in the principle of 
+adaptive management [@Walters1978] and existing practices such as Management Strategy Evaluation (MSE)
+[@Punt2006] used in fisheries. Given this ability to decrease uncertainty and avoid over-fitting,
+such iterative updating approaches seem irreproachable from a theoretical basis, even if
+they may be difficult to execute in practice.
 
-Why then do some models which most closely match available data lead to worse outcomes than models which appear less accurate?
-This situation is almost universally considered to be the result of "over-fitting" [@], which much effort is dedicated to avoiding 
-in selecting models. Because the defining flaw of an over-fit model is the failure to correctly predict additional data,
-
-<!-- even in a world absent of power politics, conflicting incentives, imperfect implementation, this is not an issue that can be addressed by statistics and data alone.  The
-
--->
-
-making approaches to uncertainty a central part
-of effective decision-making [@Polasky2011; @Clark2001]. Reducing that uncertainty through additional data collection 
-and evaluation of outcomes is thus essential for effective management [@walters1978].
-
-While some uncertainty is intrinsic to the underlying processes (stochasticity) or by
-limits to what variables we can observe and how accurately we measure them (measurement uncertainty), much uncertainty 
-is the result of our imperfect knowledge of the processes involved, as expressed by the structure and
-parameters of mathematical models used to approximate those processes (model uncertainty). Unlike the other forms of
-uncertainty, model uncertainty can be 
-reduced by gathering additional data, which may rule out certain models or parameter values as implausible[^1]. 
-When model parameters are estimated directly from available data, there is a risk that the best-fitting models may 
-_overfit_ to patterns arising from by chance from the stochasticity or measurement error which do not reflect the
-underlying process [@Ginzburg2004]. Researchers have often favored simpler models which are less prone to over-fitting, sometimes
-through explicit penalties such as information criteria [@], though these can be misleading [@]. With the rapid 
-expansion of available ecological and environmental data [@], it is increasingly possible to rely instead on comparisons
-between model predictions and data purposely excluded from model estimation (cross-validation), or better, 
-comparisons between model predictions and data collected in the future (forecasting, @Clark2001; @Dietze2018]). 
-Over sufficiently long timescales, comparing model forecasts to future observations will select the models with
-greatest predictive accuracy.  Because model-based forecasts frequently play an important role in decision-making,
-it is commonly assumed [e.g. @Walters1981; @Clark2001; @Dietze2018] that this approach for addressing model uncertainty
-will also improve decision-making outcomes. Here, I illustrate that this need not be the case: it is quiet possible
-that any approach with improves the forecast accuracy of the model(s) over time can simultaneously lead to worse 
-decision outcomes. This example illustrates that knowing when increased forecast accuracy will or will not also improve 
-decision outcomes in general remains an open problem, as is the challenge of what to do about it.
+However, as long as the true process model is unknown (effectively always), there is no
+such theoretical guarantee that iterative learning over possible models or parameters will
+lead to better outcomes.  This is not -- cannot -- be because some models have over-fit the 
+data, as least as long as proper scoring rules are used for evaluation.  As I illustrate here,
+given a set of candidate models, iterative learning either through assessment of forecast
+skill or through formal adaptive management can lead to worse outcomes than forgoing any learning step.
+Put another way, the value of information [@Howard1966], as measured by the expected utility given that 
+information minus the utility without it, can be negative: a rational decision maker ought to pay
+not to know information even when that information is perfectly accurate.  
+This situation is a direct consequence of the fact that a model need only capture certain
+features of the underlying process to lead to an effective decision.
 
@@
 
-<!--Gonna be two examples.  Shorten me? -->
+<!-- Haven't really queued up the idea that I will present two distinct approaches that both fail.  need to more explicitly introduce and justify the two approaches-->
+
+<!--
+
 A wide range of paradigms are available for approaching the issue of decision-making under uncertainty.
@@
 
-Models are assessed by comparison to observation and the best model or models are used to inform management decisions.
-This process of model evaluation may be repeated iteratively after observing the outcome of the management decision compared to the predictions under the model.
-This approach is implicit to most statistical methods of model estimation and assessment [e.g.
-information criteria, @Burnham1998]
-
-There are a wide range of techniques within each, and it is also possible to blend approaches.
-The key distinction is that methods in the first group do not involve any direct consideration of the possible actions or the utility that may result from those actions in how they select models [statistical approaches such as information criteria e.g.
-@Burnham1998]; and in particular, forecasting evaluation @Clark2001; @Dietze2018], while those in the second group require a more explicit statement of possible actions and the desired objectives [Decision theoretic approaches, for which @Polasky2011 provides an excellent and accessible review].
-I illustrate how the most promising techniques from each of these groups would be applied to this simple problem, and demonstrate that it both cases they lead us away from the model that produces the most desirable decisions towards worse outcomes.
-In retrospect, it will become obvious that neither these nor any other widely applied methods will select the model that leads to the best outcomes from the set of models considered.
-
-<!-- Proper scores-->
 In the first group, I focus on evaluation using "proper" scoring rules as defined by @Gneiting2007, rather than potentially more familiar model comparisons such as $r^2$, likelihood, or information criteria [@Burnham1998] because proper scoring rules are more robust  (see @Gneiting2007 for comparison and how the approaches relate under appropriate assumptions).
@@
 
-<!--Adaptive management-->
 For the second group, decision-theoretic approaches, I focus on the example of adaptive management
@@
 In practice, many managers are reluctant to employ exploratory strategies, so passive adaptive management approaches such as as Management Strategy Evaluation (MSE) are more common [@Punt2016].
+
+-->
 

### 2021-03-20 manuscript/appendix_B.Rmd
:pen:

@@
 ```{r}
+# Try 6 log-spaced values
+rs <- 10 ^ seq(0.01, 2, length=6)
+Ks <- 10 ^ seq(-.3, 1.2, length=6)
+
+# compare over fixed set of candidate r and K
 rs <- c(0.01, 0.1, 0.2, 0.3, 0.4, 0.5, 2)
@@
 ```
-
-On a discrete grid of possible states and actions, we can define the growth rate of a given state $X_t$ subject to harvest $H_t$,
-$f(X_t,H_t)$ as set of matrices.  Each matrix $i$ gives the transition probabilities for any current state to any future state, 
-given that action $i$ is taken.
 
@@
 ```{r}
+Tmax <- max(am_many$df$time)
+best_i <- which.max(am_many$posterior[Tmax,])
+good_i <- am_many$posterior[Tmax,] > 0.1
+pars[good_i,] %>% mutate(final_prob = as.numeric(am_many$posterior[Tmax,good_i])) 
+```
+
+```{r}
+model1 <- dim(am_many$posterior)[2]
+
+am_multi <- am_many$df %>% 
+  mutate(belief = am_many$posterior[,model1],
+         stock = states[state],
+         quota = actions[action]) %>%
+  select(time, belief, stock, quota) 
+write_csv(am_multi, "../data/am_multi.csv")
+```
+
+```{r}
+am_multi %>%
+  ggplot(aes(time, stock, col = belief)) + 
+  geom_line() + 
+  geom_point() +
+  geom_point(aes(time, quota), col="darkgreen", shape=2) + 
+  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) + 
+  ylab("fish stock") + labs(caption = "Belief in model 1 falls off rapidly as magagement pushes stocks to lower and lower sizes.")
+
+```
+
+```{r}
 posteriors <- data.frame(r = rs, K = Ks, t(am_many$posterior)) %>%
@@
 
-i <- 10:20
+i <- 20
 r_marginals <- posteriors %>% group_by(time,r) %>% 
@@
 
-r_marginals %>%  ggplot(aes(r, prob, group=time)) + geom_line() + facet_wrap(~time)
+r_marginals %>%  ggplot(aes(r, prob, group=time)) + geom_bar(stat="identity") + facet_wrap(~time)
 ```
@@
 ```{r}
-i = 19
+i = 20
 K_marginals <- posteriors %>% group_by(time,K) %>% 
@@
 K_marginals %>%
-  ggplot(aes(K, prob, group=time)) + geom_path() + geom_point() + facet_wrap(~time)
+  ggplot(aes(K, prob, group=time)) + geom_bar(stat="identity") + facet_wrap(~time) + scale_x_log10()
 
@@
 ```{r}
-Tmax <- max(am_many$df$time)
-best_i <- which.max(am_many$posterior[Tmax,])
-good_i <- am_many$posterior[Tmax,] > 0.1
-pars[good_i,] %>% mutate(final_prob = as.numeric(am_many$posterior[Tmax,good_i])) %>% knitr::kable("latex")
-```
+policy <- mdp_compute_policy(transitions[-1],  
+                           reward = reward, 
+                           discount = discount,
+                           model_prior = model_prior,
+                           Tmax = 20)
+non_adaptive <- memoise::memoise(mdp_planning, cache = memoise::cache_filesystem("cache/"))
+x0 <- which.min(abs(states - 6))
+## start with 90% weight on 'model 1'
+model_prior <- c(rep(.1, length(models)-2) / (length(models) - 2), .9)
 
-```{r, fig.cap = ""}
-model1 <- dim(am_many$posterior)[2]
-am_many$df %>% 
-  mutate(belief = am_many$posterior[,model1]) %>%
-  ggplot(aes(time, states[state], col = belief)) + 
-  geom_line() + 
-  geom_point() +
-  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) + 
-  ylab("fish stock") + labs(caption = "Belief in model 1 falls off rapidly as magagement pushes stocks to lower and lower sizes.")
-
+non_am <- non_adaptive(transitions[-1],  
+                           reward = reward, 
+                           discount = discount,
+                           model_prior = model_prior,
+                           x0 = x0, 
+                           policy = policy,
+                           Tmax = 20, 
+                           true_transition = transitions[[true_model]], 
+                           epsilon = 0.001, 
+                           max_iter = 2000)
 ```

### 2021-03-20 manuscript/manuscript.Rmd
:pen:

@@
 lineno: true
-
+csl: ecology-letters.csl
 output: rticles::elsevier_article
 layout: 3p
+journal: TBD
 ---
@@
 the model may over-fit the data: making the data seem even more likely than it would be
-under the true process [@Burnham2002].  Over-fitting is a central concern of any model comparison
+under the true process [@Burnham1998].  Over-fitting is a central concern of any model comparison
 because over-fit models may underestimate uncertainty and make poor predictions about future
@@
 such iterative learning particularly appealing approach to decision making in an uncertain
-world [@Deitze2018], and the approach is already well established in the principle of 
+world [@Dietze2018], and the approach is already well established in the principle of 
 adaptive management [@Walters1978] and existing practices such as Management Strategy Evaluation (MSE)
-[@Punt2006] used in fisheries. Given this ability to decrease uncertainty and avoid over-fitting,
+[@Punt2016] used in fisheries. Given this ability to decrease uncertainty and avoid over-fitting,
 such iterative updating approaches seem irreproachable from a theoretical basis, even if
@@
 skill or through formal adaptive management can lead to worse outcomes than forgoing any learning step.
-Put another way, the value of information [@Howard1966], as measured by the expected utility given that 
+Put another way, the value of information [VOI; @Howard1966; @Katz1987], as measured by the expected utility given that 
 information minus the utility without it, can be negative: a rational decision maker ought to pay
@@
 
-```{r figure3, fig.cap="Adaptive management under model uncertainty, color indicates the belief that Model 2 is correct (red).
+```{r figure3, fig.cap="Adaptive management under model uncertainty, color indicates the belief that Model 1 is correct (blue).
 Each model is assigned an initial prior belief.
@@
 am <- read_csv("../data/am.csv")
-am %>%  filter(time <30) %>%
-  ggplot(aes(time, states[state], col = belief)) + 
+am %>%  filter(time <=20) %>%
+  ggplot(aes(time, stock, col = belief)) + 
   geom_line() + 
   geom_point() +
-  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4]) +
+  geom_point(aes(time, quota), col="darkgreen", shape=2) + 
+  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) +
   ylab("stock size")
@@
 
-Increasing the space of possible models to cover a whole plausible range of parameters $r$ and $K$ does nothing to resolve the problem [Fig 4].
+Increasing the space of possible models to cover a whole plausible range of parameters $r$ and $K$ does nothing to resolve this problem.
+Learning is somewhat slower over the larger range of possibilities, but nevertheless converges towards parameter values with lower growth rates and lower carrying capacity, for which the target biomass is far lower [Fig 4].
+The manager becomes trapped in vicious cycle as biomass declines further and further, subsequent observations seem more and more consistent with models which most favor over-fishing in the first place.
+As biomass declines, annual harvests and profits also plunge, and yet our manager becomes only more confident in their decisions.
+
+```{r}
+am_multi <- read_csv("../data/am_multi.csv")
+am_multi %>%
+  ggplot(aes(time, stock, col = belief)) + 
+  geom_line() + 
+  geom_point() +
+  geom_point(aes(time, harvest), col="darkgreen", shape=2) + 
+  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) + 
+  ylab("fish stock") + labs(caption = "Belief in model 1 falls off rapidly as magagement pushes stocks to lower and lower sizes.")
+```
 
@@
 <!-- Not just overfitting -->
-Note that the mechanism shown here has nothing to do with the much more familiar issue of over-fitting, in which a better-fitting model will also lead to worse outcomes [@overfit].
+Note that the mechanism shown here has nothing to do with the much more familiar issue of over-fitting, in which a better-fitting model will also lead to worse outcomes [@Burnham1998].
 In fact, an model which has been over-fit will gradually be rejected by either the iterative forecasting or adaptive management approaches shown here, as these approaches continually confront the models with new data to which they had not been previously fit.
@@
 The better decision performance of Model 1 is a consequence of capturing the key attribute needed for a good decision, which is perhaps a surprisingly an all-together different criteria than fit.
-Note that other methods of model selection not considered here, including goodness-of-fit metrics such as $r^2$  or information criteria [@Burnham], will all prefer Model 2 over Model 1 for the same reason.
+Note that other methods of model selection not considered here, including goodness-of-fit metrics such as $r^2$  or information criteria [@Burnham1998], will all prefer Model 2 over Model 1 for the same reason.
 
@@
 Model 1 approximates the key feature, giving nearly optimal outcomes.
-Previous literature has underscored the importance of Knightian "unknown unknowns" two alternative models examined here fail to fully reflect our uncertainty in the underlying dynamics, and the consequences of underestimating model uncertainty are well understood [].
+Previous literature has underscored the importance of Knightian "unknown unknowns" two alternative models examined here fail to fully reflect our uncertainty in the underlying dynamics, and the consequences of underestimating model uncertainty are well understood [@Polasky2011; @Wintle2010].
 Had we included the true model in the set of possibilities, the techniques illustrated would have had little difficulty in distinguishing it from Models 1 and 2 after sufficient iterations.

### 2021-03-26 manuscript/appendix.Rmd
:pen:

@@
 ---
-title: 'Appendix for: "Bad Forecast, Good Decision"'
+title: 'Appendix A'
 date: "`r Sys.Date()`"
@@
 This appendix provides a fully reproducible walk-through of the implementation of the analyses considered in this paper.
-None of the methods applied here: Stochastic Dynamic Programming, Iterative Forecasts assessed with Proper Scoring Rules, or Adaptive Management, are novel
+Because the methods applied here: Stochastic Dynamic Programming, Iterative Forecasts assessed with Proper Scoring Rules, or Adaptive Management, are not themselves novel, the reader is referred to previous reviews and text books which can provide a thorough introduction to methods, in particular, @Marescot2013 and @Smith1981.
+However, implementation of these approaches generally depends on numerical methods, for which even precise mathematical formulae or pseudo-code descriptions can be inadequate to specify unambiguously or to properly facilitate replication [@Barnes2010].
+As such, this appendix includes computer code in the R language, which can also be found in the corresponding R-Markdown document in the paper's GitHub repository, <https://github.com/cboettig/bad-forecast-good-decision>.
+While computer code can be more verbose and difficult to read than mathematical formulae presented here, it is also less ambiguous.
+Though R is already widely used among ecologists, the implementations concern mostly matrix algebra that can be translated to other languages, and the `MDPtoolbox` package used here is also implemented in Matlab [@Marescot2013; @MDPtoolbox].
 
-This appendix includes computer code in the R language, which can also be found in the corresponding R-Markdown document in the paper's GitHub repository, <https://github.com/cboettig/bad-forecast-good-decision>.
-While computer code can be more verbose and difficult to read than mathematical formulae presented here, it is also less ambiguous.
+## Optimal quotas as a Markov Decision Process (MDP) problem
 
-The question of optimal harvest given a population growth model can be posed as a Markov Decision Process (MDP).
+The question of optimal harvest quotas given a population growth model can be posed as a Markov Decision Process (MDP).
 @Marescot2013 provides an accessible overview of MDP problems and their solutions in conservation context.
@@
 As @Reed1979 proves, the optimal strategy can be characterized by a policy of "constant escapement," in which the manager adjusts the harvest effort each year in an effort to ensure the same number or biomass of fish are left in the sea each year.
-
-The optimization problem can then be solved by dynamic programming, using the recursive Bellman equation as,
-
-\begin{equation}
-\end{equation}
-
-This technique, known as stochastic dynamic programming (SDP) is widely used in conservation decision problems [e.g. @Mangel1985, @Marescot2013].
+The optimization problem can then be solved by dynamic programming, using the recursive Bellman equation.
+This technique, known as stochastic dynamic programming (SDP) is widely used in conservation decision problems [e.g.
+@Mangel1985, @Marescot2013].
 
@@
 Whether we use SDP or some other way to determine the optimal policy, given the model, is immaterial to the conclusion.
+
+## Solving MDP via SDP
 
@@
 Passive adaptive management using a Bayesian learning scheme still learns the wrong model.
+Adaptive management for MDP problems is particularly computationally intensive, and not directly supported by the `MDPtoolbox` implementation, though the principle involves only a relatively straightforward, if not particularly efficient extension to the task of solving and MDP problem for a single model shown above.
+Given a set of transition probability matrices $P_i$ and a belief probability $b_i$ that the $i$th model is correct, it simple to define the transition matrices over possible models by the laws of conditional probability, $P = \sum_i P_i b_i$, given $\sum_i b_i = 1$.
+After each subsequent action and then new observation x_j, the $b_i$ probabilities for each model are updated according to Bayes law: if $b_i = P(M_i)$ is the prior probability that the model $M_i$ is correct, then the posterior probability that model $M_i$ is correct given observation $x_j$ is:
+
+$$b_i' = P(M_i | x_j) = P(M_i) P(x_j | M_i) = b_i P(x_j | M_i) $$
+
+Note that because the transition matrices are now altered after every action and observation, it is necessary to repeat the complete SDP calculation to determine the optimal action given the updated uncertainty.
+While straight forward, this is what makes adaptive management for MDP problems particularly computationally intensive and thus poorly suited for handling a large space of candidate models.
+Nevertheless, our example of two models is most feasible, and even the larger example of 36 models (appendix B) is not unreasonable.
+Here I make use of the `mdplearning` R package, which provides a straight-forward implementation of this algorithm.
 
@@
 ```{r}
-am <- am1$df %>% mutate(belief = am1$posterior$V2)
+am <- am1$df %>% mutate(belief = am1$posterior$V1,
+                        stock = states[state],
+                        quota = actions[action])
 write_csv(am, "../data/am.csv")

### 2021-03-26 manuscript/appendix_B.Rmd
:pen:

@@
 ---
-title: 'Appendix for: "Bad Forecast, Good Decision"'
+title: 'Appendix B'
 date: "`r Sys.Date()`"

### 2021-03-26 manuscript/manuscript.Rmd
:pen:

@@
 ---
-title: 'Are we learning to build better models at the cost of better decisions?'
+title: 'When to ignore perfect information: how iterative learning can lead to worse decisions'
 author:
@@
 To guide decisions in an uncertainty world, researchers and decision makers 
-may consider a range of alternative plausible models[^1] to better reflect 
+may consider a range of alternative plausible models to better reflect 
 what we do and do not know about the processes involved [@Polasky2011].  Forecasts or
 predictions from possible models can indicate what outcomes are most likely to result
-under what decisions or actions. This has made model-based forecasts a cornerstone for scientifically
-based decision making.  But, do better models always lead to better decisions?
+under what decisions or actions. This has made model-based forecasts a cornerstone 
+for scientifically based decision making.  
+By comparing outcomes predicted by a model to future observations, 
+a decision maker can not only _plan_ for the uncertainty, but also _learn_ which models are most trustworthy.  The value of iterative learning has long been reflected in the theory of 
+adaptive management [@Walters1978] as well as in actual adaptive management practices such as 
+Management Strategy Evaluation (MSE) [@Punt2016] used in fisheries, and is a central tenant
+of a rapidly growing interest in ecological forecasting [@Dietze2018].
+But, do iterative learning approaches always lead to better decisions?
 
-Whenever models are estimated from the data they seek to describe, there is a concern that
-the model may over-fit the data: making the data seem even more likely than it would be
-under the true process [@Burnham1998].  Over-fitting is a central concern of any model comparison
-because over-fit models may underestimate uncertainty and make poor predictions about future
-outcomes [@Ginzburg2004]. By comparing outcomes predicted by a model to future observations, 
-a decision maker can not only _plan_ for the uncertainty, but also _learn_ which models are most trustworthy.  
-This iterative learning process of learning will not only decrease uncertainty by narrowing 
-beliefs over plausible model or parameter values, but can also eliminate models which overfit the data.
-Because any iterative learning compares models to future outcomes, overfitting those outcomes
-is impossible: a concept formalized rigorously by @Gneiting2007's proof of "proper" scoring rules.
-By definition, under a proper scoring rule, no probabilistic prediction $Q(x)$ can score better, on
+In this paper, I will demonstrate that iterative learning or iterative forecasting can also be counter-productive to management outcomes:
+that in some cases, a manager would be better to ignore new information which could reduce
+uncertainty over available models. Put another way, the value of information (VOI, as
+measured by the expected utility given that information minus the utility without it; see @Howard1966; @Katz1987), can actually be negative. This is not a consequence of imperfect information
+(e.g. observation error) or the result of model over-fitting.  The issue is an intrisic 
+consequence of the fact that many models can lead to the same decisions, and no generic
+solution to the problem exists. However, as this example shows, by better understanding
+what aspects of model are most and least important to shaping a particular decision, 
+we can at least anticipate the circumstances that make this problem more likely.
+
+Iterative learning approaches such as adaptive management or iterative forecasting are
+particularly compelling because they address a different reason why seemingly-accurate
+models can lead to poor outcomes: the problem of overfitting.  When models are estimated
+from the data they seek to describe, the model may over-fit the data, making the data
+seem even more likely than it would be under the true process [@Burnham1998].
+Over-fit models may underestimate uncertainty and make poor predictions about future
+outcomes [@Ginzburg2004]. Because any iterative learning compares models to future outcomes,
+overfitting those outcomes is impossible.  This concept was rigorously formalized
+by @Gneiting2007's proof of "proper" scoring rules. By definition, under a proper 
+scoring rule, no probabilistic prediction $Q(x)$ can score better, on
 average, than that of the underlying process, $P(x)$.  @Gneiting2007 further shows which
 techniques for evaluating forecast predictions do an do not satisfy this condition, i.e. which
-cannot be over-fit.  The ability to decrease uncertainty while avoiding over-fitting makes
-such iterative learning particularly appealing approach to decision making in an uncertain
-world [@Dietze2018], and the approach is already well established in the principle of 
-adaptive management [@Walters1978] and existing practices such as Management Strategy Evaluation (MSE)
-[@Punt2016] used in fisheries. Given this ability to decrease uncertainty and avoid over-fitting,
-such iterative updating approaches seem irreproachable from a theoretical basis, even if
-they may be difficult to execute in practice.
-
-However, as long as the true process model is unknown (effectively always), there is no
-such theoretical guarantee that iterative learning over possible models or parameters will
-lead to better outcomes.  This is not -- cannot -- be because some models have over-fit the 
-data, as least as long as proper scoring rules are used for evaluation.  As I illustrate here,
-given a set of candidate models, iterative learning either through assessment of forecast
-skill or through formal adaptive management can lead to worse outcomes than forgoing any learning step.
-Put another way, the value of information [VOI; @Howard1966; @Katz1987], as measured by the expected utility given that 
-information minus the utility without it, can be negative: a rational decision maker ought to pay
-not to know information even when that information is perfectly accurate.  
-This situation is a direct consequence of the fact that a model need only capture certain
-features of the underlying process to lead to an effective decision.
-
-[^1]: While some authors distinguish between model uncertainty and parameter uncertainty, I will refer collectively to
-any uncertainty that arises from our imperfect knowledge of the system as "model uncertainty," because the distinction
-is largely a consequence of notation rather something more intrinsic.  For example, two structurally different models, $f(x) = a x$
-and $f(x) = a x^2$ can be considered merely different parameterizations of $f(x) = a x^b$, or vice versa.
+cannot be over-fit. Adaptive management is based on equally secure footing because
+model predictions are again compared to future outcomes not used to fit the model.
+Instead of scoring and selecting among alternative models, adaptive management 
+considers probabilities over all models [@Walters1978].  These two approaches are
+particularly compelling because they can both reduce uncertainty and avoid over-fitting.
+My example is not meant to undercut the importance of these approaches.  Previous
+work has long acknowledged the panelope of ways in which iterative learning or other
+model-based decision making can go astray due to conflicting incentives,
+implementation errors, or lack of resources for monitoring and updating [e.g. @Ludwig1993].
+Here, the problem is more fundamental.
 
@@
 
+For illustrative purposes, we will focus on the simplest example of model uncertainty,
+considering two alternative models of equal complexity but differing in the estimated
+value of certain parameters.
+While some authors distinguish between model uncertainty and parameter uncertainty, I will refer collectively to any uncertainty that arises from our imperfect knowledge of the system as "model uncertainty," because the distinction is largely a consequence of notation rather something more intrinsic.
+(For example, two structurally different models, $f(x) = a x$ and $f(x) = a x^2$ can be considered merely different parameterizations of $f(x) = a x^b$, or vice versa.)
+
 # Ecological Models
@@
 
-A manager seeks each year to select the harvest $H_t$ which will maximize the sum
+A manager seeks each year to select the harvest quota $H_t$ which will maximize the sum
 of the utility derived from such a harvest and such a state, $U(X_t,H_t)$, over all time,
@@
 
-Further we imagine that the function $f$ is not known precisely, and so we will rely on an evaluation of forecasting skill across a set of candidate models to determine which one to use to manage the fishery.
-Again for simplicity, we will restrict ourselves to two simple candidate models $f_1$ and $f_2$.
+For simplicity and comparison with prior work, we will assume a fixed price of fish $p$ with no additional cost on additional harvest, $U(X_t, H_t) = p \min(H_t, X_t)$ (noting that realized harvest cannot exceed the stock size).
+Without loss of generality we will set the price $p = 1$ and modest discount $\delta = 0.99$.
+
+We further imagine that the function $f$ is not known precisely.
+Again for simplicity, we restrict ourselves to two simple candidate models $f_1$ and $f_2$.
 Both share the same underlying structure of logistic recruitment (known as the Gordon-Schaefer model in fisheries context owing to groundbreaking work independently by @Schaefer1954 and @Gordon1954), differing only in their choice of certain parameters:
@@
 
-Where $\xi_t(\sigma)$ represents log-normal random noise with mean of unity and log-standard-deviation $\sigma$.
+For simplicity, we will assume $\xi_t(\sigma)$ represents log-normal random noise with mean of unity and log-standard-deviation $\sigma_i$ for each model.
 Model 1 is given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, Model 2 by $r_2 = 0.5$,  $K_2 = 10$, $\sigma = 0.075$ (in dimensionless units).
@@
 
-<!-- I feel like parts of the below paragrph should be in the abstract (i.e.
-"I illustrate how the most promising techniques from statistical approaches and decision theoretic approaches to decision making under uncertainty would be applied to this simple problem, and demonstrate that it both cases they lead us away from the model that produces the most desirable decisions towards worse outcome") -->
+We will use this example to illustrate two alternative approaches for iterative learning over model uncertainty: iterative forecasting and adaptive management.
+The central difference in the approaches
+is that iterative forecasting is premised on the ability to score the predictions of alternative models.
+Iterative forecasting is silent on the issue of what to do with those scores, this is left up to the 
+decision-maker.  Adaptive management approaches, by contrast, explicitly seek to integrate probabilities over all candidate models to reach a decision.  I consider each in turn.
 
@@
 In Scenario A (passive observation) the fish stock is unharvested and allowed to recover towards carrying capacity (as simulated under our "true" model, Model 3) while comparing the observed stock size in each subsequent time step to the distribution predicted under Model 1 and Model 2 respectively [Fig 1].
-The mean, $\mu$ and variance, $\sigma$ of the forecast are compared against the true observation $x$ using a Proper scoring rule of @Gneiting2007,
+The mean, $\mu_t$ and variance, $\sigma_t$ of the forecast are compared against the true observation $x_t$ using a Proper scoring rule of @Gneiting2007,
 
 \begin{equation}
--(\mu - x )^2 / \sigma^2  - \log(\sigma) \label{proper}
+S(x_t|\mu_t,\sigma_t) = -(\mu_t - x_t )^2 / \sigma_t^2  - \log(\sigma_t) \label{proper}
 \end{equation}
@@
 After a single iteration of learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct  [Fig 3].
-As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating [Fig 2, achieving a net present value over 100 time steps that only 31% that expected under management using Model 1 alone].
+As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating, which achieves a net present value over 100 time steps that only 31% that expected under management using Model 1 alone [Fig 2].
 
-```{r figure3, fig.cap="Adaptive management under model uncertainty, color indicates the belief that Model 1 is correct (blue).
-Each model is assigned an initial prior belief.
+```{r figure3, fig.cap="Adaptive management under model uncertainty, color indicating the belief that Model 1 is correct (blue), lines indicating stock biomass over time, green triangles indicating harvest quotas set by adaptive management.
+Panel A: Each model is assigned an initial prior belief.
 To ensure that any issues with passive exploration are not the cause of preferring Model 2, the initial belief in Model 2 is set to 1%.
-Within a single iteration of adaptive management, the belief over models is updated to near certainty in Model 2, resulting in higher harvests and lower stock sizes similar to managing under Model 2 alone, Fig 2."}
+Within a single iteration of adaptive management, the belief over models is updated to near certainty in Model 2, resulting in higher harvests and lower stock sizes similar to managing under Model 2 alone, Fig 2A.
+Panel B: given 36 candidate models over a broad range of parameters, adaptive management leads to models that overharvest even more, leading to even lower stock sizes and lower harvest quotas."}
 am <- read_csv("../data/am.csv")
-am %>%  filter(time <=20) %>%
+fig3a <- am %>%  filter(time <=20) %>%
+  ggplot(aes(time, stock, col = belief)) + 
+  geom_line(show.legend = FALSE) + 
+  geom_point(show.legend = FALSE) +
+  geom_point(aes(time, quota), col="darkgreen", shape=2) + 
+  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) +
+  ylab("stock size") + labs(subtitle="A. Two models") + ylim(c(0,11))
+
+am_multi <- read_csv("../data/am_multi.csv")
+fig3b <- am_multi %>%
   ggplot(aes(time, stock, col = belief)) + 
@@
   scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) + 
-  ylab("stock size")
+  ylab("fish stock") + labs(subtitle="B. 36 models")+ ylim(c(0,11))
+
+fig3a + fig3b
 ```
@@
 Increasing the space of possible models to cover a whole plausible range of parameters $r$ and $K$ does nothing to resolve this problem.
-Learning is somewhat slower over the larger range of possibilities, but nevertheless converges towards parameter values with lower growth rates and lower carrying capacity, for which the target biomass is far lower [Fig 4].
+Learning is somewhat slower over the larger range of possibilities, but nevertheless converges towards parameter values with lower growth rates and lower carrying capacity, for which the target biomass is far lower [Fig 3B; details in appendix B].
 The manager becomes trapped in vicious cycle as biomass declines further and further, subsequent observations seem more and more consistent with models which most favor over-fishing in the first place.
 As biomass declines, annual harvests and profits also plunge, and yet our manager becomes only more confident in their decisions.
-
-```{r}
-am_multi <- read_csv("../data/am_multi.csv")
-am_multi %>%
-  ggplot(aes(time, stock, col = belief)) + 
-  geom_line() + 
-  geom_point() +
-  geom_point(aes(time, harvest), col="darkgreen", shape=2) + 
-  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) + 
-  ylab("fish stock") + labs(caption = "Belief in model 1 falls off rapidly as magagement pushes stocks to lower and lower sizes.")
-```
 
@@
 In both approaches, the manager quickly concludes that Model 1 is entirely implausible, while finding that Model 2 is remarkably accurate at predicting future values.
-This problem is not addressed by re-estimating parameters (equivalently, considering a larger suite of models with all possible $K$ and $r$ values): in fact leads to a model with slightly lower $K$ and even higher level of over-fishing and under-performance than Model 2 (Fig 4).
+This problem is not addressed by re-estimating parameters (equivalently, considering a larger suite of models with all possible $K$ and $r$ values): in fact leads to a model with slightly lower $K$ and even higher level of over-fishing and under-performance than Model 2 (Fig 3B).
 
-<!-- Intuition, Figure 5 -->
+<!-- Intuition, Figure 4 -->
 The reason for Model 1's seemingly contradictory ability to make good decisions but bad forecasts becomes obvious once we compare both curves to that of the underlying model, Model 3.
-Plotting the growth rate functions of each model, [Fig 5A], it is hardly surprising that no method exists which would not prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3.
-However, Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 5B], while Model 2 leads to over-harvesting.
+Plotting the growth rate functions of each model, [Fig 4A], it is hardly surprising that no method exists which would not prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3.
+However, Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 4B], while Model 2 leads to over-harvesting.
 The explanation comes from noticing that the stock size corresponding to the maximum growth rate under Model 1 (the peak of the curve) falls at almost exactly the same stock size as that of the peak growth rate for Model 3.
@@
 Had we included the true model in the set of possibilities, the techniques illustrated would have had little difficulty in distinguishing it from Models 1 and 2 after sufficient iterations.
-In practice, we never have the true model, and even a much more thorough suite of candidate models, such as considering any possible values of $r$, $K$ [Fig 4], or even among common alternative models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980], would fail to do any better, since all these models still have symmetric growth rates with a peak growth rate at half the steady-state population size).
+In practice, we never have the true model, and even a much more thorough suite of candidate models, such as considering any possible values of $r$, $K$ [Fig 3B], or even among common alternative models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980], would fail to do any better, since all these models still have symmetric growth rates with a peak growth rate at half the steady-state population size).
 Models will always be simpler than reality.

### 2021-03-26 manuscript/manuscript.Rmd
:pen:

@@
 
-```{r figure5,  fig.cap="Panel A: Population growth curves of each model.
+```{r figure4,  fig.cap="Panel A: Population growth curves of each model.
 The positive equilibrium of each model occurs where the curve crosses the horizontal axis.
@@
 Panel B: The optimal control policy under Model 1 is nearly identical to that under the true Model 3, while the optimal policy under Model 2 supresses stock to a much lower escapement level.", fig.width=8}
-# Fig 5
+# Fig 4
 model_curves <- read_csv("../data/model_curves.csv") %>% mutate(model = as.character(model))
@@
 But that is a difficult assertion to prove.
-Examples of serious forecasts that widely fail to predict future observations abound wherever forecasts are common, from elections to economics to environmental change [@].
+Examples of serious forecasts that widely fail to predict future observations abound wherever forecasts are common, from elections to economics to environmental change [e.g.
+@Tetlock2015].
 If there is no shortage of bad forecasts, then could any of them be useful?
 The premise that model need not perfectly capture reality to be useful is at the very heart of modeling.
-Model 1 successfully captures the one key feature driving decisions in optimal harvest control problems: the stock size at which the maximum growth rate occurs [Fig 5A].
+Model 1 successfully captures the one key feature driving decisions in optimal harvest control problems: the stock size at which the maximum growth rate occurs [Fig 4A].
 Capturing only the essential aspects as simply as possible is the goal of any model building exercise.
@@
 Approaches such as iterative forecasting or adaptive management that can reduce model uncertainty over time remain promising and important techniques, but because value of information can be negative, we must revise model uncertainty only cautiously.
-More importantly, we have seen that  the intuition offered by decision theory [Fig 5] can help us better understand what features of a model are essential to decision outcomes and what are not.
+More importantly, we have seen that  the intuition offered by decision theory [Fig 4] can help us better understand what features of a model are essential to decision outcomes and what are not.
 Conversely, models which appear to lead to accurate predictions (like Model 2) can result in outcomes that are far from optimal.

### 2021-03-30 manuscript/appendix_B.Rmd
tidy up
tweak

--- 2021-03-18 manuscript/appendix_B.Rmd
@@
 
-knitr::opts_chunk$set(cache=FALSE, tidy = "styler", message = FALSE, warning = FALSE, echo = FALSE)
+knitr::opts_chunk$set(cache=FALSE, tidy = "styler", message = FALSE, warning = FALSE, echo = TRUE)
 
@@
 }
-```
 
-```{r}
-rs <- c(0.01, 0.1, 0.2, 0.3, 0.4, 0.5, 2)
-Ks <- c(0.5, 1, 1.5, 4, 10, 16)
-n_models <- length(rs)*length(Ks)
-sigmas <- 0.1
-pars <- expand.grid(rs, Ks)
-names(pars) <- c("r", "K")
-pars$sigma <- sigmas
+n_models <- 30
+rs <- seq(0.05, 2, length.out = n_models)
+Ks <- seq(0.1, 10*8/5, length.out = n_models)
+sigmas <- seq(0.05, 1.5*0.05, length.out = n_models)
+pars <- data.frame(r = rs, K = Ks, sigma = sigmas)
 
@@
 
+This follows the standard setup for standard stochastic dynamic programming, see @Marescot2013.
+Having defined a function to compute the transition matrix, we can use it to create matrices corresponding to each of the three models:
+
 ```{r}
@@
 ```
+
+Likewise, a corresponding matrix defining the rewards associated with each state $X$ and each harvest action $H$ can also be defined.
 
@@
 
-```{r plot_models, fig.width=6, fig.height=4}
+```{r plot_models}
 model_set <- models
-names(model_set) <- c("true", 1:n_models)
+names(model_set) <- c("true", as.character(round(Ks, 2)))
 d <- 
@@
 
-d %>% pivot_longer(names(model_set[-1]), "model") %>%
+d  %>%
+  pivot_longer(names(model_set[-1]), "model") %>%
   mutate(model = as.numeric(model)) %>%
-  ggplot(aes(state, value, col = model)) +
+  ggplot(aes(state, value, col = model, group=model)) +
   geom_hline(aes(yintercept = 0), lwd=1) + 
-  geom_point(alpha = .5) + 
+  geom_line(alpha = .8) + 
   geom_line(aes(state, true), col = "red") +
@@
   ylab(bquote(f(x) - x)) + xlab("x") + viridis::scale_color_viridis()
-
 ```
@@
 x0 <- which.min(abs(states - 6))
-## start with 90% weight on 'model 1'
-model_prior <- c(rep(.1, length(models)-2) / (length(models) - 2), .9)
 
 am_many <- adaptive_management(transitions[-1], 
-                           reward = reward, 
-                           discount = discount,
-                           model_prior = model_prior,
+                           reward, 
+                           discount, 
                            x0 = x0, 
-                           Tmax = 20, 
+                           Tmax = 25, 
                            true_transition = transitions[[true_model]], 
@@
 ```{r}
-posteriors <- data.frame(r = rs, K = Ks, t(am_many$posterior)) %>%
+i <- 1:20
+posteriors <- data.frame(r = rs, K = Ks, t(am_many$posterior[i,])) %>%
   pivot_longer(starts_with("X"), 
@@
                names_to = "time") %>% 
-  mutate(time = as.integer(as.factor(time)))
+  mutate(time = as.integer(as.factor(time))) %>%
+  pivot_longer(c("r", "K"), names_to="parameter")
 
-i <- 10:20
-r_marginals <- posteriors %>% group_by(time,r) %>% 
-  summarise(prob = sum(probability), .groups = "drop") %>% 
-  filter(time %in% i)
-
-r_marginals %>%  ggplot(aes(r, prob, group=time)) + geom_line() + facet_wrap(~time)
-```
-
-## Marginal Posterior Probability evolution over K
-
-```{r}
-i = 19
-K_marginals <- posteriors %>% group_by(time,K) %>% 
-  summarise(prob = sum(probability), .groups = "drop") %>% 
-  filter(time %in% i) 
-K_marginals %>%
-  ggplot(aes(K, prob, group=time)) + geom_path() + geom_point() + facet_wrap(~time)
-
+posteriors %>%
+  ggplot(aes(value, probability, group=time, col=time)) +
+  geom_line(show.legend = FALSE) +
+  facet_grid(~parameter, scales="free")
 ```
@@
 ```{r}
-Tmax <- max(am_many$df$time)
-best_i <- which.max(am_many$posterior[Tmax,])
-good_i <- am_many$posterior[Tmax,] > 0.1
-pars[good_i,] %>% mutate(final_prob = as.numeric(am_many$posterior[Tmax,good_i])) %>% knitr::kable("latex")
+best_model <- which.max(am_many$posterior[20,])
 ```
@@
 ```{r, fig.cap = ""}
-model1 <- dim(am_many$posterior)[2]
 am_many$df %>% 
-  mutate(belief = am_many$posterior[,model1]) %>%
+  mutate(belief = am_many$posterior[,best_model]) %>%
   ggplot(aes(time, states[state], col = belief)) + 
@@
   geom_point() +
-  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) + 
-  ylab("fish stock") + labs(caption = "Belief in model 1 falls off rapidly as magagement pushes stocks to lower and lower sizes.")
-
+  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4]) + 
+  xlab("fish stock")
 ```
@@
 ```{r}
-am_npv <- sum(am_many$df$value * discount ^ am_many$df$time)
-am_npv
+am_npv <- sum(am_many$value * discount ^ am$time)
+am_npv / npv[[1,"npv"]]
 ```
+
+```{r include = FALSE}
+as_tibble(am1$posterior) %>% 
+  mutate(time = seq_along(V1)) %>%
+  select(time = time, "1" = V1, "2" = V2) %>% 
+  pivot_longer(c("1", "2"), values_to="belief", names_to = "model") %>%
+  ggplot(aes(time, belief, col=model)) + geom_line(col = pal[4])
+```

### 2021-03-31 manuscript/appendix_B.Rmd
re-run

--- 2021-03-26 manuscript/appendix_B.Rmd
@@
 ```{r}
-# K is at twice max of f3; 8 * K_3 / 5
-f1 <- function(x, h = 0, r = 2, K = 10 * 8 / 5){
-  s <- pmax(x - h, 0)
-  s + s * (r * (1 - s / K) )
-}
-f2 <- function(x, h = 0, r = 0.5, K = 10){
-  s <- pmax(x - h, 0)
-  s + s * (r * (1 - s / K) )
-}
-
 # max is at 4 * K / 5 
@@
 sigma_g <- 0.05
-models <- c(f3, models)
-model_sigmas <- c(sigma_g, pars$sigma)
-true_model <- 1
+model_sigmas <- pars$sigma
 ```
@@
 
+true_transition <- transition_matrices(f3, states, actions, sigma_g)
 ```
@@
 ```{r plot_models, fig.width=6, fig.height=4}
-model_set <- models
-names(model_set) <- c("true", 1:n_models)
+model_set <-c(f3,  models)
+names(model_set) <- c("true", 1:length(models))
 d <- 
@@
 
+# Non-adaptive management over model uncertainty
+
+With strong prior probabilities weighted to Model 1
+
+```{r}
+x0 <- which.min(abs(states - 6))
+## start with 99% weight on 'model 1'
+model_prior <- c(rep(.01, length(models)-1) / (length(models) - 1), .99)
+
+```
+
+```{r}
+policy_sdp <- memoise::memoise(mdp_compute_policy, cache = memoise::cache_filesystem("cache/"))
+
+policy <- policy_sdp(transitions,  
+                     reward = reward, 
+                     discount = discount,
+                     model_prior = model_prior,
+                     Tmax = 20, 
+                     epsilon = 0.001, 
+                     max_iter = 2000)
+
+policy %>% mutate(state = states[state], quota = actions[policy]) %>% mutate(escapement = state - quota) %>% ggplot(aes(state, escapement)) + geom_line()
+```
+
+```{r}
+
+non_am <- mdp_planning(true_transition,  
+                       reward = reward, 
+                       discount = discount,
+                       model_prior = model_prior,
+                       x0 = x0, 
+                       a0 = 1,
+                       policy = policy$policy,
+                       Tmax = 20)
+
+non_am %>% mutate(state = states[state]) %>% ggplot(aes(time, state)) + geom_point() + geom_line()
+
+```
+
 # Adaptive Management
 
-Passive adaptive management using a Bayesian learning scheme still learns the wrong model.
+Passive adaptive management, with initial strong prior probabilities weighted to Model 1.
 
@@
 adaptive_management <- memoise::memoise(mdp_learning, cache = memoise::cache_filesystem("cache/"))
-x0 <- which.min(abs(states - 6))
-## start with 90% weight on 'model 1'
-model_prior <- c(rep(.1, length(models)-2) / (length(models) - 2), .9)
 
-am_many <- adaptive_management(transitions[-1],  
+am_many <- adaptive_management(transitions,  
                            reward = reward, 
@@
                            Tmax = 20, 
-                           true_transition = transitions[[true_model]], 
+                           true_transition = true_transition, 
                            epsilon = 0.001, 
@@
 ```{r}
-policy <- mdp_compute_policy(transitions[-1],  
-                           reward = reward, 
-                           discount = discount,
-                           model_prior = model_prior,
-                           Tmax = 20)
-non_adaptive <- memoise::memoise(mdp_planning, cache = memoise::cache_filesystem("cache/"))
-x0 <- which.min(abs(states - 6))
-## start with 90% weight on 'model 1'
-model_prior <- c(rep(.1, length(models)-2) / (length(models) - 2), .9)
-
-non_am <- non_adaptive(transitions[-1],  
-                           reward = reward, 
-                           discount = discount,
-                           model_prior = model_prior,
-                           x0 = x0, 
-                           policy = policy,
-                           Tmax = 20, 
-                           true_transition = transitions[[true_model]], 
-                           epsilon = 0.001, 
-                           max_iter = 2000)
-```
-
-```{r}
 am_npv <- sum(am_many$df$value * discount ^ am_many$df$time)

### 2021-03-31 manuscript/appendix.Rmd
consolidate appendix

@@
 ---
-
-This appendix provides a fully reproducible walk-through of the implementation of the analyses considered in this paper.
-Because the methods applied here: Stochastic Dynamic Programming, Iterative Forecasts assessed with Proper Scoring Rules, or Adaptive Management, are not themselves novel, the reader is referred to previous reviews and text books which can provide a thorough introduction to methods, in particular, @Marescot2013 and @Smith1981.
-However, implementation of these approaches generally depends on numerical methods, for which even precise mathematical formulae or pseudo-code descriptions can be inadequate to specify unambiguously or to properly facilitate replication [@Barnes2010].
-As such, this appendix includes computer code in the R language, which can also be found in the corresponding R-Markdown document in the paper's GitHub repository, <https://github.com/cboettig/bad-forecast-good-decision>.
-While computer code can be more verbose and difficult to read than mathematical formulae presented here, it is also less ambiguous.
-Though R is already widely used among ecologists, the implementations concern mostly matrix algebra that can be translated to other languages, and the `MDPtoolbox` package used here is also implemented in Matlab [@Marescot2013; @MDPtoolbox].
 
@@
 
-Here we provide annotated code necessary to completely reproduce all of the analysis presented in the main paper.
+Because the methods applied here: Stochastic Dynamic Programming, Iterative Forecasts assessed with Proper Scoring Rules, or Adaptive Management, are not themselves novel, the reader is referred to previous reviews and text books which can provide a thorough introduction to methods, in particular, @Marescot2013 and @Smith1981.
+However, implementation of these approaches generally depends on numerical methods, for which even precise mathematical formulae or pseudo-code descriptions can be inadequate to specify unambiguously or to properly facilitate replication [@Barnes2010].
+As such, this appendix includes computer code in the R language, which can also be found in the corresponding R-Markdown document in the paper's GitHub repository, <https://github.com/cboettig/bad-forecast-good-decision>.
+While computer code can be more verbose and difficult to read than mathematical formulae presented here, it is also less ambiguous.
+Though R is already widely used among ecologists, the implementations concern mostly matrix algebra that can be translated to other languages, and the `MDPtoolbox` package used here is also implemented in Matlab [@Marescot2013; @MDPtoolbox].
+
+Below, I provide annotated code necessary to completely reproduce all of the analysis presented in the main paper.
 This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving Markov Decision Processes (MDP) using stochastic dynamic programming functionality, `expm` for matrix exponentials [@expm], and a few custom MDP functions provided by our package, `mdplearning` [@mdplearning].
@@
 This file is also available as an RMarkdown document [@rmarkdown] at  <https://github.com/cboettig/bad-forecast-good-decision>.
+The code below generates the data tables required for each plot shown in the main paper.
+Data are stored as `.csv` tables to be accessible to any relevant software program.
+Plotting commands used in the main paper are embedded in the paper's RMarkdown file at <https://github.com/cboettig/bad-forecast-good-decision>.
 
@@
 
+# Model definitions
+
 ```{r}
@@
 On a discrete grid of possible states and actions, we can define the growth rate of a given state $X_t$ subject to harvest $H_t$, $f(X_t,H_t)$ as set of matrices.
-Each matrix $i$ gives the transition probabilities for any current state to any future state,
-given that action $i$ is taken.
+Each matrix $i$ gives the transition probabilities for any current state to any future state, given that action $i$ is taken.
 
@@
 
+We store the replicate predictions under each model at each timestep, along with the scores, in a `predictions.csv` table for plotting.
+
 ```{r}
@@
 While straight forward, this is what makes adaptive management for MDP problems particularly computationally intensive and thus poorly suited for handling a large space of candidate models.
-Nevertheless, our example of two models is most feasible, and even the larger example of 36 models (appendix B) is not unreasonable.
+Nevertheless, our example of two models is most feasible, and even the larger example of 36 models is not unreasonable.
 Here I make use of the `mdplearning` R package, which provides a straight-forward implementation of this algorithm.
@@
 adaptive_management <- memoise::memoise(mdp_learning, cache = memoise::cache_filesystem("cache/"))
-x0 <- which.min(abs(states - 6))
-
 am1 <- adaptive_management(transitions[1:2], reward, discount, 
@@
 
+I also compare the adaptive management solution to an approach which still integrates uncertainty over both models, but treats the resulting policy as fixed for the duration of management, rather than adaptive _learning_ by updating probabilities.
+This approach is sometimes referred to as uncertainty "planning," as it accounts for the same initial uncertainty over models but does not attempt to learn.
+For consistency, I use the same model prior as in the learning case, which heavily favors model 1 to begin with.
+Not surprisingly, the resulting policy is nearly identical to that under model 1 alone.
+
 ```{r}
-am <- am1$df %>% mutate(belief = am1$posterior$V1,
-                        stock = states[state],
-                        quota = actions[action])
+# Compute policy using SDP over both models, given fixed prior beliefs
+policy_planning <- memoise::memoise(mdp_compute_policy, cache = memoise::cache_filesystem("cache/"))
+policy <- policy_planning(transitions[1:2],  
+                          reward = reward, 
+                          discount = discount,
+                          model_prior = c(0.99, 0.01),
+                          Tmax = 50, 
+                          epsilon = 0.001, 
+                          max_iter = 2000)
+```
+
+```{r}
+# Simulate under true model using fixed planning policy
+non_am <- mdp_planning(transitions[[3]],  
+                       reward = reward, 
+                       discount = discount,
+                       model_prior = c(0.99, 0.01),
+                       x0 = x0, 
+                       a0 = 1,
+                       policy = policy$policy,
+                       Tmax = 50)  %>% 
+  mutate(belief = 0.99, # belief in model 1 is fixed
+         method = "planning")
+```
+
+We combine results of the two methods and store the resulting `data.frame` of stock size, quota, and belief in Model 1 over time as `am.csv` for later plotting.
+
+```{r}
+am <- am1$df %>%
+  mutate(belief = am1$posterior$V1,
+         method = "learning") %>%
+  bind_rows(non_am) %>%
+  mutate(stock = states[state],
+         quota = actions[action]) %>%
+  select(time, stock, quota, belief, method)
+
 write_csv(am, "../data/am.csv")
@@
 
-### Additional calculations for plots
+## Adaptive management over many models
+
+We repeat the analysis above over a grid of $r$ and $K$ values.
+
+```{r}
+# compare over grid set of candidate r and K
+rs <- c(0.01, 0.1, 0.2, 0.3, 0.4, 0.5, 2)
+Ks <- c(0.5, 1, 1.5, 4, 10, 16)
+sigma <- 0.1
+n_models <- length(rs)*length(Ks)
+pars <- expand.grid(rs, Ks)
+names(pars) <- c("r", "K")
+pars$sigma <- 0.1
+closure <- function(r, K){
+  function(x, h = 0){
+  s <- pmax(x - h, 0)
+  s + s * (r * (1 - s / K) )
+  }
+}
+many_models <-  lapply(1:n_models, function(i) closure(pars[i,"r"], pars[i,"K"]))
+
+# transition matrices
+many_transitions <- lapply(seq_along(many_models), function(i) 
+  transition_matrices(many_models[[i]], states,  actions, sigma))
+names(many_transitions) <- as.character(seq_along(many_models))
+true_transition <- transition_matrices(f3, states, actions, sigma_g)
+
+## Once again, initial belief favors "model 1"
+model_prior <- c(rep(.01, n_models-1) / (n_models - 1), .99)
+
+```
+
+With the transition matrices defined for all 42 models, we perform adaptive management as before:
+
+```{r}
+adaptive_management <- memoise::memoise(mdp_learning, cache = memoise::cache_filesystem("cache/"))
+am_many <- adaptive_management(many_transitions,  
+                               reward = reward, 
+                               discount = discount,
+                               model_prior = model_prior,
+                               x0 = x0, 
+                               Tmax = 20, 
+                               true_transition = true_transition, 
+                               epsilon = 0.001, 
+                               max_iter = 2000)
+
+```
+
+As before, we can compare the adaptive learning case to the policy which accounts for the initial uncertainty over all `r n_models` models, but does not seek to learn.
+
+```{r}
+policy_many <- policy_planning(many_transitions, reward = reward,  discount = discount,
+                          model_prior = model_prior, Tmax = 20, 
+                          epsilon = 0.001, max_iter = 2000)
+
+non_am_many <- mdp_planning(true_transition,  
+                       reward = reward, 
+                       discount = discount,
+                       model_prior = model_prior,
+                       x0 = x0, 
+                       a0 = 1,
+                       policy = policy_many$policy,
+                       Tmax = 20) %>% 
+  mutate(belief = 0.99, # belief in model 1 is fixed
+         method = "planning")
+
+model1 <- dim(am_many$posterior)[2]
+
+am_multi <- am_many$df %>% 
+  mutate(belief = am_many$posterior[,model1],
+         method = "learning") %>%
+  bind_rows(non_am_many) %>%
+  mutate(stock = states[state],
+         quota = actions[action]) %>%
+  select(time, belief, stock, quota, method)
+
+write_csv(am_multi, "../data/am_multi.csv")
+```
+
+# Posteriors
+
+```{r}
+posteriors <- data.frame(r = rs, K = Ks, t(am_many$posterior)) %>%
+  pivot_longer(starts_with("X"), 
+               values_to = "probability",
+               names_to = "time") %>% 
+  mutate(time = as.integer(as.factor(time)))
+
+i <- 20
+r_marginals <- posteriors %>% group_by(time,r) %>% 
+  summarise(prob = sum(probability), .groups = "drop") %>% 
+  filter(time %in% i)
+
+r_marginals %>%  ggplot(aes(r, prob, group=time)) + geom_bar(stat="identity") + facet_wrap(~time)
+```
+
+## Marginal Posterior Probability evolution over K
+
+```{r}
+i = 20
+K_marginals <- posteriors %>% group_by(time,K) %>% 
+  summarise(prob = sum(probability), .groups = "drop") %>% 
+  filter(time %in% i) 
+K_marginals %>%
+  ggplot(aes(K, prob, group=time)) + geom_bar(stat="identity") + facet_wrap(~time) + scale_x_log10()
+
+```
+
+# Economic value tabulation
 
@@
 
+# Growth models
+
 To plot growth curves of individual models (Fig 4a), we evaluate
@@
 
+## Growth models in the many-models case:
+
+```{r plot_models, fig.width=6, fig.height=4}
+final_belief <- t(am_many$posterior)
+final_belief <- final_belief[,dim(final_belief)[[2]]]
+posteriors <- data.frame(model = as.character(1:length(many_models)), belief = final_belief)
+
+model_set <-c(f3,  many_models)
+names(model_set) <- c("true", 1:length(many_models))
+d <- 
+  map_dfc(model_set, function(f) f(states) - states) %>%
+  mutate(state = states) %>% pivot_longer(names(model_set[-1]), "model") %>%
+  left_join(posteriors)
+
+d %>% 
+  ggplot(aes(state, value, group=model, alpha = belief)) +
+  #geom_hline(aes(yintercept = 0), lwd=1, lty=2) + 
+  geom_line(lwd=1.5) + 
+  geom_line(aes(state, true, alpha=1), col = "red", show.legend = FALSE) +
+  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
+  ylab(bquote(f(x) - x)) + xlab("x") 
+```
+
 # References

### 2021-03-31 manuscript/appendix_B.Rmd
consolidate appendix

@@
   coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
-  ylab(bquote(f(x) - x)) + xlab("x") + viridis::scale_color_viridis()
-
+  ylab(bquote(f(x) - x)) + xlab("x") + scale_color_grey()
 ```

### 2021-03-31 manuscript/manuscript.Rmd
consolidate appendix

@@
 abstract: |
-  Model-based forecasts have been enabled by recent explosion of data and computational methods and spurred on by decision-makers appetite for forecasts in everything from elections to pandemic response.
+  Encouraged by decision makers appetite for future information on topics ranging from elections to pandemics, and enabled by the explosion of data and computational methods,  model based forecasts have garnered increasing influence on a breadth of decisions in modern society.
 Using a classic example from fisheries management, I demonstrate that selecting the model that produces the most accurate and precise forecast can lead to decidedly worse outcomes.
@@
 Within a single iteration of adaptive management, the belief over models is updated to near certainty in Model 2, resulting in higher harvests and lower stock sizes similar to managing under Model 2 alone, Fig 2A.
-Panel B: given 36 candidate models over a broad range of parameters, adaptive management leads to models that overharvest even more, leading to even lower stock sizes and lower harvest quotas."}
+Panel B: given 42 candidate models over a broad range of parameters, adaptive management leads to models that overharvest even more, leading to even lower stock sizes and lower harvest quotas."}
 am <- read_csv("../data/am.csv")
-fig3a <- am %>%  filter(time <=20) %>%
-  ggplot(aes(time, stock, col = belief)) + 
-  geom_line(show.legend = FALSE) + 
-  geom_point(show.legend = FALSE) +
-  geom_point(aes(time, quota), col="darkgreen", shape=2) + 
+fig3a <- am %>%  filter(time <=20) %>% #
+  ggplot(aes(time, stock, group = method)) + 
+  geom_line(aes(lty=method), col = "gray40") + 
+  geom_point(aes(col = belief), show.legend = FALSE, size=2) +
+  geom_line(aes(y = quota, lty=method), col = "gray40") + 
+  geom_point(aes(time, quota, col = belief), shape=2, size=2) + 
   scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) +
-  ylab("stock size") + labs(subtitle="A. Two models") + ylim(c(0,11))
+  ylab("fish stock") + labs(subtitle="A.")
 
@@
 fig3b <- am_multi %>%
-  ggplot(aes(time, stock, col = belief)) + 
-  geom_line() + 
-  geom_point() +  
-  geom_point(aes(time, quota), col="darkgreen", shape=2) + 
+   ggplot(aes(time, stock, group = method)) + 
+  geom_line(aes(lty=method), col = "gray40") + 
+  geom_point(aes(col = belief), show.legend = FALSE, size=2) +
+  geom_line(aes(y = quota, lty=method), col = "gray40") + 
+  geom_point(aes(time, quota, col = belief), shape=2, size=2) + 
   scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) +
-  ylab("fish stock") + labs(subtitle="B. 36 models")+ ylim(c(0,11))
+  ylab("fish stock") + labs(subtitle="B.")
 
@@
 Plotting the growth rate functions of each model, [Fig 4A], it is hardly surprising that no method exists which would not prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3.
-However, Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 4B], while Model 2 leads to over-harvesting.
+Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 4B], while Model 2 leads to over-harvesting.
 The explanation comes from noticing that the stock size corresponding to the maximum growth rate under Model 1 (the peak of the curve) falls at almost exactly the same stock size as that of the peak growth rate for Model 3.

### 2021-04-02 manuscript/appendix.Rmd
:pen:

@@
 
-```{r plot_models, fig.width=6, fig.height=4}
+```{r plot_many_models, fig.width=6, fig.height=4}
 final_belief <- t(am_many$posterior)

### 2021-04-02 manuscript/appendix_B.Rmd
:pen:

@@
 ```{r}
-# Try 6 log-spaced values
-rs <- 10 ^ seq(0.01, 2, length=6)
-Ks <- 10 ^ seq(-.3, 1.2, length=6)
+library(nimble)
+
+logistic  <- nimble::nimbleCode({
+
+  x[1] <- x0
+  for(t in 1:(N-1)){
+    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)  
+    x[t+1] ~ dlnorm(log(mu[t]), sd = sigma)
+  }
+  r ~ dunif(0,10);
+  K ~ dunif(0,20);
+  sigma ~ dunif(0,0.1);
+
+})
+l_model <- nimbleModel(logistic,constants = list(N = 1e3, x0 = 6.0))
+
+f3_nimble  <- nimble::nimbleCode({
+  x[1] <- x0
+  for(t in 1:(N-1)){
+    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)  
+    x[t+1] ~ dlnorm(log(mu[t]), sd = sigma)
+  }
+})
+model <- nimbleModel(f3_nimble, constants = list(r = .002, K = 10, sigma = .05, N = 1e3), inits = list(x0 = 6.0))
+
+simulate(model)
+ggplot(data.frame(t = 1:1e3, x = model$x), aes(t,x)) + geom_line()
+```
+```{r}
+l_model$setData(list(x = model$x))
+c_model <- compileNimble(l_model)
+l_mcmc <- buildMCMC(c_model)
+c_mcmc <- compileNimble(l_mcmc)
+df <- runMCMC(c_mcmc, niter = 10000, nburnin = 1000)
+
+A <- df %>% as_tibble() %>% summarise(across(everything(), list(mean = mean, sd = sd)))
+```
+
+```{r}
+rs <-A$r_mean #10 ^ seq(log10(.01), log10(2), length=60)
+Ks <- A$K_mean # seq(1,16,length=50) # 10^ seq(log10(1), log10(16), length=7)
 
 # compare over fixed set of candidate r and K
-rs <- c(0.01, 0.1, 0.2, 0.3, 0.4, 0.5, 2)
-Ks <- c(0.5, 1, 1.5, 4, 10, 16)
 n_models <- length(rs)*length(Ks)
-sigmas <- 0.1
+sigmas <- 0.05
 pars <- expand.grid(rs, Ks)
@@
 
-```{r plot_models, fig.width=6, fig.height=4}
+```{r plot_models, fig.width=10, fig.height=7}
 model_set <-c(f3,  models)
@@
 d %>% pivot_longer(names(model_set[-1]), "model") %>%
-  mutate(model = as.numeric(model)) %>%
   ggplot(aes(state, value, col = model)) +
   geom_hline(aes(yintercept = 0), lwd=1) + 
-  geom_point(alpha = .5) + 
+  geom_line(alpha = .5, show.legend = FALSE) + 
   geom_line(aes(state, true), col = "red") +

### 2021-04-02 manuscript/manuscript.Rmd
:pen:

@@
 ---
-title: 'When to ignore perfect information: how iterative learning can lead to worse decisions'
+title: 'Iterative forecasting and the trap of self-fulfilling prophecies'
 author:
@@
   Encouraged by decision makers appetite for future information on topics ranging from elections to pandemics, and enabled by the explosion of data and computational methods, model based forecasts have garnered increasing influence on a breadth of decisions in modern society.
-Using a classic example from fisheries management, I demonstrate that selecting the model that produces the most accurate and precise forecast can lead to decidedly worse outcomes.
+Using a classic example from fisheries management, I demonstrate that selecting the model or models that produce the most accurate and precise forecast can lead to decidedly worse outcomes.
+This creates a cycle of self-fulfilling prophecy, in which the manager not only over-exploits the fishery leading to declining biomass and declining yields, but also becomes increasingly convinced that these actions are not over-exploitation but in fact consistent with the best models and data for sustainable management.
 
@@
 In this paper, I will demonstrate that iterative learning or iterative forecasting can also be counter-productive to management outcomes:
-that in some cases, a manager would be better to ignore new information which could reduce
-uncertainty over available models. Put another way, the value of information (VOI, as
-measured by the expected utility given that information minus the utility without it; see @Howard1966; @Katz1987), can actually be negative. This is not a consequence of imperfect information
-(e.g. observation error) or the result of model over-fitting.  The issue is an intrisic 
-consequence of the fact that many models can lead to the same decisions, and no generic
-solution to the problem exists. However, as this example shows, by better understanding
+that in some cases, a manager would be better to ignore new information which could reduce uncertainty over available models.
+Put another way, the value of information (VOI, as measured by the expected utility given that information minus the utility without it; see @Howard1966; @Katz1987), can actually be negative.
+This is not a consequence of imperfect information (e.g. observation error) or the result of model over-fitting.
+The issue is an intrinsic consequence of the fact that many models can lead to the same decisions, and no generic solution to the problem exists. 
+Worse, whenever this situation arises, the decision-maker becomes trapped by a form of self-fulfilling prophecy: 
+the model(s) which the decision-makers have learned to favor lead to decisions and outcomes that closely match model expectations.
+Consequently, the decision-maker can become ever more firmly entrenched in the belief that their new models are accurate and the mediocre outcomes they experience are as good as it gets,
+while confidently rejecting earlier models which would have led to much better outcomes.  
+As this example will show, by better understanding
 what aspects of model are most and least important to shaping a particular decision, 
@@
 I illustrate this problem using an example from fisheries management. 
-Fisheries are a significant economic and conservation concern world
-wide and their management remains an important debate [e.g. @Worm2006;
-@Worm2009; @Costello2016]. Moreover, their management has been a proving 
-grounds for theoretical and practical decision-making issues which are widely
-applicable in other areas of ecology and conservation [@Ludwig1993; @Lande1994],
-and one that has long wrestled with issues of uncertainty in the context
-of management decisions [e.g. @Clark1973; @Reed1979; @Walters1981; @Ludwig1982].
-While modern fisheries management frequently relies on complex models which may
-contain scores of parameters to reflect the specific age or stage structure of a
-specific fish stock [@ramlegacy2018; @ram], I will rely on simple, well-studied models which
-permit greater intuition and generalization [@Levins1966]. Consistent with 
-such previous work [@Schaefer1954; @Clark1973; @Reed1979; @Walters1981; @Ludwig1982; @Costello2016],
-let us consider the problem of determining the optimal harvest policy given a 
-measurement of the current stock size.
+Fisheries are a significant economic and conservation concern world wide and their management remains an important debate [e.g. @Worm2006; @Worm2009; @Costello2016].
+Moreover, their management has been a proving grounds for theoretical and practical decision-making issues which are widely applicable in other areas of ecology and conservation [@Ludwig1993; @Lande1994], and one that has long wrestled with issues of uncertainty in the context of management decisions [e.g.
+@Clark1973; @Reed1979; @Walters1981; @Ludwig1982].
+While modern fisheries management frequently relies on complex models which may contain scores of parameters to reflect the specific age or stage structure of a specific fish stock [@ramlegacy2018; @ram], I will rely on simple, well-studied models which permit greater intuition and generalization [@Levins1966].
+Consistent with such previous work [@Schaefer1954; @Clark1973; @Reed1979; @Walters1981; @Ludwig1982; @Costello2016], let us consider the problem of determining the optimal harvest policy given a measurement of the current stock size.
 
-For illustrative purposes, we will focus on the simplest example of model uncertainty,
-considering two alternative models of equal complexity but differing in the estimated
-value of certain parameters.
+For illustrative purposes, we will focus on the simplest example of model uncertainty: two alternative models of equal complexity but differing in the estimated value of certain parameters.
 While some authors distinguish between model uncertainty and parameter uncertainty, I will refer collectively to any uncertainty that arises from our imperfect knowledge of the system as "model uncertainty," because the distinction is largely a consequence of notation rather something more intrinsic.
-(For example, two structurally different models, $f(x) = a x$ and $f(x) = a x^2$ can be considered merely different parameterizations of $f(x) = a x^b$, or vice versa.)
+For example, two structurally different models, $f(x) = a x$ and $f(x) = a x^2$ can be considered merely different parameterizations of $f(x) = a x^b$, or vice versa.
 
@@
 We further imagine that the function $f$ is not known precisely.
-Again for simplicity, we restrict ourselves to two simple candidate models $f_1$ and $f_2$.
+We restrict ourselves to two simple candidate models $f_1$ and $f_2$.
 Both share the same underlying structure of logistic recruitment (known as the Gordon-Schaefer model in fisheries context owing to groundbreaking work independently by @Schaefer1954 and @Gordon1954), differing only in their choice of certain parameters:
@@
 
-For simplicity, we will assume $\xi_t(\sigma)$ represents log-normal random noise with mean of unity and log-standard-deviation $\sigma_i$ for each model.
+Let us assume $\xi_t(\sigma)$ represents log-normal random noise with mean of unity and log-standard-deviation $\sigma_i$ for each model.
 Model 1 is given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, Model 2 by $r_2 = 0.5$,  $K_2 = 10$, $\sigma = 0.075$ (in dimensionless units).
@@
 
-The task of deciding whether Model 1 or Model 2 would be the better choice for decision making is 
-thus perhaps the simplest example of the much studied issue of model uncertainty that we can pose. 
-As in any real world scenario, neither model is the true model, but nevertheless this model set
-contains a good enough approximation of the true model to make good decisions.  However, any of
-the well-developed approaches for decision-making under model uncertainty will prefer Model 2 over
-Model 1, despite the fact that the optimal policy under Model 2 leads to much worse outcomes
-ecologically and economically.
+The task of deciding whether Model 1 or Model 2 would be the better choice for decision making is thus perhaps the simplest example of the much studied issue of model uncertainty that we can pose. 
+As in any real world scenario, neither model is the true model, but nevertheless this model set contains a good enough approximation of the true model to make good decisions.
+However, any of the well-developed approaches for decision-making under model uncertainty will prefer Model 2 over Model 1, despite the fact that the optimal policy under Model 2 leads to much worse outcomes ecologically and economically.
 
@@
 
-We will use this example to illustrate two alternative approaches for iterative learning over model uncertainty: iterative forecasting and adaptive management.
-The central difference in the approaches
-is that iterative forecasting is premised on the ability to score the predictions of alternative models.
-Iterative forecasting is silent on the issue of what to do with those scores, this is left up to the 
-decision-maker.  Adaptive management approaches, by contrast, explicitly seek to integrate probabilities over all candidate models to reach a decision.  I consider each in turn.
+I will use this example to illustrate two alternative approaches for iterative learning over model uncertainty: iterative forecasting and adaptive management.
+The central difference in the approaches is that iterative forecasting is premised on the ability to score the predictions of alternative models.
+Iterative forecasting is silent on the issue of what to do with those scores, this is left up to the decision-maker.
+Adaptive management approaches, by contrast, explicitly seek to integrate probabilities over all candidate models to reach a decision.
+I consider each in turn.
 
@@
 In Scenario A (passive observation) the fish stock is unharvested and allowed to recover towards carrying capacity (as simulated under our "true" model, Model 3) while comparing the observed stock size in each subsequent time step to the distribution predicted under Model 1 and Model 2 respectively [Fig 1].
-The mean, $\mu_t$ and variance, $\sigma_t$ of the forecast are compared against the true observation $x_t$ using a Proper scoring rule of @Gneiting2007,
+The mean, $\mu_t$ and variance, $\sigma_t$ of the forecast are compared against the true observation $x_t$ using a proper scoring rule given by @Gneiting2007,
 
@@
 
+In both scenarios, the careful comparison through proper scoring rules has led us to select the worse-performing model.
+Crucially, a manager operating under this selection would have little indication that their model was flawed: both future stock sizes and expected harvest yields consistently match model predictions.
 Had we been able to include Model 3 in our forecast comparisons, it would equal or outperform the forecasting skill of both Model 1 and Model 2 (as guaranteed by the theorem of @Gneiting2007), while also matching or out-performing their economic utility (as guaranteed by the theorem of @Reed1979).
-In practice, we never have access to the generating model.
+In practice, we never have access to the generating model, so it is reasonable to expect model selection to determine the best approximation.
+As we see here, the best approximation for forecasting future states does not in fact lead to better outcomes.
 
 One obvious limitation in this comparison is that scenario B treats each model as fixed over the entire course of the simulation.
-It would be possible to generate forecasts for the next $n$ steps, but these forecasts would also be conditional on the management action (i.e.
-harvest quota) selected at each step.
+It would be possible to generate forecasts for the next $n$ steps, but these forecasts would also be conditional on the management action (i.e. harvest quota) selected at each step.
 To evaluate two-step-ahead predictions we must consider each possible action under each possible state predicted by the step-ahead forecast, weighted by probability of the state given the model and the probability of the model.
@@
 To demonstrate that the behavior is not driven by failure to explore sufficiently, (which might be addressed by an active adaptive management), I will assign initial probability that model 2 is true at 1%.
-After a single iteration of learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct  [Fig 3].
-As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating, which achieves a net present value over 100 time steps that only 31% that expected under management using Model 1 alone [Fig 2].
+After a single iteration of learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct  [Fig 3A].
+I compare these results to 
+As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating, which achieves a net present value that only 31% that expected under management using Model 1 alone [Fig 2].
 
-```{r figure3, fig.cap="Adaptive management under model uncertainty, color indicating the belief that Model 1 is correct (blue), lines indicating stock biomass over time, green triangles indicating harvest quotas set by adaptive management.
-Panel A: Each model is assigned an initial prior belief.
-To ensure that any issues with passive exploration are not the cause of preferring Model 2, the initial belief in Model 2 is set to 1%.
-Within a single iteration of adaptive management, the belief over models is updated to near certainty in Model 2, resulting in higher harvests and lower stock sizes similar to managing under Model 2 alone, Fig 2A.
-Panel B: given 42 candidate models over a broad range of parameters, adaptive management leads to models that overharvest even more, leading to even lower stock sizes and lower harvest quotas."}
+```{r figure3, fig.height=5, fig.width=8, fig.cap="Adaptive management under model uncertainty.
+Solid lines trace the trajectories of the state (fish stock, circles) and action (harvest quota, triangles), under adaptive management (learning).
+Dotted lines trace the corresponding trajectories if iterative learning is omitted, leaving the prior belief fixed throughout the simulation (planning).
+Color indicates the belief that Model 1 is correct (blue), with an intial prior belief of 99%.
+Panel A: Management over the two candidate models, Model 1 and Model 2.
+Within a single iteration of adaptive management, the belief over models switches from a prior belief that heavily favored Model 1 to a posterior that favors Model 2 with near certainty.
+Future iterations reinfornce the belief in Model 2, resulting in both depressed harvests and low stock sizes (solid lines).
+If no iterative learning updates are performed, stock sizes and realized harvests (and thus economic profit) are both higher.
+Panel B: given 42 candidate models over a broad range of parameters, adaptive management quickly reduces the probability on model 1, and substantially underperforms management without learning (dotted lines), though not over-harvesting quite as badly as under the two-model scenario."}
 am <- read_csv("../data/am.csv")
-fig3a <- am %>%  filter(time <=20) %>% #
-  ggplot(aes(time, stock, group = method)) + 
-  geom_line(aes(lty=method), col = "gray40") + 
-  geom_point(aes(col = belief), show.legend = FALSE, size=2) +
-  geom_line(aes(y = quota, lty=method), col = "gray40") + 
-  geom_point(aes(time, quota, col = belief), shape=2, size=2) + 
+fig3a <- am %>%  filter(time <=20) %>% 
+  pivot_longer(c(stock, quota), values_to="stock", names_to="key") %>%
+  mutate(key = factor(key, levels = c("stock", "quota"))) %>%
+  ggplot(aes(time, stock, shape = key, group = interaction(method,key))) + 
+  geom_line(aes(lty=method), col = "gray40", show.legend = FALSE) + 
+  geom_point(aes(col = belief), size=2, show.legend = FALSE) +
   scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) +
@@
 fig3b <- am_multi %>% 
-   ggplot(aes(time, stock, group = method)) + 
+  pivot_longer(c(stock, quota), values_to="stock", names_to="key") %>%
+  mutate(key = factor(key, levels = c("stock", "quota"))) %>%
+  ggplot(aes(time, stock, shape = key, group = interaction(method,key))) + 
     geom_line(aes(lty=method), col = "gray40") + 
-  geom_point(aes(col = belief), show.legend = FALSE, size=2) +
-  geom_line(aes(y = quota, lty=method), col = "gray40") + 
-  geom_point(aes(time, quota, col = belief), shape=2, size=2) + 
+    geom_point(aes(col = belief), size=2) +
     scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) +
@@
 
-Increasing the space of possible models to cover a whole plausible range of parameters $r$ and $K$ does nothing to resolve this problem.
+Increasing the space of possible models to cover a whole plausible range of parameters $r$ and $K$ does not resolve this problem.
 Learning is somewhat slower over the larger range of possibilities, but nevertheless converges towards parameter values with lower growth rates and lower carrying capacity, for which the target biomass is far lower [Fig 3B; details in appendix B].

### 2021-04-02 manuscript/appendix.Rmd
:pen:

--- 2021-03-31 manuscript/appendix.Rmd
@@
 ---
-title: 'Appendix A'
+title: 'Appendix'
 date: "`r Sys.Date()`"
 author: Carl Boettiger
-journal: Appendix
+journal: TBD
 layout: 3p
@@
 ```{r}
-transitions <- lapply(seq_along(models), 
-                      function(i) transition_matrices(models[[i]], 
-                                                      states, 
-                                                      actions, 
-                                                      model_sigmas[[i]]))
+transitions <- lapply(seq_along(models), function(i) 
+  transition_matrices(models[[i]], states, actions, model_sigmas[[i]]))
 names(transitions) <- c("1", "2", "3")
@@
 
-# Simulations and step-ahead forecasts
+# Simulations and step-ahead forecasts (Fig 1, 2A)
 
@@
 
-# Proper scores
+# Proper scores (Fig 1)
 
@@
 
+# Economic value tabulation (Fig 2B)
+
+To plot the economic value over time, we must sum up the discounted values
+at each time step, and then average over replicate simulations of each model:
+
+```{r plot_npv}
+##  Net Present Value accumulates over time
+npv_df <- sims %>% 
+  group_by(model, reps) %>%
+  mutate(npv = cumsum(value * discount ^ time)) %>%
+  group_by(time, model)  %>% 
+  summarise(mean_npv = mean(npv), .groups="drop") %>% 
+  arrange(model, time)
+
+write_csv(npv_df, "../data/npv_df.csv")
+```
+
 # Adaptive Management
@@
 Passive adaptive management using a Bayesian learning scheme still learns the wrong model.
-Adaptive management for MDP problems is particularly computationally intensive, and not directly supported by the `MDPtoolbox` implementation, though the principle involves only a relatively straightforward, if not particularly efficient extension to the task of solving and MDP problem for a single model shown above.
+Adaptive management for MDP problems is particularly computationally intensive, and not directly supported by the `MDPtoolbox` implementation.
+Nevertheless, the principle involves only a relatively straightforward, if not particularly efficient extension to the task of solving and MDP problem for a single model shown above.
 Given a set of transition probability matrices $P_i$ and a belief probability $b_i$ that the $i$th model is correct, it simple to define the transition matrices over possible models by the laws of conditional probability, $P = \sum_i P_i b_i$, given $\sum_i b_i = 1$.
@@
 
+We can also compare these results to the net economic value achieved under Model 1 alone:
+
+```{r}
+# tabular comparisons
+npv <- npv_df %>% group_by(model) %>% summarize(npv = max(mean_npv))
+
+am_npv <- sum(am$value * discount ^ am$time)
+am_economics_percent <- round(am_npv / npv[[1,"npv"]] * 100)
+
+mean_state <- sims %>% group_by(model) %>% summarize(state = mean(state))
+am_ecology_percent <- round(mean(states[am$state])/mean_state[[1, "state"]]*100)
+```
+
+Under adaptive management, the manager realizes only `r am_economics_percent`%
+of the economic value that would be achieved under Model 1 alone, and only 
+`r am_ecology_percent`% of the spawning stock biomass that would have been 
+achieved under Model 1 alone.
+
 ## Adaptive management over many models
@@
 
-With the transition matrices defined for all 42 models, we perform adaptive management as before:
+With the transition matrices defined for all `r n_models` models, we perform adaptive management as before:
 
@@
          method = "planning")
+```
 
+Again we record the posterior belief in the 'model 1' parameter values, and stack the data frames for learning and planning-only cases into a single file used to generate figure 3B.
+
+```{r}
 model1 <- dim(am_many$posterior)[2]
-
 am_multi <- am_many$df %>% 
@@
 
+With `r n_models` possible models instead of two, the probability weight (belief) in Model 1 does not fully capture the evolution of the belief probabilities over time.
+To better illustrate how these beliefs evolve over the course of the simulation, I plot the marginal probabilities over values of $r$ and $K$ across the models over time.
+Note that Model 1 values, $r = 2$, $K = 16$, are quickly disfavored.
+
 ```{r}
-posteriors <- data.frame(r = rs, K = Ks, t(am_many$posterior)) %>%
+posteriors <- 
+  data.frame(model = as.character(1:n_models), t(am_many$posterior)) %>%
+  left_join(mutate(pars, model = as.character(1:n_models))) %>%
   pivot_longer(starts_with("X"), 
@@
 
-i <- 20
-r_marginals <- posteriors %>% group_by(time,r) %>% 
-  summarise(prob = sum(probability), .groups = "drop") %>% 
-  filter(time %in% i)
+r_marginals <- posteriors %>% 
+  group_by(time,r) %>% 
+  summarise(prob = sum(probability), .groups = "drop")
 
-r_marginals %>%  ggplot(aes(r, prob, group=time)) + geom_bar(stat="identity") + facet_wrap(~time)
+K_marginals <- posteriors %>% group_by(time,K) %>% 
+  summarise(prob = sum(probability), .groups = "drop")
 ```
 
-## Marginal Posterior Probability evolution over K
+From the table of marginal probabilities, we generate the corresponding plots
 
-```{r}
-i = 20
-K_marginals <- posteriors %>% group_by(time,K) %>% 
-  summarise(prob = sum(probability), .groups = "drop") %>% 
-  filter(time %in% i) 
-K_marginals %>%
-  ggplot(aes(K, prob, group=time)) + geom_bar(stat="identity") + facet_wrap(~time) + scale_x_log10()
+```{r fig.width=8, fig.height=4, fig.cap = "Fig S1: Marginal probabilities of parameters $r$ and $K$ over the course of adaptive management over all 42 models." }
+belief_r <- r_marginals %>% 
+  mutate(r = as.character(r)) %>%
+  ggplot(aes(time, prob, group = r, col = r)) + geom_line() +
+  geom_point()
 
+belief_K <- K_marginals %>% 
+  mutate(K = as.character(K)) %>%
+  ggplot(aes(time, prob, group = K, col = K)) + geom_line() +
+  geom_point()
+
+belief_r + belief_K 
 ```
 
-# Economic value tabulation
-
-To plot the economic value over time, we must sum up the discounted values
-at each time step, and then average over replicate simulations of each model:
-
-```{r plot_npv}
-##  Net Present Value accumulates over time
-npv_df <- sims %>% 
-  group_by(model, reps) %>%
-  mutate(npv = cumsum(value * discount ^ time)) %>%
-  group_by(time, model)  %>% 
-  summarise(mean_npv = mean(npv), .groups="drop") %>% 
-  arrange(model, time)
-
-write_csv(npv_df, "../data/npv_df.csv")
-```
-
-```{r}
-# tabular comparisons
-npv <- npv_df %>% group_by(model) %>% summarize(npv = max(mean_npv))
-
-am_npv <- sum(am$value * discount ^ am$time)
-am_economics_percent <- round(am_npv / npv[[1,"npv"]] * 100)
-
-mean_state <- sims %>% group_by(model) %>% summarize(state = mean(state))
-am_ecology_percent <- round(mean(states[am$state])/mean_state[[1, "state"]]*100)
-```
-
-Under adaptive management, the manager realizes only `r am_economics_percent`%
-of the economic value that would be achieved under Model 1 alone, and only 
-`r am_ecology_percent`% of the spawning stock biomass that would have been 
-achieved under Model 1 alone.
+For most of the simulation, the most probable $r$ value is around 0.3, and the most probable $K$ value around 1.5.
 
@@
 
-## Growth models in the many-models case:
+The resulting model curves are shown in the main paper, Fig 4A.
 
-```{r plot_models, fig.width=6, fig.height=4}
-final_belief <- t(am_many$posterior)
-final_belief <- final_belief[,dim(final_belief)[[2]]]
-posteriors <- data.frame(model = as.character(1:length(many_models)), belief = final_belief)
+## Growth models in the many-models case
+
+The corresponding figure for the case of `r n_models` models.
+
+```{r plot_many_models, fig.width=7, fig.height=5, fig.cap="growth curves for 42 candidate models, shaded by belief at timestep 20, relative to the true model.
+The curve with parameter values corresponding to 'model 1' are recognized in the highest arc, with belief weight of less than 1%."}
+
+final_belief <- posteriors %>% filter(time == max(time)) %>%
+  select(model, belief = probability) %>% mutate(type = "candidate") %>%
+  bind_rows(data.frame(belief = 1, model="true", type="true"))
 
@@
   map_dfc(model_set, function(f) f(states) - states) %>%
-  mutate(state = states) %>% pivot_longer(names(model_set[-1]), "model") %>%
-  left_join(posteriors)
+  mutate(state = states) %>% pivot_longer(names(model_set), "model") %>%
+  left_join(final_belief)
 
 d %>% 
-  ggplot(aes(state, value, group=model, alpha = belief)) +
-  #geom_hline(aes(yintercept = 0), lwd=1, lty=2) + 
-  geom_line(lwd=1.5) + 
-  geom_line(aes(state, true, alpha=1), col = "red", show.legend = FALSE) +
+  ggplot(aes(state, value, group=model, col = type, alpha = belief)) +
+  geom_line(lwd=1) + 
   coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
@@
 
+\pagebreak
+
 # References

### 2021-04-02 manuscript/manuscript.Rmd
:pen:

@@
 Model 1 is given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, Model 2 by $r_2 = 0.5$,  $K_2 = 10$, $\sigma = 0.075$ (in dimensionless units).
-Having both the larger growth rate and the larger carrying capacity, Model 1 is clearly the more optimistic of the two choices.
+Having both the larger growth rate and the larger carrying capacity, model 1 is clearly the more optimistic of the two choices.
 
-Selecting between Model 1 and Model 2 can thus be considered the simplest illustration of the model uncertainty problem.
+Selecting between model 1 and model 2 can thus be considered the simplest illustration of the model uncertainty problem.
 This is a subset of the more general problem of selecting model parameters, assuming a logistic growth, which itself is a subset of estimating the best structural from (e.g.
@@
 
-The task of deciding whether Model 1 or Model 2 would be the better choice for decision making is thus perhaps the simplest example of the much studied issue of model uncertainty that we can pose. 
+The task of deciding whether Model 1 or model 2 would be the better choice for decision making is thus perhaps the simplest example of the much studied issue of model uncertainty that we can pose. 
 As in any real world scenario, neither model is the true model, but nevertheless this model set contains a good enough approximation of the true model to make good decisions.
-However, any of the well-developed approaches for decision-making under model uncertainty will prefer Model 2 over Model 1, despite the fact that the optimal policy under Model 2 leads to much worse outcomes ecologically and economically.
+However, any of the well-developed approaches for decision-making under model uncertainty will prefer model 2 over model 1, despite the fact that the optimal policy under model 2 leads to much worse outcomes ecologically and economically.
 
@@
 I illustrate the process of model selection by strictly proper scoring rules using two scenarios.
-In Scenario A (passive observation) the fish stock is unharvested and allowed to recover towards carrying capacity (as simulated under our "true" model, Model 3) while comparing the observed stock size in each subsequent time step to the distribution predicted under Model 1 and Model 2 respectively [Fig 1].
+In Scenario A (passive observation) the fish stock is unharvested and allowed to recover towards carrying capacity (as simulated under our "true" model, Model 3) while comparing the observed stock size in each subsequent time step to the distribution predicted under model 1 and model 2 respectively [Fig 1].
 The mean, $\mu_t$ and variance, $\sigma_t$ of the forecast are compared against the true observation $x_t$ using a proper scoring rule given by @Gneiting2007,
@@
 
-In Scenario B (actively harvest), I have first solved for the optimal management strategy for both Model 1 and Model 2 [Fig 1b].
+In Scenario B (actively harvest), I have first solved for the optimal management strategy for both model 1 and model 2 [Fig 1b].
 For small noise and concave functions with linear reward structure this can be done analytically [see proof in @Reed1979], or solved more generally by stochastic dynamic programming [see review by @Marescot2013, details in Appendix].
-Under this scenario, replicate simulations of the stock are harvested at each time step using the optimal quota dictated by either Model 1 and Model 2.
+Under this scenario, replicate simulations of the stock are harvested at each time step using the optimal quota dictated by either model 1 and model 2.
 The resulting stock sizes in the time-step following this harvest are once again compared to the probabilities predicted by each model using Eq \eqref{proper}.
-Model 2 unequivocally outperforms Model 1 in both scenarios of passive observation and active harvest.
+Model 2 unequivocally outperforms model 1 in both scenarios of passive observation and active harvest.
 
-Despite the clearly superior predictive accuracy of Model 2 in both scenarios, the outcomes from management under Model 2 are substantially worse.
+Despite the clearly superior predictive accuracy of model 2 in both scenarios, the outcomes from management under model 2 are substantially worse.
 We can assess such outcomes in less abstract terms than forecasting skill, such as economic value (in dollars) or the ecological value (unharvested biomass).
@@
 While it is possible to include such contributions directly in the utility function being optimized [e.g.
-@Halpern2013], even without doing so, Model 1 maintains both a higher unharvested biomass and also leads to higher economic returns throughout [Fig 2].
+@Halpern2013], even without doing so, model 1 maintains both a higher unharvested biomass and also leads to higher economic returns throughout [Fig 2].
 
 ```{r figure2, message = FALSE, fig.cap = "Ecological and economic performance of each forecast.
-Harvest quotas derived from Model 1 result in a significantly higher fish stock size than under Model 2 (panel A).
-Economic returns under Model 1 are also substantially higher (panel B)"}
+Harvest quotas derived from model 1 result in a significantly higher fish stock size than under Model 2 (panel A).
+Economic returns under model 1 are also substantially higher (panel B)"}
 
@@
 Crucially, a manager operating under this selection would have little indication that their model was flawed: both future stock sizes and expected harvest yields consistently match model predictions.
-Had we been able to include Model 3 in our forecast comparisons, it would equal or outperform the forecasting skill of both Model 1 and Model 2 (as guaranteed by the theorem of @Gneiting2007), while also matching or out-performing their economic utility (as guaranteed by the theorem of @Reed1979).
+Had we been able to include Model 3 in our forecast comparisons, it would equal or outperform the forecasting skill of both model 1 and model 2 (as guaranteed by the theorem of @Gneiting2007), while also matching or out-performing their economic utility (as guaranteed by the theorem of @Reed1979).
 In practice, we never have access to the generating model, so it is reasonable to expect model selection to determine the best approximation.
@@
 I compare these results to 
-As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating, which achieves a net present value that only 31% that expected under management using Model 1 alone [Fig 2].
+As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating, which achieves a net present value that only 31% that expected under management using model 1 alone [Fig 2].
 
@@
 Dotted lines trace the corresponding trajectories if iterative learning is omitted, leaving the prior belief fixed throughout the simulation (planning).
-Color indicates the belief that Model 1 is correct (blue), with an intial prior belief of 99%.
+Color indicates the belief that model 1 is correct (blue), with an intial prior belief of 99%.
 Panel A: Management over the two candidate models, Model 1 and Model 2.
-Within a single iteration of adaptive management, the belief over models switches from a prior belief that heavily favored Model 1 to a posterior that favors Model 2 with near certainty.
-Future iterations reinfornce the belief in Model 2, resulting in both depressed harvests and low stock sizes (solid lines).
+Within a single iteration of adaptive management, the belief over models switches from a prior belief that heavily favored model 1 to a posterior that favors model 2 with near certainty.
+Future iterations reinfornce the belief in model 2, resulting in both depressed harvests and low stock sizes (solid lines).
 If no iterative learning updates are performed, stock sizes and realized harvests (and thus economic profit) are both higher.
-Panel B: given 42 candidate models over a broad range of parameters, adaptive management quickly reduces the probability on model 1, and substantially underperforms management without learning (dotted lines), though not over-harvesting quite as badly as under the two-model scenario."}
+Panel B: given 42 candidate models over a broad range of parameter values, adaptive management quickly reduces the probability on model 1, and substantially underperforms management without learning (dotted lines).
+While outcomes improve marginally relative to the two-model case (figure A) they remain significantly worse than had no iterative learning been included."}
 am <- read_csv("../data/am.csv")
@@
 
-Increasing the space of possible models to cover a whole plausible range of parameters $r$ and $K$ does not resolve this problem.
-Learning is somewhat slower over the larger range of possibilities, but nevertheless converges towards parameter values with lower growth rates and lower carrying capacity, for which the target biomass is far lower [Fig 3B; details in appendix B].
-The manager becomes trapped in vicious cycle as biomass declines further and further, subsequent observations seem more and more consistent with models which most favor over-fishing in the first place.
-As biomass declines, annual harvests and profits also plunge, and yet our manager becomes only more confident in their decisions.
+Increasing the space of possible models to cover a whole plausible range of parameters $r$ and $K$ does little to resolve this problem [Fig 3B]. 
+Iterative updates again quickly dismiss the parameter values assumed by model 1, though with more options to choose from, this probability is spead over a range of seemingly plausible candidate models instead of a single alternative model (see Appendix).
+While the adaptive management of additional actions and observations slowly narrow this subset of plausible models, decisions based on this uncertainty prevent fish stocks from recovering fully, and realize lower harvests as a result.
+Note that learning under either adaptive management approach (using two models or 42), the decision-maker becomes ever-increasingly convinced that they are using the right model or models.
+Future stock sizes fall consistently in the range predicted by the model(s), and consistently outside the range predicted by model 1.
+Consequently, each iteration the managers are only more firmly convinced that they are maintaining the fish stock near the biomass that supports maximum sustainable yield, when in fact they are sustaining a harvest regime that is preventing recovery of the stock to the much higher productivity regime which would have been achieved under model 1.
 
@@
 Moreover, a decision maker employing an adaptive management or iterative forecasting assessment such as those considered here would have no way of realizing that the outcomes they experienced were in fact sub-optimal.
-In both approaches, the manager quickly concludes that Model 1 is entirely implausible, while finding that Model 2 is remarkably accurate at predicting future values.
-This problem is not addressed by re-estimating parameters (equivalently, considering a larger suite of models with all possible $K$ and $r$ values): in fact leads to a model with slightly lower $K$ and even higher level of over-fishing and under-performance than Model 2 (Fig 3B).
+In both approaches, the manager quickly concludes that model 1 is entirely implausible, while finding that model 2 is remarkably accurate at predicting future values.
+This problem is not addressed by re-estimating parameters (equivalently, considering a larger suite of models spanning a wide range of possible $K$ and $r$ values): which improves only marginally over the outcome of considering only two models [Fig 3B].
+Both iterative forecast evaluation and the adaptive management approaches to reducing model uncertainty lead to management becoming trapped in a self-fulfilling prophecy:
+holding the fish population at an unproductively low biomass suggested by model 2 while becoming ever more convinced that model is suitable. 
+Without access to the true model, none of our methods for model choice or reducing model uncertainty can break this trap.
 
 <!-- Intuition, Figure 4 -->
-The reason for Model 1's seemingly contradictory ability to make good decisions but bad forecasts becomes obvious once we compare both curves to that of the underlying model, Model 3.
-Plotting the growth rate functions of each model, [Fig 4A], it is hardly surprising that no method exists which would not prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3.
-Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 4B], while Model 2 leads to over-harvesting.
-The explanation comes from noticing that the stock size corresponding to the maximum growth rate under Model 1 (the peak of the curve) falls at almost exactly the same stock size as that of the peak growth rate for Model 3.
-Meanwhile, the peak of Model 2 occurs at a substantially lower stock size.
+The reason for model 1's seemingly contradictory ability to make good decisions but bad forecasts becomes obvious once we compare both curves to that of the underlying model, model 3.
+Plotting the growth rate functions of each model, [Fig 4A], it is hardly surprising that no method exists which would not prefer the closely overlapping model 2 to the no-where-close model 1 as the better approximation of model 3.
+Nevertheless, decisions based on model 1 are nearly indistinguishable from those based on the true model [Fig 4B], while model 2 leads to over-harvesting.
+The explanation comes from noticing that the stock size corresponding to the maximum growth rate under model 1 (the peak of the curve) falls at almost exactly the same stock size as that of the peak growth rate for model 3.
+Meanwhile, the peak of model 2 occurs at a substantially lower stock size.
 While the optimal control solution appears to depend only on a step-ahead forecast accuracy (indeed, the SDP solution method used here takes only step-ahead forecast probabilities as input, [@Marescot2013]), mathematical analysis showed long ago [@Reed1979] that the optimal solution for this problem depends only on keeping the biomass at the value responsible for the maximum growth rate.
@@
 This phenomenon is not unique to sequential decision problems or optimal control solutions.
-For example, Management Strategy Evaluation [@Punt2016] in fisheries seeks to evaluate pre-specified strategies rather solve for the best possible strategy through optimal control.
-This approach is well justified -- optimal control techniques such as stochastic dynamic programming illustrated here quickly become intractable for more complex models [@Marescot2013] typically used in fisheries stock assessments.
+For example, Management Strategy Evaluation [MSE; @Punt2016] in fisheries seeks to evaluate pre-specified strategies rather solve for the best possible strategy through optimal control.
+That approach is well justified -- optimal control techniques such as stochastic dynamic programming illustrated here quickly become intractable for more complex models [@Marescot2013] typically used in fisheries stock assessments.
 Constraining the search to only strategies that impose a constant mortality (defined as harvest per unit biomass, $F = H/B$) means we only have to evaluate those $N$ strategies each iteration, not solve a sequential decision problem.
-Doing so, we would find the best constant-mortality solution under Model 1 performs much worse than the best constant-mortality solution under Model 2.
+Doing so, we would find the best constant-mortality solution under model 1 performs much worse than the best constant-mortality solution under model 2.
 Does this mean MSE is not susceptible to this issue?  
 No, it does not.
-It is just as easy to construct an alternative Model 1 with the same properties of leading to nearly optimal decisions while being rejected by any method of iterative learning, such as forecasting or adaptive management.
+It is just as easy to construct an alternative model 1 with the same properties of leading to nearly optimal decisions while being rejected by any method of iterative learning, such as forecasting or adaptive management.
 Just as the optimal control policy for a Gordon-Schaefer (logistic) model depends only on parameter $K$ [@Reed1979], the optimum constant escapement policy depends only on parameter $r$ [@Schaefer1954].
-If Model 1 has a value of $r$ such that it happens to match the best possible constant-mortality solution for the (unobserved) true model, then $K$ can be set arbitrarily high, ensuring the any model selection, forecasting, or adaptive management approach would lead away from Model 1 and towards worse-performing models once again.
-Many optimization problems share this feature in which the optimal policy depends only on a subset or ratio of model parameters, such that it is usually easy to find a similar Model 1.
+If model 1 has a value of $r$ such that it happens to match the best possible constant-mortality solution for the (unobserved) true model, then $K$ can be set arbitrarily high, ensuring the any model selection, forecasting, or adaptive management approach would lead away from model 1 and towards worse-performing models once again.
+Many optimization problems share this feature in which the optimal policy depends only on a subset or ratio of model parameters, such that it is usually easy to find a similar model 1.
 This example also illustrates the importance of decision constraints -- a model that gives very good outcomes when we are free to vary the havest quota each year (the optimal control problem here) may give very poor results under the constraint of constant mortality, and vice versa.
@@
 <!-- Not just "get more models" -->
-This case cannot be dismissed merely as being dealt a bad hand in having to pick only between Model 1 and Model 2.
+This case cannot be dismissed merely as being dealt a bad hand in having to pick only between model 1 and model 2.
 Model 1 approximates the key feature, giving nearly optimal outcomes.
 Previous literature has underscored the importance of Knightian "unknown unknowns" two alternative models examined here fail to fully reflect our uncertainty in the underlying dynamics, and the consequences of underestimating model uncertainty are well understood [@Polasky2011; @Wintle2010].
-Had we included the true model in the set of possibilities, the techniques illustrated would have had little difficulty in distinguishing it from Models 1 and 2 after sufficient iterations.
-In practice, we never have the true model, and even a much more thorough suite of candidate models, such as considering any possible values of $r$, $K$ [Fig 3B], or even among common alternative models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980], would fail to do any better, since all these models still have symmetric growth rates with a peak growth rate at half the steady-state population size).
+Had we included the true model in the set of possibilities, the techniques illustrated would have had little difficulty in distinguishing it from models 1 and 2 after sufficient iterations.
+In practice, we never have the true model.
+We have also seen that including a much larger suite of possible values of $r$, $K$ [Fig 3B] still leaves us far from the optimal performance of model 1 alone.
+Would a more comprehensive grid of $r$ and $K$ values, or the consideration common alternative models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980], succeed in closing that gap?
+No it would not, as the comparison to the true model makes clear [Fig 4].
+Regardless of parameter value, all of these models still have symmetric growth rates with a peak growth rate at half their steady-state population size.
+By contrast, the true model has a peak growth rate much closer to the steady-state size.
+Because the decision in this problem is driven by the location of the peak growth rate [@Reed1979], any symmetric model which places its peak at the correct stock size also overestimates the carrying capacity, as is the case with our Model 1.
+If such additional structurally different models are not enough, then perhaps more complex models would serve?
 Models will always be simpler than reality.
@@
 
-Previous work has acknowledged that iterative learning may not always be feasible and that it may not always be beneficial, but has failed to recognize the possibility that it can potentially be detrimental.
+The potential for long-term deterioration in management outcomes due to adaptive learning has not been widely acknowledged before.
+Certainly the limitations of any modeling approach to improve management outcomes due to the human context of competing interests, imperfect enforcement, etc., is readily appreciated [@Ludwig1993].
+The problem here goes deeper, to the limits of models and learning from data in even in a toy example free from other real world challenges.
+Previous work has acknowledged that iterative learning may not always be feasible and that it may not always be beneficial [@Polasky2011], but has failed to recognize the possibility that it can potentially be detrimental.
 For example, active adaptive management is premised on the observation that future observations may be too uninformative to distinguish between alternatives [@Walters1978].
@@
 
+Worse, absent the counterfactual of non-adaptive management to compare to, this detioriation in outcomes is invisible. 
+All the information available to our manager reinforces the conviction that this deteriorated ecological state is in fact the natural, sustainably managed equilibrium of our fish population.
+By responding in any way to future observations to reduce model uncertainty, our manager is caught in a trap of self-fulfilling prophecy:
+believing ever more firmly in models that make more accurate predictions while driving decisions that prevent the fish stock from recovering.
 The only way to avoid this trap is to _forego learning_ with each new observation, to avoid iterative updates.
@@
 This raises several questions.
-Do such methods of learning and reducing model uncertainty ever lead us towards worse long-term outcomes in the real world?
-If so, How would we know, and what should we do about it?
+Do such methods of learning and reducing model uncertainty ever lead us towards worse long-term outcomes in the real world? If so, How would we know, and what should we do about it?
 
@@
 Do models which give nearly optimal performance while at the same time making wildly wrong predictions really exist?
-It is tempting to argue that Model 1 is a kind of unicorn, a mythical creature existing only in theory.
-But that is a difficult assertion to prove.
-Examples of serious forecasts that widely fail to predict future observations abound wherever forecasts are common, from elections to economics to environmental change [e.g.
-@Tetlock2015].
-If there is no shortage of bad forecasts, then could any of them be useful?
-The premise that model need not perfectly capture reality to be useful is at the very heart of modeling.
+It is tempting to argue that model 1 is a kind of unicorn, a mythical creature existing only in theory. But that is a difficult assertion to prove.
+Examples of serious forecasts that widely fail to predict future observations abound wherever forecasts are common, from elections to economics to environmental change [e.g. @Tetlock2015].
+If there is no shortage of bad forecasts, then could any of them be useful?  The premise that model need not perfectly capture reality to be useful is at the very heart of modeling. 
 Model 1 successfully captures the one key feature driving decisions in optimal harvest control problems: the stock size at which the maximum growth rate occurs [Fig 4A].
@@
 We have also seen how not updating our uncertainty estimates, or doing so less frequently, can reveal a unicorn model before it is discounted by further iterations.
-Approaches such as iterative forecasting or adaptive management that can reduce model uncertainty over time remain promising and important techniques, but because value of information can be negative, we must revise model uncertainty only cautiously.
-More importantly, we have seen that  the intuition offered by decision theory [Fig 4] can help us better understand what features of a model are essential to decision outcomes and what are not.
+Approaches such as iterative forecasting or adaptive management that can reduce model uncertainty over time remain promising and important techniques, but because value of information can be negative, we must be careful not to update models without a keen understanding of what aspects of a model drive a decision [Fig 4].
+We have seen that  the intuition offered by decision theory [Fig 4] can help us better understand what features of a model are essential to decision outcomes and what are not.
 Conversely, models which appear to lead to accurate predictions (like Model 2) can result in outcomes that are far from optimal.
 Building on such understanding, it may possible to identify strategies for learning that are more agnostic to the details of the model that are not important, or perhaps not reliant on model-based predictions at all.
-How that is done is an open question.
+How that is best done in general is an open question.
 

### 2021-04-02 manuscript/manuscript.Rmd
:pen:

@@
 ```
+
 ```{r, include=FALSE}
@@
 knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
-                      warning = FALSE, dev="cairo_pdf")
+                      warning = FALSE, dev="cairo_pdf", fig.height=5, fig.width=7.5)
 ```
 
-<!-- I think the introduction could have a clearer storyline and better introduce some of the ideas you bring up later on in the methods/results. A quick outline idea:
-
-  - Paragraph1: Model-based forecasts have become a cornerstone of applied science in disciplines as diverse as atmospheric science [weather?], economics [stock markets?], and ecology [dietz].
-Seeking to inform a variety of decisions, near-term forecasting in these diverse fields has  simultaneously been encouraged by policy-makers [CITE?] and enabled by the rapid and recent increase in data availability and computational methods leading to a rapid expansion of thier use (or something...)
-
-   - Paragraph 2: Introduce ecological decision making under uncertainty, rapid global change, etc. (kind of like your current opening bit)
-
-  - Paragraph 3: But do better forecast always lead to better decisions?
-Because model-based forecasts frequently play an important role in decision-making,it is commonly assumed [e.g.
-@Walters1981; @Clark2001; @Dietze2018] that this approach for addressing model uncertainty will also improve decision-making outcomes.
-            - While the literature has largely addressed issues of overfittting that might lead forecasts to misguide decisions, proper scoring methods blah blah [...introduce this concept]. 
-            - Beyond overfitted models,... [maybe add something about Value of Information literature here?]
-
-  - Paragraph 4: Your current paragraph on "I illustrate..." (I'd add into that paragraph something about the fact that you explore adaptive management?)
-
-A second idea on framing: this maybe is a to divergent from the main point (or leads you into a post-modernist narrative that you'd prefer to avoid...) but it seems like you could reframe the whole thing as an argument for embracing an inclusive process of resource management that acknowledges pluralism.
-  (1) Policy makers often think "scientists" hold knowledge of "an environmental reality"
-  (2) Despite this not being true (scientist disagree on a lot of things), even when "science" points to a more likely reality, that might be off in ways that can fundementally misguide decisions. 
-  (3) The problem is deeper than acknowledging a "rashamon set" of models or multiple possible descriptions of of reality.
-Sometimes what appears closer to reality is actually off in a way that is critical to the decision.
--->
-
-<!-- Open on real-world issues, uncertainty, particularly model uncertainty as key theme.  -->
 Global change issues are complex and outcomes are difficult to predict  [@Clark2001].
-To guide decisions in an uncertainty world, researchers and decision makers 
-may consider a range of alternative plausible models to better reflect 
-what we do and do not know about the processes involved [@Polasky2011].  Forecasts or
-predictions from possible models can indicate what outcomes are most likely to result
-under what decisions or actions. This has made model-based forecasts a cornerstone 
-for scientifically based decision making.  
-By comparing outcomes predicted by a model to future observations, 
-a decision maker can not only _plan_ for the uncertainty, but also _learn_ which models are most trustworthy.  The value of iterative learning has long been reflected in the theory of 
-adaptive management [@Walters1978] as well as in actual adaptive management practices such as 
-Management Strategy Evaluation (MSE) [@Punt2016] used in fisheries, and is a central tenant
-of a rapidly growing interest in ecological forecasting [@Dietze2018].
+To guide decisions in an uncertainty world, researchers and decision makers may consider a range of alternative plausible models to better reflect what we do and do not know about the processes involved [@Polasky2011].
+Forecasts or predictions from possible models can indicate what outcomes are most likely to result under what decisions or actions.
+This has made model-based forecasts a cornerstone for scientifically based decision making.
+By comparing outcomes predicted by a model to future observations, a decision maker can not only _plan_ for the uncertainty, but also _learn_ which models are most trustworthy. 
+The value of iterative learning has long been reflected in the theory of adaptive management [@Walters1978] as well as in actual adaptive management practices such as Management Strategy Evaluation (MSE) [@Punt2016] used in fisheries, and is a central tenant of a rapidly growing interest in ecological forecasting [@Dietze2018].
 But, do iterative learning approaches always lead to better decisions?
@@
 the model(s) which the decision-makers have learned to favor lead to decisions and outcomes that closely match model expectations.
-Consequently, the decision-maker can become ever more firmly entrenched in the belief that their new models are accurate and the mediocre outcomes they experience are as good as it gets,
-while confidently rejecting earlier models which would have led to much better outcomes.  
-As this example will show, by better understanding
-what aspects of model are most and least important to shaping a particular decision, 
-we can at least anticipate the circumstances that make this problem more likely.
+Consequently, the decision-maker can become ever more firmly entrenched in the belief that their new models are accurate and the mediocre outcomes they experience are as good as it gets, while confidently rejecting earlier models which would have led to much better outcomes.
+As this example will show, by better understanding what aspects of model are most and least important to shaping a particular decision, we can at least anticipate the circumstances that make this problem more likely.
 
-Iterative learning approaches such as adaptive management or iterative forecasting are
-particularly compelling because they address a different reason why seemingly-accurate
-models can lead to poor outcomes: the problem of overfitting.  When models are estimated
-from the data they seek to describe, the model may over-fit the data, making the data
-seem even more likely than it would be under the true process [@Burnham1998].
-Over-fit models may underestimate uncertainty and make poor predictions about future
-outcomes [@Ginzburg2004]. Because any iterative learning compares models to future outcomes,
-overfitting those outcomes is impossible.  This concept was rigorously formalized
-by @Gneiting2007's proof of "proper" scoring rules. By definition, under a proper 
-scoring rule, no probabilistic prediction $Q(x)$ can score better, on
-average, than that of the underlying process, $P(x)$.  @Gneiting2007 further shows which
-techniques for evaluating forecast predictions do an do not satisfy this condition, i.e. which
-cannot be over-fit. Adaptive management is based on equally secure footing because
-model predictions are again compared to future outcomes not used to fit the model.
-Instead of scoring and selecting among alternative models, adaptive management 
-considers probabilities over all models [@Walters1978].  These two approaches are
-particularly compelling because they can both reduce uncertainty and avoid over-fitting.
-My example is not meant to undercut the importance of these approaches.  Previous
-work has long acknowledged the panelope of ways in which iterative learning or other
-model-based decision making can go astray due to conflicting incentives,
-implementation errors, or lack of resources for monitoring and updating [e.g. @Ludwig1993].
+Iterative learning approaches such as adaptive management or iterative forecasting [@Dietze2018] are particularly compelling because they address a different reason why seemingly-accurate models can lead to poor outcomes: the problem of over-fitting.
+When models are estimated from the data they seek to describe, the model may over-fit the data, making the data seem even more likely than it would be under the true process [@Burnham1998].
+Over-fit models may underestimate uncertainty and make poor predictions about future outcomes [@Ginzburg2004].
+Because any iterative learning compares models to future outcomes, overfitting those outcomes is impossible.
+This concept was rigorously formalized by @Gneiting2007's proof of "proper" scoring rules.
+By definition, under a proper scoring rule, no probabilistic prediction $Q(x)$ can score better, on average, than that of the underlying process, $P(x)$.
+@Gneiting2007 further shows which techniques for evaluating forecast predictions do an do not satisfy this condition, i.e. which cannot be over-fit.
+Adaptive management is based on equally secure footing because model predictions are again compared to future outcomes not used to fit the model.
+Instead of scoring and selecting among alternative models, adaptive management considers probabilities over all models [@Walters1978].
+These two approaches are particularly compelling because they can both reduce uncertainty and avoid over-fitting.
+My example is not meant to undercut the importance of these approaches.
+Previous work has long acknowledged the panelope of ways in which iterative learning or other model-based decision making can go astray due to conflicting incentives implementation errors, or lack of resources for monitoring and updating [e.g.
+@Ludwig1993].
 Here, the problem is more fundamental.
-
-<!-- Haven't really queued up the idea that I will present two distinct approaches that both fail.  need to more explicitly introduce and justify the two approaches-->
-
-<!--
-
-A wide range of paradigms are available for approaching the issue of decision-making under uncertainty.
-These approaches can roughly be divided into two groups: statistical approaches and adaptive management approaches.
-The first group, "statistical approaches," seeks to maximize some statistical metric (such as goodness-of-fit or predictive accuracy) in determining which model(s) will drive decisions.
-In contrast, adaptive management approaches seek to select actions which maximize the decision-maker's expected net utility over
-
-In the first group, I focus on evaluation using "proper" scoring rules as defined by @Gneiting2007, rather than potentially more familiar model comparisons such as $r^2$, likelihood, or information criteria [@Burnham1998] because proper scoring rules are more robust  (see @Gneiting2007 for comparison and how the approaches relate under appropriate assumptions).
-@Gneiting2007 provides a rigorous proof of scoring rules that have the desirable property of being "proper", that is, rules for which no model predicting the distribution of future outcomes, $Q(x)$ can achieve a better average score than the true model $P(x)$.
-Unlike likelihood or other goodness-of-fit scores alone, it is impossible (in expectation) to overfit the proper score -- no model model can out-perform the true model.
-Note that strictly proper scoring rules score _probabilistic forecasts_ and not just point predictions, favoring models which accurately reflect the uncertainty over those which under-estimate it.
-These features have made proper scoring rules for probabilistic forecasts a successful and popular approach for addressing model uncertainty in many other areas [@Gneiting2014; @Raftery2016] and a promising tool for evaluation of ecological forecasts [@Dietze2018].
-
-For the second group, decision-theoretic approaches, I focus on the example of adaptive management
-Decision-theoretic approaches include scenario analysis, resilience thinking, optimal control and related methods [@Polasky2011], which have a long history in ecology and particularly in fisheries management.
-One of the most influential of these is _adaptive management_ [sensu @Walters1978], in which a manager seeks to both reduce uncertainty over time while also achieving the best outcomes given current knowledge.
-While the term today is frequently used in a looser sense, adaptive management as originally developed can be considered an example of optimal control under model uncertainty [@Walters1981; @Smith1981].
-In this approach, a manager assigns probabilities to each of the possible models under consideration, i.e.
-models 1 & 2.
-The expected utility of any action reflects both the intrinsic uncertainty over future states under either model (process uncertainty) and the manager's uncertainty over the choice of models (or parameter values).
-Unlike the forecast comparison, an adaptive management approach explicitly considers the utility function, Eqn \eqref{utility}, which it seeks to maximize by integrating over model uncertainty.
-After observing the consequences of the action, the manager updates the posterior probabilities that each model is correct, typically by Bayes rule [e.g.
-@Walters1981; @Smith1981; @Ludwig1982; @Punt2016].
-The decision strategy is then re-evaluated in light of this revised uncertainty.
-Adaptive management may be either passive or active.
-Under passive adaptive management, the decision-maker chooses the action which maximizes that expected future utility.
-Under active adaptive management, the decision-maker may chose 'exploratory' actions which are not optimal under the current uncertainty but may lead to more rapidly decreasing the uncertainty in the future [@Walters1978].
-In practice, many managers are reluctant to employ exploratory strategies, so passive adaptive management approaches such as as Management Strategy Evaluation (MSE) are more common [@Punt2016].
-
--->
 
@@
 
-```{r figure1, fig.width=7, fig.height=5, fig.cap = "Forecast performance of each model.
+```{r figure1, fig.cap = "Forecast performance of each model.
 Panels A, B: Step ahead predictions of stock size under unfished (A) and fished (B) scenarios.
@@
 To evaluate two-step-ahead predictions we must consider each possible action under each possible state predicted by the step-ahead forecast, weighted by probability of the state given the model and the probability of the model.
-This rapidly expanding set of possibilities is addressed by adaptive management for sequential decision problems [e.g.
-@Smith1981], which I employ in the next section.
-
-<!-- replacing step-ahead with n-step ahead would not matter-->
+This rapidly expanding set of possibilities is addressed by adaptive management for sequential decision problems [e.g. @Smith1981], which I employ in the next section.
 
 ## Decision-Theoretic Approaches
-
-<!--
-Any of these adaptive management approaches proves to be maladaptive under the simple example of model uncertainty considered here.
-That is, the manager derives lower utility (here, economic yield), and the fish stock is also surpressed to a lower biomass than would be achieved if an identical management approach without any adaptive updating to model uncertainty were employed.
-Put another way, the "value of information" gained from future observations of stock in response to any management action is actually negative.
-
-Owing to the greater complexity involved in many fisheries models, a typical MSE would evaluate a fixed set of possible strategies rather than solve a sequential decision problem [@Marescot2013] for the optimal sequence of actions, but the central premise of iteratively updating model posterior probabilities remains the same.
-By limiting ourselves to the simpler models considered here, we are able to perform a passive adaptive management analysis of the full sequential decision problem, analgous to earlier examples [@Walters1981; @Smith1981; @Ludwig1982], which ensures the result is not an artifact of considering too narrow a set of possible policies (e.g.
-only constant-mortality policies).
-
--->
 
@@
 
-```{r figure3, fig.height=5, fig.width=8, fig.cap="Adaptive management under model uncertainty.
+```{r figure3, fig.cap="Adaptive management under model uncertainty.
 Solid lines trace the trajectories of the state (fish stock, circles) and action (harvest quota, triangles), under adaptive management (learning).
@@
 
-<!-- 
-First: Intuition for the result.
-  - See that model 1 captures the key feature, while model 2 is just 'closer to the curve'
-  - See that this is a generic truth about decision-making: there's always a Model 1
-  - This is not about over-fitting.
--->
-
 <!-- quick summary of what was demonstrated -->
@@
 Without access to the true model, none of our methods for model choice or reducing model uncertainty can break this trap.
+This should not be mistaken as a critique of adaptive management or iterative forecasting specifically -- other model choice approaches such as goodness-of-fit, information criteria or cross-validation would all prefer model 2 as well.
 
@@
 Note that while Model 2 is a better approximation to the truth (Model 3), Model 1 better approximates the stock size which leads to maximum growth.
-Panel B: The optimal control policy under Model 1 is nearly identical to that under the true Model 3, while the optimal policy under Model 2 supresses stock to a much lower escapement level.", fig.width=8}
+Panel B: The optimal control policy under Model 1 is nearly identical to that under the true Model 3, while the optimal policy under Model 2 supresses stock to a much lower escapement level."}
 # Fig 4
@@
 To insist that this issue can and will be avoided by always including more and better models of the process misses the point.
-
-<!--
-So what?
-- This is different than previous work: e.g. VOI is not zero, it's negative here.
-- Model 1 is not a unicorn: we cannot rule out these situations in real world (particularly ecology where we don't actually do much iterative updating in practice)
-- 
--->
 
@@
 
-<!-- Model free management: what would it look like? -->
-<!-- For example, a simple approach could gradually increase harvest from 0 as long as the observed recruitment increases relative to previous values.
-This corresponds to a local search for the peak growth-rate from above.
-This strategy is optimal for the model considered here, at least under small noise, but is not robust to growth rates with local optima, nor is it optimal during a possibly-very-long transient period of the search.
--->
-
-<!-- Correlary: Neither accurate forecasting nor effective outcomes are sufficient grounds to assert model validity. (meh, that's just 'all models are wrong, some are useful' for certain things)   -->
-
 ## Acknowledgements
@@
 
-<!--
-Is this result merely an artifact of differing objectives?
-@Gneiting2007's proper scoring rules are designed to select models which make the best forecast, and Model 2 really is the best choice for that task.
-However, the adaptive management approach explicitly seeks to maximize expected utility, integrating over the uncertainty in the model, and yet it evolves away from Model 1 which would achieve that goal and settles on Model 2.
-Note that this issue is not resolved by updating the learning step to reflect only the immediate reward -- it is in fact easy to see that so-called greedy algorithms would take the largest harvest possible and never reach the sustainable yield achieved by lower harvest rates of Model 1.
-To select Model 1 over alternatives like Model 2 which make better predictions but lead to worse outcomes requires a more long-term approach that forgoes iterative updating of probabilities as new data becomes available.
-Note that this is the opposite recommendation of _adaptive_ management [@Walters1978] or _iterative_ forecasting [@Dietze2018].
--->
-
-<!--
-
-The open problem is simply to propose an approach to model uncertainty which would select model 1
-over model 2 (i.e. choosing among the suite of available models, the one that gives the better 
-decisions.)
-
--->
-
 \pagebreak

### 2021-04-02 manuscript/scratch.Rmd
:pen:

@@
+<!--
+
+A wide range of paradigms are available for approaching the issue of decision-making under uncertainty.
+These approaches can roughly be divided into two groups: statistical approaches and adaptive management approaches.
+The first group, "statistical approaches," seeks to maximize some statistical metric (such as goodness-of-fit or predictive accuracy) in determining which model(s) will drive decisions.
+In contrast, adaptive management approaches seek to select actions which maximize the decision-maker's expected net utility over
+
+In the first group, I focus on evaluation using "proper" scoring rules as defined by @Gneiting2007, rather than potentially more familiar model comparisons such as $r^2$, likelihood, or information criteria [@Burnham1998] because proper scoring rules are more robust  (see @Gneiting2007 for comparison and how the approaches relate under appropriate assumptions).
+@Gneiting2007 provides a rigorous proof of scoring rules that have the desirable property of being "proper", that is, rules for which no model predicting the distribution of future outcomes, $Q(x)$ can achieve a better average score than the true model $P(x)$.
+Unlike likelihood or other goodness-of-fit scores alone, it is impossible (in expectation) to overfit the proper score -- no model model can out-perform the true model.
+Note that strictly proper scoring rules score _probabilistic forecasts_ and not just point predictions, favoring models which accurately reflect the uncertainty over those which under-estimate it.
+These features have made proper scoring rules for probabilistic forecasts a successful and popular approach for addressing model uncertainty in many other areas [@Gneiting2014; @Raftery2016] and a promising tool for evaluation of ecological forecasts [@Dietze2018].
+
+For the second group, decision-theoretic approaches, I focus on the example of adaptive management
+Decision-theoretic approaches include scenario analysis, resilience thinking, optimal control and related methods [@Polasky2011], which have a long history in ecology and particularly in fisheries management.
+One of the most influential of these is _adaptive management_ [sensu @Walters1978], in which a manager seeks to both reduce uncertainty over time while also achieving the best outcomes given current knowledge.
+While the term today is frequently used in a looser sense, adaptive management as originally developed can be considered an example of optimal control under model uncertainty [@Walters1981; @Smith1981].
+In this approach, a manager assigns probabilities to each of the possible models under consideration, i.e.
+models 1 & 2.
+The expected utility of any action reflects both the intrinsic uncertainty over future states under either model (process uncertainty) and the manager's uncertainty over the choice of models (or parameter values).
+Unlike the forecast comparison, an adaptive management approach explicitly considers the utility function, Eqn \eqref{utility}, which it seeks to maximize by integrating over model uncertainty.
+After observing the consequences of the action, the manager updates the posterior probabilities that each model is correct, typically by Bayes rule [e.g.
+@Walters1981; @Smith1981; @Ludwig1982; @Punt2016].
+The decision strategy is then re-evaluated in light of this revised uncertainty.
+Adaptive management may be either passive or active.
+Under passive adaptive management, the decision-maker chooses the action which maximizes that expected future utility.
+Under active adaptive management, the decision-maker may chose 'exploratory' actions which are not optimal under the current uncertainty but may lead to more rapidly decreasing the uncertainty in the future [@Walters1978].
+In practice, many managers are reluctant to employ exploratory strategies, so passive adaptive management approaches such as as Management Strategy Evaluation (MSE) are more common [@Punt2016].
+
+-->
+
+<!--
+Any of these adaptive management approaches proves to be maladaptive under the simple example of model uncertainty considered here.
+That is, the manager derives lower utility (here, economic yield), and the fish stock is also surpressed to a lower biomass than would be achieved if an identical management approach without any adaptive updating to model uncertainty were employed.
+Put another way, the "value of information" gained from future observations of stock in response to any management action is actually negative.
+
+Owing to the greater complexity involved in many fisheries models, a typical MSE would evaluate a fixed set of possible strategies rather than solve a sequential decision problem [@Marescot2013] for the optimal sequence of actions, but the central premise of iteratively updating model posterior probabilities remains the same.
+By limiting ourselves to the simpler models considered here, we are able to perform a passive adaptive management analysis of the full sequential decision problem, analgous to earlier examples [@Walters1981; @Smith1981; @Ludwig1982], which ensures the result is not an artifact of considering too narrow a set of possible policies (e.g.
+only constant-mortality policies).
+
+-->
+
 ```{r prob_scores, message = FALSE}
@@
 -->
+
+<!--
+Is this result merely an artifact of differing objectives?
+@Gneiting2007's proper scoring rules are designed to select models which make the best forecast, and Model 2 really is the best choice for that task.
+However, the adaptive management approach explicitly seeks to maximize expected utility, integrating over the uncertainty in the model, and yet it evolves away from Model 1 which would achieve that goal and settles on Model 2.
+Note that this issue is not resolved by updating the learning step to reflect only the immediate reward -- it is in fact easy to see that so-called greedy algorithms would take the largest harvest possible and never reach the sustainable yield achieved by lower harvest rates of Model 1.
+To select Model 1 over alternatives like Model 2 which make better predictions but lead to worse outcomes requires a more long-term approach that forgoes iterative updating of probabilities as new data becomes available.
+Note that this is the opposite recommendation of _adaptive_ management [@Walters1978] or _iterative_ forecasting [@Dietze2018].
+-->
+
+<!--
+
+The open problem is simply to propose an approach to model uncertainty which would select model 1
+over model 2 (i.e. choosing among the suite of available models, the one that gives the better 
+decisions.)
+
+-->

### 2021-04-02 manuscript/appendix.Rmd
tidy :broom:

@@
 bibliography: refs.bib
+csl: csl/ecology-letters.csl
 #output:
@@
 
+header-includes:
+  - \usepackage[left]{lineno}
+  - \linenumbers
+
 ---
 
-## Optimal quotas as a Markov Decision Process (MDP) problem
+Because the methods applied here: Stochastic Dynamic Programming, Iterative Forecasts assessed with Proper Scoring Rules, or Adaptive Management, are not themselves novel, the reader is referred to previous reviews and text books which can provide a thorough introduction to methods, in particular, @Marescot2013 and @Smith1981.
+However, implementation of these approaches generally depends on numerical methods, for which even precise mathematical formulas or pseudo-code descriptions can be inadequate to specify unambiguously or to properly facilitate replication [@Barnes2010].
+As such, this appendix includes computer code in the R language, which can also be found in the corresponding R-Markdown document in the paper's GitHub repository, <https://github.com/cboettig/bad-forecast-good-decision>.
+While computer code can be more verbose and difficult to read than mathematical formulas presented here, it is also less ambiguous.
+Though R is already widely used among ecologists, the implementations concern mostly matrix algebra that can be translated to other languages, and the `MDPtoolbox` package used here is also implemented in MatLab [@Marescot2013; @MDPtoolbox].
+
+Below, I provide carefully annotated code necessary to completely reproduce all of the analysis presented in the main paper.
+This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving Markov Decision Processes (MDP) using stochastic dynamic programming functionality, `expm` for matrix exponentials [@expm], and a few custom MDP functions provided by our package, `mdplearning` [@mdplearning].
+We will also use `tidyverse`  packages for basic manipulation and plotting [@tidyverse].
+This file is also available as an RMarkdown document [@rmarkdown] at  <https://github.com/cboettig/bad-forecast-good-decision>.
+The code below generates the data tables required for each plot shown in the main paper.
+Data are stored as `.csv` tables to be accessible to any relevant software program.
+Plotting commands used in the main paper are embedded in the paper's RMarkdown file at <https://github.com/cboettig/bad-forecast-good-decision>.
+
+```{r setup, message=FALSE}
+library(tidyverse)
+library(MDPtoolbox)
+library(expm)
+# remotes::install_github("boettiger-lab/mdplearning")
+library(mdplearning)
+```
+
+# Optimal quotas as a Markov Decision Process (MDP) problem
 
@@
 
-# Ecological Models
+## Ecological Model Definitions
 
@@
 
-where $\xi_i(t)$ is a multiplicative log-normal noise term with mean 1 and log-standar-deviation $\sigma_i$.
+where $\xi_i(t)$ is a multiplicative log-normal noise term with mean 1 and log-standard-deviation $\sigma_i$.
 
@@
 with $r_3 = 0.002$, $K_3 = 10$ and $\sigma_3 = 0.05$.
-
-Because the methods applied here: Stochastic Dynamic Programming, Iterative Forecasts assessed with Proper Scoring Rules, or Adaptive Management, are not themselves novel, the reader is referred to previous reviews and text books which can provide a thorough introduction to methods, in particular, @Marescot2013 and @Smith1981.
-However, implementation of these approaches generally depends on numerical methods, for which even precise mathematical formulae or pseudo-code descriptions can be inadequate to specify unambiguously or to properly facilitate replication [@Barnes2010].
-As such, this appendix includes computer code in the R language, which can also be found in the corresponding R-Markdown document in the paper's GitHub repository, <https://github.com/cboettig/bad-forecast-good-decision>.
-While computer code can be more verbose and difficult to read than mathematical formulae presented here, it is also less ambiguous.
-Though R is already widely used among ecologists, the implementations concern mostly matrix algebra that can be translated to other languages, and the `MDPtoolbox` package used here is also implemented in Matlab [@Marescot2013; @MDPtoolbox].
-
-Below, I provide annotated code necessary to completely reproduce all of the analysis presented in the main paper.
-This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving Markov Decision Processes (MDP) using stochastic dynamic programming functionality, `expm` for matrix exponentials [@expm], and a few custom MDP functions provided by our package, `mdplearning` [@mdplearning].
-We will also use `tidyverse`  packages for basic manipulation and plotting [@tidyverse].
-This file is also available as an RMarkdown document [@rmarkdown] at  <https://github.com/cboettig/bad-forecast-good-decision>.
-The code below generates the data tables required for each plot shown in the main paper.
-Data are stored as `.csv` tables to be accessible to any relevant software program.
-Plotting commands used in the main paper are embedded in the paper's RMarkdown file at <https://github.com/cboettig/bad-forecast-good-decision>.
 
@@
 ```
-
-```{r setup, message=FALSE}
-library(tidyverse)
-library(MDPtoolbox)
-library(expm)
-# remotes::install_github("boettiger-lab/mdplearning")
-library(mdplearning)
-```
-
-# Model definitions
 
@@
 
-# Optimal control solutions
+## Optimal control solutions
 
@@
 based on the first two moments of the distribution to score the respective
-forecasts under both the unfished and actively managed scenario sfor each model:
+forecasts under both the un-fished and actively managed scenario for each model:
 
@@
 
-# Posteriors
+## Posteriors
 

### 2021-04-02 manuscript/manuscript.Rmd
tidy :broom:

@@
 
-## change to true to add optional line numbering
-lineno: true
-csl: ecology-letters.csl
+header-includes:
+  - \linenumbers
+
+csl: csl/ecology-letters.csl
 output: rticles::elsevier_article
@@
 Over-fit models may underestimate uncertainty and make poor predictions about future outcomes [@Ginzburg2004].
-Because any iterative learning compares models to future outcomes, overfitting those outcomes is impossible.
+Because any iterative learning compares models to future outcomes, over-fitting those outcomes is impossible.
 This concept was rigorously formalized by @Gneiting2007's proof of "proper" scoring rules.
@@
 My example is not meant to undercut the importance of these approaches.
-Previous work has long acknowledged the panelope of ways in which iterative learning or other model-based decision making can go astray due to conflicting incentives implementation errors, or lack of resources for monitoring and updating [e.g.
+Previous work has long acknowledged the panoply of ways in which iterative learning or other model-based decision making can go astray due to conflicting incentives implementation errors, or lack of resources for monitoring and updating [e.g.
 @Ludwig1993].
@@
 Increasing the space of possible models to cover a whole plausible range of parameters $r$ and $K$ does little to resolve this problem [Fig 3B]. 
-Iterative updates again quickly dismiss the parameter values assumed by model 1, though with more options to choose from, this probability is spead over a range of seemingly plausible candidate models instead of a single alternative model (see Appendix).
+Iterative updates again quickly dismiss the parameter values assumed by model 1, though with more options to choose from, this probability is spread over a range of seemingly plausible candidate models instead of a single alternative model (see Appendix).
 While the adaptive management of additional actions and observations slowly narrow this subset of plausible models, decisions based on this uncertainty prevent fish stocks from recovering fully, and realize lower harvests as a result.
@@
 Many optimization problems share this feature in which the optimal policy depends only on a subset or ratio of model parameters, such that it is usually easy to find a similar model 1.
-This example also illustrates the importance of decision constraints -- a model that gives very good outcomes when we are free to vary the havest quota each year (the optimal control problem here) may give very poor results under the constraint of constant mortality, and vice versa.
+This example also illustrates the importance of decision constraints -- a model that gives very good outcomes when we are free to vary the harvest quota each year (the optimal control problem here) may give very poor results under the constraint of constant mortality, and vice versa.
 
@@
 
-Worse, absent the counterfactual of non-adaptive management to compare to, this detioriation in outcomes is invisible. 
+Worse, absent the counter-factual of non-adaptive management to compare to, this deterioration in outcomes is invisible. 
 All the information available to our manager reinforces the conviction that this deteriorated ecological state is in fact the natural, sustainably managed equilibrium of our fish population.
@@
 For example, so-called greedy optimization techniques that favor actions with higher immediate reward do even worse in this context, since such an algorithm would obviously maximize the immediate harvest and therefore collapse the stock.
-Only by comparing the net utility derived from management under model 1 for many iterations to the net utility derived under model 2 after many iterations can we deterime that model 1 leads to better outcomes.
+Only by comparing the net utility derived from management under model 1 for many iterations to the net utility derived under model 2 after many iterations can we determine that model 1 leads to better outcomes.
 This raises several questions.

### 2021-04-02 manuscript/appendix.Rmd
:broom:

@@
 title: 'Appendix'
-date: "`r Sys.Date()`"
 author: Carl Boettiger

### 2021-04-06 manuscript/manuscript.Rmd
formatting

@@
   - \linenumbers
+  - \usepackage{endfloat}
 
@@
 layout: 3p
-journal: TBD
+journal: Ecology Letters
 ---
+
+- abbreviated running title: "The trap of self-fulfilling prophecies"
+- Authorship statement: _This is a single-author paper_
+- data accessibility statement: _All simulation data generated for analyses here, along with code required for the analysis, is available in the Zenodo Data archive, https://doi.org/10.5281/zenodo.4660621_
+- number of words in the abstract: 116
+- number of words in main text: 4998
+- number of cited references: 33
+- number of tables & figures: 4 Figures
+
+\pagebreak
 

### 2021-04-13 manuscript/manuscript.Rmd
revisions

@@
 
-In this paper, I will demonstrate that iterative learning or iterative forecasting can also be counter-productive to management outcomes:
+In this paper, I demonstrate that the model that makes the best prediction is not necessarily the model that makes the best policy.
+I show that iterative learning or iterative forecasting can be counter-productive to management outcomes:
 that in some cases, a manager would be better to ignore new information which could reduce uncertainty over available models.
@@
 @Clark1973; @Reed1979; @Walters1981; @Ludwig1982].
-While modern fisheries management frequently relies on complex models which may contain scores of parameters to reflect the specific age or stage structure of a specific fish stock [@ramlegacy2018; @ram], I will rely on simple, well-studied models which permit greater intuition and generalization [@Levins1966].
-Consistent with such previous work [@Schaefer1954; @Clark1973; @Reed1979; @Walters1981; @Ludwig1982; @Costello2016], let us consider the problem of determining the optimal harvest policy given a measurement of the current stock size.
-
-For illustrative purposes, we will focus on the simplest example of model uncertainty: two alternative models of equal complexity but differing in the estimated value of certain parameters.
-While some authors distinguish between model uncertainty and parameter uncertainty, I will refer collectively to any uncertainty that arises from our imperfect knowledge of the system as "model uncertainty," because the distinction is largely a consequence of notation rather something more intrinsic.
-For example, two structurally different models, $f(x) = a x$ and $f(x) = a x^2$ can be considered merely different parameterizations of $f(x) = a x^b$, or vice versa.
+While modern fisheries management frequently relies on complex models which may contain over 100 parameters to reflect the specific age or stage structure of a specific fish stock [@ramlegacy2018; @ram], fisheries management and research has always appreciated the central insights offered by simpler one-dimensional models which continue to underpin both the theory and practice.
+For example, the strategies of constant mortality [@Schaefer1954] and constant escapement [@Clark1973; @Reed1979] are built on the same class of models considered here, which also continue to underpin global analyses of stock-rebuilding [e.g.
+@Costello2016; @Memarzadeh2019] and the foundations of adaptive management [e.g.
+@Walters1978; @Smith1981; @Walters1981].
+While more complex models are often required for specific applications, such models both share features with and are guided by theory built on much more general, simpler models [@Levins1966],
+from the Maximum Sustainable Yield of @Schaefer1954 or the epidemiological reproductive number, $R_0$ of @Kermack1927.
 
@@
 <!-- Intuition for it's not just this example -->
-This phenomenon is not unique to sequential decision problems or optimal control solutions.
-For example, Management Strategy Evaluation [MSE; @Punt2016] in fisheries seeks to evaluate pre-specified strategies rather solve for the best possible strategy through optimal control.
-That approach is well justified -- optimal control techniques such as stochastic dynamic programming illustrated here quickly become intractable for more complex models [@Marescot2013] typically used in fisheries stock assessments.
-Constraining the search to only strategies that impose a constant mortality (defined as harvest per unit biomass, $F = H/B$) means we only have to evaluate those $N$ strategies each iteration, not solve a sequential decision problem.
-Doing so, we would find the best constant-mortality solution under model 1 performs much worse than the best constant-mortality solution under model 2.
-Does this mean MSE is not susceptible to this issue?
-No, it does not.
-It is just as easy to construct an alternative model 1 with the same properties of leading to nearly optimal decisions while being rejected by any method of iterative learning, such as forecasting or adaptive management.
-Just as the optimal control policy for a Gordon-Schaefer (logistic) model depends only on parameter $K$ [@Reed1979], the optimum constant escapement policy depends only on parameter $r$ [@Schaefer1954].
-If model 1 has a value of $r$ such that it happens to match the best possible constant-mortality solution for the (unobserved) true model, then $K$ can be set arbitrarily high, ensuring the any model selection, forecasting, or adaptive management approach would lead away from model 1 and towards worse-performing models once again.
-Many optimization problems share this feature in which the optimal policy depends only on a subset or ratio of model parameters, such that it is usually easy to find a similar model 1.
-This example also illustrates the importance of decision constraints -- a model that gives very good outcomes when we are free to vary the harvest quota each year (the optimal control problem here) may give very poor results under the constraint of constant mortality, and vice versa.
+Situations similar to this example, in which a model can lead to the optimal decision while producing a poor forecast, are likely to be widespread.
+For example, marine fisheries typically restrict a Management Strategy Evaluation [MSE; @Punt2016] to constant mortality, (defined as harvest per unit biomass, $F = H/B$).
+Constant mortality is technically only optimal in a deterministic model [@Clark1973]; but is easier to solve for than the SDP methods for constant escapement shown here,
+which in practice is used mostly in salmon fisheries and the resource economics literature [see @pomdp-intro, @Costello2016].
+While Model 1 used here will not lead to a good constant-mortality policy, it is just as easy to construct a model which gives optimal decisions and bad forecasts.
+In the example shown above, Model 1 achieves optimal results by choosing $K$ such that $K/2$ matches the peak growth rate under the true model [@Reed1979.
+Under constant mortality, we instead choose a logistic model with $r$ such that $r/4$ matches the optimal constant mortality under the true model [@Schaefer1954].
+Everything else about our choice of Model 1 can be far from the true model, and yet it will give optimal results, just as above.
+Examples in which decisions depend a subset or ratio of model parameters can be found everywhere.
+In @Dee2017, the optimal number of non-critical species to protect in order to conserve critical species is given by a ratio of model parameters.
+In the SIR model of @Kermack1927 the basic reproductive number that determines if an infection spreads is the ratio of contact rate over the recovery rate, $R_0 = \beta / \gamma$.
+In general, there will be many models that can lead to the same decision, and thus the possibility of a similar scenario to the one seen here.
+Likewise there will always be infinitely many models like Model 2, which can fit very close to the data [Fig 4] while leading to a very
+different decision policy.
 
@@
 Previous literature has underscored the importance of Knightian "unknown unknowns" two alternative models examined here fail to fully reflect our uncertainty in the underlying dynamics, and the consequences of underestimating model uncertainty are well understood [@Polasky2011; @Wintle2010].
-Had we included the true model in the set of possibilities, the techniques illustrated would have had little difficulty in distinguishing it from models 1 and 2 after sufficient iterations.
-In practice, we never have the true model.
 We have also seen that including a much larger suite of possible values of $r$, $K$ [Fig 3B] still leaves us far from the optimal performance of model 1 alone.
@@
 Because the decision in this problem is driven by the location of the peak growth rate [@Reed1979], any symmetric model which places its peak at the correct stock size also overestimates the carrying capacity, as is the case with our Model 1.
-If such additional structurally different models are not enough, then perhaps more complex models would serve?
-Models will always be simpler than reality.
-To insist that this issue can and will be avoided by always including more and better models of the process misses the point.
+If such additional structurally different models are not enough, then perhaps yet more complex models would serve?
+For this toy example, it would be possible to span all the models considered here, including our "true" model 3, using 
+non-parametric approaches such as Gaussian Process models [@Boettiger2015].
+Even this does not guarantee a solution, as iterative learning erodes belief in model 1, and prefers harvests that retain stock sizes near historic values where predictability is highest.
+Besides, reality will always be more complex; we will never have the true model.
 
@@
 I believe this is an open question, but that our first step must be to recognize it as such.
-We have seen that the problem cannot be resolved by more data, and is not the result of overfitting.
+We have seen that the problem cannot be resolved by more data, and is not the result of over-fitting.
 Nor is "creating more models" the answer: when we have a model that is good enough to get optimal results, we cannot always insist on more models.

### 2021-04-13 reviews/README.Rmd
revisions

---
output: github_document
---

Revision reply letters are encrypted on this repository.  The reply letter quotes the 
recommendations of the editors and reviewers, to which those individuals hold copyright
unless they grant express permission to reproduce their comments here.  Encrypting
the reply letter preserves this confidential correspondence in this otherwise-public repository.
The source `.Rmd` and resulting `.pdf` reply letter are then `gitignore`'d.

A template .Rmd of the reply letter, capturing my formatting settings, is included here.

Encrypt to my public key:

```{r setup}
library(gpg)
gpg::gpg_recv("3908E1CFD28B380C") # import, needed only once
```

```{r}
msg <- gpg_encrypt("review-reply.Rmd", receiver =  "3908E1CFD28B380C")
writeLines(msg, "review-reply.Rmd.gpg")
```

## decrypt

(decrypt locally offline where my private key is available from yubikey, `gpg -d review-reply.Rmd.gpg`)

### 2021-04-13 manuscript/manuscript.Rmd
sp/grammar

@@
 abstract: |
-  Encouraged by decision makers appetite for future information on topics ranging from elections to pandemics, and enabled by the explosion of data and computational methods, model based forecasts have garnered increasing influence on a breadth of decisions in modern society.
+  Encouraged by decision makers' appetite for future information on topics ranging from elections to pandemics, and enabled by the explosion of data and computational methods, model based forecasts have garnered increasing influence on a breadth of decisions in modern society.
 Using a classic example from fisheries management, I demonstrate that selecting the model or models that produce the most accurate and precise forecast can lead to decidedly worse outcomes.
@@
 Global change issues are complex and outcomes are difficult to predict  [@Clark2001].
-To guide decisions in an uncertainty world, researchers and decision makers may consider a range of alternative plausible models to better reflect what we do and do not know about the processes involved [@Polasky2011].
+To guide decisions in an uncertain world, researchers and decision makers may consider a range of alternative plausible models to better reflect what we do and do not know about the processes involved [@Polasky2011].
 Forecasts or predictions from possible models can indicate what outcomes are most likely to result under what decisions or actions.
@@
 By comparing outcomes predicted by a model to future observations, a decision maker can not only _plan_ for the uncertainty, but also _learn_ which models are most trustworthy. 
-The value of iterative learning has long been reflected in the theory of adaptive management [@Walters1978] as well as in actual adaptive management practices such as Management Strategy Evaluation (MSE) [@Punt2016] used in fisheries, and is a central tenant of a rapidly growing interest in ecological forecasting [@Dietze2018].
+The value of iterative learning has long been reflected in the theory of adaptive management [@Walters1978] as well as in actual adaptive management practices such as Management Strategy Evaluation (MSE) [@Punt2016] used in fisheries, and is a central tenet of a rapidly growing interest in ecological forecasting [@Dietze2018].
 But, do iterative learning approaches always lead to better decisions?
@@
 Consequently, the decision-maker can become ever more firmly entrenched in the belief that their new models are accurate and the mediocre outcomes they experience are as good as it gets, while confidently rejecting earlier models which would have led to much better outcomes.
-As this example will show, by better understanding what aspects of model are most and least important to shaping a particular decision, we can at least anticipate the circumstances that make this problem more likely.
+As this example will show, by better understanding what aspects of a model are most and least important to shaping a particular decision, we can at least anticipate the circumstances that make this problem more likely.
 
-Iterative learning approaches such as adaptive management or iterative forecasting [@Dietze2018] are particularly compelling because they address a different reason why seemingly-accurate models can lead to poor outcomes: the problem of over-fitting.
-When models are estimated from the data they seek to describe, the model may over-fit the data, making the data seem even more likely than it would be under the true process [@Burnham1998].
+Iterative learning approaches such as adaptive management or iterative forecasting [@Dietze2018] are particularly compelling because they address a different reason why seemingly-accurate models can lead to poor outcomes: the problem of overfitting.
+When models are estimated from the data they seek to describe, the model may overfit the data, making the data seem even more likely than it would be under the true process [@Burnham1998].
 Over-fit models may underestimate uncertainty and make poor predictions about future outcomes [@Ginzburg2004].
@@
 By definition, under a proper scoring rule, no probabilistic prediction $Q(x)$ can score better, on average, than that of the underlying process, $P(x)$.
-@Gneiting2007 further shows which techniques for evaluating forecast predictions do an do not satisfy this condition, i.e. which cannot be over-fit.
+@Gneiting2007 further shows which techniques for evaluating forecast predictions do and do not satisfy this condition, i.e. which cannot be over-fit.
 Adaptive management is based on equally secure footing because model predictions are again compared to future outcomes not used to fit the model.
@@
 My example is not meant to undercut the importance of these approaches.
-Previous work has long acknowledged the panoply of ways in which iterative learning or other model-based decision making can go astray due to conflicting incentives implementation errors, or lack of resources for monitoring and updating [e.g.
+Previous work has long acknowledged the panoply of ways in which iterative learning or other model-based decision making can go astray due to conflicting incentives, implementation errors, or lack of resources for monitoring and updating [e.g.
 @Ludwig1993].
@@
 I illustrate this problem using an example from fisheries management. 
-Fisheries are a significant economic and conservation concern world wide and their management remains an important debate [e.g. @Worm2006; @Worm2009; @Costello2016].
+Fisheries are a significant economic and conservation concern worldwide and their management remains an important debate [e.g. @Worm2006; @Worm2009; @Costello2016].
 Moreover, their management has been a proving grounds for theoretical and practical decision-making issues which are widely applicable in other areas of ecology and conservation [@Ludwig1993; @Lande1994], and one that has long wrestled with issues of uncertainty in the context of management decisions [e.g.
@@
 
-Let us assume $\xi_t(\sigma)$ represents log-normal random noise with mean of unity and log-standard-deviation $\sigma_i$ for each model.
+Let us assume $\xi_t(\sigma)$ represents log-normal random noise with a mean of unity and log-standard-deviation $\sigma_i$ for each model.
 Model 1 is given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, Model 2 by $r_2 = 0.5$, $K_2 = 10$, $\sigma = 0.075$ (in dimensionless units).
@@
 Selecting between model 1 and model 2 can thus be considered the simplest illustration of the model uncertainty problem.
-This is a subset of the more general problem of selecting model parameters, assuming a logistic growth, which itself is a subset of estimating the best structural from (e.g.
+This is a subset of the more general problem of selecting model parameters, assuming a logistic growth, which itself is a subset of estimating the best structural form (e.g.
 Ricker, Beverton-Holt, etc).
@@
 Like many decision problems, the task of setting a sustainable harvest quota appears to hinge on having an accurate forecast: if we can predict to what size the fish stock will increase next year, $X_t+1$, and we know the current stock, $X_t$, then we can sustainably harvest $X_{t+1} - X_t$ without decreasing the biomass over the long term.
-Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of over-fitting by comparing model predictions to later observations that were not used to estimate the model [@Gneiting2014].
+Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of overfitting by comparing model predictions to later observations that were not used to estimate the model [@Gneiting2014].
 
@@
   ylab("stock size")
+ 
+ 
  
@@
 I illustrate the application of a passive adaptive management strategy to this simple example, following classic examples for parameter [@Ludwig1982] or structural [@Smith1981] model uncertainty.
-Passive adaptive management for a simple sequential decision problem is straight forward to implement over a discrete set of states and actions using dynamic programming with iterative updates [@Smith1981, example code in Appendix].
+Passive adaptive management for a simple sequential decision problem is straightforward to implement over a discrete set of states and actions using dynamic programming with iterative updates [@Smith1981, example code in Appendix].
 To demonstrate that the behavior is not driven by failure to explore sufficiently, (which might be addressed by an active adaptive management), I will assign initial probability that model 2 is true at 1%.
@@
  Dotted lines trace the corresponding trajectories if iterative learning is omitted, leaving the prior belief fixed throughout the simulation (planning).
-Color indicates the belief that model 1 is correct (blue), with an intial prior belief of 99%.
+  Color indicates the belief that model 1 is correct (blue), with an initial prior belief of 99%.
  Panel A: Management over the two candidate models, Model 1 and Model 2.
 Within a single iteration of adaptive management, the belief over models switches from a prior belief that heavily favored model 1 to a posterior that favors model 2 with near certainty.
-Future iterations reinfornce the belief in model 2, resulting in both depressed harvests and low stock sizes (solid lines).
+ Future iterations reinforce the belief in model 2, resulting in both depressed harvests and low stock sizes (solid lines).
  If no iterative learning updates are performed, stock sizes and realized harvests (and thus economic profit) are both higher.
-Panel B: given 42 candidate models over a broad range of parameter values, adaptive management quickly reduces the probability on model 1, and substantially underperforms management without learning (dotted lines).
+Panel B: given 42 candidate models over a broad range of parameter values, adaptive management quickly reduces the probability of model 1, and substantially underperforms management without learning (dotted lines).
 While outcomes improve marginally relative to the two-model case (figure A) they remain significantly worse than had no iterative learning been included."}
@@
 Note that while Model 2 is a better approximation to the truth (Model 3), Model 1 better approximates the stock size which leads to maximum growth.
-Panel B: The optimal control policy under Model 1 is nearly identical to that under the true Model 3, while the optimal policy under Model 2 supresses stock to a much lower escapement level."}
+ Panel B: The optimal control policy under Model 1 is nearly identical to that under the true Model 3, while the optimal policy under Model 2 suppresses stock to a much lower escapement level."}
 # Fig 4
@@
 Note that the mechanism shown here has nothing to do with the much more familiar issue of over-fitting, in which a better-fitting model will also lead to worse outcomes [@Burnham1998].
-In fact, an model which has been over-fit will gradually be rejected by either the iterative forecasting or adaptive management approaches shown here, as these approaches continually confront the models with new data to which they had not been previously fit.
-The ability to avoid over-fitting is one of the greatest appeals of any iterative management strategy.
+In fact, a model which has been over-fit will gradually be rejected by either the iterative forecasting or adaptive management approaches shown here, as these approaches continually confront the models with new data to which they had not been previously fit.
+The ability to avoid overfitting is one of the greatest appeals of any iterative management strategy.
 In the example here, both Model 1 and Model 2 have the same structural complexity.
-The better decision performance of Model 1 is a consequence of capturing the key attribute needed for a good decision, which is perhaps a surprisingly an all-together different criteria than fit.
+The better decision performance of Model 1 is a consequence of capturing the key attribute needed for a good decision, which is perhaps surprisingly an all-together different criteria than fit.
 Note that other methods of model selection not considered here, including goodness-of-fit metrics such as $r^2$  or information criteria [@Burnham1998], will all prefer Model 2 over Model 1 for the same reason.
@@
 Certainly the limitations of any modeling approach to improve management outcomes due to the human context of competing interests, imperfect enforcement, etc., is readily appreciated [@Ludwig1993].
-The problem here goes deeper, to the limits of models and learning from data in even in a toy example free from other real world challenges.
+The problem here goes deeper, to the limits of models and learning from data even in a toy example free from other real world challenges.
 Previous work has acknowledged that iterative learning may not always be feasible and that it may not always be beneficial [@Polasky2011], but has failed to recognize the possibility that it can potentially be detrimental.
@@
 Others have also noted that even big improvements in predictive accuracy could have negligible improvement on the decision outcomes, which motivates quantifying the Value of Information (VOI) [@Katz1987].
-But in both cases, the worse outcome is wasted effort.
+But in both cases, the worst outcome is wasted effort.
 In the scenario considered here, the VOI is negative: any learning over the proposed model uncertainty leads to lower expected net utility than not learning.
@@
 Examples of serious forecasts that widely fail to predict future observations abound wherever forecasts are common, from elections to economics to environmental change [e.g. @Tetlock2015].
-If there is no shortage of bad forecasts, then could any of them be useful?  The premise that model need not perfectly capture reality to be useful is at the very heart of modeling. 
+If there is no shortage of bad forecasts, then could any of them be useful? 
+The premise that a model need not perfectly capture reality to be useful is at the very heart of modeling.
 Model 1 successfully captures the one key feature driving decisions in optimal harvest control problems: the stock size at which the maximum growth rate occurs [Fig 4A].
@@
 This makes it more likely that such models could persist in practice until now. 
-While iterations that revise this models will no doubt improve decision outcomes in many case, it is worth bearing in mind from the example here that such a connection is not guaranteed.
-
+While iterations that revise these models will no doubt improve decision outcomes in many cases, it is worth bearing in mind from the example here that such a connection is not guaranteed.
 <!-- So what? -->

### 2021-04-13 manuscript/manuscript.Rmd
double-space char
no `nbsp;`
re-knit

@@
 - code: a
-  address: Department of Environmental Science, Policy, and Management, University of California, 130 Mulford Hall Berkeley, CA
-    94720-3114, USA
+  address: Department of Environmental Science, Policy, and Management, University of California, 130 Mulford Hall Berkeley, CA 94720-3114, USA
 
@@
   ylab("stock size")
- 
- 
 
@@
 
-The author acknowledges support from NSF CAREER Award #1942280 and helpful discussions with Melissa Chapman and Jeremy Fox.
+The author acknowledges support from NSF CAREER Award #1942280 and helpful discussions with Melissa Chapman, Jeremy Fox, and anonymous reviewers.
 

### 2021-04-15 manuscript/manuscript.Rmd
update refs
revise language regarding alternate models

@@
 The ability to avoid overfitting is one of the greatest appeals of any iterative management strategy.
-In the example here, both Model 1 and Model 2 have the same structural complexity.
-The better decision performance of Model 1 is a consequence of capturing the key attribute needed for a good decision, which is perhaps surprisingly an all-together different criteria than fit.
+In the example here, both model 1 and model 2 have the same structural complexity.
+The better decision performance of model 1 is a consequence of capturing the key attribute needed for a good decision, which is perhaps surprisingly an all-together different criteria than fit.
 Note that other methods of model selection not considered here, including goodness-of-fit metrics such as $r^2$ or information criteria [@Burnham1998], will all prefer Model 2 over Model 1 for the same reason.
 
-<!-- Not just "get more models" -->
-This case cannot be dismissed merely as being dealt a bad hand in having to pick only between model 1 and model 2.
-Model 1 approximates the key feature, giving nearly optimal outcomes.
-Previous literature has underscored the importance of Knightian "unknown unknowns" two alternative models examined here fail to fully reflect our uncertainty in the underlying dynamics, and the consequences of underestimating model uncertainty are well understood [@Polasky2011; @Wintle2010].
-We have also seen that including a much larger suite of possible values of $r$, $K$ [Fig 3B] still leaves us far from the optimal performance of model 1 alone.
-Would a more comprehensive grid of $r$ and $K$ values, or the consideration common alternative models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980], succeed in closing that gap?
-No it would not, as the comparison to the true model makes clear [Fig 4].
-Regardless of parameter value, all of these models still have symmetric growth rates with a peak growth rate at half their steady-state population size.
-By contrast, the true model has a peak growth rate much closer to the steady-state size.
-Because the decision in this problem is driven by the location of the peak growth rate [@Reed1979], any symmetric model which places its peak at the correct stock size also overestimates the carrying capacity, as is the case with our Model 1.
-If such additional structurally different models are not enough, then perhaps yet more complex models would serve?
-For this toy example, it would be possible to span all the models considered here, including our "true" model 3, using 
-non-parametric approaches such as Gaussian Process models [@Boettiger2015].
-Even this does not guarantee a solution, as iterative learning erodes belief in model 1, and prefers harvests that retain stock sizes near historic values where predictability is highest.
-Besides, reality will always be more complex; we will never have the true model.
+<!-- Not just "get more models"
+wrong!
 
-The potential for long-term deterioration in management outcomes due to adaptive learning has not been widely acknowledged before.
+-->
+
+This simple example can be resolved by including a model that closely approximates both the location of the peak and the shape of the curve of our "true" model.
+Given sufficient data, either adaptive management or iterative forecasting will prefer such a model (as guaranteed by @Gneiting2007's theorem), while having the correct peak location means the model will also lead to the optimal policy in this case (as guaranteed by @Reed1979's theorem).
+Yet the larger problem remains unresolved.
+In most realistic management scenarios, we do not have the benefit of a theorem similar to Reed's to identify what features are most critical to approximate well and what are not.
+For instance, the theorem no longer holds if stock estimates are uncertain [@pomdp-intro].
+Absent the guidance of such results, iterative learning over even relatively rich possible candidate models can move us towards worse outcomes than not learning at all.
+
+That potential for long-term deterioration in management outcomes due to adaptive learning has not been widely acknowledged before.
 Certainly the limitations of any modeling approach to improve management outcomes due to the human context of competing interests, imperfect enforcement, etc., is readily appreciated [@Ludwig1993].
@@
 This raises several questions.
-Do such methods of learning and reducing model uncertainty ever lead us towards worse long-term outcomes in the real world? If so, How would we know, and what should we do about it?
+Do such methods of learning and reducing model uncertainty ever lead us towards worse long-term outcomes in the real world? 
+If so, How would we know, and what should we do about it?
 
@@
 The premise that a model need not perfectly capture reality to be useful is at the very heart of modeling. 
-Model 1 successfully captures the one key feature driving decisions in optimal harvest control problems: the stock size at which the maximum growth rate occurs [Fig 4A].
-Capturing only the essential aspects as simply as possible is the goal of any model building exercise. 
-Thus it should be no surprise that a model can drive good decisions while making poor predictions.
 It is also important to note that in many cases, any models being used in decision-making are not necessarily being subjected to the rigorous evaluation and updating steps proposed by adaptive management [@Walters1978] or iterative forecasting [@Dietze2018].
 This makes it more likely that such models could persist in practice until now. 
-While iterations that revise these models will no doubt improve decision outcomes in many cases, it is worth bearing in mind from the example here that such a connection is not guaranteed.
+While iterations that revise these models will no doubt improve decision outcomes in many cases, it is worth bearing in mind from the example here that such improvement is not guaranteed.
+
 <!-- So what? -->
@@
 I believe this is an open question, but that our first step must be to recognize it as such.
-We have seen that the problem cannot be resolved by more data, and is not the result of over-fitting.
-Nor is "creating more models" the answer: when we have a model that is good enough to get optimal results, we cannot always insist on more models.
+We have seen that the problem cannot be resolved by more data alone, and is not the result of over-fitting.
+Introducing more models into our candidate set is only helpful to the extent that such models can capture features important to the decision, which as we have seen can be quite different from the feature which matter most to prediction.
 We have also seen how not updating our uncertainty estimates, or doing so less frequently, can reveal a unicorn model before it is discounted by further iterations.
-Approaches such as iterative forecasting or adaptive management that can reduce model uncertainty over time remain promising and important techniques, but because value of information can be negative, we must be careful not to update models without a keen understanding of what aspects of a model drive a decision [Fig 4].
-We have seen that the intuition offered by decision theory [Fig 4] can help us better understand what features of a model are essential to decision outcomes and what are not.
-Conversely, models which appear to lead to accurate predictions (like Model 2) can result in outcomes that are far from optimal.
-Building on such understanding, it may possible to identify strategies for learning that are more agnostic to the details of the model that are not important, or perhaps not reliant on model-based predictions at all.
+Approaches such as iterative forecasting or adaptive management that can reduce model uncertainty over time remain promising and important techniques, and essential strategies to avoid the very real problem of model over-fitting.
+But their very success in selecting models which give the most accurate predictions can also be trap.
 How that is best done in general is an open question.

### 2021-05-18 notebook/data-sources.Rmd
in search of a better example

---
title: "Data Sources"
output: github_document
---

```{r}
library(contentid)
library(tidyverse)
library(httr)
#library(taxalight)
```

## GPDD

John Prendergast, Ellen Bazeley-White, Owen Smith, John Lawton, Pablo Inchausti, et al. 2010. The Global Population Dynamics Database. Knowledge Network for Biocomplexity. doi:10.5063/F1BZ63Z8.

```{r}
taxon <- read_csv(resolve("hash://md5/c996a91042243648fa0c803ac631803b", store=TRUE))
main <- read_csv(resolve("hash://md5/b7c5d86a5a2c44d6a303f09be1a8688a", store=TRUE))
data_source <- read_csv(resolve("hash://md5/22bf3d98318860b4cf5ea15179d5442b", store=TRUE))
ts <- read_csv(resolve("hash://md5/8eb14acec7eccfd342f11860909ce588", store=TRUE), guess_max = 1e6)
```

```{r}
# Redlist Data, from API via public token
pages <- ceiling(137210/10000) -1
query <- paste0("https://apiv3.iucnredlist.org/api/v3/species/page/", 0:pages,
"?token=9bb4facb6d23f48efbf424bb05c0c1ef1cf6f468393bc745d42179ac4aca5fee")
resp <- map(query, httr::GET)
redlist_status <- map_dfr(resp, function(r){
  x <- content(r)
  jsonlite::fromJSON(jsonlite::toJSON(x$result, auto_unbox = TRUE), flatten=TRUE)
})
is_global <- map_lgl(redlist_status$population, function(x) length(x) == 0)
redlist_endangered <- redlist_status %>% filter(category %in% c("CR", "EN"), is_global)
```

```{r}
# main timeseries data for only those species where complete Census is available
endagered <- taxon %>% 
  inner_join(redlist_endangered, by = c(TaxonName = "scientific_name")) %>%
  select(TaxonName,CommonName, TaxonID)
endagered  <- main %>% inner_join(endagered)
census <- endagered %>% filter(SamplingProtocol == "Count", grepl("Census", SamplingEffort))
census

```

```{r}
selected_ts <- ts %>% inner_join(census)

selected_ts %>%
  mutate(MainID = as.character(MainID)) %>%
  ggplot(aes(SeriesStep, Population, col=MainID)) +
  geom_point() + facet_wrap(~CommonName, scales="free")

## Focus on two populations of Macaca sylvanus, Barbary macaque, introduced in Gilbratar
selected_ts %>% filter(MainID %in% c(5019,5020)) %>% write_csv("barbary_macaque.csv")
```

```{r}
# Examine details about these studies from the original publications
data_source %>%
  inner_join(select(census, 
                    DataSourceID, 
                    MainID, 
                    TaxonName, 
                    CommonName), 
             by = c(DatasourceID = "DataSourceID"))
```

```{r}
#Access Biotime data, BioTIME Consortium. (2018). BioTIME (Version 2) [Data set]. Zenodo. http://doi.org/10.5281/zenodo.3265871
#biotime_csv <- resolve("hash://md5/e1b4b7fca5091584831d43a798086496", store=TRUE)
#biotime <- read_csv(biotime_csv, guess_max = 1e6)

```

### 2021-05-18 notebook/model-estimates.Rmd
in search of a better example

--- 2021-05-18 notebook/data-sources.Rmd
@@
 
-## GPDD
+# May model fit
 
-John Prendergast, Ellen Bazeley-White, Owen Smith, John Lawton, Pablo Inchausti, et al. 2010. The Global Population Dynamics Database. Knowledge Network for Biocomplexity. doi:10.5063/F1BZ63Z8.
+# Logistic model fit
 
-```{r}
-taxon <- read_csv(resolve("hash://md5/c996a91042243648fa0c803ac631803b", store=TRUE))
-main <- read_csv(resolve("hash://md5/b7c5d86a5a2c44d6a303f09be1a8688a", store=TRUE))
-data_source <- read_csv(resolve("hash://md5/22bf3d98318860b4cf5ea15179d5442b", store=TRUE))
-ts <- read_csv(resolve("hash://md5/8eb14acec7eccfd342f11860909ce588", store=TRUE), guess_max = 1e6)
-```
+# Allen model fit
 
-```{r}
-# Redlist Data, from API via public token
-pages <- ceiling(137210/10000) -1
-query <- paste0("https://apiv3.iucnredlist.org/api/v3/species/page/", 0:pages,
-"?token=9bb4facb6d23f48efbf424bb05c0c1ef1cf6f468393bc745d42179ac4aca5fee")
-resp <- map(query, httr::GET)
-redlist_status <- map_dfr(resp, function(r){
-  x <- content(r)
-  jsonlite::fromJSON(jsonlite::toJSON(x$result, auto_unbox = TRUE), flatten=TRUE)
-})
-is_global <- map_lgl(redlist_status$population, function(x) length(x) == 0)
-redlist_endangered <- redlist_status %>% filter(category %in% c("CR", "EN"), is_global)
-```
-
-```{r}
-# main timeseries data for only those species where complete Census is available
-endagered <- taxon %>% 
-  inner_join(redlist_endangered, by = c(TaxonName = "scientific_name")) %>%
-  select(TaxonName,CommonName, TaxonID)
-endagered  <- main %>% inner_join(endagered)
-census <- endagered %>% filter(SamplingProtocol == "Count", grepl("Census", SamplingEffort))
-census
-
-```
-
-```{r}
-selected_ts <- ts %>% inner_join(census)
-
-selected_ts %>%
-  mutate(MainID = as.character(MainID)) %>%
-  ggplot(aes(SeriesStep, Population, col=MainID)) +
-  geom_point() + facet_wrap(~CommonName, scales="free")
-
-## Focus on two populations of Macaca sylvanus, Barbary macaque, introduced in Gilbratar
-selected_ts %>% filter(MainID %in% c(5019,5020)) %>% write_csv("barbary_macaque.csv")
-```
-
-```{r}
-# Examine details about these studies from the original publications
-data_source %>%
-  inner_join(select(census, 
-                    DataSourceID, 
-                    MainID, 
-                    TaxonName, 
-                    CommonName), 
-             by = c(DatasourceID = "DataSourceID"))
-```
-
-```{r}
-#Access Biotime data, BioTIME Consortium. (2018). BioTIME (Version 2) [Data set]. Zenodo. http://doi.org/10.5281/zenodo.3265871
-#biotime_csv <- resolve("hash://md5/e1b4b7fca5091584831d43a798086496", store=TRUE)
-#biotime <- read_csv(biotime_csv, guess_max = 1e6)
-
-```
+# Transition matrices from ARIMA?

### 2021-05-18 notebook/tipping-decisions.Rmd
in search of a better example

---
output: github_document
---

```{r setup, message=FALSE}
library(MDPtoolbox)
library(sarsop)
library(tidyverse) # for plotting
library(mdplearning) # remotes::install_github("boettiger-lab/mdplearning")
```

```{r}
states <- seq(0,2, length=100)
actions <- seq(0,2, length.out = 100)
observations <- states
sigma_g <- 0.05
sigma_m <- 0.0

benefit <- 1
cost <- 1 # quadratic costs
reward_fn <- function(x,h) benefit * x - (h ^ cost)
discount <- 0.999

p <- list(r =.7, K = 1.2, q = 3, b = 0.15, a = 0.2) # lower-state peak is optimal
#p <- list(r =.7, K = 1.5, q = 3, b = 0.15, a = 0.2) # higher-state peak is optimal

may <- 
  function(x, h = 0){ # May
    y <- x +  h
    pmax(
      ## controlling h is controlling the bifurcation point directly...
      y + y * p$r * (1 - y / p$K)  - p$a * y ^ p$q / (y ^ p$q + p$b ^ p$q),  
      0)
  }

# Matches the optimal policy under these settings (depends on sigma, among others)
logistic_1 <- 
  function(x, h = 0){
    y <- x +  h
    r <- p$r - 0.09

    pmax(
      y + y * r * (1 - y / p$K),  
      0)
  }

logistic_2 <- 
  function(x, h = 0){
    y <- x +  h
    r <- p$r + .5
    pmax(
      y + y * r * (1 - y / p$K),  
      0)
  }
```

```{r}

landscape <- map_dfr(list(logistic_1, logistic_2, may), 
                     function(f) tibble(state = states, f = f(states, 0) -states),
                     .id = "model")

landscape %>% ggplot(aes(state, f, col=model)) + geom_point()
```

## Optimal management

```{r}
matrices <- lapply(list(logistic_1, logistic_2, may), 
       function(f) fisheries_matrices(states, actions, observations, reward_fn, 
                        f, sigma_g, sigma_m, noise = "lognormal")
)
```

```{r, results="hide"}
solns <- lapply(matrices, function(m) mdp_value_iteration(m$transition, m$reward, discount))
```

```{r}
df <- map_dfr(solns, function(soln)
        tibble(state = states,
               action = actions[soln$policy],
               value = soln$V),
  .id = "model")

df
```

```{r}
df %>% ggplot(aes(state, action, col=model)) + geom_point(alpha=0.5)

#+
#  geom_line(aes(state, f), col = "red")  + 
#  coord_cartesian(xlim=c(0,1.3), ylim = c(-2, 3))
```

```{r}
x0 = which.min(abs(states - 0.05))
sim <- mdp_planning(m$transition, m$reward, discount, model_prior = c(1), 
                   policy = soln$policy, x0 = x0, Tmax = 100)
sim %>% mutate(state = states[state], action = actions[action]) %>% 
  ggplot(aes(time, state)) + geom_point()+geom_path() + 
  geom_point(aes(time, action), col="blue")
```

# Tipping

```{r}

p <- list(r =.7, K = 1.2, q = 3, b = 0.15, a = 0.2) # lower-state peak is optimal
#p <- list(r =.7, K = 1.5, q = 3, b = 0.15, a = 0.2) # higher-state peak is optimal

may <- function(a) 
  function(x, h = 0){ # May
    y <- x +  h
    pmax(
      ## controlling h is controlling the bifurcation point directly...
      y + y * p$r * (1 - y / p$K)  - a * y ^ p$q / (y ^ p$q + p$b ^ p$q),  
      0)
  }

df <- seq(0.1, 0.26, by = 0.001) %>%
  map_dfr(function(a) tibble(x = states, f = may(a)(x,0) - x, a = a))
df %>% 
  ggplot(aes(x, f)) + 
    geom_line(aes(group = a, col = a), lwd = 1) +
    geom_line(data=dplyr::filter(df, near(a, 0.215)), col = "purple", lwd = 1) +
    geom_hline(aes(yintercept = 0)) + coord_cartesian(ylim = c(-.2, .1))
```

### 2021-05-19 notebook/model-estimates.Rmd
:construction:

@@
 library(tidyverse)
-library(httr)
-#library(taxalight)
 ```
@@
 # May model fit
+
+---
+title: 'Appendix B'
+date: "`r Sys.Date()`"
+author: Carl Boettiger
+journal: Appendix
+layout: 3p
+bibliography: refs.bib
+#output:
+#  tufte::tufte_handout:
+#    dev: cairo_pdf
+#    latex_engine: xelatex
+output: 
+  hrbrthemes::ipsum_pdf:
+    dev: cairo_pdf
+    latex_engine: xelatex
+
+---
+
+```{r graphics_setup, message=FALSE, warning=FALSE, include = FALSE}
+## Plotting themes, colors, and fonts 
+## aesthetics choices only, all can be omitted without consequence
+library(ggthemes)
+library(hrbrthemes)
+library(ggplot2)
+library(Cairo)
+library(extrafont)
+library(patchwork)
+library(styler)
+extrafont::loadfonts(quiet = TRUE)
+ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
+
+scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
+scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
+pal <- ggthemes::solarized_pal()(8)
+txtcolor <- "#586e75"
+
+knitr::opts_chunk$set(cache=FALSE, tidy = "styler", message = FALSE, warning = FALSE, echo = FALSE)
+
+```
+
+```{r setup, message=FALSE}
+library(tidyverse)
+library(nimble)
+```
+
+```{r}
+siteA <- read_csv("barbary_macaque.csv") %>% 
+  select(SampleYear, PopulationUntransformed, LocationID) %>%
+  mutate(LocationID = as.character(LocationID)) %>% 
+  filter(LocationID == "414")
+
+scale <- 20
+x <- siteA$PopulationUntransformed / scale
+constants <- list(N = length(siteA$SampleYear),
+                  x0 = x[[1]]
+)
+```
+
+```{r}
+logistic  <- nimble::nimbleCode({
+  r ~ dunif(0,2)
+  K ~ dunif(0,10)
+  sigma ~ dunif(0,.4)
+  x[1] <- x0
+  for(t in 1:(N-1)){
+    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)
+    x[t+1] ~ dlnorm(log(mu[t]), sd = sigma)
+  }
+
+})
+
+```
+
+```{r}
+may  <- nimble::nimbleCode({
+  x[1] <- x0
+  q <- 3
+  s <- 2
+  r ~ dlnorm(log(.7) - s^2/2, s)
+  K ~ dlnorm(log(1.2)- s^2/2, s)
+  a ~ dlnorm(log(0.2) - s^2/2,  s)
+  b ~ dlnorm(log(0.15) - s^2/2, s)
+  sigma ~ dlnorm(log(0.05) - s^2/2, s)
+
+  for(t in 1:(N-1)){
+    mu[t] <-  x[t] + x[t] * r * (1 - x[t] / K)  - a * x[t] ^ q / (x[t] ^ q + b ^ q)
+    x[t+1] ~ dnorm(log( mu[t] ), sd = sigma)
+  }
+
+})
+
+```
+
+```{r}
+mcmc <- function(model, x, constants, niter = 100000, nburin = 1000){
+  r_model <- nimbleModel(model, constants = constants)
+  r_model$setData(list(x = x))
+  c_model <- compileNimble(r_model)
+  r_mcmc <- buildMCMC(c_model)
+  c_mcmc <- compileNimble(r_mcmc)
+  df <- runMCMC(c_mcmc, niter = niter, nburnin = nburin)
+  tibble::as_tibble(df)
+}
+```
+
+```{r}
+l_draws <- mcmc(logistic, x, constants)
+l_draws %>% summarise(across(everything(), list(mean = mean, sd = sd)))
+
+```
+```{r}
+bayesplot::mcmc_trace(l_draws)
+```
+
+```{r}
+m_draws <- mcmc(may, x, constants)
+m_draws %>% summarise(across(everything(), list(mean = mean, sd = sd)))
+bayesplot::mcmc_trace(m_draws)
+
+```
 

### 2021-05-19 notebook/tipping-decisions.Rmd
:construction:

@@
 ```{r}
+m <- matrices[[3]]
 x0 = which.min(abs(states - 0.05))
 sim <- mdp_planning(m$transition, m$reward, discount, model_prior = c(1), 
-                   policy = soln$policy, x0 = x0, Tmax = 100)
+                   policy = solns[[3]]$policy, x0 = x0, Tmax = 100)
 sim %>% mutate(state = states[state], action = actions[action]) %>% 
@@
   geom_point(aes(time, action), col="blue")
+
+```
+
+unmanaged
+
+```{r}
+nothing <- rep(1, length(states))
+m <- matrices[[3]]
+x0 = which.min(abs(states - 0.7))
+sim <- mdp_planning(m$transition, m$reward, discount, model_prior = c(1), 
+                   policy = nothing, x0 = x0, Tmax = 1000)
+sim %>% mutate(state = states[state], action = actions[action]) %>% 
+  ggplot(aes(time, state)) + geom_point()+geom_path() + 
+  geom_point(aes(time, action), col="blue")
+
 ```

### 2021-05-19 notebook/model-estimates.Rmd
...

@@
 
-scale <- 20
+#scale <- 20
 x <- siteA$PopulationUntransformed / scale
@@
 logistic  <- nimble::nimbleCode({
-  r ~ dunif(0,2)
-  K ~ dunif(0,10)
+  r ~ dunif(0,5)
+  K ~ dunif(0,30)
   sigma ~ dunif(0,.4)
@@
 ```{r}
+#p <- list(r = 2, K = 25, q = 4, b = 2, a = 7) # higher-state peak is optimal
+
 may  <- nimble::nimbleCode({
@@
   s <- 2
-  r ~ dlnorm(log(.7) - s^2/2, s)
-  K ~ dlnorm(log(1.2)- s^2/2, s)
-  a ~ dlnorm(log(0.2) - s^2/2,  s)
-  b ~ dlnorm(log(0.15) - s^2/2, s)
+  r ~ dunif(0,5) #dlnorm(log(1) - s^2/2, s)
+  K ~ dunif(0,40) #dlnorm(log(2)- s^2/2, s)
+  a ~ dunif(0,5) # dlnorm(log(7) - s^2/2,  s)
+  b ~ dunif(1.5,2.5) #dlnorm(log(1) - s^2/2, s)
   sigma ~ dlnorm(log(0.05) - s^2/2, s)
@@
 l_draws <- mcmc(logistic, x, constants)
-l_draws %>% summarise(across(everything(), list(mean = mean, sd = sd)))
-
-```
-```{r}
+l_p <- l_draws %>% summarise(across(everything(), list(mean = mean, sd = sd)))
 bayesplot::mcmc_trace(l_draws)
@@
 m_draws <- mcmc(may, x, constants)
-m_draws %>% summarise(across(everything(), list(mean = mean, sd = sd)))
+p <- m_draws %>% summarise(across(everything(), list(mean = mean, sd = sd)))
 bayesplot::mcmc_trace(m_draws)
 
+```
+
+```{r}
+
+may_f <- function(x){ # May
+      x + x * p$r_mean * (1 - x / p$K_mean)  - (p$a_mean) * x ^ 3 / (x ^ 3 + (p$b_mean) ^ 3)
+}
+
+# Matches the optimal policy under these settings (depends on sigma, among others)
+logistic_f <- 
+  function(x){
+      x + x * l_p$r_mean * (1 - x / l_p$K_mean)
+  }
+
+states <- seq(0,30, length=200)
+landscape <- map_dfr(list(may_f, logistic_f), 
+                     function(f) tibble(state = states, f = f(states) -states),
+                     .id = "model")
+
+landscape %>% ggplot(aes(state, f, col=model)) + geom_point()
 ```

### 2021-05-19 notebook/tipping-decisions.Rmd
...

@@
 ```{r}
-states <- seq(0,2, length=100)
-actions <- seq(0,2, length.out = 100)
+states <- seq(0,30, length=100)
+actions <- seq(0,30, length.out = 100)
 observations <- states
@@
 
-p <- list(r =.7, K = 1.2, q = 3, b = 0.15, a = 0.2) # lower-state peak is optimal
-#p <- list(r =.7, K = 1.5, q = 3, b = 0.15, a = 0.2) # higher-state peak is optimal
+#p <- list(r =.7, K = 1.2, q = 3, b = 0.15, a = 0.2) # lower-state peak is optimal
+p <- list(r = 2, K = 25, q = 4, b = 2, a = 7) # higher-state peak is optimal
 
@@
     y <- x +  h
-    r <- p$r - 0.09
+    r <- p$r - 0.2
 
@@
 m <- matrices[[3]]
-x0 = which.min(abs(states - 0.7))
+x0 = which.min(abs(states - 17))
 sim <- mdp_planning(m$transition, m$reward, discount, model_prior = c(1), 

### 2021-05-19 notebook/model-estimates.Rmd
:construction:

@@
 #scale <- 20
-x <- siteA$PopulationUntransformed / scale
+x <- siteA$PopulationUntransformed
 constants <- list(N = length(siteA$SampleYear),
@@
 ```{r}
-#p <- list(r = 2, K = 25, q = 4, b = 2, a = 7) # higher-state peak is optimal
+p <- list(r = 2, K = 25, q = 4, b = 2, a = 10)
 
@@
   x[1] <- x0
-  q <- 3
-  s <- 2
-  r ~ dunif(0,5) #dlnorm(log(1) - s^2/2, s)
-  K ~ dunif(0,40) #dlnorm(log(2)- s^2/2, s)
-  a ~ dunif(0,5) # dlnorm(log(7) - s^2/2,  s)
-  b ~ dunif(1.5,2.5) #dlnorm(log(1) - s^2/2, s)
-  sigma ~ dlnorm(log(0.05) - s^2/2, s)
+  q <- 4
+  r ~ dnorm(2, 0.5)
+  K ~ dnorm(25, 8)
+  a ~ dnorm(10, 3)
+  b ~ dnorm(2, .5)
+  sigma ~ dunif(0.0, .1)
 
@@
     mu[t] <-  x[t] + x[t] * r * (1 - x[t] / K)  - a * x[t] ^ q / (x[t] ^ q + b ^ q)
-    x[t+1] ~ dnorm(log( mu[t] ), sd = sigma)
+    x[t+1] ~ dlnorm(log( mu[t] ), sd = sigma)
   }
@@
 ```{r}
-
-may_f <- function(x){ # May
-      x + x * p$r_mean * (1 - x / p$K_mean)  - (p$a_mean) * x ^ 3 / (x ^ 3 + (p$b_mean) ^ 3)
+states <- seq(0,30, length=100)
+p <- list(r = 2, K = 25, q = 4, b = 2, a = 10)
+may_f <- function(y,p){ # May
+  p$q <- 4
+  pmax(y + y * p$r * (1 - y / p$K)  - p$a * y ^ p$q / (y ^ p$q + p$b ^ p$q), 0)
 }
 
-# Matches the optimal policy under these settings (depends on sigma, among others)
-logistic_f <- 
-  function(x){
-      x + x * l_p$r_mean * (1 - x / l_p$K_mean)
+logistic_f <- function(y,p){
+    pmax(y + y * p$r * (1 - y / p$K), 0)
 }
+# draw and plot samples from the mcmc
+landscape <- 
+  slice_sample(m_draws, n=1000) %>% 
+  transpose() %>%
+  map_dfr(function(p) 
+    map_dfr(list(may_f, logistic_f), 
+            function(fun) tibble(state = states, f = fun(states,p)-states),
+            .id = "model"),
+          .id = "sample")
 
-states <- seq(0,30, length=200)
-landscape <- map_dfr(list(may_f, logistic_f), 
-                     function(f) tibble(state = states, f = f(states) -states),
+landscape %>% ggplot(aes(state, f, group=interaction(sample,model), col=model)) + geom_line(alpha=0.01) #+ ylim(c(-5,1))
+```
+
+## Approximations
+
+controlling h is actually controlling the bifurcation point directly...
+
+```{r}
+landscape <- map_dfr(list(logistic_1, logistic_2, may), 
+                     function(f) tibble(state = states, f = f(states, p)),
                      .id = "model")
+monkeys <- data.frame(state = x, f = lead(x), model = "data") 
+bind_rows(landscape, monkeys) %>% ggplot(aes(state, f, col=model)) + geom_point()
+```
 
-landscape %>% ggplot(aes(state, f, col=model)) + geom_point()
+```{r}
+
+apes %>% ggplot(aes(x,x_t1)) + geom_point()
 ```

### 2021-05-19 notebook/state-tipping-decisions.Rmd
:construction:

--- 2021-05-19 notebook/tipping-decisions.Rmd
@@
 output: github_document
+title: The Forecasting Trap -- notes
 ---
@@
   function(x, h = 0){
-    y <- x +  h
-    r <- p$r - 0.2
-
+    y = x+h
     pmax(
-      y + y * r * (1 - y / p$K),  
+      y + y * (p$r - .65) * (1 - y / p$K),  
       0)
@@
 
-logistic_2 <- 
+alt  <- function(x, h = 0, r = .095, K = 20){
+  s <- pmax(x - h, 0)
+  s + s ^ 2 * r * (1 - s / K)
+}
+
+bh <- 
   function(x, h = 0){
-    y <- x +  h
-    r <- p$r + .5
+    y = x+h
+    A <- 1.3 * p$r
+    B <- .6 * p$K / (p$r - 1)
     pmax(
-      y + y * r * (1 - y / p$K),  
+      y * A / (1 + y / B),  
       0)
   }
+# Matches the optimal policy under these settings (depends on sigma, among others)
+logistic <- 
+  function(x, h = 0){
+    y = x + h
+    r <- p$r - .9
+    K <- p$K - 4
+    pmax(
+      y + y * r * (1 - y / K),  
+      0)
+  }
+logistic_2 <- function(x, h = 2) logistic(x,h)
+may_2 <- function(x, h = 2) may(x,h)
+may_10 <- function(x, h = 10) may(x,h)
 ```
@@
 
-landscape <- map_dfr(list(logistic_1, logistic_2, may), 
-                     function(f) tibble(state = states, f = f(states, 0) -states),
+landscape <- map_dfr(list(logistic_1, alt, may), 
+                     function(f) tibble(state = states, f = f(states) -states),
                      .id = "model")
 
-landscape %>% ggplot(aes(state, f, col=model)) + geom_point()
+landscape %>% ggplot(aes(state, f, col=model)) + geom_point() + ylim(-5,10) + xlim(0,5)
 ```
@@
 ```{r}
-matrices <- lapply(list(logistic_1, logistic_2, may), 
+matrices <- lapply(list(logistic_1, alt, may), 
        function(f) fisheries_matrices(states, actions, observations, reward_fn, 
@@
 ```{r, results="hide"}
-solns <- lapply(matrices, function(m) mdp_value_iteration(m$transition, m$reward, discount))
+#solns <- lapply(matrices, function(m) mdp_value_iteration(m$transition, m$reward, discount))
+solns <- lapply(matrices, function(m) mdp_policy_iteration(m$transition, m$reward, discount, policy0 = rep(1,length(states)), max_iter = 5000))
+
 ```
@@
 ```{r}
-df %>% ggplot(aes(state, action, col=model)) + geom_point(alpha=0.5)
+df %>% ggplot(aes(state, action, col=model)) + geom_line(alpha=0.5)
 

### 2021-05-19 notebook/tipping_point_appendix.Rmd
:construction:

--- 2021-04-02 manuscript/appendix.Rmd
@@
 ---
-title: 'Appendix'
+title: 'Appendix for: the Forecasting Trap'
 author: Carl Boettiger
@@
 
-The question of optimal harvest quotas given a population growth model can be posed as a Markov Decision Process (MDP).
-@Marescot2013 provides an accessible overview of MDP problems and their solutions in conservation context.
-@Reed1979 provides a formal proof of optimal harvest strategy for stochastic dynamics under the assumption of any concave growth function.
-The problem can be summarized as follows:
-
-A manager seeks to maximize the sum of the utility derived from such a harvest
-and such a state, $U(X_t,H_t)$, over all time, subject to discount rate $\delta$:
-
-\begin{equation}
-\sum_{t=0}^{t=\infty} U(X_t, H_t) \delta^t
-\end{equation}
-
-While in principle the utility could reflect many things, including the cost of fishing, market responses to supply and demand, the value of recreational fishing, the intrinsic value fish left in the sea [see @Halpern2013], for simplicity we will assume utility is merely a linear function of the harvest quota set by the manager, i.e.
-a fixed price $p$ per kilogram of fish harvested: $U(X_t, H_t) = p \min(H_t, X_t)$ (noting that realized harvest cannot exceed the available stock).
-As units are already specified in dimensionless quantities in this example, so without loss of generality we will set $p=1$.
-
-To solve this optimization, we must also know the dynamics of $X_t$, describing how future stock $X_{t+1}$ changes from the present state, $X_t$, after to harvest $H_t$:
-
-\begin{equation}
-X_{t+1} = f(X_t, H_t)
-\end{equation}
-
-This problem is already well studied, and it is worth noting that even under the simple utility function which assigns no value to fish that are not harvested, the optimal strategy still seeks to sustain the fish population indefinitely; as @Clark1973 shows for the deterministic function $f$ and @Reed1979 extended to the stochastic case.
-As @Reed1979 proves, the optimal strategy can be characterized by a policy of "constant escapement," in which the manager adjusts the harvest effort each year in an effort to ensure the same number or biomass of fish are left in the sea each year.
-The optimization problem can then be solved by dynamic programming, using the recursive Bellman equation.
-This technique, known as stochastic dynamic programming (SDP) is widely used in conservation decision problems [e.g.
-@Mangel1985, @Marescot2013].
-
-It is worth noting that while constant escapement is used in the management of several important fish stocks such as salmon, most marine fisheries are not managed using this optimal solution for a fluctuating population, but instead rely on a 'constant harvest effort' (or constant "yield") policy.
-In the case of a deterministic model at equilibrium, these policies are identical.
-Though constant-effort policies are technically sub-optimal for the problem as stated above with stochastic growth, in practice they can be more robust, such as when measurements of the current stock size $X_t$ are uncertain.
-These issues are discussed in detail in @pomdp-intro.
-For our purposes, it suffices to note that constant escapement policy is relevant both from a theoretical standpoint as the optimal solution for the problem under consideration, and an applied standpoint as a policy that guides management of many salmon fisheries.
-The optimal control approach using SDP as illustrated here is used in a wide and growing number of conservation problems.
-
-Bear in mind that in the example presented here, SDP merely serves the role of a convenient and established way for a manager to determine the optimal action, given some model estimation.
-Whether we use SDP or some other way to determine the optimal policy, given the model, is immaterial to the conclusion.
-
 ## Solving MDP via SDP
 
-To solve the decision problem using SDP, we define the state space and the action space on a discrete grid of 240 points.
-The maximum state is set well above the largest carrying capacity used in the models, which limits the influence of boundary effects introduced by the transformation to a discrete, finite grid.
-Available harvest actions match the state space, effectively allowing any harvest level to be possible.
-
 ```{r model_definitions}
-states <- seq(0,24, length.out = 240)
-actions <- states
+states <- seq(0,30, length=100)
+actions <- seq(0,30, length.out = 100)
 obs <- states
@@
 
-The utility (reward) of an action is set to the (realized) harvest (e.g.
-a fixed price per unit fish, with no cost applied to harvest effort).  Future
-utility is discounted by fixed factor of $0.99$. Classic maximum sustainable 
-yield models [@Schaefer1954] ignore discounting, while modern economic optimization
-models insist on it, so a small discount conforms to the latter while reasonably
-approximating the former.  The qualitative conclusion is not sensitive to the 
-discounting rate.
+```{r}
 
-```{r}
-reward_fn <- function(x,h) pmin(x,h)
-discount <- 0.99
+benefit <- 1
+cost <- 1 # quadratic costs
+reward_fn <- function(x,h) benefit * x - (h ^ cost)
+discount <- 0.999
 ```
-
-Alternate reward functions with varying price and cost structures are possible but do not qualitatively impact the conclusions.
-This reward function is a limiting case of any more complex reward, and corresponds both to classic work that does not model utility explicitly [e.g.
-MSY theory, @Schaefer1954], as well as explicit assumptions typically made in more recent models [@Reed1979].
-Moreover, using a simple reward function makes it clear that model that performs best is not doing so merely because of particular features baked into the a carefully chosen reward rule.
-
-## Ecological Model Definitions
-
-The manager chooses between two different logistic growth models, each of which
-can be thought of as an approximation to the underlying "true" population model:
-
-\begin{equation}
-f_i(Y) = Y + Y r_i \left( 1 - \frac{Y}{K_i} \right) \xi_i(t)
-\end{equation}
-
-where $\xi_i(t)$ is a multiplicative log-normal noise term with mean 1 and log-standard-deviation $\sigma_i$.
-
-Model 1 has $r_1 = 2$, $K = 16$, and $\sigma_1 = 0.05$.
-Model 2 has $r_2 = 0.5, K_2 = 10$, and $\sigma_2 = 0.075$.
-
-Meanwhile, the true population growth rate is simulated using a function with
-non-linear per-capita growth:
-
-\begin{equation}
-f_i(Y) = Y + Y^4 r_i \left( 1 - \frac{Y}{K_i} \right)
-\end{equation}
-
-with $r_3 = 0.002$, $K_3 = 10$ and $\sigma_3 = 0.05$.
 
@@
 ```{r}
-# K is at twice max of f3; 8 * K_3 / 5
-f1 <- function(x, h = 0, r = 2, K = 10 * 8 / 5){
-  s <- pmax(x - h, 0)
-  s + s * (r * (1 - s / K) )
-}
-f2 <- function(x, h = 0, r = 0.5, K = 10){
-  s <- pmax(x - h, 0)
-  s + s * (r * (1 - s / K) )
+
+p <- list(r = 2, K = 25, q = 4, b = 2, a = 7) # higher-state peak is optimal
+
+# Matches the optimal policy under these settings (depends on sigma, among others)
+f1 <- 
+  function(x, h = 0){
+    y = x+h
+    pmax(
+      y + y * (p$r - .65) * (1 - y / p$K),  
+      0)
   }
 
-# max is at 4 * K / 5 
-f3  <- function(x, h = 0, r = .002, K = 10){
+f2  <- function(x, h = 0, r = .095, K = 20){
   s <- pmax(x - h, 0)
-  s + s ^ 4 * r * (1 - s / K)
+  s + s ^ 2 * r * (1 - s / K)
+}
+
+f3 <- 
+  function(x, h = 0){ # May
+    y <- x +  h
+    pmax(
+      ## controlling h is controlling the bifurcation point directly...
+      y + y * p$r * (1 - y / p$K)  - p$a * y ^ p$q / (y ^ p$q + p$b ^ p$q),  
+      0)
   }
@@
 models <- list("1" = f1, "2" = f2, "3" = f3)
-model_sigmas <- c(sigma_g, 1.5 * sigma_g, sigma_g)
+model_sigmas <- c(sigma_g, sigma_g, sigma_g)
 true_model <- "3"
@@
 ```{r sdp, results="hide"}
-mdp <- memoise::memoise(mdp_value_iteration,
+mdp <- memoise::memoise(mdp_policy_iteration,
                         cache = memoise::cache_filesystem("cache/"))
@@
           function(P){
-    soln <- mdp(P, reward, discount = discount, 
-                epsilon = 0.01, max_iter = 2000, V0 = rep(0,dim(P)[[1]]))
-    escapement <- states - actions[soln$policy]
-    tibble(states, policy = soln$policy, escapement)
+    soln <- mdp(P, reward, discount = discount, max_iter = 5000, policy0 = rep(1, length(states)))
+    tibble(states, policy = soln$policy, actions = actions[soln$policy])
     }, 
   .id = "model") 
-
-write_csv(policies, "../data/policies.csv")
 ```
@@
 Tmax <- 100
-x0 <- which.min(abs(states - 6))
+x0 <- which.min(abs(states - 1.5))
 reps <- 100
@@
          )
+```
 
-write_csv(sims, "../data/sims.csv")
+```{r}
+## add predictions from each model.  
+do_nothing <- rep(1, length(states))
+unmanaged <- mdp_planning(transitions[[true_model]], reward, discount,
+           policy = do_nothing, x0 = x0, Tmax = Tmax) %>%
+    select(value, state_index = state, time, action_index = action)  %>% 
+    mutate(state = states[state_index])
+
 ```
@@
 confidence interval.
-
-```{r stepahead_unfished}
-stepahead_unfished <- sims
-stepahead_unfished$state_index <- rep(sims$state_index[sims$model == "1"],3)
-
-stepahead_unfished <- stepahead_unfished  %>% 
-  filter(model != "3") %>% 
-  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
-  rowwise() %>%
-  mutate(expected = transitions[[model]][state_index, , 1]  %*% states,
-         var = transitions[[model]][state_index, , 1]  %*% states ^ 2 - expected ^ 2,
-         low = states[max(which(cumsum(transitions[[model]][state_index,,1]) < 0.025)) ],
-         high = states[min(which(cumsum(transitions[[model]][state_index,,1]) > 0.975)) ],
-         true = states[next_state ]) 
-```
 
@@
 
-stepahead_unfished <- stepahead_unfished %>%
-  mutate(sd = sqrt(var),
-         score = scoring_fn(expected, true, sd))
-
 stepahead_fished <- stepahead_fished %>%
@@
 predictions <- 
-  stepahead_unfished %>% 
+  stepahead_fished  %>% 
   select(time, model, reps, expected, low, high, true, score)  %>% 
-  mutate(scenario = "A_unfished") %>% 
-  bind_rows(stepahead_fished  %>% 
-          select(time, model, reps, expected, low, high, true, score) %>% 
-          mutate(scenario = "B_fished")
-  ) %>% 
   mutate(model = as.character(model))
 
-write_csv(predictions, "../data/predictions.csv")
+```
+
+```{r}
+
+fig1cd <- predictions %>%
+  ggplot(aes(x = score, group = model, fill = model)) +
+  geom_histogram(binwidth = 2, show.legend = FALSE) +
+  coord_cartesian(xlim = c(-100, 1), ylim = c(0,4000)) +
+  xlab("Proper score") +
+  facet_wrap(~scenario, ncol = 1, labeller =as_labeller(c(
+    A_unfished = "C.",
+    B_fished = "D."))) + 
+  scale_x_continuous(breaks = c(-100, 0)) +
+  theme(axis.text.y = element_blank(),
+        plot.margin = margin(0, 0, 0, 0, "cm"),
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank()
+        )
+
+fig1ab <- predictions %>% 
+  filter(reps == "2", time < 10) %>%
+  ggplot(aes(time, col = model, fill = model)) + 
+  geom_point(aes(y = expected)) + 
+  geom_errorbar(aes(ymin = low, ymax = high)) +
+  geom_point(aes(y = true), pch = "*", size = 12, 
+             alpha = 0.6) +
+  ylab("stock size")
+
+fig1ab + fig1cd + 
+  plot_layout(widths = c(3,1), guides = 'collect') + 
+  theme(plot.margin = margin(0, 0, .1, .1, "cm"))
 ```
@@
   arrange(model, time)
+```
 
-write_csv(npv_df, "../data/npv_df.csv")
+```{r figure2, message = FALSE, fig.cap = "Ecological and economic performance of each forecast."}
+plot_ecology <- 
+  sims %>%
+  mutate(model = as.character(model), resp = as.character(reps)) %>%
+  group_by(model, time) %>%
+  summarise(mean_state = mean(state), sd = sd(state), .groups = "drop") %>%
+  filter(time < 25, model != "3") %>%
+  ggplot(aes(time, mean_state)) + 
+  geom_line(aes(col = model), lwd=1.5, show.legend = FALSE) +
+  geom_ribbon(aes(ymin = mean_state - 2*sd, 
+                    ymax = mean_state + 2*sd,
+                    fill = model),
+                alpha = 0.2, show.legend = FALSE) +
+  ylab("state") + 
+  labs(subtitle = "A. Ecological outcomes")
+
+optimal <- select(filter(npv_df, model == "3"), time, mean_npv)
+
+plot_econ <- 
+npv_df %>%
+  mutate(model = as.character(model), resp = as.character(reps)) %>%
+  filter(model != "3", time %in% seq(1,100, by = 5)) %>%
+  ggplot(aes(time, mean_npv)) +
+  geom_line(data = optimal, lwd = 1.5, col = "grey20") +
+  geom_point(aes(col=model), size = 4, alpha = 0.8) + 
+  ylab("Net present value") + xlab("time") + 
+  labs(subtitle = "B. Economic outcomes")
+
+plot_ecology + plot_econ
 ```
@@
   select(time, stock, quota, belief, method) 
-
-write_csv(am, "../data/am.csv")
-
 ```
@@
 
-## Adaptive management over many models
-
-We repeat the analysis above over a grid of $r$ and $K$ values.
-
-```{r}
-# compare over grid set of candidate r and K
-rs <- c(0.01, 0.1, 0.2, 0.3, 0.4, 0.5, 2)
-Ks <- c(0.5, 1, 1.5, 4, 10, 16)
-sigma <- 0.1
-n_models <- length(rs)*length(Ks)
-pars <- expand.grid(rs, Ks)
-names(pars) <- c("r", "K")
-pars$sigma <- 0.1
-closure <- function(r, K){
-  function(x, h = 0){
-  s <- pmax(x - h, 0)
-  s + s * (r * (1 - s / K) )
-  }
-}
-many_models <-  lapply(1:n_models, function(i) closure(pars[i,"r"], pars[i,"K"]))
-
-# transition matrices
-many_transitions <- lapply(seq_along(many_models), function(i) 
-  transition_matrices(many_models[[i]], states,  actions, sigma))
-names(many_transitions) <- as.character(seq_along(many_models))
-true_transition <- transition_matrices(f3, states, actions, sigma_g)
-
-## Once again, initial belief favors "model 1"
-model_prior <- c(rep(.01, n_models-1) / (n_models - 1), .99)
-
-```
-
-With the transition matrices defined for all `r n_models` models, we perform adaptive management as before:
-
-```{r}
-adaptive_management <- memoise::memoise(mdp_learning, cache = memoise::cache_filesystem("cache/"))
-am_many <- adaptive_management(many_transitions,  
-                               reward = reward, 
-                               discount = discount,
-                               model_prior = model_prior,
-                               x0 = x0, 
-                               Tmax = 20, 
-                               true_transition = true_transition, 
-                               epsilon = 0.001, 
-                               max_iter = 2000)
-
-```
-
-As before, we can compare the adaptive learning case to the policy which accounts for the initial uncertainty over all `r n_models` models, but does not seek to learn.
-
-```{r}
-policy_many <- policy_planning(many_transitions, reward = reward,  discount = discount,
-                          model_prior = model_prior, Tmax = 20, 
-                          epsilon = 0.001, max_iter = 2000)
-
-non_am_many <- mdp_planning(true_transition,  
-                       reward = reward, 
-                       discount = discount,
-                       model_prior = model_prior,
-                       x0 = x0, 
-                       a0 = 1,
-                       policy = policy_many$policy,
-                       Tmax = 20) %>% 
-  mutate(belief = 0.99, # belief in model 1 is fixed
-         method = "planning")
-```
-
-Again we record the posterior belief in the 'model 1' parameter values, and stack the data frames for learning and planning-only cases into a single file used to generate figure 3B.
-
-```{r}
-model1 <- dim(am_many$posterior)[2]
-am_multi <- am_many$df %>% 
-  mutate(belief = am_many$posterior[,model1],
-         method = "learning") %>%
-  bind_rows(non_am_many) %>%
-  mutate(stock = states[state],
-         quota = actions[action]) %>%
-  select(time, belief, stock, quota, method)
-
-write_csv(am_multi, "../data/am_multi.csv")
-```
-
-## Posteriors
-
-With `r n_models` possible models instead of two, the probability weight (belief) in Model 1 does not fully capture the evolution of the belief probabilities over time.
-To better illustrate how these beliefs evolve over the course of the simulation, I plot the marginal probabilities over values of $r$ and $K$ across the models over time.
-Note that Model 1 values, $r = 2$, $K = 16$, are quickly disfavored.
-
-```{r}
-posteriors <- 
-  data.frame(model = as.character(1:n_models), t(am_many$posterior)) %>%
-  left_join(mutate(pars, model = as.character(1:n_models))) %>%
-  pivot_longer(starts_with("X"), 
-               values_to = "probability",
-               names_to = "time") %>% 
-  mutate(time = as.integer(as.factor(time)))
-
-r_marginals <- posteriors %>% 
-  group_by(time,r) %>% 
-  summarise(prob = sum(probability), .groups = "drop")
-
-K_marginals <- posteriors %>% group_by(time,K) %>% 
-  summarise(prob = sum(probability), .groups = "drop")
-```
-
-From the table of marginal probabilities, we generate the corresponding plots
-
-```{r fig.width=8, fig.height=4, fig.cap = "Fig S1: Marginal probabilities of parameters $r$ and $K$ over the course of adaptive management over all 42 models." }
-belief_r <- r_marginals %>% 
-  mutate(r = as.character(r)) %>%
-  ggplot(aes(time, prob, group = r, col = r)) + geom_line() +
-  geom_point()
-
-belief_K <- K_marginals %>% 
-  mutate(K = as.character(K)) %>%
-  ggplot(aes(time, prob, group = K, col = K)) + geom_line() +
-  geom_point()
-
-belief_r + belief_K 
-```
-
-For most of the simulation, the most probable $r$ value is around 0.3, and the most probable $K$ value around 1.5.
-
 # Growth models
@@
 
-write_csv(model_curves, "../data/model_curves.csv")
-
 ```
 
-The resulting model curves are shown in the main paper, Fig 4A.
+```{r}
+plot_models <- 
+  model_curves %>% 
+  ggplot(aes(state, value, col=model, lty=model)) +
+  geom_hline(aes(yintercept = 0)) + 
+  geom_line(show.legend = FALSE) + 
+  coord_cartesian(ylim = c(-5, 8)) +
+  ylab(bquote(f(x) - x)) + xlab("x")
 
-## Growth models in the many-models case
+plot_policies <- policies %>%
+  ggplot(aes(states, actions, col=model, lty=model)) + 
+  geom_line() + xlab("state")
 
-The corresponding figure for the case of `r n_models` models.
+plot_models + labs(subtitle="A") + plot_policies + labs(subtitle="B")
+```
 
-```{r plot_many_models, fig.width=7, fig.height=5, fig.cap="growth curves for 42 candidate models, shaded by belief at timestep 20, relative to the true model.
-The curve with parameter values corresponding to 'model 1' are recognized in the highest arc, with belief weight of less than 1%."}
-
-final_belief <- posteriors %>% filter(time == max(time)) %>%
-  select(model, belief = probability) %>% mutate(type = "candidate") %>%
-  bind_rows(data.frame(belief = 1, model="true", type="true"))
-
-model_set <-c(f3,  many_models)
-names(model_set) <- c("true", 1:length(many_models))
-d <- 
-  map_dfc(model_set, function(f) f(states) - states) %>%
-  mutate(state = states) %>% pivot_longer(names(model_set), "model") %>%
-  left_join(final_belief)
-
-d %>% 
-  ggplot(aes(state, value, group=model, col = type, alpha = belief)) +
-  geom_line(lwd=1) + 
-  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
-  ylab(bquote(f(x) - x)) + xlab("x")
+```{r}
 ```

### 2021-07-06 manuscript/appendix.Rmd
:pen:

--- 2021-04-02 manuscript/appendix.Rmd
@@
 # compare over grid set of candidate r and K
-rs <- c(0.01, 0.1, 0.2, 0.3, 0.4, 0.5, 2)
-Ks <- c(0.5, 1, 1.5, 4, 10, 16)
+rs <- seq(.1, 2, length=6)
+Ks <- seq(1,19,by=3)
 sigma <- 0.1

### 2021-08-12 manuscript/manuscript.Rmd
edits

@@
 ---
-title: 'Iterative forecasting and the trap of self-fulfilling prophecies'
-author:
+title: 'The forecast trap'
 - name: Carl Boettiger
@@
   Encouraged by decision makers' appetite for future information on topics ranging from elections to pandemics, and enabled by the explosion of data and computational methods, model based forecasts have garnered increasing influence on a breadth of decisions in modern society.
-Using a classic example from fisheries management, I demonstrate that selecting the model or models that produce the most accurate and precise forecast can lead to decidedly worse outcomes.
-This creates a cycle of self-fulfilling prophecy, in which the manager not only over-exploits the fishery leading to declining biomass and declining yields, but also becomes increasingly convinced that these actions are not over-exploitation but in fact consistent with the best models and data for sustainable management.
+Using a classic example from fisheries management, I demonstrate that selecting the model or models that produce the most accurate and precise forecast (as measured by most widely used statistical scores) can lead to decidedly worse outcomes (as measured by real-world objectives such as dollars or fish).
+This can create a forecast trap: in which the manager not only over-exploits the fishery leading to declining biomass and declining yields, but also becomes increasingly convinced that these actions are not over-exploitation but in fact consistent with the best models and data for sustainable management.
+The forecast trap is not unique to this example, but possible whenever (1) the optimal management policy is not unique to the generative process, and (2) the generative process is not itself in our candidate set of models.
 
@@
 
-- abbreviated running title: "The trap of self-fulfilling prophecies"
+- abbreviated running title: "The forecast trap"
 - Authorship statement: _This is a single-author paper_
@@
 
-In this paper, I demonstrate that the model that makes the best prediction is not necessarily the model that makes the best policy.
-I show that iterative learning or iterative forecasting can be counter-productive to management outcomes:
-that in some cases, a manager would be better to ignore new information which could reduce uncertainty over available models.
+In this paper, I demonstrate that the model that makes the better prediction (rigorously defined as a strictly proper score, @Gneiting2007) is not necessarily the model that makes the better policy (rigorously defined in terms of utility, e.g.
+expected net present value, @Clark1990).
+I show that methods for learning about model structure or parameters by repeatedly comparing forecasts to observations can be counter-productive.
 Put another way, the value of information (VOI, as measured by the expected utility given that information minus the utility without it; see @Howard1966; @Katz1987), can actually be negative.
-This is not a consequence of imperfect information (e.g. observation error) or the result of model over-fitting.
-The issue is an intrinsic consequence of the fact that many models can lead to the same decisions, and no generic solution to the problem exists. 
-Worse, whenever this situation arises, the decision-maker becomes trapped by a form of self-fulfilling prophecy: 
-the model(s) which the decision-makers have learned to favor lead to decisions and outcomes that closely match model expectations.
-Consequently, the decision-maker can become ever more firmly entrenched in the belief that their new models are accurate and the mediocre outcomes they experience are as good as it gets, while confidently rejecting earlier models which would have led to much better outcomes.
-As this example will show, by better understanding what aspects of a model are most and least important to shaping a particular decision, we can at least anticipate the circumstances that make this problem more likely.
+When VOI is negative, the decision-maker may become trapped into accepting mediocre outcomes derived from a model that makes accurate forecasts, even when a less accurate model that would generate better outcomes is available.
+In our example, the manager will decide the fishery in question simply has low productivity, because such a model yields better predictions, rather than realizing that the low productivity observed is in fact a consequence of the over-harvesting.
+This disconcerting situation can arise whenever two conditions are met: (1) the optimal management policy is not unique to the generative process, and (2) the generative process is not included in candidate set of models.
+These conditions do not guarantee the trap will occur, only the circumstances in which it cannot be ruled out entirely.
 
-Iterative learning approaches such as adaptive management or iterative forecasting [@Dietze2018] are particularly compelling because they address a different reason why seemingly-accurate models can lead to poor outcomes: the problem of overfitting.
-When models are estimated from the data they seek to describe, the model may overfit the data, making the data seem even more likely than it would be under the true process [@Burnham1998].
-Over-fit models may underestimate uncertainty and make poor predictions about future outcomes [@Ginzburg2004].
-Because any iterative learning compares models to future outcomes, over-fitting those outcomes is impossible.
-This concept was rigorously formalized by @Gneiting2007's proof of "proper" scoring rules.
-By definition, under a proper scoring rule, no probabilistic prediction $Q(x)$ can score better, on average, than that of the underlying process, $P(x)$.
-@Gneiting2007 further shows which techniques for evaluating forecast predictions do and do not satisfy this condition, i.e. which cannot be over-fit.
-Adaptive management is based on equally secure footing because model predictions are again compared to future outcomes not used to fit the model.
-Instead of scoring and selecting among alternative models, adaptive management considers probabilities over all models [@Walters1978].
-These two approaches are particularly compelling because they can both reduce uncertainty and avoid over-fitting.
-My example is not meant to undercut the importance of these approaches.
-Previous work has long acknowledged the panoply of ways in which iterative learning or other model-based decision making can go astray due to conflicting incentives, implementation errors, or lack of resources for monitoring and updating [e.g.
+The forecast trap is not the only mechanism by which some model-choice methods lead to worse outcomes.
+Previous work has long acknowledged the panoply of ways in which model-based decision making can go astray due to conflicting incentives, implementation errors, or lack of resources for monitoring and updating [e.g.
 @Ludwig1993].
-Here, the problem is more fundamental.
+Another widely recognized problem is that of over-fitting [@Burnham1998], in which the model that best fits historical data fails to best predict future data [@Ginzburg2004]. 
+Under such circumstances, it is easy to see how an over-fit model would also lead to bad outcomes.
+However, over-fitting plays no role in the forecast trap, where model predictions are assessed only using probabilistic forecasts, and not observations which had previously been used to fit the models.
+Formally, these scores satisfy the 'proper scoring' rule of @Gneiting2007, which proves no other probabilistic prediction $Q(x)$ will have a better expected score than that of the generative process $P(x)$.
+@Gneiting2007's proof of proper scoring has since become a critical tool to avoid over-fitting when choosing models to make decisions, but as I illustrate, will not prevent the forecast trap.
+
+# From Predictive Models to Decision Policies
+
+How do we translate a model-based forecast into a decision?
+It is impossible to discuss outcomes associated with a forecast without first agreeing on this process.
+In practice, decision-makers may use a forecast in a wide variety of ways in selecting a course of action, including ways which may run counter to the stated objectives of management [@Ludwig1993].
+In principle at least, the field of decision theory provides a formal mechanism for determining the optimal strategy given a model forecast.
+For instance, a wide range of ecological conservation and management problems can be expressed as a Markov Decision Process (MDP) problems [@Marescot2013].
+Existing computer algorithms such as stochastic dynamic programming (SDP) take a probabilistic model _forecast_ (more precisely, the probability $P(x_{t+1} | x_t, a_t)$ of the system being in state $x_{t+1}$ in the next iteration given that it was previously in state $x_t$ and the manager selected action $a_t$) and the _desired management objective_ (i.e.
+the maximize the expected biomass of species protected or the expected dollar profit of a fishery [see @Clark1990; @Halpern2013]) as input, and return the _decision policy_ which maximizes that objective [@Marescot2013].
+This provides a principled way to associate a decision policy with any given forecast model.
+
+Two features of this approach are worth emphasizing. 
+First, the resulting decision is derived directly from the forecast model and the desired objective.
+The SDP algorithm is a reasonable description of the approach any ideal manager would use -- considering all possible outcomes from all possible sequences of actions and selecting the best sequence.
+For complex models this process is too laborious even for a computer, and is often simplified by considering only a selection of predetermined policies (as in Management Strategy Evaluation, MSE, @Punt2016), or scenarios (as in scenario analysis, @Polasky2011).
+Such shortcuts are often necessary for complex real-world models, but open additional room for error: the policy we derive from a given forecast may perform poorly not because the model forecast was at fault, but because of those simplifying assumptions about possible policies.
+To ensure that the forecast trap is not a result of such assumptions about possible policies, we will consider a problem simple enough to solve directly with SDP. 
+This leads to the second point: the resulting decision policy is optimal, so long as the forecast model is correct.
+In this way, the SDP merely stands in for a mathematically precise way in which forecasts are turned into decisions.
+Recognizing that the SDP-derived policy (A) comes directly from the forecast model, and (B) gives the optimal policy for said forecast, seems to suggest that the whatever model makes the better forecast will surely also lead to better outcomes (as measured in terms of whatever utility we have chosen to maximize).
+While this intuition is no doubt _often_ accurate, our purpose here is to demonstrate that it is by no means _guaranteed_:
+it is also possible for the model which makes the better forecast to lead to worse outcomes.
+
+# Ecological Models
 
@@
 from the Maximum Sustainable Yield of @Schaefer1954 or the epidemiological reproductive number, $R_0$ of @Kermack1927.
-
-# Ecological Models
 
@@
 
-In Scenario B (actively harvest), I have first solved for the optimal management strategy for both model 1 and model 2 [Fig 1b].
-For small noise and concave functions with linear reward structure this can be done analytically [see proof in @Reed1979], or solved more generally by stochastic dynamic programming [see review by @Marescot2013, details in Appendix].
+In Scenario B (actively harvest), I have first solved for the optimal management strategy for both model 1 and model 2 [Fig 1b] using SDP, [see @Marescot2013 or code in the Appendix].
 Under this scenario, replicate simulations of the stock are harvested at each time step using the optimal quota dictated by either model 1 and model 2.

### 2021-08-23 manuscript/manuscript.Rmd
[WIP] :pen:

@@
 expected net present value, @Clark1990).
-I show that methods for learning about model structure or parameters by repeatedly comparing forecasts to observations can be counter-productive.
+I show that our best methods for learning about model structure or parameters by repeatedly comparing forecasts to observations can be counter-productive.
 Put another way, the value of information (VOI, as measured by the expected utility given that information minus the utility without it; see @Howard1966; @Katz1987), can actually be negative.
@@
 @Clark1973; @Reed1979; @Walters1981; @Ludwig1982].
+
+While methods such as iterative forecasting [@Deitze2018] and adaptive management [@Walters1978] can be _applied_ to real-world using empirical data, we can only _evaluate_ their potential in hypothetical examples when the true model is known, e.g.
+through numerical simulation.
+That approach allows us to compare both predictions and outcomes across implementations in independent identical replicate worlds.
+As noted above, we will assume our underlying model simple enough to solve by SDP, ensuring any poor outcomes from a given forecast are not merely an artifact of an imperfect decision process.
+Simple models also have the virtue in being accessible to closed form analysis, which, as we shall see, can give greater insight into when and why this forecast trap arises.
+That insight will in turn will allow us to examine if the same problem is likely to arise under more complex models.
+
+<!--
 While modern fisheries management frequently relies on complex models which may contain over 100 parameters to reflect the specific age or stage structure of a specific fish stock [@ramlegacy2018; @ram], fisheries management and research has always appreciated the central insights offered by simpler one-dimensional models which continue to underpin both the theory and practice.
@@
 from the Maximum Sustainable Yield of @Schaefer1954 or the epidemiological reproductive number, $R_0$ of @Kermack1927. 
+-->
 
@@
 
-For simplicity and comparison with prior work, we will assume a fixed price of fish $p$ with no additional cost on additional harvest, $U(X_t, H_t) = p \min(H_t, X_t)$ (noting that realized harvest cannot exceed the stock size).
-Without loss of generality we will set the price $p = 1$ and modest discount $\delta = 0.99$.
+We will assume we have been given a fixed price of fish $p=1$ with no additional cost on additional harvest, $U(X_t, H_t) = p \min(H_t, X_t)$ modest discount $\delta = 0.99$.
 
-We further imagine that the function $f$ is not known precisely.
-We restrict ourselves to two simple candidate models $f_1$ and $f_2$.
-Both share the same underlying structure of logistic recruitment (known as the Gordon-Schaefer model in fisheries context owing to groundbreaking work independently by @Schaefer1954 and @Gordon1954), differing only in their choice of certain parameters:
+Let us assume that our set of candidate models are simply the possible parameterizations of a stochastic version of the classic Gordon-Schaefer model [@Schaefer1954; @Gordon1954]:
 
 \begin{equation}
-f_i(Y) = Y + r_i Y \left( 1 - \frac{Y}{K_i} \right) * \xi_t(\sigma)
+f_i(Y_{t+1}) = Y_t + r_i Y_t \left( 1 - \frac{Y_t}{K_i} \right) * \xi_t(\sigma)
 \end{equation}
 
-Let us assume $\xi_t(\sigma)$ represents log-normal random noise with a mean of unity and log-standard-deviation $\sigma_i$ for each model.
-Model 1 is given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, Model 2 by $r_2 = 0.5$, $K_2 = 10$, $\sigma = 0.075$ (in dimensionless units).
-Having both the larger growth rate and the larger carrying capacity, model 1 is clearly the more optimistic of the two choices.
+Where $Y$ is the population size after harvest, $Y_t = X_t-H_t$  and $\xi_t(\sigma)$ represents log-normal random noise with a mean of unity and log-standard-deviation $\sigma_i$.
 
-Selecting between model 1 and model 2 can thus be considered the simplest illustration of the model uncertainty problem.
-This is a subset of the more general problem of selecting model parameters, assuming a logistic growth, which itself is a subset of estimating the best structural form (e.g.
-Ricker, Beverton-Holt, etc).
-There is no need to consider these more complicated versions of the model uncertainty problem here, since they all inherit the same issue.
-Reducing the model selection problem to these two models simplifies the presentation and will aid intuition at no loss of generality.
+<!--
+This is where most fisheries biologists will probably stop reading.
+Modern stock assessment models frequently have over 100 parameters.
+However, the goal of exercise understand how a good forecast can lead to a bad decision even when we use these iterative forecasting methods correctly, not predict the safe harvest levels of Pacific Anchovy in 2022.
+If any method fails to select the better model in the simplest cases, it does not often do better in more complex ones, even though the problems are harder to detect.
+-->
 
-The only additional assumption we will need is that the "true" model is not among the suite of models under consideration.
-Mathematical models are, at best, approximations of the underlying processes.
-Ecological processes are much too complex to ever be modeled exactly.
+Each possible parameterization of the model will generate a different forecast, and thus require a it's own SDP solution to determine the corresponding policy.
+We will consider a larger suite of these logistic models in a moment, but first let us draw out only two such parameterizations from this larger suite.
+This will let us focus our attention on two of the most interesting regions of that are already included within that larger parameter space of all possible values for $r$, $K$, and $\sigma$.
+Thus, let us take "Model 1" as being given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, "Model 2" by $r_2 = 0.5$, $K_2 = 10$, $\sigma_2 = 0.075$. 
+We can imagine our comparison of these two models as a microcosm of the larger comparison between all possible paremeterizations.
+
+Ecologists will rightly scoff at the simplicity of these models -- the real world is much more complicated.
+So it is important to bear in mind that these are not models that seek to approximate the stock dynamics of real world fisheries, only to approximate whatever "true model" we are using to drive the simulation.
+In recognition of the fact that real world is always more complex than even our best ecological models, we will assume a "true model" for the simulations that is not in the Gordon-Schaefer class (i.e.
+our candidate models will never contain the true model), but is not so rich that a Gordon-Schaefer curve would seem a hopelessly poor approximation.
+
 For illustrative purposes, we will thus assume the "true" process to be given by Model 3, which is unknown to the decision-maker, but similar enough to at least one of the candidate models might be considered a reasonable approximation:
@@
 
-The task of deciding whether Model 1 or model 2 would be the better choice for decision making is thus perhaps the simplest example of the much studied issue of model uncertainty that we can pose. 
-As in any real world scenario, neither model is the true model, but nevertheless this model set contains a good enough approximation of the true model to make good decisions.
-However, any of the well-developed approaches for decision-making under model uncertainty will prefer model 2 over model 1, despite the fact that the optimal policy under model 2 leads to much worse outcomes ecologically and economically.
+Certainly, the challenge of choosing which model to base a decision policy on in the real world is much harder than this binary choice between two models, and yet it is sufficient to illustrate the trap.
+We will see later why making the models much more complex does not guarantee that the task becomes easier or that the trap may be ruled out.
 
@@
 Moreover, a decision maker employing an adaptive management or iterative forecasting assessment such as those considered here would have no way of realizing that the outcomes they experienced were in fact sub-optimal.
-In both approaches, the manager quickly concludes that model 1 is entirely implausible, while finding that model 2 is remarkably accurate at predicting future values.
+In both approaches, the manager quickly concludes that model 1 is entirely implausible, while finding that model 2 is remarkably accurate at predicting future values -- the manager is unknowing trapped into bad outcomes.
 This problem is not addressed by re-estimating parameters (equivalently, considering a larger suite of models spanning a wide range of possible $K$ and $r$ values): which improves only marginally over the outcome of considering only two models [Fig 3B].
-Both iterative forecast evaluation and the adaptive management approaches to reducing model uncertainty lead to management becoming trapped in a self-fulfilling prophecy:
-holding the fish population at an unproductively low biomass suggested by model 2 while becoming ever more convinced that model is suitable. 
 Without access to the true model, none of our methods for model choice or reducing model uncertainty can break this trap.
@@
 Meanwhile, the peak of model 2 occurs at a substantially lower stock size.
-While the optimal control solution appears to depend only on a step-ahead forecast accuracy (indeed, the SDP solution method used here takes only step-ahead forecast probabilities as input, [@Marescot2013]), mathematical analysis showed long ago [@Reed1979] that the optimal solution for this problem depends only on keeping the biomass at the value responsible for the maximum growth rate.
+While the optimal control solution appears to depend only on a step-ahead forecast accuracy (recall that the SDP solution method used here takes only step-ahead forecast probabilities as input, [@Marescot2013]), mathematical analysis showed long ago [@Reed1979] that the optimal solution for this problem depends only on keeping the biomass at the value responsible for the maximum growth rate.[^1]
+Consequently, many models can share the same optimal solution, regardless of their forecast performance.
+
+This result may appear surprising, and begs the question as to whether the non-uniqueness of the optimal solution is not merely an artifact of the simple models considered here. 
+Surely more any realistically complex model would require a comparably more complex management policy such that the likelihood that any of our candidate models shares the same optimal policy would be vanishingly small?
+The intuition sounds reasonable
+
+[^1]: This is only approximately accurate. @Reed1979 demonstrates
+
 This realization is quite general: for most decision problems, simple models will exist under which the optimal decision is the same as it would be for the true model, even when that simple model is wrong in most other ways.
@@
 Situations similar to this example, in which a model can lead to the optimal decision while producing a poor forecast, are likely to be widespread.
+
+<!--
 For example, marine fisheries typically restrict a Management Strategy Evaluation [MSE; @Punt2016] to constant mortality, (defined as harvest per unit biomass, $F = H/B$).
@@
 Everything else about our choice of Model 1 can be far from the true model, and yet it will give optimal results, just as above.
-Examples in which decisions depend a subset or ratio of model parameters can be found everywhere.
-In @Dee2017, the optimal number of non-critical species to protect in order to conserve critical species is given by a ratio of model parameters.
-In the SIR model of @Kermack1927 the basic reproductive number that determines if an infection spreads is the ratio of contact rate over the recovery rate, $R_0 = \beta / \gamma$.
-In general, there will be many models that can lead to the same decision, and thus the possibility of a similar scenario to the one seen here.
-Likewise there will always be infinitely many models like Model 2, which can fit very close to the data [Fig 4] while leading to a very
-different decision policy.
+-->
 
@@
 
-<!-- Not just overfitting -->
-Note that the mechanism shown here has nothing to do with the much more familiar issue of over-fitting, in which a better-fitting model will also lead to worse outcomes [@Burnham1998].
-In fact, a model which has been over-fit will gradually be rejected by either the iterative forecasting or adaptive management approaches shown here, as these approaches continually confront the models with new data to which they had not been previously fit.
-The ability to avoid overfitting is one of the greatest appeals of any iterative management strategy.
-In the example here, both model 1 and model 2 have the same structural complexity.
-The better decision performance of model 1 is a consequence of capturing the key attribute needed for a good decision, which is perhaps surprisingly an all-together different criteria than fit.
-Note that other methods of model selection not considered here, including goodness-of-fit metrics such as $r^2$ or information criteria [@Burnham1998], will all prefer Model 2 over Model 1 for the same reason.
+This simple example can be resolved by including our true model in the candidate set, or any class of models flexible enough to 'span' the true model, such as a Gaussian Process [e.g.
+see @Boettiger2015].
+Given sufficient data, either approach shown here will prefer the true model over any other (as guaranteed by @Gneiting2007's theorem), and thus discover the optimal policy.
+While it is not difficult to image a candidate set of models flexible enough to include the very simple "true model" from Eq (4), it is much more unlikely that our model candidates truly span the generative processes of the real world.
 
-<!-- Not just "get more models"
-wrong!
+This example demonstrates that it is possible for a model to give quite good forecasts while leading to marginal outcomes, and conversely, that is possible for a model to give optimal outcomes while making terrible forecasts.
+This result may be particularly surprising given that the policy is derived directly from the forecast.
 
--->
+In most realistic management scenarios, we do not have the benefit of a theorem similar to Reed's to identify what features are most critical to approximate well and what are not.
 
-This simple example can be resolved by including a model that closely approximates both the location of the peak and the shape of the curve of our "true" model.
-Given sufficient data, either adaptive management or iterative forecasting will prefer such a model (as guaranteed by @Gneiting2007's theorem), while having the correct peak location means the model will also lead to the optimal policy in this case (as guaranteed by @Reed1979's theorem).
-Yet the larger problem remains unresolved.
-In most realistic management scenarios, we do not have the benefit of a theorem similar to Reed's to identify what features are most critical to approximate well and what are not.
 For instance, the theorem no longer holds if stock estimates are uncertain [@pomdp-intro].

### 2021-09-10 manuscript/manuscript.Rmd
WIP

@@
 
-In Scenario B (actively harvest), I have first solved for the optimal management strategy for both model 1 and model 2 [Fig 1b] using SDP, [see @Marescot2013 or code in the Appendix].
-Under this scenario, replicate simulations of the stock are harvested at each time step using the optimal quota dictated by either model 1 and model 2.
-The resulting stock sizes in the time-step following this harvest are once again compared to the probabilities predicted by each model using Eq \eqref{proper}.
+In Scenario B (actively harvest), I have first solved for the optimal management strategy using the forecast-matrices of both model 1 and model 2 [Fig 1b] using SDP [@Marescot2013; code in the Appendix].
+Replicate simulations of the stock are harvested at each time step using the optimal quota dictated by either model's forecasts, according to the SDP. 
+The resulting stock sizes in the subsequent timestep are scored against the forecast probabilities of each model using Eq \eqref{proper}.
 Model 2 unequivocally outperforms model 1 in both scenarios of passive observation and active harvest.
@@
 Had we been able to include Model 3 in our forecast comparisons, it would equal or outperform the forecasting skill of both model 1 and model 2 (as guaranteed by the theorem of @Gneiting2007), while also matching or out-performing their economic utility (as guaranteed by the theorem of @Reed1979).
-In practice, we never have access to the generating model, so it is reasonable to expect model selection to determine the best approximation.
-As we see here, the best approximation for forecasting future states does not in fact lead to better outcomes.
+In practice, we never have access to the generating model, so it is reasonable to expect model selection to determine the better approximation.
+As we see here, the better approximation for forecasting future states does not in fact lead to better outcomes.
 
-One obvious limitation in this comparison is that scenario B treats each model as fixed over the entire course of the simulation.
-It would be possible to generate forecasts for the next $n$ steps, but these forecasts would also be conditional on the management action (i.e. harvest quota) selected at each step.
-To evaluate two-step-ahead predictions we must consider each possible action under each possible state predicted by the step-ahead forecast, weighted by probability of the state given the model and the probability of the model.
-This rapidly expanding set of possibilities is addressed by adaptive management for sequential decision problems [e.g. @Smith1981], which I employ in the next section.
+One obvious limitation in this comparison is that scenario B treats each model as fixed over the entire course of the management simulation.
+In reality, managers will typically re-estimate model parameters after each subsequent observation.
+And rather than consider each model/parameter combination in isolation, managers will generate forecasts which reflect the current uncertainty as to which model or parameter values are most likely. 
+Re-estimating model parameters results in adjusting those probabilities.
+This approach is characterized by adaptive management for sequential decision problems [e.g. @Smith1981], which I employ in the next section.
 
@@
 Any adaptive management strategy updates posterior distributions over model uncertainty [@Punt2016; @Ludwig1982].
-Unfortunately, any such adaptive updating leads to worse outcomes than the equivalent non-adaptive strategy, in which model uncertainty is held fixed.
+Unfortunately, in this case, any such adaptive updating leads to worse outcomes than the equivalent non-adaptive strategy, in which model uncertainty is held fixed.
 I illustrate the application of a passive adaptive management strategy to this simple example, following classic examples for parameter [@Ludwig1982] or structural [@Smith1981] model uncertainty.
@@
 To demonstrate that the behavior is not driven by failure to explore sufficiently, (which might be addressed by an active adaptive management), I will assign initial probability that model 2 is true at 1%.
-After a single iteration of learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct [Fig 3A].
-I compare these results to 
+After a single iteration of adaptive learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct [Fig 3A].
+
 As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating, which achieves a net present value that only 31% that expected under management using model 1 alone [Fig 2].
@@
 
+So far we have considered only two alternative combinations of the parameters $r$, $K$ and $\sigma$. 
+This simplifies the calculations, because each unique parameter value combination requires a new run of the SDP algorithm to determine the optimal policy from the corresponding forecast.
 Increasing the space of possible models to cover a whole plausible range of parameters $r$ and $K$ does little to resolve this problem [Fig 3B]. 
@@
 <!-- quick summary of what was demonstrated -->
-Given this simple decision problem in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariably choose the other.
-Moreover, a decision maker employing an adaptive management or iterative forecasting assessment such as those considered here would have no way of realizing that the outcomes they experienced were in fact sub-optimal.
-In both approaches, the manager quickly concludes that model 1 is entirely implausible, while finding that model 2 is remarkably accurate at predicting future values -- the manager is unknowing trapped into bad outcomes.
-This problem is not addressed by re-estimating parameters (equivalently, considering a larger suite of models spanning a wide range of possible $K$ and $r$ values): which improves only marginally over the outcome of considering only two models [Fig 3B].
-Without access to the true model, none of our methods for model choice or reducing model uncertainty can break this trap.
-This should not be mistaken as a critique of adaptive management or iterative forecasting specifically -- other model choice approaches such as goodness-of-fit, information criteria or cross-validation would all prefer model 2 as well.
+Given this simple decision problem in which one of the two models leads to better ecological and economic outcomes, current approaches invariably choose the wrong one.
+Moreover, despite continuing to collect new observations, the decision maker has no way of realizing their mistake.
+The manager is trapped into believing whichever model produces the better forecast, even when this results in decidedly worse objective outcomes.
+Re-estimating parameters with as new observations accumulate only reinforces the problem [Fig 3A], and
+introducing a larger suite of models, such as our wide range of $r$ and $K$ values, does not escape this trap either [Fig 3B].
+Other model choice approaches such as goodness-of-fit, information criteria or cross-validation would all prefer model 2 as well.
+Only by including the true model in our set of candidates can we be certain that forecast-based methods will converge on optimal outcomes.
 
@@
 The reason for model 1's seemingly contradictory ability to make good decisions but bad forecasts becomes obvious once we compare both curves to that of the underlying model, model 3.
-Plotting the growth rate functions of each model, [Fig 4A], it is hardly surprising that no method exists which would not prefer the closely overlapping model 2 to the no-where-close model 1 as the better approximation of model 3.
-Nevertheless, decisions based on model 1 are nearly indistinguishable from those based on the true model [Fig 4B], while model 2 leads to over-harvesting.
+Looking at plots of the growth rate curves for each model [Fig 4A], it is hardly surprising that all model selection approaches prefer the closely overlapping curve of model 2 to the no-where-close curve of model 1 as the better approximation of model 3.
+Nevertheless, the decision policy derived from model 1 forecasts is indistinguishable from that based on the true model [Fig 4B], while the policy derived from model 2 forecasts lead to over-harvesting.
+Being closest to the true model's forecast skill never guarantees that we are closest to the true model's optimal policy.
+
+Perhaps this should not be surprising: ecologists have long observed that all models are wrong and the choice of better model depends on the the modeling goals [@Levins1966; @Walters1978; @Ludwig1993; @Getz2017].
+And we are clearly considering different goals: forecast skill (a unitless statistical measure) vs policy outcomes (be they measured in dollars or fish in the ocean).
+Yet the result is surprising all the same. 
+The forecast isn't just some other arbitrary modeling objective; it is a central input into the decision making process, both in the real world [@Clark2001; @Deitze2018] and in our idealized decision-making algorithm, SDP [@Marescot2013].
+Nor can we say the same model can never be best at both goals -- obviously the 'true' model always optimizes both objectives [^1]. 
+It is natural to assume from this that the candidate with the closest forecast will also be the one with the closest policy.
+The example presented here proves this is by no means guaranteed, but also begs the question -- how common is this forecast trap?
+
+[^1]: With enough data from enough of the state space, an SDP algorithm using a Gaussian Process prior [@Boettiger2015], which spans the "true model" given by Eq 4, will escape the forecast trap, as guaranteed by @Gneiting2007's theorem.
+However, real ecological systems are much more complex than Eq 4, and not so easily spanned by mathematical models.
+
+The forecast trap can occur whenever two conditions are met: 
+(1) our candidate models do not include the true model, and
+(2) the optimal policy of the true model is not unique to that model.
+The first condition is widely accepted as common-place -- ecological processes will always be more complex than our models [@Getz2017].  
+The second condition is more subtle: the fact that both curves for model 1 and model 3 produce exactly the same management policy (Fig 4B) under the SDP algorithm may be surprising.
+Fortunately, the models we have chosen to focus on here are simple enough to be analytically tractable -- that is, we can solve for the optimal policy without the brute-force numerical approach of SDP.
+That solution was first provided by @Reed1979, who demonstrated the optimal policy would be a so-called "bang-bang" strategy, in which the manager seeks to bring the fish stock to some target biomass as quickly as possible whenever it deviates from that size.
+The precise value of that target biomass depends of course on the details of the model -- but importantly not all the details.
+Intuitively one can think of this as maintaining the biomass at the most productive size: the maximum population growth rate (position of the peak of the growth curves in Fig 4A), though this is only precisely true without discounting ($\delta = 1$): the optimal stock size $\hat x$ is the solution to $f(\hat x) = \hat x/\delta$ when stochasticity is sufficiently small [@Reed1979].
+Thus, all models in which the peak growth rate occurs at the same stock size will have the same optimal policy.
+While the SDP is able to discover this same policy numerically by examining the forecast probabilities, Reed's theorem makes it intuitive why that policy is shared between models 1 and 3.
+
+However, Reed's proof considered only one-dimensional growth models $f(x)$, and so tells us nothing about models with stage structure or interacting species.
+It is easy to believe that the simple and thus non-unique optimal policy studied by @Reed1979 is an artifact of the simple growth model: surely the intuition provided will not extend to two or more species?
+Fortunately, recent work in the mathematics literature has finally been able to extend Reed's results.
+@Hening2021 proves under comparable assumptions that the optimal strategy
+
+<!---
 The explanation comes from noticing that the stock size corresponding to the maximum growth rate under model 1 (the peak of the curve) falls at almost exactly the same stock size as that of the peak growth rate for model 3.
@@
 [^1]: This is only approximately accurate. @Reed1979 demonstrates 
+-->
 
@@
 ```
-
-This simple example can be resolved by including our true model in the candidate set, or any class of models flexible enough to 'span' the true model, such as a Gaussian Process [e.g.
-see @Boettiger2015].
-Given sufficient data, either approach shown here will prefer the true model over any other (as guaranteed by @Gneiting2007's theorem), and thus discover the optimal policy.
-While it is not difficult to image a candidate set of models flexible enough to include the very simple "true model" from Eq (4), it is much more unlikely that our model candidates truly span the generative processes of the real world.
 

### 2021-09-11 manuscript/manuscript.Rmd
:pen:

@@
 Fortunately, the models we have chosen to focus on here are simple enough to be analytically tractable -- that is, we can solve for the optimal policy without the brute-force numerical approach of SDP.
-That solution was first provided by @Reed1979, who demonstrated the optimal policy would be a so-called "bang-bang" strategy, in which the manager seeks to bring the fish stock to some target biomass as quickly as possible whenever it deviates from that size.
-The precise value of that target biomass depends of course on the details of the model -- but importantly not all the details.
+
+Analytical solutions offer more insight as to when and why very different forecasts can generate the identical policy.
+Such a solution was first provided by @Reed1979, who demonstrated the optimal policy would be a so-called "bang-bang", in which the manager always seeks to bring the fish stock to some fixed target biomass as quickly as possible.
+The precise value of that target biomass depends of course on the details of the model -- but crucially, only on some of those details.
 Intuitively one can think of this as maintaining the biomass at the most productive size: the maximum population growth rate (position of the peak of the growth curves in Fig 4A), though this is only precisely true without discounting ($\delta = 1$): the optimal stock size $\hat x$ is the solution to $f(\hat x) = \hat x/\delta$ when stochasticity is sufficiently small [@Reed1979].
 Thus, all models in which the peak growth rate occurs at the same stock size will have the same optimal policy.
-While the SDP is able to discover this same policy numerically by examining the forecast probabilities, Reed's theorem makes it intuitive why that policy is shared between models 1 and 3.
+While the SDP is able to discover this same policy numerically by examining the forecast probabilities, Reed's theorem explains why that policy is shared between models 1 and 3.
+Because Reed's theorem leads to a policy that is so obviously independent of forecast skill, it is easy to forget the connection and think of it as nothing more than some arbitrary rule-of-thumb policy.
+But Reed's theorem shows this intuitive strategy is neither more nor less than the best possible policy an idealized decision maker would come up with given a perfect probabilistic forecast.
 
-However, Reed's proof considered only one-dimensional growth models $f(x)$, and so tells us nothing about models with stage structure or interacting species.
-It is easy to believe that the simple and thus non-unique optimal policy studied by @Reed1979 is an artifact of the simple growth model: surely the intuition provided will not extend to two or more species?
-Fortunately, recent work in the mathematics literature has finally been able to extend Reed's results.
-@Hening2021 proves under comparable assumptions that the optimal strategy
+It is tempting to think that the simplicity of the optimal control policy in this case is a direct consequence of the simple models involved.
+We might expect that more biologically realistic models than Eq 1-4 would always have more complex optimal control rules, which might make it much more unlikely that we would ever consider a model such as model 1 which just happened to share the optimal solution of the underlying process.
+Non-uniqueness is certainly harder to prove in more complex models. 
+For instance, Reed's results, while quite general, are nevertheless limited to one-dimensional models and therefore tell us nothing about the all-important models with stage structure or interacting species.
+Perhaps the forecast trap is merely an artifact of overly simple models -- surely we could never achieve optimal management of a real-world system with an obviously simple model?
+Without rigorous analysis, the veracity of that assertion is difficult to establish either way.
+Fortunately, recent work in mathematics has finally been able to extend Reed's results to stage structured and multi-species models [@Holden2015; @Hening2019; @Hening2021].
+Like @Reed1979, these results make it clear that the optimal control solution is not unique to the generative model: both very simple and more complex models will share the same optimal policy despite producing very different dynamics.
+Thus, the forecast trap is not limited to the one-dimensional models of @Reed1979, but can also exist when the unknown true model is a stage structured, predator prey, or competition model.
+
+In fact, non-uniqueness of the optimal solution is even more likely in real world systems due to the constraints under which most managers operate.
+While ecological processes often involve a very high-dimensional space, the space of possible actions a manager can take is often much more limited. 
+Managers may be limited to relatively static strategies, because policy adjustments are often costly [@Boettiger2016].
+Such constraints make it only more likely that the optimal solution is not unique.
+For instance, for any optimal policy is expressed as an ann total allowable catch,
+
+Such constraints only make it easier to discover simple models that would make poor forecasts but lead to the same optimal policy.
+For any value of an annual total allowable catch, there will be plenty of simple models like Model 1 whose forecasts could generate precisely the same decision policy despite terrible forecast accuracy.
+
+Reed's insights illustrate that this happens not because "bad" models sometimes get lucky, but because they capture the key features of the decision process.
+This is the goal of all modeling. 
+Models are simplifying abstractions which capture the details important to the task at hand.
+In Reed's case, that key feature is quiet intuitive -- the peak growth rate of the curve (if we ignore discounting). 
+But we do not need to know and understand that feature for it to exist.
+For example, if the true model were the predator-prey model of @Hening2021 or the stage-structured model of @Holden2015;
+the simple heuristic would not apply
+Even in the results of @Holden2015 and @Hening2021, the target stock-size of the bang-bang solution does not permit quite so concise and intuitive explanation as @Reed1979's.
+While analytic solutions for even more complex models are likely intractable, it would be rash to assume
+
+But for the reasons outlined above, it is very possible, even likely, that our candidate models include those which have the same or nearly the same optimal policy as the true model.
+And we cannot be certain that these will also be the models which give the best forecasts.
+This problem may become more likely as better and better forecasts can be achieved by statistical and machine learning models, which may be simultaneously less likely to reflect those unknown key features than the more mechanistic models they out-forecast.
+Filtering or weighting models based on forecast skill alone
+
+Other fields have already recognized the potential to achieve optimal decisions without the pursuit of ever more accurate predictive models. 
+The leading artificial intelligence algorithms used in robotic automation and game-playing [@starcraftii; @atari] no longer attempt to estimate a predictive model which is then used to select an action.
+In the past several years, so-called "Model-Free" approaches have come to dominate Deep Reinforcement Learning.
+These algorithms have no predictive capacity at all; instead, they try to learn the most effective policy rules directly.
+
+Instead of estimating a predictive model the cutting edge, algorithms which make no attempt to learn the policy directly.
+Decision makers do not need to predict the future, they only need to know what to do about it.
 

### 2021-09-15 manuscript/manuscript.Rmd
:pen:

--- 2021-09-10 manuscript/manuscript.Rmd
@@
 ---
-title: 'The forecast trap'
-- name: Carl Boettiger
-  email: cboettig@berkeley.edu
+title: "The forecast trap"
+name: Carl Boettiger
+email: "cboettig@berkeley.edu"
 affiliation: a,1
@@
 
-Each possible parameterization of the model will generate a different forecast, and thus require a it's own SDP solution to determine the corresponding policy.
-We will consider a larger suite of these logistic models in a moment, but first let us draw out only two such parameterizations from this larger suite.
-This will let us focus our attention on two of the most interesting regions of that are already included within that larger parameter space of all possible values for $r$, $K$, and $\sigma$.
-Thus, let us take "Model 1" as being given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, "Model 2" by $r_2 = 0.5$, $K_2 = 10$, $\sigma_2 = 0.075$. 
+Before this or any other model can generate a forecast, we must first come up with some parameter estimates.
+Because the model includes (log-normal) stochastic growth, no amount of data will make any parameter combination impossible, though certain parameter values are more likely than others.
+Remember too that parameter estimates may be derived in other ways than than model fitting, especially when parameters are amenable to biological interpretation. 
+We will consider the whole range of possible parameters in a moment, but for simplicity, let us begin by focusing in around two of the most interesting regions of that are already included within that larger parameter space of all possible values for $r$, $K$, and $\sigma$.
+Let us take "Model 1" as being given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, "Model 2" by $r_2 = 0.5$, $K_2 = 10$, $\sigma_2 = 0.075$. 
 We can imagine our comparison of these two models as a microcosm of the larger comparison between all possible paremeterizations.
@@
 Passive adaptive management for a simple sequential decision problem is straightforward to implement over a discrete set of states and actions using dynamic programming with iterative updates [@Smith1981, example code in Appendix].
-To demonstrate that the behavior is not driven by failure to explore sufficiently, (which might be addressed by an active adaptive management), I will assign initial probability that model 2 is true at 1%.
+To demonstrate that the behavior is not driven by failure to explore sufficiently, (active adaptive management), I will assign initial probability that model 2 is true at 1%.
 After a single iteration of adaptive learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct [Fig 3A].
@@
 
-Perhaps this should not be surprising: ecologists have long observed that all models are wrong and the choice of better model depends on the the modeling goals [@Levins1966; @Walters1978; @Ludwig1993; @Getz2017].
-And we are clearly considering different goals: forecast skill (a unitless statistical measure) vs policy outcomes (be they measured in dollars or fish in the ocean).
-Yet the result is surprising all the same. 
-The forecast isn't just some other arbitrary modeling objective; it is a central input into the decision making process, both in the real world [@Clark2001; @Deitze2018] and in our idealized decision-making algorithm, SDP [@Marescot2013].
-Nor can we say the same model can never be best at both goals -- obviously the 'true' model always optimizes both objectives [^1]. 
-It is natural to assume from this that the candidate with the closest forecast will also be the one with the closest policy.
-The example presented here proves this is by no means guaranteed, but also begs the question -- how common is this forecast trap?
-
-[^1]: With enough data from enough of the state space, an SDP algorithm using a Gaussian Process prior [@Boettiger2015], which spans the "true model" given by Eq 4, will escape the forecast trap, as guaranteed by @Gneiting2007's theorem.
-However, real ecological systems are much more complex than Eq 4, and not so easily spanned by mathematical models.
-
-The forecast trap can occur whenever two conditions are met: 
-(1) our candidate models do not include the true model, and
-(2) the optimal policy of the true model is not unique to that model.
-The first condition is widely accepted as common-place -- ecological processes will always be more complex than our models [@Getz2017].  
-The second condition is more subtle: the fact that both curves for model 1 and model 3 produce exactly the same management policy (Fig 4B) under the SDP algorithm may be surprising.
-Fortunately, the models we have chosen to focus on here are simple enough to be analytically tractable -- that is, we can solve for the optimal policy without the brute-force numerical approach of SDP.
-That solution was first provided by @Reed1979, who demonstrated the optimal policy would be a so-called "bang-bang" strategy, in which the manager seeks to bring the fish stock to some target biomass as quickly as possible whenever it deviates from that size.
-The precise value of that target biomass depends of course on the details of the model -- but importantly not all the details.
-Intuitively one can think of this as maintaining the biomass at the most productive size: the maximum population growth rate (position of the peak of the growth curves in Fig 4A), though this is only precisely true without discounting ($\delta = 1$): the optimal stock size $\hat x$ is the solution to $f(\hat x) = \hat x/\delta$ when stochasticity is sufficiently small [@Reed1979].
-Thus, all models in which the peak growth rate occurs at the same stock size will have the same optimal policy.
-While the SDP is able to discover this same policy numerically by examining the forecast probabilities, Reed's theorem makes it intuitive why that policy is shared between models 1 and 3.
-
-However, Reed's proof considered only one-dimensional growth models $f(x)$, and so tells us nothing about models with stage structure or interacting species.
-It is easy to believe that the simple and thus non-unique optimal policy studied by @Reed1979 is an artifact of the simple growth model: surely the intuition provided will not extend to two or more species?
-Fortunately, recent work in the mathematics literature has finally been able to extend Reed's results.
-@Hening2021 proves under comparable assumptions that the optimal strategy
-
-<!---
-The explanation comes from noticing that the stock size corresponding to the maximum growth rate under model 1 (the peak of the curve) falls at almost exactly the same stock size as that of the peak growth rate for model 3.
-Meanwhile, the peak of model 2 occurs at a substantially lower stock size.
-While the optimal control solution appears to depend only on a step-ahead forecast accuracy (recall that the SDP solution method used here takes only step-ahead forecast probabilities as input, [@Marescot2013]), mathematical analysis showed long ago [@Reed1979] that the optimal solution for this problem depends only on keeping the biomass at the value responsible for the maximum growth rate.[^1]
-Consequently, many models can share the same optimal solution, regardless of their forecast performance.
-
-This result may appear surprising, and begs the question as to whether the non-uniqueness of the optimal solution is not merely an artifact of the simple models considered here. 
-Surely more any realistically complex model would require a comparably more complex management policy such that the likelihood that any of our candidate models shares the same optimal policy would be vanishingly small?
-The intuition sounds reasonable
-
-[^1]: This is only approximately accurate. @Reed1979 demonstrates 
--->
-
-This realization is quite general: for most decision problems, simple models will exist under which the optimal decision is the same as it would be for the true model, even when that simple model is wrong in most other ways.
-
-<!-- Intuition for it's not just this example -->
-Situations similar to this example, in which a model can lead to the optimal decision while producing a poor forecast, are likely to be widespread.
-
-<!--
-For example, marine fisheries typically restrict a Management Strategy Evaluation [MSE; @Punt2016] to constant mortality, (defined as harvest per unit biomass, $F = H/B$).
-Constant mortality is technically only optimal in a deterministic model [@Clark1973]; but is easier to solve for than the SDP methods for constant escapement shown here,
-which in practice is used mostly in salmon fisheries and the resource economics literature [see @pomdp-intro, @Costello2016].
-While Model 1 used here will not lead to a good constant-mortality policy, it is just as easy to construct a model which gives optimal decisions and bad forecasts.
-In the example shown above, Model 1 achieves optimal results by choosing $K$ such that $K/2$ matches the peak growth rate under the true model [@Reed1979.
-Under constant mortality, we instead choose a logistic model with $r$ such that $r/4$ matches the optimal constant mortality under the true model [@Schaefer1954].
-Everything else about our choice of Model 1 can be far from the true model, and yet it will give optimal results, just as above.
--->
-
 ```{r figure4, fig.cap="Panel A: Population growth curves of each model.
@@
 
-This example demonstrates that it is possible for a model to give quite good forecasts while leading to marginal outcomes, and conversely, that is possible for a model to give optimal outcomes while making terrible forecasts.
-This result may be particularly surprising given that the policy is derived directly from the forecast.
+Perhaps this should not be surprising: ecologists have long observed that all models are wrong and the choice of better model depends on the the modeling goals [@Levins1966; @Walters1978; @Ludwig1993; @Getz2017].
+And we are clearly considering different goals: forecast skill (a unitless statistical measure) vs policy outcomes (be they measured in dollars or fish in the ocean).
+Yet the result is surprising all the same. 
+The forecast isn't just some other arbitrary modeling objective; it is a central input into the decision making process, both in the real world [@Clark2001; @Deitze2018] and in our idealized decision-making algorithm, SDP [@Marescot2013].
+Nor can we say the same model can never be best at both goals -- obviously the 'true' model always optimizes both objectives [^1]. 
+It is natural to assume from this that the candidate with the closest forecast will also be the one with the closest policy.
+The example presented here proves this is by no means guaranteed, but also begs the question -- how common is this forecast trap?
 
-In most realistic management scenarios, we do not have the benefit of a theorem similar to Reed's to identify what features are most critical to approximate well and what are not.
+[^1]: With enough data from enough of the state space, an SDP algorithm using a Gaussian Process prior [@Boettiger2015], which spans the "true model" given by Eq 4, will escape the forecast trap, as guaranteed by @Gneiting2007's theorem.
+However, real ecological systems are much more complex than Eq 4, and not so easily spanned by mathematical models.
 
-For instance, the theorem no longer holds if stock estimates are uncertain [@pomdp-intro].
-Absent the guidance of such results, iterative learning over even relatively rich possible candidate models can move us towards worse outcomes than not learning at all.
+It may seem reasonable to expect that the forecast trap would be rare in real world situations:
+the chance that the candidate set of models would include anything coming close to the optimal policy of the (unknown) true model without also providing a good forecast seems like it ought to be vanishingly small.
+In our example, we were only able to capture the optimal policy with a Gordon-Schafer model because it turned out the optimal policy boiled down to a very simple rule. 
+Surely this does not happen in the more complex models of the real world?
+Surprisingly, more complex models offer no such guarantee, while real-world constraints make this situation *more* likely, not less.
+
+How can the very different forecasts from model 1 and model 3 could produce exactly the same optimal management policy (Fig 4B) under the SDP algorithm?
+Analytic solutions offer more insight as to when and why very different forecasts can generate the identical policy.
+Such a solution was first provided by @Reed1979, who demonstrated the optimal policy in the case considered here would be a so-called "bang-bang" policy.
+Intuitively one can think of this as maintaining the biomass at the most productive size: the maximum population growth rate (position of the peak of the growth curves in Fig 4A), though this is only precisely true without discounting ($\delta = 1$): the optimal stock size $\hat x$ is the solution to $f(\hat x) = \hat x/\delta$ when stochasticity is sufficiently small [@Reed1979].
+Thus, all models in which the peak growth rate occurs at the same stock size will have the same optimal policy.
+These are not merely "bad models getting lucky" -- all such models correctly capture the crucial feature relevant to the decision.
+In more complex models, such features are more difficult or impossible to identify analytically; but just because we cannot intuit the optimal policy does not mean it is uniquely complex.
+
+Do more complex models lead to more complex control rules?
+The theorems of @Reed1979, while quite general, say nothing about structured models or those with predator-prey or competitive interactions.
+Yet recent mathematical breakthroughs such as @Holden2015, @Hennig2019, and @Henning2021 have finally been able to extend Reed's theorem to such cases more generally. 
+As with Reed's result, these recent proofs make it clear that the optimal strategy for managing these more complex models is not unique, but will be shared by many much simpler models.
+So while the true population dynamics may be given by very complex non-linear functions of the interacting species, there will be simple two species models and even un-coupled, one-species models which would lead to the identical optimal policy.
+It may be more likely that our candidate set contains a model which matches the optimal policy than that it contains a model which matches the generative process.
+
+Real world considerations actually make this situation more likely, not less likely.
+Managers cannot resort to arbitrarily complex policies, regardless of how complex their models.  
+Policy adjustments are costly [@Boettiger2016] and the space of available actions is usually far more limited than the space of available states.
+In some ecological decision-making contexts, a manager may only be able to select between a handful of alternative actions.
+Such constraints make it much more likely that the optimal policy will be replicated by a much simpler model.
+Most well-managed marine fisheries are constrained to constant or simple piecewise-linear harvest control rules [@Punt2010]. 
+This ensures that an infinite number of possible models will share the optimal solution with the true model.
+
+<!--
+Reed's insights illustrate that this happens not because "bad" models sometimes get lucky, but because they capture the key features of the decision process.
+This is the goal of all modeling: 
+models are simplifying abstractions which capture the details important to the task at hand.
+In Reed's case, that key feature is quiet intuitive -- the peak growth rate of the curve (if we ignore discounting). 
+That simple intuition does not extend to the more complex cases examined by @Hening2021 and certainly in even richer models where the optimal policy is not tractable.
+Consequently, these results do not provide a short-cut for finding an optimal policy without knowing the true model: we cannot merely look for the harvest that generates the peak growth rate (even in Reed's case that would not be optimal due to the role of discounting).
+All the same, they are a reminder of why simple, processed based models can lead us to the optimal policy without providing accurate forecasts.
+Meanwhile, filtering out or weighing models based on forecast skill _alone_ can precipitate the forecast trap by decreasing any prior belief in such simple but effective models in favor of those with greater forecast skill.
+-->
+
+The forecast trap may become a more acute issue as the simple, process-based models that have historically underpinned ecology and conservation policy are challenged by more accurate forecasts from statistical and machine learning tools.
+Buoyed by the rapid expansion of available data and computational power, complex and increasingly opaque models are becoming more common [].
+Evaluating such models based on forecast skill will not only reduce concerns about over-fitting, but will make a compelling illustration of their viability as a tool for informing policy.
+In many cases, these more accurate forecasts may very well prove invaluable in delivering better (or less bad) real-world outcomes.
+But this paper is a reminder that such outcomes are by no means an inevitable consequence of better forecasts.
+These more accurate predictions are still not the true model, and it so it is always possible to improve predictive accuracy while less accurately reflecting the unknown key features that really drive the policy decision.
+I hope this simple and intuitive logic of Figure 4 will provide a ready reminder as to why the model that produces the best forecast will not always produce the best decision.
+
+<!--
+Other fields have already recognized the potential to achieve optimal decisions without the pursuit of ever more accurate predictive models. 
+The leading artificial intelligence algorithms used in robotic automation and game-playing [@starcraftii; @atari] no longer attempt to estimate a predictive model which is then used to select an action.
+In the past several years, so-called "Model-Free" approaches have come to dominate Deep Reinforcement Learning.
+These algorithms have no predictive capacity at all; instead, they try to learn the most effective policy rules directly.
+Instead of estimating a predictive model the cutting edge, algorithms which make no attempt to learn the policy directly.
+Decision makers do not need to predict the future, they only need to know what to do about it.
+-->
+
+<!--
+For example, marine fisheries typically restrict a Management Strategy Evaluation [MSE; @Punt2016] to constant mortality, (defined as harvest per unit biomass, $F = H/B$).
+Constant mortality is technically only optimal in a deterministic model [@Clark1973]; but is easier to solve for than the SDP methods for constant escapement shown here,
+which in practice is used mostly in salmon fisheries and the resource economics literature [see @pomdp-intro, @Costello2016].
+While Model 1 used here will not lead to a good constant-mortality policy, it is just as easy to construct a model which gives optimal decisions and bad forecasts.
+In the example shown above, Model 1 achieves optimal results by choosing $K$ such that $K/2$ matches the peak growth rate under the true model [@Reed1979.
+Under constant mortality, we instead choose a logistic model with $r$ such that $r/4$ matches the optimal constant mortality under the true model [@Schaefer1954].
+Everything else about our choice of Model 1 can be far from the true model, and yet it will give optimal results, just as above.
+-->
+
+<!--
 
@@
 In the scenario considered here, the VOI is negative: any learning over the proposed model uncertainty leads to lower expected net utility than not learning.
+-->
 
+<!--
 Worse, absent the counter-factual of non-adaptive management to compare to, this deterioration in outcomes is invisible. 
@@
 If so, How would we know, and what should we do about it?
-
-<!-- **Is it just a unicorn?** -->
-Do models which give nearly optimal performance while at the same time making wildly wrong predictions really exist?
-It is tempting to argue that model 1 is a kind of unicorn, a mythical creature existing only in theory. But that is a difficult assertion to prove.
-Examples of serious forecasts that widely fail to predict future observations abound wherever forecasts are common, from elections to economics to environmental change [e.g. @Tetlock2015].
-If there is no shortage of bad forecasts, then could any of them be useful? 
-The premise that a model need not perfectly capture reality to be useful is at the very heart of modeling. 
-It is also important to note that in many cases, any models being used in decision-making are not necessarily being subjected to the rigorous evaluation and updating steps proposed by adaptive management [@Walters1978] or iterative forecasting [@Dietze2018].
-This makes it more likely that such models could persist in practice until now. 
-While iterations that revise these models will no doubt improve decision outcomes in many cases, it is worth bearing in mind from the example here that such improvement is not guaranteed.
-
-<!-- So what? -->
-If these unicorns do exist in real world management, then what do we do about them?
-I believe this is an open question, but that our first step must be to recognize it as such.
-We have seen that the problem cannot be resolved by more data alone, and is not the result of over-fitting.
-Introducing more models into our candidate set is only helpful to the extent that such models can capture features important to the decision, which as we have seen can be quite different from the feature which matter most to prediction.
-We have also seen how not updating our uncertainty estimates, or doing so less frequently, can reveal a unicorn model before it is discounted by further iterations.
-Approaches such as iterative forecasting or adaptive management that can reduce model uncertainty over time remain promising and important techniques, and essential strategies to avoid the very real problem of model over-fitting.
-But their very success in selecting models which give the most accurate predictions can also be trap.
-How that is best done in general is an open question.
+-->
 

### 2021-09-15 manuscript/manuscript.Rmd
:pen:

@@
 
-While methods such as iterative forecasting [@Deitze2018] and adaptive management [@Walters1978] can be _applied_ to real-world using empirical data, we can only _evaluate_ their potential in hypothetical examples when the true model is known, e.g.
+While methods such as iterative forecasting [@Dietze2018] and adaptive management [@Walters1978] can be _applied_ to real-world using empirical data, we can only _evaluate_ their potential in hypothetical examples when the true model is known, e.g.
 through numerical simulation.
@@
 Yet the result is surprising all the same. 
-The forecast isn't just some other arbitrary modeling objective; it is a central input into the decision making process, both in the real world [@Clark2001; @Deitze2018] and in our idealized decision-making algorithm, SDP [@Marescot2013].
+The forecast isn't just some other arbitrary modeling objective; it is a central input into the decision making process, both in the real world [@Clark2001; @Dietze2018] and in our idealized decision-making algorithm, SDP [@Marescot2013].
 Nor can we say the same model can never be best at both goals -- obviously the 'true' model always optimizes both objectives [^1]. 
@@
 The theorems of @Reed1979, while quite general, say nothing about structured models or those with predator-prey or competitive interactions.
-Yet recent mathematical breakthroughs such as @Holden2015, @Hennig2019, and @Henning2021 have finally been able to extend Reed's theorem to such cases more generally. 
+Yet recent mathematical breakthroughs such as @Holden2015, @Hening2019, and @Hening2021 have finally been able to extend Reed's theorem to such cases more generally. 
 As with Reed's result, these recent proofs make it clear that the optimal strategy for managing these more complex models is not unique, but will be shared by many much simpler models.

### 2021-09-15 notebook/state-tipping-decisions.Rmd
closing branch :construction:

@@
   }
-# Matches the optimal policy under these settings (depends on sigma, among others)
-logistic <- 
-  function(x, h = 0){
-    y = x + h
-    r <- p$r - .9
-    K <- p$K - 4
-    pmax(
-      y + y * r * (1 - y / K),  
-      0)
-  }
+
 logistic_2 <- function(x, h = 2) logistic(x,h)
@@
 ```{r}
+## True model is May,
+alt2  <- function(x, h = 0, r = .05, K = 10){
+  s <- pmax(x - h, 0)
+  s + s ^ 2 * r * (1 - s / K)
+}
 
-landscape <- map_dfr(list(logistic_1, alt, may), 
+models <- list(logistic_1 = logistic_1, alt = alt, may =  may, alt2=alt2)
+landscape <- map_dfr(models, 
                      function(f) tibble(state = states, f = f(states) -states),
@@
 
-landscape %>% ggplot(aes(state, f, col=model)) + geom_point() + ylim(-5,10) + xlim(0,5)
+landscape %>% ggplot(aes(state, f, col=model)) + geom_point(alpha=0.8) + geom_line() + ylim(-5,10) + xlim(0,5)
 ```
@@
 ```{r}
-matrices <- lapply(list(logistic_1, alt, may), 
+matrices <- lapply(models, 
        function(f) fisheries_matrices(states, actions, observations, reward_fn, 
@@
 ```{r}
-df %>% ggplot(aes(state, action, col=model)) + geom_line(alpha=0.5)
+df %>% ggplot(aes(state, action, col=model)) + geom_line(alpha=0.7) + geom_point(alpha=0.6) + ggtitle("Policy Functions")
 
@@
 
-```{r}
+```{r fig.cap="Under the true model, agressive conservation in Year 1 is optimal, restoring the system to the desirable stable point"}
 m <- matrices[[3]]
-x0 = which.min(abs(states - 0.05))
+x0 = which.min(abs(states - 3.5))
 sim <- mdp_planning(m$transition, m$reward, discount, model_prior = c(1), 
@@
   ggplot(aes(time, state)) + geom_point()+geom_path() + 
-  geom_point(aes(time, action), col="blue")
+  geom_point(aes(time, action), col="blue") +
+  ggtitle("Optimal management under the true model")
 
@@
 
-unmanaged
+```{r }
+m <- matrices[[3]] # generative process
+x0 = which.min(abs(states - 3.5))
+sim <- mdp_planning(m$transition, m$reward, discount, model_prior = c(1), 
+                   policy = solns[[2]]$policy, x0 = x0, Tmax = 100)
 
-```{r}
-nothing <- rep(1, length(states))
-m <- matrices[[3]]
-x0 = which.min(abs(states - 17))
-sim <- mdp_planning(m$transition, m$reward, discount, model_prior = c(1), 
-                   policy = nothing, x0 = x0, Tmax = 1000)
 sim %>% mutate(state = states[state], action = actions[action]) %>% 
   ggplot(aes(time, state)) + geom_point()+geom_path() + 
-  geom_point(aes(time, action), col="blue")
+  geom_point(aes(time, action), col="blue") +
+  ggtitle("Optimal management under the true model")
 
@@
     geom_line(data=dplyr::filter(df, near(a, 0.215)), col = "purple", lwd = 1) +
-    geom_hline(aes(yintercept = 0)) + coord_cartesian(ylim = c(-.2, .1))
+    geom_hline(aes(yintercept = 0)) + coord_cartesian(ylim = c(-.2, .1), xlim = c(0,1.5))
 ```

### 2021-09-15 notebook/tipping-decisions.Rmd
closing branch :construction:

--- 2021-05-18 notebook/tipping-decisions.Rmd
@@
 ```{r}
-states <- seq(0,2, length=100)
-actions <- seq(0,2, length.out = 100)
+states <- seq(0,30, length=200)
+actions <- seq(0,1, length.out = 100)
 observations <- states
@@
 benefit <- 1
-cost <- 1 # quadratic costs
+cost <- 1
 reward_fn <- function(x,h) benefit * x - (h ^ cost)
-discount <- 0.999
+discount <- 0.99
 
-p <- list(r =.7, K = 1.2, q = 3, b = 0.15, a = 0.2) # lower-state peak is optimal
-#p <- list(r =.7, K = 1.5, q = 3, b = 0.15, a = 0.2) # higher-state peak is optimal
+#p <- list(r =.7, K = 1.2, q = 3, b = 0.15, a = 0.2) # lower-state peak is optimal
+p <- list(r = 2, K = 25, q = 4, b = 2, a = 7) # higher-state peak is optimal
 
 may <- 
-  function(x, h = 0){ # May
-    y <- x +  h
+  function(y, h=0){ # May
+
     pmax(
       ## controlling h is controlling the bifurcation point directly...
-      y + y * p$r * (1 - y / p$K)  - p$a * y ^ p$q / (y ^ p$q + p$b ^ p$q),  
+      y + y * (p$r+h) * (1 - y / p$K)  - (p$a + h) * y ^ p$q / (y ^ p$q + p$b ^ p$q),
       0)
@@
 logistic_1 <- 
-  function(x, h = 0){
-    y <- x +  h
-    r <- p$r - 0.09
-
+  function(y, h = 0){
     pmax(
-      y + y * r * (1 - y / p$K),  
+      y + y * (p$r+h) * (1 - y / p$K),
       0)
@@
 
-logistic_2 <- 
-  function(x, h = 0){
-    y <- x +  h
-    r <- p$r + .5
-    pmax(
-      y + y * r * (1 - y / p$K),  
-      0)
-  }
+may_2 <- function(x, h = .5) may(x,h)
+may_10 <- function(x, h = 1) may(x,h)
 ```
@@
 
-landscape <- map_dfr(list(logistic_1, logistic_2, may), 
-                     function(f) tibble(state = states, f = f(states, 0) -states),
+landscape <- map_dfr(list(logistic_1, may, may_2, may_10), 
+                     function(f) tibble(state = states, f = f(states) -states),
                      .id = "model")
 
-landscape %>% ggplot(aes(state, f, col=model)) + geom_point()
+landscape %>% ggplot(aes(state, f, col=model)) + geom_line()
 ```
@@
 ```{r}
-matrices <- lapply(list(logistic_1, logistic_2, may), 
+matrices <- lapply(list(logistic_1, may), 
        function(f) fisheries_matrices(states, actions, observations, reward_fn, 
@@
 ```{r, results="hide"}
-solns <- lapply(matrices, function(m) mdp_value_iteration(m$transition, m$reward, discount))
+#solns <- lapply(matrices, function(m) mdp_value_iteration(m$transition, m$reward, discount, epsilon = 0.01, max_iter = 5000))
+
+solns <- lapply(matrices, function(m) mdp_policy_iteration(m$transition, m$reward, discount, policy0 = rep(1,length(states)), max_iter = 5000))
 ```
@@
   .id = "model")
-
-df
 ```
@@
 ```{r}
-x0 = which.min(abs(states - 0.05))
+m <- matrices[[2]]
+x0 = 100
 sim <- mdp_planning(m$transition, m$reward, discount, model_prior = c(1), 
-                   policy = soln$policy, x0 = x0, Tmax = 100)
-sim %>% mutate(state = states[state], action = actions[action]) %>% 
+                   policy = solns[[2]]$policy, x0 = x0, Tmax = 100)
+sim %>% mutate(state = states[state], action = 20*actions[action]) %>% 
   ggplot(aes(time, state)) + geom_point()+geom_path() + 
   geom_point(aes(time, action), col="blue")
+
 ```
@@
     geom_line(data=dplyr::filter(df, near(a, 0.215)), col = "purple", lwd = 1) +
-    geom_hline(aes(yintercept = 0)) + coord_cartesian(ylim = c(-.2, .1))
+    geom_hline(aes(yintercept = 0)) + coord_cartesian(ylim = c(-.2, .1), xlim=c(0,1.5))
 ```

### 2021-09-16 manuscript/manuscript.Rmd

@@
 
-- abbreviated running title: "The forecast trap"
+- abbreviated running title: "The Forecast Trap"
 - Authorship statement: _This is a single-author paper_
 - data accessibility statement: _All simulation data generated for analyses here, along with code required for the analysis, is available in the Zenodo Data archive, https://doi.org/10.5281/zenodo.4660621_
-- number of words in the abstract: 116
-- number of words in main text: 4998
+- number of words in the abstract: 172
+- number of words in main text: 4967
 - number of cited references: 33
@@
 ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
-
 scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
@@
 After a single iteration of adaptive learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct [Fig 3A].
-
 As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating, which achieves a net present value that only 31% that expected under management using model 1 alone [Fig 2].
@@
 
-Perhaps this should not be surprising: ecologists have long observed that all models are wrong and the choice of better model depends on the the modeling goals [@Levins1966; @Walters1978; @Ludwig1993; @Getz2017].
+Perhaps this should not be surprising: ecologists have long observed that all models are wrong and the choice of better model depends on the the modeling goals [@Levins1966; @Walters1978; @Ludwig1993; @Getz2018].
 And we are clearly considering different goals: forecast skill (a unitless statistical measure) vs policy outcomes (be they measured in dollars or fish in the ocean).
@@
 Thus, all models in which the peak growth rate occurs at the same stock size will have the same optimal policy.
-These are not merely "bad models getting lucky" -- all such models correctly capture the crucial feature relevant to the decision.
+These are not merely bad models getting lucky -- all such models correctly capture the crucial feature relevant to the decision.
 In more complex models, such features are more difficult or impossible to identify analytically; but just because we cannot intuit the optimal policy does not mean it is uniquely complex.
@@
 
-<!--
-Reed's insights illustrate that this happens not because "bad" models sometimes get lucky, but because they capture the key features of the decision process.
-This is the goal of all modeling: 
-models are simplifying abstractions which capture the details important to the task at hand.
-In Reed's case, that key feature is quiet intuitive -- the peak growth rate of the curve (if we ignore discounting). 
-That simple intuition does not extend to the more complex cases examined by @Hening2021 and certainly in even richer models where the optimal policy is not tractable.
-Consequently, these results do not provide a short-cut for finding an optimal policy without knowing the true model: we cannot merely look for the harvest that generates the peak growth rate (even in Reed's case that would not be optimal due to the role of discounting).
-All the same, they are a reminder of why simple, processed based models can lead us to the optimal policy without providing accurate forecasts.
-Meanwhile, filtering out or weighing models based on forecast skill _alone_ can precipitate the forecast trap by decreasing any prior belief in such simple but effective models in favor of those with greater forecast skill.
--->
+Because even complex models frequently have simple control rules or else we are constrained to simple control rules as a practical matter, it is much more likely that any set of candidate models includes parameterizations that reproduce the true optimal policy than that they reproduce the optimal forecast.
+When data is initially limited, those simple models that could generate optimal outcomes may have non-negligible probabilities associated with them.
+This sets the stage for the forecast trap, adjustments with additional data or comparisons against alternative models that produce more accurate forecasts erodes those probabilities in favor of models which lead to less desirable outcomes.
 
 The forecast trap may become a more acute issue as the simple, process-based models that have historically underpinned ecology and conservation policy are challenged by more accurate forecasts from statistical and machine learning tools.
-Buoyed by the rapid expansion of available data and computational power, complex and increasingly opaque models are becoming more common [].
+Buoyed by the rapid expansion of available data and computational power, complex and increasingly opaque models are becoming more common [@Desjardins2019].
 Evaluating such models based on forecast skill will not only reduce concerns about over-fitting, but will make a compelling illustration of their viability as a tool for informing policy.
@@
 These more accurate predictions are still not the true model, and it so it is always possible to improve predictive accuracy while less accurately reflecting the unknown key features that really drive the policy decision.
-I hope this simple and intuitive logic of Figure 4 will provide a ready reminder as to why the model that produces the best forecast will not always produce the best decision.
+I hope this simple and intuitive example will provide a ready reminder as to why the model that produces the best forecast will not always produce the best decision.
 
@@
 
-<!--
-For example, marine fisheries typically restrict a Management Strategy Evaluation [MSE; @Punt2016] to constant mortality, (defined as harvest per unit biomass, $F = H/B$).
-Constant mortality is technically only optimal in a deterministic model [@Clark1973]; but is easier to solve for than the SDP methods for constant escapement shown here,
-which in practice is used mostly in salmon fisheries and the resource economics literature [see @pomdp-intro, @Costello2016].
-While Model 1 used here will not lead to a good constant-mortality policy, it is just as easy to construct a model which gives optimal decisions and bad forecasts.
-In the example shown above, Model 1 achieves optimal results by choosing $K$ such that $K/2$ matches the peak growth rate under the true model [@Reed1979.
-Under constant mortality, we instead choose a logistic model with $r$ such that $r/4$ matches the optimal constant mortality under the true model [@Schaefer1954].
-Everything else about our choice of Model 1 can be far from the true model, and yet it will give optimal results, just as above.
--->
-
-<!--
-
-That potential for long-term deterioration in management outcomes due to adaptive learning has not been widely acknowledged before.
-Certainly the limitations of any modeling approach to improve management outcomes due to the human context of competing interests, imperfect enforcement, etc., is readily appreciated [@Ludwig1993].
-The problem here goes deeper, to the limits of models and learning from data even in a toy example free from other real world challenges.
-Previous work has acknowledged that iterative learning may not always be feasible and that it may not always be beneficial [@Polasky2011], but has failed to recognize the possibility that it can potentially be detrimental.
-For example, active adaptive management is premised on the observation that future observations may be too uninformative to distinguish between alternatives [@Walters1978].
-Others have also noted that even big improvements in predictive accuracy could have negligible improvement on the decision outcomes, which motivates quantifying the Value of Information (VOI) [@Katz1987].
-But in both cases, the worst outcome is wasted effort.
-In the scenario considered here, the VOI is negative: any learning over the proposed model uncertainty leads to lower expected net utility than not learning.
--->
-
-<!--
-Worse, absent the counter-factual of non-adaptive management to compare to, this deterioration in outcomes is invisible. 
-All the information available to our manager reinforces the conviction that this deteriorated ecological state is in fact the natural, sustainably managed equilibrium of our fish population.
-By responding in any way to future observations to reduce model uncertainty, our manager is caught in a trap of self-fulfilling prophecy:
-believing ever more firmly in models that make more accurate predictions while driving decisions that prevent the fish stock from recovering.
-The only way to avoid this trap is to _forego learning_ with each new observation, to avoid iterative updates.
-It may seem that the adaptive management approach fails because posterior probabilities are updated according to Bayes rule.
-Like iterative forecasting, this favors models which better predict the data, despite the fact that under the adaptive management approach, the overall optimization is conditioned on actual utility and not model fit.
-However, this issue is not easily avoided.
-For example, so-called greedy optimization techniques that favor actions with higher immediate reward do even worse in this context, since such an algorithm would obviously maximize the immediate harvest and therefore collapse the stock.
-Only by comparing the net utility derived from management under model 1 for many iterations to the net utility derived under model 2 after many iterations can we determine that model 1 leads to better outcomes.
-This raises several questions.
-Do such methods of learning and reducing model uncertainty ever lead us towards worse long-term outcomes in the real world? 
-If so, How would we know, and what should we do about it?
--->
-
 ## Acknowledgements

### 2021-09-17 manuscript/manuscript.Rmd
:robot:
caps
trigger Rmd build
update abstract

@@
 ---
-title: "The forecast trap"
+title: "The Forecast Trap"
 name: Carl Boettiger
@@
   Encouraged by decision makers' appetite for future information on topics ranging from elections to pandemics, and enabled by the explosion of data and computational methods, model based forecasts have garnered increasing influence on a breadth of decisions in modern society.
-Using a classic example from fisheries management, I demonstrate that selecting the model or models that produce the most accurate and precise forecast (as measured by most widely used statistical scores) can lead to decidedly worse outcomes (as measured by real-world objectives such as dollars or fish).
-This can create a forecast trap: in which the manager not only over-exploits the fishery leading to declining biomass and declining yields, but also becomes increasingly convinced that these actions are not over-exploitation but in fact consistent with the best models and data for sustainable management.
-The forecast trap is not unique to this example, but possible whenever (1) the optimal management policy is not unique to the generative process, and (2) the generative process is not itself in our candidate set of models.
+Using a classic example from fisheries management, I demonstrate that selecting the model or models that produce the most accurate and precise forecast (measured by statistical scores) can lead to decidedly worse outcomes (measured by real-world objectives).
+This can create a forecast trap, in which the outcomes such as fish biomass or economic yield decline while the manager becomes increasingly convinced that these actions are consistent with the best models and data available.
+The forecast trap is not unique to this example, but possible whenever (1) the optimal management policy is not unique to the generative process, and (2) the generative process is not in our candidate set of models.
 

### 2021-09-17 manuscript/manuscript.Rmd
abstract < 150 words

@@
 - data accessibility statement: _All simulation data generated for analyses here, along with code required for the analysis, is available in the Zenodo Data archive, https://doi.org/10.5281/zenodo.4660621_
-- number of words in the abstract: 172
+- number of words in the abstract: 149
 - number of words in main text: 4967

### 2022-01-05 notebook/alternate-models.Rmd
here we go

---
output: github_document
---

- Define a model
- Define action space. Under the assumption of fixed action, problem is not MDP and trivial to solve by sweep through possible fixed actions.

```{r}
library(tidyverse)
```

```{r}
# these dynamics are nonsense
# predator pop goes to Inf if prey is extinct
# prey = sp 1, predator= sp 2
R1 <- R2 <- 3
A1 <- A2 <- 2
B1 <- 0.5
B2 <- -0.3
Tmax <- 50
omega1 <- rnorm(Tmax, 0, 0.001)
omega2 <- rnorm(Tmax, 0, 0.001)

u1 <- u2 <- s1 <- s2 <- x1 <- x2 <- numeric(Tmax)
x1[1] <- 1
x2[1] <- 1

# or, constant strategy
u1 <- rep(.2, Tmax)
u2 <- rep(.5, Tmax)

for (t in 1:(Tmax-1)) {
    s1[t] <- x1[t] * (1 - u1[t])
    s2[t] <- x2[t] * (1 - u2[t])
    x1[t+1] <- s1[t] * exp(R1 - A1 * s1[t] - B1 * s2[t] + omega1[t])
    x2[t+1] <- s2[t] * exp(R2 - A2 * s1[t] - B2 * s1[t] + omega2[t])
}

df <- data.frame(time = 1:Tmax, x1, x2, u1, u2)

```

```{r}
ggplot(df) +
 geom_path(aes(time, x1)) + 
 geom_path(aes(time, x2), col="red")
```

## Five species

```{r}

model_5sp <- function(u_prey = 0.45, u_pred = .66) {
    Tmax <- 50
    x1 <- x2 <- x3 <- x4 <- x5 <- numeric(Tmax)
    s1 <- s2 <- s3 <- s4 <- numeric(Tmax)
    xi1 <- rnorm(Tmax, 0, 0.01)
    xi2 <- rnorm(Tmax, 0, 0.01)
    xi3 <- rnorm(Tmax, 0, 0.01)
    xi4 <- rnorm(Tmax, 0, 0.01)
    xi5 <- rnorm(Tmax, 0, 0.01)

    x1[1] <- x2[1] <- x3[1] <- 1
    x4[1] <- x5[1] <- .5

    for (t in 1:(Tmax-1)) {
        s1[t] <- x1[t] * (1 - u_prey)
        s2[t] <- x2[t] * (1 - u_prey)
        s3[t] <- x3[t] * (1 - u_prey)
        s4[t] <- x4[t] * (1 - u_pred)

        x1[t+1] = s1[t]*exp(1.0213 - s1[t] - 0.0861*s2[t] - 0.3141*s3[t] - 0.7252*s4[t] - 0.2445*x5[t] + xi1[t])
        x2[t+1] = s2[t]*exp(1.0289 - 0.4765*s1[t] - s2[t] - 0.1370*s3[t] - 0.9811*s4[t] - 0.0915*x5[t] + xi2[t])
        x3[t+1] = s3[t]*exp(1.0207 - 0.3193*s1[t] - 0.3461*s2[t] - s3[t] - 0.6367*s4[t] - 0.8716*x5[t] + xi3[t])
        x4[t+1] = s4[t]*exp(0.7252*s1[t] + 0.9811*s2[t] + 0.6367*s3[t] - s4[t] + xi4[t])
        x5[t+1] = x5[t]*exp(0.2445*s1[t] + 0.0915*s2[t] + 0.8716*s3[t] - x5[t] +  xi5[t])

    }
    df <- tibble(t = 1:Tmax, x1, x2, x3, x4, x5)
    df
}

df <- model_5sp()

df %>%
    pivot_longer(tidyselect::starts_with("x"),
                    names_to = "species",
                    values_to = "population") %>%
    ggplot(aes(t, population, col= species)) + geom_path()
```

```{r}
clip <- function(x, lower=0, upper=1) {
    x[x <= lower] <- lower
    x[x >= upper] <- upper
    x
}

utility <- function(x) {
    delta <- 0.99
    w1 <- 0.25
    w2 <- 0.25
    w3 <- 1 - (w1 + w2)
    u_prey <- clip(x[1], 0, 1)
    u_pred <- clip(x[2], 0, 1)
    df <- model_5sp(u_prey, u_pred)
    R <- 
        w1 * u_prey * (df$x1 + df$x2 + df$x3) +
        w2 * u_pred * df$x4 +
        w3 * x5
    t <- seq_along(R)
    - as.numeric(R %*% delta ^ t)
    }

-utility(c(0.5,0.5))
-utility(c(0,1))
-utility(c(1,0))

o <- optim(c(0.5,0.5), utility, method="L-BFGS-B", lower = c(0,0), upper = c(1,1))
o$par
o$value

o <- optim(c(0.5,0.5), utility, method="SANN")
clip(o$par, 0, 1)
o$value

```

- Determine n alternate models over the same state space. Consider structurally different models, ideally with structures consistent with
  common forecast / inference models

### 2022-01-05 notebook/alternate-models.Rmd
alternate "model 1" example added

@@
 
+## Five species
+
 ```{r}
-# these dynamics are nonsense
-# predator pop goes to Inf if prey is extinct
-# prey = sp 1, predator= sp 2
+
+model_5sp <- function(u_prey = 0.5,
+                      u_pred = 0.5,
+                      sigma = 0.05,
+                      x1o = 0.44, # unfushed equib
+                      x2o = 0.17,
+                      x3o = 0.186,
+                      x4o = 0.600,
+                      x5o = 0.280,
+                      Tmax = 50) {
+    x1 <- x2 <- x3 <- x4 <- x5 <- numeric(Tmax)
+    s1 <- s2 <- s3 <- s4 <- numeric(Tmax)
+    xi1 <- rnorm(Tmax, 0, sigma)
+    xi2 <- rnorm(Tmax, 0, sigma)
+    xi3 <- rnorm(Tmax, 0, sigma)
+    xi4 <- rnorm(Tmax, 0, sigma)
+    xi5 <- rnorm(Tmax, 0, sigma)
+
+    x1[1] <- x1o
+    x2[1] <- x2o
+    x3[1] <- x3o
+    x4[1] <- x4o
+    x5[1] <- x5o
+
+    for (t in 1:(Tmax-1)) {
+        s1[t] <- x1[t] * (1 - u_prey)
+        s2[t] <- x2[t] * (1 - u_prey)
+        s3[t] <- x3[t] * (1 - u_prey)
+        s4[t] <- x4[t] * (1 - u_pred)
+
+        x1[t+1] = s1[t]*exp(1.0213 - s1[t] - 0.0861*s2[t] - 0.3141*s3[t] - 0.7252*s4[t] - 0.2445*x5[t] + xi1[t])
+        x2[t+1] = s2[t]*exp(1.0289 - 0.4765*s1[t] - s2[t] - 0.1370*s3[t] - 0.9811*s4[t] - 0.0915*x5[t] + xi2[t])
+        x3[t+1] = s3[t]*exp(1.0207 - 0.3193*s1[t] - 0.3461*s2[t] - s3[t] - 0.6367*s4[t] - 0.8716*x5[t] + xi3[t])
+        x4[t+1] = s4[t]*exp(0.7252*s1[t] + 0.9811*s2[t] + 0.6367*s3[t] - s4[t] + xi4[t])
+        x5[t+1] = x5[t]*exp(0.2445*s1[t] + 0.0915*s2[t] + 0.8716*s3[t] - x5[t] +  xi5[t])
+
+    }
+    df <- tibble(t = 1:Tmax, x1, x2, x3, x4, x5)
+    df
+}
+
+```
+
+```{r}
+clip <- function(x, lower=0, upper=1) {
+    x[x <= lower] <- lower
+    x[x >= upper] <- upper
+    x
+}
+
+utility <- function(x) {
+    delta <- 0.999
+    w3 <- 0.5
+    w2 <- 0.25
+    w1 <- 1 - (w2 + w3)
+    u_prey <- clip(x[1], 0, 1)
+    u_pred <- clip(x[2], 0, 1)
+    df <- model_5sp(u_prey, u_pred)
+    R <- 
+        w1 * u_prey * (df$x1 + df$x2 + df$x3) +
+        w2 * u_pred * df$x4 +
+        w3 * df$x5
+    t <- seq_along(R)
+    - as.numeric(R %*% delta ^ t)
+    }
+
+-utility(c(0.5,0.5))
+-utility(c(0,1))
+-utility(c(1,0))
+-utility(c(0,0))
+
+o <- optim(c(0.,0.), utility, method="L-BFGS-B", lower = c(0,0), upper = c(1,1))
+o$par
+o$value
+
+o <- optim(c(0.5,0.1), utility)
+clip(o$par, 0, 1)
+o$value
+
+o <- optim(c(0.5,0.5), utility, method="SANN")
+clip(o$par, 0, 1)
+o$value
+
+```
+
+For the sake of argument, let us allow the model of Brias & Munch to represent the "true" model, which is unknown.
+As researchers, we only have access to data derived from observations from the model. 
+Let us consider a sample simulation of such data:
+
+```{r}
+df <- model_5sp(0, .6, sigma = 0.01)
+
+df %>%
+    pivot_longer(tidyselect::starts_with("x"),
+                    names_to = "species",
+                    values_to = "biomass") %>%
+    ggplot(aes(t, biomass, col= species)) + geom_path()
+```
+
+There are many other models, including process-based models, statistical models or even black-box machine learning models
+which 5 dimensional state-space and two-dimensional action space.
+
+```{r}
+df <- model_5sp(0, 0., sigma = 0.01)
+
+df %>%
+    transmute(time = t, cormorant = x5, bass = x4, herring = x1+x2+x3) %>%
+    pivot_longer(any_of(c("cormorant", "bass", "herring")),
+                    names_to = "species",
+                    values_to = "biomass") %>%
+    ggplot(aes(time, biomass, col= species)) + geom_path()
+```
+
+A reasonable single-species model for Bass achieves the optimal solution.
+The model is reasonable in the sense that it is a simple Ricker model 
+that comparably reproduces the unfished dynamics of the Bass.
+Nevertheless, it is actually quite poor at forecasting fished stock sizes.
+
+```{r}
+
+model_1sp <- function(u = 0, A = 2, R = 1.6, Tmax=50, x1o = .8, sigma = 0.01) {
+
+    omega1 <- rnorm(Tmax, 0, sigma)
+
+    s1 <- x1 <- numeric(Tmax)
+    x1[1] <- x1o
+
+    for (t in 1:(Tmax-1)) {
+        s1[t] <- x1[t] * (1 - u)
+        x1[t+1] <- s1[t] * exp(R - s1[t] * A + omega1[t])
+    }
+
+    df <- tibble(time = 1:Tmax, x1)
+}
+
+df <- model_1sp(0)
+df %>% ggplot(aes(t, x1)) + geom_path()
+
+```
+
+```{r}
+utility <- function(x) {
+    delta <- 0.999
+    u <- clip(x, 0, 1)
+    df <- model_1sp(u)
+    R <- u * df$x1 
+    t <- seq_along(R)
+    - as.numeric(R %*% delta ^ t)
+    }
+
+o <- optimize(utility, c(0,1))
+o$minimum
+```
+
+```{r}
+df <- model_1sp(o$minimum)
+df %>% ggplot(aes(t, x1)) + geom_path()
+```
+
+```{r}
 R1 <- R2 <- 3
@@
 df <- data.frame(time = 1:Tmax, x1, x2, u1, u2)
-
 ```
-
-```{r}
-ggplot(df) +
- geom_path(aes(time, x1)) + 
- geom_path(aes(time, x2), col="red")
-```
-
-## Five species
-
-```{r}
-
-model_5sp <- function(u_prey = 0.45, u_pred = .66) {
-    Tmax <- 50
-    x1 <- x2 <- x3 <- x4 <- x5 <- numeric(Tmax)
-    s1 <- s2 <- s3 <- s4 <- numeric(Tmax)
-    xi1 <- rnorm(Tmax, 0, 0.01)
-    xi2 <- rnorm(Tmax, 0, 0.01)
-    xi3 <- rnorm(Tmax, 0, 0.01)
-    xi4 <- rnorm(Tmax, 0, 0.01)
-    xi5 <- rnorm(Tmax, 0, 0.01)
-
-    x1[1] <- x2[1] <- x3[1] <- 1
-    x4[1] <- x5[1] <- .5
-
-    for (t in 1:(Tmax-1)) {
-        s1[t] <- x1[t] * (1 - u_prey)
-        s2[t] <- x2[t] * (1 - u_prey)
-        s3[t] <- x3[t] * (1 - u_prey)
-        s4[t] <- x4[t] * (1 - u_pred)
-
-        x1[t+1] = s1[t]*exp(1.0213 - s1[t] - 0.0861*s2[t] - 0.3141*s3[t] - 0.7252*s4[t] - 0.2445*x5[t] + xi1[t])
-        x2[t+1] = s2[t]*exp(1.0289 - 0.4765*s1[t] - s2[t] - 0.1370*s3[t] - 0.9811*s4[t] - 0.0915*x5[t] + xi2[t])
-        x3[t+1] = s3[t]*exp(1.0207 - 0.3193*s1[t] - 0.3461*s2[t] - s3[t] - 0.6367*s4[t] - 0.8716*x5[t] + xi3[t])
-        x4[t+1] = s4[t]*exp(0.7252*s1[t] + 0.9811*s2[t] + 0.6367*s3[t] - s4[t] + xi4[t])
-        x5[t+1] = x5[t]*exp(0.2445*s1[t] + 0.0915*s2[t] + 0.8716*s3[t] - x5[t] +  xi5[t])
-
-    }
-    df <- tibble(t = 1:Tmax, x1, x2, x3, x4, x5)
-    df
-}
-
-df <- model_5sp()
-
-df %>%
-    pivot_longer(tidyselect::starts_with("x"),
-                    names_to = "species",
-                    values_to = "population") %>%
-    ggplot(aes(t, population, col= species)) + geom_path()
-```
-
-```{r}
-clip <- function(x, lower=0, upper=1) {
-    x[x <= lower] <- lower
-    x[x >= upper] <- upper
-    x
-}
-
-utility <- function(x) {
-    delta <- 0.99
-    w1 <- 0.25
-    w2 <- 0.25
-    w3 <- 1 - (w1 + w2)
-    u_prey <- clip(x[1], 0, 1)
-    u_pred <- clip(x[2], 0, 1)
-    df <- model_5sp(u_prey, u_pred)
-    R <- 
-        w1 * u_prey * (df$x1 + df$x2 + df$x3) +
-        w2 * u_pred * df$x4 +
-        w3 * x5
-    t <- seq_along(R)
-    - as.numeric(R %*% delta ^ t)
-    }
-
--utility(c(0.5,0.5))
--utility(c(0,1))
--utility(c(1,0))
-
-o <- optim(c(0.5,0.5), utility, method="L-BFGS-B", lower = c(0,0), upper = c(1,1))
-o$par
-o$value
-
-o <- optim(c(0.5,0.5), utility, method="SANN")
-clip(o$par, 0, 1)
-o$value
-
-```
-
-- Determine n alternate models over the same state space. Consider structurally different models, ideally with structures consistent with
-  common forecast / inference models

### 2022-02-04 notebook/alternate-models.Rmd
WIP :construction: alternate examples

@@
     }
-    df <- tibble(t = 1:Tmax, x1, x2, x3, x4, x5)
+    df <- tibble(time = 1:Tmax, x1, x2, x3, x4, x5)
     df
@@
 }
+```
 
-utility <- function(x) {
-    delta <- 0.999
-    w3 <- 0.5
-    w2 <- 0.25
-    w1 <- 1 - (w2 + w3)
+```{r}
+
+reward <- function(df,
+                   x,
+                   w3 = 0.5,
+                   w2 = 0.25,
+                   w1 = 1 - (w2+w3),
+                   delta = 0.999
+                  ) {
     u_prey <- clip(x[1], 0, 1)
     u_pred <- clip(x[2], 0, 1)
-    df <- model_5sp(u_prey, u_pred)
     R <-
@@
     t <- seq_along(R)
-    - as.numeric(R %*% delta ^ t)
+    as.numeric(R %*% delta ^ t)
 }
 
+utility <- function(x, f = model_5sp) {
+
+    u_prey <- clip(x[1], 0, 1)
+    u_pred <- clip(x[2], 0, 1)
+    ## ought to average over many replicates!
+    df <- f(u_prey, u_pred)
+
+    # negative since optimizer minimizes
+    - reward(df, x)
+}
+
+```
+
+```{r}
 -utility(c(0.5,0.5))
@@
 
-o <- optim(c(0.5,0.5), utility, method="SANN")
-clip(o$par, 0, 1)
-o$value
+#o <- optim(c(0.5,0.5), utility, method="SANN")
+#clip(o$par, 0, 1)
+#o$value
 
@@
                     values_to = "biomass") %>%
-    ggplot(aes(t, biomass, col= species)) + geom_path()
+    ggplot(aes(time, biomass, col= species)) + geom_path()
 ```
@@
 ```{r}
-df <- model_5sp(0, 0., sigma = 0.01)
+## Three-species 'view' of the underlying dynamics, unfished
 
+df <- model_5sp(0, 0., sigma = 0.01, x1o =.1, x2o = .1, x3o = .1, x4o = .1, x5o = .1)
 df %>%
-    transmute(time = t, cormorant = x5, bass = x4, herring = x1+x2+x3) %>%
+    transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3) %>%
     pivot_longer(any_of(c("cormorant", "bass", "herring")),
@@
 
-A reasonable single-species model for Bass achieves the optimal solution.
-The model is reasonable in the sense that it is a simple Ricker model 
-that comparably reproduces the unfished dynamics of the Bass.
-Nevertheless, it is actually quite poor at forecasting fished stock sizes.
+Now consider an alternative model which ignores some of this biological complexity:
+Our alternate model has several oversimplifications:
+(1) it treat all three herring species as a single species,
+(2) it fails to capture the coupled predator-prey dynamics between bass and herring, 
+(3) it oversimplifies the cormorant dynamics, assuming the cormorant population is determined to be a fixed fraction of the herring,
+and ignoring the impact of cormorant's predation on the herring itself.
+(4) Lastly, our model will overestimate the mortality introduced on herring by a given fishing effort.
+These elements are all obviously wrong, but not so arbitrary as to be inconceivable as a candidate model.
+Researchers frequently consider models which make oversimplifications all the time, and rely on model choice processes to weed them out.
+
+The parameterization chosen for the oversimplified 3 species model can reasonably reproduce the unfished dynamics.
+Under any harvesting regime or other influence that perturbs the system significantly far from the unfished equilibrium co-existence state of the model quickly reveals the poor forecasting ability of this model.
+Nevertheless, solving for the optimal policy (under the same constraints of constant harvest fractions as we consider for the true model) results in a harvest policy which provides nearly optimal performance.
 
 ```{r}
-
-model_1sp <- function(u = 0, A = 2, R = 1.6, Tmax=50, x1o = .8, sigma = 0.01) {
+model_3sp <- function(u_prey = 0,
+                      u_predator = 0,
+                      A1 = 2,
+                      R1 = 1.6,
+                      A2 = 2.75,
+                      R2 = 1.6,
+                      Tmax = 50,
+                      x1o = .8, # Herring
+                      x2o = .6, # Bass
+                      x3o = 0.15, # Cormorant
+                      sigma = 0.01) {
 
     omega1 <- rnorm(Tmax, 0, sigma)
+    omega2 <- rnorm(Tmax, 0, sigma)
+    omega3 <- rnorm(Tmax, 0, sigma)
 
     s1 <- x1 <- numeric(Tmax)
+    s2 <- x2 <- numeric(Tmax)
+    x3 <- numeric(Tmax)
     x1[1] <- x1o
+    x2[1] <- x2o
+    x3[1] <- x3o
 
     for (t in 1:(Tmax-1)) {
-        s1[t] <- x1[t] * (1 - u)
-        x1[t+1] <- s1[t] * exp(R - s1[t] * A + omega1[t])
+        s1[t] <- x1[t] * (1 - 10*u_prey)
+        s2[t] <- x2[t] * (1 - u_predator)
+
+        x1[t+1] <- s1[t] * exp(R1 - s1[t] * A1 + omega1[t])
+        x2[t+1] <- s2[t] * exp(R2 - s2[t] * A2 + omega2[t])
+        x3[t+1] <- .25 * x1[t]
     }
 
-    df <- tibble(time = 1:Tmax, x1)
+    df <- tibble(time = 1:Tmax, x1, x4 = x2, x5 = x3, x2 = 0, x3 = 0)
 }
+```
 
-df <- model_1sp(0)
-df %>% ggplot(aes(t, x1)) + geom_path()
+```{r}
+# unfished equilibrium
+df <- model_3sp(0, 0)
+df %>%
+    transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3) %>%
+    pivot_longer(any_of(c("cormorant", "bass", "herring")),
+                    names_to = "species",
+                    values_to = "biomass") %>%
+    ggplot(aes(time, biomass, col= species)) + geom_path()
+```
+
+```{r}
+# Optimal policy under identical utility
+u_3sp <- function(x) utility(x, f = model_3sp)
+
+m3sp <- optim(c(0.5,0.1), u_3sp)
+m3sp$value
+clip(m3sp$par, 0, 1)
+
+```
+
+Alternate model, better forecast, worse decision:
+
+```{r}
+model_B <- function(u_prey = 0.5,
+                      u_pred = 0.5,
+                      sigma = 0.05,
+                      x1o = 0.796, # unfushed equib
+                      x4o = 0.600,
+                      x5o = 0.280,
+                      Tmax = 50,
+                      R1 = 1.0213,
+                      A11 = 1,
+                      A14 = 0.7252,
+                      A15 = 0.2445,
+                      A41 = 0.7252,
+                      A51 = 0.2445) {
+    x1 <- x4 <- x5 <- numeric(Tmax)
+    s1 <-  s4 <- numeric(Tmax)
+    xi1 <- rnorm(Tmax, 0, sigma)
+    xi4 <- rnorm(Tmax, 0, sigma)
+    xi5 <- rnorm(Tmax, 0, sigma)
+
+    x1[1] <- x1o
+    x4[1] <- x4o
+    x5[1] <- x5o
+
+    for (t in 1:(Tmax-1)) {
+        s1[t] <- x1[t] * (1 - u_prey)
+        s4[t] <- x4[t] * (1 - u_pred)
+        x1[t+1] = s1[t]*exp(R1 - A11*s1[t] - A14*s4[t] - A15 * x5[t] + xi1[t])
+        x4[t+1] = s4[t]*exp(A41*s1[t] - s4[t] + xi4[t])
+        x5[t+1] = x5[t]*exp(A51*s1[t] - x5[t] +  xi5[t])
+
+    }
+    df <- tibble(time = 1:Tmax, x1, x4, x5, x2 =0, x3 = 0)
+    df
+}
+```
+
+```{r}
+# Optimal policy under identical utility
+u_B <- function(x) utility(x, f = model_B)
+u_B(c(0, .63))
+#optim(c(0.5,0.1), u_B, method = "L-BFGS-B", lower=0, upper=1)
+
+B <- optim(c(0.5,0.1), u_B)
+B$value
+clip(B$par, 0, 1)
 
@@
 ```{r}
-utility <- function(x) {
-    delta <- 0.999
-    u <- clip(x, 0, 1)
-    df <- model_1sp(u)
-    R <- u * df$x1 
-    t <- seq_along(R)
-    - as.numeric(R %*% delta ^ t)
+
+utility( c(clip(o$par[[1]]), clip(o$par[[2]])) )
+
+utility( c(clip(m3sp$par[[1]]), clip(m3sp$par[[2]])) )
+
+utility( c(clip(B$par[[1]]), clip(B$par[[2]])) )
+
+```
+
+Run the simulation using management policy of each
+
+```{r}
+
+## Generate replicate predictions from each model at each step.
+## compute CRPS of an observation from the "true" model vs those distributions
+
+function(est, f) {
+    u_prey <- clip(est$par[[1]])
+    u_pred <- clip(est$par[[2]])
+    map_dfr(1:100, function(i) {
+    df <- f(u_prey, u_pred, x1o = .1, x2o = .1, x3o = .1, x4o = .1, x5o = .1)
+    df %>%
+        transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3) %>%
+        pivot_longer(any_of(c("cormorant", "bass", "herring")),
+                        names_to = "species",
+                        values_to = "biomass")
+    }, .id = "rep")
+
 }
 
-o <- optimize(utility, c(0,1))
-o$minimum
 ```
@@
 ```{r}
-df <- model_1sp(o$minimum)
-df %>% ggplot(aes(t, x1)) + geom_path()
+# management of true model under optimal pars for model_3sp
+df <- model_5sp(clip(m3sp$par[[1]],0,1),
+                clip(m3sp$par[[2]],0,1),
+                 x1o = .1, x4o = .1, x5o = .1)
+df %>%
+    transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3) %>%
+    pivot_longer(any_of(c("cormorant", "bass", "herring")),
+                    names_to = "species",
+                    values_to = "biomass") %>%
+    ggplot(aes(time, biomass, col= species)) + geom_path()
 ```
-
-```{r}
-R1 <- R2 <- 3
-A1 <- A2 <- 2
-B1 <- 0.5
-B2 <- -0.3
-Tmax <- 50
-omega1 <- rnorm(Tmax, 0, 0.001)
-omega2 <- rnorm(Tmax, 0, 0.001)
-
-u1 <- u2 <- s1 <- s2 <- x1 <- x2 <- numeric(Tmax)
-x1[1] <- 1
-x2[1] <- 1
-
-# or, constant strategy
-u1 <- rep(.2, Tmax)
-u2 <- rep(.5, Tmax)
-
-for (t in 1:(Tmax-1)) {
-    s1[t] <- x1[t] * (1 - u1[t])
-    s2[t] <- x2[t] * (1 - u2[t])
-    x1[t+1] <- s1[t] * exp(R1 - A1 * s1[t] - B1 * s2[t] + omega1[t])
-    x2[t+1] <- s2[t] * exp(R2 - A2 * s1[t] - B2 * s1[t] + omega2[t])
-}
-
-df <- data.frame(time = 1:Tmax, x1, x2, u1, u2)
-```

### 2022-02-11 notebook/alternate-models.Rmd
pretty close

--- 2022-01-05 notebook/alternate-models.Rmd
@@
 ---
+
+- Model 1: A reasonable approximation model that does not produce very accurate forecasts, but nevertheless leads to good management decisions.
+- Model 2: A reasonable approximation model that produces more accurate forecasts but leads to substantially worse management outcomes
+- Model 3: The "True" Model: five-species model; functional forms and parameter values are never disclosed to the 'manager' but are used to simulate all data and management scenarios.
 
@@
 ```{r}
-# these dynamics are nonsense
-# predator pop goes to Inf if prey is extinct
-# prey = sp 1, predator= sp 2
-R1 <- R2 <- 3
-A1 <- A2 <- 2
-B1 <- 0.5
-B2 <- -0.3
-Tmax <- 50
-omega1 <- rnorm(Tmax, 0, 0.001)
-omega2 <- rnorm(Tmax, 0, 0.001)
-
-u1 <- u2 <- s1 <- s2 <- x1 <- x2 <- numeric(Tmax)
-x1[1] <- 1
-x2[1] <- 1
-
-# or, constant strategy
-u1 <- rep(.2, Tmax)
-u2 <- rep(.5, Tmax)
-
-for (t in 1:(Tmax-1)) {
-    s1[t] <- x1[t] * (1 - u1[t])
-    s2[t] <- x2[t] * (1 - u2[t])
-    x1[t+1] <- s1[t] * exp(R1 - A1 * s1[t] - B1 * s2[t] + omega1[t])
-    x2[t+1] <- s2[t] * exp(R2 - A2 * s1[t] - B2 * s1[t] + omega2[t])
+## helper funs
+format_sim <- function(df) {
+  df %>%  
+  transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3, r=r) %>%
+  pivot_longer(any_of(c("cormorant", "bass", "herring")),
+                  names_to = "species",
+                  values_to = "biomass")
 }
-
-df <- data.frame(time = 1:Tmax, x1, x2, u1, u2)
-
-```
-
-```{r}
-ggplot(df) +
- geom_path(aes(time, x1)) + 
- geom_path(aes(time, x2), col="red")
 ```
@@
 
-```{r}
-
-model_5sp <- function(u_prey = 0.45, u_pred = .66) {
-    Tmax <- 50
+```{r model_5sp}
+# x1,x2,x3 Herring (Prey)
+# x4 Bass (predator)
+# x5 Cormorant (conservation target, predator)
+model_5sp <- function(u_prey = 0.5,
+                      u_pred = 0.5,
+                      sigma = 0.01,
+                      x1o = 0.44, # unfished equib
+                      x2o = 0.17,
+                      x3o = 0.186,
+                      x4o = 0.600,
+                      x5o = 0.280,
+                      Tmax = 50,
+                      reps = 40) {
+    df <- tibble()
+    for(r in 1:reps){
       x1 <- x2 <- x3 <- x4 <- x5 <- numeric(Tmax)
       s1 <- s2 <- s3 <- s4 <- numeric(Tmax)
-    xi1 <- rnorm(Tmax, 0, 0.01)
-    xi2 <- rnorm(Tmax, 0, 0.01)
-    xi3 <- rnorm(Tmax, 0, 0.01)
-    xi4 <- rnorm(Tmax, 0, 0.01)
-    xi5 <- rnorm(Tmax, 0, 0.01)
+      xi1 <- rnorm(Tmax, 0, sigma)
+      xi2 <- rnorm(Tmax, 0, sigma)
+      xi3 <- rnorm(Tmax, 0, sigma)
+      xi4 <- rnorm(Tmax, 0, sigma)
+      xi5 <- rnorm(Tmax, 0, sigma)
 
-    x1[1] <- x2[1] <- x3[1] <- 1
-    x4[1] <- x5[1] <- .5
+      x1[1] <- x1o
+      x2[1] <- x2o
+      x3[1] <- x3o
+      x4[1] <- x4o
+      x5[1] <- x5o
 
@@
       }
-    df <- tibble(t = 1:Tmax, x1, x2, x3, x4, x5)
+      df <- dplyr::bind_rows(tibble(time = 1:Tmax, x1, x2, x3, x4, x5, r), df)
+    }
     df
@@
 
-df <- model_5sp()
-
-df %>%
-    pivot_longer(tidyselect::starts_with("x"),
-                    names_to = "species",
-                    values_to = "population") %>%
-    ggplot(aes(t, population, col= species)) + geom_path()
 ```
 
-```{r}
+```{r utility}
 clip <- function(x, lower=0, upper=1) {
@@
 
-utility <- function(x) {
-    delta <- 0.99
-    w1 <- 0.25
-    w2 <- 0.25
-    w3 <- 1 - (w1 + w2)
+## Compute the reward associated with a given simulation df given 
+## rates for prey and predator species (x[1],x[2])
+reward <- function(df,
+                   x,
+                   w3 = 0.5,              # Conservation value
+                   w2 = 0.25,             # Predator harvest value
+                   w1 = 1 - (w2+w3),      # Prey harvest value
+                   delta = 0.999
+                  ) {
     u_prey <- clip(x[1], 0, 1)
     u_pred <- clip(x[2], 0, 1)
-    df <- model_5sp(u_prey, u_pred)
     R <-
@@
         w2 * u_pred * df$x4 +
-        w3 * x5
+        w3 * df$x5
     t <- seq_along(R)
-    - as.numeric(R %*% delta ^ t)
+    as.numeric(R %*% delta ^ t)
 }
 
--utility(c(0.5,0.5))
--utility(c(0,1))
--utility(c(1,0))
+# Function to be optimized for fixed choice of harvest rates x[1],x[2]
+utility <- function(x, f = model_5sp, ...) {
 
-o <- optim(c(0.5,0.5), utility, method="L-BFGS-B", lower = c(0,0), upper = c(1,1))
+    u_prey <- clip(x[1], 0, 1)
+    u_pred <- clip(x[2], 0, 1)
+    df <- f(u_prey, u_pred, ...)
+
+    # average accross the reps    
+    df <- df %>% group_by(time) %>% summarise(across(.fns=mean))
+
+    # negative since optimizer minimizes
+    - reward(df, x)
+}
+
+```
+
+```{r}
+
+o <- optim(c(0.01,0.5), utility, method="L-BFGS-B", lower = c(0,0), upper = c(1,1))
 o$par
@@
 
-o <- optim(c(0.5,0.5), utility, method="SANN")
+o <- optim(c(0.01,0.5), utility)
 clip(o$par, 0, 1)
@@
 
+x <- c(0.0,.58)
+utility(x)
+
 ```
 
-- Determine n alternate models over the same state space. Consider structurally different models, ideally with structures consistent with
-  common forecast / inference models
+For the sake of argument, let us allow the model of Brias & Munch to represent the "true" model, which is unknown.
+As researchers, we only have access to data derived from observations from the model. 
+Let us consider a sample simulation of such data:
+
+```{r}
+df <- model_5sp(0, .0, sigma = 0.01)
+
+df %>%
+    pivot_longer(tidyselect::starts_with("x"),
+                    names_to = "species",
+                    values_to = "biomass") %>%
+    ggplot(aes(time, biomass, group= interaction(species,r), col=species)) + geom_path() + ggtitle("Unfished")
+```
+
+Starting from the unfished equilibrium,
+
+```{r}
+df <- model_5sp(0, .58, sigma = 0.01)
+
+df %>%
+    pivot_longer(tidyselect::starts_with("x"),
+                    names_to = "species",
+                    values_to = "biomass") %>%
+    ggplot(aes(time, biomass, group= interaction(species,r), col=species)) + 
+  geom_path() + ggtitle("Optimally managed")
+```
+
+There are many other models, including process-based models, statistical models or even black-box machine learning models
+which 5 dimensional state-space and two-dimensional action space.
+
+```{r}
+## Three-species 'view' of the underlying dynamics, without fishing, starting all species from a low level
+
+#true_observations <- model_5sp(0, 0., sigma = 0.05, x1o =.033, x2o = .033, x3o = .033, x4o = .1, x5o = .1, Tmax = 50)
+true_observations <- model_5sp(0, 0, sigma = 0.05)
+
+obs <- true_observations %>% 
+  transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3, r=r) %>%
+    pivot_longer(any_of(c("cormorant", "bass", "herring")),
+                    names_to = "species",
+                    values_to = "biomass")
+
+obs %>%
+    ggplot(aes(time, biomass, col = species, group = interaction(species,r))) +
+    geom_path(alpha=0.4)
+```
+
+Now consider an alternative model which ignores some of this biological complexity:
+Our alternate model has several oversimplifications:
+(1) it treat all three herring species as a single species,
+(2) it fails to capture the coupled predator-prey dynamics between bass and herring, 
+(3) it oversimplifies the cormorant dynamics, assuming the cormorant population is determined to be a fixed fraction of the herring,
+and ignoring the impact of cormorant's predation on the herring itself.
+(4) Lastly, our model will overestimate the mortality introduced on herring by a given fishing effort.
+These elements are all obviously wrong, but not so arbitrary as to be inconceivable as a candidate model.
+Researchers frequently consider models which make oversimplifications all the time, and rely on model choice processes to weed them out.
+
+The parameterization chosen for the oversimplified 3 species model can reasonably reproduce the unfished dynamics.
+Under any harvesting regime or other influence that perturbs the system significantly far from the unfished equilibrium co-existence state of the model quickly reveals the poor forecasting ability of this model.
+Nevertheless, solving for the optimal policy (under the same constraints of constant harvest fractions as we consider for the true model) results in a harvest policy which provides nearly optimal performance.
+
+```{r}
+model_3sp <- function(u_prey = 0,
+                      u_predator = 0,
+                      A1 = 2,
+                      R1 = 1.6,
+                      A2 = 2.75,
+                      R2 = 1.45,
+                      Tmax = 50,
+                      x1o = 0.44, # unfished equib
+                      x2o = 0.17,
+                      x3o = 0.186,
+                      x4o = 0.600,
+                      x5o = 0.280,
+                      sigma = 0.01,
+                      reps=40) {
+    df <- tibble()
+    for(r in 1:reps){
+    omega1 <- rnorm(Tmax, 0, sigma)
+    omega2 <- rnorm(Tmax, 0, sigma)
+    omega3 <- rnorm(Tmax, 0, sigma)
+
+    s1 <- x1 <- numeric(Tmax)
+    s4 <- x4 <- numeric(Tmax)
+    x2 <- x3 <- numeric(Tmax)
+    x5 <- numeric(Tmax)
+
+    x1[1] <- x1o + x2o + x3o
+    x4[1] <- x4o
+    x5[1] <- x5o
+
+    for (t in 1:(Tmax-1)) {
+        s1[t] <- x1[t] * (1 - 10*u_prey)
+        s4[t] <- x4[t] * (1 - u_predator)
+
+        x1[t+1] <- s1[t] * exp(R1 - s1[t] * A1 + omega1[t])
+        x4[t+1] <- s4[t] * exp(R2 - s4[t] * A2 + omega2[t])
+        x5[t+1] <- .25 * x1[t]
+    }
+
+    df <- bind_rows(df, tibble(time = 1:Tmax, x1, x2, x3, x4, x5, r=r))
+    }
+  df
+}
+```
+
+Does this model produce good forecasts?
+
+```{r}
+# unfished equilibrium
+m1_forecast <- model_3sp(0, 0.0, sigma = 0.05,  Tmax = 50)
+
+predictions <- m1_forecast %>%
+  transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3, r=r) %>%
+  pivot_longer(any_of(c("cormorant", "bass", "herring")),
+                  names_to = "species",
+                  values_to = "biomass") %>% 
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass))
+
+predictions %>%
+  ggplot(aes(time, biomass, col= species)) + 
+  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd)) +
+  geom_point(data = filter(obs, r==1))
+```
+
+```{r}
+# unfished equilibrium
+m1_forecast <- model_3sp(0, 0.58, sigma = 0.05,  Tmax = 50)
+
+predictions <-   
+m1_forecast %>%
+    transmute(time = time, cormorant = x5, bass = x4, herring = x1, r=r) %>%
+    pivot_longer(any_of(c("cormorant", "bass", "herring")),
+                    names_to = "species",
+                    values_to = "biomass") %>% 
+    group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass))
+
+predictions %>%
+ggplot(aes(time, biomass, col= species)) + 
+  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd)) +
+  geom_point(data = filter(obs, r==1))
+```
+
+```{r}
+# Optimal policy under identical utility
+u_3sp <- function(x) utility(x, f = model_3sp)
+
+m3sp <- optim(c(0.5,0.1), u_3sp)
+m3sp$value
+clip(m3sp$par, 0, 1)
+
+```
+
+Alternate model, also mistakenly treats the three species of herring as a single species, but does a better job capturing the predator-prey relationships.
+This model also happens to produce better forecasts, but leads to worse decision:
+
+```{r}
+model_B <- function(u_prey = 0.5,
+                      u_pred = 0.5,
+                      sigma = 0.05,
+                      x1o = 0.44, # unfished equib
+                      x2o = 0.17,
+                      x3o = 0.186,
+                      x4o = 0.600,
+                      x5o = 0.280,
+                      Tmax = 50,
+                      R1 =  1.15, #1.0213,
+                      A11 = 1,
+                      A14 = 0.1, # 0.7252,
+                      A15 = 0.1, # 0.2445,
+                      A41 =  0.8, #0.7252,
+                      A51 = 0.3, # 0.2445,
+                    reps = 40) {
+
+    df <- tibble()
+    for(r in 1:reps){
+
+      x1 <- x4 <- x5 <- numeric(Tmax)
+      s1 <-  s4 <- numeric(Tmax)
+      x2 <- x3 <- numeric(Tmax)
+      xi1 <- rnorm(Tmax, 0, sigma)
+      xi4 <- rnorm(Tmax, 0, sigma)
+      xi5 <- rnorm(Tmax, 0, sigma)
+
+      x1[1] <- x1o + x2o + x3o
+      x4[1] <- x4o
+      x5[1] <- x5o
+
+      for (t in 1:(Tmax-1)) {
+          s1[t] <- x1[t] * (1 - u_prey)
+          s4[t] <- x4[t] * (1 - u_pred)
+          x1[t+1] = s1[t]*exp(R1 - A11*s1[t] - A14*s4[t] - A15 * x5[t] + xi1[t])
+          x4[t+1] = s4[t]*exp(A41*s1[t] - s4[t] + xi4[t])
+          x5[t+1] = x5[t]*exp(A51*s1[t] - x5[t] +  xi5[t])
+
+      }
+      df <- bind_rows(df, tibble(time = 1:Tmax, x1, x2, x3, x4, x5, r=r))
+    }
+    df
+}
+```
+
+```{r}
+# Optimal policy under identical utility
+u_B <- function(x) utility(x, f = model_B)
+u_B(c(0, .58))
+#optim(c(0.5,0.1), u_B, method = "L-BFGS-B", lower=0, upper=1)
+
+B <- optim(c(0.5,0.1), u_B)
+B$value
+model_B_policy <- clip(B$par, 0, 1)
+model_B_policy
+
+```
+
+Run the simulation using management policy of each
+
+```{r}
+
+Tmax <- 20
+x <- model_B_policy
+#x <- c(0.,0.58)  # Following the optimal policy under model_B
+
+obs <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
+predict_B <- model_B(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
+
+obs_stats <- obs %>%
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass))
+
+predict_stats_B <- predict_B %>% 
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass))
+
+predict_stats_B %>%
+  ggplot(aes(time, biomass, col= species)) + 
+  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd), lty=2) + 
+  geom_errorbar(data = obs_stats, aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd))
+
+```
+
+```{r}
+## True observations vs forecasts, 5 steps ahead, while following the Optimal / Model 1 policy, starting with default initial condition
+Tmax <- 5
+x <- c(0,0.58)
+
+obs <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
+predict <- model_3sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
+
+obs_stats <- obs %>%
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass))
+
+predict_stats <- predict %>% 
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass))
+
+## dotted are forecasted values, solid are observed values
+
+predict_stats %>%
+  ggplot(aes(time, biomass, col= species)) + 
+  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd), lty=2) + 
+  geom_errorbar(data = obs_stats, aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd))
+
+```
+
+```{r}
+score_me <- function(x = c(0,0.58), process=model_5sp, forecast=model3p, Tmax=5) {
+
+  obs <- process(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
+  predict <- forecast(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
+
+  #scoringRules::crps_sample(predict$x1, )
+
+}
+
+```

### 2022-02-11 manuscript/appendix.Rmd

--- 2021-04-02 manuscript/appendix.Rmd
@@
 
-Because the methods applied here: Stochastic Dynamic Programming, Iterative Forecasts assessed with Proper Scoring Rules, or Adaptive Management, are not themselves novel, the reader is referred to previous reviews and text books which can provide a thorough introduction to methods, in particular, @Marescot2013 and @Smith1981.
+The methods applied here -- stochastic dynamic programming, iterative forecasts assessed with proper scoring rules, and adaptive management -- are not themselves novel, thus the reader is referred to previous reviews and text books which can provide a thorough introduction to methods, in particular, @Marescot2013 and @Smith1981.
 However, implementation of these approaches generally depends on numerical methods, for which even precise mathematical formulas or pseudo-code descriptions can be inadequate to specify unambiguously or to properly facilitate replication [@Barnes2010].
@@
   rowwise() %>%
-  mutate(expected = transitions[[model]][state_index, , 1]  %*% states,
+  mutate(expected = as.numeric( transitions[[model]][state_index, , 1]  %*% states),
          var = transitions[[model]][state_index, , 1]  %*% states ^ 2 - expected ^ 2,
@@
   mutate(prob =  transitions[[model]][state_index, next_state, action_index],
-         expected = transitions[[model]][state_index, , action_index]  %*% states,
+         expected = as.numeric( transitions[[model]][state_index, , action_index]  %*% states ),
          var = transitions[[model]][state_index, , action_index]  %*% states ^ 2 - expected ^ 2,
@@
 # Gneiting & Raferty (2007), eq27
-scoring_fn <- function(x, mu, sigma){ -(mu - x )^2 / sigma^2  - log(sigma)}
+scoring_fn <- function(x, mu, sigma){ as.numeric( -(mu - x )^2 / sigma^2  - log(sigma) )}
 

### 2022-02-15 manuscript/round3-manuscript.Rmd
:pen:

--- 2021-09-17 manuscript/manuscript.Rmd
@@
 @Gneiting2007's proof of proper scoring has since become a critical tool to avoid over-fitting when choosing models to make decisions, but as I illustrate, will not prevent the forecast trap.
+
+First, I will introduce a motivating example in which we will consider two reasonable process-based models, A and B. 
+Model B will produce very accurate forecasts, but lead to much worse outcomes than Model A.
+These accurate forecasts in Model B are not the result of chance or of over-fitting the data.
+
+# A motivating example
+
+To better understand how a model can produce a more accurate forecast and yet still lead to a worse decision, it may be helpful to start with a concrete example.
+
+I will use the term "model" to refer to any set of equations or code that can be used to produce a forecast.
+This term thus includes not only traditional process-based models, but could also includes statistical forecasting methods (e.g.
+ARIMA, @arima), non-parametric approaches such as empirical dynamical modeling [@Ye2015], or machine learning.
+Most such models must first be calibrated to historical data before they can produce a forecast, e.g. by parameter fitting, expert knowledge, or some other means. 
+Because different choices for those parameters create different forecasts, I will refer to those different parameterizations as different models.
+It is of course possible for a decision-maker to consider forecasts coming from multiple structurally different models simultaneously, and potentially assigning different weights to each model.
+As more data becomes available, it is possible to update model parameters or the weights assigned to models.
+I will examine such approaches for model ensembles and model updating further on.
+
+In these examples, we will focus on situations in which our 'data' comes from a model simulation rather than empirical sources.
+This enables us to conduct experiments that would be often impossible or unethical to perform in the real world: for instance, does a given fishery experience better long-term outcomes when managed according to forecasts derived from model 1 or from model 2?
+The ability to perform many replicate simulations will also allow us to avoid results that could be due to chance alone.
+
+We will base this data-generating simulation on the five-species model of @Brias2021.
+In this model, three competing species of river herring are preyed upon by by striped bass and double-crested cormorants.
+Managers face the multiple competing objectives of promoting economically valuable fisheries for striped bass and herring, while the double-crested cormorants are a target of conservation.
+Following the formulation of @Brias2021, let us assume managers seek to maximize a weighted sum of composed of the total number of cormorants and the value of harvest from both herring fishery, and the striped bass fishery.
+We will assume the manager has decided to balance conservation and economic objectives, with 50% weight allocated to cormorant population and 25% to each fishery.
+We will assume that the manager must select a fixed fishing effort $u_{\textrm prey}$ and $u_{\textrm predator}$ that determines the annual mortality of herring and bass respectively. 
+In practice, managers may have flexibility to adjust fishing effort annually, increasing the complexity of the decision involved.
+We will explore more complex decision-processes later, but note that in many ecological settings, choices available to managers are highly constrained and frequent adjustment of policies is costly or impossible [@ @ @].
+
+Reality is always more complex than even our richest models [@Schindler2015].
+In the real world, we can never have the 'true model' among our candidates -- after all, models seek to represent only the most salient features for their purpose [@Getz2018].
+To reflect this gap between model and real-world complexity, we assume our manager does not know the underlying process (Eq 1), but must choose between two alternative three-dimensional models.
+Both models will treat the herring as if it were a single species, rather than three distinct species.
+This lumping of data is typical of one of the many assumptions any model must make, and can be justified in terms of ecological similarities and/or data limitations. 
+A second important aspect of real-world systems and realistic models is that not all states can be measured perfectly. 
+To reflect this, we will assume the manager has only estimates of the bass and cormorant populations.
+
+Beyond this, each of our alternative models makes somewhat different assumptions about the biology, as seen in comparing the dynamical equations for Model A to Model B.  
+Model A does not explicitly capture reflect the role of predation on the herring population, and oversimplifies the dynamics of the cormorant population as tracking at a fixed fraction of the herring population.
+These assumptions are not strictly true, but are justifiable in certain limits -- modeling is the science of approximation.
+In contrast, under Model B, the trophic structure more closely reflects the true model, with the herring population growth rate decreasing when either cormorant or bass populations increase.
+
+Model parameters are selected for each model consistent with limited 'historical' data from stochastic simulation.
+Given either forecast model, it straight forward to determine the harvest policy  $\lbrace u_{\textrm prey}, u_{\textrm predator} \rbrace$ which maximizes the stated management objective. 
+We perform a standard Nelder-Meade optimization routine on each forecast to determine the optimal policy.
+We also note the forecasts each model predicts it will see using policy $T_\{max}$ years into the future.
+
+We can simulate what the actual management outcomes and observed species dynamics are under the policy derived from each forecast. 
+We compare how well each model did in predicting the resulting population dynamics across each species (considering the herring population as the sum across all three herring species), as well as the realized utility defined by the management objective.
+Because we are using a simulation, we can also go one step further and compare the policy and mean net utility to that of the optimal strategy for the true model.
 

### 2022-02-15 notebook/alternate-models.Rmd
:pen:

@@
 ---
+
+The model which gives the best forecast does not always lead to the best decision.
 
@@
 
+```{r}
+# Optimal policy under identical utility
+u_3sp <- function(x) utility(x, f = model_3sp)
+
+m3sp <- optim(c(0.5,0.1), u_3sp)
+m3sp$value
+clip(m3sp$par, 0, 1)
+
+```
+
+Alternate model, also mistakenly treats the three species of herring as a single species, but does a better job capturing the predator-prey relationships.
+This model also happens to produce better forecasts, but leads to worse decision:
+
+```{r}
+model_B <- function(u_prey = 0.5,
+                      u_pred = 0.5,
+                      sigma = 0.05,
+                      x1o = 0.44, # unfished equib
+                      x2o = 0.17,
+                      x3o = 0.186,
+                      x4o = 0.600,
+                      x5o = 0.280,
+                      Tmax = 50,
+                      R1 =  3, #1.15, #1.0213,
+                      A11 = 8,
+                      A14 = 0.9, # 0.7252,
+                      A15 = 0.9, # 0.2445,
+                      A41 =  0.1, #0.7252,
+                      A51 = 0.3, # 0.2445,
+                    reps = 40) {
+
+    df <- tibble()
+    for(r in 1:reps){
+
+      x1 <- x4 <- x5 <- numeric(Tmax)
+      s1 <-  s4 <- numeric(Tmax)
+      x2 <- x3 <- numeric(Tmax)
+      xi1 <- rnorm(Tmax, 0, sigma)
+      xi4 <- rnorm(Tmax, 0, sigma)
+      xi5 <- rnorm(Tmax, 0, sigma)
+
+      x1[1] <- x1o + x2o + x3o
+      x4[1] <- x4o
+      x5[1] <- x5o
+
+      for (t in 1:(Tmax-1)) {
+          s1[t] <- x1[t] * (1 - u_prey)
+          s4[t] <- x4[t] * (1 - u_pred)
+          x1[t+1] = s1[t]*exp(R1 - A11*s1[t] - A14*s4[t] - A15 * x5[t] + xi1[t])
+          x4[t+1] = s4[t]*exp(A41*s1[t] - s4[t] + xi4[t])
+          x5[t+1] = x5[t]*exp(A51*s1[t] - x5[t] +  xi5[t])
+
+      }
+      df <- bind_rows(df, tibble(time = 1:Tmax, x1, x2, x3, x4, x5, r=r))
+    }
+    df
+}
+```
+
+```{r}
+# Optimal policy under identical utility
+u_B <- function(x) utility(x, f = model_B)
+u_B(c(0, .58))
+#optim(c(0.5,0.1), u_B, method = "L-BFGS-B", lower=0, upper=1)
+
+B <- optim(c(.2,0.01), u_B)
+model_B_policy <- clip(B$par, 0, 1)
+model_B_policy
+
+```
+
+Run the simulation using management policy of each
+
+```{r}
+
+Tmax <- 20
+x <- model_B_policy
+#x <- c(0.,0.58)  # Following the optimal policy under model_B
+
+sim <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax, sigma= 0.05)
+obs <- sim %>% format_sim()
+predict_B <- model_B(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
+
+reward(df = sim, x = x) / 40
+```
+
+```{r}
+obs_stats <- obs %>%
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass), type = "observed")
+
+predict_stats_B <- predict_B %>% 
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass), type = "predicted_B")
+
+bind_rows(obs_stats, predict_stats_B) %>% 
+  filter(species != "herring") %>%
+  ggplot(aes(time, biomass, col= species)) + 
+  geom_point(aes(shape=type)) +
+  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd, lty=type))
+
+```
+
+```{r}
+## True observations vs forecasts, 5 steps ahead, while following the Optimal / Model 1 policy, starting with default initial condition
+Tmax <- 10
+x <- c(0,0.58)
+
+sim <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax, sigma=0.05)
+obs <- sim %>% format_sim()
+predict <- model_3sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
+
+reward(df = sim, x = x) / 40
+```
+
+```{r}
+obs_stats <- obs %>%
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass), type = "observed")
+
+predict_stats <- predict %>% 
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass), type = "predicted_A")
+
+bind_rows(obs_stats, predict_stats) %>% 
+  filter(species != "herring") %>%
+  ggplot(aes(time, biomass, col= species)) + 
+  geom_point(aes(shape=type)) +
+  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd, lty=type))
+```
+
+```{r}
+score_me <- function(x = c(0,0.58), process=model_5sp, forecast=model3p, Tmax=5) {
+
+  obs <- process(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
+  predict <- forecast(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
+
+}
+```
+
+## unfished
+
+```{r}
+## True observations vs forecasts, 5 steps ahead, unfished, starting with default initial condition
+Tmax <- 20
+x <- c(0,0.)
+
+obs <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], sigma=0.05, Tmax = Tmax) %>% format_sim()
+predict <- model_3sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
+
+obs_stats <- obs %>%
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass))
+
+predict_stats <- predict %>% 
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass))
+
+## dotted are forecasted values, solid are observed values
+
+predict_stats %>%
+  ggplot(aes(time, biomass, col= species)) + 
+  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd), lty=2) + 
+  geom_errorbar(data = obs_stats, aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd))
+
+```
+
+```{r}
+
+Tmax <- 5
+x <- c(0,0)
+x <- model_B_policy
+#x <- c(0.,0.58)  # Following the optimal policy under model_B
+
+obs <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
+predict_B <- model_B(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
+
+obs_stats <- obs %>%
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass))
+
+predict_stats_B <- predict_B %>% 
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass))
+
+predict_stats_B %>%
+  ggplot(aes(time, biomass, col= species)) + 
+  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd), lty=2) + 
+  geom_errorbar(data = obs_stats, aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd))
+
+```
+
 Does this model produce good forecasts?
@@
 ```
-
-```{r}
-# Optimal policy under identical utility
-u_3sp <- function(x) utility(x, f = model_3sp)
-
-m3sp <- optim(c(0.5,0.1), u_3sp)
-m3sp$value
-clip(m3sp$par, 0, 1)
-
-```
-
-Alternate model, also mistakenly treats the three species of herring as a single species, but does a better job capturing the predator-prey relationships.
-This model also happens to produce better forecasts, but leads to worse decision:
-
-```{r}
-model_B <- function(u_prey = 0.5,
-                      u_pred = 0.5,
-                      sigma = 0.05,
-                      x1o = 0.44, # unfished equib
-                      x2o = 0.17,
-                      x3o = 0.186,
-                      x4o = 0.600,
-                      x5o = 0.280,
-                      Tmax = 50,
-                      R1 =  1.15, #1.0213,
-                      A11 = 1,
-                      A14 = 0.1, # 0.7252,
-                      A15 = 0.1, # 0.2445,
-                      A41 =  0.8, #0.7252,
-                      A51 = 0.3, # 0.2445,
-                    reps = 40) {
-
-    df <- tibble()
-    for(r in 1:reps){
-
-      x1 <- x4 <- x5 <- numeric(Tmax)
-      s1 <-  s4 <- numeric(Tmax)
-      x2 <- x3 <- numeric(Tmax)
-      xi1 <- rnorm(Tmax, 0, sigma)
-      xi4 <- rnorm(Tmax, 0, sigma)
-      xi5 <- rnorm(Tmax, 0, sigma)
-
-      x1[1] <- x1o + x2o + x3o
-      x4[1] <- x4o
-      x5[1] <- x5o
-
-      for (t in 1:(Tmax-1)) {
-          s1[t] <- x1[t] * (1 - u_prey)
-          s4[t] <- x4[t] * (1 - u_pred)
-          x1[t+1] = s1[t]*exp(R1 - A11*s1[t] - A14*s4[t] - A15 * x5[t] + xi1[t])
-          x4[t+1] = s4[t]*exp(A41*s1[t] - s4[t] + xi4[t])
-          x5[t+1] = x5[t]*exp(A51*s1[t] - x5[t] +  xi5[t])
-
-      }
-      df <- bind_rows(df, tibble(time = 1:Tmax, x1, x2, x3, x4, x5, r=r))
-    }
-    df
-}
-```
-
-```{r}
-# Optimal policy under identical utility
-u_B <- function(x) utility(x, f = model_B)
-u_B(c(0, .58))
-#optim(c(0.5,0.1), u_B, method = "L-BFGS-B", lower=0, upper=1)
-
-B <- optim(c(0.5,0.1), u_B)
-B$value
-model_B_policy <- clip(B$par, 0, 1)
-model_B_policy
-
-```
-
-Run the simulation using management policy of each
-
-```{r}
-
-Tmax <- 20
-x <- model_B_policy
-#x <- c(0.,0.58)  # Following the optimal policy under model_B
-
-obs <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
-predict_B <- model_B(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
-
-obs_stats <- obs %>%
-  group_by(time, species) %>% 
-  summarize(sd = sd(biomass), biomass = mean(biomass))
-
-predict_stats_B <- predict_B %>% 
-  group_by(time, species) %>% 
-  summarize(sd = sd(biomass), biomass = mean(biomass))
-
-predict_stats_B %>%
-  ggplot(aes(time, biomass, col= species)) + 
-  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd), lty=2) + 
-  geom_errorbar(data = obs_stats, aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd))
-
-```
-
-```{r}
-## True observations vs forecasts, 5 steps ahead, while following the Optimal / Model 1 policy, starting with default initial condition
-Tmax <- 5
-x <- c(0,0.58)
-
-obs <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
-predict <- model_3sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
-
-obs_stats <- obs %>%
-  group_by(time, species) %>% 
-  summarize(sd = sd(biomass), biomass = mean(biomass))
-
-predict_stats <- predict %>% 
-  group_by(time, species) %>% 
-  summarize(sd = sd(biomass), biomass = mean(biomass))
-
-## dotted are forecasted values, solid are observed values
-
-predict_stats %>%
-  ggplot(aes(time, biomass, col= species)) + 
-  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd), lty=2) + 
-  geom_errorbar(data = obs_stats, aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd))
-
-```
-
-```{r}
-score_me <- function(x = c(0,0.58), process=model_5sp, forecast=model3p, Tmax=5) {
-
-  obs <- process(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
-  predict <- forecast(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
-
-  #scoringRules::crps_sample(predict$x1, )
-
-}
-
-```

### 2022-02-23 manuscript/appendix.Rmd
:pen:

@@
 
-```{r setup, message=FALSE}
-library(tidyverse)
-library(MDPtoolbox)
-library(expm)
-# remotes::install_github("boettiger-lab/mdplearning")
-library(mdplearning)
-```
-
 # Optimal quotas as a Markov Decision Process (MDP) problem
@@
 
-This problem is already well studied, and it is worth noting that even under the simple utility function which assigns no value to fish that are not harvested, the optimal strategy still seeks to sustain the fish population indefinitely; as @Clark1973 shows for the deterministic function $f$ and @Reed1979 extended to the stochastic case.
+This problem is already well studied in 1 dimension (i.e. a single species without age structure):
+@Clark1973 solves the optimization when $f$ is a deterministic function $f$ and @Reed1979 extended that result to the stochastic case.  
+It is worth noting that even under the simple utility function which assigns no value to fish that are not harvested, the optimal strategy still seeks to sustain the fish population indefinitely.
 As @Reed1979 proves, the optimal strategy can be characterized by a policy of "constant escapement," in which the manager adjusts the harvest effort each year in an effort to ensure the same number or biomass of fish are left in the sea each year.
 The optimization problem can then be solved by dynamic programming, using the recursive Bellman equation. 
-This technique, known as stochastic dynamic programming (SDP) is widely used in conservation decision problems [e.g.
-@Mangel1985, @Marescot2013].
+This technique, known as stochastic dynamic programming (SDP) is widely used in conservation decision problems [e.g. @Mangel1985, @Marescot2013].
 
@@
 These issues are discussed in detail in @pomdp-intro.
-For our purposes, it suffices to note that constant escapement policy is relevant both from a theoretical standpoint as the optimal solution for the problem under consideration, and an applied standpoint as a policy that guides management of many salmon fisheries.
-The optimal control approach using SDP as illustrated here is used in a wide and growing number of conservation problems.
+For our purposes, it suffices to note that constant escapement policy is relevant both from a theoretical standpoint as the optimal solution for the problem under consideration, and an applied standpoint as a policy that guides management of many salmon fisheries.The optimal control approach using SDP as illustrated here is used in a wide and growing number of conservation problems.
 
@@
 
-```{r model_definitions}
-states <- seq(0,24, length.out = 240)
-actions <- states
-obs <- states
-```
+The utility (reward) of an action is set to the (realized) harvest (e.g. a fixed price per unit fish, with no cost applied to harvest effort), ensuring no harvest can exceed the available stock.
 
-The utility (reward) of an action is set to the (realized) harvest (e.g.
-a fixed price per unit fish, with no cost applied to harvest effort).  Future
-utility is discounted by fixed factor of $0.99$. Classic maximum sustainable 
+$$U(x_t, h_t) = p \min(x_t, h_t) $$
+
+Future utility is discounted by fixed factor of $\delta = 0.99$. Classic maximum sustainable 
 yield models [@Schaefer1954] ignore discounting, while modern economic optimization
@@
 discounting rate.
-
-```{r}
-reward_fn <- function(x,h) pmin(x,h)
-discount <- 0.99
-```
 
@@
 
-The manager chooses between two different logistic growth models, each of which
-can be thought of as an approximation to the underlying "true" population model:
+Let us assume that our set of candidate models are simply the possible parameterizations of a stochastic version of the classic Gordon-Schaefer model [@Schaefer1954; @Gordon1954]:
 
 \begin{equation}
-f_i(Y) = Y + Y r_i \left( 1 - \frac{Y}{K_i} \right) \xi_i(t)
+f_i(Y_{t+1}) = Y_t + r_i Y_t \left( 1 - \frac{Y_t}{K_i} \right) * \xi_t(\sigma)
 \end{equation}
@@
 
-Model 1 has $r_1 = 2$, $K = 16$, and $\sigma_1 = 0.05$.
-Model 2 has $r_2 = 0.5, K_2 = 10$, and $\sigma_2 = 0.075$.
+Before this or any other model can generate a forecast, we must first come up with some parameter estimates.
+Because the model includes (log-normal) stochastic growth, no amount of data will make any parameter combination impossible, though certain parameter values are more likely than others.
+Remember too that parameter estimates may be derived in other ways than than model fitting, especially when parameters are amenable to biological interpretation. 
+We will consider the whole range of possible parameters in a moment, but for simplicity, let us begin by focusing in around two of the most interesting regions of that are already included within that larger parameter space of all possible values for $r$, $K$, and $\sigma$.
+Let us take "Model 1" as being given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, "Model 2" by $r_2 = 0.5$, $K_2 = 10$, $\sigma_2 = 0.075$. 
+We can imagine our comparison of these two models as a microcosm of the larger comparison between all possible paremeterizations.
 
-Meanwhile, the true population growth rate is simulated using a function with
-non-linear per-capita growth:
+Ecologists will rightly scoff at the simplicity of these models -- the real world is much more complicated.
+So it is important to bear in mind that these are not models that seek to approximate the stock dynamics of real world fisheries, only to approximate whatever "true model" we are using to drive the simulation.
+In recognition of the fact that real world is always more complex than even our best ecological models, we will assume a "true model" for the simulations that is not in the Gordon-Schaefer class (i.e.
+our candidate models will never contain the true model), but is not so rich that a Gordon-Schaefer curve would seem a hopelessly poor approximation.
+
+For illustrative purposes, we will thus assume the "true" process to be given by Model 3, which is unknown to the decision-maker, but similar enough to at least one of the candidate models might be considered a reasonable approximation:
 
 \begin{equation}
-f_i(Y) = Y + Y^4 r_i \left( 1 - \frac{Y}{K_i} \right)
+f_3(Y) = Y + r_3 Y^4 \left( 1 - \frac{Y}{K_3} \right)
 \end{equation}
@@
 with $r_3 = 0.002$, $K_3 = 10$ and $\sigma_3 = 0.05$.
+
+Certainly, the challenge of choosing which model to base a decision policy on in the real world is much harder than this binary choice between two models, and yet it is sufficient to illustrate the trap.
+We will see later why making the models much more complex does not guarantee that the task becomes easier or that the trap may be ruled out.
+
+# Methods for Managing Under Model Uncertainty
+
+I will use this example to illustrate two alternative approaches for iterative learning over model uncertainty: iterative forecasting and adaptive management.
+The central difference in the approaches is that iterative forecasting is premised on the ability to score the predictions of alternative models.
+Iterative forecasting is silent on the issue of what to do with those scores, this is left up to the decision-maker.
+Adaptive management approaches, by contrast, explicitly seek to integrate probabilities over all candidate models to reach a decision.
+I consider each in turn.
+
+## Statistical approaches: Forecasting under "Proper" Scoring Rules
+
+Like many decision problems, the task of setting a sustainable harvest quota appears to hinge on having an accurate forecast: if we can predict to what size the fish stock will increase next year, $X_t+1$, and we know the current stock, $X_t$, then we can sustainably harvest $X_{t+1} - X_t$ without decreasing the biomass over the long term.
+Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of overfitting by comparing model predictions to later observations that were not used to estimate the model [@Gneiting2014].
+
+I illustrate the process of model selection by strictly proper scoring rules using two scenarios.
+In Scenario A (passive observation) the fish stock is unharvested and allowed to recover towards carrying capacity (as simulated under our "true" model, Model 3) while comparing the observed stock size in each subsequent time step to the distribution predicted under model 1 and model 2 respectively [Fig 1].
+The mean, $\mu_t$ and variance, $\sigma_t$ of the forecast are compared against the true observation $x_t$ using a proper scoring rule given by @Gneiting2007,
+
+\begin{equation}
+S(x_t|\mu_t,\sigma_t) = -(\mu_t - x_t )^2 / \sigma_t^2 - \log(\sigma_t) \label{proper}
+\end{equation}
+
+for each prediction over 100 replicate simulations of 100 time steps each [Fig 1].
+
+# Analysis
 
@@
 
-```{r}
+```{r setup, message=FALSE}
+library(tidyverse)
+library(MDPtoolbox)
+library(expm)
+# remotes::install_github("boettiger-lab/mdplearning")
+library(mdplearning)
+```
+
+```{r model_definitions}
+states <- seq(0,24, length.out = 240)
+actions <- states
+obs <- states
+reward_fn <- function(x,h) pmin(x,h)
+discount <- 0.99
+
 # K is at twice max of f3; 8 * K_3 / 5
@@
 ```{r}
-transitions <- lapply(seq_along(models), function(i) 
+P <- lapply(seq_along(models), function(i) 
   transition_matrices(models[[i]], states, actions, model_sigmas[[i]]))
-names(transitions) <- c("1", "2", "3")
+names(P) <- c("1", "2", "3")
 
@@
 policies <- 
-  map_dfr(transitions, 
+  map_dfr(P, 
           function(P){
@@
 simulate_policy <- function(i, policy){
-  mdp_planning(transitions[[true_model]], reward, discount,
+  mdp_planning(P[[true_model]], reward, discount,
            policy = policy, x0 = x0, Tmax = Tmax) %>%
@@
 sims <- 
-  map_dfr(names(transitions), 
+  map_dfr(names(P), 
           function(m){
@@
   rowwise() %>%
-  mutate(expected = as.numeric( transitions[[model]][state_index, , 1]  %*% states),
-         var = transitions[[model]][state_index, , 1]  %*% states ^ 2 - expected ^ 2,
-         low = states[max(which(cumsum(transitions[[model]][state_index,,1]) < 0.025)) ],
-         high = states[min(which(cumsum(transitions[[model]][state_index,,1]) > 0.975)) ],
+  mutate(expected = as.numeric( P[[model]][state_index, , 1]  %*% states),
+         var = P[[model]][state_index, , 1]  %*% states ^ 2 - expected ^ 2,
+         low = states[max(which(cumsum(P[[model]][state_index,,1]) < 0.025)) ],
+         high = states[min(which(cumsum(P[[model]][state_index,,1]) > 0.975)) ],
          true = states[next_state ]) 
@@
   rowwise() %>%
-  mutate(prob =  transitions[[model]][state_index, next_state, action_index],
-         expected = as.numeric( transitions[[model]][state_index, , action_index]  %*% states ),
-         var = transitions[[model]][state_index, , action_index]  %*% states ^ 2 - expected ^ 2,
-         low = states[max(which(cumsum(transitions[[model]][state_index,,action_index]) < 0.025)) ],
-         high = states[min(which(cumsum(transitions[[model]][state_index,,action_index]) > 0.975)) ],
+  mutate(prob =  P[[model]][state_index, next_state, action_index],
+         expected = as.numeric( P[[model]][state_index, , action_index]  %*% states ),
+         var = P[[model]][state_index, , action_index]  %*% states ^ 2 - expected ^ 2,
+         low = states[max(which(cumsum(P[[model]][state_index,,action_index]) < 0.025)) ],
+         high = states[min(which(cumsum(P[[model]][state_index,,action_index]) > 0.975)) ],
          true = states[next_state]) %>%
@@
 adaptive_management <- memoise::memoise(mdp_learning, cache = memoise::cache_filesystem("cache/"))
-am1 <- adaptive_management(transitions[1:2], reward, discount, 
+am1 <- adaptive_management(P[1:2], reward, discount, 
                            model_prior = c(0.99, 0.01), x0 = x0, 
-                           Tmax = 50, true_transition = transitions[[3]], 
+                           Tmax = 50, true_transition = P[[3]], 
                             epsilon = 0.001, max_iter = 2000)
@@
 policy_planning <- memoise::memoise(mdp_compute_policy, cache = memoise::cache_filesystem("cache/"))
-policy <- policy_planning(transitions[1:2],  
+policy <- policy_planning(P[1:2],  
                           reward = reward, 
@@
 # Simulate under true model using fixed planning policy
-non_am <- mdp_planning(transitions[[3]],  
+non_am <- mdp_planning(P[[3]],  
                        reward = reward, 
@@
 
-Under adaptive management, the manager realizes only `r am_economics_percent`%
-of the economic value that would be achieved under Model 1 alone, and only 
-`r am_ecology_percent`% of the spawning stock biomass that would have been 
-achieved under Model 1 alone.
+Under adaptive management, the manager realizes only `r am_economics_percent`% of the economic value that would be achieved under Model 1 alone, and only `r am_ecology_percent`% of the spawning stock biomass that would have been achieved under Model 1 alone.
 
@@
 write_csv(model_curves, "../data/model_curves.csv")
-
 ```

### 2022-02-23 manuscript/round3-manuscript.Rmd
:pen:

--- 2021-09-17 manuscript/manuscript.Rmd
@@
   Encouraged by decision makers' appetite for future information on topics ranging from elections to pandemics, and enabled by the explosion of data and computational methods, model based forecasts have garnered increasing influence on a breadth of decisions in modern society.
-Using a classic example from fisheries management, I demonstrate that selecting the model or models that produce the most accurate and precise forecast (measured by statistical scores) can lead to decidedly worse outcomes (measured by real-world objectives).
+Using several classic examples from fisheries management, I demonstrate that selecting the model or models that produce the most accurate and precise forecast (measured by statistical scores) can lead to decidedly worse outcomes (measured by real-world objectives).
 This can create a forecast trap, in which the outcomes such as fish biomass or economic yield decline while the manager becomes increasingly convinced that these actions are consistent with the best models and data available.
@@
 
-In this paper, I demonstrate that the model that makes the better prediction (rigorously defined as a strictly proper score, @Gneiting2007) is not necessarily the model that makes the better policy (rigorously defined in terms of utility, e.g.
+In this paper, I demonstrate that the model that makes the better prediction (defined as a strictly proper score, @Gneiting2007) is not necessarily the model that makes the better policy (defined in terms of utility, e.g.
 expected net present value, @Clark1990).
@@
 When VOI is negative, the decision-maker may become trapped into accepting mediocre outcomes derived from a model that makes accurate forecasts, even when a less accurate model that would generate better outcomes is available.
-In our example, the manager will decide the fishery in question simply has low productivity, because such a model yields better predictions, rather than realizing that the low productivity observed is in fact a consequence of the over-harvesting.
-This disconcerting situation can arise whenever two conditions are met: (1) the optimal management policy is not unique to the generative process, and (2) the generative process is not included in candidate set of models.
+I present two examples of this "forecast trap" and examine how it arises.
+This disconcerting situation is possible whenever two conditions are met: (1) the optimal management policy is not unique to the generative process, and (2) the generative process is not included in candidate set of models.
 These conditions do not guarantee the trap will occur, only the circumstances in which it cannot be ruled out entirely.
@@
 
-# From Predictive Models to Decision Policies
+First, I will introduce a motivating example in which we will consider two reasonable process-based models, A and B. 
+Model A will produce very accurate forecasts, but lead to much worse outcomes than Model B.
+Though I will establish that these accurate forecasts in Model A are not the result of chance or of over-fitting the data, this example may raise more questions than answers. 
+To get a better understanding of when the forecast trap arises, I will turn to a simpler ecological model, to which we may apply more sophisticated decision tools of iterative forecasting and adaptive management.
+We will see that these approaches do not avoid the forecast trap either. 
+No collection of such examples can establish precisely how common the forecast trap may be in real-world applications.
+The examples do establish unequivocally that achieving incrementally ever-more-accurate forecasts does _not guarantee_ better decisions.
+I conclude by pointing to a range of established and emerging approaches to quantitative decision-making which are not based on forecasts.
+As sophisticated forecasting techniques become more common-place in conservation and ecology, the forecast trap is a reminder that we should not forget about these alternatives.
 
-How do we translate a model-based forecast into a decision?
+**A note on models and data**
+
+I will use the term "model" to refer to any set of equations or code that can be used to produce a forecast.
+This term thus includes not only process-based models, but could also statistical forecasting methods (e.g.
+ARIMA, @arima), non-parametric approaches such as empirical dynamical modeling [@Ye2015], or machine learning.
+Most such models must first be calibrated to historical data before they can produce a forecast, e.g. by parameter fitting, expert knowledge, or some other means. 
+Different choices for those parameters create different forecasts, I will refer to those different parameterizations as different models.
+It is of course possible for a decision-maker to consider forecasts coming from multiple structurally different models simultaneously, and potentially assigning different weights to each model.
+As more data becomes available, it is possible to update model parameters, or equivalently, update the weights assigned across models.
+I will examine such approaches for model ensembles and model updating further on.
+
+In these examples, we will focus on situations in which our 'data' comes from a model simulation rather than empirical sources.
+Simulations are simplifications of the real world -- just because a method works in a simulation is no guarantee that it works in reality.
+Conversely, if decision methods are not reliable even when applied to simulated cases, we should be even more cautious in how we use them.
+Simulations also allow us to consider many replicates and
+conduct experiments that would be often impossible or unethical to perform in the real world: for instance, does a given fishery experience better long-term outcomes when managed according to forecasts derived from model 1 or from model 2?
+
+I shall focus on examples involving fisheries management to illustrate principles shared in many ecological systems.
+Fisheries are a significant economic and conservation concern worldwide and their management remains an important debate [e.g. @Worm2006; @Worm2009; @Costello2016].
+Moreover, their management has been a proving grounds for theoretical and practical decision-making issues which are widely applicable in other areas of ecology and conservation [@Ludwig1993; @Lande1994], and one that has long wrestled with issues of uncertainty in the context of management decisions [e.g.
+@Clark1973; @Reed1979; @Walters1981; @Ludwig1982].
+
+# A motivating example
+
+To better understand how a model can produce a more accurate forecast and yet still lead to a worse decision, it may be helpful to start with a concrete example.
+Figure \ref{fig1} shows both the forecasts and realized management outcomes of two alternative three-species models, "A" and "B" (see Appendix A) in predicting the population dynamics of striped bass (an economically important fishery) and double-crested cormorants (a target species for conservation) which both feed on a population of river herring (whose abundance we assume is not measured).
+Model A accurately forecasts the abundance of both bass and cormorants well into the future, but the optimal management strategy derived from its forecasts leads to steadily declining species abundances and overall disappointing net utility.
+Model B produces substantially less accurate forecasts, but Model B's forecasts also lead to better outcomes, seen both in the population abundances and overall net utility.
+
+This management problem is motivated by a real world example of a herring fishery as described in @Brias2021, 
+in which our manager seeks to balance multiple objectives of sustaining the cormorant population while maximizing the economic value from harvesting both herring and striped bass.
+In the scenario depicted in Figure \ref{fig1}, I have used a the richer five-species model introduced @Brias2021 to drive the underlying dynamics, which includes three competing species of herring that are preyed upon by both the bass and the cormorants.
+Here, the manager seeks to maximize utility given by the weighted sum of the individual objectives [@Brias2021], placing 50% of the weight on the conservation objective and splits the remaining weight evenly over the harvests for predator (bass) and prey (herring) species.
+I have assumed a partially observable system - in this case, the manager measures only the abundance of bass and cormorant species, and not of the three herring species.
+In this scenario, I have further assumed the manager must choose a fixed fishing effort for herring and for the bass harvest,
+I will consider more dynamic decision-processes later, but it is worth noting that in many real world conservation settings policy choices are highly constrained and frequent adjustment of those policies may be costly or impossible.
+Likewise, the assumptions of partially observable system and imperfect models are characteristic of ecological decision-making.
+Equations and code for all models in this example are presented in Appendix A.
+
+Both models A & B can be seen as alternative attempts to approximate the "true" model (generative process), which in real systems is always unknown and more complex than any model thereof. 
+Model A assumes a three-species Ricker model which closely matches the trophic structure of the "true" five-species model, lumping the three competing herring species into a single variable. 
+Because herring abundance is not observed directly, the model parameters related to herring growth are less accurate than other parameters.
+Model B also lumps herring species together into a single variable, but fails to reflect the trophic relationship between bass and herring.
+Model B also oversimplifies the relationship between cormorant and herring population.
+This does not make Model B an unreasonable model out-of-hand -- all models contain such simplifications (e.g.
+our "true" model does not model the trophic relationship between the herring and its food sources or environmental conditions explicitly either).
+Both models are consistent with the limited historical data available to them.
+
+# Iterative updating: _Sequential Decision Processes_
+
+While this example demonstrates that the model which provides the better forecast does not necessarily lead to a better decision, it may raise more questions than it answers.
+Why does this happen? 
+Is this an isolated example or not?
+Can this forecast trap be resolved by more sophisticated approaches to model selection and decision-making?
+I now consider scenarios involving iterative forecasts and adaptive management: in which the manager monitors outcomes, compares forecasts to observations and updates model estimates.
+Such _sequential decision processes_ are not merely iterative versions of single-decision problems, but are much more challenging.
+In the opening example, the manager had to choose the harvest policy for each fish species at the start of the scenario and stick with it.
+The ability to select new actions in response to new observations turns that decision into a game of chess: each turn, the manager must consider not only their next move but all possible series of moves.
+
+How do we translate a model-based forecast into a decision policy?
 It is impossible to discuss outcomes associated with a forecast without first agreeing on this process.
@@
 Two features of this approach are worth emphasizing.
-First, the resulting decision is derived directly from the forecast model and the desired objective.
+As before, the resulting decision is derived directly from the forecast model and the desired objective.
 The SDP algorithm is a reasonable description of the approach any ideal manager would use -- considering all possible outcomes from all possible sequences of actions and selecting the best sequence.
@@
 To ensure that the forecast trap is not a result of such assumptions about possible policies, we will consider a problem simple enough to solve directly with SDP.
-This leads to the second point: the resulting decision policy is optimal, so long as the forecast model is correct.
+The resulting decision policy is optimal, so long as the forecast model is correct.
 In this way, the SDP merely stands in for a mathematically precise way in which forecasts are turned into decisions.
@@
 
-# Ecological Models
+Let us consider the management of single species in which we seek to maximize the long-term net harvest.
+In this scenario, the manager estimates the population size each year and must set the total allowable catch (TAC) for that season.
+The underlying dynamics are unknown, but the manager is presented with any of a variety of forecast models which can predict the future stock sizes given the current population size and proposed TAC. 
+Our manager does not know which of these models is the most accurate a priori. 
+Instead, the manager will be able to compare the population size predicted by each forecast (under the chosen TAC) to the measurement of the population size in the following year before coming up with the next year's catch limit.
 
-I illustrate this problem using an example from fisheries management.
-Fisheries are a significant economic and conservation concern worldwide and their management remains an important debate [e.g. @Worm2006; @Worm2009; @Costello2016].
-Moreover, their management has been a proving grounds for theoretical and practical decision-making issues which are widely applicable in other areas of ecology and conservation [@Ludwig1993; @Lande1994], and one that has long wrestled with issues of uncertainty in the context of management decisions [e.g.
-@Clark1973; @Reed1979; @Walters1981; @Ludwig1982].
+Faced with a collection of models, a manager can seek either to identify the best model to use, or to consider an integrative assessment which uses the whole ensemble of models to represent the manager's uncertainty about the underlying process.
+I will consider both approaches in turn. 
+Figure 2 compares forecasts generated by two of the candidate models to observations drawn from simulations of the underlying process.
+As before, these are true forecasts: the model forecasts are generated first, they have not been fit to these observations.
+In the un-fished scenario (top panels), both models try to predicting the same un-fished equilibrium dynamics.
+In the second scenario (lower panel), the manager uses the optimal SDP policy derived from each forecast to determine the TAC for the following year, and compares the observed stock size to that which the model predicted given that fishing quota.
+In both cases, model 2 provides far more accurate forecasts, as seen in the error bars and confirmed by the distribution of proper scores [@Gneiting2007].
+Model and simulation details are provided in Appendix B.
 
-While methods such as iterative forecasting [@Dietze2018] and adaptive management [@Walters1978] can be _applied_ to real-world using empirical data, we can only _evaluate_ their potential in hypothetical examples when the true model is known, e.g.
-through numerical simulation.
-That approach allows us to compare both predictions and outcomes across implementations in independent identical replicate worlds.
-As noted above, we will assume our underlying model simple enough to solve by SDP, ensuring any poor outcomes from a given forecast are not merely an artifact of an imperfect decision process.
-Simple models also have the virtue in being accessible to closed form analysis, which, as we shall see, can give greater insight into when and why this forecast trap arises.
-That insight will in turn will allow us to examine if the same problem is likely to arise under more complex models.
-
-<!--
-While modern fisheries management frequently relies on complex models which may contain over 100 parameters to reflect the specific age or stage structure of a specific fish stock [@ramlegacy2018; @ram], fisheries management and research has always appreciated the central insights offered by simpler one-dimensional models which continue to underpin both the theory and practice.
-For example, the strategies of constant mortality [@Schaefer1954] and constant escapement [@Clark1973; @Reed1979] are built on the same class of models considered here, which also continue to underpin global analyses of stock-rebuilding [e.g.
-@Costello2016; @Memarzadeh2019] and the foundations of adaptive management [e.g.
-@Walters1978; @Smith1981; @Walters1981].
-While more complex models are often required for specific applications, such models both share features with and are guided by theory built on much more general, simpler models [@Levins1966],
-from the Maximum Sustainable Yield of @Schaefer1954 or the epidemiological reproductive number, $R_0$ of @Kermack1927. 
--->
-
-The sustainable harvest decision problem can be stated as follows: The fish stock is observed to be in state $X_t$ at time $t$, and is then subjected to some harvest $H_t$ before recruiting new fish, subject to stochastic environmental noise $\xi_t$, to bring the stock to $X_t+1$,
-
-\begin{equation}
-X_{t+1} = f(X_t - H_t, \xi_t) 
-\end{equation}
-
-A manager seeks each year to select the harvest quota $H_t$ which will maximize the sum
-of the utility derived from such a harvest and such a state, $U(X_t,H_t)$, over all time,
-subject to discount rate $\delta$ [@Clark1973]:
-
-\begin{equation}
-\sum_{t=0}^{t=\infty} U(X_t, H_t) \delta^t \label{utility}
-\end{equation}
-
-We will assume we have been given a fixed price of fish $p=1$ with no additional cost on additional harvest, $U(X_t, H_t) = p \min(H_t, X_t)$ modest discount $\delta = 0.99$.
-
-Let us assume that our set of candidate models are simply the possible parameterizations of a stochastic version of the classic Gordon-Schaefer model [@Schaefer1954; @Gordon1954]:
-
-\begin{equation}
-f_i(Y_{t+1}) = Y_t + r_i Y_t \left( 1 - \frac{Y_t}{K_i} \right) * \xi_t(\sigma)
-\end{equation}
-
-Where $Y$ is the population size after harvest, $Y_t = X_t-H_t$ and $\xi_t(\sigma)$ represents log-normal random noise with a mean of unity and log-standard-deviation $\sigma_i$.
-
-<!--
-This is where most fisheries biologists will probably stop reading.
-Modern stock assessment models frequently have over 100 parameters.
-However, the goal of exercise understand how a good forecast can lead to a bad decision even when we use these iterative forecasting methods correctly, not predict the safe harvest levels of Pacific Anchovy in 2022.
-If any method fails to select the better model in the simplest cases, it does not often do better in more complex ones, even though the problems are harder to detect.
--->
-
-Before this or any other model can generate a forecast, we must first come up with some parameter estimates.
-Because the model includes (log-normal) stochastic growth, no amount of data will make any parameter combination impossible, though certain parameter values are more likely than others.
-Remember too that parameter estimates may be derived in other ways than than model fitting, especially when parameters are amenable to biological interpretation. 
-We will consider the whole range of possible parameters in a moment, but for simplicity, let us begin by focusing in around two of the most interesting regions of that are already included within that larger parameter space of all possible values for $r$, $K$, and $\sigma$.
-Let us take "Model 1" as being given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, "Model 2" by $r_2 = 0.5$, $K_2 = 10$, $\sigma_2 = 0.075$. 
-We can imagine our comparison of these two models as a microcosm of the larger comparison between all possible paremeterizations.
-
-Ecologists will rightly scoff at the simplicity of these models -- the real world is much more complicated.
-So it is important to bear in mind that these are not models that seek to approximate the stock dynamics of real world fisheries, only to approximate whatever "true model" we are using to drive the simulation.
-In recognition of the fact that real world is always more complex than even our best ecological models, we will assume a "true model" for the simulations that is not in the Gordon-Schaefer class (i.e.
-our candidate models will never contain the true model), but is not so rich that a Gordon-Schaefer curve would seem a hopelessly poor approximation.
-
-For illustrative purposes, we will thus assume the "true" process to be given by Model 3, which is unknown to the decision-maker, but similar enough to at least one of the candidate models might be considered a reasonable approximation:
-
-\begin{equation}
-f_3(Y) = Y + r_3 Y^4 \left( 1 - \frac{Y}{K_3} \right)
-\end{equation}
-
-with $r_3 = 0.002$, $K_3 = 10$ and $\sigma_3 = 0.05$.
-
-Certainly, the challenge of choosing which model to base a decision policy on in the real world is much harder than this binary choice between two models, and yet it is sufficient to illustrate the trap.
-We will see later why making the models much more complex does not guarantee that the task becomes easier or that the trap may be ruled out.
-
-# Methods for Managing Under Model Uncertainty
-
-I will use this example to illustrate two alternative approaches for iterative learning over model uncertainty: iterative forecasting and adaptive management.
-The central difference in the approaches is that iterative forecasting is premised on the ability to score the predictions of alternative models.
-Iterative forecasting is silent on the issue of what to do with those scores, this is left up to the decision-maker.
-Adaptive management approaches, by contrast, explicitly seek to integrate probabilities over all candidate models to reach a decision.
-I consider each in turn.
-
-## Statistical approaches: Forecasting under "Proper" Scoring Rules
-
-Like many decision problems, the task of setting a sustainable harvest quota appears to hinge on having an accurate forecast: if we can predict to what size the fish stock will increase next year, $X_t+1$, and we know the current stock, $X_t$, then we can sustainably harvest $X_{t+1} - X_t$ without decreasing the biomass over the long term.
-Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of overfitting by comparing model predictions to later observations that were not used to estimate the model [@Gneiting2014].
-
-I illustrate the process of model selection by strictly proper scoring rules using two scenarios.
-In Scenario A (passive observation) the fish stock is unharvested and allowed to recover towards carrying capacity (as simulated under our "true" model, Model 3) while comparing the observed stock size in each subsequent time step to the distribution predicted under model 1 and model 2 respectively [Fig 1].
-The mean, $\mu_t$ and variance, $\sigma_t$ of the forecast are compared against the true observation $x_t$ using a proper scoring rule given by @Gneiting2007,
-
-\begin{equation}
-S(x_t|\mu_t,\sigma_t) = -(\mu_t - x_t )^2 / \sigma_t^2 - \log(\sigma_t) \label{proper}
-\end{equation}
-
-for each prediction over 100 replicate simulations of 100 time steps each [Fig 1].
-
-```{r figure1, fig.cap = "Forecast performance of each model.
+```{r figure2, fig.cap = "Forecast performance of each model.
 Panels A, B: Step ahead predictions of stock size under unfished (A) and fished (B) scenarios.
@@
 
-In Scenario B (actively harvest), I have first solved for the optimal management strategy using the forecast-matrices of both model 1 and model 2 [Fig 1b] using SDP [@Marescot2013; code in the Appendix].
-Replicate simulations of the stock are harvested at each time step using the optimal quota dictated by either model's forecasts, according to the SDP. 
-The resulting stock sizes in the subsequent timestep are scored against the forecast probabilities of each model using Eq \eqref{proper}.
-Model 2 unequivocally outperforms model 1 in both scenarios of passive observation and active harvest.
+Despite the clearly superior predictive accuracy of model 2 in both scenarios, the outcomes from management under model 2 are substantially worse.
+We can assess such outcomes in less abstract terms than forecasting skill, such as economic value them manager sought to optimize (in dollars) or the ecological value (unharvested biomass) [Fig 3].
 
-Despite the clearly superior predictive accuracy of model 2 in both scenarios, the outcomes from management under model 2 are substantially worse.
-We can assess such outcomes in less abstract terms than forecasting skill, such as economic value (in dollars) or the ecological value (unharvested biomass).
-In our simple formulation of the decision problem, the "utility" the manager seeks to maximize is simply the economic value (net present value: the discounted sum of all profits from future harvests, Eq \eqref{utility}) of harvested fish.
-This formulation ignores any utility provided by fish that are not harvested, beyond their contribution to future potential harvests.
-While it is possible to include such contributions directly in the utility function being optimized [e.g.
-@Halpern2013], even without doing so, model 1 maintains both a higher unharvested biomass and also leads to higher economic returns throughout [Fig 2].
-
-```{r figure2, message = FALSE, fig.cap = "Ecological and economic performance of each forecast.
+```{r figure3, message = FALSE, fig.cap = "Ecological and economic performance of each forecast.
 Harvest quotas derived from model 1 result in a significantly higher fish stock size than under Model 2 (panel A).
@@
 
-In both scenarios, the careful comparison through proper scoring rules has led us to select the worse-performing model.
-Crucially, a manager operating under this selection would have little indication that their model was flawed: both future stock sizes and expected harvest yields consistently match model predictions.
-Had we been able to include Model 3 in our forecast comparisons, it would equal or outperform the forecasting skill of both model 1 and model 2 (as guaranteed by the theorem of @Gneiting2007), while also matching or out-performing their economic utility (as guaranteed by the theorem of @Reed1979).
+A manager operating under model 2 would have little indication that the model was flawed: both future stock sizes and expected harvest yields consistently match model predictions.
+This manager would be stuck in the forecast trap, incorrectly mistaking the a degraded ecological state and reduced economic outcomes as the best that can be achieved because future observations continue to validate this model.
+Had we been able to include Model 3 in our forecast comparisons, it would equal or outperform the forecasting skill of both model 1 and model 2 (as guaranteed by the theorem of @Gneiting2007), while also matching the economic utility of model 1 (as guaranteed by the theorem of @Reed1979).
 In practice, we never have access to the generating model, so it is reasonable to expect model selection to determine the better approximation.
@@
 
-One obvious limitation in this comparison is that scenario B treats each model as fixed over the entire course of the management simulation.
-In reality, managers will typically re-estimate model parameters after each subsequent observation.
-And rather than consider each model/parameter combination in isolation, managers will generate forecasts which reflect the current uncertainty as to which model or parameter values are most likely. 
-Re-estimating model parameters results in adjusting those probabilities.
-This approach is characterized by adaptive management for sequential decision problems [e.g. @Smith1981], which I employ in the next section.
+## Integrated assessments and adaptive management
 
-## Decision-Theoretic Approaches
-
-Any adaptive management strategy updates posterior distributions over model uncertainty [@Punt2016; @Ludwig1982].
-Unfortunately, in this case, any such adaptive updating leads to worse outcomes than the equivalent non-adaptive strategy, in which model uncertainty is held fixed.
-I illustrate the application of a passive adaptive management strategy to this simple example, following classic examples for parameter [@Ludwig1982] or structural [@Smith1981] model uncertainty.
-Passive adaptive management for a simple sequential decision problem is straightforward to implement over a discrete set of states and actions using dynamic programming with iterative updates [@Smith1981, example code in Appendix].
-To demonstrate that the behavior is not driven by failure to explore sufficiently, (active adaptive management), I will assign initial probability that model 2 is true at 1%.
-After a single iteration of adaptive learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct [Fig 3A].
-As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating, which achieves a net present value that only 31% that expected under management using model 1 alone [Fig 2].
+Rather than select a single best model (or best parameter value), a manager could choose to integrate over possible outcomes generated from all candidate models.
+Updating posterior distributions over parameters and/or weights assigned to different models are examples of this kind of adaptive management [@Punt2016; @Ludwig1982].
+I illustrate the application of such an adaptive management strategy, following classic examples for parameter [@Ludwig1982] or structural [@Smith1981] model uncertainty.
+I first consider only the same two models considered in the previous example. 
+I later consider a larger suite of 42 models, spanning the parameter space of Gordon-Schaefer curves.
+To avoid failure to explore sufficiently, [see exploration-exploitation tradeoff, e.g. @Walters1981],
+I assign prior belief of 99% weight on the optimally performing model, model 1.  
+For comparison, I consider the baseline case in which the manager does not update posterior distribution over which models/parameter values are correct (the manager still chooses a new TAC after each observation, but does not update their belief in the model, i.e.
+does not _learn_ over time).
+The difference between the performance with and without learning is known as the "Value of Information".
+In both 2-model and 42-model scenarios, the value of information is strongly negative.
+The 2-model case achieves a net present value to -58% of the value of having used model 1 alone [Fig 4]. 
+Including all 42 models reduces this to a value of -32%.
+Both harvests and fish biomass remain significantly lower under adaptive learning scenarios.
 
@@
 
+out <- am %>% group_by(method) %>% summarize(total = sum(quota*.99^time))
+( out[out$method=="learning","total"][[1]] -  out[out$method=="planning","total"][[1]]) / out[out$method=="planning","total"][[1]]
+
+out <- am_multi %>% group_by(method) %>% summarize(total = sum(quota*.99^time))
+( out[out$method=="learning","total"][[1]] -  out[out$method=="planning","total"][[1]]) / out[out$method=="planning","total"][[1]]
+
 fig3a + fig3b
@@
 
-So far we have considered only two alternative combinations of the parameters $r$, $K$ and $\sigma$. 
-This simplifies the calculations, because each unique parameter value combination requires a new run of the SDP algorithm to determine the optimal policy from the corresponding forecast.
-Increasing the space of possible models to cover a whole plausible range of parameters $r$ and $K$ does little to resolve this problem [Fig 3B]. 
-Iterative updates again quickly dismiss the parameter values assumed by model 1, though with more options to choose from, this probability is spread over a range of seemingly plausible candidate models instead of a single alternative model (see Appendix).
-While the adaptive management of additional actions and observations slowly narrow this subset of plausible models, decisions based on this uncertainty prevent fish stocks from recovering fully, and realize lower harvests as a result.
-Note that learning under either adaptive management approach (using two models or 42), the decision-maker becomes ever-increasingly convinced that they are using the right model or models.
-Future stock sizes fall consistently in the range predicted by the model(s), and consistently outside the range predicted by model 1.
-Consequently, each iteration the managers are only more firmly convinced that they are maintaining the fish stock near the biomass that supports maximum sustainable yield, when in fact they are sustaining a harvest regime that is preventing recovery of the stock to the much higher productivity regime which would have been achieved under model 1.
+<!-- Intuition, Figure 4 -->
 
-# Discussion
-
-<!-- quick summary of what was demonstrated -->
-Given this simple decision problem in which one of the two models leads to better ecological and economic outcomes, current approaches invariably choose the wrong one.
-Moreover, despite continuing to collect new observations, the decision maker has no way of realizing their mistake.
-The manager is trapped into believing whichever model produces the better forecast, even when this results in decidedly worse objective outcomes.
-Re-estimating parameters with as new observations accumulate only reinforces the problem [Fig 3A], and
-introducing a larger suite of models, such as our wide range of $r$ and $K$ values, does not escape this trap either [Fig 3B].
-Other model choice approaches such as goodness-of-fit, information criteria or cross-validation would all prefer model 2 as well.
-Only by including the true model in our set of candidates can we be certain that forecast-based methods will converge on optimal outcomes.
-
-<!-- Intuition, Figure 4 -->
 The reason for model 1's seemingly contradictory ability to make good decisions but bad forecasts becomes obvious once we compare both curves to that of the underlying model, model 3.
@@
 
+How can the very different forecasts from model 1 and model 3 could produce exactly the same optimal management policy (Fig 5B) under the SDP algorithm?
+Analytic solutions offer more insight as to when and why very different forecasts can generate the identical policy.
+Such a solution was first provided by @Reed1979, who demonstrated the optimal policy in the case considered here would be a so-called "bang-bang" policy.
+Intuitively one can think of this as maintaining the biomass at the most productive size: the maximum population growth rate (position of the peak of the growth curves in Fig 5A), though this is only precisely true without discounting ($\delta = 1$): the optimal stock size $\hat x$ is the solution to $f(\hat x) = \hat x/\delta$ when stochasticity is sufficiently small [@Reed1979].
+Thus, all models in which the peak growth rate occurs at the same stock size will have the same optimal policy.
+These are not merely bad models getting lucky -- all such models correctly capture the crucial feature relevant to the decision.
+In more complex models, such features are more difficult or impossible to identify analytically; but this does not mean they do not exist.
+For instance, Recent mathematical breakthroughs such as @Holden2015, @Hening2019, and @Hening2021 have proven that the optimal harvest control rule in age-structured and predator-prey systems maintain similar bang-bang dynamics.
+This means that the optimal policy of such very complex models will once again be shared by infinite number of simpler models.
+
+# Discussion
+
+The forecast trap illustrated in both examples can best be understood as a problem of _non-uniqueness_.
+Even modestly complex models can successfully predict the observed dynamics, but for wrong mechanistic reasons [@Schnute2001; @Schindler2015].
+In both examples, a decision-maker who accepts the model which leads to very accurate predictions as the basis for their decision-making winds up in the forecast trap: accepting poor ecological and economic outcomes as the best possible option.
+The space of possible models is infinitely large when measured against any a scalar metric such as mean forecast skill or expected net utility.
+Perhaps it should be no surprise then that many models will achieve the same policy outcomes or achieve comparable predictive accuracy.
+Just as the forecast skill is not unique, both examples also demonstrated that the optimal policy is not unique to the "true model" -- many models will result in the same policy and achieve the same outcomes; despite making very different forecasts.
+
+The forecast trap is likely to be more common in contexts in which systems are more complex, partially observed, and available actions are constrained -- all features which are particularly common to ecological management and conservation.
+Because simplicity of the second example allows analytic theory to reveal a precise explanation, it is tempting to assume the trap is only a consequence of examining overly simple models. 
+In fact, the opposite is true.
+In the second example, the "true" model is simple enough to be covered by a suitable candidate set of models (for instance, a Gaussian Process, @Boettiger2015).
+Given sufficient data, @Gneiting2007's theorem regarding strictly proper scores guarantees we will eventually select the underlying model and achieve the optimal forecast and optimal decision.
+In reality, our models never span the true process.
+Partially-observed systems increase the space of possible models that achieve comparable predictive accuracy.
+Constraints on action space such as adjustment costs [@Boettiger2016] or piecewise-linear control rules [@Punt2010] increase the space of models which will result in the same policy.
+Both of these aspects make the forecast trap easier to encounter in our opening example.
+
+The forecast trap demonstrates that over a given ensemble of candidate models, the value of information (VOI) can be negative. 
+Consequently, methods to select models or re-estimate parameters can lead to worse outcomes than had these new observations simply been ignored.
+Crucially, a manager implementing the optimal policy from the most predictive model sees no indication that their models are wrong -- the declining ecosystem and economic returns observed under model A in the first example or under iterative learning in the second example are completely consistent with and predicted by the models.
+Only by winding back the clock, making decisions based over the original uncertainty without learning (Fig 4), can we see that better outcomes could have been achieved.
+The policy derived without learning reflects greater uncertainty: it is more _robust_.
+
+## A way forward
+
+In practice, managers rarely rely only on forecasting skill to assess models, nor determine policies directly from forecasts alone, as in the SDP  [@Schindler2015].
+Three principles already established in management practice but too often overlooked in research [@Schindler2015] can also help avert the forecast trap: (1) methods for articulating uncertainty
+
+ (a) better articulation of uncertainty; (b) more attention to which elements of uncertainty really matter to the decision; (c) methods to detect when the truth falls outside the model set; and (d) methods for generating strategies that are robust to the sort of uncertainty described here.
+
+@Schindler2015 contrasts 
+Many disciplines use scenario planning to assist policy in accounting for irreducible uncertainties [@Peterson2003], in which a range of methods, including model-based forecasts, can be used to determine plausible scenarios.
+
+It is worth noting that quantitative and algorithmic approaches to decision-making need not be predicated on forecasting skill. 
+Recently developed reinforcement learning which have been applied to Atari video-games, car racing, and nuclear fusion are based on "model-free" algorithms which make no attempt to learn and predict future states of the systems they control.
+Instead, the algorithm's neural networks are trained only to learn a  robust "policy", mapping observations to actions.
+While a search space of policies: say, $n$ possible states by $m$ possible actions is potentially very large, it is always far smaller than the corresponding search space of forecasts: which for a time horizon of $T$ is $(m \cdot n^2)^T$.
+
+a type of machine learning, rely on so-called "model free" approaches which do not involve
+
+Several ways to avoid the forecast trap are
+
+The forecast trap can be avoided by seeking policies which are robust to uncertainty [e.g. @Schindler2015].
+
+For instance,
+
+A quantitative decision process does not need to be derived from a forecast
+
+The forecast trap is a consequence of model non-uniqueness [@Konikow1992] of models with respect either to forecast skill or policy outcomes.
+
+Models of any natural system can never be validated, because natural systems are never closed and models are not unique [@Oreskes1994].
+
+@Schindler2015
+
+<!--
 Perhaps this should not be surprising: ecologists have long observed that all models are wrong and the choice of better model depends on the the modeling goals [@Levins1966; @Walters1978; @Ludwig1993; @Getz2018].
-And we are clearly considering different goals: forecast skill (a unitless statistical measure) vs policy outcomes (be they measured in dollars or fish in the ocean).
+Forecast skill (a unitless statistical measure) is clearly a different objective than policy outcomes (be they measured in dollars or fish in the ocean).
 Yet the result is surprising all the same. 
-The forecast isn't just some other arbitrary modeling objective; it is a central input into the decision making process, both in the real world [@Clark2001; @Dietze2018] and in our idealized decision-making algorithm, SDP [@Marescot2013].
+The forecast is not just some other arbitrary modeling objective; it is a central input into the decision making process, both in the real world [@Clark2001; @Dietze2018] and in our idealized decision-making algorithm, SDP [@Marescot2013].
 Nor can we say the same model can never be best at both goals -- obviously the 'true' model always optimizes both objectives [^1]. 
@@
 The example presented here proves this is by no means guaranteed, but also begs the question -- how common is this forecast trap?
-
-[^1]: With enough data from enough of the state space, an SDP algorithm using a Gaussian Process prior [@Boettiger2015], which spans the "true model" given by Eq 4, will escape the forecast trap, as guaranteed by @Gneiting2007's theorem.
-However, real ecological systems are much more complex than Eq 4, and not so easily spanned by mathematical models.
-
-It may seem reasonable to expect that the forecast trap would be rare in real world situations:
-the chance that the candidate set of models would include anything coming close to the optimal policy of the (unknown) true model without also providing a good forecast seems like it ought to be vanishingly small.
-In our example, we were only able to capture the optimal policy with a Gordon-Schafer model because it turned out the optimal policy boiled down to a very simple rule. 
-Surely this does not happen in the more complex models of the real world?
-Surprisingly, more complex models offer no such guarantee, while real-world constraints make this situation *more* likely, not less.
-
-How can the very different forecasts from model 1 and model 3 could produce exactly the same optimal management policy (Fig 4B) under the SDP algorithm?
-Analytic solutions offer more insight as to when and why very different forecasts can generate the identical policy.
-Such a solution was first provided by @Reed1979, who demonstrated the optimal policy in the case considered here would be a so-called "bang-bang" policy.
-Intuitively one can think of this as maintaining the biomass at the most productive size: the maximum population growth rate (position of the peak of the growth curves in Fig 4A), though this is only precisely true without discounting ($\delta = 1$): the optimal stock size $\hat x$ is the solution to $f(\hat x) = \hat x/\delta$ when stochasticity is sufficiently small [@Reed1979].
-Thus, all models in which the peak growth rate occurs at the same stock size will have the same optimal policy.
-These are not merely bad models getting lucky -- all such models correctly capture the crucial feature relevant to the decision.
-In more complex models, such features are more difficult or impossible to identify analytically; but just because we cannot intuit the optimal policy does not mean it is uniquely complex.
-
-Do more complex models lead to more complex control rules?
-The theorems of @Reed1979, while quite general, say nothing about structured models or those with predator-prey or competitive interactions.
-Yet recent mathematical breakthroughs such as @Holden2015, @Hening2019, and @Hening2021 have finally been able to extend Reed's theorem to such cases more generally. 
-As with Reed's result, these recent proofs make it clear that the optimal strategy for managing these more complex models is not unique, but will be shared by many much simpler models.
-So while the true population dynamics may be given by very complex non-linear functions of the interacting species, there will be simple two species models and even un-coupled, one-species models which would lead to the identical optimal policy.
-It may be more likely that our candidate set contains a model which matches the optimal policy than that it contains a model which matches the generative process.
+-->
 
@@
 
-Because even complex models frequently have simple control rules or else we are constrained to simple control rules as a practical matter, it is much more likely that any set of candidate models includes parameterizations that reproduce the true optimal policy than that they reproduce the optimal forecast.
-When data is initially limited, those simple models that could generate optimal outcomes may have non-negligible probabilities associated with them.
-This sets the stage for the forecast trap, adjustments with additional data or comparisons against alternative models that produce more accurate forecasts erodes those probabilities in favor of models which lead to less desirable outcomes.
-
 The forecast trap may become a more acute issue as the simple, process-based models that have historically underpinned ecology and conservation policy are challenged by more accurate forecasts from statistical and machine learning tools.
@@
 I hope this simple and intuitive example will provide a ready reminder as to why the model that produces the best forecast will not always produce the best decision.
+
+# A way forward
 

### 2022-02-23 notebook/appendix-A.Rmd
:pen:

--- 2022-02-15 notebook/alternate-models.Rmd
@@
 ---
-output: github_document
+output: pdf_document
+title: Appendix A: Multispecies models
 ---
@@
 
-- Model 1: A reasonable approximation model that does not produce very accurate forecasts, but nevertheless leads to good management decisions.
-- Model 2: A reasonable approximation model that produces more accurate forecasts but leads to substantially worse management outcomes
-- Model 3: The "True" Model: five-species model; functional forms and parameter values are never disclosed to the 'manager' but are used to simulate all data and management scenarios.
+We consider a simulated system described by the 5-dimensional model of a herring fishery presented in @Brias2021:
+
+\begin{align}
+\end{align}
+
+We will assume this is the "true" model, which is unknown to managers.
+
+While richer numerical models can successfully capture the full complexity of this five-dimensional model, we must remember that reality is always more complex than our models.
+As the term implies, models intentional simplifications of reality.
+
+- Model A: A reasonable approximation model that produces more accurate forecasts but leads to substantially worse management outcomes
+- Model B: A reasonable approximation model that does not produce very accurate forecasts, but nevertheless leads to good management decisions.
+- Model C: The "True" Model: five-species model; functional forms and parameter values are never disclosed to the 'manager' but are used to simulate all data and management scenarios.
 
@@
 
-```{r}
-df <- model_5sp(0, .0, sigma = 0.01)
-
-df %>%
-    pivot_longer(tidyselect::starts_with("x"),
-                    names_to = "species",
-                    values_to = "biomass") %>%
-    ggplot(aes(time, biomass, group= interaction(species,r), col=species)) + geom_path() + ggtitle("Unfished")
-```
-
-Starting from the unfished equilibrium,
-
-```{r}
-df <- model_5sp(0, .58, sigma = 0.01)
-
-df %>%
-    pivot_longer(tidyselect::starts_with("x"),
-                    names_to = "species",
-                    values_to = "biomass") %>%
-    ggplot(aes(time, biomass, group= interaction(species,r), col=species)) + 
-  geom_path() + ggtitle("Optimally managed")
-```
-
 There are many other models, including process-based models, statistical models or even black-box machine learning models
 which 5 dimensional state-space and two-dimensional action space.
-
-```{r}
-## Three-species 'view' of the underlying dynamics, without fishing, starting all species from a low level
-
-#true_observations <- model_5sp(0, 0., sigma = 0.05, x1o =.033, x2o = .033, x3o = .033, x4o = .1, x5o = .1, Tmax = 50)
-true_observations <- model_5sp(0, 0, sigma = 0.05)
-
-obs <- true_observations %>% 
-  transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3, r=r) %>%
-    pivot_longer(any_of(c("cormorant", "bass", "herring")),
-                    names_to = "species",
-                    values_to = "biomass")
-
-obs %>%
-    ggplot(aes(time, biomass, col = species, group = interaction(species,r))) +
-    geom_path(alpha=0.4)
-```
 

### 2022-02-25 manuscript/appendix-B.Rmd
appendix A

--- 2022-02-23 manuscript/appendix.Rmd
@@
 ---
-title: 'Appendix'
+title: 'Appendix B: Forecast Trap in Sequential Decision-Making'
 author: Carl Boettiger
-journal: TBD
+journal: Ecology Letters
 layout: 3p

### 2022-02-25 manuscript/round3-manuscript.Rmd
appendix A

@@
 The forecast trap is not unique to this example, but possible whenever (1) the optimal management policy is not unique to the generative process, and (2) the generative process is not in our candidate set of models.
+Existing practices
 
@@
 Model A accurately forecasts the abundance of both bass and cormorants well into the future, but the optimal management strategy derived from its forecasts leads to steadily declining species abundances and overall disappointing net utility.
-Model B produces substantially less accurate forecasts, but Model B's forecasts also lead to better outcomes, seen both in the population abundances and overall net utility.
+Model B produces substantially less accurate forecasts, but nevertheless achieves better outcomes.
+The net utility under model B is in fact nearly identical to the maximum utility attainable given the true model, while management under model A achieves only 38% of that utility.
 
@@
 Both models are consistent with the limited historical data available to them.
+
+```{r example1, fig.cap="Forecast performance and realized outcomes from management of a 5-species system under either model.
+Based on forecast performance alone, Model A clearly performs better, accurately predicting steady declines.
+Model B predicts overly optimistic outcomes for Bass population levels, and overly pessimistic outcomes for cormorants, with observed dynamics falling well outside the predicted range.
+Despite this, net utility achieved under this management regime is virtually optimal, while the declines under model A result in net utility that is only 38% of optimal."}
+example1 <- read_csv("../data/example1.csv")
+
+example1 %>% 
+  ggplot(aes(time, biomass, col= species)) + 
+  geom_point(aes(shape=type)) +
+  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd, lty=type)) +
+  facet_wrap(~scenario)
+```
 
@@
 
-The forecast trap illustrated in both examples can best be understood as a problem of _non-uniqueness_.
+The forecast trap illustrated in both examples can best be understood as a problem of _non-uniqueness_ [@Oreskes1994].
 Even modestly complex models can successfully predict the observed dynamics, but for wrong mechanistic reasons [@Schnute2001; @Schindler2015].
@@
 Only by winding back the clock, making decisions based over the original uncertainty without learning (Fig 4), can we see that better outcomes could have been achieved.
-The policy derived without learning reflects greater uncertainty: it is more _robust_.
+The policy derived without learning reflects greater uncertainty: it is thus more _robust_.
 
@@
 
-In practice, managers rarely rely only on forecasting skill to assess models, nor determine policies directly from forecasts alone, as in the SDP  [@Schindler2015].
-Three principles already established in management practice but too often overlooked in research [@Schindler2015] can also help avert the forecast trap: (1) methods for articulating uncertainty
+In practice, managers rarely rely only on forecasting skill to assess models, nor determine policies directly from forecasts alone, as in the SDP.
+Such existing principles, frequently reflected in management practice can help avert the forecast trap, but are too often overlooked in research predicated on predict-and-prescribe approaches [@Schindler2015].
+These include 
+(a) emphasizing a better articulation of uncertainty a priori; 
+(b) active exploration of new policies
+(c) methods for generating strategies that are robust to the sort of uncertainty described here.
 
- (a) better articulation of uncertainty; (b) more attention to which elements of uncertainty really matter to the decision; (c) methods to detect when the truth falls outside the model set; and (d) methods for generating strategies that are robust to the sort of uncertainty described here.
+A central role of models is to help articulate uncertainty around possible outcomes rather than make precise predictions.
 
-@Schindler2015 contrasts 
-Many disciplines use scenario planning to assist policy in accounting for irreducible uncertainties [@Peterson2003], in which a range of methods, including model-based forecasts, can be used to determine plausible scenarios.
+The danger of an insufficiently broad model ensemble is well understood in many disciplines which use _scenario planning_ to assist policy in accounting for irreducible uncertainties [@Peterson2003].
+Forecasts based on non-mechanistic models such as empirical dynamical modeling [EDM, Ye2015; @Brias2021] may help in articulating a broader ensemble of scenarios.
 
-It is worth noting that quantitative and algorithmic approaches to decision-making need not be predicated on forecasting skill. 
-Recently developed reinforcement learning which have been applied to Atari video-games, car racing, and nuclear fusion are based on "model-free" algorithms which make no attempt to learn and predict future states of the systems they control.
-Instead, the algorithm's neural networks are trained only to learn a  robust "policy", mapping observations to actions.
-While a search space of policies: say, $n$ possible states by $m$ possible actions is potentially very large, it is always far smaller than the corresponding search space of forecasts: which for a time horizon of $T$ is $(m \cdot n^2)^T$.
+A manager may also escape the forecast trap by exploring actions which are not recommended from any of the available forecasts.
+For example, a manager looking at the low stocks and poor harvests achieved under model 2, could decide experimentally to reduce fishing quota. 
+This will allow the population to enter a range of state-space where discrepancies between model 2 and observations are more obvious.
+@Runge2016 describes how such "double-loop learning" can be used to identify when the entire model set is inaccurate, a problem which is not solved by "single-loop" adaptive-management in our examples above.
+@Schindler2015 also underscores the value of flexible policies; rather than "managing solely within the range of past variation; active probing is usually needed."
+@Schindler2015 contrasts this to a typical interpretation of the 'precautionary principle' often cited as a reason to avoid exploratory actions.
+However, just because active exploration can escape the forecast trap does mean it is always a good idea.
 
-a type of machine learning, rely on so-called "model free" approaches which do not involve
+Third, managers may emphasize policy robustness over expected net utility. 
+In many formal treatments, this is not qualitatively different to the analysis considered here: 
+a manager simply chooses a different utility function, such as minimizing 'regret' rather than maximizing expected value [@Polasky2011]. 
+Such approaches are just as vulnerable to bad outcomes (as defined by their own utility functions) whenever models are selected only on the basis of forecast skill.
+But in practice, robust design may emphasize acceptable performance across the widest possible array of scenarios (e.g. candidate models).
+This acts more like a sensitivity analysis of utility with respect to underlying assumptions, rather than an optimization routine.
+Computationally, the former is much simpler, allowing researchers to evaluate the performance of a policy on more complex simulations for which calculating the optimal policy would be prohibitively difficult.
 
-Several ways to avoid the forecast trap are
-
-The forecast trap can be avoided by seeking policies which are robust to uncertainty [e.g. @Schindler2015].
-
-For instance,
-
-A quantitative decision process does not need to be derived from a forecast
-
-The forecast trap is a consequence of model non-uniqueness [@Konikow1992] of models with respect either to forecast skill or policy outcomes.
-
-Models of any natural system can never be validated, because natural systems are never closed and models are not unique [@Oreskes1994].
-
-@Schindler2015
-
-<!--
-Perhaps this should not be surprising: ecologists have long observed that all models are wrong and the choice of better model depends on the the modeling goals [@Levins1966; @Walters1978; @Ludwig1993; @Getz2018].
-Forecast skill (a unitless statistical measure) is clearly a different objective than policy outcomes (be they measured in dollars or fish in the ocean).
-Yet the result is surprising all the same. 
-The forecast is not just some other arbitrary modeling objective; it is a central input into the decision making process, both in the real world [@Clark2001; @Dietze2018] and in our idealized decision-making algorithm, SDP [@Marescot2013].
-Nor can we say the same model can never be best at both goals -- obviously the 'true' model always optimizes both objectives [^1]. 
-It is natural to assume from this that the candidate with the closest forecast will also be the one with the closest policy.
-The example presented here proves this is by no means guaranteed, but also begs the question -- how common is this forecast trap?
--->
-
-Real world considerations actually make this situation more likely, not less likely.
-Managers cannot resort to arbitrarily complex policies, regardless of how complex their models.
-Policy adjustments are costly [@Boettiger2016] and the space of available actions is usually far more limited than the space of available states.
-In some ecological decision-making contexts, a manager may only be able to select between a handful of alternative actions.
-Such constraints make it much more likely that the optimal policy will be replicated by a much simpler model.
-Most well-managed marine fisheries are constrained to constant or simple piecewise-linear harvest control rules [@Punt2010]. 
-This ensures that an infinite number of possible models will share the optimal solution with the true model.
-
-The forecast trap may become a more acute issue as the simple, process-based models that have historically underpinned ecology and conservation policy are challenged by more accurate forecasts from statistical and machine learning tools.
-Buoyed by the rapid expansion of available data and computational power, complex and increasingly opaque models are becoming more common [@Desjardins2019].
-Evaluating such models based on forecast skill will not only reduce concerns about over-fitting, but will make a compelling illustration of their viability as a tool for informing policy.
-In many cases, these more accurate forecasts may very well prove invaluable in delivering better (or less bad) real-world outcomes.
-But this paper is a reminder that such outcomes are by no means an inevitable consequence of better forecasts.
-These more accurate predictions are still not the true model, and it so it is always possible to improve predictive accuracy while less accurately reflecting the unknown key features that really drive the policy decision.
-I hope this simple and intuitive example will provide a ready reminder as to why the model that produces the best forecast will not always produce the best decision.
-
-# A way forward
-
-<!--
-Other fields have already recognized the potential to achieve optimal decisions without the pursuit of ever more accurate predictive models. 
-The leading artificial intelligence algorithms used in robotic automation and game-playing [@starcraftii; @atari] no longer attempt to estimate a predictive model which is then used to select an action.
-In the past several years, so-called "Model-Free" approaches have come to dominate Deep Reinforcement Learning.
-These algorithms have no predictive capacity at all; instead, they try to learn the most effective policy rules directly.
-Instead of estimating a predictive model the cutting edge, algorithms which make no attempt to learn the policy directly.
-Decision makers do not need to predict the future, they only need to know what to do about it.
--->
+Finally, it is worth noting that decisions do not need to be premised on a forecast at all, but can be premised entirely on the basis of past experience.
+If the stock size has gone up, increase harvest slightly, otherwise, decrease slightly.
+Such a policy is not optimal, but it is robust across a wide range of unimodal stock-recruitment curves without ever estimating a predictive model.
+This is the basis of so-called 'model-free' reinforcement learning algorithms such as DQN [@DQN] and SAC [@SAC], which train deep neural networks to learn a policy without ever attempting to predict future states of the underlying process.
+Training such a artificial intelligence agents across a wide suite of simulations,
+a process known as curriculum learning [@], mimics the scenario analyses and search for .
+Such approaches have been used to train agents to play 2600 Atari console games at superhuman ability [@], outperform race-car drivers [@] and control nuclear fusion reactions [@].
+Such approaches are also not yet well understood, and can fail dramatically [@].
 
@@
 
-The author acknowledges support from NSF CAREER Award #1942280 and helpful discussions with Melissa Chapman, Jeremy Fox, and anonymous reviewers.
+The author acknowledges support from NSF CAREER Award #1942280 and helpful discussions with Melissa Chapman.
+The author is also deeply grateful to the Michael Runge and three other anonymous reviewers whose detailed feedback and insights have greatly shaped the paper and my own understanding of these issues.
 

### 2022-02-25 notebook/appendix-A.Rmd
appendix A

--- 2022-02-11 notebook/alternate-models.Rmd
@@
 ---
-output: github_document
+title: 'Appendix A: Forecast Trap in a multi-species model'
+author: Carl Boettiger
+journal: Ecology Letters
+layout: 3p
+bibliography: refs.bib
+csl: csl/ecology-letters.csl
+#output:
+#  tufte::tufte_handout:
+#    dev: cairo_pdf
+#    latex_engine: xelatex
+output: 
+  hrbrthemes::ipsum_pdf:
+    dev: cairo_pdf
+    latex_engine: xelatex
+
+header-includes:
+  - \usepackage[left]{lineno}
+  - \linenumbers
+
 ---
 
-- Model 1: A reasonable approximation model that does not produce very accurate forecasts, but nevertheless leads to good management decisions.
-- Model 2: A reasonable approximation model that produces more accurate forecasts but leads to substantially worse management outcomes
-- Model 3: The "True" Model: five-species model; functional forms and parameter values are never disclosed to the 'manager' but are used to simulate all data and management scenarios.
+This appendix illustrates how the forecast trap can emerge in choosing between two alternative three-species models which may be used to manage a partially observed five-species system. 
+The underlying dynamics for the simulated "true" system are based on the five-dimensional model of a herring fishery presented in @Brias2021.
+Following @Brias2021, we assume three species of herring, $X_1$, $X_2$, $X_3$ are preyed upon by
 
-- Define a model
-- Define action space. Under the assumption of fixed action, problem is not MDP and trivial to solve by sweep through possible fixed actions.
+\begin{align}
+x_{1,t+1} &= s_{1,t} \exp{(1.0213 -        s_{1,t} - 0.0861 s_{2,t} - 0.3141 s_{3,t} - 0.7252 s_{4,t} -0.2445 x_{5,t} +\espilon_{1,t} )}
+x_{2,t+1} &= s_{2,t} \exp{(1.0289 - 0.4765 s_{1,t} -        s_{2,t} - 0.1370 s_{3,t} - 0.9811 s_{4,t} -0.0915 x_{5,t} +\espilon_{2,t} )}
+x_{3,t+1} &= s_{3,t} \exp{(1.0207 - 0.3193 s_{1,t} - 0.3461 s_{2,t} -        s_{3,t} - 0.6367 s_{4,t} -0.8716 x_{5,t} +\espilon_{3,t} )}
+x_{4,t+1} &= s_{4,t} \exp{(         0.7252 s_{1,t} + 0.9811 s_{2,t} + 0.6367 s_{3,t} -        s_{4,t}                 +\espilon_{4,t})}
+x_{5,t+1} &= x_{5,t} \exp{(         0.2445 s_{1,t} + 0.0915 s {2,t} + 0.8716 s {3,t} -                        x_{5,t} +\espilon_{5,t} )}
+\end{align}
+
+where
+
+\begin{equation}
+s_{i,t} = x_{i_t} \left(1 - u_{i,t}\right)
+\end{equation}
+
+\begin{equation}
+\epsilon_{i,t} \stackrel{\texrm{iid}}{\tilde} \mathcal{N(0, \sigma_i)}
+\end{equation}
+
+We will assume this is the "true" model, which is unknown to managers.
+
+Herring are harvested together, $u_1 = u_2 = u_3 = u_{\textrm prey}$, while bass are harvested at effort $u_{\textrm predator}$.
+The manager's utility in year $t$ is given by
+
+\begin{equation}
+R(\vec x_t, \vec u_t) = w_3 x_5 + w_2 x_4 + \sum_i^{1,2,3} w_1 x_i
+\end{equation}
+
+We will consider the case which places 50% of the weight on the cormorant conservation objective, $w_3 = 0.5$, and 25% weight on the harvest of bass and herring respectively, $w_2 = w_1 = 0.25$.
+
+In contrast to @Brias2021, we will assume that the manager must simply select the harvest efforts $u_{\textrm prey} \in [0,1]$ and $u_{\textrm predator} \in [0,1]$ to be used over the duration of the management scenario, rather than being able to adjust these annually.
+In most real-world systems it is not possible to measure all variables which influence the dynamics, some dimensions are treated as latent variables of the model.
+To illustrate this, we will assume that the manager observes only the population abundance of bass $x_{4,t}$, and cormorants, $x_{5,t}$, but not the population size of any of the three species of herring.
+
+### Model A
+
+Model A is a possible 3-species approximation that produces more accurate forecasts but leads to substantially worse management outcomes
+This model treats the three species of herring collectively as $x_1$. 
+Parameters are estimated consistent with limited historical data of the observed variables, bass biomass $x_4$ and herring biomass $x_5$.
+
+\begin{equation}
+x_{1,t+1} &= s_{1,t} \exp(3 - 8 s_{1,t} - 0.9 s_4 - 0.9 * x_{5,t} + \espilon_{1,t} )
+x_{4,t+1} &= s_{4,t} \exp(0.1 s_{1,t} - s_{4,t} + \espilon_{4,t}  )
+x_{5,t+1} &= x_{5,t} \exp(0.3 s_{1,t} - x_5  + \espilon_{5,t}  )
+\end{equation}
+
+### Model B
+
+Model B is a structurally different model which postulates a further simplification of the dynamics.
+This slightly over-simplified approximation does not produce very accurate forecasts, but nevertheless leads to good management decisions.
+
+\begin{equation}
+x_{1,t+1} &= s_{1,t} \exp(1.60 - 2.00 s_{1,t}  + \espilon_{1,t} )
+x_{4,t+1} &= s_{4,t} \exp(1.45 - 2.75 s_{1,t}  + \espilon_{4,t}  )
+x_{5,t+1} &= 0.25 x_{5,t} 
+\end{equation}
+
+Our alternate model has several oversimplifications:
+(1) it treat all three herring species as a single species,
+(2) it fails to capture the coupled predator-prey dynamics between bass and herring, 
+(3) it oversimplifies the cormorant dynamics, assuming the cormorant population is determined to be a fixed fraction of the herring,
+and ignoring the impact of cormorant's predation on the herring itself.
+(4) Lastly, our model will overestimate the mortality introduced on herring by a given fishing effort.
+These elements are all obviously wrong, but not so arbitrary as to be inconceivable as a candidate model.
+Researchers frequently consider models which make oversimplifications all the time, and rely on model choice processes to weed them out.
+
+The parameterization chosen for the oversimplified 3 species model can reasonably reproduce the un-fished dynamics.
+Under any harvesting regime or other influence that perturbs the system significantly far from the un-fished equilibrium co-existence state of the model quickly reveals the poor forecasting ability of this model.
+Nevertheless, solving for the optimal policy (under the same constraints of constant harvest fractions as we consider for the true model) results in a harvest policy which provides nearly optimal performance.
+
+# Software implementation
+
+```{r graphics_setup, message=FALSE, warning=FALSE, include = FALSE}
+## Plotting themes, colors, and fonts 
+## aesthetics choices only, all can be omitted without consequence
+library(ggthemes)
+library(hrbrthemes)
+library(ggplot2)
+library(Cairo)
+library(extrafont)
+library(patchwork)
+library(styler)
+extrafont::loadfonts(quiet = TRUE)
+ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
+
+scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
+scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
+pal <- ggthemes::solarized_pal()(8)
+txtcolor <- "#586e75"
+
+knitr::opts_chunk$set(cache=FALSE, tidy = "styler", message = FALSE, warning = FALSE, echo = TRUE)
+
+```
 
@@
 }
+
+clip <- function(x, lower=0, upper=1) {
+    x[x <= lower] <- lower
+    x[x >= upper] <- upper
+    x
+}
 ```
 
-## Five species
+Define the reward for a given state,action pair and the net utility
+
+```{r utility}
+## Compute the reward associated with a given simulation df given 
+## rates for prey and predator species (x[1],x[2])
+reward <- function(df,
+                   x,
+                   w3 = 0.5,              # Conservation value
+                   w2 = 0.25,             # Predator harvest value
+                   w1 = 1 - (w2+w3),      # Prey harvest value
+                   delta = 0.999
+                  ) {
+    u_prey <- clip(x[1], 0, 1)
+    u_pred <- clip(x[2], 0, 1)
+    R <-
+        w1 * u_prey * (df$x1 + df$x2 + df$x3) +
+        w2 * u_pred * df$x4 +
+        w3 * df$x5
+    t <- seq_along(R)
+    as.numeric(R %*% delta ^ t)
+}
+
+# Function to be optimized for fixed choice of harvest rates x[1],x[2]
+utility <- function(x, f, ...) {
+
+    u_prey <- clip(x[1], 0, 1)
+    u_pred <- clip(x[2], 0, 1)
+    df <- f(u_prey, u_pred, ...)
+
+    # average accross the reps    
+    df <- df %>% group_by(time) %>% summarise(across(.fns=mean))
+
+    # negative since optimizer minimizes
+    - reward(df, x)
+}
+
+```
+
+## Model definitions
+
+### True five species-model
 
@@
 
-```{r utility}
-clip <- function(x, lower=0, upper=1) {
-    x[x <= lower] <- lower
-    x[x >= upper] <- upper
-    x
+### Model A
+
+```{r}
+model_A <- function(u_prey = 0.5,
+                      u_pred = 0.5,
+                      sigma = 0.05,
+                      x1o = 0.44, # unfished equib
+                      x2o = 0.17,
+                      x3o = 0.186,
+                      x4o = 0.600,
+                      x5o = 0.280,
+                      Tmax = 50,
+                      R1 =  3, #1.15, #1.0213,
+                      A11 = 8,
+                      A14 = 0.9, # 0.7252,
+                      A15 = 0.9, # 0.2445,
+                      A41 =  0.1, #0.7252,
+                      A51 = 0.3, # 0.2445,
+                    reps = 40) {
+
+    df <- tibble()
+    for(r in 1:reps){
+
+      x1 <- x4 <- x5 <- numeric(Tmax)
+      s1 <-  s4 <- numeric(Tmax)
+      x2 <- x3 <- numeric(Tmax)
+      xi1 <- rnorm(Tmax, 0, sigma)
+      xi4 <- rnorm(Tmax, 0, sigma)
+      xi5 <- rnorm(Tmax, 0, sigma)
+
+      x1[1] <- x1o + x2o + x3o
+      x4[1] <- x4o
+      x5[1] <- x5o
+
+      for (t in 1:(Tmax-1)) {
+          s1[t] <- x1[t] * (1 - u_prey)
+          s4[t] <- x4[t] * (1 - u_pred)
+          x1[t+1] = s1[t]*exp(R1 - A11*s1[t] - A14*s4[t] - A15 * x5[t] + xi1[t])
+          x4[t+1] = s4[t]*exp(A41*s1[t] - s4[t] + xi4[t])
+          x5[t+1] = x5[t]*exp(A51*s1[t] - x5[t] +  xi5[t])
+
       }
-
-## Compute the reward associated with a given simulation df given 
-## rates for prey and predator species (x[1],x[2])
-reward <- function(df,
-                   x,
-                   w3 = 0.5,              # Conservation value
-                   w2 = 0.25,             # Predator harvest value
-                   w1 = 1 - (w2+w3),      # Prey harvest value
-                   delta = 0.999
-                  ) {
-    u_prey <- clip(x[1], 0, 1)
-    u_pred <- clip(x[2], 0, 1)
-    R <-
-        w1 * u_prey * (df$x1 + df$x2 + df$x3) +
-        w2 * u_pred * df$x4 +
-        w3 * df$x5
-    t <- seq_along(R)
-    as.numeric(R %*% delta ^ t)
+      df <- bind_rows(df, tibble(time = 1:Tmax, x1, x2, x3, x4, x5, r=r))
     }
-
-# Function to be optimized for fixed choice of harvest rates x[1],x[2]
-utility <- function(x, f = model_5sp, ...) {
-
-    u_prey <- clip(x[1], 0, 1)
-    u_pred <- clip(x[2], 0, 1)
-    df <- f(u_prey, u_pred, ...)
-
-    # average accross the reps    
-    df <- df %>% group_by(time) %>% summarise(across(.fns=mean))
-
-    # negative since optimizer minimizes
-    - reward(df, x)
+    df
 }
-
 ```
 
-```{r}
-
-o <- optim(c(0.01,0.5), utility, method="L-BFGS-B", lower = c(0,0), upper = c(1,1))
-o$par
-o$value
-
-o <- optim(c(0.01,0.5), utility)
-clip(o$par, 0, 1)
-o$value
-
-x <- c(0.0,.58)
-utility(x)
-
-```
-
-For the sake of argument, let us allow the model of Brias & Munch to represent the "true" model, which is unknown.
-As researchers, we only have access to data derived from observations from the model. 
-Let us consider a sample simulation of such data:
+### Model B
 
 ```{r}
-df <- model_5sp(0, .0, sigma = 0.01)
-
-df %>%
-    pivot_longer(tidyselect::starts_with("x"),
-                    names_to = "species",
-                    values_to = "biomass") %>%
-    ggplot(aes(time, biomass, group= interaction(species,r), col=species)) + geom_path() + ggtitle("Unfished")
-```
-
-Starting from the unfished equilibrium,
-
-```{r}
-df <- model_5sp(0, .58, sigma = 0.01)
-
-df %>%
-    pivot_longer(tidyselect::starts_with("x"),
-                    names_to = "species",
-                    values_to = "biomass") %>%
-    ggplot(aes(time, biomass, group= interaction(species,r), col=species)) + 
-  geom_path() + ggtitle("Optimally managed")
-```
-
-There are many other models, including process-based models, statistical models or even black-box machine learning models
-which 5 dimensional state-space and two-dimensional action space.
-
-```{r}
-## Three-species 'view' of the underlying dynamics, without fishing, starting all species from a low level
-
-#true_observations <- model_5sp(0, 0., sigma = 0.05, x1o =.033, x2o = .033, x3o = .033, x4o = .1, x5o = .1, Tmax = 50)
-true_observations <- model_5sp(0, 0, sigma = 0.05)
-
-obs <- true_observations %>% 
-  transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3, r=r) %>%
-    pivot_longer(any_of(c("cormorant", "bass", "herring")),
-                    names_to = "species",
-                    values_to = "biomass")
-
-obs %>%
-    ggplot(aes(time, biomass, col = species, group = interaction(species,r))) +
-    geom_path(alpha=0.4)
-```
-
-Now consider an alternative model which ignores some of this biological complexity:
-Our alternate model has several oversimplifications:
-(1) it treat all three herring species as a single species,
-(2) it fails to capture the coupled predator-prey dynamics between bass and herring, 
-(3) it oversimplifies the cormorant dynamics, assuming the cormorant population is determined to be a fixed fraction of the herring,
-and ignoring the impact of cormorant's predation on the herring itself.
-(4) Lastly, our model will overestimate the mortality introduced on herring by a given fishing effort.
-These elements are all obviously wrong, but not so arbitrary as to be inconceivable as a candidate model.
-Researchers frequently consider models which make oversimplifications all the time, and rely on model choice processes to weed them out.
-
-The parameterization chosen for the oversimplified 3 species model can reasonably reproduce the unfished dynamics.
-Under any harvesting regime or other influence that perturbs the system significantly far from the unfished equilibrium co-existence state of the model quickly reveals the poor forecasting ability of this model.
-Nevertheless, solving for the optimal policy (under the same constraints of constant harvest fractions as we consider for the true model) results in a harvest policy which provides nearly optimal performance.
-
-```{r}
-model_3sp <- function(u_prey = 0,
+model_B <- function(u_prey = 0,
                       u_predator = 0,
@@
 
-Does this model produce good forecasts?
+## Optimal decision
+
+Determine the optimal policy using Nelder-Meade simplex method provided by the `optim` function:
 
 ```{r}
-# unfished equilibrium
-m1_forecast <- model_3sp(0, 0.0, sigma = 0.05,  Tmax = 50)
-
-predictions <- m1_forecast %>%
-  transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3, r=r) %>%
-  pivot_longer(any_of(c("cormorant", "bass", "herring")),
-                  names_to = "species",
-                  values_to = "biomass") %>% 
-  group_by(time, species) %>% 
-  summarize(sd = sd(biomass), biomass = mean(biomass))
-
-predictions %>%
-  ggplot(aes(time, biomass, col= species)) + 
-  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd)) +
-  geom_point(data = filter(obs, r==1))
+u_A <- function(x) utility(x, f = model_A)
+A <- optim(c(.2,0.01), u_A)
+model_A_policy <- clip(A$par, 0, 1)
 ```
@@
 ```{r}
-# unfished equilibrium
-m1_forecast <- model_3sp(0, 0.58, sigma = 0.05,  Tmax = 50)
-
-predictions <-   
-m1_forecast %>%
-    transmute(time = time, cormorant = x5, bass = x4, herring = x1, r=r) %>%
-    pivot_longer(any_of(c("cormorant", "bass", "herring")),
-                    names_to = "species",
-                    values_to = "biomass") %>% 
-    group_by(time, species) %>% 
-  summarize(sd = sd(biomass), biomass = mean(biomass))
-
-predictions %>%
-ggplot(aes(time, biomass, col= species)) + 
-  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd)) +
-  geom_point(data = filter(obs, r==1))
+u_B <- function(x) utility(x, f = model_B)
+B <- optim(c(0.5,0.1), u_B)
+model_B_policy <- clip(B$par, 0, 1)
 ```
@@
 ```{r}
-# Optimal policy under identical utility
-u_3sp <- function(x) utility(x, f = model_3sp)
-
-m3sp <- optim(c(0.5,0.1), u_3sp)
-m3sp$value
-clip(m3sp$par, 0, 1)
-
+u_true <- function(x) utility(x, f = model_5sp)
+o <- optim(c(0.01,0.5), u_true)
+true_policy <- clip(o$par, 0, 1)
 ```
 
-Alternate model, also mistakenly treats the three species of herring as a single species, but does a better job capturing the predator-prey relationships.
-This model also happens to produce better forecasts, but leads to worse decision:
+## Simulate scenarios
+
+We simulate the system using the true model, while managing from the policy derived from forecasts of model A:
 
 ```{r}
-model_B <- function(u_prey = 0.5,
-                      u_pred = 0.5,
-                      sigma = 0.05,
-                      x1o = 0.44, # unfished equib
-                      x2o = 0.17,
-                      x3o = 0.186,
-                      x4o = 0.600,
-                      x5o = 0.280,
-                      Tmax = 50,
-                      R1 =  1.15, #1.0213,
-                      A11 = 1,
-                      A14 = 0.1, # 0.7252,
-                      A15 = 0.1, # 0.2445,
-                      A41 =  0.8, #0.7252,
-                      A51 = 0.3, # 0.2445,
-                    reps = 40) {
+Tmax <- 10
+resp <- 40
 
-    df <- tibble()
-    for(r in 1:reps){
+x <- model_A_policy
+sim <- model_5sp(u_prey = x[[1]], u_pred = x[[2]],
+                 Tmax = Tmax, sigma= 0.05, reps = reps)
+obs <- sim %>% 
+  format_sim() %>%
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass), type = "observed")
 
-      x1 <- x4 <- x5 <- numeric(Tmax)
-      s1 <-  s4 <- numeric(Tmax)
-      x2 <- x3 <- numeric(Tmax)
-      xi1 <- rnorm(Tmax, 0, sigma)
-      xi4 <- rnorm(Tmax, 0, sigma)
-      xi5 <- rnorm(Tmax, 0, sigma)
+predict <- model_A(u_prey = x[[1]], u_pred = x[[2]],
+                     Tmax = Tmax, reps = reps) %>%
+  format_sim() %>% 
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass), type = "predicted")
 
-      x1[1] <- x1o + x2o + x3o
-      x4[1] <- x4o
-      x5[1] <- x5o
+scenario_A <- bind_rows(obs, predict) %>% 
+  filter(species != "herring") %>% mutate(scenario = "Model A")
+mean_utility_A <- reward(df = sim, x = x) / reps
+```
 
-      for (t in 1:(Tmax-1)) {
-          s1[t] <- x1[t] * (1 - u_prey)
-          s4[t] <- x4[t] * (1 - u_pred)
-          x1[t+1] = s1[t]*exp(R1 - A11*s1[t] - A14*s4[t] - A15 * x5[t] + xi1[t])
-          x4[t+1] = s4[t]*exp(A41*s1[t] - s4[t] + xi4[t])
-          x5[t+1] = x5[t]*exp(A51*s1[t] - x5[t] +  xi5[t])
+We repeat the simulation over 40 replicates using the policy derived from model B forecasts:
 
-      }
-      df <- bind_rows(df, tibble(time = 1:Tmax, x1, x2, x3, x4, x5, r=r))
-    }
-    df
-}
+```{r}
+x <- model_B_policy
+sim <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], 
+                 Tmax = Tmax, sigma=0.05, reps = reps)
+obs <- sim %>%
+  format_sim() %>%
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass), type = "observed")
+
+predict <- model_B(u_prey = x[[1]], u_pred = x[[2]], 
+                   Tmax = Tmax, reps = reps) %>% 
+  format_sim()  %>% 
+  group_by(time, species) %>% 
+  summarize(sd = sd(biomass), biomass = mean(biomass), type = "predicted")
+
+scenario_B  <- bind_rows(obs, predict) %>% 
+  filter(species != "herring")  %>% mutate(scenario = "Model B")
+
+mean_utility_B <- reward(df = sim, x = x) / reps
 ```
@@
 ```{r}
-# Optimal policy under identical utility
-u_B <- function(x) utility(x, f = model_B)
-u_B(c(0, .58))
-#optim(c(0.5,0.1), u_B, method = "L-BFGS-B", lower=0, upper=1)
 
-B <- optim(c(0.5,0.1), u_B)
-B$value
-model_B_policy <- clip(B$par, 0, 1)
-model_B_policy
+example1 <- bind_rows(scenario_A, scenario_B)
+write_csv(example1, "../data/example1.csv")
 
-```
-
-Run the simulation using management policy of each
-
-```{r}
-
-Tmax <- 20
-x <- model_B_policy
-#x <- c(0.,0.58)  # Following the optimal policy under model_B
-
-obs <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
-predict_B <- model_B(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
-
-obs_stats <- obs %>%
-  group_by(time, species) %>% 
-  summarize(sd = sd(biomass), biomass = mean(biomass))
-
-predict_stats_B <- predict_B %>% 
-  group_by(time, species) %>% 
-  summarize(sd = sd(biomass), biomass = mean(biomass))
-
-predict_stats_B %>%
+example1 %>% 
   ggplot(aes(time, biomass, col= species)) + 
-  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd), lty=2) + 
-  geom_errorbar(data = obs_stats, aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd))
+  geom_point(aes(shape=type)) +
+  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd, lty=type)) +
+  facet_wrap(~scenario)
 
@@
 ```{r}
-## True observations vs forecasts, 5 steps ahead, while following the Optimal / Model 1 policy, starting with default initial condition
-Tmax <- 5
-x <- c(0,0.58)
-
-obs <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
-predict <- model_3sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
-
-obs_stats <- obs %>%
-  group_by(time, species) %>% 
-  summarize(sd = sd(biomass), biomass = mean(biomass))
-
-predict_stats <- predict %>% 
-  group_by(time, species) %>% 
-  summarize(sd = sd(biomass), biomass = mean(biomass))
-
-## dotted are forecasted values, solid are observed values
-
-predict_stats %>%
-  ggplot(aes(time, biomass, col= species)) + 
-  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd), lty=2) + 
-  geom_errorbar(data = obs_stats, aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd))
-
+mean_utility_A / mean_utility_B
 ```
-
-```{r}
-score_me <- function(x = c(0,0.58), process=model_5sp, forecast=model3p, Tmax=5) {
-
-  obs <- process(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
-  predict <- forecast(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
-
-  #scoringRules::crps_sample(predict$x1, )
-
-}
-
-```

### 2022-02-26 manuscript/appendix-B.Rmd
:pen:

@@
 In Scenario A (passive observation) the fish stock is unharvested and allowed to recover towards carrying capacity (as simulated under our "true" model, Model 3) while comparing the observed stock size in each subsequent time step to the distribution predicted under model 1 and model 2 respectively [Fig 1].
-The mean, $\mu_t$ and variance, $\sigma_t$ of the forecast are compared against the true observation $x_t$ using a proper scoring rule given by @Gneiting2007,
+The mean, $\mu_t$ and variance, $\sigma_t$ of the forecast are compared against the true observation $x_t$ using the log-likelihood proper scoring rule given by @Gneiting2007,
 

### 2022-02-26 manuscript/round3-manuscript.Rmd
:pen:

@@
   Encouraged by decision makers' appetite for future information on topics ranging from elections to pandemics, and enabled by the explosion of data and computational methods, model based forecasts have garnered increasing influence on a breadth of decisions in modern society.
-Using several classic examples from fisheries management, I demonstrate that selecting the model or models that produce the most accurate and precise forecast (measured by statistical scores) can lead to decidedly worse outcomes (measured by real-world objectives).
+Using several classic examples from fisheries management, I demonstrate that selecting the model or models that produce the most accurate and precise forecast (measured by statistical scores) can sometimes lead to worse outcomes (measured by real-world objectives).
 This can create a forecast trap, in which the outcomes such as fish biomass or economic yield decline while the manager becomes increasingly convinced that these actions are consistent with the best models and data available.
-The forecast trap is not unique to this example, but possible whenever (1) the optimal management policy is not unique to the generative process, and (2) the generative process is not in our candidate set of models.
-Existing practices
+The forecast trap is not unique to this example, but a fundamental consequence of non-uniqueness of models.
+Existing practices that promote the consideration of a broader set of models _a priori_ rather than rely on forecast skill alone are the best way to avoid the trap.
 
@@
 When VOI is negative, the decision-maker may become trapped into accepting mediocre outcomes derived from a model that makes accurate forecasts, even when a less accurate model that would generate better outcomes is available.
-I present two examples of this "forecast trap" and examine how it arises.
-This disconcerting situation is possible whenever two conditions are met: (1) the optimal management policy is not unique to the generative process, and (2) the generative process is not included in candidate set of models.
-These conditions do not guarantee the trap will occur, only the circumstances in which it cannot be ruled out entirely.
+I will present two examples of this "forecast trap" and examine how it arises.
 
@@
 However, over-fitting plays no role in the forecast trap, where model predictions are assessed only using probabilistic forecasts, and not observations which had previously been used to fit the models.
-Formally, these scores satisfy the 'proper scoring' rule of @Gneiting2007, which proves no other probabilistic prediction $Q(x)$ will have a better expected score than that of the generative process $P(x)$.
+Formally, these scores satisfy the 'proper scoring' rule of @Gneiting2007, which proves no other probabilistic prediction $Q(x)$ will have a better expected score than that of the true model (i.e.
+generative process), $P(x)$.
 @Gneiting2007's proof of proper scoring has since become a critical tool to avoid over-fitting when choosing models to make decisions, but as I illustrate, will not prevent the forecast trap.
@@
 In this way, the SDP merely stands in for a mathematically precise way in which forecasts are turned into decisions.
-Recognizing that the SDP-derived policy (A) comes directly from the forecast model, and (B) gives the optimal policy for said forecast, seems to suggest that the whatever model makes the better forecast will surely also lead to better outcomes (as measured in terms of whatever utility we have chosen to maximize).
+Recognizing the SDP-derived policy (A) comes directly from the forecast model, and (B) gives the optimal policy for said forecast, seems to suggest that whichever model makes the better forecast will surely also lead to better outcomes (as measured in terms of whatever utility we have chosen to maximize).
 While this intuition is no doubt _often_ accurate, our purpose here is to demonstrate that it is by no means _guaranteed_:
@@
 In the second scenario (lower panel), the manager uses the optimal SDP policy derived from each forecast to determine the TAC for the following year, and compares the observed stock size to that which the model predicted given that fishing quota.
-In both cases, model 2 provides far more accurate forecasts, as seen in the error bars and confirmed by the distribution of proper scores [@Gneiting2007].
+In both cases, model 2 provides far more accurate forecasts, as seen in the error bars and confirmed by the distribution of proper scores [Fig 3C-D; @Gneiting2007].
 Model and simulation details are provided in Appendix B.
@@
 
-## Integrated assessments and adaptive management
+## Iterative upates and adaptive management
 
@@
 In fact, the opposite is true.
-In the second example, the "true" model is simple enough to be covered by a suitable candidate set of models (for instance, a Gaussian Process, @Boettiger2015).
-Given sufficient data, @Gneiting2007's theorem regarding strictly proper scores guarantees we will eventually select the underlying model and achieve the optimal forecast and optimal decision.
+In the second example, the "true" model is simple enough to be covered by a suitable candidate set of models (e.g.
+a Gaussian Process, @Boettiger2015), which would resolve the trap given sufficient data.
 In reality, our models never span the true process.
@@
 
-The forecast trap demonstrates that over a given ensemble of candidate models, the value of information (VOI) can be negative. 
+The forecast trap demonstrates that for certain ensembles of candidate models, the value of information (VOI) [@Howard1966; @Katz1987] can in fact be negative. 
 Consequently, methods to select models or re-estimate parameters can lead to worse outcomes than had these new observations simply been ignored.
@@
 
-In practice, managers rarely rely only on forecasting skill to assess models, nor determine policies directly from forecasts alone, as in the SDP.
-Such existing principles, frequently reflected in management practice can help avert the forecast trap, but are too often overlooked in research predicated on predict-and-prescribe approaches [@Schindler2015].
-These include 
-(a) emphasizing a better articulation of uncertainty a priori; 
-(b) active exploration of new policies
+In practice, managers rarely rely only on forecasting skill to assess models, nor determine policies directly from forecasts alone.
+In both examples presented above, the forecast trap is most likely in circumstances where the collection of candidate models is insufficiently broad.
+Management practice tends to emphasize approaches which _broaden_ rather than _narrow down_ this candidate set.
+This reflects the view that "the primary values of ecosystem models are as heuristic tools for communication and for developing scenarios to express uncertainties and test policies" rather than as a source of reliable forecasts [@Schindler2015].
+The practices include:
+
+(a) emphasizing a better articulation of uncertainty _a priori_; 
+(b) active exploration of alternative policies can reveal when the model set is inadequate
 (c) methods for generating strategies that are robust to the sort of uncertainty described here.
@@
 A central role of models is to help articulate uncertainty around possible outcomes rather than make precise predictions.
+As @Schindler2015 notes, "current approaches to verification and validation of ecosystem models likely produce overly optimistic impressions of the reliability of forecasts underlying management and conservation prescriptions."
+Forecasts based on non-mechanistic models such as empirical dynamical modeling [EDM, @Ye2015; @Brias2021] may help in articulating a broader ensemble of scenarios [@Boettiger2015].
+In contrast, if such approaches are selected solely on forecasting skill, they may increase the probability of the forecast trap.
+The danger of an insufficiently broad model ensemble is well understood in many disciplines which use _scenario planning_ to assist policy in accounting for irreducible uncertainties [@Peterson2003].
 
-The danger of an insufficiently broad model ensemble is well understood in many disciplines which use _scenario planning_ to assist policy in accounting for irreducible uncertainties [@Peterson2003].
-Forecasts based on non-mechanistic models such as empirical dynamical modeling [EDM, Ye2015; @Brias2021] may help in articulating a broader ensemble of scenarios.
-
-A manager may also escape the forecast trap by exploring actions which are not recommended from any of the available forecasts.
+A manager may also escape the forecast trap by exploring actions which are never recommended from any of the available forecasts.
 For example, a manager looking at the low stocks and poor harvests achieved under model 2, could decide experimentally to reduce fishing quota. 
@@
 @Runge2016 describes how such "double-loop learning" can be used to identify when the entire model set is inaccurate, a problem which is not solved by "single-loop" adaptive-management in our examples above.
-@Schindler2015 also underscores the value of flexible policies; rather than "managing solely within the range of past variation; active probing is usually needed."
-@Schindler2015 contrasts this to a typical interpretation of the 'precautionary principle' often cited as a reason to avoid exploratory actions.
+@Schindler2015 also underscores the value of flexible policies; rather than "managing solely within the range of past variation; active probing is usually needed," and contrasts this to a typical interpretation of the 'precautionary principle' often cited as a reason to avoid exploratory actions.
 However, just because active exploration can escape the forecast trap does mean it is always a good idea.
@@
 
-Finally, it is worth noting that decisions do not need to be premised on a forecast at all, but can be premised entirely on the basis of past experience.
-If the stock size has gone up, increase harvest slightly, otherwise, decrease slightly.
+Finally, it is worth noting that decisions do not need to be premised on a forecast at all, but can be premised entirely on the basis of past experience:
+'If the fish stock size has gone up, increase harvest slightly, otherwise, decrease slightly.'
 Such a policy is not optimal, but it is robust across a wide range of unimodal stock-recruitment curves without ever estimating a predictive model.
@@
 Training such a artificial intelligence agents across a wide suite of simulations,
-a process known as curriculum learning [@], mimics the scenario analyses and search for .
-Such approaches have been used to train agents to play 2600 Atari console games at superhuman ability [@], outperform race-car drivers [@] and control nuclear fusion reactions [@].
+a process known as curriculum learning [@curriculum-learning], mimics the scenario analyses and search for .
+Such approaches have been used to train agents to play 2600 Atari console games at superhuman ability [@], outperform race-car drivers [@sony] and control nuclear fusion reactions [@deepmind-fusion].
 Such approaches are also not yet well understood, and can fail dramatically [@].

### 2022-02-28 manuscript/appendix-A.Rmd
:pen:

--- 2022-02-25 notebook/appendix-A.Rmd
@@
 ---
-title: 'Appendix A: Forecast Trap in a multi-species model'
+title: 'Appendix A: Forecast Trap in a Multi-Species Model'
 author: Carl Boettiger
@@
 Tmax <- 10
-resp <- 40
+reps <- 40
 

### 2022-02-28 manuscript/appendix-B.Rmd
:pen:

@@
 ---
-title: 'Appendix B: Forecast Trap in Sequential Decision-Making'
+title: 'Appendix B: The Forecast Trap in a Sequential Decision-Making Problem'
 author: Carl Boettiger
@@
 
-# Simulations and step-ahead forecasts (Fig 1, 2A)
+# Simulations and step-ahead forecasts (Fig 2, 3A)
 

### 2022-02-28 manuscript/round3-manuscript.Rmd
:pen:

@@
 The forecast trap is not unique to this example, but a fundamental consequence of non-uniqueness of models.
-Existing practices that promote the consideration of a broader set of models _a priori_ rather than rely on forecast skill alone are the best way to avoid the trap.
+Existing practices promoting a broader set of models are the best way to avoid the trap.
 
@@
 - data accessibility statement: _All simulation data generated for analyses here, along with code required for the analysis, is available in the Zenodo Data archive, https://doi.org/10.5281/zenodo.4660621_
-- number of words in the abstract: 149
-- number of words in main text: 4967
+- number of words in the abstract: 148
+- number of words in main text: 4900
 - number of cited references: 33
-- number of tables & figures: 4 Figures
+- number of tables & figures: 5 Figures
 
@@
 ```{r, include=FALSE}
-rmarkdown::render("appendix.Rmd")
+rmarkdown::render("appendix-A.Rmd")
+rmarkdown::render("appendix-B.Rmd")
 knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
@@
 I will use the term "model" to refer to any set of equations or code that can be used to produce a forecast.
-This term thus includes not only process-based models, but could also statistical forecasting methods (e.g.
-ARIMA, @arima), non-parametric approaches such as empirical dynamical modeling [@Ye2015], or machine learning.
+This term thus includes not only process-based models, but could also statistical forecasting methods, non-parametric approaches such as empirical dynamical modeling [@Ye2015], or machine learning. 
 Most such models must first be calibrated to historical data before they can produce a forecast, e.g. by parameter fitting, expert knowledge, or some other means. 
@@
 
+I shall focus on examples involving fisheries management to illustrate principles shared in many ecological systems.
+Fisheries are a significant economic and conservation concern worldwide and their management remains an important debate [e.g. @Worm2006; @Worm2009; @Costello2016].
+Moreover, their management has been a proving grounds for theoretical and practical decision-making issues which are widely applicable in other areas of ecology and conservation [@Ludwig1993; @Lande1994], and one that has long wrestled with issues of uncertainty in the context of management decisions [e.g.
+@Clark1973; @Reed1979; @Walters1981; @Ludwig1982].
+
 In these examples, we will focus on situations in which our 'data' comes from a model simulation rather than empirical sources.
@@
 conduct experiments that would be often impossible or unethical to perform in the real world: for instance, does a given fishery experience better long-term outcomes when managed according to forecasts derived from model 1 or from model 2?
-
-I shall focus on examples involving fisheries management to illustrate principles shared in many ecological systems.
-Fisheries are a significant economic and conservation concern worldwide and their management remains an important debate [e.g. @Worm2006; @Worm2009; @Costello2016].
-Moreover, their management has been a proving grounds for theoretical and practical decision-making issues which are widely applicable in other areas of ecology and conservation [@Ludwig1993; @Lande1994], and one that has long wrestled with issues of uncertainty in the context of management decisions [e.g.
-@Clark1973; @Reed1979; @Walters1981; @Ludwig1982].
 
@@
 Despite the clearly superior predictive accuracy of model 2 in both scenarios, the outcomes from management under model 2 are substantially worse.
-We can assess such outcomes in less abstract terms than forecasting skill, such as economic value them manager sought to optimize (in dollars) or the ecological value (unharvested biomass) [Fig 3].
+We can assess such outcomes in less abstract terms than forecasting skill, such as economic value them manager sought to optimize (in dollars) or the ecological value (unharvested biomass) [Fig 4].
 
@@
 
-```{r figure3, fig.cap="Adaptive management under model uncertainty.
+```{r figure4, fig.cap="Adaptive management under model uncertainty.
 Solid lines trace the trajectories of the state (fish stock, circles) and action (harvest quota, triangles), under adaptive management (learning).
@@
 
-<!-- Intuition, Figure 4 -->
+<!-- Intuition, Figure 5 -->
 
 The reason for model 1's seemingly contradictory ability to make good decisions but bad forecasts becomes obvious once we compare both curves to that of the underlying model, model 3.
-Looking at plots of the growth rate curves for each model [Fig 4A], it is hardly surprising that all model selection approaches prefer the closely overlapping curve of model 2 to the no-where-close curve of model 1 as the better approximation of model 3.
-Nevertheless, the decision policy derived from model 1 forecasts is indistinguishable from that based on the true model [Fig 4B], while the policy derived from model 2 forecasts lead to over-harvesting.
+Looking at plots of the growth rate curves for each model [Fig 5A], it is hardly surprising that all model selection approaches prefer the closely overlapping curve of model 2 to the no-where-close curve of model 1 as the better approximation of model 3.
+Nevertheless, the decision policy derived from model 1 forecasts is indistinguishable from that based on the true model [Fig 5B], while the policy derived from model 2 forecasts lead to over-harvesting.
 Being closest to the true model's forecast skill never guarantees that we are closest to the true model's optimal policy.
 
-```{r figure4, fig.cap="Panel A: Population growth curves of each model.
+```{r figure5, fig.cap="Panel A: Population growth curves of each model.
 The positive equilibrium of each model occurs where the curve crosses the horizontal axis.
@@
 Panel B: The optimal control policy under Model 1 is nearly identical to that under the true Model 3, while the optimal policy under Model 2 suppresses stock to a much lower escapement level."}
-# Fig 4
+# Fig 5
 model_curves <- read_csv("../data/model_curves.csv") %>% mutate(model = as.character(model))
@@
 Training such a artificial intelligence agents across a wide suite of simulations,
-a process known as curriculum learning [@curriculum-learning], mimics the scenario analyses and search for .
-Such approaches have been used to train agents to play 2600 Atari console games at superhuman ability [@], outperform race-car drivers [@sony] and control nuclear fusion reactions [@deepmind-fusion].
-Such approaches are also not yet well understood, and can fail dramatically [@].
+a process known as curriculum learning [@curriculum_learning], mimics the scenario analyses and search for robust policies.
+Such approaches have been used to train agents to play 2600 Atari console games at superhuman ability [@DQN], outperform race-car drivers [@sony] and control nuclear fusion reactions [@deepmind_fusion].
+Such approaches are also not yet well understood, and can fail dramatically [@Henderson2019; @Dulac-Arnold_2019].
 

### 2022-02-28 manuscript/appendix-B.Rmd

@@
 \begin{equation}
-f_i(Y_{t+1}) = Y_t + r_i Y_t \left( 1 - \frac{Y_t}{K_i} \right) * \xi_t(\sigma)
+X_{t+1} = f_i(Y_t) = Y_t + r_i Y_t \left( 1 - \frac{Y_t}{K_i} \right) * \xi_t(\sigma)
 \end{equation}
 
-where $\xi_i(t)$ is a multiplicative log-normal noise term with mean 1 and log-standard-deviation $\sigma_i$.
+where harvest occurs before recruitment, $Y_t = \min(X_t - H_t, 0)$,  $\xi_i(t)$ is a multiplicative log-normal noise term with mean 1 and log-standard-deviation $\sigma_i$.
 
@@
 
-for each prediction over 100 replicate simulations of 100 time steps each [Fig 1].
+for each prediction over 100 replicate simulations of 100 time steps each [Fig 2].
 
@@
 
-````{r}
+```{r}
 ## Compute reward matrix (shared across all models)
@@
 
-# Simulations and step-ahead forecasts (Fig 2, 3A)
+# Simulations and step-ahead forecasts (Fig 3, 4A)
 
@@
 occurs (note that for each model, we use the transition matrix that corresponds
-to 'no fishing', `model[[state_index, ,1]]`) (Fig 1a, main text).
+to 'no fishing', `model[[state_index, ,1]]`) (Fig 2A, main text).
 
@@
 
-# Proper scores (Fig 1)
+# Proper scores (Fig 2C-D)
 
@@
 # Gneiting & Raferty (2007), eq27
-scoring_fn <- function(x, mu, sigma){ as.numeric( -(mu - x )^2 / sigma^2  - log(sigma) )}
+scoring_fn <- function(x, mu, sigma){
+  as.numeric( -(mu - x )^2 / sigma^2  - log(sigma) )
+  }
 
@@
 
-# Economic value tabulation (Fig 2B)
+# Economic value tabulation (Fig 3B)
 
@@
 ```{r}
-adaptive_management <- memoise::memoise(mdp_learning, cache = memoise::cache_filesystem("cache/"))
+adaptive_management <- memoise::memoise(mdp_learning, 
+                                        cache = memoise::cache_filesystem("cache/"))
 am_many <- adaptive_management(many_transitions,  
@@
 
-Again we record the posterior belief in the 'model 1' parameter values, and stack the data frames for learning and planning-only cases into a single file used to generate figure 3B.
+Again we record the posterior belief in the 'model 1' parameter values, and stack the data frames for learning and planning-only cases into a single file used to generate figure 4B.
 
@@
 
-For most of the simulation, the most probable $r$ value is around 0.3, and the most probable $K$ value around 1.5.
+For most of the simulation, the most probable $r$ value is around 0.3, and the most probable $K$ value around 10.
 
@@
 
-To plot growth curves of individual models (Fig 4a), we evaluate
+To plot growth curves of individual models (Fig 5A), we evaluate
 
@@
 
-The resulting model curves are shown in the main paper, Fig 4A.
+The resulting model curves are shown in the main paper, Fig 5A.
 

### 2022-02-28 manuscript/round3-manuscript.Rmd

@@
 - number of words in main text: 4900
-- number of cited references: 33
+- number of cited references: 44
 - number of tables & figures: 5 Figures
@@
 
-```{r, include=FALSE}
+```{r renderA, include=FALSE}
 rmarkdown::render("appendix-A.Rmd")
+
+```
+
+```{r renderB, include=FALSE, eval=FALSE}
 rmarkdown::render("appendix-B.Rmd")
+```
+
+```{r knit_global, include=FALSE}
+library(tidyverse)
 knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
@@
 When VOI is negative, the decision-maker may become trapped into accepting mediocre outcomes derived from a model that makes accurate forecasts, even when a less accurate model that would generate better outcomes is available.
-I will present two examples of this "forecast trap" and examine how it arises.
+I will present two examples of this "forecast trap" and examine how it arises as a result of non-uniqueness of models [@Oreskes1994; @Schindler2015] with respect to either of these objectives.
 
@@
 To better understand how a model can produce a more accurate forecast and yet still lead to a worse decision, it may be helpful to start with a concrete example.
-Figure \ref{fig1} shows both the forecasts and realized management outcomes of two alternative three-species models, "A" and "B" (see Appendix A) in predicting the population dynamics of striped bass (an economically important fishery) and double-crested cormorants (a target species for conservation) which both feed on a population of river herring (whose abundance we assume is not measured).
+Fig 1 shows both the forecasts and realized management outcomes of two alternative three-species models, "A" and "B" (see Appendix A) in predicting the population dynamics of striped bass (an economically important fishery) and double-crested cormorants (a target species for conservation) which both feed on a population of river herring (whose abundance we assume is not measured).
 Model A accurately forecasts the abundance of both bass and cormorants well into the future, but the optimal management strategy derived from its forecasts leads to steadily declining species abundances and overall disappointing net utility.
@@
 in which our manager seeks to balance multiple objectives of sustaining the cormorant population while maximizing the economic value from harvesting both herring and striped bass.
-In the scenario depicted in Figure \ref{fig1}, I have used a the richer five-species model introduced @Brias2021 to drive the underlying dynamics, which includes three competing species of herring that are preyed upon by both the bass and the cormorants.
+In the scenarios depicted in Fig 1, I have used a the richer five-species model introduced @Brias2021 to drive the underlying dynamics, which includes three competing species of herring that are preyed upon by both the bass and the cormorants.
 Here, the manager seeks to maximize utility given by the weighted sum of the individual objectives [@Brias2021], placing 50% of the weight on the conservation objective and splits the remaining weight evenly over the harvests for predator (bass) and prey (herring) species.
@@
 Panel B: given 42 candidate models over a broad range of parameter values, adaptive management quickly reduces the probability of model 1, and substantially underperforms management without learning (dotted lines).
-While outcomes improve marginally relative to the two-model case (figure A) they remain significantly worse than had no iterative learning been included."}
+While outcomes improve marginally relative to the two-model case (panel A) they remain significantly worse than had no iterative learning been included."}
 am <- read_csv("../data/am.csv")
@@
 
-Third, managers may emphasize policy robustness over expected net utility. 
+Third, managers may emphasize policy robustness over forecast skill [@Schindler2015]. 
 In many formal treatments, this is not qualitatively different to the analysis considered here: 
@@
 But in practice, robust design may emphasize acceptable performance across the widest possible array of scenarios (e.g. candidate models).
-This acts more like a sensitivity analysis of utility with respect to underlying assumptions, rather than an optimization routine.
+This acts more like a sensitivity analysis of utility with respect to underlying assumptions, rather than an optimization routine [e.g. @Punt2016; @Fischer2009].
 Computationally, the former is much simpler, allowing researchers to evaluate the performance of a policy on more complex simulations for which calculating the optimal policy would be prohibitively difficult.

### 2022-02-28 manuscript/round3-manuscript.Rmd
:pen:

@@
 
-**A note on models and data**
+## A note on models and data
 
@@
 
-# Iterative updating: _Sequential Decision Processes_
+# An Iterative Decision Example
 
@@
 
-## Iterative upates and adaptive management
+## Adaptive Management
 
@@
 I later consider a larger suite of 42 models, spanning the parameter space of Gordon-Schaefer curves.
-To avoid failure to explore sufficiently, [see exploration-exploitation tradeoff, e.g. @Walters1981],
+To avoid failure to explore sufficiently, [see exploration-exploitation trade-off, e.g. @Walters1981],
 I assign prior belief of 99% weight on the optimally performing model, model 1.  
@@
 does not _learn_ over time).
-The difference between the performance with and without learning is known as the "Value of Information".
+The difference between the performance with and without learning is known as the "Value of Information" [@Howard1966].
 In both 2-model and 42-model scenarios, the value of information is strongly negative.
@@
 out <- am %>% group_by(method) %>% summarize(total = sum(quota*.99^time))
-( out[out$method=="learning","total"][[1]] -  out[out$method=="planning","total"][[1]]) / out[out$method=="planning","total"][[1]]
+VOI_2models <- ( out[out$method=="learning","total"][[1]] -  out[out$method=="planning","total"][[1]]) / out[out$method=="planning","total"][[1]]
 
 out <- am_multi %>% group_by(method) %>% summarize(total = sum(quota*.99^time))
-( out[out$method=="learning","total"][[1]] -  out[out$method=="planning","total"][[1]]) / out[out$method=="planning","total"][[1]]
+VOI_42models <- ( out[out$method=="learning","total"][[1]] -  out[out$method=="planning","total"][[1]]) / out[out$method=="planning","total"][[1]]
 
@@
 
-A manager may also escape the forecast trap by exploring actions which are never recommended from any of the available forecasts.
+Second, A manager may also escape the forecast trap by exploring actions which are never recommended from any of the available forecasts.
 For example, a manager looking at the low stocks and poor harvests achieved under model 2, could decide experimentally to reduce fishing quota. 
@@
 Such approaches have been used to train agents to play 2600 Atari console games at superhuman ability [@DQN], outperform race-car drivers [@sony] and control nuclear fusion reactions [@deepmind_fusion].
-Such approaches are also not yet well understood, and can fail dramatically [@Henderson2019; @Dulac-Arnold_2019].
+Such approaches are also not yet well understood, introducing new risks as well as new possibilities[@Henderson2019; @Dulac-Arnold_2019].
 

### 2022-02-28 manuscript/appendix-B.Rmd

@@
 These issues are discussed in detail in @pomdp-intro.
-For our purposes, it suffices to note that constant escapement policy is relevant both from a theoretical standpoint as the optimal solution for the problem under consideration, and an applied standpoint as a policy that guides management of many salmon fisheries.The optimal control approach using SDP as illustrated here is used in a wide and growing number of conservation problems.
+For our purposes, it suffices to note that constant escapement policy is relevant both from a theoretical standpoint as the optimal solution for the problem under consideration, and an applied standpoint as a policy that guides management of many salmon fisheries.
+The optimal control approach using SDP as illustrated here is used in a wide and growing number of conservation problems.
 
@@
 
-Future utility is discounted by fixed factor of $\delta = 0.99$. Classic maximum sustainable 
-yield models [@Schaefer1954] ignore discounting, while modern economic optimization
-models insist on it, so a small discount conforms to the latter while reasonably
-approximating the former.  The qualitative conclusion is not sensitive to the 
-discounting rate.
+Future utility is discounted by fixed factor of $\delta = 0.99$.
+Classic maximum sustainable yield models [@Schaefer1954] ignore discounting, while modern economic optimization models insist on it, so a small discount conforms to the latter while reasonably approximating the former.
+The qualitative conclusion is not sensitive to the discounting rate.
 
@@
 to 'no fishing', `model[[state_index, ,1]]`) (Fig 2A, main text).
-
 The transition matrices give the full (discretized) probability distribution,

### 2022-04-09 manuscript/round4-manuscript.Rmd
here we go

--- 2022-02-28 manuscript/round3-manuscript.Rmd
@@
 Fisheries are a significant economic and conservation concern worldwide and their management remains an important debate [e.g. @Worm2006; @Worm2009; @Costello2016].
-Moreover, their management has been a proving grounds for theoretical and practical decision-making issues which are widely applicable in other areas of ecology and conservation [@Ludwig1993; @Lande1994], and one that has long wrestled with issues of uncertainty in the context of management decisions [e.g.
-@Clark1973; @Reed1979; @Walters1981; @Ludwig1982].
+Moreover, their management has been a proving grounds for theoretical and practical decision-making issues [e.g.
+@Clark1973; @Reed1979; @Walters1981; @Ludwig1982] arising in a wide range of other contexts, including
+invasive species [@Boettiger2021], infectious diseases [@Li2017; Shea2014], fire management [@Richards1999] conservation planning and priortization [@Chades2008; @Wilson2006] climate policy [@] and much else [@Ludwig1993; @Lande1994; @Polasky2011].
 
@@
 The author acknowledges support from NSF CAREER Award #1942280 and helpful discussions with Melissa Chapman.
-The author is also deeply grateful to the Michael Runge and three other anonymous reviewers whose detailed feedback and insights have greatly shaped the paper and my own understanding of these issues.
+The author is also deeply grateful to the Michael Runge, Chih-hao Hsieh, Antoine Brias, and other anonymous reviewers whose detailed feedback and insights have greatly shaped the paper and my own understanding of these issues.
 

### 2022-04-12 manuscript/appendix-A.Rmd
:pen:
here we go

@@
 The underlying dynamics for the simulated "true" system are based on the five-dimensional model of a herring fishery presented in @Brias2021.
+I have used the model parameters from @Brias2021, which were selected to ensure co-existence of the five species as outlined in their paper.
 Following @Brias2021, we assume three species of herring, $X_1$, $X_2$, $X_3$ are preyed upon by
@@
 \begin{align}
-x_{1,t+1} &= s_{1,t} \exp{(1.0213 -        s_{1,t} - 0.0861 s_{2,t} - 0.3141 s_{3,t} - 0.7252 s_{4,t} -0.2445 x_{5,t} +\espilon_{1,t} )}
-x_{2,t+1} &= s_{2,t} \exp{(1.0289 - 0.4765 s_{1,t} -        s_{2,t} - 0.1370 s_{3,t} - 0.9811 s_{4,t} -0.0915 x_{5,t} +\espilon_{2,t} )}
-x_{3,t+1} &= s_{3,t} \exp{(1.0207 - 0.3193 s_{1,t} - 0.3461 s_{2,t} -        s_{3,t} - 0.6367 s_{4,t} -0.8716 x_{5,t} +\espilon_{3,t} )}
-x_{4,t+1} &= s_{4,t} \exp{(         0.7252 s_{1,t} + 0.9811 s_{2,t} + 0.6367 s_{3,t} -        s_{4,t}                 +\espilon_{4,t})}
-x_{5,t+1} &= x_{5,t} \exp{(         0.2445 s_{1,t} + 0.0915 s {2,t} + 0.8716 s {3,t} -                        x_{5,t} +\espilon_{5,t} )}
+x_{1,t+1} &= s_{1,t} \exp{(1.0213 -        s_{1,t} - 0.0861 s_{2,t} - 0.3141 s_{3,t} - 0.7252 s_{4,t} -0.2445 x_{5,t} +\epsilon_{1,t} )} \\
+x_{2,t+1} &= s_{2,t} \exp{(1.0289 - 0.4765 s_{1,t} -        s_{2,t} - 0.1370 s_{3,t} - 0.9811 s_{4,t} -0.0915 x_{5,t} +\epsilon_{2,t} )} \\
+x_{3,t+1} &= s_{3,t} \exp{(1.0207 - 0.3193 s_{1,t} - 0.3461 s_{2,t} -        s_{3,t} - 0.6367 s_{4,t} -0.8716 x_{5,t} +\epsilon_{3,t} )} \\
+x_{4,t+1} &= s_{4,t} \exp{(         0.7252 s_{1,t} + 0.9811 s_{2,t} + 0.6367 s_{3,t} -        s_{4,t}                 +\epsilon_{4,t})} \\
+x_{5,t+1} &= x_{5,t} \exp{(         0.2445 s_{1,t} + 0.0915 s {2,t} + 0.8716 s {3,t} -                        x_{5,t} +\epsilon_{5,t} )}
 \end{align}
@@
 
-\begin{equation}
-s_{i,t} = x_{i_t} \left(1 - u_{i,t}\right)
-\end{equation}
+$$
+s_{i,t} = x_{i_t} \left( 1 - u_{i,t} \right)
+$$
 
-\begin{equation}
-\epsilon_{i,t} \stackrel{\texrm{iid}}{\tilde} \mathcal{N(0, \sigma_i)}
-\end{equation}
+$$
+\epsilon_{i,t} \stackrel{iid}{\sim} \mathcal{N}(0, \sigma_i)
+$$
 
@@
 This model treats the three species of herring collectively as $x_1$. 
-Parameters are estimated consistent with limited historical data of the observed variables, bass biomass $x_4$ and herring biomass $x_5$.
+As with the original @Brias2021 model, parameters are not based directly on historical fisheries data, but have been selected to be consistent with simulations of the observed variables from the "true" model of @Brias2021, bass biomass $x_4$ and herring biomass $x_5$.
 
-\begin{equation}
-x_{1,t+1} &= s_{1,t} \exp(3 - 8 s_{1,t} - 0.9 s_4 - 0.9 * x_{5,t} + \espilon_{1,t} )
-x_{4,t+1} &= s_{4,t} \exp(0.1 s_{1,t} - s_{4,t} + \espilon_{4,t}  )
-x_{5,t+1} &= x_{5,t} \exp(0.3 s_{1,t} - x_5  + \espilon_{5,t}  )
-\end{equation}
+\begin{align}
+x_{1,t+1} &= s_{1,t} \exp(3 - 8 s_{1,t} - 0.9 s_4 - 0.9 * x_{5,t} + \epsilon_{1,t} ) \\
+x_{4,t+1} &= s_{4,t} \exp(0.1 s_{1,t} - s_{4,t} + \epsilon_{4,t}  ) \\
+x_{5,t+1} &= x_{5,t} \exp(0.3 s_{1,t} - x_5  + \epsilon_{5,t}  )
+\end{align}
 
@@
 
-\begin{equation}
-x_{1,t+1} &= s_{1,t} \exp(1.60 - 2.00 s_{1,t}  + \espilon_{1,t} )
-x_{4,t+1} &= s_{4,t} \exp(1.45 - 2.75 s_{1,t}  + \espilon_{4,t}  )
+\begin{align}
+x_{1,t+1} &= s_{1,t} \exp(0.1 - 2.00 s_{1,t}  + \epsilon_{1,t} ) \\
+x_{4,t+1} &= s_{4,t} \exp(1.45 - 2.75 s_{1,t}  + \epsilon_{4,t}  ) \\
 x_{5,t+1} &= 0.25 x_{5,t} 
-\end{equation}
+\end{align}
 
@@
 
-knitr::opts_chunk$set(cache=FALSE, tidy = "styler", message = FALSE, warning = FALSE, echo = TRUE)
+knitr::opts_chunk$set(cache=TRUE, tidy = "styler", message = FALSE, warning = FALSE, echo = TRUE)
 
+set.seed(123)
 ```
@@
 
-          x1[t+1] = s1[t]*exp(1.0213 - s1[t] - 0.0861*s2[t] - 0.3141*s3[t] - 0.7252*s4[t] - 0.2445*x5[t] + xi1[t])
-          x2[t+1] = s2[t]*exp(1.0289 - 0.4765*s1[t] - s2[t] - 0.1370*s3[t] - 0.9811*s4[t] - 0.0915*x5[t] + xi2[t])
-          x3[t+1] = s3[t]*exp(1.0207 - 0.3193*s1[t] - 0.3461*s2[t] - s3[t] - 0.6367*s4[t] - 0.8716*x5[t] + xi3[t])
-          x4[t+1] = s4[t]*exp(0.7252*s1[t] + 0.9811*s2[t] + 0.6367*s3[t] - s4[t] + xi4[t])
-          x5[t+1] = x5[t]*exp(0.2445*s1[t] + 0.0915*s2[t] + 0.8716*s3[t] - x5[t] +  xi5[t])
+          x1[t+1] = s1[t]*exp(1.0213 - s1[t] - 0.0861*s2[t] -
+                                0.3141*s3[t] - 0.7252*s4[t] -
+                                0.2445*x5[t] + xi1[t])
+          x2[t+1] = s2[t]*exp(1.0289 - 0.4765*s1[t] - s2[t] -
+                                0.1370*s3[t] - 0.9811*s4[t] -
+                                0.0915*x5[t] + xi2[t])
+          x3[t+1] = s3[t]*exp(1.0207 - 0.3193*s1[t] - 
+                                0.3461*s2[t] - s3[t] -
+                                0.6367*s4[t] - 0.8716*x5[t] + xi3[t])
+          x4[t+1] = s4[t]*exp(0.7252*s1[t] + 0.9811*s2[t] +
+                                0.6367*s3[t] - s4[t] + xi4[t])
+          x5[t+1] = x5[t]*exp(0.2445*s1[t] + 0.0915*s2[t] + 
+                                0.8716*s3[t] - x5[t] +  xi5[t])
 
@@
                       A1 = 2,
-                      R1 = 1.6,
+                      R1 = 0.1,
                       A2 = 2.75,
@@
     for (t in 1:(Tmax-1)) {
-        s1[t] <- x1[t] * (1 - 10*u_prey)
+        s1[t] <- x1[t] * (1 - u_prey)
         s4[t] <- x4[t] * (1 - u_predator)
@@
 u_A <- function(x) utility(x, f = model_A)
-A <- optim(c(.2,0.01), u_A)
+A <- optim(c(.8,0.02), u_A)
 model_A_policy <- clip(A$par, 0, 1)
@@
 u_B <- function(x) utility(x, f = model_B)
-B <- optim(c(0.5,0.1), u_B)
+B <- optim(c(0.01,0.1), u_B)
 model_B_policy <- clip(B$par, 0, 1)
@@
 ```{r}
-mean_utility_A / mean_utility_B
+ratio <- mean_utility_A / mean_utility_B
 ```
+
+The ratio of mean utility under model A to mean utility under the nearly-optimal model B is `r round(ratio *100)`%.
+
+# References

### 2022-04-12 manuscript/appendix-B.Rmd
here we go

@@
 
-Like many decision problems, the task of setting a sustainable harvest quota appears to hinge on having an accurate forecast: if we can predict to what size the fish stock will increase next year, $X_t+1$, and we know the current stock, $X_t$, then we can sustainably harvest $X_{t+1} - X_t$ without decreasing the biomass over the long term.
+Like many decision problems, the task of setting a sustainable harvest quota appears to hinge on having an accurate forecast: if we can predict to what size the fish stock will increase next year, $X_{t+1}$, and we know the current stock, $X_t$, then we can sustainably harvest $X_{t+1} - X_t$ without decreasing the biomass over the long term.
 Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of overfitting by comparing model predictions to later observations that were not used to estimate the model [@Gneiting2014].
@@
 This determines the optimal harvest policy for each possible state, given each model.
-Because this step is the most computationally intensive routine, we cache the results using memosization conditioned on the transition matrices [@memoise].
+Because this step is the most computationally intensive routine, we cache the results using memoization conditioned on the transition matrices [@memoise].
 Running this code with alternate transition matrices automatically invalidates that cache, reducing the risk of loading spurious results.
@@
 
-am_npv <- sum(am$value * discount ^ am$time)
-am_economics_percent <- round(am_npv / npv[[1,"npv"]] * 100)
+am_npv <- am %>% group_by(method) %>% summarize(npv = sum(quota * discount ^ time))
+learning <- am_npv %>% filter(method=="learning") %>% pull(npv) 
+am_economics_percent <- round(learning / npv[[1,"npv"]] * 100)
 
-mean_state <- sims %>% group_by(model) %>% summarize(state = mean(state))
-am_ecology_percent <- round(mean(states[am$state])/mean_state[[1, "state"]]*100)
+mean_state <- sims %>%
+  group_by(model) %>% 
+  summarize(state = mean(state)) %>% 
+  filter(model == 1) %>%
+  pull(state)
+learn_state <- am %>%
+  group_by(method) %>% 
+  summarize(state = mean(stock))  %>%
+  filter(method == "learning") %>%
+  pull(state)
+
+am_ecology_percent <- round(learn_state / mean_state *100)
 ```

### 2022-04-12 manuscript/round4-manuscript.Rmd

@@
 - number of words in the abstract: 148
-- number of words in main text: 4900
-- number of cited references: 44
+- number of words in main text: 4950
+- number of cited references: 51
 - number of tables & figures: 5 Figures
@@
 When VOI is negative, the decision-maker may become trapped into accepting mediocre outcomes derived from a model that makes accurate forecasts, even when a less accurate model that would generate better outcomes is available.
+This trap is invisible to the manager unless sufficient alternative models outside the original set are introduced.
 I will present two examples of this "forecast trap" and examine how it arises as a result of non-uniqueness of models [@Oreskes1994; @Schindler2015] with respect to either of these objectives.
@@
 Fisheries are a significant economic and conservation concern worldwide and their management remains an important debate [e.g. @Worm2006; @Worm2009; @Costello2016].
-Moreover, their management has been a proving grounds for theoretical and practical decision-making issues [e.g.
+Moreover, fisheries management has been a proving grounds for theoretical and practical decision-making issues [e.g.
 @Clark1973; @Reed1979; @Walters1981; @Ludwig1982] arising in a wide range of other contexts, including
-invasive species [@Boettiger2021], infectious diseases [@Li2017; Shea2014], fire management [@Richards1999] conservation planning and priortization [@Chades2008; @Wilson2006] climate policy [@] and much else [@Ludwig1993; @Lande1994; @Polasky2011].
+invasive species [@Boettiger2021], infectious diseases [@Li2017; Shea2014], fire management [@Richards1999] conservation planning and prioritization [@Chades2008; @Wilson2006] climate policy [@Nitzbon2017] and much else [@Ludwig1993; @Lande1994; @Polasky2011].
 
@@
 
-To better understand how a model can produce a more accurate forecast and yet still lead to a worse decision, it may be helpful to start with a concrete example.
+To better understand how a model can produce a more accurate forecast and yet still lead to a worse decision, it may be helpful to start with a concrete example in which a manager faces a trade-off between cormorant conservation and fish harvest.
 Fig 1 shows both the forecasts and realized management outcomes of two alternative three-species models, "A" and "B" (see Appendix A) in predicting the population dynamics of striped bass (an economically important fishery) and double-crested cormorants (a target species for conservation) which both feed on a population of river herring (whose abundance we assume is not measured).
@@
 In the second scenario (lower panel), the manager uses the optimal SDP policy derived from each forecast to determine the TAC for the following year, and compares the observed stock size to that which the model predicted given that fishing quota.
-In both cases, model 2 provides far more accurate forecasts, as seen in the error bars and confirmed by the distribution of proper scores [Fig 3C-D; @Gneiting2007].
+In both cases, model 2 provides far more accurate forecasts, as seen in the error bars and confirmed by the distribution of proper scores [Fig 2C-D; @Gneiting2007].
 Model and simulation details are provided in Appendix B.
@@
 Despite the clearly superior predictive accuracy of model 2 in both scenarios, the outcomes from management under model 2 are substantially worse.
-We can assess such outcomes in less abstract terms than forecasting skill, such as economic value them manager sought to optimize (in dollars) or the ecological value (unharvested biomass) [Fig 4].
+We can assess such outcomes in less abstract terms than forecasting skill, such as economic value them manager sought to optimize (in dollars) or the ecological value (unharvested biomass) [Fig 3].
 
@@
 A manager operating under model 2 would have little indication that the model was flawed: both future stock sizes and expected harvest yields consistently match model predictions.
-This manager would be stuck in the forecast trap, incorrectly mistaking the a degraded ecological state and reduced economic outcomes as the best that can be achieved because future observations continue to validate this model.
+This manager would be stuck in the forecast trap, incorrectly mistaking a degraded ecological state and reduced economic outcomes as the best that can be achieved because future observations continue to validate this model.
 Had we been able to include Model 3 in our forecast comparisons, it would equal or outperform the forecasting skill of both model 1 and model 2 (as guaranteed by the theorem of @Gneiting2007), while also matching the economic utility of model 1 (as guaranteed by the theorem of @Reed1979).
@@
 This reflects the view that "the primary values of ecosystem models are as heuristic tools for communication and for developing scenarios to express uncertainties and test policies" rather than as a source of reliable forecasts [@Schindler2015].
-The practices include:
+Such practices include:
 
@@
 
-Second, A manager may also escape the forecast trap by exploring actions which are never recommended from any of the available forecasts.
+Second, a manager may also escape the forecast trap by exploring actions which are never recommended from any of the available forecasts.
 For example, a manager looking at the low stocks and poor harvests achieved under model 2, could decide experimentally to reduce fishing quota. 
@@
 @Schindler2015 also underscores the value of flexible policies; rather than "managing solely within the range of past variation; active probing is usually needed," and contrasts this to a typical interpretation of the 'precautionary principle' often cited as a reason to avoid exploratory actions.
-However, just because active exploration can escape the forecast trap does mean it is always a good idea.
+However, just because active exploration can escape the forecast trap does not mean it is always a good idea.
 
@@
 Such approaches are just as vulnerable to bad outcomes (as defined by their own utility functions) whenever models are selected only on the basis of forecast skill.
-But in practice, robust design may emphasize acceptable performance across the widest possible array of scenarios (e.g. candidate models).
+Alternative approaches may not seek any such optimization, emphasize the viability [@Aubin1991] of possible policy under constraints.
+In practice, robust design may emphasize acceptable performance across the widest possible array of scenarios (e.g. candidate models).
 This acts more like a sensitivity analysis of utility with respect to underlying assumptions, rather than an optimization routine [e.g. @Punt2016; @Fischer2009].

### 2022-04-12 manuscript/round4-manuscript.Rmd
no renv, old rtciles for correct csl use

@@
 date: "`r Sys.Date()`"
+csl: csl/ecology-letters.csl
 
@@
 
-csl: csl/ecology-letters.csl
 output: rticles::elsevier_article
@@
 @Clark1973; @Reed1979; @Walters1981; @Ludwig1982] arising in a wide range of other contexts, including
-invasive species [@Boettiger2021], infectious diseases [@Li2017; Shea2014], fire management [@Richards1999] conservation planning and prioritization [@Chades2008; @Wilson2006] climate policy [@Nitzbon2017] and much else [@Ludwig1993; @Lande1994; @Polasky2011].
+invasive species [@Boettiger2021], infectious diseases [@Li2017; @Shea2014], fire management [@Richards1999] conservation planning and prioritization [@Chades2008; @Wilson2006] climate policy [@Nitzbon2017] and much else [@Ludwig1993; @Lande1994; @Polasky2011].
 

### 2022-04-26 manuscript/manuscript.tex
cache

\documentclass[3p]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{Ecology Letters} % Sets Journal name

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={The Forecast Trap},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% Pandoc header
\linenumbers
\usepackage{endfloat}

\begin{document}
\begin{frontmatter}

  \title{The Forecast Trap}
      \address[a]{Department of Environmental Science, Policy, and
Management, University of California, 130 Mulford Hall Berkeley, CA
94720-3114, USA}

  \begin{abstract}
  Encouraged by decision makers' appetite for future information on
  topics ranging from elections to pandemics, and enabled by the
  explosion of data and computational methods, model based forecasts
  have garnered increasing influence on a breadth of decisions in modern
  society. Using several classic examples from fisheries management, I
  demonstrate that selecting the model or models that produce the most
  accurate and precise forecast (measured by statistical scores) can
  sometimes lead to worse outcomes (measured by real-world objectives).
  This can create a forecast trap, in which the outcomes such as fish
  biomass or economic yield decline while the manager becomes
  increasingly convinced that these actions are consistent with the best
  models and data available. The forecast trap is not unique to this
  example, but a fundamental consequence of non-uniqueness of models.
  Existing practices promoting a broader set of models are the best way
  to avoid the trap.
  \end{abstract}
   \begin{keyword} forecasting, adaptive
management, stochasticity, uncertainty, optimal control\end{keyword}
 \end{frontmatter}

\begin{itemize}
\tightlist
\item
  abbreviated running title: ``The Forecast Trap''
\item
  Authorship statement: \emph{This is a single-author paper}
\item
  data accessibility statement: \emph{All simulation data generated for
  analyses here, along with code required for the analysis, is available
  in the Zenodo Data archive, https://doi.org/10.5281/zenodo.4660621}
\item
  number of words in the abstract: 148
\item
  number of words in main text: 4950
\item
  number of cited references: 51
\item
  number of tables \& figures: 5 Figures
\end{itemize}

\pagebreak

Global change issues are complex and outcomes are difficult to predict
(Clark \emph{et al.} 2001). To guide decisions in an uncertain world,
researchers and decision makers may consider a range of alternative
plausible models to better reflect what we do and do not know about the
processes involved (Polasky \emph{et al.} 2011). Forecasts or
predictions from possible models can indicate what outcomes are most
likely to result under what decisions or actions. This has made
model-based forecasts a cornerstone for scientifically based decision
making. By comparing outcomes predicted by a model to future
observations, a decision maker can not only \emph{plan} for the
uncertainty, but also \emph{learn} which models are most trustworthy.
The value of iterative learning has long been reflected in the theory of
adaptive management (Walters \& Hilborn 1978) as well as in actual
adaptive management practices such as Management Strategy Evaluation
(MSE) (Punt \emph{et al.} 2016) used in fisheries, and is a central
tenet of a rapidly growing interest in ecological forecasting (Dietze
\emph{et al.} 2018). But, do iterative learning approaches always lead
to better decisions?

In this paper, I demonstrate that the model that makes the better
prediction (defined as a strictly proper score, Gneiting \& Raftery
(2007)) is not necessarily the model that makes the better policy
(defined in terms of utility, e.g.~expected net present value, Clark
(1990)). I show that our best methods for learning about model structure
or parameters by repeatedly comparing forecasts to observations can be
counter-productive. Put another way, the value of information (VOI, as
measured by the expected utility given that information minus the
utility without it; see Howard (1966); Katz \emph{et al.} (1987)), can
actually be negative. When VOI is negative, the decision-maker may
become trapped into accepting mediocre outcomes derived from a model
that makes accurate forecasts, even when a less accurate model that
would generate better outcomes is available. This trap is invisible to
the manager unless sufficient alternative models outside the original
set are introduced. I will present two examples of this ``forecast
trap'' and examine how it arises as a result of non-uniqueness of models
(Oreskes \emph{et al.} 1994; Schindler \& Hilborn 2015) with respect to
either of these objectives.

The forecast trap is not the only mechanism by which some model-choice
methods lead to worse outcomes. Previous work has long acknowledged the
panoply of ways in which model-based decision making can go astray due
to conflicting incentives, implementation errors, or lack of resources
for monitoring and updating (e.g. Ludwig \emph{et al.} 1993). Another
widely recognized problem is that of over-fitting (Burnham \& Anderson
1998), in which the model that best fits historical data fails to best
predict future data (Ginzburg \& Jensen 2004). Under such circumstances,
it is easy to see how an over-fit model would also lead to bad outcomes.
However, over-fitting plays no role in the forecast trap, where model
predictions are assessed only using probabilistic forecasts, and not
observations which had previously been used to fit the models. Formally,
these scores satisfy the `proper scoring' rule of Gneiting \& Raftery
(2007), which proves no other probabilistic prediction \(Q(x)\) will
have a better expected score than that of the true model
(i.e.~generative process), \(P(x)\). Gneiting \& Raftery (2007)'s proof
of proper scoring has since become a critical tool to avoid over-fitting
when choosing models to make decisions, but as I illustrate, will not
prevent the forecast trap.

First, I will introduce a motivating example in which we will consider
two reasonable process-based models, A and B. Model A will produce very
accurate forecasts, but lead to much worse outcomes than Model B. Though
I will establish that these accurate forecasts in Model A are not the
result of chance or of over-fitting the data, this example may raise
more questions than answers. To get a better understanding of when the
forecast trap arises, I will turn to a simpler ecological model, to
which we may apply more sophisticated decision tools of iterative
forecasting and adaptive management. We will see that these approaches
do not avoid the forecast trap either. No collection of such examples
can establish precisely how common the forecast trap may be in
real-world applications. The examples do establish unequivocally that
achieving incrementally ever-more-accurate forecasts does \emph{not
guarantee} better decisions. I conclude by pointing to a range of
established and emerging approaches to quantitative decision-making
which are not based on forecasts. As sophisticated forecasting
techniques become more common-place in conservation and ecology, the
forecast trap is a reminder that we should not forget about these
alternatives.

\hypertarget{a-note-on-models-and-data}{%
\subsection{A note on models and data}\label{a-note-on-models-and-data}}

I will use the term ``model'' to refer to any set of equations or code
that can be used to produce a forecast. This term thus includes not only
process-based models, but could also statistical forecasting methods,
non-parametric approaches such as empirical dynamical modeling (Ye
\emph{et al.} 2015), or machine learning. Most such models must first be
calibrated to historical data before they can produce a forecast,
e.g.~by parameter fitting, expert knowledge, or some other means.
Different choices for those parameters create different forecasts, I
will refer to those different parameterizations as different models. It
is of course possible for a decision-maker to consider forecasts coming
from multiple structurally different models simultaneously, and
potentially assigning different weights to each model. As more data
becomes available, it is possible to update model parameters, or
equivalently, update the weights assigned across models. I will examine
such approaches for model ensembles and model updating further on.

I shall focus on examples involving fisheries management to illustrate
principles shared in many ecological systems. Fisheries are a
significant economic and conservation concern worldwide and their
management remains an important debate (e.g. Worm \emph{et al.} 2006,
2009; Costello \emph{et al.} 2016). Moreover, fisheries management has
been a proving grounds for theoretical and practical decision-making
issues (e.g. Clark 1973; Reed 1979; Walters 1981; Ludwig \& Walters
1982) arising in a wide range of other contexts, including invasive
species (Boettiger 2021), infectious diseases (Shea \emph{et al.} 2014;
Li \emph{et al.} 2017), fire management (Richards \emph{et al.} 1999)
conservation planning and prioritization (Wilson \emph{et al.} 2006;
Chadès \emph{et al.} 2008) climate policy (Nitzbon \emph{et al.} 2017)
and much else (Ludwig \emph{et al.} 1993; Lande \emph{et al.} 1994;
Polasky \emph{et al.} 2011).

In these examples, we will focus on situations in which our `data' comes
from a model simulation rather than empirical sources. Simulations are
simplifications of the real world -- just because a method works in a
simulation is no guarantee that it works in reality. Conversely, if
decision methods are not reliable even when applied to simulated cases,
we should be even more cautious in how we use them. Simulations also
allow us to consider many replicates and conduct experiments that would
be often impossible or unethical to perform in the real world: for
instance, does a given fishery experience better long-term outcomes when
managed according to forecasts derived from model 1 or from model 2?

\hypertarget{a-motivating-example}{%
\section{A motivating example}\label{a-motivating-example}}

To better understand how a model can produce a more accurate forecast
and yet still lead to a worse decision, it may be helpful to start with
a concrete example in which a manager faces a trade-off between
cormorant conservation and fish harvest. Fig 1 shows both the forecasts
and realized management outcomes of two alternative three-species
models, ``A'' and ``B'' (see Appendix A) in predicting the population
dynamics of striped bass (an economically important fishery) and
double-crested cormorants (a target species for conservation) which both
feed on a population of river herring (whose abundance we assume is not
measured). Model A accurately forecasts the abundance of both bass and
cormorants well into the future, but the optimal management strategy
derived from its forecasts leads to steadily declining species
abundances and overall disappointing net utility. Model B produces
substantially less accurate forecasts, but nevertheless achieves better
outcomes. The net utility under model B is in fact nearly identical to
the maximum utility attainable given the true model, while management
under model A achieves only 38\% of that utility.

This management problem is motivated by a real world example of a
herring fishery as described in Brias \& Munch (2021), in which our
manager seeks to balance multiple objectives of sustaining the cormorant
population while maximizing the economic value from harvesting both
herring and striped bass. In the scenarios depicted in Fig 1, I have
used a the richer five-species model introduced Brias \& Munch (2021) to
drive the underlying dynamics, which includes three competing species of
herring that are preyed upon by both the bass and the cormorants. Here,
the manager seeks to maximize utility given by the weighted sum of the
individual objectives (Brias \& Munch 2021), placing 50\% of the weight
on the conservation objective and splits the remaining weight evenly
over the harvests for predator (bass) and prey (herring) species. I have
assumed a partially observable system - in this case, the manager
measures only the abundance of bass and cormorant species, and not of
the three herring species. In this scenario, I have further assumed the
manager must choose a fixed fishing effort for herring and for the bass
harvest, I will consider more dynamic decision-processes later, but it
is worth noting that in many real world conservation settings policy
choices are highly constrained and frequent adjustment of those policies
may be costly or impossible. Likewise, the assumptions of partially
observable system and imperfect models are characteristic of ecological
decision-making. Equations and code for all models in this example are
presented in Appendix A.

Both models A \& B can be seen as alternative attempts to approximate
the ``true'' model (generative process), which in real systems is always
unknown and more complex than any model thereof. Model A assumes a
three-species Ricker model which closely matches the trophic structure
of the ``true'' five-species model, lumping the three competing herring
species into a single variable. Because herring abundance is not
observed directly, the model parameters related to herring growth are
less accurate than other parameters. Model B also lumps herring species
together into a single variable, but fails to reflect the trophic
relationship between bass and herring. Model B also oversimplifies the
relationship between cormorant and herring population. This does not
make Model B an unreasonable model out-of-hand -- all models contain
such simplifications (e.g.~our ``true'' model does not model the trophic
relationship between the herring and its food sources or environmental
conditions explicitly either). Both models are consistent with the
limited historical data available to them.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/example1-1.pdf}
\caption{Forecast performance and realized outcomes from management of a
5-species system under either model. Based on forecast performance
alone, Model A clearly performs better, accurately predicting steady
declines. Model B predicts overly optimistic outcomes for Bass
population levels, and overly pessimistic outcomes for cormorants, with
observed dynamics falling well outside the predicted range. Despite
this, net utility achieved under this management regime is virtually
optimal, while the declines under model A result in net utility that is
only 38\% of optimal.}
\end{figure}

\hypertarget{an-iterative-decision-example}{%
\section{An Iterative Decision
Example}\label{an-iterative-decision-example}}

While this example demonstrates that the model which provides the better
forecast does not necessarily lead to a better decision, it may raise
more questions than it answers. Why does this happen? Is this an
isolated example or not? Can this forecast trap be resolved by more
sophisticated approaches to model selection and decision-making? I now
consider scenarios involving iterative forecasts and adaptive
management: in which the manager monitors outcomes, compares forecasts
to observations and updates model estimates. Such \emph{sequential
decision processes} are not merely iterative versions of single-decision
problems, but are much more challenging. In the opening example, the
manager had to choose the harvest policy for each fish species at the
start of the scenario and stick with it. The ability to select new
actions in response to new observations turns that decision into a game
of chess: each turn, the manager must consider not only their next move
but all possible series of moves.

How do we translate a model-based forecast into a decision policy? It is
impossible to discuss outcomes associated with a forecast without first
agreeing on this process. In practice, decision-makers may use a
forecast in a wide variety of ways in selecting a course of action,
including ways which may run counter to the stated objectives of
management (Ludwig \emph{et al.} 1993). In principle at least, the field
of decision theory provides a formal mechanism for determining the
optimal strategy given a model forecast. For instance, a wide range of
ecological conservation and management problems can be expressed as a
Markov Decision Process (MDP) problems (Marescot \emph{et al.} 2013).
Existing computer algorithms such as stochastic dynamic programming
(SDP) take a probabilistic model \emph{forecast} (more precisely, the
probability \(P(x_{t+1} | x_t, a_t)\) of the system being in state
\(x_{t+1}\) in the next iteration given that it was previously in state
\(x_t\) and the manager selected action \(a_t\)) and the \emph{desired
management objective} (i.e.~the maximize the expected biomass of species
protected or the expected dollar profit of a fishery (see Clark 1990;
Halpern \emph{et al.} 2013)) as input, and return the \emph{decision
policy} which maximizes that objective (Marescot \emph{et al.} 2013).
This provides a principled way to associate a decision policy with any
given forecast model.

Two features of this approach are worth emphasizing. As before, the
resulting decision is derived directly from the forecast model and the
desired objective. The SDP algorithm is a reasonable description of the
approach any ideal manager would use -- considering all possible
outcomes from all possible sequences of actions and selecting the best
sequence. For complex models this process is too laborious even for a
computer, and is often simplified by considering only a selection of
predetermined policies (as in Management Strategy Evaluation, MSE, Punt
\emph{et al.} (2016)), or scenarios (as in scenario analysis, Polasky
\emph{et al.} (2011)). Such shortcuts are often necessary for complex
real-world models, but open additional room for error: the policy we
derive from a given forecast may perform poorly not because the model
forecast was at fault, but because of those simplifying assumptions
about possible policies. To ensure that the forecast trap is not a
result of such assumptions about possible policies, we will consider a
problem simple enough to solve directly with SDP. The resulting decision
policy is optimal, so long as the forecast model is correct. In this
way, the SDP merely stands in for a mathematically precise way in which
forecasts are turned into decisions. Recognizing the SDP-derived policy
(A) comes directly from the forecast model, and (B) gives the optimal
policy for said forecast, seems to suggest that whichever model makes
the better forecast will surely also lead to better outcomes (as
measured in terms of whatever utility we have chosen to maximize). While
this intuition is no doubt \emph{often} accurate, our purpose here is to
demonstrate that it is by no means \emph{guaranteed}: it is also
possible for the model which makes the better forecast to lead to worse
outcomes.

Let us consider the management of single species in which we seek to
maximize the long-term net harvest. In this scenario, the manager
estimates the population size each year and must set the total allowable
catch (TAC) for that season. The underlying dynamics are unknown, but
the manager is presented with any of a variety of forecast models which
can predict the future stock sizes given the current population size and
proposed TAC. Our manager does not know which of these models is the
most accurate a priori. Instead, the manager will be able to compare the
population size predicted by each forecast (under the chosen TAC) to the
measurement of the population size in the following year before coming
up with the next year's catch limit.

Faced with a collection of models, a manager can seek either to identify
the best model to use, or to consider an integrative assessment which
uses the whole ensemble of models to represent the manager's uncertainty
about the underlying process. I will consider both approaches in turn.
Figure 2 compares forecasts generated by two of the candidate models to
observations drawn from simulations of the underlying process. As
before, these are true forecasts: the model forecasts are generated
first, they have not been fit to these observations. In the un-fished
scenario (top panels), both models try to predicting the same un-fished
equilibrium dynamics. In the second scenario (lower panel), the manager
uses the optimal SDP policy derived from each forecast to determine the
TAC for the following year, and compares the observed stock size to that
which the model predicted given that fishing quota. In both cases, model
2 provides far more accurate forecasts, as seen in the error bars and
confirmed by the distribution of proper scores {[}Fig 2C-D; Gneiting \&
Raftery (2007){]}. Model and simulation details are provided in Appendix
B.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/figure2-1.pdf}
\caption{Forecast performance of each model. Panels A, B: Step ahead
predictions of stock size under unfished (A) and fished (B) scenarios.
Error bars indicating the 95\% confidence intervals around each
prediction, while stars denote the observed value in that year. Because
the models make different decisions each year in the fished scenario,
the observed stock size in year 2, 3, etc under the management of model
1 (blue stars) is different from that under model 2 (red stars). Panels
C, D: corresponding distribution of proper scores across all predictions
(100 replicates of 100 timesteps). Higher scores are better, confirming
that model 2 makes the better forecasts.}
\end{figure}

Despite the clearly superior predictive accuracy of model 2 in both
scenarios, the outcomes from management under model 2 are substantially
worse. We can assess such outcomes in less abstract terms than
forecasting skill, such as economic value them manager sought to
optimize (in dollars) or the ecological value (unharvested biomass)
{[}Fig 3{]}.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/figure3-1.pdf}
\caption{Ecological and economic performance of each forecast. Harvest
quotas derived from model 1 result in a significantly higher fish stock
size than under Model 2 (panel A). Economic returns under model 1 are
also substantially higher (panel B)}
\end{figure}

A manager operating under model 2 would have little indication that the
model was flawed: both future stock sizes and expected harvest yields
consistently match model predictions. This manager would be stuck in the
forecast trap, incorrectly mistaking a degraded ecological state and
reduced economic outcomes as the best that can be achieved because
future observations continue to validate this model. Had we been able to
include Model 3 in our forecast comparisons, it would equal or
outperform the forecasting skill of both model 1 and model 2 (as
guaranteed by the theorem of Gneiting \& Raftery (2007)), while also
matching the economic utility of model 1 (as guaranteed by the theorem
of Reed (1979)). In practice, we never have access to the generating
model, so it is reasonable to expect model selection to determine the
better approximation. As we see here, the better approximation for
forecasting future states does not in fact lead to better outcomes.

\hypertarget{adaptive-management}{%
\subsection{Adaptive Management}\label{adaptive-management}}

Rather than select a single best model (or best parameter value), a
manager could choose to integrate over possible outcomes generated from
all candidate models. Updating posterior distributions over parameters
and/or weights assigned to different models are examples of this kind of
adaptive management (Ludwig \& Walters 1982; Punt \emph{et al.} 2016). I
illustrate the application of such an adaptive management strategy,
following classic examples for parameter (Ludwig \& Walters 1982) or
structural (Smith \& Walters 1981) model uncertainty. I first consider
only the same two models considered in the previous example. I later
consider a larger suite of 42 models, spanning the parameter space of
Gordon-Schaefer curves. To avoid failure to explore sufficiently, (see
exploration-exploitation trade-off, e.g. Walters 1981), I assign prior
belief of 99\% weight on the optimally performing model, model 1.\\
For comparison, I consider the baseline case in which the manager does
not update posterior distribution over which models/parameter values are
correct (the manager still chooses a new TAC after each observation, but
does not update their belief in the model, i.e.~does not \emph{learn}
over time). The difference between the performance with and without
learning is known as the ``Value of Information'' (Howard 1966). In both
2-model and 42-model scenarios, the value of information is strongly
negative. The 2-model case achieves a net present value to -58\% of the
value of having used model 1 alone {[}Fig 4{]}. Including all 42 models
reduces this to a value of -32\%. Both harvests and fish biomass remain
significantly lower under adaptive learning scenarios.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/figure4-1.pdf}
\caption{Adaptive management under model uncertainty. Solid lines trace
the trajectories of the state (fish stock, circles) and action (harvest
quota, triangles), under adaptive management (learning). Dotted lines
trace the corresponding trajectories if iterative learning is omitted,
leaving the prior belief fixed throughout the simulation (planning).
Color indicates the belief that model 1 is correct (blue), with an
initial prior belief of 99\%. Panel A: Management over the two candidate
models, Model 1 and Model 2. Within a single iteration of adaptive
management, the belief over models switches from a prior belief that
heavily favored model 1 to a posterior that favors model 2 with near
certainty. Future iterations reinforce the belief in model 2, resulting
in both depressed harvests and low stock sizes (solid lines). If no
iterative learning updates are performed, stock sizes and realized
harvests (and thus economic profit) are both higher. Panel B: given 42
candidate models over a broad range of parameter values, adaptive
management quickly reduces the probability of model 1, and substantially
underperforms management without learning (dotted lines). While outcomes
improve marginally relative to the two-model case (panel A) they remain
significantly worse than had no iterative learning been included.}
\end{figure}

The reason for model 1's seemingly contradictory ability to make good
decisions but bad forecasts becomes obvious once we compare both curves
to that of the underlying model, model 3. Looking at plots of the growth
rate curves for each model {[}Fig 5A{]}, it is hardly surprising that
all model selection approaches prefer the closely overlapping curve of
model 2 to the no-where-close curve of model 1 as the better
approximation of model 3. Nevertheless, the decision policy derived from
model 1 forecasts is indistinguishable from that based on the true model
{[}Fig 5B{]}, while the policy derived from model 2 forecasts lead to
over-harvesting. Being closest to the true model's forecast skill never
guarantees that we are closest to the true model's optimal policy.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/figure5-1.pdf}
\caption{Panel A: Population growth curves of each model. The positive
equilibrium of each model occurs where the curve crosses the horizontal
axis. Note that while Model 2 is a better approximation to the truth
(Model 3), Model 1 better approximates the stock size which leads to
maximum growth. Panel B: The optimal control policy under Model 1 is
nearly identical to that under the true Model 3, while the optimal
policy under Model 2 suppresses stock to a much lower escapement level.}
\end{figure}

How can the very different forecasts from model 1 and model 3 could
produce exactly the same optimal management policy (Fig 5B) under the
SDP algorithm? Analytic solutions offer more insight as to when and why
very different forecasts can generate the identical policy. Such a
solution was first provided by Reed (1979), who demonstrated the optimal
policy in the case considered here would be a so-called ``bang-bang''
policy. Intuitively one can think of this as maintaining the biomass at
the most productive size: the maximum population growth rate (position
of the peak of the growth curves in Fig 5A), though this is only
precisely true without discounting (\(\delta = 1\)): the optimal stock
size \(\hat x\) is the solution to \(f(\hat x) = \hat x/\delta\) when
stochasticity is sufficiently small (Reed 1979). Thus, all models in
which the peak growth rate occurs at the same stock size will have the
same optimal policy. These are not merely bad models getting lucky --
all such models correctly capture the crucial feature relevant to the
decision. In more complex models, such features are more difficult or
impossible to identify analytically; but this does not mean they do not
exist. For instance, Recent mathematical breakthroughs such as Holden \&
Conrad (2015), Hening \emph{et al.} (2019), and Hening (2021) have
proven that the optimal harvest control rule in age-structured and
predator-prey systems maintain similar bang-bang dynamics. This means
that the optimal policy of such very complex models will once again be
shared by infinite number of simpler models.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

The forecast trap illustrated in both examples can best be understood as
a problem of \emph{non-uniqueness} (Oreskes \emph{et al.} 1994). Even
modestly complex models can successfully predict the observed dynamics,
but for wrong mechanistic reasons (Schnute \& Richards 2001; Schindler
\& Hilborn 2015). In both examples, a decision-maker who accepts the
model which leads to very accurate predictions as the basis for their
decision-making winds up in the forecast trap: accepting poor ecological
and economic outcomes as the best possible option. The space of possible
models is infinitely large when measured against any a scalar metric
such as mean forecast skill or expected net utility. Perhaps it should
be no surprise then that many models will achieve the same policy
outcomes or achieve comparable predictive accuracy. Just as the forecast
skill is not unique, both examples also demonstrated that the optimal
policy is not unique to the ``true model'' -- many models will result in
the same policy and achieve the same outcomes; despite making very
different forecasts.

The forecast trap is likely to be more common in contexts in which
systems are more complex, partially observed, and available actions are
constrained -- all features which are particularly common to ecological
management and conservation. Because simplicity of the second example
allows analytic theory to reveal a precise explanation, it is tempting
to assume the trap is only a consequence of examining overly simple
models. In fact, the opposite is true. In the second example, the
``true'' model is simple enough to be covered by a suitable candidate
set of models (e.g.~a Gaussian Process, Boettiger \emph{et al.} (2015)),
which would resolve the trap given sufficient data. In reality, our
models never span the true process. Partially-observed systems increase
the space of possible models that achieve comparable predictive
accuracy. Constraints on action space such as adjustment costs
(Boettiger \emph{et al.} 2016) or piecewise-linear control rules (Punt
2010) increase the space of models which will result in the same policy.
Both of these aspects make the forecast trap easier to encounter in our
opening example.

The forecast trap demonstrates that for certain ensembles of candidate
models, the value of information (VOI) (Howard 1966; Katz \emph{et al.}
1987) can in fact be negative. Consequently, methods to select models or
re-estimate parameters can lead to worse outcomes than had these new
observations simply been ignored. Crucially, a manager implementing the
optimal policy from the most predictive model sees no indication that
their models are wrong -- the declining ecosystem and economic returns
observed under model A in the first example or under iterative learning
in the second example are completely consistent with and predicted by
the models. Only by winding back the clock, making decisions based over
the original uncertainty without learning (Fig 4), can we see that
better outcomes could have been achieved. The policy derived without
learning reflects greater uncertainty: it is thus more \emph{robust}.

\hypertarget{a-way-forward}{%
\subsection{A way forward}\label{a-way-forward}}

In practice, managers rarely rely only on forecasting skill to assess
models, nor determine policies directly from forecasts alone. In both
examples presented above, the forecast trap is most likely in
circumstances where the collection of candidate models is insufficiently
broad. Management practice tends to emphasize approaches which
\emph{broaden} rather than \emph{narrow down} this candidate set. This
reflects the view that ``the primary values of ecosystem models are as
heuristic tools for communication and for developing scenarios to
express uncertainties and test policies'' rather than as a source of
reliable forecasts (Schindler \& Hilborn 2015). Such practices include:

\begin{enumerate}
\def\labelenumi{(\alph{enumi})}
\tightlist
\item
  emphasizing a better articulation of uncertainty \emph{a priori};
\item
  active exploration of alternative policies can reveal when the model
  set is inadequate
\item
  methods for generating strategies that are robust to the sort of
  uncertainty described here.
\end{enumerate}

A central role of models is to help articulate uncertainty around
possible outcomes rather than make precise predictions. As Schindler \&
Hilborn (2015) notes, ``current approaches to verification and
validation of ecosystem models likely produce overly optimistic
impressions of the reliability of forecasts underlying management and
conservation prescriptions.'' Forecasts based on non-mechanistic models
such as empirical dynamical modeling (EDM, Ye \emph{et al.} 2015; Brias
\& Munch 2021) may help in articulating a broader ensemble of scenarios
(Boettiger \emph{et al.} 2015). In contrast, if such approaches are
selected solely on forecasting skill, they may increase the probability
of the forecast trap. The danger of an insufficiently broad model
ensemble is well understood in many disciplines which use \emph{scenario
planning} to assist policy in accounting for irreducible uncertainties
(Peterson \emph{et al.} 2003).

Second, a manager may also escape the forecast trap by exploring actions
which are never recommended from any of the available forecasts. For
example, a manager looking at the low stocks and poor harvests achieved
under model 2, could decide experimentally to reduce fishing quota. This
will allow the population to enter a range of state-space where
discrepancies between model 2 and observations are more obvious. Runge
\emph{et al.} (2016) describes how such ``double-loop learning'' can be
used to identify when the entire model set is inaccurate, a problem
which is not solved by ``single-loop'' adaptive-management in our
examples above. Schindler \& Hilborn (2015) also underscores the value
of flexible policies; rather than ``managing solely within the range of
past variation; active probing is usually needed,'' and contrasts this
to a typical interpretation of the `precautionary principle' often cited
as a reason to avoid exploratory actions. However, just because active
exploration can escape the forecast trap does not mean it is always a
good idea.

Third, managers may emphasize policy robustness over forecast skill
(Schindler \& Hilborn 2015). In many formal treatments, this is not
qualitatively different to the analysis considered here: a manager
simply chooses a different utility function, such as minimizing `regret'
rather than maximizing expected value (Polasky \emph{et al.} 2011). Such
approaches are just as vulnerable to bad outcomes (as defined by their
own utility functions) whenever models are selected only on the basis of
forecast skill. Alternative approaches may not seek any such
optimization, emphasize the viability (Aubin 1991) of possible policy
under constraints. In practice, robust design may emphasize acceptable
performance across the widest possible array of scenarios
(e.g.~candidate models). This acts more like a sensitivity analysis of
utility with respect to underlying assumptions, rather than an
optimization routine (Fischer \emph{et al.} 2009; e.g. Punt \emph{et
al.} 2016). Computationally, the former is much simpler, allowing
researchers to evaluate the performance of a policy on more complex
simulations for which calculating the optimal policy would be
prohibitively difficult.

Finally, it is worth noting that decisions do not need to be premised on
a forecast at all, but can be premised entirely on the basis of past
experience: `If the fish stock size has gone up, increase harvest
slightly, otherwise, decrease slightly.' Such a policy is not optimal,
but it is robust across a wide range of unimodal stock-recruitment
curves without ever estimating a predictive model. This is the basis of
so-called `model-free' reinforcement learning algorithms such as DQN
(Mnih \emph{et al.} 2015) and SAC (Haarnoja \emph{et al.} 2018), which
train deep neural networks to learn a policy without ever attempting to
predict future states of the underlying process. Training such a
artificial intelligence agents across a wide suite of simulations, a
process known as curriculum learning (Graves \emph{et al.} 2017), mimics
the scenario analyses and search for robust policies. Such approaches
have been used to train agents to play 2600 Atari console games at
superhuman ability (Mnih \emph{et al.} 2015), outperform race-car
drivers (Wurman \emph{et al.} 2022) and control nuclear fusion reactions
(Degrave \emph{et al.} 2022). Such approaches are also not yet well
understood, introducing new risks as well as new
possibilities(Dulac-Arnold \emph{et al.} 2019; Henderson \emph{et al.}
2019).

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

The author acknowledges support from NSF CAREER Award \#1942280 and
helpful discussions with Melissa Chapman. The author is also deeply
grateful to the Michael Runge, Chih-hao Hsieh, Antoine Brias, and other
anonymous reviewers whose detailed feedback and insights have greatly
shaped the paper and my own understanding of these issues.

\pagebreak

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Aubin1991}{}}%
Aubin, J.P. (1991). \emph{Viability theory}. Systems \& control.
Birkhauser, Boston.

\leavevmode\vadjust pre{\hypertarget{ref-Boettiger2021}{}}%
Boettiger, C. (2021).
\href{https://doi.org/10.1007/s12080-020-00477-4}{Ecological management
of stochastic systems with long transients}. \emph{Theoretical Ecology},
14, 663--671.

\leavevmode\vadjust pre{\hypertarget{ref-Boettiger2016}{}}%
Boettiger, C., Bode, M., Sanchirico, J.N., LaRiviere, J., Hastings, A.
\& Armsworth, P.R. (2016).
\href{https://doi.org/10.1890/15-0236}{Optimal management of a
stochastically varying population when policy adjustment is costly}.
\emph{Ecological Applications}, 26, 808--817.

\leavevmode\vadjust pre{\hypertarget{ref-Boettiger2015}{}}%
Boettiger, C., Mangel, M. \& Munch, S. (2015).
\href{https://doi.org/10.1098/rspb.2014.1631}{Avoiding tipping points in
fisheries management through {Gaussian} process dynamic programming}.
\emph{Proceedings of the Royal Society B: Biological Sciences}, 282,
20141631--20141631.

\leavevmode\vadjust pre{\hypertarget{ref-Brias2021}{}}%
Brias, A. \& Munch, S.B. (2021).
\href{https://doi.org/10.1016/j.ecolmodel.2020.109423}{Ecosystem based
multi-species management using {Empirical} {Dynamic} {Programming}}.
\emph{Ecological Modelling}, 441, 109423.

\leavevmode\vadjust pre{\hypertarget{ref-Burnham1998}{}}%
Burnham, K.P. \& Anderson, D.R. (1998).
\href{https://doi.org/10.1007/978-1-4757-2917-7_3}{Practical {Use} of
the {Information}-{Theoretic} {Approach}}. In: \emph{Model {Selection}
and {Inference}}. Springer New York, New York, NY, pp. 75--117.

\leavevmode\vadjust pre{\hypertarget{ref-Chades2008}{}}%
Chadès, I., McDonald-Madden, E., McCarthy, M.a., Wintle, B., Linkie, M.
\& Possingham, H.P. (2008).
\href{https://doi.org/10.1073/pnas.0805265105}{When to stop managing or
surveying cryptic threatened species.} \emph{Proceedings of the National
Academy of Sciences}, 105, 13936--40.

\leavevmode\vadjust pre{\hypertarget{ref-Clark1973}{}}%
Clark, C.W. (1973). \href{https://doi.org/10.1086/260090}{Profit
maximization and the extinction of animal species}. \emph{Journal of
Political Economy}, 81, 950--961.

\leavevmode\vadjust pre{\hypertarget{ref-Clark1990}{}}%
Clark, C.W. (1990). \emph{{Mathematical Bioeconomics: The Optimal
Management of Renewable Resources, 2nd Edition}}. Wiley-Interscience.

\leavevmode\vadjust pre{\hypertarget{ref-Clark2001}{}}%
Clark, J.S., Carpenter, S.R., Barber, M., Collins, S., Dobson, A.,
Foley, J.A., \emph{et al.} (2001).
\href{https://doi.org/10.1126/science.293.5530.657}{Ecological
{Forecasts}: {An} {Emerging} {Imperative}}. \emph{Science}, 293,
657--660.

\leavevmode\vadjust pre{\hypertarget{ref-Costello2016}{}}%
Costello, C., Ovando, D., Clavelle, T., Strauss, C.K., Hilborn, R.,
Melnychuk, M.C., \emph{et al.} (2016).
\href{https://doi.org/10.1073/pnas.1520420113}{{Global fishery prospects
under contrasting management regimes}}. \emph{Proceedings of the
National Academy of Sciences}, 113, 5125--5129.

\leavevmode\vadjust pre{\hypertarget{ref-deepmind_fusion}{}}%
Degrave, J., Felici, F., Buchli, J., Neunert, M., Tracey, B., Carpanese,
F., \emph{et al.} (2022).
\href{https://doi.org/10.1038/s41586-021-04301-9}{Magnetic control of
tokamak plasmas through deep reinforcement learning}. \emph{Nature},
602, 414--419.

\leavevmode\vadjust pre{\hypertarget{ref-Dietze2018}{}}%
Dietze, M.C., Fox, A., Beck-Johnson, L.M., Betancourt, J.L., Hooten,
M.B., Jarnevich, C.S., \emph{et al.} (2018).
\href{https://doi.org/10.1073/pnas.1710231115}{Iterative near-term
ecological forecasting: {Needs}, opportunities, and challenges}.
\emph{Proceedings of the National Academy of Sciences}, 115, 1424--1432.

\leavevmode\vadjust pre{\hypertarget{ref-Dulac-Arnold_2019}{}}%
Dulac-Arnold, G., Mankowitz, D. \& Hester, T. (2019).
\href{http://arxiv.org/abs/1904.12901}{Challenges of {Real}-{World}
{Reinforcement} {Learning}}. \emph{arXiv:1904.12901 {[}cs, stat{]}}.

\leavevmode\vadjust pre{\hypertarget{ref-Fischer2009}{}}%
Fischer, J., Peterson, G.D., Gardner, T.A., Gordon, L.J., Fazey, I.,
Elmqvist, T., \emph{et al.} (2009).
\href{https://doi.org/10.1016/j.tree.2009.03.020}{Integrating resilience
thinking and optimisation for conservation.} \emph{Trends in ecology \&
evolution}, 24, 549--54.

\leavevmode\vadjust pre{\hypertarget{ref-Ginzburg2004}{}}%
Ginzburg, L.R. \& Jensen, C.X.J. (2004).
\href{https://doi.org/10.1016/j.tree.2003.11.004}{Rules of thumb for
judging ecological theories}. \emph{Trends in Ecology \& Evolution}, 19,
121--126.

\leavevmode\vadjust pre{\hypertarget{ref-Gneiting2007}{}}%
Gneiting, T. \& Raftery, A.E. (2007).
\href{https://doi.org/10.1198/016214506000001437}{Strictly {Proper}
{Scoring} {Rules}, {Prediction}, and {Estimation}}. \emph{Journal of the
American Statistical Association}, 102, 359--378.

\leavevmode\vadjust pre{\hypertarget{ref-curriculum_learning}{}}%
Graves, A., Bellemare, M.G., Menick, J., Munos, R. \& Kavukcuoglu, K.
(2017). \href{http://arxiv.org/abs/1704.03003}{Automated {Curriculum}
{Learning} for {Neural} {Networks}}. \emph{arXiv:1704.03003 {[}cs{]}}.

\leavevmode\vadjust pre{\hypertarget{ref-SAC}{}}%
Haarnoja, T., Zhou, A., Abbeel, P. \& Levine, S. (2018).
\href{http://arxiv.org/abs/1801.01290}{Soft {Actor}-{Critic}:
{Off}-{Policy} {Maximum} {Entropy} {Deep} {Reinforcement} {Learning}
with a {Stochastic} {Actor}}. \emph{arXiv:1801.01290 {[}cs, stat{]}}.

\leavevmode\vadjust pre{\hypertarget{ref-Halpern2013}{}}%
Halpern, B.S., Klein, C.J., Brown, C.J., Beger, M., Grantham, H.S.,
Mangubhai, S., \emph{et al.} (2013).
\href{https://doi.org/10.1073/pnas.1217689110}{Achieving the triple
bottom line in the face of inherent trade-offs among social equity,
economic return, and conservation.} \emph{Proceedings of the National
Academy of Sciences}, 110, 6229--34.

\leavevmode\vadjust pre{\hypertarget{ref-Henderson2019}{}}%
Henderson, P., Islam, R., Bachman, P., Pineau, J., Precup, D. \& Meger,
D. (2019). \href{http://arxiv.org/abs/1709.06560}{Deep {Reinforcement}
{Learning} that {Matters}}. \emph{arXiv:1709.06560 {[}cs, stat{]}}.

\leavevmode\vadjust pre{\hypertarget{ref-Hening2021}{}}%
Hening, A. (2021).
\href{https://doi.org/10.1007/s00332-020-09667-0}{Coexistence,
{Extinction}, and {Optimal} {Harvesting} in {Discrete}-{Time}
{Stochastic} {Population} {Models}}. \emph{Journal of Nonlinear
Science}, 31, 1.

\leavevmode\vadjust pre{\hypertarget{ref-Hening2019}{}}%
Hening, A., Nguyen, D.H., Ungureanu, S.C. \& Wong, T.K. (2019).
\href{https://doi.org/10.1007/s00285-018-1275-1}{Asymptotic harvesting
of populations in random environments}. \emph{Journal of Mathematical
Biology}, 78, 293--329.

\leavevmode\vadjust pre{\hypertarget{ref-Holden2015}{}}%
Holden, M.H. \& Conrad, J.M. (2015).
\href{http://dx.doi.org/10.1016/j.mbs.2015.08.021}{Optimal escapement in
stage-structured fisheries with environmental stochasticity}.
\emph{Mathematical Biosciences}, 269, 76--85.

\leavevmode\vadjust pre{\hypertarget{ref-Howard1966}{}}%
Howard, R. (1966).
\href{https://doi.org/10.1109/TSSC.1966.300074}{Information {Value}
{Theory}}. \emph{IEEE Transactions on Systems Science and Cybernetics},
2, 22--26.

\leavevmode\vadjust pre{\hypertarget{ref-Katz1987}{}}%
Katz, R.W., Brown, B.G. \& Murphy, A.H. (1987).
\href{https://doi.org/10.1002/for.3980060202}{Decision-analytic
assessment of the economic value of weather forecasts: {The}
fallowing/planting problem}. \emph{Journal of Forecasting}, 6, 77--89.

\leavevmode\vadjust pre{\hypertarget{ref-Lande1994}{}}%
Lande, R., Engen, S. \& Saether, B.-E. (1994).
\href{https://doi.org/10.1038/372088a0}{Optimal harvesting, economic
discounting and extinction risk in fluctuating populations}.
\emph{Nature}, 372, 88--90.

\leavevmode\vadjust pre{\hypertarget{ref-Li2017}{}}%
Li, S.-L., Bjørnstad, O.N., Ferrari, M.J., Mummah, R., Runge, M.C.,
Fonnesbeck, C.J., \emph{et al.} (2017).
\href{https://doi.org/10.1073/pnas.1617482114}{Essential information:
{Uncertainty} and optimal control of {Ebola} outbreaks}.
\emph{Proceedings of the National Academy of Sciences}, 114, 5659--5664.

\leavevmode\vadjust pre{\hypertarget{ref-Ludwig1993}{}}%
Ludwig, D., Hilborn, R. \& Walters, C. (1993).
\href{https://doi.org/10.1126/science.260.5104.17}{Uncertainty,
{Resource} {Exploitation}, and {Conservation}: {Lessons} from
{History}}. \emph{Science}, 260, 17--36.

\leavevmode\vadjust pre{\hypertarget{ref-Ludwig1982}{}}%
Ludwig, D. \& Walters, C.J. (1982).
\href{https://doi.org/10.1016/0304-3800(82)90023-0}{Optimal harvesting
with imprecise parameter estimates}. \emph{Ecological Modelling}, 14,
273--292.

\leavevmode\vadjust pre{\hypertarget{ref-Marescot2013}{}}%
Marescot, L., Chapron, G., Chadès, I., Fackler, P.L., Duchamp, C.,
Marboutin, E., \emph{et al.} (2013).
\href{https://doi.org/10.1111/2041-210X.12082}{Complex decisions made
simple: A primer on stochastic dynamic programming}. \emph{Methods in
Ecology and Evolution}, 4, 872--884.

\leavevmode\vadjust pre{\hypertarget{ref-DQN}{}}%
Mnih, V., Kavukcuoglu, K., Silver, D., Rusu, A.A., Veness, J.,
Bellemare, M.G., \emph{et al.} (2015).
\href{https://doi.org/10.1038/nature14236}{Human-level control through
deep reinforcement learning}. \emph{Nature}, 518, 529--533.

\leavevmode\vadjust pre{\hypertarget{ref-Nitzbon2017}{}}%
Nitzbon, J., Heitzig, J. \& Parlitz, U. (2017).
\href{https://doi.org/10.1088/1748-9326/aa7581}{Sustainability, collapse
and oscillations in a simple {World}-{Earth} model}. \emph{Environmental
Research Letters}, 12, 074020.

\leavevmode\vadjust pre{\hypertarget{ref-Oreskes1994}{}}%
Oreskes, N., Shrader-Frechette, K. \& Belitz, K. (1994).
\href{https://doi.org/10.1126/science.263.5147.641}{Verification,
{Validation}, and {Confirmation} of {Numerical} {Models} in the {Earth}
{Sciences}}. \emph{Science}, 263, 641--646.

\leavevmode\vadjust pre{\hypertarget{ref-Peterson2003}{}}%
Peterson, G.D., Cumming, G.S. \& Carpenter, S.R. (2003).
\href{https://doi.org/10.1046/j.1523-1739.2003.01491.x}{Scenario
{Planning}: A {Tool} for {Conservation} in an {Uncertain} {World}}.
\emph{Conservation Biology}, 17, 358--366.

\leavevmode\vadjust pre{\hypertarget{ref-Polasky2011}{}}%
Polasky, S., Carpenter, S.R., Folke, C. \& Keeler, B. (2011).
\href{https://doi.org/10.1016/j.tree.2011.04.007}{{Decision-making under
great uncertainty: environmental management in an era of global
change}}. \emph{Trends in Ecology {\&} Evolution}, 26, 398--404.

\leavevmode\vadjust pre{\hypertarget{ref-Punt2010}{}}%
Punt, A.E. (2010). Harvest control rules and fisheries management. In:
\emph{Handbook of {Marine} {Fisheries} {Conservation} and {Management}
.} (eds. Grafton, RQ, Hilborn, R, Squires, D, Tait, M \& Williams, M).
Oxford University Press.

\leavevmode\vadjust pre{\hypertarget{ref-Punt2016}{}}%
Punt, A.E., Butterworth, D.S., Moor, C.L. de, De Oliveira, J.A.A. \&
Haddon, M. (2016). \href{https://doi.org/10.1111/faf.12104}{Management
strategy evaluation: Best practices}. \emph{Fish and Fisheries}, 17,
303--334.

\leavevmode\vadjust pre{\hypertarget{ref-Reed1979}{}}%
Reed, W.J. (1979).
\href{https://doi.org/10.1016/0095-0696(79)90014-7}{{Optimal escapement
levels in stochastic and deterministic harvesting models}}.
\emph{Journal of Environmental Economics and Management}, 6, 350--363.

\leavevmode\vadjust pre{\hypertarget{ref-Richards1999}{}}%
Richards, S.A., Possingham, H.P. \& Tizard, J. (1999).
\href{https://doi.org/10.1890/1051-0761(1999)009\%5B0880:OFMFMC\%5D2.0.CO;2}{Optimal
{Fire} {Management} for {Maintaining} {Community} {Diversity}}.
\emph{Ecological Applications}, 9, 880--892.

\leavevmode\vadjust pre{\hypertarget{ref-Runge2016}{}}%
Runge, M.C., Stroeve, J.C., Barrett, A.P. \& McDonald-Madden, E. (2016).
\href{https://doi.org/10.1038/nclimate3041}{Detecting failure of climate
predictions}. \emph{Nature Climate Change}, 6, 861--864.

\leavevmode\vadjust pre{\hypertarget{ref-Schindler2015}{}}%
Schindler, D.E. \& Hilborn, R. (2015).
\href{https://doi.org/10.1126/science.1261824}{Prediction, precaution,
and policy under global change}. \emph{Science}, 347, 953--954.

\leavevmode\vadjust pre{\hypertarget{ref-Schnute2001}{}}%
Schnute, J.T. \& Richards, L.J. (2001).
\href{https://doi.org/10.1139/f00-150}{Use and abuse of fishery models}.
\emph{Canadian Journal of Fisheries and Aquatic Sciences}, 58, 10--17.

\leavevmode\vadjust pre{\hypertarget{ref-Shea2014}{}}%
Shea, K., Tildesley, M.J., Runge, M.C., Fonnesbeck, C.J. \& Ferrari,
M.J. (2014).
\href{https://doi.org/10.1371/journal.pbio.1001970}{Adaptive
{Management} and the {Value} of {Information}: {Learning} {Via}
{Intervention} in {Epidemiology}}. \emph{PLoS Biology}, 12, 9--12.

\leavevmode\vadjust pre{\hypertarget{ref-Smith1981}{}}%
Smith, A.D.M. \& Walters, C.J. (1981).
\href{https://doi.org/10.1139/f81-092}{Adaptive {Management} of
{Stock}--{Recruitment} {Systems}}. \emph{Canadian Journal of Fisheries
and Aquatic Sciences}, 38, 690--703.

\leavevmode\vadjust pre{\hypertarget{ref-Walters1981}{}}%
Walters, C.J. (1981). \href{https://doi.org/10.1139/f81-091}{Optimum
{Escapements} in the {Face} of {Alternative} {Recruitment}
{Hypotheses}}. \emph{Canadian Journal of Fisheries and Aquatic
Sciences}, 38, 678--689.

\leavevmode\vadjust pre{\hypertarget{ref-Walters1978}{}}%
Walters, C.J. \& Hilborn, R. (1978).
\href{https://doi.org/10.1146/annurev.es.09.110178.001105}{Ecological
{Optimization} and {Adaptive} {Management}}. \emph{Annual Review of
Ecology and Systematics}, 9, 157--188.

\leavevmode\vadjust pre{\hypertarget{ref-Wilson2006}{}}%
Wilson, K.A., McBride, M.F., Bode, M. \& Possingham, H.P. (2006).
\href{https://doi.org/10.1038/nature04366}{Prioritizing global
conservation efforts}. \emph{Nature}, 440, 337--340.

\leavevmode\vadjust pre{\hypertarget{ref-Worm2006}{}}%
Worm, B., Barbier, E.B., Beaumont, N., Duffy, J.E., Folke, C., Halpern,
B.S., \emph{et al.} (2006).
\href{https://doi.org/10.1126/science.1132294}{{Impacts of biodiversity
loss on ocean ecosystem services.}} \emph{Science (New York, N.Y.)},
314, 787--90.

\leavevmode\vadjust pre{\hypertarget{ref-Worm2009}{}}%
Worm, B., Hilborn, R., Baum, J.K., Branch, T.A., Collie, J.S., Costello,
C., \emph{et al.} (2009).
\href{https://doi.org/10.1126/science.1173146}{{Rebuilding global
fisheries.}} \emph{Science (New York, N.Y.)}, 325, 578--85.

\leavevmode\vadjust pre{\hypertarget{ref-sony}{}}%
Wurman, P.R., Barrett, S., Kawamoto, K., MacGlashan, J., Subramanian,
K., Walsh, T.J., \emph{et al.} (2022).
\href{https://doi.org/10.1038/s41586-021-04357-7}{Outracing champion
{Gran} {Turismo} drivers with deep reinforcement learning}.
\emph{Nature}, 602, 223--228.

\leavevmode\vadjust pre{\hypertarget{ref-Ye2015}{}}%
Ye, H., Beamish, R.J., Glaser, S.M., Grant, S.C.H., Hsieh, C., Richards,
L.J., \emph{et al.} (2015).
\href{https://doi.org/10.1073/pnas.1417063112}{Equation-free mechanistic
ecosystem forecasting using empirical dynamic modeling}.
\emph{Proceedings of the National Academy of Sciences}, 112,
E1569--E1576.

\end{CSLReferences}

\end{document}
